### 2016-09-15 Code/1_Explore_Data.Rmd
first commit

---
title: "Exploratory Data Analysis -- Supressed Code of Death"
author: "Corinne Riddell"
date: "July 22, 2016"
output: html_document
---

```{r load_libs, echo=F, warning=F, message=F}
library(plyr)
library(dplyr)
library(ggplot2)
library(tidyr)
```

*About the dataset*
The original dataset was collected by Sam using software from the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (known as the SEER*Stat Software) to access the National Vital Statistics System that is maintained by the National Center for Health Statistics.

```{r read_data}
dat <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/deaths-cause-6913.txt", header = F,sep="\t")
names(dat) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
```

```{r recode_variables}
dat <- dat %>%
  mutate(State2 = factor(State, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
                                                             "Colorado", "Connecticut", "Delaware", "Washington DC",
                                                             "Florida", "Georgia", "Hawaii","Idaho", "Illinois", 
                                                             "Indiana","Iowa", "Kansas", "Kentucky", "Louisiana", 
                                                             "Maine", "Maryland", "Massachusetts", "Michigan", 
                                                             "Minnesota", "Mississippi","Missouri", "Montana", 
                                                             "Nebraska", "Nevada", "New Hampshire", "New Jersey",
                                                             "New Mexico", "New York","North Carolina","North Dakota",
                                                             "Ohio", "Oklahoma","Oregon", "Pennsylvania", 
                                                             "Rhode Island", "South Carolina", "South Dakota", 
                                                             "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
                                                             "Washington", "West Virginia", "Wisconsin", "Wyoming")), 
         Age2 = factor(Age, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                         "15-19 years", "20-24 years", "25-29 years", "30-34 years",
                                                         "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
                                                         "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
                                                         "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
         Sex2 = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
         Race2 = factor(Race, levels = c(0, 1), labels = c("White", "Black")),
         COD2 = factor(COD, levels = c(0:5), labels = c("Cardiovascular", "Cancers", "Communicable", 
                                                        "Non-communicable", "Injuries", "All other causes")), 
         Year2 = factor(Year, levels = c(0:44), labels = as.character(1969:2013)),
         RaceSex = interaction(Race2, Sex2),
         StateYearRaceSex = interaction(State2, Year2, Race2, Sex2))
```

Some of the rows have Population==0, and we explore that and record below. These mostly correspond to rows with Unknown age groups that we cannot use in our analysis so I remove them. This leaves 654 rows with Populations of 0.

```{r explore_dataset}
table(dat$Population==0, useNA = "always")
#55734 cells have Populations of 0.

table(dat$Count[dat$Population==0]==0, useNA = "always")
#Of these, all have Counts of 0, except for 26 with counts of NA

table(dat$Age2 == "Unknown", dat$Population == 0, useNA = "always")
#The vast majority of cells with populations of 0 have unknown age groups

#remove this data from the data we analyze:
dat.clean <- dat[dat$Age2 != "Unknown", ]

table(dat.clean$Population == 0, useNA = "always")
#654 cells have Populations of 0 now.

table(dat.clean$Count[dat.clean$Population == 0] == 0, useNA = "always")
summary(dat.clean$Count[dat.clean$Population == 0])
#628 are equal to 0 and 26 are equal to NA

#replace the NA Counts with 0 for the 26 cells with Populations of 0.
dat.clean$Count[is.na(dat.clean$Count) == T & dat.clean$Population == 0] <- 0
summary(dat.clean$Count[dat.clean$Population == 0])

#now the Counts should only equal NA when the true count is between 1 and 9.

summary(dat.clean$Crude_Rate[dat.clean$Population == 0])
table(dat.clean$Crude_Rate[dat.clean$Population==0], useNA = "always")
#whenever the population is 0, the Crude_Rate is equal to NA.
#need to think: should these be equal to 0, or NA? How will it affect the calculations?
```

**Missingness of Count (of deaths)**
If there are between 1 and 9 deaths in the strata, this information is surpressed and coded as missing. Later in this document we investigate trends in supression/missingness to understand how it differs according to state, age, race, and sex.

Since our goal is to compute the difference in life expectancy between Blacks and Whites by strata, we impute death counts at their logical extremes to explore this affects the results.

- Set all missing counts to 1, 5, and 9.
- Set all missing counts to 1 for Whites and 9 for Blacks and vice versa.

```{r fill_in_missing_counts}
dat.clean <- dat.clean %>% mutate(Count_md1 = ifelse(is.na(Count)==T, 1, Count),
                                  Count_md5 = ifelse(is.na(Count)==T, 5, Count),
                                  Count_md9 = ifelse(is.na(Count)==T, 9, Count)
                                  )

dat.clean$Count_b9w1 <- dat.clean$Count
dat.clean$Count_b9w1[is.na(dat.clean$Count) & dat.clean$Race2 == "White"] <- 1
dat.clean$Count_b9w1[is.na(dat.clean$Count) & dat.clean$Race2 == "Black"] <- 9

dat.clean$Count_b1w9 <- dat.clean$Count
dat.clean$Count_b1w9[is.na(dat.clean$Count) & dat.clean$Race2 == "White"] <- 9
dat.clean$Count_b1w9[is.na(dat.clean$Count) & dat.clean$Race2 == "Black"] <- 1

#check recoding
dat.clean$Count_grouped <- cut(dat.clean$Count, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count1_grouped <- cut(dat.clean$Count_md1, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count5_grouped <- cut(dat.clean$Count_md5, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count9_grouped <- cut(dat.clean$Count_md9, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count_b9w1_grouped <- cut(dat.clean$Count_b9w1, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count_b1w9_grouped <- cut(dat.clean$Count_b1w9, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))

table(dat.clean$Count_grouped, dat.clean$Count1_grouped, useNA = "always")
table(dat.clean$Count_grouped, dat.clean$Count5_grouped, useNA = "always")
table(dat.clean$Count_grouped, dat.clean$Count9_grouped, useNA = "always")

table(dat.clean$Count_grouped, dat.clean$Count_b9w1_grouped, dat.clean$Race2, useNA = "always")
table(dat.clean$Count_grouped, dat.clean$Count_b1w9_grouped, dat.clean$Race2, useNA = "always")

dat.clean <- dat.clean[ , !names(dat.clean) %in% c("Count_grouped", "Count1_grouped", "Count5_grouped", "Count9_grouped", 
                                                   "Count_b9w1_grouped", "Count_b1w9_grouped")]
```

To calculate life expectancy, we have to aggregate the data across the causes of death.

```{r aggregate_over_COD}
dat2 <- dat.clean %>% group_by(State2, Year2, RaceSex, Age2) %>%
  arrange(State2, RaceSex, Year2, Age2) %>%
  summarise(total_deaths1 = sum(Count_md1), total_deaths5 = sum(Count_md5),
            total_deaths9 = sum(Count_md9), 
            total_deaths_b9 = sum(Count_b9w1), total_deaths_w9 = sum(Count_b1w9),
            Population = first(Population), Race2 = first(Race2), Sex2 = first(Sex2), 
            StateYearRaceSex = first(StateYearRaceSex))
```

We implemented the calculations performed in the technical supplement found here [http://www.lho.org.uk/Download/Public/7656/1/tech_supp_3.pdf], also saved in our dropbox.

Let's make sure we can do the calculation all the way through for a few of the groupings, including a grouping with a stratum with a zero population count.

```{r calc_life_expectancy1}

dat2$numYrsInt <- 5
dat2$numYrsInt[dat2$Age2 == "<1 year"] <- 1
dat2$numYrsInt[dat2$Age2 == "1-4 years"] <- 4

dat2 <- dat2 %>% mutate( 
            Death_Rate1 = ifelse(Population!=0, total_deaths1/Population, NA),
            Death_Rate5 = ifelse(Population!=0, total_deaths5/Population, NA),
            Death_Rate9 = ifelse(Population!=0, total_deaths9/Population, NA),
            Death_Rateb9 = ifelse(Population!=0, total_deaths_b9/Population, NA),
            Death_Ratew9 = ifelse(Population!=0, total_deaths_w9/Population, NA),
            Ave_Lived_by_Died = ifelse(Age2=="<1 year", 0.1, 0.5),
            Prob_Dying1 = ifelse(Population!=0, 
                                 numYrsInt*Death_Rate1/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Rate1), 
                                 NA),
            Prob_Dying5 = ifelse(Population!=0, 
                                 numYrsInt*Death_Rate5/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Rate5), 
                                 NA),
            Prob_Dying9 = ifelse(Population!=0, 
                                 numYrsInt*Death_Rate9/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Rate9), 
                                 NA),
            Prob_Dyingb9 = ifelse(Population!=0, 
                                 numYrsInt*Death_Rateb9/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Rateb9), 
                                 NA),
            Prob_Dyingw9 = ifelse(Population!=0, 
                                 numYrsInt*Death_Ratew9/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Ratew9), 
                                 NA),
            Prob_Surv1 = ifelse(Population!=0, 1 - Prob_Dying1, NA),
            Prob_Surv5 = ifelse(Population!=0, 1 - Prob_Dying5, NA),
            Prob_Surv9 = ifelse(Population!=0, 1 - Prob_Dying9, NA),            
            Prob_Survb9 = ifelse(Population!=0, 1 - Prob_Dyingb9, NA),
            Prob_Survw9 = ifelse(Population!=0, 1 - Prob_Dyingw9, NA),
            alive_at_start_int1 = ifelse(Age2 == "<1 year", 1000000, NA),
            alive_at_start_int5 = ifelse(Age2 == "<1 year", 1000000, NA),
            alive_at_start_int9 = ifelse(Age2 == "<1 year", 1000000, NA),
            alive_at_start_intb9 = ifelse(Age2 == "<1 year", 1000000, NA),
            alive_at_start_intw9 = ifelse(Age2 == "<1 year", 1000000, NA),
            deaths_during_int1 = ifelse(Age2 == "<1 year", alive_at_start_int1*Prob_Dying1, NA),
            deaths_during_int5 = ifelse(Age2 == "<1 year", alive_at_start_int5*Prob_Dying5, NA),
            deaths_during_int9 = ifelse(Age2 == "<1 year", alive_at_start_int9*Prob_Dying9, NA),
            deaths_during_intb9 = ifelse(Age2 == "<1 year", alive_at_start_intb9*Prob_Dyingb9, NA),
            deaths_during_intw9 = ifelse(Age2 == "<1 year", alive_at_start_intw9*Prob_Dyingw9, NA))
```

```{r first_five}
levels(dat2$StateYearRaceSex)[1:5]
summary(dat2$Population[dat2$StateYearRaceSex %in% levels(dat2$StateYearRaceSex)[1:5]])

timer1 <- system.time (
  for(x in levels(dat2$StateYearRaceSex)[1:5]) {
      for(row in 2:length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) { 
        dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row - 1]

          if (row != length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) {
              dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying1[dat2$StateYearRaceSex == x][row]
              dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying5[dat2$StateYearRaceSex == x][row]
              dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying9[dat2$StateYearRaceSex == x][row]
              dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingb9[dat2$StateYearRaceSex == x][row]
              dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingw9[dat2$StateYearRaceSex == x][row]
          }
          else { dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] #everyone dies in the 85+ grouping
                 dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]
                 dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]
                 dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]
                 dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]

          } 
        }

  }
)

timer1 # 25 seconds for 5 -- guess 13 hrs for all the strata
```

```{r calc_pop0}
View(dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[2789], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1")])
timer2 <- system.time (
  for(x in levels(dat2$StateYearRaceSex)[2789]) {
      for(row in 2:length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) { 
        dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row - 1]

          if (row != length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) {
              dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying1[dat2$StateYearRaceSex == x][row]
              dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying5[dat2$StateYearRaceSex == x][row]
              dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying9[dat2$StateYearRaceSex == x][row]
              dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingb9[dat2$StateYearRaceSex == x][row]
              dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingw9[dat2$StateYearRaceSex == x][row]
          }
          else { dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] #everyone dies in the 85+ grouping
                 dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]
                 dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]
                 dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]
                 dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]

          } 
        }

  }
)

timer2
```

This takes `r time1[3]` seconds to run.

```{r rest_of_life_expectancy_calc}
dat2$num_PY_int1 <- (dat2$alive_at_start_int1 - dat2$deaths_during_int1)*dat2$numYrsInt + (dat2$deaths_during_int1*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
dat2$num_PY_int5 <- (dat2$alive_at_start_int5 - dat2$deaths_during_int5)*dat2$numYrsInt + (dat2$deaths_during_int5*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
dat2$num_PY_int9 <- (dat2$alive_at_start_int9 - dat2$deaths_during_int9)*dat2$numYrsInt + (dat2$deaths_during_int9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
dat2$num_PY_intb9 <- (dat2$alive_at_start_intb9 - dat2$deaths_during_intb9)*dat2$numYrsInt + (dat2$deaths_during_intb9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
dat2$num_PY_intw9 <- (dat2$alive_at_start_intw9 - dat2$deaths_during_intw9)*dat2$numYrsInt + (dat2$deaths_during_intw9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)

dat2$num_PY_int1[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_int1[dat2$Age2 == "85+ years"])/dat2$Death_Rate1[dat2$Age2 == "85+ years"]
dat2$num_PY_int5[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_int5[dat2$Age2 == "85+ years"])/dat2$Death_Rate5[dat2$Age2 == "85+ years"]
dat2$num_PY_int9[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_int9[dat2$Age2 == "85+ years"])/dat2$Death_Rate9[dat2$Age2 == "85+ years"]
dat2$num_PY_intb9[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_intb9[dat2$Age2 == "85+ years"])/dat2$Death_Rateb9[dat2$Age2 == "85+ years"]
dat2$num_PY_intw9[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_intw9[dat2$Age2 == "85+ years"])/dat2$Death_Ratew9[dat2$Age2 == "85+ years"]

dat2 <- dat2 %>% group_by(StateYearRaceSex) %>% arrange(desc(Age2)) %>%
                 mutate(tot_PY_after1 = cumsum(num_PY_int1), tot_PY_after5 = cumsum(num_PY_int5), tot_PY_after9 = cumsum(num_PY_int9),
                        tot_PY_afterb9 = cumsum(num_PY_intb9),tot_PY_afterw9 = cumsum(num_PY_intw9),
                        life_exp1 = tot_PY_after1/alive_at_start_int1, life_exp5 = tot_PY_after5/alive_at_start_int5, 
                        life_exp9 = tot_PY_after9/alive_at_start_int9, 
                        life_expb9 = tot_PY_afterb9/alive_at_start_intb9, life_expw9 = tot_PY_afterw9/alive_at_start_intw9 
                )

dat2 <- dat2 %>% group_by(StateYearRaceSex) %>% arrange(Age2)
```

I compared the following with my calculations in the excel spreadsheet to check the calculations -- all good.
```{r}
dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[1], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1", 
                                                                       "alive_at_start_int1", "deaths_during_int1", "num_PY_int1", "tot_PY_after1", "life_exp1")]

dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[4], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1", 
                                                                       "alive_at_start_int1", "deaths_during_int1", "num_PY_int1", "tot_PY_after1", "life_exp1")]

dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[2789], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1", 
                                                                       "alive_at_start_int1", "deaths_during_int1", "num_PY_int1", "tot_PY_after1", "life_exp1")]
```

Notes on Figure 1 (below):

- Figure 1 is sorted by size of the black population in the state in 2013. This is so you can see how missingness is a function of population size. We expect that states with less black people will have more missing data and more counts of exactly zero. Thus, states with the biggest populations (of blacks and whites) will "hold constant" the effect of population size on the amount of missing data.

- The first panels you see in Figure 1 (below) are for the states with the smallest black populations. For whites, the amount of missing data increases in age until about 40 years old, and then decreases. For blacks, the amount of missing data is very low across age. This seemed strange to me at first, but then I remembered that the data is not missing if the count is equal to 0, which will be the case for many strata in these states with the lowest black population.

- As you read left to right (as the black population size increases by state), the trends in missing data across age look more similar by race, although blacks appear to have more missing data in individuals < 30 years old for the states with the most black people.

**Figure 1: Proportion missing by State (panel) and Age group (x-axis), ordered by increasing size of black population in 2013**
```{r bystate, fig.width = 12, fig.height = 37, echo=F}
bystate <- 
  dat %>%
  group_by(State2, Race2, Age2) %>%
  summarise(prop.missing = mean(is.na(Count)))

#recode a new count variable that "counts the missing data" so that these rows are used when calculating the % of data == 0.
dat$Count2 <- dat$Count
dat$Count2[is.na(dat$Count) == T] <- 1

bystate2 <- 
  dat %>%
  group_by(State2, Race2, Age2) %>%
  summarise(prop.zero = mean(Count2 == 0))

bystate <- merge(bystate, bystate2, by=c("State2", "Race2", "Age2"))

#calculate the size of the black population in 2013 by State
bystate.bpop <- 
  dat %>%
  filter(COD2 == "Cancers" & Year2 == 2013 & Race2 == "Black") %>% #to stop repeats of the population size
  group_by(State2) %>%
  summarize(pop_by_race = sum(as.numeric(Population)))

bystate <- merge(bystate, bystate.bpop, by = "State2", all.x = T)
bystate$State3 <- reorder(bystate$State2, bystate$pop_by_race)

ggplot(bystate, aes(x = Age2, y = prop.missing)) + 
  geom_point(aes(col=Race2)) + facet_wrap(~ State3, ncol = 4) +
  theme_bw(base_size = 11) +
  theme(legend.position="top") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Missing (%)") +
  xlab("Age group")
```

Notes on Figure 2 (below):

- Figure 2 examines the proportion of counts equal to zero. The purpose of this examination is to visually confirm that there are higher proportions of zero counts in the states with the least amount of blacks. This is very clear when viewing the figure.

- Question - any explantion for why the portion of zeroes stays relatively higher in blacks vs. whites in the states with the most blacks (within the younger age groups)?

**Figure 2: Proportion zero by State (panel) and Age group (x-axis), ordered by increasing size of black population in 2013**

```{r zero_by_state, fig.width = 12, fig.height = 37, echo=F}
ggplot(bystate, aes(x = Age2, y = prop.zero)) + 
  geom_point(aes(col=Race2)) + facet_wrap(~ State3, ncol = 4) +
  theme_bw(base_size = 11) +
  theme(legend.position="top") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Zero (%)") +
  xlab("Age group") 
```

###Other plots (less interesting/informative)

**How is missingness related to age, race, and sex?**

```{r missing_data, fig.width = 10, fig.height = 5, echo=F}
#table(is.na(dat$Count))

table1.whitemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "White.Male"]), 
                                           dat$Age2[dat$RaceSex == "White.Male"]),2)*100,2)
table1.blackmale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "Black.Male"]), 
                                           dat$Age2[dat$RaceSex == "Black.Male"]),2)*100,2)
table1.whitefemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "White.Female"]), 
                                           dat$Age2[dat$RaceSex == "White.Female"]),2)*100,2)
table1.blackfemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "Black.Female"]), 
                                           dat$Age2[dat$RaceSex == "Black.Female"]),2)*100,2)

miss1 <- data.frame("white.male" = table1.whitemale[2, ], 
                    "black.male"= table1.blackmale[2, ],
                    "white.female" = table1.whitefemale[2, ], 
                    "black.female"= table1.blackfemale[2, ],
                    "age" = names(table1.whitemale[2, ]))
miss1$age <- factor(miss1$age, levels(miss1$age)[c(1, 2, 11, 3:10, 12:20)])

miss1 <- gather(miss1, racesex, missingness, c(white.male, black.male, white.female, black.female))

ggplot(miss1, aes(x = age, y = missingness)) + geom_point() + geom_point(aes(col=racesex)) +
  theme_bw(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Missing (%)") +
  xlab("Age group")
```

**Let's look at the proportion with 0's**

```{r zero_data, fig.width = 10, fig.height = 5, echo=F}
table2.whitemale <- round(prop.table(table(dat$Count[dat$RaceSex == "White.Male"]==0, 
                                       dat$Age2[dat$RaceSex == "White.Male"]),2)*100,2)
table2.blackmale <- round(prop.table(table(dat$Count[dat$RaceSex == "Black.Male"]==0, 
                                       dat$Age2[dat$RaceSex == "Black.Male"]),2)*100,2)
table2.whitefemale <- round(prop.table(table(dat$Count[dat$RaceSex == "White.Female"]==0, 
                                       dat$Age2[dat$RaceSex == "White.Female"]),2)*100,2)
table2.blackfemale <- round(prop.table(table(dat$Count[dat$RaceSex == "Black.Female"]==0, 
                                       dat$Age2[dat$RaceSex == "Black.Female"]),2)*100,2)

miss2 <- data.frame("whitemale" = table2.whitemale[2, ], 
                    "blackmale"= table2.blackmale[2, ],
                    "whitefemale" = table2.whitefemale[2, ], 
                    "blackfemale"= table2.blackfemale[2, ],
                    "age" = names(table2.blackmale[2, ]))

miss2$age <- factor(miss2$age, levels(miss2$age)[c(1, 2, 11, 3:10, 12:20)])

miss2 <- gather(miss2, racesex, zeroes, c(whitemale, blackmale, whitefemale, blackfemale))

ggplot(miss2, aes(x = age, y = zeroes)) + geom_point() + geom_point(aes(col=racesex)) +
  theme_bw(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Zero (%)") +
  xlab("Age group")
```

**Missing or zero (which is equivalent to < 10)**

```{r miss_or_zero_data, fig.width = 10, fig.height = 5, echo=F}
dat$miss_or_zero <- F
dat$miss_or_zero[dat$Count == 0 | is.na(dat$Count) == T] <- T

table3.whitemale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "White.Male"]==T, 
                                      dat$Age2[dat$RaceSex == "White.Male"]),2)*100,2)
table3.blackmale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "Black.Male"]==T, 
                                       dat$Age2[dat$RaceSex == "Black.Male"]),2)*100,2)
table3.whitefemale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "White.Female"]==T, 
                                       dat$Age2[dat$RaceSex == "White.Female"]),2)*100,2)
table3.blackfemale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "Black.Female"]==T, 
                                       dat$Age2[dat$RaceSex == "Black.Female"]),2)*100,2)

miss3 <- data.frame("whitemale" = table3.whitemale[2, ], 
                    "blackmale"= table3.blackmale[2, ],
                    "whitefemale" = table3.whitefemale[2, ], 
                    "blackfemale"= table3.blackfemale[2, ],
                    "age" = names(table3.blackmale[2, ]))

miss3$age <- factor(miss3$age, levels(miss3$age)[c(1, 2, 11, 3:10, 12:20)])

miss3 <- gather(miss3, racesex, miss_or_zero, c(whitemale, blackmale, whitefemale, blackfemale))

ggplot(miss3, aes(x = age, y = miss_or_zero)) + geom_point() + geom_point(aes(col=racesex)) +
  theme_bw(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Missing or Zero (%)") +
  xlab("Age group")
```

```{r explore_data2, fig.width = 10, fig.height = 5, echo=F, eval=F}

summary(dat$Count)
ggplot(subset(dat, Count>0), aes(x = Count)) + geom_histogram(binwidth=100)

ggplot(subset(dat, State2=="Louisiana"), aes(x = Population, y = Count)) + 
  geom_point(aes(col=RaceSex), alpha=0.2) +
  facet_wrap(~COD2) + theme_bw()

```

```{r save_workspace}
save.image(file = "/Users/corinneriddell/Dropbox/SRJ/Data/corinne/Sep5_workspace_bw_gap.Rdata")
```

### 2016-09-15 Code/1_Explore_Data_v2.Rmd
first commit

--- 2016-09-15 Code/1_Explore_Data.Rmd
@@
 ---
-title: "Exploratory Data Analysis -- Supressed Code of Death"
+title: "Calculating life expectancy and exploring missingness"
 author: "Corinne Riddell"
@@
 #need to think: should these be equal to 0, or NA? How will it affect the calculations?
+
+#States with at least one strata of 0
+na.pops <- dat.clean %>% select(State2, Population) %>% filter(Population == 0)
+unique(na.pops$State2)
+
+pops.le.20 <- dat.clean %>% select(State2, Population) %>% filter(Population <= 20)
+unique(pops.le.20$State2)
 ```
@@
 
-```{r calc_pop0}
+```{r calc_pop0, eval=F, echo=F}
 View(dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[2789], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
@@
 
-This takes `r time1[3]` seconds to run.
+```{r first_thousand}
+timer3 <- system.time (
+  for(x in levels(dat2$StateYearRaceSex)[6:1000]) {
+      for(row in 2:length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) { 
+        dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row - 1]
+        dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row - 1]
+        dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row - 1]
+        dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row - 1]
+        dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row - 1]
+
+          if (row != length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) {
+              dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying1[dat2$StateYearRaceSex == x][row]
+              dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying5[dat2$StateYearRaceSex == x][row]
+              dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying9[dat2$StateYearRaceSex == x][row]
+              dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingb9[dat2$StateYearRaceSex == x][row]
+              dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingw9[dat2$StateYearRaceSex == x][row]
+          }
+          else { dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] #everyone dies in the 85+ grouping
+                 dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]
+                 dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]
+                 dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]
+                 dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]
+
+          } 
+        }
+
+  }
+)
+
+timer3
+```
 
@@
 
+```{r save_workspace1}
+save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
+```
+
+```{r rest}
+timer4 <- system.time (
+  for(x in levels(dat2$StateYearRaceSex)[1001:9180]) {
+      for(row in 2:length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) { 
+        dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row - 1]
+        dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row - 1]
+        dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row - 1]
+        dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row - 1]
+        dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row - 1]
+
+          if (row != length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) {
+              dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying1[dat2$StateYearRaceSex == x][row]
+              dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying5[dat2$StateYearRaceSex == x][row]
+              dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying9[dat2$StateYearRaceSex == x][row]
+              dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingb9[dat2$StateYearRaceSex == x][row]
+              dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingw9[dat2$StateYearRaceSex == x][row]
+          }
+          else { dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] #everyone dies in the 85+ grouping
+                 dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]
+                 dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]
+                 dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]
+                 dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]
+
+          } 
+        }
+
+  }
+)
+
+timer4
+```
+
+```{r rest_of_life_expectancy_calc2}
+dat2$num_PY_int1 <- (dat2$alive_at_start_int1 - dat2$deaths_during_int1)*dat2$numYrsInt + (dat2$deaths_during_int1*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
+dat2$num_PY_int5 <- (dat2$alive_at_start_int5 - dat2$deaths_during_int5)*dat2$numYrsInt + (dat2$deaths_during_int5*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
+dat2$num_PY_int9 <- (dat2$alive_at_start_int9 - dat2$deaths_during_int9)*dat2$numYrsInt + (dat2$deaths_during_int9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
+dat2$num_PY_intb9 <- (dat2$alive_at_start_intb9 - dat2$deaths_during_intb9)*dat2$numYrsInt + (dat2$deaths_during_intb9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
+dat2$num_PY_intw9 <- (dat2$alive_at_start_intw9 - dat2$deaths_during_intw9)*dat2$numYrsInt + (dat2$deaths_during_intw9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
+
+dat2$num_PY_int1[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_int1[dat2$Age2 == "85+ years"])/dat2$Death_Rate1[dat2$Age2 == "85+ years"]
+dat2$num_PY_int5[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_int5[dat2$Age2 == "85+ years"])/dat2$Death_Rate5[dat2$Age2 == "85+ years"]
+dat2$num_PY_int9[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_int9[dat2$Age2 == "85+ years"])/dat2$Death_Rate9[dat2$Age2 == "85+ years"]
+dat2$num_PY_intb9[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_intb9[dat2$Age2 == "85+ years"])/dat2$Death_Rateb9[dat2$Age2 == "85+ years"]
+dat2$num_PY_intw9[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_intw9[dat2$Age2 == "85+ years"])/dat2$Death_Ratew9[dat2$Age2 == "85+ years"]
+
+dat2 <- dat2 %>% group_by(StateYearRaceSex) %>% arrange(desc(Age2)) %>%
+                 mutate(tot_PY_after1 = cumsum(num_PY_int1), tot_PY_after5 = cumsum(num_PY_int5), tot_PY_after9 = cumsum(num_PY_int9),
+                        tot_PY_afterb9 = cumsum(num_PY_intb9),tot_PY_afterw9 = cumsum(num_PY_intw9),
+                        life_exp1 = tot_PY_after1/alive_at_start_int1, life_exp5 = tot_PY_after5/alive_at_start_int5, 
+                        life_exp9 = tot_PY_after9/alive_at_start_int9, 
+                        life_expb9 = tot_PY_afterb9/alive_at_start_intb9, life_expw9 = tot_PY_afterw9/alive_at_start_intw9 
+                )
+
+dat2 <- dat2 %>% group_by(StateYearRaceSex) %>% arrange(Age2)
+```
+
+```{r save_workspace2}
+save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
+```
+
 I compared the following with my calculations in the excel spreadsheet to check the calculations -- all good.
@@
 
+###Initial graphs on missing data
 Notes on Figure 1 (below):
@@
 
+#calculate the size of the black population in 1969 by State
+bystate.bpop.1969 <- 
+  dat %>%
+  filter(COD2 == "Cancers" & Year2 == 1969 & Race2 == "Black") %>% #to stop repeats of the population size
+  group_by(State2) %>%
+  summarize(pop_by_race_1969 = sum(as.numeric(Population)))
+
 bystate <- merge(bystate, bystate.bpop, by = "State2", all.x = T)
+bystate <- merge(bystate, bystate.bpop.1969, by = "State2", all.x = T)
 bystate$State3 <- reorder(bystate$State2, bystate$pop_by_race)
@@
   xlab("Age group")
+
+ggplot(subset(bystate, Age2 == "<1 year"), aes(pop_by_race_1969, pop_by_race)) + geom_point() + geom_abline(intercept = 0, slope = 1)
+
+unique(bystate$State2[bystate$Age2 == "<1 year" & bystate$pop_by_race_1969 < 10000])
 ```
@@
 
-```{r save_workspace}
-save.image(file = "/Users/corinneriddell/Dropbox/SRJ/Data/corinne/Sep5_workspace_bw_gap.Rdata")
+```{r save_workspace3}
+save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
 ```

### 2016-09-15 Code/2_Examine_LifeExpectancy.Rmd
first commit

---
title: "Examining differences in life expectancy across sensitivity analyses"
author: "Corinne Riddell"
date: "July 22, 2016"
output: html_document
---

```{r load_libs, echo=F, warning=F, message=F}
library(plyr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
```

```{r load_workspace}
load(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
```

```{r}

dat3 <- 
dat2 %>% group_by(StateYearRaceSex) %>%
  arrange(Age2) %>%  
  summarise(State2 = first(State2), Year2 = first(Year2), Race2 = first(Race2), Sex2 = first(Sex2),
            Pop_across_age = sum(Population, na.rm = "always"), Pop_has_NA = sum(Population), 
            LE_birth1 = first(life_exp1), LE_birth5 = first(life_exp5), LE_birth9 = first(life_exp9), 
            LE_birthb9 = first(life_expb9), LE_birthw9 = first(life_expw9))

dat3$grouping <- interaction(dat3$State2, dat3$Race2, dat3$Sex2)
dat3$Year <- as.numeric(as.character(dat3$Year2))

#life expectancy over time under two sensitivities for imputing the NA death counts
ggplot(subset(dat3, State2 == "Alabama"), aes(y = LE_birth1, x = Year)) + 
  geom_line(aes(col = grouping, lty = "Imputed 1")) +
  geom_line(aes(y = LE_birth9, col = grouping, lty = "Imputed 9")) +
 # geom_line(aes(y = LE_birthb9, col = grouping, lty = "Imputed w=1, b=9")) +  
  ylab("Life Expectancy at Birth") + scale_x_continuous(breaks = seq(1969, 2013, by = 4)) +
  scale_y_continuous(breaks = seq(55, 90, by = 2)) +
  facet_wrap(~ Sex2)

#graph of population size over time 
#important because gives us an idea of how stable/precise the life expectancy calculation are 
ggplot(subset(dat3, State2 == "Alabama"), aes(y = Pop_across_age, x = Year)) + geom_line(aes(col = grouping)) +
  ylab("Population Size") + scale_y_continuous(label = scales::comma)

ggplot(subset(dat3, State2 == "Massachusetts"), aes(y = LE_birth1, x = Year)) + 
  geom_line(aes(col = grouping, lty = "Imputed 1")) +
  geom_line(aes(y = LE_birth9, col = grouping, lty = "Imputed 9")) +
  geom_line(aes(y = LE_birthb9, col = grouping, lty = "Imputed w=1, b=9")) +  
  ylab("Life Expectancy at Birth") + scale_x_continuous(breaks = seq(1969, 2013, by = 4)) +
  scale_y_continuous(breaks = seq(55, 90, by = 2)) +
  facet_wrap(~ Sex2)

ggplot(subset(dat3, State2 == "Massachusetts"), aes(y = Pop_across_age, x = Year)) + geom_line(aes(col = grouping)) +
  ylab("Population Size") + scale_y_continuous(label = scales::comma)

ggplot(subset(dat3, State2 == "New York"), aes(y = LE_birth1, x = Year)) + 
  geom_line(aes(col = grouping, lty = "Imputed 1")) +
  geom_line(aes(y = LE_birth9, col = grouping, lty = "Imputed 9")) +
  geom_line(aes(y = LE_birthb9, col = grouping, lty = "Imputed w=1, b=9")) +  
  ylab("Life Expectancy at Birth") + scale_x_continuous(breaks = seq(1969, 2013, by = 4)) +
  scale_y_continuous(breaks = seq(55, 90, by = 2)) +
  facet_wrap(~ Sex2)

ggplot(subset(dat3, State2 == "New York"), aes(y = Pop_across_age, x = Year)) + geom_line(aes(col = grouping)) +
  ylab("Population Size") + scale_y_continuous(label = scales::comma)

ggplot(subset(dat3, State2 == "Mississippi"), aes(y = LE_birth1, x = Year)) + 
  geom_line(aes(col = grouping, lty = "Imputed 1")) +
  geom_line(aes(y = LE_birth9, col = grouping, lty = "Imputed 9")) +
  geom_line(aes(y = LE_birthb9, col = grouping, lty = "Imputed w=1, b=9")) +  
  ylab("Life Expectancy at Birth") + scale_x_continuous(breaks = seq(1969, 2013, by = 4)) +
  scale_y_continuous(breaks = seq(55, 90, by = 2)) +
  facet_wrap(~ Sex2)

ggplot(subset(dat3, State2 == "Mississippi"), aes(y = Pop_across_age, x = Year)) + geom_line(aes(col = grouping)) +
  ylab("Population Size") + scale_y_continuous(label = scales::comma)
```

```{r}
dat4_Black <- dat3[dat3$Race2 == "Black", ]
dat4_Black <- dat4_Black %>% 
  select(State2, Year, Sex2, Pop_across_age, Pop_has_NA, LE_birth1, LE_birth5, LE_birth9) %>%
  rename(Pop_black = Pop_across_age, Pop_NA_black = Pop_has_NA, LE1_black = LE_birth1, LE5_black = LE_birth5, LE9_black = LE_birth9)

dat4_White <- dat3[dat3$Race2 == "White", ]
dat4_White <- dat4_White %>% 
  select(State2, Year, Sex2, Pop_across_age, Pop_has_NA, LE_birth1, LE_birth5, LE_birth9) %>%
    rename(Pop_white = Pop_across_age, Pop_NA_white = Pop_has_NA, LE1_white = LE_birth1, LE5_white = LE_birth5, LE9_white = LE_birth9)

dat4 <- merge(dat4_Black, dat4_White, by = c("State2", "Year", "Sex2"))
dim(dat4)
View(dat4)

dat4 <- dat4 %>% mutate(gap_w1b1 = LE1_white - LE1_black, 
                gap_w5b5 = LE5_white - LE5_black,
                gap_w9b9 = LE9_white - LE9_black,
                gap_w1b9 = LE1_white - LE9_black,
                gap_w9b1 = LE9_white - LE1_black)

summary(dat4$Pop_white)
summary(dat4$Pop_black)

#Life expectancy gap, under the different sensitivity analyses
ggplot(subset(dat4, State2 == "Alabama"), aes(x = Year, y = gap_w1b1)) + 
  geom_line(aes(y = gap_w1b9, col = "Imputed w=1, b=9")) +   
  geom_line(aes(col = "Imputed 1")) + 
  geom_line(aes(y = gap_w5b5, col = "Imputed 5")) +
  geom_line(aes(y = gap_w9b9, col = "Imputed 9")) +
  geom_line(aes(y = gap_w9b1, col = "Imputed w=9, b=1")) +
  facet_wrap( ~ Sex2) +
  ylab("Life Expectancy Gap") + 
  scale_x_continuous(breaks = seq(1969, 2013, by = 4)) +
  scale_y_continuous(breaks = seq(-1, 11, by = 1))
```

### 2016-09-15 Code/3_Prelim_Smoothing.R

@@ -1,0 +1,82 @@
+
+# Preamble
+
+require(rjags)
+require(R2jags) 
+require(devtools)
+require(jagstools)
+
+load('/Users/kathryn/Dropbox/BlackWhiteGap/Data/alabama_only.Rdata')
+head(dat.clean.alabama)
+
+selected_vars = c('Age2', 'Sex2', 'Race2', 'COD2', 'Year2', 'Count', 'Population')
+ds = dat.clean.alabama[ dat.clean.alabama$COD2=='Injuries', selected_vars] 
+head(ds) ; length(ds$Age2) 
+
+ds1 = ds[(ds$Race2=='Black' & ds$Age2=='60-64 years'), ] 
+nyears = length(ds1$Age2)
+ds_jags = list( deaths=ds1$Count, pop = ds1$Population, nyears = nyears) 
+
+
+# Exploring the data a bit
+summary(ds1$Count/ds1$Population)
+
+# More reasonable prior?  
+
+# Just smoothing by years, for one age group
+
+ds_jags2 = ds_jags1 = ds_jags
+ds_jags1$pop =  ds_jags$pop/100
+
+
+ds_jags$pop = seq(100,500,by=10)
+exp(ds_jags$pop)
+ds_jags$deaths = seq(10,50,by=1) 
+ds_jags$nyears = length(ds_jags$deaths)
+
+ds_jags2 = ds_jags
+ds_jags2$pop = ds_jags$pop/5
+
+
+model = function() {  
+  for(year in 1:nyears) {
+    deaths[year] ~ dpois(mu[year]) 
+    log(mu[year]) <- (pop[year] + b[year])
+    mort_rate[year] <- exp(b[year])
+    
+  }
+  
+  for(year in 2:nyears) {
+    b[year] ~ dnorm(b[year-1], tau) 
+  }
+  
+  b[1] ~ dnorm(-4, 0.1) 
+  beta ~ dnorm(0, 0.1) 
+  tau ~ dgamma(0.01, 0.01)
+}  
+
+params = c('mu', 'b', 'tau', 'mort_rate')
+
+nchains = 2
+inits = function() list('tau' = runif(1)) 
+
+# run the model 
+jags_model = jags(data=ds_jags, param=params, n.chains=nchains, inits=inits,
+                  n.iter=10000, n.burnin=2000, model.file=model) 
+
+jags_model2 = jags(data=ds_jags2, param=params, n.chains=nchains, inits=inits,
[…]

### 2016-09-15 Code/3b_Stdz_pops.R

@@ -1,0 +1,86 @@
+
+require(rjags)
+require(R2jags) 
+require(devtools)
+require(jagstools)
+
+# Simulate data 
+set.seed(1001)
+
+nyears = 100
+ds_jags1 = data.frame(pop = rpois(nyears, lambda=200))
+
+ds_jags1$deaths = rpois(nyears, lambda=5)
+
+ds_jags2 = ds_jags1
+ds_jags2$pop = ds_jags1$pop/10
+
+summary(ds_jags1$deaths/ds_jags1$pop) ; log(mean(ds_jags1$deaths/ds_jags1$pop))
+summary(ds_jags2$deaths/ds_jags2$pop) ; log(mean(ds_jags2$deaths/ds_jags2$pop))
+
+ds_jags1 = list(pop=ds_jags1$pop, deaths=ds_jags1$deaths, nyears=nyears)
+ds_jags2 = list(pop=ds_jags2$pop, deaths=ds_jags2$deaths, nyears=nyears)
+
+model1 = function() {  
+  for(year in 1:nyears) {
+    deaths[year] ~ dpois(mu[year]) 
+    log(mu[year]) <- (pop[year] + b[year])
+    mort_rate[year] <- exp(b[year])
+    
+  }
+  
+  for(year in 2:nyears) {
+    b[year] ~ dnorm(b[year-1], tau) 
+  }
+  
+  b[1] ~ dnorm(-3.75, 0.1) 
+  tau ~ dgamma(0.01, 0.01)
+}  
+
+
+model2 = function() {  
+  for(year in 1:nyears) {
+    deaths[year] ~ dpois(mu[year]) 
+    log(mu[year]) <- (pop[year] + b[year])
+    mort_rate[year] <- exp(b[year]*10)
+  }
+  
+  for(year in 2:nyears) {
+    b[year] ~ dnorm(b[year-1], tau) 
+  }
+  
+  b[1] ~ dnorm(-1.45, 0.1) 
+  tau ~ dgamma(0.01, 0.01)
+}  
+
+params = c('mu', 'b', 'tau', 'mort_rate')
+nchains = 2
+inits = function() list('tau' = runif(1)) 
+
+jags_model1 = jags(data=ds_jags1, param=params, n.chains=nchains, inits=inits,
+                  n.iter=10000, n.burnin=2000, model.file=model1) 
+
+jags_model2 = jags(data=ds_jags2, param=params, n.chains=nchains, inits=inits,
[…]

### 2016-09-15 Code/4_smoothing_models.R

@@ -1,0 +1,25 @@
+
+
+
+source('/Users/kathryn/Dropbox/BlackWhiteGap/Code/4_smoothing_models_src.R') 
+
+
+data = subset_data(race='Black', sex='Female', cod='Injuries') 
+head(data) 
+
+age_cats = unique(data$Age) 
+
+ds = data[data$Age2==age_cats[3], ] 
+
+m = run_jags_model(ds, pop_div=100) ; m 
+
+smoothed_deaths = jagsresults(x=m, params=c('mu'))  
+mort_rate = jagsresults(x=m, params=c('b'))
+
+
+r = data.frame(deaths=ds$Count, smoothed_deaths=round(smoothed_deaths[,'50%'],0)) 
+
+r
+
+
+

### 2016-09-15 Code/4_smoothing_models_fun.R

@@ -1,0 +1,155 @@
+
+
+
+# Subset dataset 
+
+subset_data = function(race='Black', sex='Female', cod='Injuries') {
+
+  selected_vars = c('Age', 'Sex2', 'Race2', 'COD2', 'Year', 'Count', 'Population')
+
+  ds = dat.clean.alabama[(dat.clean.alabama$Race2==race & 
+                         dat.clean.alabama$Sex2==sex & 
+                          dat.clean.alabama$COD2==cod), selected_vars]
+  return(ds) 
+}
+
+
+# Run smoothing model v1: smoothing by year 
+
+jags_smooth_year = function(ds, pop_div=100) { 
+
+  df = data.frame(deaths=ds$Count, pop = ds$Population/pop_div, age=(ds$Age+1), year=(ds$Year+1)) 
+  
+  # Need wide format for jags   
+  
+  nagecats = length(unique(ds$Age))  
+  nyears = length(unique(ds$Year))  
+  jags_deaths = as.data.frame(matrix(NA, nrow=nyears, ncol=nagecats)) 
+  jags_pop = as.data.frame(matrix(NA, nrow=nyears, ncol=nagecats)) 
+  
+  for(year in 1:nyears) {
+    for(agecat in 1:nagecats) {
+      jags_deaths[year, agecat] = df$deaths[df$age==agecat& df$year==year]
+      jags_pop[year, agecat] = df$pop[df$age==agecat& df$year==year]
+    }
+  }
+  
+  colnames(jags_deaths) = colnames(jags_pop) = unique(df$age)
+  
+  ds_jags = list(deaths = jags_deaths, 
+                 pop = jags_pop) 
+                  
+  ds_jags$nagecats = length(unique(ds$Age)) 
+  ds_jags$nyears = length(unique(ds$Year)) 
+
+  for(year in 1:ds_jags$nyears) {
+    for(agecat in 1:ds_jags$nagecats) {
+    if ( exp(ds_jags$pop[year, agecat]) == Inf) { 
+      return('exp(pop) does not exist') 
+      }
+    }
+  }
+  
+  
+  model = function() {  
+    for(year in 1:nyears) {
+      for(agecat in 1:nagecats) {
+          deaths[year, agecat] ~ dpois(mu[year, agecat]) 
+          log(mu[year, agecat]) <- (pop[year, agecat] + b[year, agecat])
+      }
+    }
[…]

### 2016-09-15 Code/4_smoothing_models_src.R

@@ -1,0 +1,14 @@
+
+#Packages 
+
+require(rjags)
+require(R2jags) 
+require(devtools)
+require(jagstools)
+
+source('/Users/kathryn/Dropbox/BlackWhiteGap/Code/4_smoothing_models_fun.R')
+
+load('/Users/kathryn/Dropbox/BlackWhiteGap/Data/alabama_only.Rdata')
+
+
+

### 2016-09-15 Code/life_expectancy_function.R

@@ -1,0 +1,24 @@
+load("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/alabama_only.Rdata")
+
+#example data
+alabama.by.strata <- 
+  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
+  arrange(RaceSex, Year2, Age2) %>%
+  summarise(total_deaths5 = sum(Count_md5),
+            Population = first(Population), 
+            Race2 = first(Race2), 
+            Sex2 = first(Sex2), 
+            StateYearRaceSex = first(StateYearRaceSex))
+
+
+life.table <- function(data, calendar.years, age.groups, num.ages.in.group, death.counts, population.counts){
+  #Ave_Lived_by_Died = ifelse(data[age.groups]=="<1 year", 0.1, 0.5)
+  
+  temp.data <- data %>% mutate(age.spec.death.rate = data[death.counts]/data[population.counts])
+  
+  return(head(data[years]))
+  
+  
+} 
+
+life.table(alabama.by.strata, "Year2", "Age2", "total_deaths5", "Population")

### 2016-09-15 dataverse_files/CR edit state-le-analysis.R

@@ -1,0 +1,357 @@
+ # Decomposition analysis
+ #
+ ##############################################################################################
+ 
+ #Functions
+ #################################################################
+ # a couple of functions for timing the runs
+ tic <- function(gcFirst = TRUE, type=c("elapsed", "user.self", "sys.self"))
+ {
+    type <- match.arg(type)
+    assign(".type", type, envir=baseenv())
+    if(gcFirst) gc(FALSE)
+    tic <- proc.time()[type]         
+    assign(".tic", tic, envir=baseenv())
+    invisible(tic)
+ }
+ 
+ toc <- function()
+ {
+    type <- get(".type", envir=baseenv())
+    toc <- proc.time()[type]
+    tic <- get(".tic", envir=baseenv())
+    print(toc - tic)
+    invisible(toc)
+ }
+
+#functions to convert risks to rates and vice versa
+#risktorate<-function(x,t) {
+#    -log(1-x)/t
+#}
+ratetorisk<-function(x,t) {
+    1-exp(-x*t)
+}
+##################################################################################################
+
+library(BRugs)
+setwd('c:/bugsR/4')
+
+#Matrix to collect ouput of Bayes decomp
+Out <- matrix(0, 15200, 46)
+colnames(Out) <- c("state","gender","age","year",
+                "total","totall","totalu",
+                "direct","directl","directu",
+                "indirect","indirectl","indirectu",
+                "atotal","atotall","atotalu",
+                "adirect","adirectl","adirectu",
+                "aindirect","aindirectl","aindirectu",                
+                "lew","lewl","lewu",
+                "leb","lebl","lebu",
+                "lediff","lediffl","lediffu",
+                "rw","rwl","rwu",
+                "rb","rbl","rbu",
+                "changeb","changebl","changebu",
+                "changew","changewl","changewu",
+                "change","changel","changeu"                
+                )
+count <- 1
+#Flag any instances of risks > 1 among blacks and whites
+flagb <- rep(0, 3)
+flagw <- rep(0, 3)
+
[…]

### 2016-09-15 dataverse_files/state-le-analysis-nhisp.R

@@ -1,0 +1,356 @@
+ # Decomposition analysis
+ # 2/21/2013 - fixed to use non hispanic data
+ # ignoring arriaga decomp to speed things up a little. just putting zeros in those cells of output
+ ##############################################################################################
+ 
+ #Functions
+ #################################################################
+ # a couple of functions for timing the runs
+ tic <- function(gcFirst = TRUE, type=c("elapsed", "user.self", "sys.self"))
+ {
+    type <- match.arg(type)
+    assign(".type", type, envir=baseenv())
+    if(gcFirst) gc(FALSE)
+    tic <- proc.time()[type]         
+    assign(".tic", tic, envir=baseenv())
+    invisible(tic)
+ }
+ 
+ toc <- function()
+ {
+    type <- get(".type", envir=baseenv())
+    toc <- proc.time()[type]
+    tic <- get(".tic", envir=baseenv())
+    print(toc - tic)
+    invisible(toc)
+ }
+
+#functions to convert risks to rates and vice versa
+#risktorate<-function(x,t) {
+#    -log(1-x)/t
+#}
+ratetorisk<-function(x,t) {
+    1-exp(-x*t)
+}
+##################################################################################################
+
+library(BRugs)
+setwd('c:/bugsR/4')
+
+#Matrix to collect ouput of Bayes decomp
+Out<-matrix(0,15200,46)
+colnames(Out)<-c("state","gender","age","year",
+                "total","totall","totalu",
+                "direct","directl","directu",
+                "indirect","indirectl","indirectu",
+                "atotal","atotall","atotalu",
+                "adirect","adirectl","adirectu",
+                "aindirect","aindirectl","aindirectu",                
+                "lew","lewl","lewu",
[…]

### 2016-09-15 dataverse_files/state-le-analysis.R

@@ -1,0 +1,355 @@
+ # Decomposition analysis
+ #
+ ##############################################################################################
+ 
+ #Functions
+ #################################################################
+ # a couple of functions for timing the runs
+ tic <- function(gcFirst = TRUE, type=c("elapsed", "user.self", "sys.self"))
+ {
+    type <- match.arg(type)
+    assign(".type", type, envir=baseenv())
+    if(gcFirst) gc(FALSE)
+    tic <- proc.time()[type]         
+    assign(".tic", tic, envir=baseenv())
+    invisible(tic)
+ }
+ 
+ toc <- function()
+ {
+    type <- get(".type", envir=baseenv())
+    toc <- proc.time()[type]
+    tic <- get(".tic", envir=baseenv())
+    print(toc - tic)
+    invisible(toc)
+ }
+
+#functions to convert risks to rates and vice versa
+#risktorate<-function(x,t) {
+#    -log(1-x)/t
+#}
+ratetorisk<-function(x,t) {
+    1-exp(-x*t)
+}
+##################################################################################################
+
+library(BRugs)
+setwd('c:/bugsR/4')
+
+#Matrix to collect ouput of Bayes decomp
+Out <- matrix(0, 15200, 46)
+colnames(Out) <- c("state","gender","age","year",
+                "total","totall","totalu",
+                "direct","directl","directu",
+                "indirect","indirectl","indirectu",
+                "atotal","atotall","atotalu",
+                "adirect","adirectl","adirectu",
+                "aindirect","aindirectl","aindirectu",                
+                "lew","lewl","lewu",
+                "leb","lebl","lebu",
+                "lediff","lediffl","lediffu",
+                "rw","rwl","rwu",
+                "rb","rbl","rbu",
+                "changeb","changebl","changebu",
+                "changew","changewl","changewu",
+                "change","changel","changeu"                
+                )
+count <- 1
[…]

### 2016-09-15 dataverse_files/state-le-plots.R

@@ -1,0 +1,108 @@
+##  program:  state-le-plots.r
+##  task:     life expectancy plots
+## 	input:	  state-le-results.dta					
+##	output:		various figures
+##  project:  state differences in black-white life expectancy
+##  author:   sam harper \ 1aug2014
+
+## users will need to specify working directory with life expectancy output
+# setwd(".......")  
+
+# setup
+setwd("/Users/samharper/Dropbox/SRJ/Data/replication files")
+library(ggplot2)
+library(foreign)
+library(gridExtra)
+library(reshape2)
+library(directlabels)
+library(doBy)
+library(proto)
+    
+# Grab results from decomposition
+decomp <- read.dta("state-le-results.dta")
+
+# subset to life expectancy at birth
+tg <- subset(decomp,age=="00-00 years")
+
+stabblist <- unique(tg$stabb)
+stnamelist <- unique(tg$stname)
+stnum <- unique(tg$stseer)
+rnamelist <- unique(tg$regname)
+rnum <- unique(tg$reg)
+dnamelist <- unique(tg$divname)
+dnum <- unique(tg$div)
+
+# labels for state, region, division
+tg$gender <- as.numeric(tg$gender)
+tg$abb <- factor(tg$stseer,levels=c(stnum),labels=c(stabblist))
+tg$name <- factor(tg$stseer,levels=c(stnum),labels=c(stnamelist))
+tg$regn <- factor(tg$reg,levels=c(rnum),labels=c(rnamelist))
+tg$regn <- factor(tg$reg,levels=c(1L,2L,3L,4L),labels=c("Northeast","Midwest","South","West"))
+tg$divn <- factor(tg$div,levels=c(dnum),labels=c(dnamelist))
+tg$divn <- factor(tg$div,levels=c(1L,2L,3L,4L,5L,6L,7L,8L,9L),labels=c("New England","Middle Atlantic","East North Central","West North Central","South Atlantic","East South Central","West South Central","Mountain","Pacific"))
+
+### Need to just estimate a weighted average of each state-specific trend rather than using geom_smooth (too variable depending on scale)
+
+# aggregate by division, year and gender
+tg$lediffse <- abs(tg$lediffu - tg$lediffl)/(2*1.96) #standard err for gap in each state
[…]

### 2016-09-16 working_teaching_examples/how_to_abridged_life_table.Rmd
wrote life expectancy function

---
title: "How to make an abridged life table and calculate life expectancy"
author: "Corinne Riddell"
date: "September 15, 2016"
output: html_document
bibliography: "/Users/corinneriddell/Documents/repos/BlackWhiteGap/papers/bibliography.bib"
---

##What is a life table?
"A life table is a system way of keeping track of the mortality experience of a group." [@selvin_text, Chapter 9 Life Tables: An Introduction].

**Cohort life table**

A cohort life table is constructed by following a cohort of individuals through time and recording their exact age at death for all members of the cohort [@selvin_text]. For example, you could take the cohort of individuals born in 1998 in Montreal and each member's year of death. However, this would require waiting until the last member died and would therefore take a very long time.

**Current life table**

As a practical alternative, a life table can be made using current mortality rates for a population. This is called a current life table [@selvin_text].

**Complete vs abridged life table**

If mortality rates can be calculated for each age, then the life table is called complete, compared to life tables calculated based on data categorized by age groupings (i.e., <1 years old, 1-4 years old, 5-9 years old, etc) which are called abridged.

**Elements used to construct a complete (current) life table**

1. Age interval ($x$ to $x+1$): For a complete life table all age groups are 1 year in length, except for the last group, which is left open-ended (e.g, 85 years or older).

2. Number alive ($l_x$): The number of individuals alive at age $x$. For every interval, $l_x$ defined the population at risk of death during the interval. To begin, we set $l_0=100000$.

3. Number of deaths ($d_x$): The number of individuals who die between the ages of $x$ and $x+1$.

4. Probability of death ($q_x$): Prob(death before x + 1 | alive at x) = $d_x/l_x$.

5. Number of years lived ($L_x$): The person-time-at-risk of death lived by the cohort between ages $x$ and $x+1$. 
    a. Each individual who does not die during the interval contributes 1 person-year, and those who die contribute the proportion of the year up until their death (which is approximated to be 9% of the year for the <1 age group and half a year (50%) for all other groups). This "average time contributed by those who die" is denoted by $\bar{a_x}$.
    b. Therefore, $L_x=(l_x-d_x)+\bar{a_x}d_x$.

6. Total time lived ($T_x$): This is the accumulation of person-time-at-risk at age x and beyond -- the total number of person years lived by those alive at age $x$.

7. Expectation of life ($e_x$): The average number of additional years expected to be lived by those individuals alive at age $x$. $e_x=T_x/l_x$.

To construct the life table, we first need to calculate the probabilities of death, $q_x$. To find a formula for $q_x$ we first write the formula for $R_x$, the age-specific mortality rate, and re-arrange to isolate $q_x$:

$R_x=d_x/L_x$

$R_x=d_x/(l_x-d_x)+\bar{a_x}d_x$, dividing the numerator and denominator by l_x:

$R_x=q_x/(1-q_x)+\bar{a_x}q_x$, rearranging to solve for q_x:

$q_x=\frac{R_x}{1+(1-\bar{a_x})R_x}$.

Observed mortality rates $R_x$ can therefore be used to calculate the probability of dying within each age-specific interval. The observed mortality rates are calculated as $d_x/Popn_x$, where $Popn_x$ is the number of individuals alive during age x. While this looks like a proportion, it calculates a rate when we assume the population is stationary, thereby implying there are $Popn_x$ person-years at risk of death during the age bracket.

After $q_x$ is calculated based on $R_x$, all of the other quantities defined above can also be calculated, except for the $L_x$, the number of years lived, in the last age group. In this age group, $\bar{a_x}$ is unknown, and all the individuals in the age bracket will die (i.e., $q_x=1$). Because $d_x/L_x=R_x$ and $d_x=l_x$ in this age group, we can substitute $l_x$ for $d_x$ and write $l_x/L_x=R_x$, implying that $L_x=l_x/R_x$ in this age group.

##Assumptions of the life table
1. The number of births occurring each year is constant.
2. Deaths are uniformly distributed during the interval after 1 years old.
3. There is no population growth.
Together, these assumptions imply that the population must be stationary.
3.

**Construction of an abridged (current) life table**

Recall that an abridged life table includes age intervals that are longer than a year. In practice, these intervals are often 5 years in length and can be represented by $n_x$, where $x$ is the age at the beginning of the interval. The formula for $q_x$ can be generalized to this context, where: $q_x=\frac{n_xR_x}{1+(1-\bar(a_x))n_xR_x}$.

**An example**

This data was used in a previously published paper [@auger2014mortality] and we use it here to show how to create an abridged life table and calculate life expectancy at birth.

First, I create the dataset in R:
```{r create_dataset}
canada.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90),
"obs.deaths" = c(4054, 625, 445, 622, 2568, 3770, 3434,
  3681, 5155, 8449, 13421, 19371, 25448, 
  31465, 37441, 48796, 64763, 72512, 
  60109, 43812),
"obs.popn" = c(719953, 2835604, 3654546, 4071400, 4464549, 4564679, 4296613, 4237402,
               4532151, 5085092, 5163532, 4597916, 3937984, 3120981, 2323629, 1862971,
               1499907, 997492, 492445, 196109))

head(canada.men)
```

I then add an additional column that denotes how many years are in each age bracket:
```{r add_n_x}
canada.men$n_x = 5
canada.men$n_x[canada.men$ages == 0] = 1
canada.men$n_x[canada.men$ages == 1] = 4
```

```{r}
#dataset needs to be in order of age
life.table <- function(data, calendar.years, age.groups, num.ages.in.group, death.counts, population.counts){

  data["R_x"] <- data[death.counts]/data[population.counts] #mortality rates
  data["a_x"] <- 0.5 #average proportion of interval lived
  data[1, "a_x"] <- 0.09
  data["q_x"] <- data[num.ages.in.group]*data["R_x"]/(1 + (1 - data["a_x"])*data[num.ages.in.group]*data["R_x"]) #probability of dying
  data[dim(data)[1], "q_x"] <- 1 #100% die in last age group
  data["p_x"] <- 1 - data["q_x"] #probability of survival

  data["l_x"] <- NA #l_x: is the number of individuals alive at the start of the interval
  data[1, "l_x"] <- 100000

  for(i in 2:dim(data)[1]) {
    data[i, "l_x"] <- data[i - 1, "l_x"]*data[i-1, "p_x"]
  }

  data["d_x"] <- data["l_x"]*data["q_x"] #number of deaths
  data["L_x"] <- (data["l_x"] - data["d_x"])*data["n_x"] + data["d_x"]*data["a_x"]*data["n_x"] #person time in the age bracket
  data[dim(data)[1], "L_x"] <- data[dim(data)[1], "l_x"]/data[dim(data)[1], "R_x"] #different for last age bracket

  #now to calculate the total person time lived at or after age x, T_x:
  data["cumsum"] <- cumsum(data["L_x"])
  data["l.cumsum"] <- c(0, data[1:(dim(data)[1] - 1), "cumsum"])
  data["T_x"] <- sum(data["L_x"]) - data["l.cumsum"]
  data["e_x"] <- data["T_x"]/data["l_x"]

  return(data)

}

life.table.cm <- life.table(data = canada.men, age.groups = "ages", num.ages.in.group = "n_x", death.counts = "obs.deaths", population.counts = "obs.popn")

round(life.table.cm, 2)
```

##References

### 2016-09-16 Code/life_expectancy_function.R

@@ -4,21 +4,45 @@ load("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/alabama_only.Rdata")
 alabama.by.strata <- 
   dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
   arrange(RaceSex, Year2, Age2) %>%
+  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years")) %>%
   summarise(total_deaths5 = sum(Count_md5),
             Population = first(Population), 
             Race2 = first(Race2), 
             Sex2 = first(Sex2), 
-            StateYearRaceSex = first(StateYearRaceSex))
+            StateYearRaceSex = first(StateYearRaceSex),
+            nx = first(nx))
 
 
+#dataset needs to be in order of age
 life.table <- function(data, calendar.years, age.groups, num.ages.in.group, death.counts, population.counts){
-  #Ave_Lived_by_Died = ifelse(data[age.groups]=="<1 year", 0.1, 0.5)
   
-  temp.data <- data %>% mutate(age.spec.death.rate = data[death.counts]/data[population.counts])
+  data["R_x"] <- data[death.counts]/data[population.counts] #mortality rates
+  data["a_x"] <- 0.5 #average proportion of interval lived
+  data[1, "a_x"] <- 0.09
+  data["q_x"] <- data[num.ages.in.group]*data["R_x"]/(1 + (1 - data["a_x"])*data[num.ages.in.group]*data["R_x"]) #probability of dying
+  data[dim(data)[1], "q_x"] <- 1 #100% die in last age group
+  data["p_x"] <- 1 - data["q_x"] #probability of survival
   
-  return(head(data[years]))
+  data["l_x"] <- NA #l_x: is the number of individuals alive at the start of the interval
+  data[1, "l_x"] <- 100000 
   
+  for(i in 2:dim(data)[1]) {
+    data[i, "l_x"] <- data[i - 1, "l_x"]*data[i-1, "p_x"]
+  }
+  
+  data["d_x"] <- data["l_x"]*data["q_x"] #number of deaths
+  data["L_x"] <- (data["l_x"] - data["d_x"])*data[num.ages.in.group] + data["d_x"]*data["a_x"]*data[num.ages.in.group] #person time in the age bracket
[…]

### 2016-09-16 working_teaching_examples/how_to_abridged_life_table.Rmd
made folder an rstudio project and changed life table code

@@
 
-```{r}
-#dataset needs to be in order of age
-life.table <- function(data, calendar.years, age.groups, num.ages.in.group, death.counts, population.counts){
-
-  data["R_x"] <- data[death.counts]/data[population.counts] #mortality rates
-  data["a_x"] <- 0.5 #average proportion of interval lived
-  data[1, "a_x"] <- 0.09
-  data["q_x"] <- data[num.ages.in.group]*data["R_x"]/(1 + (1 - data["a_x"])*data[num.ages.in.group]*data["R_x"]) #probability of dying
-  data[dim(data)[1], "q_x"] <- 1 #100% die in last age group
-  data["p_x"] <- 1 - data["q_x"] #probability of survival
-
-  data["l_x"] <- NA #l_x: is the number of individuals alive at the start of the interval
-  data[1, "l_x"] <- 100000
-
-  for(i in 2:dim(data)[1]) {
-    data[i, "l_x"] <- data[i - 1, "l_x"]*data[i-1, "p_x"]
-  }
-
-  data["d_x"] <- data["l_x"]*data["q_x"] #number of deaths
-  data["L_x"] <- (data["l_x"] - data["d_x"])*data["n_x"] + data["d_x"]*data["a_x"]*data["n_x"] #person time in the age bracket
-  data[dim(data)[1], "L_x"] <- data[dim(data)[1], "l_x"]/data[dim(data)[1], "R_x"] #different for last age bracket
-
-  #now to calculate the total person time lived at or after age x, T_x:
-  data["cumsum"] <- cumsum(data["L_x"])
-  data["l.cumsum"] <- c(0, data[1:(dim(data)[1] - 1), "cumsum"])
-  data["T_x"] <- sum(data["L_x"]) - data["l.cumsum"]
-  data["e_x"] <- data["T_x"]/data["l_x"]
-
-  return(data)
-
-}
-
+```{r calculate_life_expectancy}
 life.table.cm <- life.table(data = canada.men, age.groups = "ages", num.ages.in.group = "n_x", death.counts = "obs.deaths", population.counts = "obs.popn")

### 2016-09-16 Code/life_expectancy_function.R

@@ -44,5 +44,7 @@ life.table <- function(data, calendar.years, age.groups, num.ages.in.group, deat
   
 } 
 
-lt <- life.table(data = subset(alabama.by.strata, StateYearRaceSex = "Alabama.1969.White.Male"), age.groups = "Age2", 
+onestrat <- subset(alabama.by.strata, StateYearRaceSex == "Alabama.1969.White.Male")
+lt <- life.table(data = subset(alabama.by.strata, StateYearRaceSex == "Alabama.1969.White.Male"), age.groups = "Age2", 
            num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population")
+

### 2016-09-16 working_teaching_examples/how_to_abridged_life_table.Rmd
added life expectancy examples

@@
 ##What is a life table?
-"A life table is a system way of keeping track of the mortality experience of a group." [@selvin_text, Chapter 9 Life Tables: An Introduction].
+"A life table is a systematic way of keeping track of the mortality experience of a group." [@selvin_text, Chapter 9 Life Tables: An Introduction].
 
@@
 
-This data was used in a previously published paper [@auger2014mortality] and we use it here to show how to create an abridged life table and calculate life expectancy at birth.
+This data was used in a previously published paper [@auger2014mortality]. We use it to show how the life expectancy function works.
 
@@
 canada.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90),
-"obs.deaths" = c(4054, 625, 445, 622, 2568, 3770, 3434,
-  3681, 5155, 8449, 13421, 19371, 25448, 
-  31465, 37441, 48796, 64763, 72512, 
-  60109, 43812),
+                         "obs.deaths" = c(4054, 625, 445, 622, 2568, 3770, 3434, 3681, 5155, 8449, 13421, 19371, 
+                                          25448, 31465, 37441, 48796, 64763, 72512, 60109, 43812),
                          "obs.popn" = c(719953, 2835604, 3654546, 4071400, 4464549, 4564679, 4296613, 4237402,
@@
 
-I then add an additional column that denotes how many years are in each age bracket:
+I then add a column that denotes how many years are in each age bracket:
 ```{r add_n_x}
@@
 canada.men$n_x[canada.men$ages == 1] = 4
+```
+
+```{r function_lifeexpectancy}
+#dataset needs to be in order of age
+life.table <- function(data, age.groups, num.ages.in.group, death.counts, population.counts, ave.prop.lived = NA, calendar.years = NA){
+
+  data["R_x"] <- data[death.counts]/data[population.counts] #mortality rates
+
+  if(is.na(ave.prop.lived) == T) {
+      data["a_x"] <- 0.5 #average proportion of interval lived
+      data[1, "a_x"] <- 0.09
+  } else{
+      data["a_x"] <- data[ave.prop.lived]
+  }
+
+  data["q_x"] <- data[num.ages.in.group]*data["R_x"]/(1 + (1 - data["a_x"])*data[num.ages.in.group]*data["R_x"]) #probability of dying
+  data[dim(data)[1], "q_x"] <- 1 #100% die in last age group
+  data["p_x"] <- 1 - data["q_x"] #probability of survival
+
+  data["l_x"] <- NA #l_x: is the number of individuals alive at the start of the interval
+  data[1, "l_x"] <- 100000
+
+  for(i in 2:dim(data)[1]) {
+    data[i, "l_x"] <- data[i - 1, "l_x"]*data[i-1, "p_x"]
+  }
+
+  data["d_x"] <- data["l_x"]*data["q_x"] #number of deaths
+  data["L_x"] <- (data["l_x"] - data["d_x"])*data[num.ages.in.group] + data["d_x"]*data["a_x"]*data[num.ages.in.group] #person time in the age bracket
+  data[dim(data)[1], "L_x"] <- data[dim(data)[1], "l_x"]/data[dim(data)[1], "R_x"] #different for last age bracket
+
+  #now to calculate the total person time lived at or after age x, T_x:
+  data["cumsum"] <- cumsum(data["L_x"])
+  data["l.cumsum"] <- c(0, data[1:(dim(data)[1] - 1), "cumsum"])
+  data["T_x"] <- sum(data["L_x"]) - data["l.cumsum"]
+  data["e_x"] <- data["T_x"]/data["l_x"]
+
+  return(data)
+
+} 
 ```
@@
 
+**A second example**
+
+In this example we use the data that Sam entered into Stata to illustrate the life expectancy concept to me.
+I'll replicate the life table here to ensure that I can arrive at the same results as he does in Stata.
+
+```{r lifetable_cali_abridged}
+cali.abridged <- data.frame("ages" = c("0-1y", "1-4y", "5-9y", "10-14y", "15+y"), "nx" = c(1, 4, 5, 5, 5),
+                         "obs.deaths" = c(1635, 168, 251, 371, 23300), #last number is fictitious
+                         "obs.popn" = c(123342, 435167, 578964, 649693, 143600))
+
+cali.abridged
+
+life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
+           death.counts = "obs.deaths", population.counts = "obs.popn")
+```
+
+```{r lifetable_cali_complete}
+cali.complete <- data.frame("ages" = c(0:15), "nx" = rep(1,16),
+                         "obs.deaths" = c(1635, 64, 41, 22, 41, 55, 42, 58, 44, 52, 48, 60, 52, 82, 129, 23300), #last number is fictitious
+                         "obs.popn" = c(123342, 111520, 109200, 108749, 105698, 110548, 106857, 112184, 116423,
+                                        132952, 134266, 128938, 125502, 128212, 132775, 143600))
+
+head(cali.complete)
+
+life.table(data = cali.complete, age.groups = "ages", num.ages.in.group = "nx", 
+           death.counts = "obs.deaths", population.counts = "obs.popn")
+
+cali.complete$avelived <- 0.5
+cali.complete$avelived[cali.complete$ages==0] <- 0.1
+
+life.table(data = cali.complete, age.groups = "ages", num.ages.in.group = "nx", ave.prop.lived = "avelived",
+           death.counts = "obs.deaths", population.counts = "obs.popn")
+```
+
 ##References

### 2016-09-19 working_teaching_examples/how_to_abridged_life_table.Rmd
change life expectancy calc 2

@@
 
-As a practical alternative, a life table can be made using current mortality rates for a population. This is called a current life table [@selvin_text].
+As a practical alternative, a life table can be made using current mortality rates for a population.
+This is called a current life table [@selvin_text] or a period life table [@preston_demography_text].
 
@@
 
-6. Total time lived ($T_x$): This is the accumulation of person-time-at-risk at age x and beyond -- the total number of person years lived by those alive at age $x$.
+6. Cumulative time lived ($T_x$): This is the accumulation of person-time-at-risk at age x and beyond -- the total number of person years lived by those alive at age $x$.
 
@@
 ##Assumptions of the life table
+
 1. The number of births occurring each year is constant.
@@
 3. There is no population growth.
+
 Together, these assumptions imply that the population must be stationary.
-3.
 
@@
 The formula for $q_x$ can be generalized to this context, where: $q_x=\frac{n_xR_x}{1+(1-\bar(a_x))n_xR_x}$.
+
+**Derivation of the formula for $q_x$ when the life table is abridged**
+
+This derivation was modified from the Preston demography text [@preston_demography_text].
+Note that in the Preston book $a_x$ represents *person-years* not the proportion of the year, which is why the formulas look different (because we have to write $n_x\bar{a_x}$ in place of Preston's $a_x$).
+
+We start with the simple formula for q_x. As before:
+$q_x=d_x/l_x$.
+
+Now, $L_x = n_x*(l_x - d_x) + n_x\bar{a_x}d_x$
+$L_x = n_xl_x - n_xd_x + n_x\bar{a_x}d_x$
+$l_x = \frac{L_x + n_xd_x - n_x\bar{a_x}d_x}{n_x}$
+$l_x = \frac{L_x + d_xn_x(1 - \bar{a_x}){n_x}$
+
+We can then substitute this expression for $l_x$ into the equations for $q_x$:
+
+$q_x=\frac{d_xn_x}{L_x + d_xn_x(1 - \bar{a_x}}$, and dividing numerator and denominator by L_x, we get
+$q_x=\frac{R_xn_x}{1 + R_xn_x(1 - \bar{a_x}}$.
 
@@
 ```{r function_lifeexpectancy}
-#dataset needs to be in order of age
-life.table <- function(data, age.groups, num.ages.in.group, death.counts, population.counts, ave.prop.lived = NA, calendar.years = NA){
+#dataset needs to be in order of age.
+life.table <- function(data, age.groups, num.ages.in.group, death.counts, population.counts, ave.prop.lived = NA){
 
@@
 
-  data["q_x"] <- data[num.ages.in.group]*data["R_x"]/(1 + (1 - data["a_x"])*data[num.ages.in.group]*data["R_x"]) #probability of dying
+  data["q_x"] <- data[num.ages.in.group]*data["R_x"]/(1 + (1 - data["a_x"])*data[num.ages.in.group]*data["R_x"]) 
+  #probability of dying
+
   data[dim(data)[1], "q_x"] <- 1 #100% die in last age group
+
   data["p_x"] <- 1 - data["q_x"] #probability of survival
@@
   data["cumsum"] <- cumsum(data["L_x"])
-  data["l.cumsum"] <- c(0, data[1:(dim(data)[1] - 1), "cumsum"])
+  data["l.cumsum"] <- c(0, unlist(data[1:(dim(data)[1] - 1), "cumsum"]))
   data["T_x"] <- sum(data["L_x"]) - data["l.cumsum"]
@@
 
-```{r lifetable_cali_abridged}
-cali.abridged <- data.frame("ages" = c("0-1y", "1-4y", "5-9y", "10-14y", "15+y"), "nx" = c(1, 4, 5, 5, 5),
-                         "obs.deaths" = c(1635, 168, 251, 371, 23300), #last number is fictitious
-                         "obs.popn" = c(123342, 435167, 578964, 649693, 143600))
-
-cali.abridged
-
-life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
-           death.counts = "obs.deaths", population.counts = "obs.popn")
-```
-
 ```{r lifetable_cali_complete}
@@
 
+```{r lifetable_cali_abridged}
+cali.complete$group <- cut(cali.complete$ages, breaks=c(0, 1, 5, 10, 15, 20), 
+                           labels = c("<1 years old", "1-4 years old", "5-9 years old", 
+                                      "10-14 years old", "15 or older"), include.lowest = T, right = F)
+
+table(cali.complete$group, cali.complete$ages)
+
+library(dplyr)
+cali.abridged <- cali.complete %>% group_by(group) %>% 
+                                   summarise(nx = sum(nx), ages = first(group), popn = sum(obs.popn), deaths = sum(obs.deaths),
+                                             avelived = first(avelived))
+
+cali.abridged
+
+life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
+           death.counts = "deaths", population.counts = "popn")
+
+life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
+           death.counts = "deaths", population.counts = "popn", ave.prop.lived = "avelived")
+```
+
+```{r}
+load("Data/alabama_only.Rdata")
+
+#example data
+alabama.by.strata <- 
+  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
+  arrange(RaceSex, Year2, Age2) %>%
+  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years")) %>%
+  summarise(total_deaths5 = sum(Count_md5),
+            Population = first(Population), 
+            Race2 = first(Race2), 
+            Sex2 = first(Sex2), 
+            StateYearRaceSex = first(StateYearRaceSex),
+            nx = first(nx))
+
+first.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.White.Male", ]
+
+system.time(first.strata.life.table <- life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
+
+View(first.strata.life.table)
+
+system.time(
+for(stratum.id in levels(droplevels(alabama.by.strata$StateYearRaceSex))) {
+  assign(paste0("lt.", as.character(stratum.id)), 
+         life.table(data = subset(alabama.by.strata, StateYearRaceSex == stratum.id), 
+                                  age.groups = "Age2", num.ages.in.group = "nx", 
+                                  death.counts = "total_deaths5", population.counts = "Population"))
+  print(stratum.id)
+}
+)
+```
+
+Stopped here. For next time: 
+need to change T_x to T_x+n in this formula and add the last term in the empty brackets after leading the number alive variable
+see the auger paper pg 576 for the formula to finish programming.
+data1 <- Quebec
+data2 <- Canada
+
+```{r function_le_age_decomp}
+
+le_age_decomp <- function(data1, data2, num.alive, interval.years.lived, accumulated.years.lived) {
+
+  C_x <- 
+  (data1[num.alive]/data1[1, num.alive])*((data2[interval.years.lived]/data2[num.alive]) - (data1[interval.years.lived]/data1[num.alive])) +
+    ((data2[accumulated.years.lived]/data2[1, num.alive])*((data1[num.alive]/data2[num.alive]) - ))
+}
+
+```
+
 ##References

### 2016-09-19 working_teaching_examples/how_to_abridged_life_table.Rmd

@@
 head(canada.men)
+
+quebec.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90), 
+                         "obs.deaths" = c(1057, 164, 116, 176, 659, 956, 1056, 1110, 1498, 2470, 4116, 6054, 8426,
+                                          11175, 13558, 16567, 21349, 22396, 16851,10878),
+                         "obs.popn" = c(215200, 789710, 1008266, 1179423, 1249360, 1261780, 1349857, 1293521,
+                                        1313834, 1551684, 1622835, 1485249, 1288014, 1077253, 796557, 609341, 
+                                        476211, 300506, 134353, 49678))
+head(quebec.men)
 ```
@@
 ```{r add_n_x}
-canada.men$n_x = 5
-canada.men$n_x[canada.men$ages == 0] = 1
-canada.men$n_x[canada.men$ages == 1] = 4
+canada.men$n_x <- 5
+canada.men$n_x[canada.men$ages == 0] <- 1
+canada.men$n_x[canada.men$ages == 1] <- 4
+
+quebec.men$n_x <- 5
+quebec.men$n_x[quebec.men$ages == 0] <- 1
+quebec.men$n_x[quebec.men$ages == 1] <- 4
 ```
@@
 round(life.table.cm, 2)
+
+life.table.qm <- life.table(data = quebec.men, age.groups = "ages", num.ages.in.group = "n_x", death.counts = "obs.deaths", population.counts = "obs.popn")
+
+round(life.table.qm, 2)
 ```
@@
 
-Stopped here. For next time: 
-need to change T_x to T_x+n in this formula and add the last term in the empty brackets after leading the number alive variable
-see the auger paper pg 576 for the formula to finish programming.
-data1 <- Quebec
-data2 <- Canada
-
-```{r function_le_age_decomp}
-
-le_age_decomp <- function(data1, data2, num.alive, interval.years.lived, accumulated.years.lived) {
-
-  C_x <- 
-  (data1[num.alive]/data1[1, num.alive])*((data2[interval.years.lived]/data2[num.alive]) - (data1[interval.years.lived]/data1[num.alive])) +
-    ((data2[accumulated.years.lived]/data2[1, num.alive])*((data1[num.alive]/data2[num.alive]) - ))
-}
-
-```
-
 ##References

### 2016-09-20 Code/life_expectancy_functions.R
cleaned up life exp fx

@@ -38,12 +38,6 @@ life.table <- function(data, age.groups, num.ages.in.group, death.counts, popula
   
 } 
 
-#Stopped here. For next time: 
-#  need to change T_x to T_x+n in this formula and add the last term in the empty brackets after leading the number alive variable
-#see the auger paper pg 576 for the formula to finish programming.
-#data1 <- Quebec
-#data2 <- Canada
-
 le_age_decomp <- function(data1, data2, num.alive, interval.years.lived, accumulated.years.lived) {
   
   data2["accumulated.lived.after"] <- c(unlist(data2[2:dim(data2)[1], accumulated.years.lived]), 0) #T_(x+n) in Auger's formula on pg 576 (step 1)
@@ -56,14 +50,6 @@ le_age_decomp <- function(data1, data2, num.alive, interval.years.lived, accumul
     (data1[num.alive]/data1[1, num.alive])*((data2[interval.years.lived]/data2[num.alive]) - (data1[interval.years.lived]/data1[num.alive])) +
     ((data2["accumulated.lived.after"]/data2[1, num.alive])*((data1[num.alive]/data2[num.alive]) - (data1["num.alive.next.interval"]/data2["num.alive.next.interval"])))
 
-#  C_x[dim(data1)[1]] <- 
-#    (data1[dim(data1)[1], num.alive]/data1[1, num.alive])*((data2[dim(data2)[1], interval.years.lived]/data2[dim(data2)[1], num.alive]) - (data1[dim(data1)[1], interval.years.lived]/data1[dim(data1)[1], num.alive])) 
-  
   return(data1)
 }
-
-age_decomp <- le_age_decomp(life.table.qm, life.table.cm, num.alive="l_x", interval.years.lived="L_x", accumulated.years.lived = "T_x")
-
-age_decomp2 <- le_age_decomp(life.table.cm, life.table.qm, num.alive="l_x", interval.years.lived="L_x", accumulated.years.lived = "T_x")
-
 

### 2016-09-20 working_teaching_examples/how_to_abridged_life_table.Rmd

@@
 
-Now, $L_x = n_x*(l_x - d_x) + n_x\bar{a_x}d_x$
+Now, $L_x = n_x(l_x - d_x) + n_x\bar{a_x}d_x$
+
 $L_x = n_xl_x - n_xd_x + n_x\bar{a_x}d_x$
+
 $l_x = \frac{L_x + n_xd_x - n_x\bar{a_x}d_x}{n_x}$
-$l_x = \frac{L_x + d_xn_x(1 - \bar{a_x}){n_x}$
+
+$l_x = \frac{L_x + d_xn_x(1 - \bar{a_x})}{n_x}$
 
@@
 
-$q_x=\frac{d_xn_x}{L_x + d_xn_x(1 - \bar{a_x}}$, and dividing numerator and denominator by L_x, we get
-$q_x=\frac{R_xn_x}{1 + R_xn_x(1 - \bar{a_x}}$.
+$q_x=\frac{d_xn_x}{L_x + d_xn_x(1 - \bar{a_x})}$, and dividing numerator and denominator by $L_x$, we get
+
+$q_x=\frac{R_xn_x}{1 + R_xn_x(1 - \bar{a_x})}$.
 
@@
 
-head(canada.men)
-
 quebec.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90), 
@@
                                         476211, 300506, 134353, 49678))
-head(quebec.men)
 ```
@@
 quebec.men$n_x[quebec.men$ages == 1] <- 4
+
+head(canada.men)
+head(quebec.men)
 ```
 
-```{r function_lifeexpectancy}
-#dataset needs to be in order of age.
-life.table <- function(data, age.groups, num.ages.in.group, death.counts, population.counts, ave.prop.lived = NA){
-
-  data["R_x"] <- data[death.counts]/data[population.counts] #mortality rates
-
-  if(is.na(ave.prop.lived) == T) {
-      data["a_x"] <- 0.5 #average proportion of interval lived
-      data[1, "a_x"] <- 0.09
-  } else{
-      data["a_x"] <- data[ave.prop.lived]
-  }
-
-  data["q_x"] <- data[num.ages.in.group]*data["R_x"]/(1 + (1 - data["a_x"])*data[num.ages.in.group]*data["R_x"]) 
-  #probability of dying
-
-  data[dim(data)[1], "q_x"] <- 1 #100% die in last age group
-
-  data["p_x"] <- 1 - data["q_x"] #probability of survival
-
-  data["l_x"] <- NA #l_x: is the number of individuals alive at the start of the interval
-  data[1, "l_x"] <- 100000
-
-  for(i in 2:dim(data)[1]) {
-    data[i, "l_x"] <- data[i - 1, "l_x"]*data[i-1, "p_x"]
-  }
-
-  data["d_x"] <- data["l_x"]*data["q_x"] #number of deaths
-  data["L_x"] <- (data["l_x"] - data["d_x"])*data[num.ages.in.group] + data["d_x"]*data["a_x"]*data[num.ages.in.group] #person time in the age bracket
-  data[dim(data)[1], "L_x"] <- data[dim(data)[1], "l_x"]/data[dim(data)[1], "R_x"] #different for last age bracket
-
-  #now to calculate the total person time lived at or after age x, T_x:
-  data["cumsum"] <- cumsum(data["L_x"])
-  data["l.cumsum"] <- c(0, unlist(data[1:(dim(data)[1] - 1), "cumsum"]))
-  data["T_x"] <- sum(data["L_x"]) - data["l.cumsum"]
-  data["e_x"] <- data["T_x"]/data["l_x"]
-
-  return(data)
-
-} 
+This source file contains the life expectancy functions used in this project:
+```{r load_functions}
+source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
 ```
@@
 round(life.table.qm, 2)
+```
+
+```{r calc_age_decomp}
+age_decomp <- le_age_decomp(life.table.qm, life.table.cm, num.alive="l_x", interval.years.lived="L_x", accumulated.years.lived = "T_x")
+
+age_decomp2 <- le_age_decomp(life.table.cm, life.table.qm, num.alive="l_x", interval.years.lived="L_x", accumulated.years.lived = "T_x")
 ```
@@
 
-life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
-           death.counts = "deaths", population.counts = "popn")
+data.frame(life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
+           death.counts = "deaths", population.counts = "popn"))
 
-life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
-           death.counts = "deaths", population.counts = "popn", ave.prop.lived = "avelived")
+data.frame(life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
+           death.counts = "deaths", population.counts = "popn", ave.prop.lived = "avelived"))
 ```
@@
 ```{r}
-load("Data/alabama_only.Rdata")
-
+#load("Data/alabama_only.Rdata")
+load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
 #example data
@@
 
-system.time(first.strata.life.table <- life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
+system.time(first.strata.life.table <- data.frame(life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population")))
 
-View(first.strata.life.table)
+first.strata.life.table[, c("Age2", "total_deaths5", "Population", "R_x", "q_x", "l_x", "d_x", "L_x", "T_x", "e_x")]
 
@@
                                   death.counts = "total_deaths5", population.counts = "Population"))
-  print(stratum.id)
 }

### 2016-09-20 working_teaching_examples/how_to_abridged_life_table.Rmd
changed output of the age decomp fx

@@
 age_decomp <- le_age_decomp(life.table.qm, life.table.cm, num.alive="l_x", interval.years.lived="L_x", accumulated.years.lived = "T_x")
+age_decomp
+#check:
+sum(age_decomp$C_x)
+age_decomp$Life_Expectancy_2[age_decomp$Ages ==0] - age_decomp$Life_Expectancy_1[age_decomp$Ages ==0]
 
+#also compare to the decomposition if the data is entered in the reverse order:
 age_decomp2 <- le_age_decomp(life.table.cm, life.table.qm, num.alive="l_x", interval.years.lived="L_x", accumulated.years.lived = "T_x")
+age_decomp2
+#check:
+sum(age_decomp2$C_x)
+age_decomp2$Life_Expectancy_2[age_decomp2$Ages ==0] - age_decomp2$Life_Expectancy_1[age_decomp2$Ages ==0]
+
 ```

### 2016-09-20 Code/life_expectancy_functions.R

@@ -1,4 +1,6 @@
 #dataset needs to be in order of age
+#none of the population.counts can equal 0. If there are rows with population counts of 0 then first need 
+#to run the collapse age brackets function.
 life.table <- function(data, age.groups, num.ages.in.group, death.counts, population.counts, ave.prop.lived = NA){
   
   data["R_x"] <- data[death.counts]/data[population.counts] #mortality rates
@@ -38,18 +40,26 @@ life.table <- function(data, age.groups, num.ages.in.group, death.counts, popula
   
 } 
 
-le_age_decomp <- function(data1, data2, num.alive, interval.years.lived, accumulated.years.lived) {
+le_age_decomp <- function(life.table1, life.table2, num.alive, interval.years.lived, accumulated.years.lived) {
   
-  data2["accumulated.lived.after"] <- c(unlist(data2[2:dim(data2)[1], accumulated.years.lived]), 0) #T_(x+n) in Auger's formula on pg 576 (step 1)
+  life.table2["accumulated.lived.after"] <- c(unlist(life.table2[2:dim(life.table2)[1], accumulated.years.lived]), 0) #T_(x+n) in Auger's formula on pg 576 (step 1)
   
-  data1["num.alive.next.interval"] <- c(unlist(data1[2:dim(data2)[1], num.alive]), 0) #l_(x+n) 
-  data2["num.alive.next.interval"] <- c(unlist(data2[2:dim(data2)[1], num.alive]), 1) #--> end in 1 to fix calculation error of dividing by 0
-  #this *only* holds when data2 is the one with the last term in the denominator
-  
-  data1["C_x"] <- 
-    (data1[num.alive]/data1[1, num.alive])*((data2[interval.years.lived]/data2[num.alive]) - (data1[interval.years.lived]/data1[num.alive])) +
-    ((data2["accumulated.lived.after"]/data2[1, num.alive])*((data1[num.alive]/data2[num.alive]) - (data1["num.alive.next.interval"]/data2["num.alive.next.interval"])))
+  life.table1["num.alive.next.interval"] <- c(unlist(life.table1[2:dim(life.table2)[1], num.alive]), 0) #l_(x+n) 
+  life.table2["num.alive.next.interval"] <- c(unlist(life.table2[2:dim(life.table2)[1], num.alive]), 1) #--> end in 1 to fix calculation error of dividing by 0
 
-  return(data1)
[…]

### 2016-09-20 working_teaching_examples/how_to_abridged_life_table.Rmd
prep work for the cod decomp fx

@@
 
-```{r lifetable_cali_abridged}
+```{r lifetable_cali_abridged, warning=F, message=F}
 cali.complete$group <- cut(cali.complete$ages, breaks=c(0, 1, 5, 10, 15, 20), 
@@
 
+alabama.by.strata.cod <- dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
+                                               arrange(RaceSex, Year2, Age2) %>%
+                                               mutate(total_deaths5 = sum(Count_md5),
+                                                      cod_prop_death5 = Count_md5/total_deaths5)
+
+first.strata.cod <- alabama.by.strata.cod[alabama.by.strata$StateYearRaceSex == "Alabama.1969.White.Male", ]
+
 system.time(first.strata.life.table <- data.frame(life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population")))

### 2016-09-20 Code/life_expectancy_functions.R

@@ -40,6 +40,8 @@ life.table <- function(data, age.groups, num.ages.in.group, death.counts, popula
   
 } 
 
+#takes the life table output from the first function -- need two life tables to perform the comparison
+#ask KTM about if I should remove the last three params? ... I am thinking that would be "cleaner" but want a second opinion.
 le_age_decomp <- function(life.table1, life.table2, num.alive, interval.years.lived, accumulated.years.lived) {
   
   life.table2["accumulated.lived.after"] <- c(unlist(life.table2[2:dim(life.table2)[1], accumulated.years.lived]), 0) #T_(x+n) in Auger's formula on pg 576 (step 1)
@@ -47,7 +49,7 @@ le_age_decomp <- function(life.table1, life.table2, num.alive, interval.years.li
   life.table1["num.alive.next.interval"] <- c(unlist(life.table1[2:dim(life.table2)[1], num.alive]), 0) #l_(x+n) 
   life.table2["num.alive.next.interval"] <- c(unlist(life.table2[2:dim(life.table2)[1], num.alive]), 1) #--> end in 1 to fix calculation error of dividing by 0
 
-  decomp.table <- data.frame("Ages" = life.table1[ ,1], "Life_Expectancy_1" = life.table1["e_x"], "Life_Expectancy_2" = life.table2["e_x"],
+  decomp.table <- data.frame("Ages" = life.table1[ ,1], life.table1["e_x"], life.table2["e_x"],
                              "C_x" = rep(NA, dim(life.table1)[1]))
   
   names(decomp.table)[2] <- "Life_Expectancy_1"
@@ -60,6 +62,11 @@ le_age_decomp <- function(life.table1, life.table2, num.alive, interval.years.li
   return(decomp.table)
 }
 
-cause_of_death_decomp <- function() {
-  
+#need to decide whether cod.proportions should be a vector (long format) or matrix (wide format) 
+#matrix is in line with Auger's spreadsheet. would use apply function to go across the variables in the data frame
+#vector is easier for me to think about, although need to think about how to perform the vectorization since not all vectors
+#are the same length (the cod proportions vector is 114 long for alabama whereas the other will be 114/6 since they aren't
+#code specific)
+cause_of_death_decomp <- function(cod.proportions, life.table1, life.table2, decomp.table) {
+  decomp.table["C_x"]*((life.table2[]*life.table2["R_x"] - life.table1[]*life.table1["R_x"])/(life.table2[]*life.table2[]))
 }

### 2016-09-20 working_teaching_examples/how_to_abridged_life_table.Rmd
fixed up latex equations

@@
 
-$R_x=d_x/(l_x-d_x)+\bar{a_x}d_x$, dividing the numerator and denominator by l_x:
+$R_x=d_x/(l_x-d_x)+\bar{a_x}d_x$, dividing the numerator and denominator by $l_x$:
 
-$R_x=q_x/(1-q_x)+\bar{a_x}q_x$, rearranging to solve for q_x:
+$R_x=q_x/(1-q_x)+\bar{a_x}q_x$, rearranging to solve for $q_x$:
 
@@
 Observed mortality rates $R_x$ can therefore be used to calculate the probability of dying within each age-specific interval.
-The observed mortality rates are calculated as $d_x/Popn_x$, where $Popn_x$ is the number of individuals alive during age x.
-While this looks like a proportion, it calculates a rate when we assume the population is stationary, thereby implying there are $Popn_x$ person-years at risk of death during the age bracket.
+The observed mortality rates are calculated as $d_x/Pop_x$, where $Pop_x$ is the number of individuals alive during age $x$.
+While this looks like a proportion, it calculates a *rate* when we assume the population is stationary, thereby implying there are $Pop_x$ person-years at risk of death during the age bracket.
 
@@
 In practice, these intervals are often 5 years in length and can be represented by $n_x$, where $x$ is the age at the beginning of the interval.
-The formula for $q_x$ can be generalized to this context, where: $q_x=\frac{n_xR_x}{1+(1-\bar(a_x))n_xR_x}$.
+The formula for $q_x$ can be generalized to this context, where: $q_x=\frac{n_xR_x}{1+(1-\bar{a_x})n_xR_x}$.
 
@@
 
-We can then substitute this expression for $l_x$ into the equations for $q_x$:
+We can then substitute this expression for $l_x$ into the equation for $q_x$:
+
+$q_x=\frac{d_x}{l_x}$
 
@@
 
-```{r}
+```{r alabama_example}
 #load("Data/alabama_only.Rdata")

### 2016-09-20 working_teaching_examples/how_to_abridged_life_table.Rmd
cleaned up age decomp fx

@@
 ```{r calc_age_decomp}
-age_decomp <- le_age_decomp(life.table.qm, life.table.cm, num.alive="l_x", interval.years.lived="L_x", accumulated.years.lived = "T_x")
+age_decomp <- le_age_decomp(life.table.qm, life.table.cm)
 age_decomp
@@
 #also compare to the decomposition if the data is entered in the reverse order:
-age_decomp2 <- le_age_decomp(life.table.cm, life.table.qm, num.alive="l_x", interval.years.lived="L_x", accumulated.years.lived = "T_x")
+age_decomp2 <- le_age_decomp(life.table.cm, life.table.qm)
 age_decomp2

### 2016-09-20 Code/life_expectancy_functions.R

@@ -1,6 +1,9 @@
 #dataset needs to be in order of age
 #none of the population.counts can equal 0. If there are rows with population counts of 0 then first need 
 #to run the collapse age brackets function.
+
+#change names to something like col.name.death.counts or death.counts.in.data
+
 life.table <- function(data, age.groups, num.ages.in.group, death.counts, population.counts, ave.prop.lived = NA){
   
   data["R_x"] <- data[death.counts]/data[population.counts] #mortality rates
@@ -41,13 +44,12 @@ life.table <- function(data, age.groups, num.ages.in.group, death.counts, popula
 } 
 
 #takes the life table output from the first function -- need two life tables to perform the comparison
-#ask KTM about if I should remove the last three params? ... I am thinking that would be "cleaner" but want a second opinion.
-le_age_decomp <- function(life.table1, life.table2, num.alive, interval.years.lived, accumulated.years.lived) {
+le_age_decomp <- function(life.table1, life.table2) {
   
-  life.table2["accumulated.lived.after"] <- c(unlist(life.table2[2:dim(life.table2)[1], accumulated.years.lived]), 0) #T_(x+n) in Auger's formula on pg 576 (step 1)
+  life.table2["accumulated.lived.after"] <- c(unlist(life.table2[2:dim(life.table2)[1], "T_x"]), 0) #T_(x+n) in Auger's formula on pg 576 (step 1)
   
-  life.table1["num.alive.next.interval"] <- c(unlist(life.table1[2:dim(life.table2)[1], num.alive]), 0) #l_(x+n) 
-  life.table2["num.alive.next.interval"] <- c(unlist(life.table2[2:dim(life.table2)[1], num.alive]), 1) #--> end in 1 to fix calculation error of dividing by 0
+  life.table1["num.alive.next.interval"] <- c(unlist(life.table1[2:dim(life.table2)[1], "l_x"]), 0) #l_(x+n) 
+  life.table2["num.alive.next.interval"] <- c(unlist(life.table2[2:dim(life.table2)[1], "l_x"]), 1) #--> end in 1 to fix calculation error of dividing by 0
 
   decomp.table <- data.frame("Ages" = life.table1[ ,1], life.table1["e_x"], life.table2["e_x"],
                              "C_x" = rep(NA, dim(life.table1)[1]))
[…]

### 2016-09-20 working_teaching_examples/how_to_abridged_life_table.Rmd
added plots to compare proportion dying by COD

@@
 first.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.White.Male", ]
+second.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.Black.Male", ]
 
@@
 
-first.strata.cod <- alabama.by.strata.cod[alabama.by.strata$StateYearRaceSex == "Alabama.1969.White.Male", ]
+first.strata.cod <- alabama.by.strata.cod[alabama.by.strata.cod$StateYearRaceSex == "Alabama.1969.White.Male", ]
+second.strata.cod <- alabama.by.strata.cod[alabama.by.strata.cod$StateYearRaceSex == "Alabama.1969.Black.Male", ]
+
+first.strata.cod <- alabama.by.strata.cod[alabama.by.strata.cod$StateYearRaceSex == "Alabama.2013.White.Male", ]
+second.strata.cod <- alabama.by.strata.cod[alabama.by.strata.cod$StateYearRaceSex == "Alabama.2013.Black.Male", ]
+
+cod.tab.alabama.male.1969 <- first.strata.cod %>% ungroup() %>%
+                                                  select(Age2, COD2, cod_prop_death5) %>%
+                                                  rename(prop2 = cod_prop_death5)
+
+cod.tab.alabama.male.1969$prop1 <- second.strata.cod$cod_prop_death5
+
+ggplot(cod.tab.alabama.male.1969, aes(x=COD2, y=prop1)) + 
+  geom_point(aes(col="White")) + geom_point(aes(y=prop2, col="Black")) +
+  scale_y_continuous(name = "Proportion dying from specified cause") +
+  facet_wrap(~Age2) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
 system.time(first.strata.life.table <- data.frame(life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population")))
+
+second.strata.life.table <- data.frame(life.table(data = second.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
 

### 2016-09-21 working_teaching_examples/how_to_abridged_life_table.Rmd
added cod decomp examples

@@
 head(quebec.men)
+
+cod.props <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteGap/working_teaching_examples/auger_cod_proportions.csv")
+
 ```
@@
 
+```
+
+```{r calc_cod_decomp}
+canada.cod.decomp <- cause_of_death_decomp(life.table1 = life.table.qm, life.table2 = life.table.cm, decomp.table = age_decomp,
+                      cod.table = cod.props,age.colname.cod.table = "Age", COD.colname.cod.table = "cod", 
+                      prop1.colname.cod.table = "prop1", prop2.colname.cod.table = "prop2")
+
+cod.contribution$Cause.of.death <- factor(cod.contribution$Cause.of.death, 
+                                          levels = cod.contribution$Cause.of.death[order(cod.contribution$cod.contribution,
+                                                                                         decreasing = T)])
+
+ggplot(cod.contribution, aes(x = Cause.of.death, y = cod.contribution)) + 
+  geom_point() + 
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  geom_abline(intercept = 0, slope = 0)
 ```
@@
 
-first.strata.cod <- alabama.by.strata.cod[alabama.by.strata.cod$StateYearRaceSex == "Alabama.2013.White.Male", ]
-second.strata.cod <- alabama.by.strata.cod[alabama.by.strata.cod$StateYearRaceSex == "Alabama.2013.Black.Male", ]
-
 cod.tab.alabama.male.1969 <- first.strata.cod %>% ungroup() %>%
@@
 cod.tab.alabama.male.1969$prop1 <- second.strata.cod$cod_prop_death5
+cod.tab.alabama.male.1969 <- cod.tab.alabama.male.1969 %>% arrange(COD2, Age2)
 
@@
 
-system.time(first.strata.life.table <- data.frame(life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population")))
+system.time(life.table.whites <- data.frame(life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population")))
 
-second.strata.life.table <- data.frame(life.table(data = second.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
+life.table.blacks <- data.frame(life.table(data = second.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
 
-first.strata.life.table[, c("Age2", "total_deaths5", "Population", "R_x", "q_x", "l_x", "d_x", "L_x", "T_x", "e_x")]
+life.table.whites[, c("Age2", "total_deaths5", "Population", "R_x", "q_x", "l_x", "d_x", "L_x", "T_x", "e_x")]
 
@@
 )
+
+alabama_age_decomp <- le_age_decomp(life.table.blacks, life.table.whites, "Age2")
+alabama_age_decomp
+
+ggplot(alabama_age_decomp, aes(y = C_x, x = Ages)) + geom_point() + scale_y_continuous("Contribution of Age to Gap")
+sum(alabama_age_decomp$C_x)
+life.table.whites$e_x[1] - life.table.blacks$e_x[1]
+
+ggplot(alabama_age_decomp, aes(y = Life_Expectancy_1, x = Ages)) + geom_point(aes(col="Black")) + scale_y_continuous("Life Expectancy (Yrs)") +
+  geom_point(aes(y = Life_Expectancy_2, col="White"))
 ```

### 2016-09-22 Code/error checking the LE decomp fxs.Rmd
split LE explanation and examples into 2 files

--- 2016-09-21 working_teaching_examples/how_to_abridged_life_table.Rmd
@@
 ---
-title: "How to make an abridged life table and calculate life expectancy"
+title: "error checking the LE decomp functions"
 author: "Corinne Riddell"
-date: "September 15, 2016"
+date: "September 22, 2016"
 output: html_document
-bibliography: "/Users/corinneriddell/Documents/repos/BlackWhiteGap/papers/bibliography.bib"
 ---
-
-##What is a life table?
-"A life table is a systematic way of keeping track of the mortality experience of a group." [@selvin_text, Chapter 9 Life Tables: An Introduction].
-
-**Cohort life table**
-
-A cohort life table is constructed by following a cohort of individuals through time and recording their exact age at death for all members of the cohort [@selvin_text].
-For example, you could take the cohort of individuals born in 1998 in Montreal and each member's year of death.
-However, this would require waiting until the last member died and would therefore take a very long time.
-
-**Current life table**
-
-As a practical alternative, a life table can be made using current mortality rates for a population.
-This is called a current life table [@selvin_text] or a period life table [@preston_demography_text].
-
-**Complete vs abridged life table**
-
-If mortality rates can be calculated for each age, then the life table is called complete, compared to life tables calculated based on data categorized by age groupings (i.e., <1 years old, 1-4 years old, 5-9 years old, etc) which are called abridged.
-
-**Elements used to construct a complete (current) life table**
-
-1. Age interval ($x$ to $x+1$): For a complete life table all age groups are 1 year in length, except for the last group, which is left open-ended (e.g, 85 years or older).
-
-2. Number alive ($l_x$): The number of individuals alive at age $x$. For every interval, $l_x$ defined the population at risk of death during the interval. To begin, we set $l_0=100000$.
-
-3. Number of deaths ($d_x$): The number of individuals who die between the ages of $x$ and $x+1$.
-
-4. Probability of death ($q_x$): Prob(death before x + 1 | alive at x) = $d_x/l_x$.
-
-5. Number of years lived ($L_x$): The person-time-at-risk of death lived by the cohort between ages $x$ and $x+1$. 
-    a.
-Each individual who does not die during the interval contributes 1 person-year, and those who die contribute the proportion of the year up until their death (which is approximated to be 9% of the year for the <1 age group and half a year (50%) for all other groups).
-This "average time contributed by those who die" is denoted by $\bar{a_x}$.
-    b. Therefore, $L_x=(l_x-d_x)+\bar{a_x}d_x$.
-
-6. Cumulative time lived ($T_x$): This is the accumulation of person-time-at-risk at age x and beyond -- the total number of person years lived by those alive at age $x$.
-
-7. Expectation of life ($e_x$): The average number of additional years expected to be lived by those individuals alive at age $x$. $e_x=T_x/l_x$.
-
-To construct the life table, we first need to calculate the probabilities of death, $q_x$.
-To find a formula for $q_x$ we first write the formula for $R_x$, the age-specific mortality rate, and re-arrange to isolate $q_x$:
-
-$R_x=d_x/L_x$
-
-$R_x=d_x/(l_x-d_x)+\bar{a_x}d_x$, dividing the numerator and denominator by $l_x$:
-
-$R_x=q_x/(1-q_x)+\bar{a_x}q_x$, rearranging to solve for $q_x$:
-
-$q_x=\frac{R_x}{1+(1-\bar{a_x})R_x}$.
-
-Observed mortality rates $R_x$ can therefore be used to calculate the probability of dying within each age-specific interval.
-The observed mortality rates are calculated as $d_x/Pop_x$, where $Pop_x$ is the number of individuals alive during age $x$.
-While this looks like a proportion, it calculates a *rate* when we assume the population is stationary, thereby implying there are $Pop_x$ person-years at risk of death during the age bracket.
-
-After $q_x$ is calculated based on $R_x$, all of the other quantities defined above can also be calculated, except for the $L_x$, the number of years lived, in the last age group.
-In this age group, $\bar{a_x}$ is unknown, and all the individuals in the age bracket will die (i.e., $q_x=1$).
-Because $d_x/L_x=R_x$ and $d_x=l_x$ in this age group, we can substitute $l_x$ for $d_x$ and write $l_x/L_x=R_x$, implying that $L_x=l_x/R_x$ in this age group.
-
-##Assumptions of the life table
-
-1. The number of births occurring each year is constant.
-2. Deaths are uniformly distributed during the interval after 1 years old.
-3. There is no population growth.
-
-Together, these assumptions imply that the population must be stationary.
-
-**Construction of an abridged (current) life table**
-
-Recall that an abridged life table includes age intervals that are longer than a year.
-In practice, these intervals are often 5 years in length and can be represented by $n_x$, where $x$ is the age at the beginning of the interval.
-The formula for $q_x$ can be generalized to this context, where: $q_x=\frac{n_xR_x}{1+(1-\bar{a_x})n_xR_x}$.
-
-**Derivation of the formula for $q_x$ when the life table is abridged**
-
-This derivation was modified from the Preston demography text [@preston_demography_text].
-Note that in the Preston book $a_x$ represents *person-years* not the proportion of the year, which is why the formulas look different (because we have to write $n_x\bar{a_x}$ in place of Preston's $a_x$).
-
-We start with the simple formula for q_x. As before:
-$q_x=d_x/l_x$.
-
-Now, $L_x = n_x(l_x - d_x) + n_x\bar{a_x}d_x$
-
-$L_x = n_xl_x - n_xd_x + n_x\bar{a_x}d_x$
-
-$l_x = \frac{L_x + n_xd_x - n_x\bar{a_x}d_x}{n_x}$
-
-$l_x = \frac{L_x + d_xn_x(1 - \bar{a_x})}{n_x}$
-
-We can then substitute this expression for $l_x$ into the equation for $q_x$:
-
-$q_x=\frac{d_x}{l_x}$
-
-$q_x=\frac{d_xn_x}{L_x + d_xn_x(1 - \bar{a_x})}$, and dividing numerator and denominator by $L_x$, we get
-
-$q_x=\frac{R_xn_x}{1 + R_xn_x(1 - \bar{a_x})}$.
 
@@
 first.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.White.Male", ]
+
 second.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.Black.Male", ]
@@
 
+ggplot(life.table.blacks, aes(x = Age2, y = q_x)) + geom_point() + geom_point(aes(y = R_x), col="red")
+
 life.table.whites[, c("Age2", "total_deaths5", "Population", "R_x", "q_x", "l_x", "d_x", "L_x", "T_x", "e_x")]
@@
 ```
-
-##References

### 2016-09-22 working_teaching_examples/how_to_abridged_life_table.Rmd
split LE into two files 2

--- 2016-09-20 working_teaching_examples/how_to_abridged_life_table.Rmd
@@
 
-**An example**
-
-This data was used in a previously published paper [@auger2014mortality]. We use it to show how the life expectancy function works.
-
-First, I create the dataset in R:
-```{r create_dataset}
-canada.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90),
-                         "obs.deaths" = c(4054, 625, 445, 622, 2568, 3770, 3434, 3681, 5155, 8449, 13421, 19371, 
-                                          25448, 31465, 37441, 48796, 64763, 72512, 60109, 43812),
-                         "obs.popn" = c(719953, 2835604, 3654546, 4071400, 4464549, 4564679, 4296613, 4237402,
-                                        4532151, 5085092, 5163532, 4597916, 3937984, 3120981, 2323629, 1862971,
-                                        1499907, 997492, 492445, 196109))
-
-quebec.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90), 
-                         "obs.deaths" = c(1057, 164, 116, 176, 659, 956, 1056, 1110, 1498, 2470, 4116, 6054, 8426,
-                                          11175, 13558, 16567, 21349, 22396, 16851,10878),
-                         "obs.popn" = c(215200, 789710, 1008266, 1179423, 1249360, 1261780, 1349857, 1293521,
-                                        1313834, 1551684, 1622835, 1485249, 1288014, 1077253, 796557, 609341, 
-                                        476211, 300506, 134353, 49678))
-```
-
-I then add a column that denotes how many years are in each age bracket:
-```{r add_n_x}
-canada.men$n_x <- 5
-canada.men$n_x[canada.men$ages == 0] <- 1
-canada.men$n_x[canada.men$ages == 1] <- 4
-
-quebec.men$n_x <- 5
-quebec.men$n_x[quebec.men$ages == 0] <- 1
-quebec.men$n_x[quebec.men$ages == 1] <- 4
-
-head(canada.men)
-head(quebec.men)
-```
-
-This source file contains the life expectancy functions used in this project:
-```{r load_functions}
-source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
-```
-
-```{r calculate_life_expectancy}
-life.table.cm <- life.table(data = canada.men, age.groups = "ages", num.ages.in.group = "n_x", death.counts = "obs.deaths", population.counts = "obs.popn")
-
-round(life.table.cm, 2)
-
-life.table.qm <- life.table(data = quebec.men, age.groups = "ages", num.ages.in.group = "n_x", death.counts = "obs.deaths", population.counts = "obs.popn")
-
-round(life.table.qm, 2)
-```
-
-```{r calc_age_decomp}
-age_decomp <- le_age_decomp(life.table.qm, life.table.cm)
-age_decomp
-#check:
-sum(age_decomp$C_x)
-age_decomp$Life_Expectancy_2[age_decomp$Ages ==0] - age_decomp$Life_Expectancy_1[age_decomp$Ages ==0]
-
-#also compare to the decomposition if the data is entered in the reverse order:
-age_decomp2 <- le_age_decomp(life.table.cm, life.table.qm)
-age_decomp2
-#check:
-sum(age_decomp2$C_x)
-age_decomp2$Life_Expectancy_2[age_decomp2$Ages ==0] - age_decomp2$Life_Expectancy_1[age_decomp2$Ages ==0]
-
-```
-
-**A second example**
-
-In this example we use the data that Sam entered into Stata to illustrate the life expectancy concept to me.
-I'll replicate the life table here to ensure that I can arrive at the same results as he does in Stata.
-
-```{r lifetable_cali_complete}
-cali.complete <- data.frame("ages" = c(0:15), "nx" = rep(1,16),
-                         "obs.deaths" = c(1635, 64, 41, 22, 41, 55, 42, 58, 44, 52, 48, 60, 52, 82, 129, 23300), #last number is fictitious
-                         "obs.popn" = c(123342, 111520, 109200, 108749, 105698, 110548, 106857, 112184, 116423,
-                                        132952, 134266, 128938, 125502, 128212, 132775, 143600))
-
-head(cali.complete)
-
-life.table(data = cali.complete, age.groups = "ages", num.ages.in.group = "nx", 
-           death.counts = "obs.deaths", population.counts = "obs.popn")
-
-cali.complete$avelived <- 0.5
-cali.complete$avelived[cali.complete$ages==0] <- 0.1
-
-life.table(data = cali.complete, age.groups = "ages", num.ages.in.group = "nx", ave.prop.lived = "avelived",
-           death.counts = "obs.deaths", population.counts = "obs.popn")
-```
-
-```{r lifetable_cali_abridged, warning=F, message=F}
-cali.complete$group <- cut(cali.complete$ages, breaks=c(0, 1, 5, 10, 15, 20), 
-                           labels = c("<1 years old", "1-4 years old", "5-9 years old", 
-                                      "10-14 years old", "15 or older"), include.lowest = T, right = F)
-
-table(cali.complete$group, cali.complete$ages)
-
-library(dplyr)
-cali.abridged <- cali.complete %>% group_by(group) %>% 
-                                   summarise(nx = sum(nx), ages = first(group), popn = sum(obs.popn), deaths = sum(obs.deaths),
-                                             avelived = first(avelived))
-
-cali.abridged
-
-data.frame(life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
-           death.counts = "deaths", population.counts = "popn"))
-
-data.frame(life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
-           death.counts = "deaths", population.counts = "popn", ave.prop.lived = "avelived"))
-```
-
-```{r alabama_example}
-#load("Data/alabama_only.Rdata")
-load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
-#example data
-alabama.by.strata <- 
-  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
-  arrange(RaceSex, Year2, Age2) %>%
-  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years")) %>%
-  summarise(total_deaths5 = sum(Count_md5),
-            Population = first(Population), 
-            Race2 = first(Race2), 
-            Sex2 = first(Sex2), 
-            StateYearRaceSex = first(StateYearRaceSex),
-            nx = first(nx))
-
-first.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.White.Male", ]
-
-alabama.by.strata.cod <- dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
-                                               arrange(RaceSex, Year2, Age2) %>%
-                                               mutate(total_deaths5 = sum(Count_md5),
-                                                      cod_prop_death5 = Count_md5/total_deaths5)
-
-first.strata.cod <- alabama.by.strata.cod[alabama.by.strata$StateYearRaceSex == "Alabama.1969.White.Male", ]
-
-system.time(first.strata.life.table <- data.frame(life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population")))
-
-first.strata.life.table[, c("Age2", "total_deaths5", "Population", "R_x", "q_x", "l_x", "d_x", "L_x", "T_x", "e_x")]
-
-system.time(
-for(stratum.id in levels(droplevels(alabama.by.strata$StateYearRaceSex))) {
-  assign(paste0("lt.", as.character(stratum.id)), 
-         life.table(data = subset(alabama.by.strata, StateYearRaceSex == stratum.id), 
-                                  age.groups = "Age2", num.ages.in.group = "nx", 
-                                  death.counts = "total_deaths5", population.counts = "Population"))
-}
-)
-```
-
 ##References

### 2016-09-22 Code/Test life expectancy fxs on Alabama.Rmd
improved LE functions and split files for testing

--- 2016-09-22 Code/error checking the LE decomp fxs.Rmd
@@
 ---
-title: "error checking the LE decomp functions"
+title: "Test functions on Alabama data"
 author: "Corinne Riddell"
@@
 
-**An example**
-
-This data was used in a previously published paper [@auger2014mortality]. We use it to show how the life expectancy function works.
-
-First, I create the dataset in R:
-```{r create_dataset}
-canada.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90),
-                         "obs.deaths" = c(4054, 625, 445, 622, 2568, 3770, 3434, 3681, 5155, 8449, 13421, 19371, 
-                                          25448, 31465, 37441, 48796, 64763, 72512, 60109, 43812),
-                         "obs.popn" = c(719953, 2835604, 3654546, 4071400, 4464549, 4564679, 4296613, 4237402,
-                                        4532151, 5085092, 5163532, 4597916, 3937984, 3120981, 2323629, 1862971,
-                                        1499907, 997492, 492445, 196109))
-
-quebec.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90), 
-                         "obs.deaths" = c(1057, 164, 116, 176, 659, 956, 1056, 1110, 1498, 2470, 4116, 6054, 8426,
-                                          11175, 13558, 16567, 21349, 22396, 16851,10878),
-                         "obs.popn" = c(215200, 789710, 1008266, 1179423, 1249360, 1261780, 1349857, 1293521,
-                                        1313834, 1551684, 1622835, 1485249, 1288014, 1077253, 796557, 609341, 
-                                        476211, 300506, 134353, 49678))
-```
-
-I then add a column that denotes how many years are in each age bracket:
-```{r add_n_x}
-canada.men$n_x <- 5
-canada.men$n_x[canada.men$ages == 0] <- 1
-canada.men$n_x[canada.men$ages == 1] <- 4
-
-quebec.men$n_x <- 5
-quebec.men$n_x[quebec.men$ages == 0] <- 1
-quebec.men$n_x[quebec.men$ages == 1] <- 4
-
-head(canada.men)
-head(quebec.men)
-
-cod.props <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteGap/working_teaching_examples/auger_cod_proportions.csv")
-
-```
-
-This source file contains the life expectancy functions used in this project:
-```{r load_functions}
-source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
-```
-
-```{r calculate_life_expectancy}
-life.table.cm <- life.table(data = canada.men, age.groups = "ages", num.ages.in.group = "n_x", death.counts = "obs.deaths", population.counts = "obs.popn")
-
-round(life.table.cm, 2)
-
-life.table.qm <- life.table(data = quebec.men, age.groups = "ages", num.ages.in.group = "n_x", death.counts = "obs.deaths", population.counts = "obs.popn")
-
-round(life.table.qm, 2)
-```
-
-```{r calc_age_decomp}
-age_decomp <- le_age_decomp(life.table.qm, life.table.cm)
-age_decomp
-#check:
-sum(age_decomp$C_x)
-age_decomp$Life_Expectancy_2[age_decomp$Ages ==0] - age_decomp$Life_Expectancy_1[age_decomp$Ages ==0]
-
-#also compare to the decomposition if the data is entered in the reverse order:
-age_decomp2 <- le_age_decomp(life.table.cm, life.table.qm)
-age_decomp2
-#check:
-sum(age_decomp2$C_x)
-age_decomp2$Life_Expectancy_2[age_decomp2$Ages ==0] - age_decomp2$Life_Expectancy_1[age_decomp2$Ages ==0]
-
-```
-
-```{r calc_cod_decomp}
-canada.cod.decomp <- cause_of_death_decomp(life.table1 = life.table.qm, life.table2 = life.table.cm, decomp.table = age_decomp,
-                      cod.table = cod.props,age.colname.cod.table = "Age", COD.colname.cod.table = "cod", 
-                      prop1.colname.cod.table = "prop1", prop2.colname.cod.table = "prop2")
-
-cod.contribution$Cause.of.death <- factor(cod.contribution$Cause.of.death, 
-                                          levels = cod.contribution$Cause.of.death[order(cod.contribution$cod.contribution,
-                                                                                         decreasing = T)])
-
-ggplot(cod.contribution, aes(x = Cause.of.death, y = cod.contribution)) + 
-  geom_point() + 
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  geom_abline(intercept = 0, slope = 0)
-```
-
-**A second example**
-
-In this example we use the data that Sam entered into Stata to illustrate the life expectancy concept to me.
-I'll replicate the life table here to ensure that I can arrive at the same results as he does in Stata.
-
-```{r lifetable_cali_complete}
-cali.complete <- data.frame("ages" = c(0:15), "nx" = rep(1,16),
-                         "obs.deaths" = c(1635, 64, 41, 22, 41, 55, 42, 58, 44, 52, 48, 60, 52, 82, 129, 23300), #last number is fictitious
-                         "obs.popn" = c(123342, 111520, 109200, 108749, 105698, 110548, 106857, 112184, 116423,
-                                        132952, 134266, 128938, 125502, 128212, 132775, 143600))
-
-head(cali.complete)
-
-life.table(data = cali.complete, age.groups = "ages", num.ages.in.group = "nx", 
-           death.counts = "obs.deaths", population.counts = "obs.popn")
-
-cali.complete$avelived <- 0.5
-cali.complete$avelived[cali.complete$ages==0] <- 0.1
-
-life.table(data = cali.complete, age.groups = "ages", num.ages.in.group = "nx", ave.prop.lived = "avelived",
-           death.counts = "obs.deaths", population.counts = "obs.popn")
-```
-
-```{r lifetable_cali_abridged, warning=F, message=F}
-cali.complete$group <- cut(cali.complete$ages, breaks=c(0, 1, 5, 10, 15, 20), 
-                           labels = c("<1 years old", "1-4 years old", "5-9 years old", 
-                                      "10-14 years old", "15 or older"), include.lowest = T, right = F)
-
-table(cali.complete$group, cali.complete$ages)
-
-library(dplyr)
-cali.abridged <- cali.complete %>% group_by(group) %>% 
-                                   summarise(nx = sum(nx), ages = first(group), popn = sum(obs.popn), deaths = sum(obs.deaths),
-                                             avelived = first(avelived))
-
-cali.abridged
-
-data.frame(life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
-           death.counts = "deaths", population.counts = "popn"))
-
-data.frame(life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
-           death.counts = "deaths", population.counts = "popn", ave.prop.lived = "avelived"))
-```
-
 ```{r alabama_example}
-#load("Data/alabama_only.Rdata")
 load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
@@
   geom_point(aes(y = Life_Expectancy_2, col="White"))
+
+alabama_cod_decomp <- cause_of_death_decomp(life.table1 = life.table.blacks, life.table2 = life.table.whites, decomp.table = alabama_age_decomp,
+                      cod.table = cod.tab.alabama.male.1969, "Age2", "COD2", "prop1", "prop2")
 ```

### 2016-09-22 Code/error checking the LE decomp fxs.Rmd
improved LE functions and split files for testing

@@
 
-**An example**
+In this file we apply the functions created to calculate and decompose life expectancy by way of example, and as a way to cross-check against results from published findings (to ensure my functions work properly).
 
-This data was used in a previously published paper [@auger2014mortality]. We use it to show how the life expectancy function works.
+Our functions:
+```{r source_LE_fxs}
+library(dplyr)
+library(ggplot2)
+source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
+```
 
-First, I create the dataset in R:
+**First example**
+
+We use data from a paper published by Auger et al and compare our findings to those found by the authors.
+
+Here is the data need to calculate life expectancy, as provided by the authors in a supplementary appendix (xls file) to their paper:
 ```{r create_dataset}
@@
 head(quebec.men)
-
-cod.props <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteGap/working_teaching_examples/auger_cod_proportions.csv")
-
 ```
 
-This source file contains the life expectancy functions used in this project:
-```{r load_functions}
-source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
+We also need data on the proportion of deaths within each age bracket due to each cause of death.
+```{r data_cod_proportions}
+cod.props <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteGap/working_teaching_examples/auger_cod_proportions.csv")
+names(cod.props)[3] <- "prop.QC"
+names(cod.props)[4] <- "prop.CAD"
+head(cod.props)
 ```
 
+Calculate and display the life tables for the Canada and Quebec data:
 ```{r calculate_life_expectancy}
-life.table.cm <- life.table(data = canada.men, age.groups = "ages", num.ages.in.group = "n_x", death.counts = "obs.deaths", population.counts = "obs.popn")
+life.table.cm <- life.table(data = canada.men, age.groups = "ages", num.ages.in.group = "n_x", 
+                            death.counts = "obs.deaths", population.counts = "obs.popn")
 
@@
 
-life.table.qm <- life.table(data = quebec.men, age.groups = "ages", num.ages.in.group = "n_x", death.counts = "obs.deaths", population.counts = "obs.popn")
+life.table.qm <- life.table(data = quebec.men, age.groups = "ages", num.ages.in.group = "n_x", 
+                            death.counts = "obs.deaths", population.counts = "obs.popn")
 
@@
 
+Life expectancy at birth in Canadian men is `r round(life.table.cm$e_x[1], 2)` vs.
+`r round(life.table.qm$e_x[1], 2)` in Quebec men, for a life expectancy gap of `r life.table.cm$e_x[1] - life.table.qm$e_x[1]` years.
+
+Decompose the life expectancy gap by age:
+
 ```{r calc_age_decomp}
-age_decomp <- le_age_decomp(life.table.qm, life.table.cm)
+age_decomp <- le_age_decomp(life.table1 = life.table.qm, name.lt1 = "Quebec", 
+                            life.table.cm, name.lt2 = "Canada", 
+                            age.groups = "ages")
 age_decomp
-#check:
 sum(age_decomp$C_x) 
-age_decomp$Life_Expectancy_2[age_decomp$Ages ==0] - age_decomp$Life_Expectancy_1[age_decomp$Ages ==0]
+```
 
-#also compare to the decomposition if the data is entered in the reverse order:
-age_decomp2 <- le_age_decomp(life.table.cm, life.table.qm)
+Check that the function still works when the life tables are provided in the reverse order:
+
+```{r calc_age_decomp_reverse}
+age_decomp2 <- le_age_decomp(life.table1 = life.table.cm, name.lt1 = "Canada", 
+                             life.table2 = life.table.qm, name.lt2 = "Quebec", 
+                             age.groups = "ages")
 age_decomp2
-#check:
 sum(age_decomp2$C_x)
-age_decomp2$Life_Expectancy_2[age_decomp2$Ages ==0] - age_decomp2$Life_Expectancy_1[age_decomp2$Ages ==0]
+```
 
+Compare the output of the above two functions -- the contribution of age to the LE gap are slightly different based on the order of operation:
+
+```{r compare decomp_output}
+round(cbind(age_decomp$C_x, age_decomp2$C_x),4)
 ```
@@
 canada.cod.decomp <- cause_of_death_decomp(life.table1 = life.table.qm, life.table2 = life.table.cm, decomp.table = age_decomp,
-                      cod.table = cod.props,age.colname.cod.table = "Age", COD.colname.cod.table = "cod", 
-                      prop1.colname.cod.table = "prop1", prop2.colname.cod.table = "prop2")
+                      cod.table = cod.props, age.colname.cod.table = "Age", COD.colname.cod.table = "cod", 
+                      prop1.colname.cod.table = "prop.QC", prop2.colname.cod.table = "prop.CAD")
+```
 
+Collapse by cause of death to calculate the overall contribution of each cause of death to the gap in life expectnancy:
+```{r }
+cod.contribution <- canada.cod.decomp %>% group_by(Cause.of.death) %>% summarise(cod.contribution = sum(C_xi))
+
+#order by contribution amount for graphing purposes
 cod.contribution$Cause.of.death <- factor(cod.contribution$Cause.of.death, 
@@
 ```
-
-```{r alabama_example}
-#load("Data/alabama_only.Rdata")
-load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
-#example data
-alabama.by.strata <- 
-  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
-  arrange(RaceSex, Year2, Age2) %>%
-  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years")) %>%
-  summarise(total_deaths5 = sum(Count_md5),
-            Population = first(Population), 
-            Race2 = first(Race2), 
-            Sex2 = first(Sex2), 
-            StateYearRaceSex = first(StateYearRaceSex),
-            nx = first(nx))
-
-first.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.White.Male", ]
-
-second.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.Black.Male", ]
-
-alabama.by.strata.cod <- dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
-                                               arrange(RaceSex, Year2, Age2) %>%
-                                               mutate(total_deaths5 = sum(Count_md5),
-                                                      cod_prop_death5 = Count_md5/total_deaths5)
-
-first.strata.cod <- alabama.by.strata.cod[alabama.by.strata.cod$StateYearRaceSex == "Alabama.1969.White.Male", ]
-second.strata.cod <- alabama.by.strata.cod[alabama.by.strata.cod$StateYearRaceSex == "Alabama.1969.Black.Male", ]
-
-cod.tab.alabama.male.1969 <- first.strata.cod %>% ungroup() %>%
-                                                  select(Age2, COD2, cod_prop_death5) %>%
-                                                  rename(prop2 = cod_prop_death5)
-
-cod.tab.alabama.male.1969$prop1 <- second.strata.cod$cod_prop_death5
-cod.tab.alabama.male.1969 <- cod.tab.alabama.male.1969 %>% arrange(COD2, Age2)
-
-ggplot(cod.tab.alabama.male.1969, aes(x=COD2, y=prop1)) + 
-  geom_point(aes(col="White")) + geom_point(aes(y=prop2, col="Black")) +
-  scale_y_continuous(name = "Proportion dying from specified cause") +
-  facet_wrap(~Age2) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-system.time(life.table.whites <- data.frame(life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population")))
-
-life.table.blacks <- data.frame(life.table(data = second.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
-
-ggplot(life.table.blacks, aes(x = Age2, y = q_x)) + geom_point() + geom_point(aes(y = R_x), col="red")
-
-life.table.whites[, c("Age2", "total_deaths5", "Population", "R_x", "q_x", "l_x", "d_x", "L_x", "T_x", "e_x")]
-
-system.time(
-for(stratum.id in levels(droplevels(alabama.by.strata$StateYearRaceSex))) {
-  assign(paste0("lt.", as.character(stratum.id)), 
-         life.table(data = subset(alabama.by.strata, StateYearRaceSex == stratum.id), 
-                                  age.groups = "Age2", num.ages.in.group = "nx", 
-                                  death.counts = "total_deaths5", population.counts = "Population"))
-}
-)
-
-alabama_age_decomp <- le_age_decomp(life.table.blacks, life.table.whites, "Age2")
-alabama_age_decomp
-
-ggplot(alabama_age_decomp, aes(y = C_x, x = Ages)) + geom_point() + scale_y_continuous("Contribution of Age to Gap")
-sum(alabama_age_decomp$C_x)
-life.table.whites$e_x[1] - life.table.blacks$e_x[1]
-
-ggplot(alabama_age_decomp, aes(y = Life_Expectancy_1, x = Ages)) + geom_point(aes(col="Black")) + scale_y_continuous("Life Expectancy (Yrs)") +
-  geom_point(aes(y = Life_Expectancy_2, col="White"))
-```

### 2016-09-23 Code/Test life expectancy fxs on Alabama.Rmd
heat maps looking at mortality and COD in alabama

@@
 
+```{r}
+black.males.alabama <- alabama.by.strata[alabama.by.strata$Race2 == "Black" &
+                                           alabama.by.strata$Sex2 == "Male", ]
+
+males.alabama <- alabama.by.strata[ alabama.by.strata$Sex2 == "Male", ]
+
+ggplot(black.males.alabama, aes(x = Year , y = Age3, z = log(mortality_rate, 10))) + 
+ # stat_contour(geom = "polygon", aes(fill = ..level..), bins = 19) 
+#         geom_tile(aes(fill = mortality_rate)) 
+         stat_contour(bins = 10, aes(colour = ..level..) )
+
+plot_ly(data = black.males.alabama, x = Year, y = Age3, z = log(mortality_rate,10), type = "heatmap")
+
+alabama.by.strata$id <- as.integer(alabama.by.strata$RaceSex)
+
+test <- plot_ly(data = alabama.by.strata, x = Year, y = Age3, z = log(mortality_rate,10),
+        group = RaceSex, type = "heatmap", xaxis = paste0("x", id))
+
+subplot(test)
+
+ggplot(data = alabama.by.strata, aes(x=Year2, y = Age3)) + geom_tile(aes(fill = log(mortality_rate,10))) +
+  facet_grid(Race2 ~ Sex2)
+
+first.strata.cod2 <- alabama.by.strata.cod[alabama.by.strata.cod$RaceSex == "White.Male", ]
+first.strata.cod2$Year3 <- as.numeric(as.character(first.strata.cod2$Year2))
+
+plot_ly(data = subset(first.strata.cod2, COD2 == "Injuries"), x = Year3, y = Age3, z = cod_prop_death5, type = "heatmap")
+
+plot_ly(data = subset(first.strata.cod2, COD2 == "Communicable"), x = Year3, y = Age3, z = cod_prop_death5, type = "heatmap")
+
+plot_ly(data = subset(first.strata.cod2, COD2 == "Cancers"), x = Year3, y = Age3, z = cod_prop_death5, type = "heatmap")
+
+plot_ly(data = subset(first.strata.cod2, COD2 == "Non-communicable"), x = Year3, y = Age3, z =cod_prop_death5, type = "heatmap")
+
+#this is wrong - need to create the dataset...
+alabama.by.strata.cod2$id <- as.integer(alabama.by.strata.cod2$RaceSex)
+subplot(plot_ly(data = subset(alabama.by.strata.cod2, COD2 == "Communicable"), x = Year3, y = Age3, z = cod_prop_death5, type = "heatmap", group = RaceSex, xaxis = paste0("x", id)))
+
+first.strata.cod$id <- as.integer(first.strata.cod$COD2)
+plot_ly(data = first.strata.cod, x = Year, y = Age3, z = log(cod_prop_death5, 10), type = "heatmap", group = COD2,
+        xaxis = paste0("x", id))
+```
+
 ```{r alabama_example}
+library(dplyr)
+library(ggplot2)
+library(plotly)
 load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
 #example data
+
+dat.clean.alabama <- dat.clean.alabama %>% mutate(Age3 = 0*(Age2 == "<1 year") + 1*(Age2 == "1-4 years") + 5*(Age2 == "5-9 years") +
+                10*(Age2 == "10-14 years") + 15*(Age2 == "15-19 years") + 20*(Age2 == "20-24 years") + 
+                25*(Age2 == "25-29 years") + 30*(Age2 == "30-34 years") + 35*(Age2 == "35-39 years") +
+                40*(Age2 == "40-44 years") + 45*(Age2 == "45-49 years") + 50*(Age2 == "50-54 years") + 
+                55*(Age2 == "55-59 years") + 60*(Age2 == "60-64 years") + 65*(Age2 == "65-69 years") +   
+                70*(Age2 == "70-74 years") + 75*(Age2 == "75-79 years") + 80*(Age2 == "80-84 years") + 
+                85*(Age2 == "85+ years"))
+
 alabama.by.strata <- 
@@
             Sex2 = first(Sex2), 
+            Year = first(Year),
+            Age = first(Age), Age3 = first(Age3),
             StateYearRaceSex = first(StateYearRaceSex),
-            nx = first(nx))
+            nx = first(nx)) %>%
+  mutate(mortality_rate = total_deaths5/Population)
 

### 2016-09-23 Code/1_Explore_Data_v2.Rmd

--- 2016-09-15 Code/1_Explore_Data.Rmd
@@
                                                          "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
+         Age3 = 0*(Age2 == "<1 year") + 1*(Age2 == "1-4 years") + 5*(Age2 == "5-9 years") +
+                10*(Age2 == "10-14 years") + 15*(Age2 == "15-19 years") + 20*(Age2 == "20-24 years") + 
+                25*(Age2 == "25-29 years") + 30*(Age2 == "30-34 years") + 35*(Age2 == "35-39 years") +
+                40*(Age2 == "40-44 years") + 45*(Age2 == "45-49 years") + 50*(Age2 == "50-54 years") + 
+                55*(Age2 == "55-59 years") + 60*(Age2 == "60-64 years") + 65*(Age2 == "65-69 years") +   
+                70*(Age2 == "70-74 years") + 75*(Age2 == "75-79 years") + 80*(Age2 == "80-84 years") + 
+                85*(Age2 == "85+ years"),
+         nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years"),
          Sex2 = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
@@
          Year2 = factor(Year, levels = c(0:44), labels = as.character(1969:2013)),
+         Year3 = as.numeric(as.character(Year2)),
          RaceSex = interaction(Race2, Sex2),
@@
   arrange(State2, RaceSex, Year2, Age2) %>%
-  summarise(total_deaths1 = sum(Count_md1), total_deaths5 = sum(Count_md5),
+  summarise(total_deaths1 = sum(Count_md1), 
+            total_deaths5 = sum(Count_md5),
             total_deaths9 = sum(Count_md9), 
-            total_deaths_b9 = sum(Count_b9w1), total_deaths_w9 = sum(Count_b1w9),
-            Population = first(Population), Race2 = first(Race2), Sex2 = first(Sex2), 
-            StateYearRaceSex = first(StateYearRaceSex))
+            total_deaths_b9 = sum(Count_b9w1), 
+            total_deaths_w9 = sum(Count_b1w9),
+            Population = first(Population), 
+            Race2 = first(Race2), 
+            Sex2 = first(Sex2), 
+            StateYearRaceSex = first(StateYearRaceSex),
+            nx = first(nx),
+            Age3 = first(Age3),
+            Year3 = first(Year3))
+
+#return stratum ids with any age-specific population counts of 0.
+#unique(dat2$StateYearRaceSex[dat2$Population == 0])
+non.zero.stratums <- setdiff(levels(dat2$StateYearRaceSex), unique(dat2$StateYearRaceSex[dat2$Population == 0]))
 ```
 
-We implemented the calculations performed in the technical supplement found here [http://www.lho.org.uk/Download/Public/7656/1/tech_supp_3.pdf], also saved in our dropbox.
+```{r}
+source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
 
-Let's make sure we can do the calculation all the way through for a few of the groupings, including a grouping with a stratum with a zero population count.
+list.life.tables <- vector(mode = "list", length = length(non.zero.stratums))
+list.index <- 1
 
-```{r calc_life_expectancy1}
+system.time(
+for(stratum.id in non.zero.stratums[1:10]) {
 
-dat2$numYrsInt <- 5
-dat2$numYrsInt[dat2$Age2 == "<1 year"] <- 1
-dat2$numYrsInt[dat2$Age2 == "1-4 years"] <- 4
-
-dat2 <- dat2 %>% mutate( 
-            Death_Rate1 = ifelse(Population!=0, total_deaths1/Population, NA),
-            Death_Rate5 = ifelse(Population!=0, total_deaths5/Population, NA),
-            Death_Rate9 = ifelse(Population!=0, total_deaths9/Population, NA),
-            Death_Rateb9 = ifelse(Population!=0, total_deaths_b9/Population, NA),
-            Death_Ratew9 = ifelse(Population!=0, total_deaths_w9/Population, NA),
-            Ave_Lived_by_Died = ifelse(Age2=="<1 year", 0.1, 0.5),
-            Prob_Dying1 = ifelse(Population!=0, 
-                                 numYrsInt*Death_Rate1/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Rate1), 
-                                 NA),
-            Prob_Dying5 = ifelse(Population!=0, 
-                                 numYrsInt*Death_Rate5/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Rate5), 
-                                 NA),
-            Prob_Dying9 = ifelse(Population!=0, 
-                                 numYrsInt*Death_Rate9/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Rate9), 
-                                 NA),
-            Prob_Dyingb9 = ifelse(Population!=0, 
-                                 numYrsInt*Death_Rateb9/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Rateb9), 
-                                 NA),
-            Prob_Dyingw9 = ifelse(Population!=0, 
-                                 numYrsInt*Death_Ratew9/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Ratew9), 
-                                 NA),
-            Prob_Surv1 = ifelse(Population!=0, 1 - Prob_Dying1, NA),
-            Prob_Surv5 = ifelse(Population!=0, 1 - Prob_Dying5, NA),
-            Prob_Surv9 = ifelse(Population!=0, 1 - Prob_Dying9, NA),            
-            Prob_Survb9 = ifelse(Population!=0, 1 - Prob_Dyingb9, NA),
-            Prob_Survw9 = ifelse(Population!=0, 1 - Prob_Dyingw9, NA),
-            alive_at_start_int1 = ifelse(Age2 == "<1 year", 1000000, NA),
-            alive_at_start_int5 = ifelse(Age2 == "<1 year", 1000000, NA),
-            alive_at_start_int9 = ifelse(Age2 == "<1 year", 1000000, NA),
-            alive_at_start_intb9 = ifelse(Age2 == "<1 year", 1000000, NA),
-            alive_at_start_intw9 = ifelse(Age2 == "<1 year", 1000000, NA),
-            deaths_during_int1 = ifelse(Age2 == "<1 year", alive_at_start_int1*Prob_Dying1, NA),
-            deaths_during_int5 = ifelse(Age2 == "<1 year", alive_at_start_int5*Prob_Dying5, NA),
-            deaths_during_int9 = ifelse(Age2 == "<1 year", alive_at_start_int9*Prob_Dying9, NA),
-            deaths_during_intb9 = ifelse(Age2 == "<1 year", alive_at_start_intb9*Prob_Dyingb9, NA),
-            deaths_during_intw9 = ifelse(Age2 == "<1 year", alive_at_start_intw9*Prob_Dyingw9, NA))
-```
-
-```{r first_five}
-levels(dat2$StateYearRaceSex)[1:5]
-summary(dat2$Population[dat2$StateYearRaceSex %in% levels(dat2$StateYearRaceSex)[1:5]])
-
-timer1 <- system.time (
-  for(x in levels(dat2$StateYearRaceSex)[1:5]) {
-      for(row in 2:length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) { 
-        dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row - 1]
-
-          if (row != length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) {
-              dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying1[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying5[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying9[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingb9[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingw9[dat2$StateYearRaceSex == x][row]
-          }
-          else { dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] #everyone dies in the 85+ grouping
-                 dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]
-                 dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]
-                 dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]
-                 dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]
-
-          } 
-        }
-
+  list.life.tables[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id), 
+                                               age.groups = "Age2", num.ages.in.group = "nx", 
+                                               death.counts = "total_deaths5", 
+                                               population.counts = "Population")
+  list.index <- list.index + 1
+  print(list.index)
   }
@@
 
-timer1 # 25 seconds for 5 -- guess 13 hrs for all the strata
 ```
 
-```{r calc_pop0}
-View(dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[2789], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
-                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1")])
-timer2 <- system.time (
-  for(x in levels(dat2$StateYearRaceSex)[2789]) {
-      for(row in 2:length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) { 
-        dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row - 1]
-
-          if (row != length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) {
-              dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying1[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying5[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying9[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingb9[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingw9[dat2$StateYearRaceSex == x][row]
-          }
-          else { dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] #everyone dies in the 85+ grouping
-                 dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]
-                 dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]
-                 dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]
-                 dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]
-
-          } 
-        }
-
-  }
-)
-
-timer2
-```
-
-This takes `r time1[3]` seconds to run.
-
-```{r rest_of_life_expectancy_calc}
-dat2$num_PY_int1 <- (dat2$alive_at_start_int1 - dat2$deaths_during_int1)*dat2$numYrsInt + (dat2$deaths_during_int1*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
-dat2$num_PY_int5 <- (dat2$alive_at_start_int5 - dat2$deaths_during_int5)*dat2$numYrsInt + (dat2$deaths_during_int5*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
-dat2$num_PY_int9 <- (dat2$alive_at_start_int9 - dat2$deaths_during_int9)*dat2$numYrsInt + (dat2$deaths_during_int9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
-dat2$num_PY_intb9 <- (dat2$alive_at_start_intb9 - dat2$deaths_during_intb9)*dat2$numYrsInt + (dat2$deaths_during_intb9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
-dat2$num_PY_intw9 <- (dat2$alive_at_start_intw9 - dat2$deaths_during_intw9)*dat2$numYrsInt + (dat2$deaths_during_intw9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
-
-dat2$num_PY_int1[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_int1[dat2$Age2 == "85+ years"])/dat2$Death_Rate1[dat2$Age2 == "85+ years"]
-dat2$num_PY_int5[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_int5[dat2$Age2 == "85+ years"])/dat2$Death_Rate5[dat2$Age2 == "85+ years"]
-dat2$num_PY_int9[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_int9[dat2$Age2 == "85+ years"])/dat2$Death_Rate9[dat2$Age2 == "85+ years"]
-dat2$num_PY_intb9[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_intb9[dat2$Age2 == "85+ years"])/dat2$Death_Rateb9[dat2$Age2 == "85+ years"]
-dat2$num_PY_intw9[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_intw9[dat2$Age2 == "85+ years"])/dat2$Death_Ratew9[dat2$Age2 == "85+ years"]
-
-dat2 <- dat2 %>% group_by(StateYearRaceSex) %>% arrange(desc(Age2)) %>%
-                 mutate(tot_PY_after1 = cumsum(num_PY_int1), tot_PY_after5 = cumsum(num_PY_int5), tot_PY_after9 = cumsum(num_PY_int9),
-                        tot_PY_afterb9 = cumsum(num_PY_intb9),tot_PY_afterw9 = cumsum(num_PY_intw9),
-                        life_exp1 = tot_PY_after1/alive_at_start_int1, life_exp5 = tot_PY_after5/alive_at_start_int5, 
-                        life_exp9 = tot_PY_after9/alive_at_start_int9, 
-                        life_expb9 = tot_PY_afterb9/alive_at_start_intb9, life_expw9 = tot_PY_afterw9/alive_at_start_intw9 
-                )
-
-dat2 <- dat2 %>% group_by(StateYearRaceSex) %>% arrange(Age2)
-```
-
-I compared the following with my calculations in the excel spreadsheet to check the calculations -- all good.
-```{r}
-dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[1], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
-                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1", 
-                                                                       "alive_at_start_int1", "deaths_during_int1", "num_PY_int1", "tot_PY_after1", "life_exp1")]
-
-dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[4], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
-                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1", 
-                                                                       "alive_at_start_int1", "deaths_during_int1", "num_PY_int1", "tot_PY_after1", "life_exp1")]
-
-dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[2789], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
-                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1", 
-                                                                       "alive_at_start_int1", "deaths_during_int1", "num_PY_int1", "tot_PY_after1", "life_exp1")]
-```
-
+###Initial graphs on missing data
 Notes on Figure 1 (below):

### 2016-09-28 shinyapp/app.R
added shiny app to view mortality

@@ -1,0 +1,52 @@
+library(shiny)
+library(plotly)
+library(viridis)
+library(scales)
+#
+#load("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
+load("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/alabama_only.Rdata")
+
+dat.clean.alabama$id_cod <- as.integer(dat.clean.alabama$COD2)
+
+ui1 <- fluidPage("Explore cause of death by age and year in the United States",
+                 
+                 # *Input() functions,
+                 selectInput(inputId = "sex", label = "Select a sex", choices = unique(levels(dat.clean.alabama$Sex2))),
+                 selectInput(inputId = "race", label = "Select a race", choices = unique(levels(dat.clean.alabama$Race2))),
+                 selectInput(inputId = "cod", label = "Select a cause of death", choices = unique(levels(dat.clean.alabama$COD2))),
+                 #                selectInput(inputId = state1, label = "Select a state", choices = dat.clean$State2)
+                 #*Output() functions,
+                 plotlyOutput("heatmapCOD"),
+                 plotlyOutput("heatmapsCOD6")
+                 
+)
+
+#need them to share a scale - 0 - 90% say
+#need the panels to be named
+
+server <- function(input, output) {
+  
+  output$heatmapCOD <- renderPlotly({
+    plot_ly(data = subset(dat.clean.alabama, Race2 == input$race & Sex2 == input$sex & COD2 == input$cod),
+            x = Year3, y = Age3, z = cod_prop_death5, type = "heatmap") %>% 
+      layout(xaxis = list(title = "Year"), yaxis = list(title = "Age (years)"))
+  })
+  
+  output$heatmapsCOD6 <- renderPlotly({  
+    cods <- unique(levels(dat.clean.alabama$COD2))
+#    vals <- unique(dat.clean.alabama$cod_prop_death5)
+#    o <- order(vals, decreasing = FALSE)
+ #   cols <- scales::col_numeric("Blues", domain = NULL)(vals)
+ #   colz <- setNames(data.frame(vals[o], cols[o]), NULL)
+    plots <- lapply(cods, function(cod){
+      plot_ly(subset(dat.clean.alabama, Race2 == input$race & Sex2 == input$sex & COD2 == cod),
[…]

### 2016-09-28 Code/1_Explore_Data_v2.Rmd

@@
          RaceSex = interaction(Race2, Sex2),
+         StateYearSex = interaction(State2, Year2, Sex2),
          StateYearRaceSex = interaction(State2, Year2, Race2, Sex2))
@@
 system.time(
-for(stratum.id in non.zero.stratums[1:10]) {
+for(stratum.id in non.zero.stratums) {
 
@@
   list.index <- list.index + 1
-  print(list.index)
   }

### 2016-09-28 Code/1_Explore_Data_v2.Rmd
more updates

@@
 
+```{r addvar_cod_proportions}
+dat.clean <- dat.clean %>% group_by(State2, RaceSex, Year3, Age3) %>%
+                           arrange(State2, RaceSex, Year3, Age3) %>%
+                           mutate(total_deaths1 = sum(Count_md1),
+                                  total_deaths5 = sum(Count_md5),
+                                  total_deaths9 = sum(Count_md9),
+                                  cod_prop_death1 = Count_md1/total_deaths1, 
+                                  cod_prop_death5 = Count_md5/total_deaths5,
+                                  cod_prop_death9 = Count_md9/total_deaths9)
+```
+
 To calculate life expectancy, we have to aggregate the data across the causes of death.
@@
   arrange(State2, RaceSex, Year2, Age2) %>%
-  summarise(total_deaths1 = sum(Count_md1), 
-            total_deaths5 = sum(Count_md5),
-            total_deaths9 = sum(Count_md9), 
-            total_deaths_b9 = sum(Count_b9w1), 
-            total_deaths_w9 = sum(Count_b1w9),
-            Population = first(Population), 
+  summarise(Population = first(Population), 
             Race2 = first(Race2), 
@@
             Age3 = first(Age3),
-            Year3 = first(Year3))
+            Year3 = first(Year3),
+            total_deaths1 = first(total_deaths1),
+            total_deaths5 = first(total_deaths5),
+            total_deaths9 = first(total_deaths9),
+            cod_prop_death1 = first(cod_prop_death1),
+            cod_prop_death5 = first(cod_prop_death5),
+            cod_prop_death9 = first(cod_prop_death9))
 

### 2016-09-28 Code/Test life expectancy fxs on Alabama.Rmd
more updates

@@
 ---
+
+```{r initialize}
+library(dplyr)
+library(ggplot2)
+library(plotly)
+load("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/alabama_only.Rdata")
+#example data
+```
 
@@
 ```{r alabama_example}
-library(dplyr)
-library(ggplot2)
-library(plotly)
-load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
-#example data
-
-dat.clean.alabama <- dat.clean.alabama %>% mutate(Age3 = 0*(Age2 == "<1 year") + 1*(Age2 == "1-4 years") + 5*(Age2 == "5-9 years") +
-                10*(Age2 == "10-14 years") + 15*(Age2 == "15-19 years") + 20*(Age2 == "20-24 years") + 
-                25*(Age2 == "25-29 years") + 30*(Age2 == "30-34 years") + 35*(Age2 == "35-39 years") +
-                40*(Age2 == "40-44 years") + 45*(Age2 == "45-49 years") + 50*(Age2 == "50-54 years") + 
-                55*(Age2 == "55-59 years") + 60*(Age2 == "60-64 years") + 65*(Age2 == "65-69 years") +   
-                70*(Age2 == "70-74 years") + 75*(Age2 == "75-79 years") + 80*(Age2 == "80-84 years") + 
-                85*(Age2 == "85+ years"))
-
 alabama.by.strata <- 
@@
   arrange(RaceSex, Year2, Age2) %>%
-  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years"))  %>%
   summarise(total_deaths5 = sum(Count_md5),

### 2016-09-28 Code/Test life expectancy fxs on Alabama.Rmd
changes to LE example alabama

@@
 load("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/alabama_only.Rdata")
-#example data
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+```
+
+```{r alabama_example}
+alabama.by.strata <- dat.clean.alabama %>% 
+  filter(COD2 == "Cardiovascular") %>% #to only get one row per year, race, sex, age strata
+  select(total_deaths5, Population, Race2, Sex2, Year, Age, StateYearRaceSex, nx) %>%
+  mutate(mortality_rate = total_deaths5/Population)
+
+first.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.White.Male", ]
+second.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.Black.Male", ]
+
+first.strata.cod <- dat.clean.alabama[dat.clean.alabama$StateYearRaceSex == "Alabama.1969.White.Male", ]
+second.strata.cod <- dat.clean.alabama[dat.clean.alabama$StateYearRaceSex == "Alabama.1969.Black.Male", ]
+
+#this is a table needed for the arriaga cod decomp function
+cod.tab.alabama.male.1969 <- first.strata.cod %>% 
+                             ungroup() %>%
+                             select(Age2, COD2, cod_prop_death5) %>%
+                             rename(prop2 = cod_prop_death5)
+
+cod.tab.alabama.male.1969$prop1 <- second.strata.cod$cod_prop_death5
+cod.tab.alabama.male.1969 <- cod.tab.alabama.male.1969 %>% arrange(COD2, Age2)
+
+ggplot(cod.tab.alabama.male.1969, aes(x=COD2, y=prop1)) + 
+  geom_point(aes(col="White")) + geom_point(aes(y=prop2, col="Black")) +
+  scale_y_continuous(name = "Proportion dying from specified cause") +
+  facet_wrap(~Age2) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+```{r}
+life.table.whites <- data.frame(life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
+
+life.table.blacks <- data.frame(life.table(data = second.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
+
+ggplot(life.table.blacks, aes(x = Age3, y = q_x, col = "q_x: Probability")) + geom_point() + geom_point(aes(y = R_x, col="R_x: Mortality rate")) + scale_y_continuous(name = "Value") + scale_x_continuous(name = "Age") + ggtitle("Examining the relationship between the probability of dying and the mortality rate")
+
+alabama_age_decomp <- le_age_decomp(life.table.blacks, life.table.whites, "Age2")
+alabama_age_decomp
+
+ggplot(alabama_age_decomp, aes(y = C_x, x = Ages)) + geom_point() + scale_y_continuous("Contribution of Age to Gap")
+sum(alabama_age_decomp$C_x)
+life.table.whites$e_x[1] - life.table.blacks$e_x[1]
+
+ggplot(alabama_age_decomp, aes(y = Life_Expectancy_1, x = Ages)) + geom_point(aes(col="Black")) + scale_y_continuous("Life Expectancy (Yrs)") +
+  geom_point(aes(y = Life_Expectancy_2, col="White"))
+
+alabama_cod_decomp <- cause_of_death_decomp(life.table1 = life.table.blacks, life.table2 = life.table.whites, decomp.table = alabama_age_decomp,
+                      cod.table = cod.tab.alabama.male.1969, "Age2", "COD2", "prop1", "prop2")
 ```
@@
 ```
-
-```{r alabama_example}
-alabama.by.strata <- 
-  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
-  arrange(RaceSex, Year2, Age2) %>%
-  summarise(total_deaths5 = sum(Count_md5),
-            Population = first(Population), 
-            Race2 = first(Race2), 
-            Sex2 = first(Sex2), 
-            Year = first(Year),
-            Age = first(Age), Age3 = first(Age3),
-            StateYearRaceSex = first(StateYearRaceSex),
-            nx = first(nx)) %>%
-  mutate(mortality_rate = total_deaths5/Population)
-
-first.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.White.Male", ]
-
-second.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.Black.Male", ]
-
-alabama.by.strata.cod <- dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
-                                               arrange(RaceSex, Year2, Age2) %>%
-                                               mutate(total_deaths5 = sum(Count_md5),
-                                                      cod_prop_death5 = Count_md5/total_deaths5)
-
-first.strata.cod <- alabama.by.strata.cod[alabama.by.strata.cod$StateYearRaceSex == "Alabama.1969.White.Male", ]
-second.strata.cod <- alabama.by.strata.cod[alabama.by.strata.cod$StateYearRaceSex == "Alabama.1969.Black.Male", ]
-
-cod.tab.alabama.male.1969 <- first.strata.cod %>% ungroup() %>%
-                                                  select(Age2, COD2, cod_prop_death5) %>%
-                                                  rename(prop2 = cod_prop_death5)
-
-cod.tab.alabama.male.1969$prop1 <- second.strata.cod$cod_prop_death5
-cod.tab.alabama.male.1969 <- cod.tab.alabama.male.1969 %>% arrange(COD2, Age2)
-
-ggplot(cod.tab.alabama.male.1969, aes(x=COD2, y=prop1)) + 
-  geom_point(aes(col="White")) + geom_point(aes(y=prop2, col="Black")) +
-  scale_y_continuous(name = "Proportion dying from specified cause") +
-  facet_wrap(~Age2) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-system.time(life.table.whites <- data.frame(life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population")))
-
-life.table.blacks <- data.frame(life.table(data = second.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
-
-ggplot(life.table.blacks, aes(x = Age2, y = q_x)) + geom_point() + geom_point(aes(y = R_x), col="red")
-
-life.table.whites[, c("Age2", "total_deaths5", "Population", "R_x", "q_x", "l_x", "d_x", "L_x", "T_x", "e_x")]
-
-system.time(
-for(stratum.id in levels(droplevels(alabama.by.strata$StateYearRaceSex))) {
-  assign(paste0("lt.", as.character(stratum.id)), 
-         life.table(data = subset(alabama.by.strata, StateYearRaceSex == stratum.id), 
-                                  age.groups = "Age2", num.ages.in.group = "nx", 
-                                  death.counts = "total_deaths5", population.counts = "Population"))
-}
-)
-
-alabama_age_decomp <- le_age_decomp(life.table.blacks, life.table.whites, "Age2")
-alabama_age_decomp
-
-ggplot(alabama_age_decomp, aes(y = C_x, x = Ages)) + geom_point() + scale_y_continuous("Contribution of Age to Gap")
-sum(alabama_age_decomp$C_x)
-life.table.whites$e_x[1] - life.table.blacks$e_x[1]
-
-ggplot(alabama_age_decomp, aes(y = Life_Expectancy_1, x = Ages)) + geom_point(aes(col="Black")) + scale_y_continuous("Life Expectancy (Yrs)") +
-  geom_point(aes(y = Life_Expectancy_2, col="White"))
-
-alabama_cod_decomp <- cause_of_death_decomp(life.table1 = life.table.blacks, life.table2 = life.table.whites, decomp.table = alabama_age_decomp,
-                      cod.table = cod.tab.alabama.male.1969, "Age2", "COD2", "prop1", "prop2")
-```

### 2016-09-28 Code/Test life expectancy fxs on Alabama.Rmd
more changes to the alabama example

@@
 cod.tab.alabama.male.1969 <- cod.tab.alabama.male.1969 %>% arrange(COD2, Age2)
+```
 
+```{r exploratory graphs}
 ggplot(cod.tab.alabama.male.1969, aes(x=COD2, y=prop1)) + 
@@
   facet_wrap(~Age2) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+ggplot(life.table.blacks, aes(x = Age3, y = q_x, col = "q_x: Probability")) + geom_point() + geom_point(aes(y = R_x, col="R_x: Mortality rate")) + scale_y_continuous(name = "Value") + scale_x_continuous(name = "Age") + ggtitle("Examining the relationship between the probability of dying and the mortality rate")
 ```
@@
 ```{r}
-life.table.whites <- data.frame(life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
+life.table.whites <- data.frame(life.table(data = first.strata, num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
 
-life.table.blacks <- data.frame(life.table(data = second.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
+life.table.blacks <- data.frame(life.table(data = second.strata, num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
 
-ggplot(life.table.blacks, aes(x = Age3, y = q_x, col = "q_x: Probability")) + geom_point() + geom_point(aes(y = R_x, col="R_x: Mortality rate")) + scale_y_continuous(name = "Value") + scale_x_continuous(name = "Age") + ggtitle("Examining the relationship between the probability of dying and the mortality rate")
-
-alabama_age_decomp <- le_age_decomp(life.table.blacks, life.table.whites, "Age2")
+alabama_age_decomp <- le_age_decomp(life.table1 = life.table.blacks, name.lt1 = "Blacks", 
+                                    life.table2 = life.table.whites, name.lt2 = "Whites", age.groups = "Age3")
 alabama_age_decomp

### 2016-09-28 Code/life_expectancy_functions.R

@@ -4,7 +4,7 @@
 
 #change names to something like col.name.death.counts or death.counts.in.data
 
-life.table <- function(data, age.groups, num.ages.in.group, death.counts, population.counts, ave.prop.lived = NA){
+life.table <- function(data, num.ages.in.group, death.counts, population.counts, ave.prop.lived = NA){
   
   data["R_x"] <- data[death.counts]/data[population.counts] #mortality rates
   
@@ -45,6 +45,8 @@ life.table <- function(data, age.groups, num.ages.in.group, death.counts, popula
 
 #takes the life table output from the first function -- need two life tables to perform the comparison
 #relies on the names of the columns kept as-is in the life tables -- user should not change the column names!
+#name.lt1 and name.lt2 are used to create variable names so these shouldn't have spaces in them and they can be 
+#made up (they don't have to replicate the name of the life table object)
 le_age_decomp <- function(life.table1, name.lt1 = 1, life.table2 = 2, name.lt2, age.groups) {
 
   life.table2["accumulated.lived.after"] <- c(unlist(life.table2[2:dim(life.table2)[1], "T_x"]), 0) #T_(x+n) in Auger's formula on pg 576 (step 1)

### 2016-09-28 Code/Test life expectancy fxs on Alabama.Rmd

@@
 
-```{r exploratory graphs}
+```{r exploratory_graphs}
 ggplot(cod.tab.alabama.male.1969, aes(x=COD2, y=prop1)) + 
@@
 
-ggplot(life.table.blacks, aes(x = Age3, y = q_x, col = "q_x: Probability")) + geom_point() + geom_point(aes(y = R_x, col="R_x: Mortality rate")) + scale_y_continuous(name = "Value") + scale_x_continuous(name = "Age") + ggtitle("Examining the relationship between the probability of dying and the mortality rate")
+ggplot(life.table.blacks, aes(x = Age3, y = q_x, col = "q_x: Probability")) + 
+  geom_point() + geom_point(aes(y = R_x, col="R_x: Mortality rate")) + 
+  scale_y_continuous(name = "Value") + 
+  scale_x_continuous(name = "Age") + 
+  ggtitle("Examining the relationship between the probability of dying and the mortality rate")
 ```
 
-```{r}
+```{r life-tables}
 life.table.whites <- data.frame(life.table(data = first.strata, num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
@@
 life.table.blacks <- data.frame(life.table(data = second.strata, num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))
+```
 
+The life expectancy at birth for white males in Alabama in 1969 is `r life.table.whites$e_x[1]` years vs.
+`r life.table.blacks$e_x[1]` for black males, for a gap of `r life.table.whites$e_x[1] - life.table.blacks$e_x[1]` years.
+
+```{r age-decomp}
 alabama_age_decomp <- le_age_decomp(life.table1 = life.table.blacks, name.lt1 = "Blacks", 
@@
 
-ggplot(alabama_age_decomp, aes(y = C_x, x = Ages)) + geom_point() + scale_y_continuous("Contribution of Age to Gap")
 sum(alabama_age_decomp$C_x)
-life.table.whites$e_x[1] - life.table.blacks$e_x[1]
 
-ggplot(alabama_age_decomp, aes(y = Life_Expectancy_1, x = Ages)) + geom_point(aes(col="Black")) + scale_y_continuous("Life Expectancy (Yrs)") +
-  geom_point(aes(y = Life_Expectancy_2, col="White"))
+ggplot(alabama_age_decomp, aes(y = C_x, x = Ages)) + 
+  geom_point() + 
+  scale_y_continuous("Contribution of Age to Gap") +
+  geom_abline(intercept = 0, slope = 0)
 
-alabama_cod_decomp <- cause_of_death_decomp(life.table1 = life.table.blacks, life.table2 = life.table.whites, decomp.table = alabama_age_decomp,
+ggplot(alabama_age_decomp, aes(y = LE_Blacks, x = Ages)) + 
+  geom_point(aes(col="Black")) + 
+  scale_y_continuous("Life Expectancy (Yrs)") +
+  geom_point(aes(y = LE_Whites, col="White")) +
+  ggtitle("Life expectancy by age in black and white males in Alabama in 1969")
+```
+
+```{r cod-decomp}
+alabama_cod_decomp <- cause_of_death_decomp(life.table1 = life.table.blacks, 
+                                            life.table2 = life.table.whites, 
+                                            decomp.table = alabama_age_decomp,
                                             cod.table = cod.tab.alabama.male.1969, "Age2", "COD2", "prop1", "prop2")
+
+ggplot(alabama_cod_decomp, aes(y = C_xi, x = COD2)) + 
+  geom_point() + 
+  scale_y_continuous("Contribution of Age to Gap") +
+  geom_abline(intercept = 0, slope = 0) +
+  facet_wrap(~Age2) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
 ```

### 2016-09-29 Code/Test life expectancy fxs on Alabama.Rmd
playing with the plots

@@
 
-ggplot(alabama_age_decomp, aes(y = C_x, x = Ages)) + 
-  geom_point() + 
-  scale_y_continuous("Contribution of Age to Gap") +
-  geom_abline(intercept = 0, slope = 0)
-
 ggplot(alabama_age_decomp, aes(y = LE_Blacks, x = Ages)) + 
@@
   ggtitle("Life expectancy by age in black and white males in Alabama in 1969")
+
+ggplot(alabama_age_decomp, aes(y = C_x, x = Ages)) + 
+  geom_point() + 
+  scale_y_continuous("Contribution of Age to Gap") +
+  geom_abline(intercept = 0, slope = 0)
 ```
@@
   geom_point() + 
-  scale_y_continuous("Contribution of Age to Gap") +
-  geom_abline(intercept = 0, slope = 0) +
-  facet_wrap(~Age2) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  scale_y_continuous("Contribution of cause of death to Gap") +
+  geom_abline(intercept = 0, slope = 0) + facet_wrap(~ Age2) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
 ```
@@
 
-test <- plot_ly(data = alabama.by.strata, x = Year, y = Age3, z = log(mortality_rate,10),
-        group = RaceSex, type = "heatmap", xaxis = paste0("x", id))
+#test <- plot_ly(data = alabama.by.strata, x = Year, y = Age3, z = log(mortality_rate,10),
+#        group = RaceSex, xaxis = paste0("x", id), type = "heatmap")
 
-subplot(test)
+#subplot(test)
 
-ggplot(data = alabama.by.strata, aes(x=Year2, y = Age3)) + geom_tile(aes(fill = log(mortality_rate,10))) +
-  facet_grid(Race2 ~ Sex2)
+#ggplot(data = alabama.by.strata, aes(x=Year3, y = Age3)) + geom_tile(aes(fill = log(mortality_rate,10))) +
+#  facet_grid(Race2 ~ Sex2)
 
-first.strata.cod2 <- alabama.by.strata.cod[alabama.by.strata.cod$RaceSex == "White.Male", ]
-first.strata.cod2$Year3 <- as.numeric(as.character(first.strata.cod2$Year2))
+#first.strata.cod2 <- alabama.by.strata.cod[alabama.by.strata.cod$RaceSex == "White.Male", ]
+first.strata.cod2 <- dat.clean.alabama[dat.clean.alabama$RaceSex == "White.Male", ]
 
@@
 
+first.strata.cod2$id <- as.integer(first.strata.cod2$COD2)
+plot_ly(data = first.strata.cod2, x = Year, y = Age3, z = cod_prop_death5, type = "heatmap", group = COD2,
+        xaxis = paste0("x", id))
+
 #this is wrong - need to create the dataset...
-alabama.by.strata.cod2$id <- as.integer(alabama.by.strata.cod2$RaceSex)
+#alabama.by.strata.cod2$id <- as.integer(alabama.by.strata.cod2$RaceSex)
 subplot(plot_ly(data = subset(alabama.by.strata.cod2, COD2 == "Communicable"), x = Year3, y = Age3, z = cod_prop_death5, type = "heatmap", group = RaceSex, xaxis = paste0("x", id)))
 
-first.strata.cod$id <- as.integer(first.strata.cod$COD2)
-plot_ly(data = first.strata.cod, x = Year, y = Age3, z = log(cod_prop_death5, 10), type = "heatmap", group = COD2,
-        xaxis = paste0("x", id))
 ```

### 2016-09-30 Code/1_Explore_Data_v2.Rmd
new app

@@
          Year3 = as.numeric(as.character(Year2)),
-         RaceSex = interaction(Race2, Sex2),
+         RaceSex = interaction(Race2, Sex2, sep = " "),
          StateYearSex = interaction(State2, Year2, Sex2),
@@
 
+Add the total population (across age brackets) to the dat2 dataset
 ```{r}
+dat3 <- dat2 %>% ungroup() %>% 
+  group_by(StateYearRaceSex) %>%
+  summarise(Pop_across_age = sum(Population))
+
+dat2 <- merge(dat2, dat3, by = "StateYearRaceSex")
+```
+
+In those strata that do not have any age-specific population counts of 0, create a life table and store it in a list.
+
+```{r calculate_life_expectancy}
 source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
@@
 )
+
+```
+
+Add the life expectancy at birth to the data file
+
+```{r add_le_to_dataset}
+le <- rep(NA, length(non.zero.stratums))
+
+for (i in 1:length(non.zero.stratums)) {
+  le[i] <- list.life.tables[[i]]$e_x[1]
+
+}
+
+les <- data.frame(life.expectancy.birth = le, StateYearRaceSex = non.zero.stratums)
+
+dat2 <- merge(dat2, les, by = "StateYearRaceSex", all.x =T)
 
@@
 ```
+
+```{r}
+save(dat.clean, dat2, file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
+```

### 2016-09-30 Code/2_Examine_LifeExpectancy.Rmd
new app

@@
 
-dat3 <- 
-dat2 %>% group_by(StateYearRaceSex) %>%
-  arrange(Age2) %>%  
-  summarise(State2 = first(State2), Year2 = first(Year2), Race2 = first(Race2), Sex2 = first(Sex2),
-            Pop_across_age = sum(Population, na.rm = "always"), Pop_has_NA = sum(Population), 
-            LE_birth1 = first(life_exp1), LE_birth5 = first(life_exp5), LE_birth9 = first(life_exp9), 
-            LE_birthb9 = first(life_expb9), LE_birthw9 = first(life_expw9))
-
 dat3$grouping <- interaction(dat3$State2, dat3$Race2, dat3$Sex2)
-dat3$Year <- as.numeric(as.character(dat3$Year2))
 

### 2016-09-30 Code/1_Explore_Data_v2.Rmd
more plots added to the app

@@
 
+make a new dataset that contains information on the gap
+
+```{r}
+Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
+                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth) %>%
+                  rename(Pop_black = Pop_across_age,  le.birth.black = life.expectancy.birth)
+
+White <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
+                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth) %>%
+                  rename(Pop_white = Pop_across_age,  le.birth.white = life.expectancy.birth)
+
+BlackWhite <- merge(Black, White, by = c("State2", "Year3", "Sex2"))
+
+BlackWhite <- BlackWhite %>% mutate(WBgap = le.birth.white - le.birth.black)
+
+```
+
 ###Initial graphs on missing data
@@
 ```{r}
-save(dat.clean, dat2, file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
+save(dat.clean, dat2, BlackWhite, file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
 ```

### 2016-10-05 shinyapp2/app.R
added all age decomp tables

@@ -11,13 +11,15 @@ ui1 <- fluidPage("Explore cause of death by age and year in the United States",
                  #selectInput(inputId = "race", label = "Select a race", choices = unique(levels(dat.clean.alabama$Race2))),
                  #selectInput(inputId = "cod", label = "Select a cause of death", choices = unique(levels(dat.clean.alabama$COD2))),
                  selectInput(inputId = "state", label = "Select a state", choices = unique(levels(dat.clean$State2))),
+                 selectInput(inputId = "year1", label = "Select first comparison year", choices = unique(dat.clean$Year3)),
+                 selectInput(inputId = "year2", label = "Select second comparison year", choices = unique(dat.clean$Year3)),
                  #*Output() functions,
                  #plotlyOutput("heatmapCOD"),
                  #plotlyOutput("heatmapsCOD6")
                  plotlyOutput("population_trend"),
 #add the new output object for our LE graphs -- see my notes.
-                 plotlyOutput("life_expectancy")
-#                 plotlyOutput("life_expectancy_gap")
+                 plotlyOutput("life_expectancy"),
+                 plotlyOutput("age_comp")
                  )
 
 server <- function(input, output) {
@@ -41,8 +43,9 @@ server <- function(input, output) {
 
     p1 <- ggplotly(ggplot(subset(BlackWhite, State2 == input$state), aes(x = Year3, y = le.birth.black)) + 
                      geom_ribbon(aes(ymin = le.birth.black, ymax = le.birth.white), alpha = 0.3) +
-                     geom_line(aes(col = Sex2, lty = "Black")) + 
-                     geom_line(aes(y = le.birth.white, col = Sex2, lty = "White")) +
+                  
+                     geom_line(aes(y = le.birth.white, col = Sex2), lty = 1) +
+                     geom_line(aes(col = Sex2), lty = 2) + 
                      facet_wrap(~Sex2) +
                      scale_x_continuous(name = "Year") + scale_y_continuous(name = "Life expectancy at birth (years)") + 
                      theme_minimal() + theme(legend.title=element_blank()))

### 2016-10-05 Code/1_Explore_Data_v2.Rmd

@@
 ```{r calculate_life_expectancy}
-source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
 
@@
 
-make a new dataset that contains information on the gap
+Make a new dataset that contains information on the gap:
 
-```{r}
+```{r create_dataset_BlackWhite}
 Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
@@
 BlackWhite <- BlackWhite %>% mutate(WBgap = le.birth.white - le.birth.black)
+
+```
+
+Decompose the gap by age. That is, for each year-state-sex combination, compare the life table for blacks to whites and decompose the gap in life expectancy at birth by age.
+
+```{r calculate_age_decomp}
+
+#create list of stratum.ids for each race-sex combination
+for (rs in unique(levels(dat.clean$RaceSex))) {
+  assign(paste0(gsub(" ", "", rs), ".stratum.ids"), 
+      dat.clean %>% filter(dat.clean$RaceSex == rs) %>%
+        select(StateYearRaceSex, StateYearSex) %>%
+        group_by(StateYearRaceSex, StateYearSex) %>%
+        filter(row_number() == 1)# %>%
+      #  rename(paste0(gsub(" ", "", rs), ".ids") = StateYearRaceSex)
+         )
+
+}
+
+#merge the black-white lists
+Female.ids <- merge(WhiteFemale.stratum.ids, BlackFemale.stratum.ids, by = "StateYearSex")
+Female.ids <- Female.ids %>% rename(White.StateYearRaceSex = StateYearRaceSex.x, Black.StateYearRaceSex = StateYearRaceSex.y)
+Female.ids$Sex2 <- "Female"
+
+Male.ids <- merge(WhiteMale.stratum.ids, BlackMale.stratum.ids, by = "StateYearSex")
+Male.ids <- Male.ids %>% rename(White.StateYearRaceSex = StateYearRaceSex.x, Black.StateYearRaceSex = StateYearRaceSex.y)
+Male.ids$Sex2 <- "Male"
+
+paired.ids <- rbind(Male.ids, Female.ids)
+
+list.age.decomp.tables <- vector(mode = "list", length = dim(paired.ids)[1])
+
+for(i in 1:dim(paired.ids)[1]) {
+  white.index <- which(non.zero.stratums == paired.ids[i, 2])
+  black.index <- which(non.zero.stratums == paired.ids[i, 3])
+
+  if(length(white.index) != 0 & length(black.index) != 0) {
+    list.age.decomp.tables[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]), 
+                                                 name.lt1 = "Black", 
+                                                 life.table2 = as.data.frame(list.life.tables[[white.index]]), 
+                                                 name.lt2 = "White", age.groups = "Age3")
+  } else {
+    list.age.decomp.tables[[i]] <- "Couldn't compare life tables"
+  }
+  print(i)
+}
 

### 2016-10-05 Code/1_Explore_Data_v2.Rmd
added the agg decomp graphs to shiny app

@@
       dat.clean %>% filter(dat.clean$RaceSex == rs) %>%
-        select(StateYearRaceSex, StateYearSex) %>%
+        select(StateYearRaceSex, StateYearSex, Sex2, State2, Year3) %>%
         group_by(StateYearRaceSex, StateYearSex) %>%
@@
 #merge the black-white lists
-Female.ids <- merge(WhiteFemale.stratum.ids, BlackFemale.stratum.ids, by = "StateYearSex")
-Female.ids <- Female.ids %>% rename(White.StateYearRaceSex = StateYearRaceSex.x, Black.StateYearRaceSex = StateYearRaceSex.y)
-Female.ids$Sex2 <- "Female"
+Female.ids <- merge(WhiteFemale.stratum.ids, BlackFemale.stratum.ids[, -c(3:5)], 
+                    by = "StateYearSex", suffixes = c("White", "Black"))
+head(Female.ids)
 
-Male.ids <- merge(WhiteMale.stratum.ids, BlackMale.stratum.ids, by = "StateYearSex")
-Male.ids <- Male.ids %>% rename(White.StateYearRaceSex = StateYearRaceSex.x, Black.StateYearRaceSex = StateYearRaceSex.y)
-Male.ids$Sex2 <- "Male"
+Male.ids <- merge(WhiteMale.stratum.ids, BlackMale.stratum.ids[, -c(3:5)],
+                  by = "StateYearSex", suffixes = c("White", "Black"))
+head(Male.ids)
 
@@
   }
-  print(i)
 }
-
 ```
@@
 ```{r}
-save(dat.clean, dat2, BlackWhite, file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
+save(dat.clean, dat2, BlackWhite, paired.ids, list.age.decomp.tables,
+     file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
 ```

### 2016-10-05 shinyapp2/app.R

@@ -19,7 +19,8 @@ ui1 <- fluidPage("Explore cause of death by age and year in the United States",
                  plotlyOutput("population_trend"),
 #add the new output object for our LE graphs -- see my notes.
                  plotlyOutput("life_expectancy"),
-                 plotlyOutput("age_comp")
+                 plotOutput("age_comp1"),
+                 plotOutput("age_comp2")
                  )
 
 server <- function(input, output) {
@@ -56,11 +57,28 @@ server <- function(input, output) {
     subplot(p1, p2, shareX = T, nrows = 2, titleY = T)
   })
   
-#  output$life_expectancy_gap <- renderPlotly({
-#  ggplotly(ggplot(subset(BlackWhite, State2 == input$state), aes(x = Year3, y = WBgap)) + geom_line(aes(col = Sex2)) +
-#      facet_wrap(~Sex2) + scale_x_continuous(name = "Year") + scale_y_continuous(name = "Life expectancy gap (years)") + 
-#      theme_minimal() + theme(legend.title=element_blank()))
-#})
+  output$age_comp1 <- renderPlot({
+   # index <- 
+  #  dat <- 
+    ggplot(as.data.frame(list.age.decomp.tables[[which(paired.ids$State2 ==  input$state & paired.ids$Year3 == input$year1 & paired.ids$Sex2 == "Male")]]), aes(Ages, C_x)) + 
+      geom_line() + 
+      geom_abline(intercept = 0, slope = 0) + 
+      geom_point() +
+      ylab("Contribution of age group to LE Gap (years)") + theme_minimal() + xlab("Age") + 
+      ggtitle(paste0("Age decomposition for males in ", input$state, " in ", input$year1))
+  })
+
+  output$age_comp2 <- renderPlot({
+    # index <- 
+    #  dat <- 
+    ggplot(as.data.frame(list.age.decomp.tables[[which(paired.ids$State2 ==  input$state & paired.ids$Year3 == input$year2 & paired.ids$Sex2 == "Male")]]), aes(Ages, C_x)) + 
+      geom_line() + 
+      geom_abline(intercept = 0, slope = 0) + 
+      geom_point() +
+      ylab("Contribution of age group to LE Gap (years)") + theme_minimal() + xlab("Age") +
+      ggtitle(paste0("Age decomposition for males in ", input$state, " in ", input$year2))
+  })    
+    
     }
 
 shinyApp(ui = ui1, server = server)

### 2016-10-18 Code/1_Explore_Data_v2.Rmd
added to the age decomp fx and fixed some coding errors

@@
 
+rm(les)
 ```
@@
 ```{r create_dataset_BlackWhite}
+#this dataset contains the black-white data within the same row (wide format) for each year-state-sex strata
+
 Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
@@
 
-```{r calculate_age_decomp}
+```{r create_paired_ids_dataset}
+#paired.ids will contain the index names that are sent to the age decomposition function in the next code chunk
 
-#create list of stratum.ids for each race-sex combination
+#first, create list of stratum.ids for each race-sex combination
 for (rs in unique(levels(dat.clean$RaceSex))) {
@@
          group_by(StateYearRaceSex, StateYearSex) %>%
-        filter(row_number() == 1)# %>%
-      #  rename(paste0(gsub(" ", "", rs), ".ids") = StateYearRaceSex)
+         filter(row_number() == 1)
          )
@@
 
-#merge the black-white lists
+#then, merge the black-white lists
 Female.ids <- merge(WhiteFemale.stratum.ids, BlackFemale.stratum.ids[, -c(3:5)], 
@@
 paired.ids <- rbind(Male.ids, Female.ids)
+```
+
+```{r calculate_age_decomp}
+#runs in about a minute
 
@@
 for(i in 1:dim(paired.ids)[1]) {
-  white.index <- which(non.zero.stratums == paired.ids[i, 2])
-  black.index <- which(non.zero.stratums == paired.ids[i, 3])
+  white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
+  black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])
 

### 2016-10-18 Code/life_expectancy_functions.R

@@ -65,6 +65,8 @@ le_age_decomp <- function(life.table1, name.lt1 = 1, life.table2 = 2, name.lt2, 
     ((life.table2["accumulated.lived.after"]/life.table2[1, "l_x"])*
      ((life.table1["l_x"]/life.table2["l_x"]) - (life.table1["num.alive.next.interval"]/life.table2["num.alive.next.interval"]))
      )
+  
+  decomp.table["adds_to_gap"] <- ifelse(decomp.table["C_x"] > 0, T, F)
 
   return(decomp.table)
 }

### 2016-10-18 shinyapp2/app.R

@@ -71,14 +71,20 @@ server <- function(input, output) {
   output$age_comp2 <- renderPlot({
     # index <- 
     #  dat <- 
-    ggplot(as.data.frame(list.age.decomp.tables[[which(paired.ids$State2 ==  input$state & paired.ids$Year3 == input$year2 & paired.ids$Sex2 == "Male")]]), aes(Ages, C_x)) + 
-      geom_segment(aes(xend = Ages, y = 0, yend = C_x), lwd=4) +
-      geom_abline(intercept = 0, slope = 0) + 
-      geom_point() +
-      ylab("Contribution of age group to LE Gap (years)") + theme_minimal() + xlab("Age") +
-      ggtitle(paste0("Age decomposition for males in ", input$state, " in ", input$year2))
-  })    
-    
+   
+  ggplot(as.data.frame(list.age.decomp.tables[[which(paired.ids$State2 ==  input$state & 
+                                                       paired.ids$Year3 == input$year2 & 
+                                                       paired.ids$Sex2 == "Male")]]), 
+         aes(y = Ages, x = C_x)) + 
+    geom_segment(aes(xend = 0, yend = Ages), lwd=8) +
+   # geom_abline(intercept = 0, slope = 0) + 
+  #  ylab("Contribution of age group to LE Gap (years)") + 
+    theme_minimal() +
+    scale_y_reverse() +  
+    xlab("Loss of gain in life expectancy in years") +
+    ggtitle(paste0("Contribution of age grouping to the black-white life expectancy gap for males in ", input$state, " in ", input$year2))
+      
+})     
     }
 
 shinyApp(ui = ui1, server = server)

### 2016-10-25 Code/1_Explore_Data_v2.Rmd

@@
 
+```{r calculate_cod_decomp}
+#still need to convince myself this works
+list.cod.decomp.tables <- vector(mode = "list", length = dim(paired.ids)[1])
+
+for(i in 1:dim(paired.ids)[1]) {
+  white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
+  black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])
+
+  if(length(white.index) != 0 & length(black.index) != 0) {
+    ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
+                ungroup() %>%
+                select(Age2, COD2, cod_prop_death5) %>%
+                rename(prop2 = cod_prop_death5)
+
+    ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
+                ungroup() %>%
+                select(Age2, COD2, cod_prop_death5) %>%
+                rename(prop1 = cod_prop_death5)
+    ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2"))
+
+    list.cod.decomp.tables[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]), 
+                                                 life.table2 = as.data.frame(list.life.tables[[white.index]]), 
+                                                 decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
+                                                 cod.table = ct.both, "Age2", "COD2", "prop1", "prop2")
+  } else {
+    list.cod.decomp.tables[[i]] <- "Couldn't compare life tables"
+  }  
+  print(i)
+}
+
+```
+
 ###Initial graphs on missing data

### 2016-10-25 Code/1_Explore_Data_v2.Rmd
still error checking this

@@
 
+test1 <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]), 
+                                                 life.table2 = as.data.frame(list.life.tables[[white.index]]), 
+                                                 decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
+                                                 cod.table = ct.both, "Age2", "COD2", "prop1", "prop2")
+
+test2 <- cause_of_death_decomp(life.table2 = as.data.frame(list.life.tables[[black.index]]), 
+                                                 life.table1 = as.data.frame(list.life.tables[[white.index]]), 
+                                                 decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
+                                                 cod.table = ct.both, "Age2", "COD2", "prop1", "prop2")
+
+age_contr1 <- test1 %>% ungroup() %>% group_by(Ages) %>% summarise(age_contribution = sum(C_xi))
+sum(age_contr1$age_contribution)
+age_contr2 <- test2 %>% ungroup() %>% group_by(Ages) %>% summarise(age_contribution = sum(C_xi))
+sum(age_contr2$age_contribution)
 ```

### 2016-10-25 Code/error checking the LE decomp fxs.Rmd
fixed error in cod decomp fx

@@
 library(ggplot2)
-source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
 ```
@@
 ```{r data_cod_proportions}
-cod.props <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteGap/working_teaching_examples/auger_cod_proportions.csv")
+cod.props <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/working_teaching_examples/auger_cod_proportions.csv")
 names(cod.props)[3] <- "prop.QC"
@@
 ```{r calculate_life_expectancy}
-life.table.cm <- life.table(data = canada.men, age.groups = "ages", num.ages.in.group = "n_x", 
+life.table.cm <- life.table(data = canada.men, num.ages.in.group = "n_x", 
                             death.counts = "obs.deaths", population.counts = "obs.popn")
@@
 
-life.table.qm <- life.table(data = quebec.men, age.groups = "ages", num.ages.in.group = "n_x", 
+life.table.qm <- life.table(data = quebec.men, num.ages.in.group = "n_x", 
                             death.counts = "obs.deaths", population.counts = "obs.popn")
@@
 age_decomp <- le_age_decomp(life.table1 = life.table.qm, name.lt1 = "Quebec", 
-                            life.table.cm, name.lt2 = "Canada", 
+                            life.table2 = life.table.cm, name.lt2 = "Canada", 
                             age.groups = "ages")
@@
 ```{r calc_cod_decomp}
-canada.cod.decomp <- cause_of_death_decomp(life.table1 = life.table.qm, life.table2 = life.table.cm, decomp.table = age_decomp,
-                      cod.table = cod.props, age.colname.cod.table = "Age", COD.colname.cod.table = "cod", 
-                      prop1.colname.cod.table = "prop.QC", prop2.colname.cod.table = "prop.CAD")
+canada.cod.decomp <- cause_of_death_decomp(life.table1 = life.table.qm, 
+                                           life.table2 = life.table.cm, 
+                                           decomp.table = age_decomp,
+                                           cod.table = cod.props, 
+                                           age.colname.cod.table = "Age", 
+                                           COD.colname.cod.table = "cod", 
+                                           prop1.colname.cod.table = "prop.QC", 
+                                           prop2.colname.cod.table = "prop.CAD")
 ```

### 2016-10-25 Code/error checking the LE decomp fxs.Rmd

@@
 
-```{r compare decomp_output}
+```{r compare_decomp_output}
 round(cbind(age_decomp$C_x, age_decomp2$C_x),4)
@@
 ```{r }
-cod.contribution <- canada.cod.decomp %>% group_by(Cause.of.death) %>% summarise(cod.contribution = sum(C_xi))
+cod.contribution <- canada.cod.decomp %>% 
+  group_by(Cause.of.death) %>% 
+  summarise(cod.contribution = sum(C_xi))
 
@@
   geom_abline(intercept = 0, slope = 0)
+```
+
+Collapse by age and confirm these line up with the marginal age decomposition:
+```{r }
+age.contribution <- canada.cod.decomp %>% group_by(Ages) %>% summarise(age.contribution = sum(C_xi))
+
+age.contribution
 ```
@@
 
-life.table(data = cali.complete, age.groups = "ages", num.ages.in.group = "nx", 
+life.table(data = cali.complete, num.ages.in.group = "nx", 
            death.counts = "obs.deaths", population.counts = "obs.popn")
@@
 
-life.table(data = cali.complete, age.groups = "ages", num.ages.in.group = "nx", ave.prop.lived = "avelived",
+life.table(data = cali.complete, num.ages.in.group = "nx", ave.prop.lived = "avelived",
            death.counts = "obs.deaths", population.counts = "obs.popn")
@@
 
-library(dplyr)
 cali.abridged <- cali.complete %>% group_by(group) %>% 
@@
 
-data.frame(life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
+data.frame(life.table(data = cali.abridged, num.ages.in.group = "nx", 
            death.counts = "deaths", population.counts = "popn"))
 
-data.frame(life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
+data.frame(life.table(data = cali.abridged, num.ages.in.group = "nx", 
            death.counts = "deaths", population.counts = "popn", ave.prop.lived = "avelived"))

### 2016-10-26 Code/1_Explore_Data_v2.Rmd
more cod decomp changes

@@
 library(tidyr)
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
 ```
@@
 
-Add the life expectancy at birth to the data file
+Add the life expectancy at birth to the data file:
 
@@
 }
+
+i <- 1
+white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
+black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])
+
+    ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
+                ungroup() %>%
+                select(Age2, COD2, cod_prop_death5) %>%
+                rename(prop.white = cod_prop_death5)
+
+    ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
+                ungroup() %>%
+                select(Age2, COD2, cod_prop_death5) %>%
+                rename(prop.black = cod_prop_death5)
+    ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2"))
 

### 2016-10-26 Code/error checking the LE decomp fxs.Rmd
more cod decomp changes
fixed error in how cod table was sorted

@@
 ```{r data_cod_proportions}
-cod.props <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/working_teaching_examples/auger_cod_proportions.csv")
+cod.props <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/working_teaching_examples/auger_cod_proportions.csv")
 names(cod.props)[3] <- "prop.QC"

### 2016-10-26 Code/life_expectancy_functions.R

@@ -73,10 +73,11 @@ le_age_decomp <- function(life.table1, name.lt1 = 1, life.table2, name.lt2 = 2, 
 
 
 #cause of death table must be organized in a very specific way
+#it needs to be sorted by COD, and then by age (within COD)
 cause_of_death_decomp <- function(life.table1, life.table2, decomp.table, 
                                   cod.table, age.colname.cod.table, COD.colname.cod.table, 
                                   prop1.colname.cod.table, prop2.colname.cod.table) {
-  C_x <- decomp.table[["C_x"]]
+ R_
   prop2 <- cod.table[[prop2.colname.cod.table]]
   R_x2 <- life.table2[["R_x"]]
   prop1 <- cod.table[[prop1.colname.cod.table]]

### 2016-10-26 Code/1_Explore_Data_v2.Rmd

@@
     ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
+    ct.both <- ct.both %>% arrange(COD2)
 
@@
                                decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
-                                                 cod.table = ct.both, "Age2", "COD2", "prop1", "prop2")
+                               cod.table = ct.both, 
+                               age.colname.cod.table = "Age2", 
+                               COD.colname.cod.table = "COD2", 
+                               prop1.colname.cod.table = "prop.black", 
+                               prop2.colname.cod.table = "prop.white")
 
@@
 age_contr1 <- test1 %>% ungroup() %>% group_by(Ages) %>% summarise(age_contribution = sum(C_xi))
-sum(age_contr1$age_contribution)
+age_contr1
+list.age.decomp.tables[[i]]
+#compare -- see that there is something wrong with the age labels in one or the other... 
+#all the same numbers but some of them are in a strange order.
+
 age_contr2 <- test2 %>% ungroup() %>% group_by(Ages) %>% summarise(age_contribution = sum(C_xi))

### 2016-10-26 Code/life_expectancy_functions.R
removed unwanted line of code

@@ -77,7 +77,7 @@ le_age_decomp <- function(life.table1, name.lt1 = 1, life.table2, name.lt2 = 2, 
 cause_of_death_decomp <- function(life.table1, life.table2, decomp.table, 
                                   cod.table, age.colname.cod.table, COD.colname.cod.table, 
                                   prop1.colname.cod.table, prop2.colname.cod.table) {
- R_
+
   prop2 <- cod.table[[prop2.colname.cod.table]]
   R_x2 <- life.table2[["R_x"]]
   prop1 <- cod.table[[prop1.colname.cod.table]]

### 2016-10-26 Code/1_Explore_Data_v2.Rmd

@@
                 ungroup() %>%
-                select(Age2, COD2, cod_prop_death5) %>%
+                select(Age2, Age3, COD2, cod_prop_death5) %>%
                 rename(prop2 = cod_prop_death5)
@@
     ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
-    ct.both <- ct.both %>% arrange(COD2)
+    ct.both <- ct.both %>% arrange(COD2, Age2)
+    View(ct.both)
 

### 2016-10-26 Code/1_Explore_Data_v2.Rmd
ran cod decomp

@@
 ```{r calculate_cod_decomp}
-#still need to convince myself this works
+#started at 12:01pm - guessing it will take 20 mins
 list.cod.decomp.tables <- vector(mode = "list", length = dim(paired.ids)[1])
@@
                 ungroup() %>%
-                select(Age2, Age3, COD2, cod_prop_death5) %>%
-                rename(prop2 = cod_prop_death5)
+                select(Age2, COD2, cod_prop_death5) %>%
+                rename(prop.white = cod_prop_death5)
 
@@
                 select(Age2, COD2, cod_prop_death5) %>%
-                rename(prop1 = cod_prop_death5)
+                rename(prop.black = cod_prop_death5)
     ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
+    ct.both <- ct.both %>% arrange(COD2, Age2)
 
@@
                                                          decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
-                                                 cod.table = ct.both, "Age2", "COD2", "prop1", "prop2")
+                                                         cod.table = ct.both, 
+                                                         age.colname.cod.table = "Age2", 
+                                                         COD.colname.cod.table = "COD2", 
+                                                         prop1.colname.cod.table = "prop.black", 
+                                                         prop2.colname.cod.table = "prop.white")
   } else {
@@
 
-test2 <- cause_of_death_decomp(life.table2 = as.data.frame(list.life.tables[[black.index]]), 
-                                                 life.table1 = as.data.frame(list.life.tables[[white.index]]), 
-                                                 decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
-                                                 cod.table = ct.both, "Age2", "COD2", "prop1", "prop2")
-
 age_contr1 <- test1 %>% ungroup() %>% group_by(Ages) %>% summarise(age_contribution = sum(C_xi))
@@
 list.age.decomp.tables[[i]]
-#compare -- see that there is something wrong with the age labels in one or the other... 
-#all the same numbers but some of them are in a strange order.
-
-age_contr2 <- test2 %>% ungroup() %>% group_by(Ages) %>% summarise(age_contribution = sum(C_xi))
-sum(age_contr2$age_contribution)
 ```

### 2016-10-26 Code/life_expectancy_functions.R

@@ -77,7 +77,8 @@ le_age_decomp <- function(life.table1, name.lt1 = 1, life.table2, name.lt2 = 2, 
 cause_of_death_decomp <- function(life.table1, life.table2, decomp.table, 
                                   cod.table, age.colname.cod.table, COD.colname.cod.table, 
                                   prop1.colname.cod.table, prop2.colname.cod.table) {
-
+  
+  C_x <- decomp.table[["C_x"]]
   prop2 <- cod.table[[prop2.colname.cod.table]]
   R_x2 <- life.table2[["R_x"]]
   prop1 <- cod.table[[prop1.colname.cod.table]]

### 2016-10-26 Code/1_Explore_Data_v2.Rmd
prepping for cod graph

@@
 ```{r calculate_cod_decomp}
-#started at 12:01pm - guessing it will take 20 mins
+#takes about 20 mins
 list.cod.decomp.tables <- vector(mode = "list", length = dim(paired.ids)[1])
@@
 }
+```
 
-i <- 1
-white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
-black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])
+```{r}
+first <- list.cod.decomp.tables[[1]]
+first$sign <- ifelse(first$C_xi < 0, "negative", "positive")
+first$group <- interaction(first$Ages, first$sign, sep = ".")
 
-    ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
-                ungroup() %>%
-                select(Age2, COD2, cod_prop_death5) %>%
-                rename(prop.white = cod_prop_death5)
+first <- first %>% 
+  arrange(Ages, sign, Cause.of.death) %>%
+  group_by(group) %>%
+  mutate(lag = lag(C_xi), one = 1, 
+         num = cumsum(one),
+         length = n(),
+         start = ifelse(is.na(lag)==T, 0, NA),
+         finish = ifelse(is.na(lag)==T, C_xi, NA))
+View(first)
 
-    ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
-                ungroup() %>%
-                select(Age2, COD2, cod_prop_death5) %>%
-                rename(prop.black = cod_prop_death5)
-    ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
-    ct.both <- ct.both %>% arrange(COD2, Age2)
-    View(ct.both)
+updated <- NULL
+for (g in unique(first$group)) {
+  subset <- first[first$group == g, ]
 
-test1 <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
-                               life.table2 = as.data.frame(list.life.tables[[white.index]]), 
-                               decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
-                               cod.table = ct.both, 
-                               age.colname.cod.table = "Age2", 
-                               COD.colname.cod.table = "COD2", 
-                               prop1.colname.cod.table = "prop.black", 
-                               prop2.colname.cod.table = "prop.white")
+  if(dim(subset)[1] > 1){
+      for (i in 2:dim(subset)[1]) {
+        subset$start[i] <- subset$finish[i - 1]
+        subset$finish[i] <- subset$start[i] + subset$C_xi[i]
+      }
+  }
 
-age_contr1 <- test1 %>% ungroup() %>% group_by(Ages) %>% summarise(age_contribution = sum(C_xi))
-age_contr1
-list.age.decomp.tables[[i]]
+  updated <- rbind(updated, subset)
+}
+updated 
 ```

### 2016-10-26 Code/1_Explore_Data_v2.Rmd

@@
          num = cumsum(one),
-         length = n(),
          start = ifelse(is.na(lag)==T, 0, NA),
@@
 }
-updated 
+View(updated)
+
+ggplotly(ggplot(updated, aes(y = Ages, x = start)) + 
+           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
+           xlab("Contribution to life expectancy gap") +
+           ylab("Age group\n") + theme_minimal() +
+           geom_vline(xintercept = 0))
 ```

### 2016-10-26 Code/1_Explore_Data_v2.Rmd

@@
 library(tidyr)
+library(plotly)
 source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
@@
            geom_vline(xintercept = 0))
+
+#add total column for graphing purposes
+total <- updated %>% ungroup() %>%
+  group_by(Ages) %>%
+  summarise(total = sum(C_xi))
+
+total2 <- dplyr::bind_rows(first, total)
+levels(total2$Cause.of.death) <- c(levels(total2$Cause.of.death), "All causes combined")
+total2$Cause.of.death[is.na(total2$total)==F] <- "All causes combined"
+
+#need to add a outlined segment for the ttoal group...
 ```

### 2016-10-27 Code/1_Explore_Data_v2.Rmd

--- 2016-10-26 Code/1_Explore_Data_v2.Rmd
@@
 
-```{r}
-first <- list.cod.decomp.tables[[1]]
-first$sign <- ifelse(first$C_xi < 0, "negative", "positive")
-first$group <- interaction(first$Ages, first$sign, sep = ".")
+```{r  update_decomp_for_plotting}
+#update the decomposition table for plotting purposes
+for(i in 1:dim(paired.ids)[1]) {
+    if(list.cod.decomp.tables[[i]][1] != "Couldn't compare life tables") { 
+        list.cod.decomp.tables[[i]] <- 
+                              make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables[[i]], 
+                              age.groups = "Ages", 
+                              cause.of.death = "Cause.of.death")  
+    } 
+  print(i)
+}  
+```
 
-first <- first %>% 
-  arrange(Ages, sign, Cause.of.death) %>%
-  group_by(group) %>%
-  mutate(lag = lag(C_xi), one = 1, 
-         num = cumsum(one),
-         length = n(),
-         start = ifelse(is.na(lag)==T, 0, NA),
-         finish = ifelse(is.na(lag)==T, C_xi, NA))
-View(first)
-
-updated <- NULL
-for (g in unique(first$group)) {
-  subset <- first[first$group == g, ]
-
-  if(dim(subset)[1] > 1){
-      for (i in 2:dim(subset)[1]) {
-        subset$start[i] <- subset$finish[i - 1]
-        subset$finish[i] <- subset$start[i] + subset$C_xi[i]
-      }
-  }
-
-  updated <- rbind(updated, subset)
-}
-updated 
+```{r pretty_cod}
+ggplotly(ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
+           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
+           xlab("Contribution to life expectancy gap") +
+           ylab("Age group\n") + theme_minimal() +
+           geom_vline(xintercept = 0))
 ```
@@
 ```{r}
-save(dat.clean, dat2, BlackWhite, paired.ids, list.age.decomp.tables,
+save(dat.clean, dat2, BlackWhite, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
      file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")

### 2016-10-27 shinyapp3/app.R
updated app to have cod plots

@@ -1,0 +1,158 @@
+library(shiny)
+library(plotly)
+library(viridis)
+library(scales)
+library(shinythemes)
+load(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
+
+ui1 <- fluidPage(#theme = shinytheme("cosmo"),
+                 #shinythemes::themeSelector(), 
+                 titlePanel("Explore the black-white life expectancy gap in the United States"),
+                 
+                 sidebarLayout(
+                   sidebarPanel(width = 4,
+                     selectInput(inputId = "state", label = "Select a state", width = 200, choices = unique(levels(dat.clean$State2))),
+                     strong("Select years to compare"),
+                     tags$style(type="text/css", ".form-group.shiny-input-container{ display: inline-block } strong{ display: block !important  } img{ margin-bottom: 30px }"),
+                  
+                     selectInput(inputId = "year1", label = NA, choices = unique(dat.clean$Year3), width = 100),
+                     selectInput(inputId = "year2", label = NA, choices = unique(dat.clean$Year3), selected = 2013, width = 100),
+                     strong("Legend\n"),
+                     img(src="legend.png"),
+                     strong("Population size over time"),
+                     plotlyOutput("population_trend")
+                      ),
+                   mainPanel(
+                     
+                     #add the new output object for our LE graphs -- see my notes.
+                     plotlyOutput("life_expectancy"),
+                     plotlyOutput("male_age_comp1"),
+                     plotlyOutput("male_age_comp2"),
+                     plotOutput("female_age_comp1"),
+                     plotOutput("female_age_comp2")                   )
+                 )
+
+                 )
+
+server <- function(input, output) {
+
+  output$population_trend <- renderPlotly({
+    p <- ggplotly(
+      ggplot(subset(dat2, State2 == input$state & Age3 == 0), aes(y = Pop_across_age, x = Year3)) + 
+      geom_line(aes(col = Sex2, lty = Race2)) +
+      scale_color_manual(values = c("#67a9cf", "#ef8a62")) +
+      ylab("Population Size") + 
+      scale_y_continuous(label = comma) + 
+      scale_x_continuous(name = "Year") + 
[…]

### 2016-10-28 Code/1_Explore_Data_v2.Rmd

@@
 ```{r calculate_cod_decomp}
+#stopped here - need to figure this out
+#if the count of deaths == 0 for all causes of death within an age band then the COD proportion is NA
+#this messes up the cod decomposition. What to do?
+#example - no black people dying between the ages of 5-9 of anything. but have whites dying in that age group so their
+#COD proportions are okay.
+
 #takes about 20 mins

### 2016-10-28 Code/1_Explore_Data_v2.Rmd

@@
 ```
+
+```{r investigate_problem}
+
+list.cod.decomp.tables[[which(paired.ids$StateYearRaceSexBlack == "Washington DC.2013.Black.Male")]]
+#index == 2115 fyi
+
+View(dat.clean[dat.clean$State2 == "Washington DC" & dat.clean$Year3 == 2013 & dat.clean$Sex2 == "Male",])
+#see the problem is that white males at 1-4 years have no deaths and prop deaths due to each cod == NaN.
+
+list.age.decomp.tables[[which(paired.ids$StateYearRaceSexBlack == "Washington DC.2013.Black.Male")]]
+
+View(list.life.tables[[which(non.zero.stratums == "Washington DC.2013.White.Male")]])
+
+```

### 2016-11-04 Code/BlackPopulationSizeHeatMap.R
heatmap to explore pop size

@@ -1,0 +1,36 @@
+library(dplyr)
+library(ggplot2)
+
+#population heat maps
+#goal: to explore population counts < 5000 to try and understand when the LE and SE calculations are meaningful.
+
+#load the datasets
+load(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
+
+#Pop_across_age is a variable in the dat2 dataset that listed the overall population for each age.
+#create a dataset with one row per state, year, sex, race strata
+
+max.pop <- max(dat2$Pop_across_age)
+
+pop_data <- dat2 %>% 
+  select(State2, Year3, Race2, Sex2, Age3, Pop_across_age) %>%
+  filter(Age3 == 0) %>% #to remove duplicate rows by age 
+  mutate(Pop_category = cut(Pop_across_age, breaks = c(0, 1000, 2500, 5000, 10000, 50000, max.pop + 1), include.lowest = T))
+
+levels(pop_data$Pop_category) <- c("0 - 1,000","1,001 - 2,500","2,501 - 5,000","5,001 - 10,000","10,001 - 50,000", "50,001+")
+
+plot <- ggplot(data = subset(pop_data, Race2 == "Black"), aes(x = Year3, y = State2, fill = Pop_category)) + 
+  geom_tile(color = "white", size = 0.025) + 
+  geom_tile() + 
+  facet_grid(. ~ Sex2) +
+  scale_fill_discrete(h = c(0, 200), guide = guide_legend(title = "Population size")) + 
+  xlab("Year") + ylab("State") +
+  ggtitle("Black population size by state over time") +
+  theme_tufte(base_family="Helvetica") +
+  theme(axis.ticks=element_blank()) 
+
+plot
+
+ggsave(filename = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/output/black_pop_heatmap", 
+       plot = last_plot(), device = NULL,
+       length = 10, width = 8, units = "in", dpi = 100)

### 2016-11-04 Code/1_Explore_Data_v2.Rmd
more changes

@@
 ```{r fill_in_missing_counts}
+##NEED TO UPDATE THIS SO THAT CAN'T IMPUTAT COUNT > POP SIZE.
+##FOLLOW UPDATE WITH RE-RUNNING THE MODELS.
 dat.clean <- dat.clean %>% mutate(Count_md1 = ifelse(is.na(Count)==T, 1, Count),
@@
 dat2 <- merge(dat2, dat3, by = "StateYearRaceSex")
+
+dat2 <- dat2 %>% arrange(State2, Year3, Race2, Sex2, Age3)
 ```
@@
 BlackWhite <- BlackWhite %>% mutate(WBgap = le.birth.white - le.birth.black)
-
 ```

### 2016-11-04 Code/1_Explore_Data_v2.Rmd

@@
 
-Some of the rows have Population==0, and we explore that and record below.
-These mostly correspond to rows with Unknown age groups that we cannot use in our analysis so I remove them.
-This leaves 654 rows with Populations of 0.
+```{r unknown_age_group}
+#there is an unknown age group for each combo of state-year-race-sex-age:
+table(dat$Age2)
+
+#Count of death is always equal to 0 for unknown age groups
+table(dat$Age, dat$Count==0)
+
+#remove these data from the data we analyze:
+dat.clean <- dat[dat$Age2 != "Unknown", ]
+```
+
+##Population counts of 0
+
+It is important to explore how many cells have population counts of zero because these break the Bayesian crank, and also currently breaks the life expectancy function.
 
 ```{r explore_dataset}
-table(dat$Population==0, useNA = "always")
-#55734 cells have Populations of 0.
-
-table(dat$Count[dat$Population==0]==0, useNA = "always")
-#Of these, all have Counts of 0, except for 26 with counts of NA
-
-table(dat$Age2 == "Unknown", dat$Population == 0, useNA = "always")
-#The vast majority of cells with populations of 0 have unknown age groups
-
-#remove this data from the data we analyze:
-dat.clean <- dat[dat$Age2 != "Unknown", ]
-
 table(dat.clean$Population == 0, useNA = "always")
-#654 cells have Populations of 0 now.
+#654 cells have Populations of 0.
 
 table(dat.clean$Count[dat.clean$Population == 0] == 0, useNA = "always")
-summary(dat.clean$Count[dat.clean$Population == 0])
-#628 are equal to 0 and 26 are equal to NA
+#Of these, all have Counts of 0, except for 26 with counts of NA
 
@@
 
-#now the Counts should only equal NA when the true count is between 1 and 9.
-
 summary(dat.clean$Crude_Rate[dat.clean$Population == 0])
-table(dat.clean$Crude_Rate[dat.clean$Population==0], useNA = "always")
 #whenever the population is 0, the Crude_Rate is equal to NA.
@@
 #States with at least one strata of 0
-na.pops <- dat.clean %>% select(State2, Population) %>% filter(Population == 0)
+na.pops <- dat.clean %>% 
+  select(State2, Population) %>% 
+  filter(Population == 0)
+
 unique(na.pops$State2)
@@
 **Missingness of Count (of deaths)**
-If there are between 1 and 9 deaths in the strata, this information is surpressed and coded as missing.
+If there are between 1 and 9 deaths in the strata, this information is 'coarsened' and coded as NA.
 Later in this document we investigate trends in supression/missingness to understand how it differs according to state, age, race, and sex.
@@
 
-- Set all missing counts to 1, 5, and 9.
-- Set all missing counts to 1 for Whites and 9 for Blacks and vice versa.
+- Set all missing counts to min({1, 5, and 9}, population_size).
 
 ```{r fill_in_missing_counts}
-##NEED TO UPDATE THIS SO THAT CAN'T IMPUTAT COUNT > POP SIZE.
-##FOLLOW UPDATE WITH RE-RUNNING THE MODELS.
-dat.clean <- dat.clean %>% mutate(Count_md1 = ifelse(is.na(Count)==T, 1, Count),
-                                  Count_md5 = ifelse(is.na(Count)==T, 5, Count),
-                                  Count_md9 = ifelse(is.na(Count)==T, 9, Count)
+dat.clean <- dat.clean %>% mutate(Count_md1 = ifelse(is.na(Count)==T, pmin(1, Population), Count),
+                                  Count_md5 = ifelse(is.na(Count)==T, pmin(5, Population), Count),
+                                  Count_md9 = ifelse(is.na(Count)==T, pmin(9, Population), Count)
                                   )
-
-dat.clean$Count_b9w1 <- dat.clean$Count
-dat.clean$Count_b9w1[is.na(dat.clean$Count) & dat.clean$Race2 == "White"] <- 1
-dat.clean$Count_b9w1[is.na(dat.clean$Count) & dat.clean$Race2 == "Black"] <- 9
-
-dat.clean$Count_b1w9 <- dat.clean$Count
-dat.clean$Count_b1w9[is.na(dat.clean$Count) & dat.clean$Race2 == "White"] <- 9
-dat.clean$Count_b1w9[is.na(dat.clean$Count) & dat.clean$Race2 == "Black"] <- 1
 
@@
 dat.clean$Count9_grouped <- cut(dat.clean$Count_md9, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
-dat.clean$Count_b9w1_grouped <- cut(dat.clean$Count_b9w1, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
-dat.clean$Count_b1w9_grouped <- cut(dat.clean$Count_b1w9, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
 
@@
 
-table(dat.clean$Count_grouped, dat.clean$Count_b9w1_grouped, dat.clean$Race2, useNA = "always")
-table(dat.clean$Count_grouped, dat.clean$Count_b1w9_grouped, dat.clean$Race2, useNA = "always")
-
-dat.clean <- dat.clean[ , !names(dat.clean) %in% c("Count_grouped", "Count1_grouped", "Count5_grouped", "Count9_grouped", 
-                                                   "Count_b9w1_grouped", "Count_b1w9_grouped")]
+dat.clean <- dat.clean[ , !names(dat.clean) %in% c("Count_grouped", "Count1_grouped", "Count5_grouped", "Count9_grouped")]
 ```

### 2016-11-04 Code/1_Explore_Data_v2.Rmd
added heatmaps to explore NA cods

@@
                                   cod_prop_death9 = Count_md9/total_deaths9)
+
+length(dat.clean$total_deaths5[dat.clean$total_deaths5 == 0])
+summary(dat.clean$cod_prop_death5[dat.clean$total_deaths5 == 0])
+#76998 strata with 0 deaths from any cause (12833 when divided by 6 -- the number of causes)
+#7% of strata have no deaths from any cause.
 ```
 
-To calculate life expectancy, we have to aggregate the data across the causes of death.
+To calculate life expectancy, we have to aggregate the data across causes of death.
 

### 2016-11-04 Code/1_Explore_Data_v2.Rmd

@@
 
-Add the total population (across age brackets) to the dat2 dataset
-```{r}
+Add the total population (across age brackets) to the dat2 dataset.
+
+```{r create_var_pop_across_age}
 dat3 <- dat2 %>% ungroup() %>% 
@@
 ```{r calculate_life_expectancy}
-source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
-
 list.life.tables <- vector(mode = "list", length = length(non.zero.stratums))
@@
   }  
-  print(i)
+  #print(i)
 }
@@
     } 
-  print(i)
+  #print(i)
 }  
@@
 
-```{r save_workspace3}
+```{r save_workspace}
 save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
@@
 
-```{r}
+```{r save_specified_files}
 save(dat.clean, dat2, BlackWhite, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
@@
 ```
-
-```{r investigate_problem}
-
-list.cod.decomp.tables[[which(paired.ids$StateYearRaceSexBlack == "Washington DC.2013.Black.Male")]]
-#index == 2115 fyi
-
-View(dat.clean[dat.clean$State2 == "Washington DC" & dat.clean$Year3 == 2013 & dat.clean$Sex2 == "Male",])
-#see the problem is that white males at 1-4 years have no deaths and prop deaths due to each cod == NaN.
-
-list.age.decomp.tables[[which(paired.ids$StateYearRaceSexBlack == "Washington DC.2013.Black.Male")]]
-
-View(list.life.tables[[which(non.zero.stratums == "Washington DC.2013.White.Male")]])
-
-```

### 2016-11-04 Code/1_Explore_Data_v2.Rmd
small cleanup

@@
   list.life.tables[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id), 
-                                               age.groups = "Age2", num.ages.in.group = "nx", 
+                                               num.ages.in.group = "nx", 
                                                death.counts = "total_deaths5", 

### 2016-11-04 Code/1_Explore_Data_v2.Rmd

@@
   assign(paste0(gsub(" ", "", rs), ".stratum.ids"), 
-         dat.clean %>% filter(dat.clean$RaceSex == rs) %>%
+         dat.clean %>% filter(RaceSex == rs) %>%
          select(StateYearRaceSex, StateYearSex, Sex2, State2, Year3) %>%

### 2016-11-04 Code/1_Explore_Data_v2.Rmd

@@
   assign(paste0(gsub(" ", "", rs), ".stratum.ids"), 
-         dat.clean %>% filter(RaceSex == rs) %>%
+         dat.clean %>% 
+           ungroup() %>%
+           filter(RaceSex == rs) %>%
            select(StateYearRaceSex, StateYearSex, Sex2, State2, Year3) %>%

### 2016-11-04 shinyapp3/app.R
added radio button for plots

@@ -17,6 +17,7 @@ ui1 <- fluidPage(#theme = shinytheme("cosmo"),
                   
                      selectInput(inputId = "year1", label = NA, choices = unique(dat.clean$Year3), width = 100),
                      selectInput(inputId = "year2", label = NA, choices = unique(dat.clean$Year3), selected = 2013, width = 100),
+                     
                      strong("Legend\n"),
                      img(src="legend.png"),
                      strong("Population size over time"),
@@ -26,6 +27,7 @@ ui1 <- fluidPage(#theme = shinytheme("cosmo"),
                      
                      #add the new output object for our LE graphs -- see my notes.
                      plotlyOutput("life_expectancy"),
+                     radioButtons(inputId = "contribution_type", label = "Display contribution in", inline = T, choices = c("Years", "Proportion (%)"), selected = "Years"),
                      plotlyOutput("male_age_comp1"),
                      plotlyOutput("male_age_comp2"),
                      plotOutput("female_age_comp1"),

### 2016-11-04 Code/1_Explore_Data_v2.Rmd
fix old coding error

@@
 
-table1.whitemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "White.Male"]), 
-                                           dat$Age2[dat$RaceSex == "White.Male"]),2)*100,2)
-table1.blackmale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "Black.Male"]), 
-                                           dat$Age2[dat$RaceSex == "Black.Male"]),2)*100,2)
-table1.whitefemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "White.Female"]), 
-                                           dat$Age2[dat$RaceSex == "White.Female"]),2)*100,2)
-table1.blackfemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "Black.Female"]), 
-                                           dat$Age2[dat$RaceSex == "Black.Female"]),2)*100,2)
+table1.whitemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "White Male"]), 
+                                           dat$Age2[dat$RaceSex == "White Male"]),2)*100,2)
+table1.blackmale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "Black Male"]), 
+                                           dat$Age2[dat$RaceSex == "Black Male"]),2)*100,2)
+table1.whitefemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "White Female"]), 
+                                           dat$Age2[dat$RaceSex == "White Female"]),2)*100,2)
+table1.blackfemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "Black Female"]), 
+                                           dat$Age2[dat$RaceSex == "Black Female"]),2)*100,2)
 
@@
 ```{r zero_data, fig.width = 10, fig.height = 5, echo=F}
-table2.whitemale <- round(prop.table(table(dat$Count[dat$RaceSex == "White.Male"]==0, 
-                                       dat$Age2[dat$RaceSex == "White.Male"]),2)*100,2)
-table2.blackmale <- round(prop.table(table(dat$Count[dat$RaceSex == "Black.Male"]==0, 
-                                       dat$Age2[dat$RaceSex == "Black.Male"]),2)*100,2)
-table2.whitefemale <- round(prop.table(table(dat$Count[dat$RaceSex == "White.Female"]==0, 
-                                       dat$Age2[dat$RaceSex == "White.Female"]),2)*100,2)
-table2.blackfemale <- round(prop.table(table(dat$Count[dat$RaceSex == "Black.Female"]==0, 
-                                       dat$Age2[dat$RaceSex == "Black.Female"]),2)*100,2)
+table2.whitemale <- round(prop.table(table(dat$Count[dat$RaceSex == "White Male"]==0, 
+                                       dat$Age2[dat$RaceSex == "White Male"]),2)*100,2)
+table2.blackmale <- round(prop.table(table(dat$Count[dat$RaceSex == "Black Male"]==0, 
+                                       dat$Age2[dat$RaceSex == "Black Male"]),2)*100,2)
+table2.whitefemale <- round(prop.table(table(dat$Count[dat$RaceSex == "White Female"]==0, 
+                                       dat$Age2[dat$RaceSex == "White Female"]),2)*100,2)
+table2.blackfemale <- round(prop.table(table(dat$Count[dat$RaceSex == "Black Female"]==0, 
+                                       dat$Age2[dat$RaceSex == "Black Female"]),2)*100,2)
 
@@
 
-table3.whitemale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "White.Male"]==T, 
-                                      dat$Age2[dat$RaceSex == "White.Male"]),2)*100,2)
-table3.blackmale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "Black.Male"]==T, 
-                                       dat$Age2[dat$RaceSex == "Black.Male"]),2)*100,2)
-table3.whitefemale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "White.Female"]==T, 
-                                       dat$Age2[dat$RaceSex == "White.Female"]),2)*100,2)
-table3.blackfemale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "Black.Female"]==T, 
-                                       dat$Age2[dat$RaceSex == "Black.Female"]),2)*100,2)
+table3.whitemale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "White Male"]==T, 
+                                      dat$Age2[dat$RaceSex == "White Male"]),2)*100,2)
+table3.blackmale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "Black Male"]==T, 
+                                       dat$Age2[dat$RaceSex == "Black Male"]),2)*100,2)
+table3.whitefemale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "White Female"]==T, 
+                                       dat$Age2[dat$RaceSex == "White Female"]),2)*100,2)
+table3.blackfemale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "Black Female"]==T, 
+                                       dat$Age2[dat$RaceSex == "Black Female"]),2)*100,2)
 

### 2016-11-05 Code/append_smoothed_results.Rmd

---
title: "Append_smoothed_results"
author: "Corinne Riddell"
date: "November 5, 2016"
output: html_document
---

```{r load_libraries}
library(plyr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

```{r load_datasets}
load(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
ktm_smoothed <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/smoothed_results.csv")
```

```{r clean_smoothed_data}
ktm_clean <- ktm_smoothed %>% 
  rename(Race2 = race, Sex2 = sex, State2 = state, COD2 = cod) %>%
  mutate(Age2 = factor(age_bin, levels = c(1:19), 
                       labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                  "15-19 years", "20-24 years", "25-29 years", "30-34 years",
                                  "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
                                  "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
                                  "75-79 years", "80-84 years","85+ years"), ordered = T),
         Year2 = factor(year, levels = c(1:45), labels = as.character(1969:2013))
         )
```

```{r merge_results}
dim(dat.clean)
dat.clean <- merge(dat.clean, ktm_clean, 
                   by = c("State2", "Year2", "Race2", "Sex2", "Age2", "COD2"), 
                   all.x = T)
dim(dat.clean)

dat.clean <- dat.clean %>% mutate(Appended_smoothed = ifelse(is.na(X), F, T))

table(dat.clean$Appended_smoothed)
```

Ensure that the merge was correct by comparing the duplicate variables (i.e., those variables in both datasets that had different names)

```{r check_merge}
dat.clean[1:10, c("smoothed_deaths", "deaths", "Count", "population", "Population", "State2", "Age2", "Sex2", "Race2", "COD2")]

diffs <- dat.clean %>% transmute(diff.pop = Population - population, diff.count = Count - deaths)
summary(diffs$diff.pop)
summary(diffs$diff.count)
rm(diffs)
```

Compare the smoothed_results to the original counts to see if they seem reasonable:

```{r compared_original_vs_smoothed}
ggplot(dat.clean, aes(x = Count, y = smoothed_deaths)) + geom_point(alpha = 0.1)
```

Can't really tell anything from this -- make another graphic:

```{r compared_original_vs_smoothed2}
dat.clean <- dat.clean %>% mutate(diff.orig.smooth = Count - smoothed_deaths)
summary(dat.clean$diff.orig.smooth)
ggplot(dat.clean, aes(x = diff.orig.smooth)) + geom_density() + geom_rug(alpha = 0.1)
```

```{r addvar_cod_proportions}
dat.clean <- dat.clean %>% group_by(State2, RaceSex, Year3, Age3) %>%
                           arrange(State2, RaceSex, Year3, Age3) %>%
                           mutate(total_smoothed_deaths = sum(smoothed_deaths),
                                  cod_prop_smoothed_deaths = smoothed_deaths/total_smoothed_deaths
                                  )
summary(dat.clean$cod_prop_smoothed_deaths[dat.clean$Appended_smoothed == F])

length(dat.clean$total_smoothed_deaths[dat.clean$Appended_smoothed == T])
length(dat.clean$total_smoothed_deaths[dat.clean$total_smoothed_deaths == 0 & dat.clean$Appended_smoothed == T])
#the smoothed results are NEVER exactly equal to 0.

#there are quite a few < 1:
length(dat.clean$total_smoothed_deaths[dat.clean$total_smoothed_deaths < 0.1 & dat.clean$Appended_smoothed == T])
length(dat.clean$total_smoothed_deaths[dat.clean$total_smoothed_deaths < 0.5 & dat.clean$Appended_smoothed == T])
length(dat.clean$total_smoothed_deaths[dat.clean$total_smoothed_deaths < 1 & dat.clean$Appended_smoothed == T])
```

```{r aggregate_over_COD}
dat.smoothed.aggregatedCOD <- dat.clean %>% 
  group_by(State2, Year2, RaceSex, Age2) %>%
  arrange(State2, RaceSex, Year2, Age2) %>%
  filter(Appended_smoothed == T) %>%
  summarise(Population = first(Population),
            Race2 = first(Race2), 
            Sex2 = first(Sex2), 
            StateYearRaceSex = first(StateYearRaceSex),
            nx = first(nx),
            Age3 = first(Age3),
            Year3 = first(Year3),
            total_smoothed_deaths = first(total_smoothed_deaths),
            cod_prop_smoothed_deaths = first(cod_prop_smoothed_deaths))

dim(dat.smoothed.aggregatedCOD)

non.zero.stratums.smoothed <-  unique(dat.smoothed.aggregatedCOD$StateYearRaceSex)
```

In those strata that KTM ran the model, create a life table and store it in a list.

```{r calculate_life_expectancy}
list.life.tables.smoothed <- vector(mode = "list", length = length(non.zero.stratums.smoothed))
list.index <- 1

system.time(
for(stratum.id in non.zero.stratums.smoothed) {

  list.life.tables.smoothed[[list.index]] <- life.table(data = subset(dat.smoothed.aggregatedCOD, StateYearRaceSex == stratum.id), 
                                               num.ages.in.group = "nx", 
                                               death.counts = "total_smoothed_deaths", 
                                               population.counts = "Population")
  list.index <- list.index + 1
  print(list.index)
  }
)

```

Add the life expectancy at birth to the dat2 data file:

```{r add_le_to_dataset}
le.smoothed <- rep(NA, length(non.zero.stratums.smoothed))

for (i in 1:length(non.zero.stratums.smoothed)) {
  le.smoothed[i] <- list.life.tables.smoothed[[i]]$e_x[1]
}

les.smoothed <- data.frame(life.expectancy.birth.smoothed = le.smoothed, StateYearRaceSex = non.zero.stratums.smoothed)

dim(dat2)
dat3 <- merge(dat2, les.smoothed, by = "StateYearRaceSex", all.x =T)
dim(dat3)

dat2 <- dat3
rm(les.smoothed, dat3)
```

```{r create_dataset_BlackWhite}
#this dataset contains the black-white data within the same row (wide format) for each year-state-sex strata

BlackSmooth <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth.smoothed) %>%
                  rename(Pop_black = Pop_across_age,  le.smoothed.black = life.expectancy.birth.smoothed)

WhiteSmooth <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth.smoothed) %>%
                  rename(Pop_white = Pop_across_age,  le.smoothed.white = life.expectancy.birth.smoothed)

BlackWhiteSmooth <- merge(BlackSmooth, WhiteSmooth, by = c("State2", "Year3", "Sex2"))

BlackWhiteSmooth <- BlackWhiteSmooth %>% mutate(WBgap = le.smoothed.white - le.smoothed.black)
```

```{r save_workspace}
save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
```

```{r save_specified_files}
save(dat.clean, dat2, BlackWhite, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
     file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
```

Curious how this affects the LE calculation 
- using fractional data
- The fact that smoothed are NEVER zero seems problematic to me because it will bias LE downwards
- Perhaps that is not the case because it also pulls some death counts down.

- Best way to investigate this? Are there any life tables where we have 0 counts (exactly known) but no NA counts?
- need to do a summary with info on num.zeroes and num.NAs and see if any strata have num.zeroes > 1 but num.NA == 0.

### 2016-11-07 shiny_compare_LE/App.R
added app to vis NA

@@ -26,7 +26,8 @@ ui1 <- fluidPage(#theme = shinytheme("cosmo"),
     mainPanel(
       
       plotlyOutput("life_expectancy"),
-      plotlyOutput("life_expectancy_smoothed"))
+      plotlyOutput("life_expectancy_smoothed"),
+      plotlyOutput("sum_missing"))
   )
   
 )
@@ -85,7 +86,22 @@ server <- function(input, output) {
                      facet_wrap(~Sex2) + scale_x_continuous(name = "Year") + scale_y_continuous(name = "Life expectancy gap (years)") + 
                      theme_minimal() + theme(legend.title=element_blank()))
     subplot(p1, p2, shareX = T, nrows = 2, titleY = T)
-  })    
+  })  
+  
+  output$sum_missing <- renderPlotly({
+    temp.dat <- (dat.clean[dat.clean$State2 == input$state & dat.clean$Race2 == "Black" & dat.clean$Sex2 == "Male", c("Age2", "Population", "Year", "Count", "COD2")])
+    
+    sums <- temp.dat %>% 
+      group_by(Year, Age2) %>% 
+      summarize(sum.na = sum(is.na(Count)), sum.zero = sum(Count == 0, na.rm=T)) %>%
+      arrange(Year, Age2)
+    
+    ggplotly(ggplot(dat = sums, aes(x = Year, y = sum.na)) + 
+             geom_line(aes(col = "Sum NA")) + geom_line(aes(y = sum.zero, col = "Sum zeroes")) +
+             facet_wrap(~Age2) +
+      ylab("Number of missing death counts") +
+      ggtitle("Black Males"))
+  })
 }
 
 shinyApp(ui = ui1, server = server)

### 2016-11-07 Code/append_smoothed_results.Rmd

@@
 ```{r save_specified_files}
-save(dat.clean, dat2, BlackWhite, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
+save(dat.clean, dat2, BlackWhite, BlackWhiteSmooth, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
      file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")

### 2016-11-08 Code/1_Explore_Data_v2.Rmd
adding in LE for 1 and 9 but not yet finished

@@
 dat2 <- merge(dat2, dat3, by = "StateYearRaceSex")
-
+rm(dat3)
 dat2 <- dat2 %>% arrange(State2, Year3, Race2, Sex2, Age3)
@@
 ```{r calculate_life_expectancy}
-list.life.tables <- vector(mode = "list", length = length(non.zero.stratums))
+list.life.tables <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
+#list.life.tables1 <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
 list.index <- 1
@@
 for(stratum.id in non.zero.stratums) {
+#for(stratum.id in head(non.zero.stratums)) {
 
-  list.life.tables[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id), 
+  list.life.tables[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2),   
+  #list.life.tables1[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
+                                               num.ages.in.group = "nx", 
+                                               death.counts = "total_deaths1", 
+                                               population.counts = "Population")
+
+  list.life.tables5[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
                                                num.ages.in.group = "nx", 
@@
                                                population.counts = "Population")
+
+  list.life.tables9[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
+                                               num.ages.in.group = "nx", 
+                                               death.counts = "total_deaths9", 
+                                               population.counts = "Population")
   list.index <- list.index + 1
+  print(list.index)
   }
 )
-
 ```
@@
 le <- rep(NA, length(non.zero.stratums))
+le5 <- rep(NA, length(non.zero.stratums))
+le9 <- rep(NA, length(non.zero.stratums))
 
@@
   le[i] <- list.life.tables[[i]]$e_x[1]
-
+  le5[i] <- list.life.tables5[[i]]$e_x[1]
+  le9[i] <- list.life.tables9[[i]]$e_x[1]
 }
 
-les <- data.frame(life.expectancy.birth = le, StateYearRaceSex = non.zero.stratums)
+les <- data.frame(life.expectancy.birth = le, life.expectancy.birth.5 = le5, life.expectancy.birth.9 = le9, StateYearRaceSex = non.zero.stratums)
 
@@
 Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
-                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth) %>%
-                  rename(Pop_black = Pop_across_age,  le.birth.black = life.expectancy.birth)
+                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth, life.expectancy.birth5, life.expectancy.birth9) %>%
+                  rename(Pop_black = Pop_across_age,  
+                         le.birth.black = life.expectancy.birth, 
+                         le.birth.black5 = life.expectancy.birth5, 
+                         le.birth.black9 = life.expectancy.birth9)
 
 White <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
-                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth) %>%
-                  rename(Pop_white = Pop_across_age,  le.birth.white = life.expectancy.birth)
+                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth, life.expectancy.birth5, life.expectancy.birth9) %>%
+                  rename(Pop_white = Pop_across_age,  
+                         le.birth.white = life.expectancy.birth,
+                         le.birth.white5 = life.expectancy.birth5,
+                         le.birth.white9 = life.expectancy.birth9)
 
@@
 
-BlackWhite <- BlackWhite %>% mutate(WBgap = le.birth.white - le.birth.black)
+BlackWhite <- BlackWhite %>% mutate(WBgap = le.birth.white - le.birth.black,
+                                    WBgap5 = le.birth.white5 - le.birth.black5,
+                                    WBgap9 = le.birth.white9 - le.birth.black9)
 ```
@@
 
-list.age.decomp.tables <- vector(mode = "list", length = dim(paired.ids)[1])
+list.age.decomp.tables <- list.age.decomp.tables5 <- list.age.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])
 
@@
                                                  name.lt2 = "White", age.groups = "Age3")
+
+    list.age.decomp.tables5[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]), 
+                                                 name.lt1 = "Black", 
+                                                 life.table2 = as.data.frame(list.life.tables5[[white.index]]), 
+                                                 name.lt2 = "White", age.groups = "Age3")
+
+    list.age.decomp.tables9[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables9[[black.index]]), 
+                                                 name.lt1 = "Black", 
+                                                 life.table2 = as.data.frame(list.life.tables9[[white.index]]), 
+                                                 name.lt2 = "White", age.groups = "Age3")
   } else {
-    list.age.decomp.tables[[i]] <- "Couldn't compare life tables"
+    list.age.decomp.tables[[i]] <- list.age.decomp.tables5[[i]] <- list.age.decomp.tables9[[i]] <-"Couldn't compare life tables"
   }
@@
 #takes about 20 mins
-list.cod.decomp.tables <- vector(mode = "list", length = dim(paired.ids)[1])
+list.cod.decomp.tables <- list.cod.decomp.tables5 <- list.cod.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])
 
@@
                 ungroup() %>%
-                select(Age2, COD2, cod_prop_death5) %>%
-                rename(prop.white = cod_prop_death5)
+                select(Age2, COD2, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
+                rename(prop.white1 = cod_prop_death1, prop.white5 = cod_prop_death5, prop.white9 = cod_prop_death9)
 
@@
                 ungroup() %>%
-                select(Age2, COD2, cod_prop_death5) %>%
-                rename(prop.black = cod_prop_death5)
+                select(Age2, COD2, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
+                rename(prop.black1 = cod_prop_death1, prop.black5 = cod_prop_death5, prop.black9 = cod_prop_death9)
     ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
@@
                                                          decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
-                                                         cod.table = ct.both, 
+                                                         cod.table = ct.both %>% select(Age2, COD2, prop.white5, prop.black5), 
                                                          age.colname.cod.table = "Age2", 

### 2016-11-08 shinyapp3/app.R

@@ -26,6 +26,7 @@ ui1 <- fluidPage(#theme = shinytheme("cosmo"),
                    mainPanel(
                      
                      #add the new output object for our LE graphs -- see my notes.
+                     radioButtons(inputId = "LE_type", label = "Use data from", inline = T, choices = c("Bayes", "Impute 1", "Impute 5", "Impute 10"), selected = "Impute 5"),
                      plotlyOutput("life_expectancy"),
                      radioButtons(inputId = "contribution_type", label = "Display contribution in", inline = T, choices = c("Years", "Proportion (%)"), selected = "Years"),
                      plotlyOutput("male_age_comp1"),

### 2016-11-08 Code/1_Explore_Data_v2.Rmd
computed LE and decomp for 1 and 5

@@
 ```{r calculate_life_expectancy}
+#takes 10 mins for all three
+
 list.life.tables <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
@@
 ```{r add_le_to_dataset}
-le <- rep(NA, length(non.zero.stratums))
-le5 <- rep(NA, length(non.zero.stratums))
-le9 <- rep(NA, length(non.zero.stratums))
+le <- le5 <- le9 <- rep(NA, length(non.zero.stratums))
 
@@
 
-les <- data.frame(life.expectancy.birth = le, life.expectancy.birth.5 = le5, life.expectancy.birth.9 = le9, StateYearRaceSex = non.zero.stratums)
+les <- data.frame(life.expectancy.birth = le, 
+                  life.expectancy.birth.5 = le5, 
+                  life.expectancy.birth.9 = le9, 
+                  StateYearRaceSex = non.zero.stratums)
 
@@
 rm(les)
+
+ggplot(dat2 %>% filter(Age2 == "<1 year"), aes(x = life.expectancy.birth)) + 
+  geom_density(aes(fill="Impute 1"), alpha = 0.3) + 
+  geom_density(aes(x = life.expectancy.birth.5, fill = "Impute 5"), alpha = 0.3) +
+  geom_density(aes(x = life.expectancy.birth.9, fill="Impute 9"), alpha = 0.3) +
+  facet_grid(Sex2 ~ Race2)
 ```
@@
 Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
-                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth, life.expectancy.birth5, life.expectancy.birth9) %>%
+                  select(State2, Year3, Sex2, Pop_across_age, 
+                         life.expectancy.birth, life.expectancy.birth.5, 
+                         life.expectancy.birth.9) %>%
                   rename(Pop_black = Pop_across_age,  
                          le.birth.black = life.expectancy.birth, 
-                         le.birth.black5 = life.expectancy.birth5, 
-                         le.birth.black9 = life.expectancy.birth9)
+                         le.birth.black5 = life.expectancy.birth.5, 
+                         le.birth.black9 = life.expectancy.birth.9)
 
 White <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
-                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth, life.expectancy.birth5, life.expectancy.birth9) %>%
+                  select(State2, Year3, Sex2, Pop_across_age, 
+                         life.expectancy.birth, life.expectancy.birth.5, 
+                         life.expectancy.birth.9) %>%
                   rename(Pop_white = Pop_across_age,  
                          le.birth.white = life.expectancy.birth,
-                         le.birth.white5 = life.expectancy.birth5,
-                         le.birth.white9 = life.expectancy.birth9)
+                         le.birth.white5 = life.expectancy.birth.5,
+                         le.birth.white9 = life.expectancy.birth.9)
 
@@
 }
+
 ```
@@
 
-#takes about 20 mins
+#takes 1.5 hours for all three
 list.cod.decomp.tables <- list.cod.decomp.tables5 <- list.cod.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])
 
+system.time(
 for(i in 1:dim(paired.ids)[1]) {
@@
 
+    if(any(is.na(ct.both$prop.black1)) == F & any(is.na(ct.both$prop.white1)) == F) { #ensures that there are no ages with 0 deaths - for these COD prop == 0 and can't compute
+
         list.cod.decomp.tables[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
+                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
+                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
+                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white1, prop.black1), 
+                                                              age.colname.cod.table = "Age2", 
+                                                              COD.colname.cod.table = "COD2", 
+                                                              prop1.colname.cod.table = "prop.black1", 
+                                                              prop2.colname.cod.table = "prop.white1")
+
+        list.cod.decomp.tables5[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
                                                               life.table2 = as.data.frame(list.life.tables[[white.index]]), 
@@
                                                               COD.colname.cod.table = "COD2", 
-                                                         prop1.colname.cod.table = "prop.black", 
-                                                         prop2.colname.cod.table = "prop.white")
+                                                              prop1.colname.cod.table = "prop.black5", 
+                                                              prop2.colname.cod.table = "prop.white5")
+
+        list.cod.decomp.tables9[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
+                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
+                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
+                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white9, prop.black9), 
+                                                              age.colname.cod.table = "Age2", 
+                                                              COD.colname.cod.table = "COD2", 
+                                                              prop1.colname.cod.table = "prop.black9", 
+                                                              prop2.colname.cod.table = "prop.white9")
+
+    } else { list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Encountered age group with no deaths"
+
+    }
   } else {
-    list.cod.decomp.tables[[i]] <- "Couldn't compare life tables"
+
+    list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Couldn't compare life tables"
   }
-  #print(i)
+
+  print(i)
 }
+)
 ```
@@
 for(i in 1:dim(paired.ids)[1]) {
-    if(list.cod.decomp.tables[[i]][1] != "Couldn't compare life tables") { 
-        list.cod.decomp.tables[[i]] <- 
-                              make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables[[i]], 
+    if(list.cod.decomp.tables[[i]][1] != "Couldn't compare life tables" & 
+       list.cod.decomp.tables[[i]][1] != "Encountered age group with no deaths") {
+
+       list.cod.decomp.tables[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables[[i]],
+                                                            age.groups = "Ages", 
+                                                            cause.of.death = "Cause.of.death")
+
+       list.cod.decomp.tables5[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables5[[i]],
+                                                            age.groups = "Ages", 
+                                                            cause.of.death = "Cause.of.death")
+
+       list.cod.decomp.tables9[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables9[[i]],
                                                             age.groups = "Ages", 
@@
     } 
-  #print(i)
+  print(i)
 }  
@@
 ggplotly(ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
+           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
+           xlab("Contribution to life expectancy gap") +
+           ylab("Age group\n") + theme_minimal() +
+           geom_vline(xintercept = 0))
+
+ggplotly(ggplot(list.cod.decomp.tables9[[1]], aes(y = Ages, x = start)) + 
            geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
@@
 save(dat.clean, dat2, BlackWhite, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
+     list.age.decomp.tables5, list.cod.decomp.tables5, list.age.decomp.tables9, list.cod.decomp.tables9,
      file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")

### 2016-11-08 Code/Explore_unaggregate_data.Rmd
computed LE and decomp for 1 and 5

---
title: "Untitled"
author: "Corinne Riddell"
date: "November 8, 2016"
output: html_document
---

```{r setup, include=FALSE}
library(plyr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

```{r load datasets}
dat.male <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/males_export.txt", header = T, sep="\t", na.strings = " ")
dat.male <- dat.male[-1, ]
names(dat.male) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
prop.table(table(dat.male$Count==" "))
prop.table(table(is.na(dat.male$Count))

head(dat.male)

dat.female <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/female_export.txt", header = T, sep="\t")
dat.female <- dat.female[-1, ]
names(dat.female) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
dat.female$Sex <- 1

prop.table(table(dat.female$Count==" "))

head(dat.female)

dat.both <- rbind(dat.male, dat.female)

rm(dat.male, dat.female)
```

```{r load_COD_dictionary}
cod.dict <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/COD_dictionary.csv")

cod.dict$COD2 <- reorder(cod.dict$COD, cod.dict$COD_num)
levels(cod.dict$COD2)
```

```{r recode_variables}
dat.both <- dat.both %>%
  mutate(State2 = factor(State, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
                                                             "Colorado", "Connecticut", "Delaware", "Washington DC",
                                                             "Florida", "Georgia", "Hawaii","Idaho", "Illinois", 
                                                             "Indiana","Iowa", "Kansas", "Kentucky", "Louisiana", 
                                                             "Maine", "Maryland", "Massachusetts", "Michigan", 
                                                             "Minnesota", "Mississippi","Missouri", "Montana", 
                                                             "Nebraska", "Nevada", "New Hampshire", "New Jersey",
                                                             "New Mexico", "New York","North Carolina","North Dakota",
                                                             "Ohio", "Oklahoma","Oregon", "Pennsylvania", 
                                                             "Rhode Island", "South Carolina", "South Dakota", 
                                                             "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
                                                             "Washington", "West Virginia", "Wisconsin", "Wyoming")), 
         Age2 = factor(Age, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                         "15-19 years", "20-24 years", "25-29 years", "30-34 years",
                                                         "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
                                                         "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
                                                         "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
         Age3 = 0*(Age2 == "<1 year") + 1*(Age2 == "1-4 years") + 5*(Age2 == "5-9 years") +
                10*(Age2 == "10-14 years") + 15*(Age2 == "15-19 years") + 20*(Age2 == "20-24 years") + 
                25*(Age2 == "25-29 years") + 30*(Age2 == "30-34 years") + 35*(Age2 == "35-39 years") +
                40*(Age2 == "40-44 years") + 45*(Age2 == "45-49 years") + 50*(Age2 == "50-54 years") + 
                55*(Age2 == "55-59 years") + 60*(Age2 == "60-64 years") + 65*(Age2 == "65-69 years") +   
                70*(Age2 == "70-74 years") + 75*(Age2 == "75-79 years") + 80*(Age2 == "80-84 years") + 
                85*(Age2 == "85+ years"),
         nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years"),
         Sex2 = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
         Race2 = factor(Race, levels = c(0, 1), labels = c("White", "Black")),
         COD2 = factor(COD, levels = c(0:92), labels = levels(cod.dict$COD2)), 
         Year2 = factor(Year, levels = c(0:44), labels = as.character(1969:2013)),
         Year3 = as.numeric(as.character(Year2)),
         RaceSex = interaction(Race2, Sex2, sep = " "),
         StateYearSex = interaction(State2, Year2, Sex2),
         StateYearRaceSex = interaction(State2, Year2, Race2, Sex2))
```

### 2016-11-08 shinyapp3/app.R

@@ -26,7 +26,10 @@ ui1 <- fluidPage(#theme = shinytheme("cosmo"),
                    mainPanel(
                      
                      #add the new output object for our LE graphs -- see my notes.
-                     radioButtons(inputId = "LE_type", label = "Use data from", inline = T, choices = c("Bayes", "Impute 1", "Impute 5", "Impute 10"), selected = "Impute 5"),
+                     radioButtons(inputId = "LE_type", label = "Use data from", inline = T, choices = c("Bayes", "Impute 1",# = "le.birth.black", 
+                                                                                                        "Impute 5",# = "le.birth.black5", 
+                                                                                                        "Impute 9"),# = "le.birth.black9"), 
+                                                                                                        selected = "Impute 5"),
                      plotlyOutput("life_expectancy"),
                      radioButtons(inputId = "contribution_type", label = "Display contribution in", inline = T, choices = c("Years", "Proportion (%)"), selected = "Years"),
                      plotlyOutput("male_age_comp1"),
@@ -53,17 +56,31 @@ server <- function(input, output) {
     p %>% layout(showlegend = F)
    
   })
+
+  yvar1 <- reactive({ switch(input$LE_type,
+                             "Impute 1" = BlackWhite$le.birth.black,
+                             "Impute 5" = BlackWhite$le.birth.black5,
+                             "Impute 9" = BlackWhite$le.birth.black9)
+  })
+  
+  
+  yvar2 <- reactive({ switch(input$LE_type,
+                             "Impute 1" = BlackWhite$le.birth.white,
+                             "Impute 5" = BlackWhite$le.birth.white5,
+                             "Impute 9" = BlackWhite$le.birth.white9)
+  })
   
   output$life_expectancy <- renderPlotly({
+
 #    p1 <- ggplotly(ggplot(subset(dat2, State2 == input$state & Age3 == 0), aes(x = Year3, y = life.expectancy.birth)) + 
 #      geom_line(aes(col = Sex2, lty = Race2)) + facet_wrap(~Sex2) +
 #      scale_x_continuous(name = "Year") + scale_y_continuous(name = "Life expectancy at birth (years)") + 
 #      theme_minimal() + theme(legend.title=element_blank()))
 
[…]

### 2016-11-09 Code/Explore_unaggregate_data.Rmd
made granular dataset for alabama

@@
 dat.male <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/males_export.txt", header = T, sep="\t", na.strings = " ")
-dat.male <- dat.male[-1, ]
 names(dat.male) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
-prop.table(table(dat.male$Count==" "))
-prop.table(table(is.na(dat.male$Count))
+prop.table(table(is.na(dat.male$Count)))
+head(dat.male)
+table(dat.male$Sex)
+str(dat.male) #note all numeric or int
 
-head(dat.male)
+dat.female <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/female_export.txt", header = T, sep="\t", na.strings = " ")
+names(dat.female) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
+table(dat.female$Sex)
+dat.female$Sex <- 1
+str(dat.male) #note all numeric or int
 
-dat.female <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/female_export.txt", header = T, sep="\t")
-dat.female <- dat.female[-1, ]
-names(dat.female) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
-dat.female$Sex <- 1
-
-prop.table(table(dat.female$Count==" "))
-
+prop.table(table(is.na(dat.female$Count)))
 head(dat.female)
@@
 levels(cod.dict$COD2)
+
+cod.dict$cod_cat <- "Cancers"
+
+cod.dict$cod_cat[cod.dict$COD2 == "Diseases of Heart" | 
+                 cod.dict$COD2 == "Hypertension without Heart Disease" |
+                 cod.dict$COD2 == "Cerebrovascular Diseases" |  
+                 cod.dict$COD2 == "Atherosclerosis" |  
+                 cod.dict$COD2 == "Aortic Aneurysm and Dissection" | 
+                 cod.dict$COD2 == "Other Diseases of Arteries, Arterioles, Capillaries"] <- "Cardiovascular"
+
+cod.dict$cod_cat[cod.dict$COD2 == "Tuberculosis" | 
+                 cod.dict$COD2 == "Syphilis" |
+                 cod.dict$COD2 == "Human Immunodeficiency Virus (HIV) (1987+)" |  
+                 cod.dict$COD2 == "Septicemia" |  
+                 cod.dict$COD2 == "Other Infectious and Parasitic Diseases" | 
+                 cod.dict$COD2 == "Pneumonia and Influenza"] <- "Communicable"
+
+cod.dict$cod_cat[cod.dict$COD2 == "Diabetes Mellitus" | 
+                 cod.dict$COD2 == "Alzheimers (ICD-9 and 10 only)" |
+                 cod.dict$COD2 == "Chronic Obstructive Pulmonary Disease and Allied Cond" |  
+                 cod.dict$COD2 == "Stomach and Duodenal Ulcers" |  
+                 cod.dict$COD2 == "Chronic Liver Disease and Cirrhosis" | 
+                 cod.dict$COD2 == "Nephritis, Nephrotic Syndrome and Nephrosis"] <- "Non-communicable"
+
+cod.dict$cod_cat[cod.dict$COD2 == "Accidents and Adverse Effects" | 
+                 cod.dict$COD2 == "Suicide and Self-Inflicted Injury" |
+                 cod.dict$COD2 == "Homicide and Legal Intervention"] <- "Injuries"
+
+cod.dict$cod_cat[cod.dict$COD2 == "Complications of Pregnancy, Childbirth, Puerperium" | 
+                 cod.dict$COD2 == "Congenital Anomalies" |
+                 cod.dict$COD2 == "Certain Conditions Originating in Perinatal Period" |  
+                 cod.dict$COD2 == "Symptoms, Signs and Ill-Defined Conditions" |  
+                 cod.dict$COD2 == "Other Cause of Death"] <- "All other causes"
+
+cod.dict <- cod.dict[ ,-2]
 ```
@@
          StateYearRaceSex = interaction(State2, Year2, Race2, Sex2))
+
+dim(dat.both)
+dat.both <- merge(dat.both, cod.dict, by = "COD2") #to add on COD_num and cod_cat
+dim(dat.both)
 ```
+
+```{r pull_alabama}
+alabama.males <- dat.both %>% filter(State2 == "Alabama", Sex2 == "Male")
+
+save(alabama.males, file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/alabama_males_granular.Rdata")
+```

### 2016-11-10 Code/append_smoothed_results.Rmd
finish stuff with bayes models

--- 2016-11-05 Code/append_smoothed_results.Rmd
@@
 ```{r calculate_life_expectancy}
+#6 mins
+
 list.life.tables.smoothed <- vector(mode = "list", length = length(non.zero.stratums.smoothed))
@@
 )
-
 ```
@@
 
-```{r create_dataset_BlackWhite}
+```{r create_dataset_BlackWhiteSmooth}
 #this dataset contains the black-white data within the same row (wide format) for each year-state-sex strata
@@
 BlackSmooth <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
-                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth.smoothed) %>%
-                  rename(Pop_black = Pop_across_age,  le.smoothed.black = life.expectancy.birth.smoothed)
+                  select(State2, Year3, Sex2, life.expectancy.birth.smoothed) %>%
+                  rename(le.smoothed.black = life.expectancy.birth.smoothed)
 
 WhiteSmooth <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
-                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth.smoothed) %>%
-                  rename(Pop_white = Pop_across_age,  le.smoothed.white = life.expectancy.birth.smoothed)
+                  select(State2, Year3, Sex2, life.expectancy.birth.smoothed) %>%
+                  rename(le.smoothed.white = life.expectancy.birth.smoothed)
 
@@
 
-BlackWhiteSmooth <- BlackWhiteSmooth %>% mutate(WBgap = le.smoothed.white - le.smoothed.black)
+BlackWhiteSmooth <- BlackWhiteSmooth %>% mutate(WBgap.s = le.smoothed.white - le.smoothed.black)
+
+dim(BlackWhite)
+BlackWhite3 <- merge(BlackWhite, BlackWhiteSmooth, by = c("State2", "Year3", "Sex2"))
+dim(BlackWhite3)
+
+BlackWhite <- BlackWhite3
+rm(BlackWhite3, BlackSmooth, WhiteSmooth)
+```
+
+```{r calculate_age_decomp_smoothed}
+#18 mins
+
+list.age.decomp.tables.smoothed <- vector(mode = "list", length = dim(paired.ids)[1])
+
+for(i in 1:dim(paired.ids)[1]) {
+  white.index <- which(non.zero.stratums.smoothed == paired.ids[i, "StateYearRaceSexWhite"])
+  black.index <- which(non.zero.stratums.smoothed == paired.ids[i, "StateYearRaceSexBlack"])
+
+  if(length(white.index) != 0 & length(black.index) != 0) {
+    list.age.decomp.tables.smoothed[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables.smoothed[[black.index]]),
+                                                          name.lt1 = "Black", 
+                                                          life.table2 = as.data.frame(list.life.tables.smoothed[[white.index]]),
+                                                          name.lt2 = "White", age.groups = "Age3")
+  } else {
+    list.age.decomp.tables.smoothed[[i]] <-"Couldn't compare life tables"
+  }
+  print(i)
+}
+```
+
+```{r calculate_cod_decomp}
+#takes 40 mins
+
+list.cod.decomp.tables.smoothed <- vector(mode = "list", length = dim(paired.ids)[1])
+
+system.time(
+for(i in 1:dim(paired.ids)[1]) {
+  white.index <- which(non.zero.stratums.smoothed == paired.ids[i, "StateYearRaceSexWhite"])
+  black.index <- which(non.zero.stratums.smoothed == paired.ids[i, "StateYearRaceSexBlack"])
+
+  if(length(white.index) != 0 & length(black.index) != 0) {
+    ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
+                ungroup() %>%
+                select(Age2, COD2, cod_prop_smoothed_deaths) %>%
+                rename(prop.white = cod_prop_smoothed_deaths)
+
+    ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
+                ungroup() %>%
+                select(Age2, COD2, cod_prop_smoothed_deaths) %>%
+                rename(prop.black = cod_prop_smoothed_deaths)
+    ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
+    ct.both <- ct.both %>% arrange(COD2, Age2)
+
+    if(any(is.na(ct.both$prop.black)) == F & any(is.na(ct.both$prop.white)) == F) { #ensures that there are no ages with 0 deaths - for these COD prop == 0 and can't compute
+
+        list.cod.decomp.tables.smoothed[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables.smoothed[[black.index]]),
+                                                                      life.table2 = as.data.frame(list.life.tables.smoothed[[white.index]]), 
+                                                                      decomp.table = as.data.frame(list.age.decomp.tables.smoothed[[i]]),
+                                                                      cod.table = ct.both, 
+                                                                      age.colname.cod.table = "Age2", 
+                                                                      COD.colname.cod.table = "COD2", 
+                                                                      prop1.colname.cod.table = "prop.black", 
+                                                                      prop2.colname.cod.table = "prop.white")
+
+    } else { list.cod.decomp.tables.smoothed[[i]] <- "Encountered age group with no deaths"
+
+    }
+  } else {
+
+    list.cod.decomp.tables.smoothed[[i]] <- "Couldn't compare life tables"
+  }
+
+  print(i)
+}
+)
+```
+
+```{r  update_decomp_for_plotting}
+#update the decomposition table for plotting purposes
+for(i in 1:dim(paired.ids)[1]) {
+    if(list.cod.decomp.tables.smoothed[[i]][1] != "Couldn't compare life tables" & 
+       list.cod.decomp.tables.smoothed[[i]][1] != "Encountered age group with no deaths") {
+
+       list.cod.decomp.tables.smoothed[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables.smoothed[[i]],
+                                                            age.groups = "Ages", 
+                                                            cause.of.death = "Cause.of.death")  
+    } 
+  print(i)
+}  
 ```
@@
 ```{r save_specified_files}
-save(dat.clean, dat2, BlackWhite, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
+save(dat.clean, dat2, BlackWhite, BlackWhiteSmooth, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
+     list.age.decomp.tables5, list.cod.decomp.tables5, list.age.decomp.tables9, list.cod.decomp.tables9,
+     list.age.decomp.tables.smoothed, list.cod.decomp.tables.smoothed,
      file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")

### 2016-11-10 Code/life_expectancy_functions.R

@@ -130,3 +130,11 @@ make_dataset_cod_plot <- function(cod.decomp.table, age.groups, cause.of.death) 
     }
   return(updated)
 }
+
+#contribution to the change in the gap
+#age.contribution.to.gap <- function(decomp.table1, decomp.table2) {
+   
+# }
+#by age
+
+#by COD

### 2016-11-10 shinyapp3/app.R

@@ -24,16 +24,15 @@ ui1 <- fluidPage(#theme = shinytheme("cosmo"),
                      plotlyOutput("population_trend")
                       ),
                    mainPanel(
-                     radioButtons(inputId = "LE_type", label = "Use data from", inline = T, 
-                                  choices = c("Bayes", "Impute 1", "Impute 5", "Impute 9"), 
+                     radioButtons(inputId = "LE_type", label = "Model choice", inline = T, 
+                                  choices = c("Bayesian Smoothing", "Impute 1", "Impute 5", "Impute 9"), 
                                   selected = "Impute 5"),
-                     textOutput("text1"),
                      plotlyOutput("life_expectancy"),
                      radioButtons(inputId = "contribution_type", label = "Display contribution in", inline = T, choices = c("Years", "Proportion (%)"), selected = "Years"),
                      plotlyOutput("male_age_comp1"),
                      plotlyOutput("male_age_comp2"),
-                     plotOutput("female_age_comp1"),
-                     plotOutput("female_age_comp2")                   
+                     plotlyOutput("female_age_comp1"),
+                     plotlyOutput("female_age_comp2")                   
                      )
                  )
 
@@ -55,103 +54,95 @@ server <- function(input, output) {
     p %>% layout(showlegend = F)
    
   })
+  
 
-#  yvar1 <- reactive({ switch(input$LE_type,
-#                             "Impute 1" = BlackWhite$le.birth.black,
-#                             "Impute 5" = BlackWhite$le.birth.black5,
-#                             "Impute 9" = BlackWhite$le.birth.black9)
-#  })
   
-  
-#  yvar2 <- reactive({ switch(input$LE_type,
-##                             "Impute 1" = BlackWhite$le.birth.white,
-##                             "Impute 5" = BlackWhite$le.birth.white5,
-#                             "Impute 9" = BlackWhite$le.birth.white9)
-#  })
-  
-  reactDat <- reactive({
-    if(input$LE_type == "Impute 1") {
-     # reactDat <- data.frame(State2 = BlackWhite[[input$state]], Year3 = BlackWhite[[input$year]], y1 = BlackWhite[["le.birth.black"]],
[…]

### 2016-11-10 Code/1_Explore_Data_v2.Rmd

@@
                                     WBgap9 = le.birth.white9 - le.birth.black9)
+
+rm(Black, White)
 ```
@@
 ```{r save_specified_files}
-save(dat.clean, dat2, BlackWhite, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
+save(dat.clean, dat2, BlackWhite, BlackWhiteSmooth, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
      list.age.decomp.tables5, list.cod.decomp.tables5, list.age.decomp.tables9, list.cod.decomp.tables9,
+     list.age.decomp.tables.smoothed, list.cod.decomp.tables.smoothed,
      file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")

### 2016-11-11 Code/1_Explore_Data_v2.Rmd
changed sshiny app for easier compare

@@
 dat.clean <- dat[dat$Age2 != "Unknown", ]
+```
+
+```{r addvar_State_abbreviation}
+dat.clean <- dat.clean %>% mutate(stabbrs = factor(State, levels = c(0:50), 
+                                      labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
+                                                 "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", 
+                                                 "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", 
+                                                 "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", 
+                                                 "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
+                                                 "VT", "VA", "WA", "WV", "WI", "WY")))
+
 ```

### 2016-11-11 shinyapp3/app.R

@@ -12,11 +12,12 @@ ui1 <- fluidPage(#theme = shinytheme("cosmo"),
                  sidebarLayout(
                    sidebarPanel(width = 4,
                      selectInput(inputId = "state", label = "Select a state", width = 200, choices = unique(levels(dat.clean$State2))),
-                     strong("Select years to compare"),
+                     #strong("Select years to compare"),
+                     strong("Select year for decomposition"),
                      tags$style(type="text/css", ".form-group.shiny-input-container{ display: inline-block } strong{ display: block !important  } img{ margin-bottom: 30px }"),
                   
                      selectInput(inputId = "year1", label = NA, choices = unique(dat.clean$Year3), width = 100),
-                     selectInput(inputId = "year2", label = NA, choices = unique(dat.clean$Year3), selected = 2013, width = 100),
+                     #selectInput(inputId = "year2", label = NA, choices = unique(dat.clean$Year3), selected = 2013, width = 100),
                      
                      strong("Legend\n"),
                      img(src="legend.png"),
@@ -29,10 +30,13 @@ ui1 <- fluidPage(#theme = shinytheme("cosmo"),
                                   selected = "Impute 5"),
                      plotlyOutput("life_expectancy"),
                      radioButtons(inputId = "contribution_type", label = "Display contribution in", inline = T, choices = c("Years", "Proportion (%)"), selected = "Years"),
+                     strong("Decomposition based on selected imputation model"),
                      plotlyOutput("male_age_comp1"),
-                     plotlyOutput("male_age_comp2"),
-                     plotlyOutput("female_age_comp1"),
-                     plotlyOutput("female_age_comp2")                   
+                     strong("Decomposition based on Bayesian model"),
+                     plotlyOutput("male_age_cod_bayes")
+                     #plotlyOutput("male_age_comp2"),
+                     #plotlyOutput("female_age_comp1"),
+                     #plotlyOutput("female_age_comp2")                   
                      )
                  )
 
@@ -140,41 +144,55 @@ server <- function(input, output) {
                    geom_vline(xintercept = 0))
   
        })
+  
+  output$male_age_cod_bayes <- renderPlotly({
+    
+    ggplotly(ggplot(list.cod.decomp.tables.smoothed[[which(paired.ids$State2 ==  input$state & 
[…]

### 2016-11-16 Code/Explore_unaggregate_data.Rmd
small change
merged suppressed and non suppressed data

@@
 
+head(dat.both)
+
+dat.both <- dat.both[dat.both$Age2 != "Unknown", ]
+
+dat.both <- dat.both %>% mutate(stabbrs = factor(State, levels = c(0:50), 
+                                      labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
+                                                 "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", 
+                                                 "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", 
+                                                 "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", 
+                                                 "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
+                                                 "VT", "VA", "WA", "WV", "WI", "WY")))
+
+dat.both$agegrp <- dat.both$Age 
+dat.both$agegrp[dat.both$Age == 0]  <- 1
+
 dim(dat.both)
@@
 dim(dat.both)
+
 ```
 
+```{r for_merge}
+
+years0610 <- dat.both[dat.both$Year3 >= 2006 & dat.both$Year3 <= 2010 & dat.both$agegrp > 1, ] 
+#added the age restriction because don't want to deal with the differences in age definition right now
+
+not.suppressed <- foreign::read.dta("/Users/corinneriddell/Dropbox/VitalStats/notsuppressed_0610.dta")
+
+not.suppressed <- not.suppressed %>% 
+  filter(agegrp > 1)
+
+years0610 <- years0610 %>% rename(COD_full = COD2)
+
+together <- merge(years0610, not.suppressed, by = c("stabbrs", "Year3", "Sex", "Race", "agegrp", "COD_full"))
+
+```
+
+```{r}
+together <- together %>% mutate(diff_count = Count - death, 
+                                diff_pop = Population - popestimate,
+                                rel_diff_pop = (Population - popestimate)/Population)
+
+prop.table(table(together$death[is.na(together$Count) == T]))
+
+ggplot(together, aes(x = diff_count)) + geom_density()
+summary(together$diff_count)
+
+ggplot(together, aes(x = diff_pop)) + geom_density()
+summary(together$diff_pop)
+
+ggplot(together, aes(x = rel_diff_pop)) + geom_density()
+summary(together$rel_diff_pop)
+
+View(together[together$rel_diff_pop > 0.6, ])
+```
+From above we see that the death counts match exactly and when the original count is NA the non-suppressed count is between 1 and 9.
+For some reason there is some discrpancy in the populational denominators. These do not match exactly and I don't know why.
+
+Since we are interested in the effect of imputing the NAs and not the effect of the different denominators, we need to use the same denominators in the analysis.
+
 ```{r pull_alabama}
-alabama.males <- dat.both %>% filter(State2 == "Alabama", Sex2 == "Male")
-
-save(alabama.males, file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/alabama_males_granular.Rdata")
+alabama.males <- together %>% filter(State2 == "Alabama", Sex2 == "Male") 
+save(alabama.males, file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/alabama_males_5plus_granular.Rdata")
 ```

### 2016-11-17 Code/life_expectancy_functions.R
updated LE function to work with SE calculation

@@ -131,6 +131,32 @@ make_dataset_cod_plot <- function(cod.decomp.table, age.groups, cause.of.death) 
   return(updated)
 }
 
+#input: data.frame from kathryn's bayesian model
+#input is specific to one state, sex, year, but across all age groups and both races
+#take this dataset and calculate LE for blacks and for whites and the gap (LE.gap = LE.white - LE.black)
+#returns these three values.
+life_expectancy_and_gap <- function(data) {
+library(sqldf)
+
+#aggegrate over COD within agegroup and race
+data.aggregated <- sqldf('select race, population, age_bin, year, sex, state, sum(smoothed_deaths) as total_smoothed_deaths from data group by race, age_bin')  
+
+#add num.ages.in.group
+data.aggregated$nx = 1*(data.aggregated$age_bin == 1) + 4*(data.aggregated$age_bin == 2) + 5*(data.aggregated$age_bin > 2)
+
+#calculate LE by race
+lt.black <- life.table(data = subset(data.aggregated, race == "Black"), num.ages.in.group = "nx" , death.counts = "total_smoothed_deaths", population.counts = "population")
+lt.white <- life.table(data = subset(data.aggregated, race == "White"), num.ages.in.group = "nx" , death.counts = "total_smoothed_deaths", population.counts = "population")
+
+return(data.frame("LE_Black" = lt.black$e_x[1], "LE_White" = lt.white$e_x[1], "LE_WBgap" = lt.white$e_x[1] - lt.black$e_x[1]))
+}
+
+
+ktm_smoothed <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/smoothed_results.csv")
+ktm_sub <- ktm_smoothed %>% filter(state == "Alabama" & sex == "Female" & year == 45)
+
+life_expectancy_and_gap(ktm_sub)
+
 #contribution to the change in the gap
 #age.contribution.to.gap <- function(decomp.table1, decomp.table2) {
    

### 2016-11-17 Code/4_bayesSE.R
major changes to things

@@ -37,14 +37,15 @@ model = function() {
 params = c('mu', 'combo', 'beta0', 'beta1')
 nchains = 2
 
-jags_model = jags(data=ds, param=params, n.thin=1, n.chains=nchains, n.iter=10000, n.burnin=3000, model.file=model) 
+jags_model = jags(data=ds, param=params, n.thin=1, n.chains=nchains, n.iter=2000, n.burnin=500, model.file=model) 
 
+class(jags_model)
 
-
-
+jags_model
+  
 #manual from posterior samples 
 
-p_mcmc = as.mcmc(jags_model)  
+p_mcmc = as.mcmc(jags_model) 
 p = data.frame(p_mcmc[[1]])   
 
 combo = (p$mu.1. + p$mu.2. + p$mu.3. + p$mu.4. +p$mu.5. ) / (p$mu.6. + p$mu.7.)

### 2016-11-17 Code/4_smoothing_time_fun.R

@@ -3,14 +3,14 @@
 # Subset dataset 
 subset_data = function(race, sex, cod, state) {
 
-  selected_vars = c('Age', 'Sex2', 'Race2', 'COD2', 'Year', 'Count', 'Population')
+  selected_vars = c('Age', 'Sex2', 'Race2', 'COD2', 'Year', 'Count', 'Population', 'upper_bound')
 
   ds = dat.clean[(dat.clean$Race2==race & 
                     dat.clean$Sex2==sex &
                     dat.clean$State2==state &
                     dat.clean$COD2==cod), selected_vars]
   
-  df = data.frame(deaths=ds$Count, pop = ds$Population, age=(ds$Age+1), year=(ds$Year+1)) 
+  df = data.frame(deaths=ds$Count, pop = ds$Population, age=(ds$Age+1), year=(ds$Year+1), upper_bound = ds$upper_bound) 
   df$censored = ifelse(is.na(df$deaths),1,0)
   
   ds2 = df[order(df$censored), ]
@@ -19,6 +19,7 @@ subset_data = function(race, sex, cod, state) {
                  lnpop = log(ds2$pop), 
                  age.bin = ds2$age, 
                  year = ds2$year, 
+                 upper_bound = ds2$upper_bound, 
                  binned = is.na(ds2$deaths), 
                  not.binned = !is.na(ds2$deaths), 
                  n.binned = sum(is.na(ds2$deaths)), 
@@ -31,6 +32,12 @@ subset_data = function(race, sex, cod, state) {
   
   return(ds_jags) 
 } 
+
+identify_cells_below_9 = function() {
+  
+  
+  }
+
 
 
 combine_race_data = function(dsb, dsw) {
@@ -74,7 +81,7 @@ model = function() {
     }
     
     for(i in n.not.binned.plus.1_b:n.rows_b) {
-      binned.id_b[i] ~ dinterval(deaths_b[i], c(0, 9)) 
+      binned.id_b[i] ~ dinterval(deaths_b[i], c(0, upper_bound)) 
       deaths_b[i] ~ dpois(mu_b[i])
       log(mu_b[i]) <- lnrate_b[age.bin_b[i], year_b[i]] + lnpop_b[i]
     }
@@ -94,7 +101,7 @@ model = function() {
     }
     
     for(i in n.not.binned.plus.1_w:n.rows_w) {
-      binned.id_w[i] ~ dinterval(deaths_w[i], c(0, 9)) 
[…]

### 2016-11-17 Code/4_smoothing_time_run.R

@@ -13,20 +13,14 @@ load('/Users/kathryn/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata')
 
 # Run models for given population subset (state and gender) 
 
-  r_m = run_smoothing_models(state='Alabama', sex='Male') 
-  r_f = run_smoothing_models(state='Alabama', sex='Female') 
-
- # Calclate LE for a given population gender and year 
-
-lt_males_69 = get_life_tables(jags_bw, sex='Male', year=1969)
-lt_males_79 = get_life_tables(r_m, year=1979)
-
+# r_m = run_smoothing_models(state='Alabama', sex='Male') 
+# r_f = run_smoothing_models(state='Alabama', sex='Female') 
 
 # Run models for all states and both genders 
 
-
-
-r_bothsex_allstates = run_smoothing_models_allstates()
+require(parallel)
+#mclapply(X=states, FUN=run_smoothing_models_allstates, mc.cores=4)
+#r_bothsex_allstates = run_smoothing_models_allstates()
 
 
 

### 2016-11-17 Code/4_smoothing_time_src.R

@@ -8,7 +8,8 @@ library(jagstools)
 library(ggplot2)
 library(reshape2)
 
-load('../Data/alabama_only.Rdata')
+#load('../Data/alabama_only.Rdata')
 
-
+dat.clean$upper_bound = ifelse(dat.clean$Population<9, dat.clean$Population, 9) 
+#table(dat.clean$upper_bound)
 

### 2016-11-17 Code/append_smoothed_results.Rmd

@@
 ktm_smoothed <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/smoothed_results.csv")
+
+ktm_rest_males <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/males_restofstates6COD.csv")
+ktm_rest_females <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/females_restofstates6COD.csv")
+
 ```

### 2016-11-22 Code/append_smoothed_results.Rmd
append all the smoothed results

--- 2016-11-10 Code/append_smoothed_results.Rmd
@@
 load(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
-ktm_smoothed <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/smoothed_results.csv")
+ktm_smoothed <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/smoothed_results.csv")
+ktm_rest <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/smoothed_results_extra_states.csv")
+ktm_virginia_males <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/results_males_Virginia_Nov21.csv")
+
+dat.clean <- dat.clean[, -c(31:40)]
+
+head(ktm_smoothed)
+head(ktm_rest)
+head(ktm_virginia_males)
+
+table(ktm_rest$state == "Virginia")
+table(ktm_rest$state)
+table(ktm_smoothed$state) #check - i think Mass is in both rest and smoothed so need to remove duplicates.
+
+dim(ktm_rest)
+ktm_rest <- ktm_rest %>% filter(state != "Massachusetts") #delete duplicate
+dim(ktm_rest)
+
+length(ktm_rest$sex[ktm_rest$state ==  "Virginia" & ktm_rest$sex == "Male"])
+dim(ktm_rest)
+ktm_rest <- ktm_rest %>% filter(!(state == "Virginia" & sex == "Male"))
+dim(ktm_rest)
+
+ktm_virginia_males <- ktm_virginia_males[ , -1]
+
+ktm_all <- rbind(ktm_smoothed, ktm_rest, ktm_virginia_males)
 ```
@@
 ```{r clean_smoothed_data}
-ktm_clean <- ktm_smoothed %>% 
+ktm_clean <- ktm_all %>% 
   rename(Race2 = race, Sex2 = sex, State2 = state, COD2 = cod) %>%
@@
 dim(dat.clean)
-
 dat.clean <- dat.clean %>% mutate(Appended_smoothed = ifelse(is.na(X), F, T))
@@
 table(dat.clean$Appended_smoothed)
+prop.table(table(dat.clean$Appended_smoothed))
 ```
@@
 
-diffs <- dat.clean %>% transmute(diff.pop = Population - population, diff.count = Count - deaths)
-summary(diffs$diff.pop)
-summary(diffs$diff.count)
-rm(diffs)
+dat.clean <- dat.clean %>% mutate(diff.pop = Population - population, diff.count = Count - deaths)
+summary(dat.clean$diff.pop)
+summary(dat.clean$diff.count)
 ```
@@
 ggplot(dat.clean, aes(x = diff.orig.smooth)) + geom_density() + geom_rug(alpha = 0.1)
+
+ggplot(subset(dat.clean, State2 == "Georgia"), aes(x = diff.orig.smooth)) + geom_histogram() + geom_rug(alpha = 0.1) + facet_grid(Sex2~Race2)
+ggplot(subset(dat.clean, State2 == "Georgia" & Sex2 == "Male" & Race2 == "Black"), aes(x = diff.orig.smooth)) + geom_histogram() + geom_rug(alpha = 0.1) + facet_wrap(~Age3)
+
+ggplot(subset(dat.clean, State2 == "California"), aes(x = diff.orig.smooth)) + geom_histogram() + geom_rug(alpha = 0.1) + facet_grid(Sex2~Race2)
+ggplot(subset(dat.clean, State2 == "California" & Sex2 == "Male" & Race2 == "Black"), aes(x = diff.orig.smooth)) + geom_histogram() + geom_rug(alpha = 0.1) + facet_wrap(~Age3)
+
+ggplot(subset(dat.clean, State2 == "Utah"), aes(x = diff.orig.smooth)) + geom_density() + geom_rug(alpha = 0.1) + facet_grid(Sex2~Race2)
+ggplot(subset(dat.clean, State2 == "Utah" & Sex2 == "Male" & Race2 == "Black"), aes(x = diff.orig.smooth)) + geom_histogram() + geom_rug(alpha = 0.1) + facet_wrap(~Age3)
 ```
@@
 ```{r calculate_life_expectancy}
-#6 mins
+#15 mins
 
@@
 ```{r calculate_age_decomp_smoothed}
-#18 mins
+#20 mins
 
@@
 ```{r calculate_cod_decomp}
-#takes 40 mins
+#takes 84 mins
 

### 2016-11-22 Code/Examine_relation_num_post_samps_and_CI.R
plots num posteriors samps vs CI

@@ -1,0 +1,55 @@
+#explore the relationship between the CI estimates and CI width as a function of the number of draws of the posterior
+#distribution.
+
+load("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/CI_vs_post_samples_results.RData")
+datasets <- c(georgia_2013_females[[3]], missouri_1985_females[[3]], utah_1969_females[[3]])
+
+temp <- georgia_2013_females[[3]] 
+temp <- temp %>% mutate(CI_width_LEW = run_975_LEW - run_025_LEW,
+                        CI_width_LEB = run_975_LEB - run_025_LEB,
+                        CI_width_LEG = run_975_LEG - run_025_LEG)
+head(temp)
+georgia_2013_females[[3]] <- temp
+
+temp <- missouri_1985_females[[3]] 
+temp <- temp %>% mutate(CI_width_LEW = run_975_LEW - run_025_LEW,
+                        CI_width_LEB = run_975_LEB - run_025_LEB,
+                        CI_width_LEG = run_975_LEG - run_025_LEG)
+head(temp)
+missouri_1985_females[[3]] <- temp
+
+temp <- utah_1969_females[[3]] 
+temp <- temp %>% mutate(CI_width_LEW = run_975_LEW - run_025_LEW,
+                        CI_width_LEB = run_975_LEB - run_025_LEB,
+                        CI_width_LEG = run_975_LEG - run_025_LEG)
+head(temp)
+utah_1969_females[[3]] <- temp
+
+#Georgia CI widths
+ggplot(georgia_2013_females[[3]][10:1000, ], aes(x = row, y = CI_width_LEW)) + geom_line() + 
+  geom_abline(intercept = georgia_2013_females[[3]]$CI_width_LEW[1000], slope = 0, lty = 2) + 
+  geom_line(aes(y = CI_width_LEB), col = "red") + 
+  geom_abline(intercept = georgia_2013_females[[3]]$CI_width_LEB[1000], slope = 0, lty = 2, col = "red")    
+
+ggplot(georgia_2013_females[[3]], aes(x = row, y = CI_width_LEG)) + geom_line(col = "blue") + 
+  geom_abline(intercept = georgia_2013_females[[3]]$CI_width_LEG[1000], slope = 0, lty = 2, col = "blue") 
+
+#Missouri CI widths
[…]

### 2016-11-22 Code/Explore_unaggregate_data.Rmd
Create AL cut of unaggregated dataset

@@
 ```{r load datasets}
+#loading in the data from NCHS, which supresses death counts between 1 and 9.
 dat.male <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/males_export.txt", header = T, sep="\t", na.strings = " ")
@@
 table(dat.male$Sex)
-str(dat.male) #note all numeric or int
+str(dat.male) #note all varinables are numeric or integer
 
@@
 dat.female$Sex <- 1
-str(dat.male) #note all numeric or int
+str(dat.male) #note all varinables are numeric or integer
 
@@
 dim(dat.both)
-
 ```
@@
 ```{r for_merge}
-
 years0610 <- dat.both[dat.both$Year3 >= 2006 & dat.both$Year3 <= 2010 & dat.both$agegrp > 1, ] 
@@
 together <- merge(years0610, not.suppressed, by = c("stabbrs", "Year3", "Sex", "Race", "agegrp", "COD_full"))
-
 ```
@@
 From above we see that the death counts match exactly and when the original count is NA the non-suppressed count is between 1 and 9.
-For some reason there is some discrpancy in the populational denominators. These do not match exactly and I don't know why.
+For some reason there is some discrepancy in the population denominators. These do not match exactly and I don't know why.
 

### 2016-11-22 Code/Prepare_Validation_Dataset.R

@@ -1,8 +1,6 @@
 
 not.suppressed <- foreign::read.dta("/Users/corinneriddell/Dropbox/VitalStats/notsuppressed_0610.dta")
 
-not.suppressed$cod_cat[not.suppressed$COD_full == "In situ, benign or unknown behavior neoplasm"] <- "All other causes"
-
 not.suppressed.grouped <- not.suppressed %>% 
   filter(agegrp > 1) %>%
   group_by(stabbrs, Year3, agegrp, Sex, Race, cod_cat) %>%
@@ -11,7 +9,6 @@ not.suppressed.grouped <- not.suppressed %>%
             division = first(division)) %>%
   mutate(COD2 = factor(cod_cat, levels = c("Cardiovascular", "Cancers", "Communicable", "Non-communicable", 
                                            "Injuries","All other causes")))
-
 
 head(not.suppressed.grouped)
 

### 2016-11-22 Code/append_smoothed_results.Rmd
made marginal COD tables

@@
 length(dat.clean$total_smoothed_deaths[dat.clean$total_smoothed_deaths < 1 & dat.clean$Appended_smoothed == T])
+
+View(dat.clean[dat.clean$total_smoothed_deaths < 1 & dat.clean$Appended_smoothed == T, ])
 ```
@@
 
+```{r compute_marginal_COD_decomposition}
+list.cod.marginal.tables.smoothed <- vector(mode = "list", length = dim(paired.ids)[1])
+
+for(i in 1:dim(paired.ids)[1]) {
+    if(list.cod.decomp.tables.smoothed[[i]][1] != "Couldn't compare life tables" & 
+       list.cod.decomp.tables.smoothed[[i]][1] != "Encountered age group with no deaths") {
+
+       list.cod.marginal.tables.smoothed[[i]] <- list.cod.decomp.tables.smoothed[[i]] %>% 
+                                                  group_by(Cause.of.death) %>% 
+                                                  summarise(C_x_COD = sum(C_xi))
+
+    }
+}
+```
+
 ```{r save_workspace}
@@
      list.age.decomp.tables5, list.cod.decomp.tables5, list.age.decomp.tables9, list.cod.decomp.tables9,
-     list.age.decomp.tables.smoothed, list.cod.decomp.tables.smoothed,
+     list.age.decomp.tables.smoothed, list.cod.decomp.tables.smoothed, list.cod.marginal.tables.smoothed,
      file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")

### 2016-11-22 Code/life_expectancy_functions.R

@@ -151,16 +151,19 @@ lt.white <- life.table(data = subset(data.aggregated, race == "White"), num.ages
 return(data.frame("LE_Black" = lt.black$e_x[1], "LE_White" = lt.white$e_x[1], "LE_WBgap" = lt.white$e_x[1] - lt.black$e_x[1]))
 }
 
-
-#ktm_smoothed <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/smoothed_results.csv")
-#ktm_sub <- ktm_smoothed %>% filter(state == "Alabama" & sex == "Female" & year == 45)
-
-#life_expectancy_and_gap(ktm_sub)
-
 #contribution to the change in the gap
-#age.contribution.to.gap <- function(decomp.table1, decomp.table2) {
-   
-# }
-#by age
-
-#by COD
+contribution.to.gap.change <- function(type.of.decomp, decomp.table1, decomp.table2) {
+  if(type.of.decomp == "Age") {
+    contribution_data <- data.frame("Ages" = decomp.table1[["Ages"]], 
+                                    "Contribution.to.change" = decomp.table1[["C_x"]] - decomp.table2[["C_x"]])
+  }
+  
+  if(type.of.decomp == "COD") {
+     contribution_data <- data.frame("COD" = decomp.table1[["Cause.of.death"]], 
+                                     "Contribution.to.change" = decomp.table1[["C_x_COD"]] - decomp.table2[["C_x_COD"]])
+  }
+  
+  contribution_data[["narrowed_gap"]] <- contribution_data[["Contribution.to.change"]] >= 0 
+  
+  return(contribution_data)
+}

### 2016-11-23 Code/append_smoothed_results.Rmd
calculate proportion contribution

@@
                                                   group_by(Cause.of.death) %>% 
-                                                  summarise(C_x_COD = sum(C_xi))
+                                                  summarise(C_x_COD = sum(C_xi)) %>%
+                                                  mutate(C_x_COD_proportion = C_x_COD/sum(C_x_COD))
 

### 2016-11-23 Code/life_expectancy_functions.R

@@ -55,7 +55,7 @@ le_age_decomp <- function(life.table1, name.lt1 = 1, life.table2, name.lt2 = 2, 
   life.table2["num.alive.next.interval"] <- c(unlist(life.table2[2:dim(life.table2)[1], "l_x"]), 1) #--> end in 1 to fix calculation error of dividing by 0
 
   decomp.table <- data.frame("Ages" = life.table1[ ,age.groups], life.table1["e_x"], life.table2["e_x"],
-                             "C_x" = rep(NA, dim(life.table1)[1]))
+                             "C_x" = rep(NA, dim(life.table1)[1]), "C_x_proportion" = rep(NA, dim(life.table1)[1]))
   
   names(decomp.table)[2] <- paste0("LE_", name.lt1)
   names(decomp.table)[3] <- paste0("LE_", name.lt2)
@@ -65,6 +65,8 @@ le_age_decomp <- function(life.table1, name.lt1 = 1, life.table2, name.lt2 = 2, 
     ((life.table2["accumulated.lived.after"]/life.table2[1, "l_x"])*
      ((life.table1["l_x"]/life.table2["l_x"]) - (life.table1["num.alive.next.interval"]/life.table2["num.alive.next.interval"]))
      )
+  
+  decomp.table["C_x_proportion"] <- decomp.table["C_x"]/sum(decomp.table["C_x"])
   
   decomp.table["adds_to_gap"] <- ifelse(decomp.table$C_x > 0, T, F)
 
@@ -91,7 +93,9 @@ cause_of_death_decomp <- function(life.table1, life.table2, decomp.table,
                                  cod.table[ , prop2.colname.cod.table],
                                  "Cause.of.death" = cod.table[ , COD.colname.cod.table],
                                  "C_xi" = C_xi)
-
+  
+  COD.decomp.table["C_xi_proportion"] <- COD.decomp.table["C_xi"]/sum(COD.decomp.table["C_xi"])
+  
   names(COD.decomp.table)[2] <- prop1.colname.cod.table
   names(COD.decomp.table)[3] <- prop2.colname.cod.table
   
@@ -112,6 +116,9 @@ make_dataset_cod_plot <- function(cod.decomp.table, age.groups, cause.of.death) 
   cod.decomp.table$start <- 0
   cod.decomp.table$finish <- cod.decomp.table$C_xi
   
+  cod.decomp.table$start2 <- 0
[…]

### 2016-11-24 Code/life_expectancy_functions.R
enhancing app interactivity

@@ -165,12 +165,14 @@ return(data.frame("LE_Black" = lt.black$e_x[1], "LE_White" = lt.white$e_x[1], "L
 contribution.to.gap.change <- function(type.of.decomp, decomp.table1, decomp.table2) {
   if(type.of.decomp == "Age") {
     contribution_data <- data.frame("Ages" = decomp.table1[["Ages"]], 
-                                    "Contribution.to.change" = decomp.table1[["C_x"]] - decomp.table2[["C_x"]])
+                                    "Contribution.to.change" = decomp.table1[["C_x"]] - decomp.table2[["C_x"]],
+                                    "Contrib.to.change.prop" = decomp.table1[["C_x_proportion"]] - decomp.table2[["C_x_proportion"]])
   }
   
   if(type.of.decomp == "COD") {
      contribution_data <- data.frame("COD" = decomp.table1[["Cause.of.death"]], 
-                                     "Contribution.to.change" = decomp.table1[["C_x_COD"]] - decomp.table2[["C_x_COD"]])
+                                     "Contribution.to.change" = decomp.table1[["C_x_COD"]] - decomp.table2[["C_x_COD"]],
+                                     "Contrib.to.change.prop" = decomp.table1[["C_x_COD_proportion"]] - decomp.table2[["C_x_COD_proportion"]])
   }
   
   contribution_data[["narrowed_gap"]] <- contribution_data[["Contribution.to.change"]] >= 0 

### 2016-11-24 shinyapp3/app.R

@@ -4,6 +4,7 @@ library(viridis)
 library(scales)
 library(shinythemes)
 load(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
 
 ui1 <- fluidPage(#theme = shinytheme("cosmo"),
                  #shinythemes::themeSelector(), 
@@ -13,30 +14,41 @@ ui1 <- fluidPage(#theme = shinytheme("cosmo"),
                    sidebarPanel(width = 4,
                      selectInput(inputId = "state", label = "Select a state", width = 200, choices = unique(levels(dat.clean$State2))),
                      #strong("Select years to compare"),
-                     strong("Select year for decomposition"),
+                     strong("Select years for decomposition"),
                      tags$style(type="text/css", ".form-group.shiny-input-container{ display: inline-block } strong{ display: block !important  } img{ margin-bottom: 30px }"),
                   
                      selectInput(inputId = "year1", label = NA, choices = unique(dat.clean$Year3), width = 100),
-                     #selectInput(inputId = "year2", label = NA, choices = unique(dat.clean$Year3), selected = 2013, width = 100),
+                     selectInput(inputId = "year2", label = NA, choices = unique(dat.clean$Year3), selected = 2013, width = 100),
                      
                      strong("Legend\n"),
                      img(src="legend.png"),
                      strong("Population size over time"),
-                     plotlyOutput("population_trend")
+                     plotlyOutput("population_trend"),
+                     radioButtons(inputId = "LE_type", label = "Model choice", inline = T, 
+                                  choices = c("Impute 1", "Impute 5", "Impute 9"), 
+                                  selected = "Impute 5"),
+                     radioButtons(inputId = "contribution_type", label = "Display contribution in", inline = T, choices = c("Years", "Proportion (%)"), selected = "Years")
+                  
                       ),
                    mainPanel(
-                     radioButtons(inputId = "LE_type", label = "Model choice", inline = T, 
-                                  choices = c("Bayesian Smoothing", "Impute 1", "Impute 5", "Impute 9"), 
-                                  selected = "Impute 5"),
-                     plotlyOutput("life_expectancy"),
[…]

### 2016-11-24 Code/append_smoothed_results.Rmd

--- 2016-11-22 Code/append_smoothed_results.Rmd
@@
                                                           life.table2 = as.data.frame(list.life.tables.smoothed[[white.index]]),
-                                                          name.lt2 = "White", age.groups = "Age3")
+                                                          name.lt2 = "White", age.groups = "Age2")
   } else {
@@
                 select(Age2, COD2, cod_prop_smoothed_deaths) %>%
-                rename(prop.white = cod_prop_smoothed_deaths)
+                dplyr::rename(prop.white = cod_prop_smoothed_deaths)
 
@@
                 select(Age2, COD2, cod_prop_smoothed_deaths) %>%
-                rename(prop.black = cod_prop_smoothed_deaths)
+                dplyr::rename(prop.black = cod_prop_smoothed_deaths)
     ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
@@
        list.cod.decomp.tables.smoothed[[i]][1] != "Encountered age group with no deaths") { 
-
        list.cod.marginal.tables.smoothed[[i]] <-  list.cod.decomp.tables.smoothed[[i]] %>%
-                                                  group_by(Cause.of.death) %>% 
-                                                  summarise(C_x_COD = sum(C_xi))
+                                                  dplyr::group_by(Cause.of.death) %>% 
+                                                  dplyr::summarise(C_x_COD = sum(C_xi)) %>%
+                                                  dplyr::mutate(C_x_COD_proportion = C_x_COD/sum(C_x_COD),
+                                                         adds_to_gap = ifelse(C_x_COD > 0, T, F))
 
@@
      list.age.decomp.tables.smoothed, list.cod.decomp.tables.smoothed, list.cod.marginal.tables.smoothed,
+     non.zero.stratums, non.zero.stratums.smoothed,
      file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")

### 2016-12-02 shinyapp3/app.R
made summary tab better

@@ -163,10 +163,13 @@ server <- function(input, output) {
       #          theme_minimal()) %>% layout(xaxis = list(title = "Life expectancy gap (years)"), yaxis = list(title = NA, autorange = "reversed")
       #                                      )
       ggplot(summary.cod.contrib.data.react(), aes(y = order.states, x = new.start)) +
-        geom_rect(aes(xmin = new.start, ymin = order.states, ymax = order.states + 0.75, xmax = new.finish, fill = COD), color = "white") + #, arrow = arrow(angle = 30, ends = "last", length = unit(0.10, "inches"))
-        geom_rect(aes(xmin = second.gap, xmax = second.gap + 0.03, ymin = order.states - 0.2, ymax = order.states + 0.75), fill = "grey40", color = "grey40") + #will need to change this - adds lots of points on top of each other
-        geom_rect(aes(xmin = first.gap, xmax = first.gap + 0.03, ymin = order.states - 0.2, ymax = order.states + 0.75), fill = "black", color = "black") + #will need to change this - adds lots of points on top of each other
-        theme_minimal() + scale_y_continuous(breaks = 1:51, labels = levels(summary.cod.contrib.data.react()$order.states2)) )%>% 
+        geom_rect(aes(xmin = new.start, ymin = order.states - 0.45, ymax = order.states + 0.45, xmax = new.finish, fill = COD), color = "white") + #, arrow = arrow(angle = 30, ends = "last", length = unit(0.10, "inches"))
+        geom_segment(aes(x = second.gap, xend= second.gap, y = order.states - 0.5, yend = order.states + 0.5), lty = 3) + #will need to change this - adds lots of points on top of each other
+        #geom_rect(aes(xmin = second.gap, xmax = second.gap + 0.03, ymin = order.states - 0.2, ymax = order.states + 0.75), fill = "grey40", color = "grey40") + #will need to change this - adds lots of points on top of each other
+        #geom_rect(aes(xmin = first.gap, xmax = first.gap + 0.03, ymin = order.states - 0.2, ymax = order.states + 0.75), fill = "black", color = "black") + #will need to change this - adds lots of points on top of each other
+        geom_segment(aes(x = first.gap, xend= first.gap, y = order.states - 0.5, yend = order.states + 0.5)) + #will need to change this - adds lots of points on top of each other
+        
[…]

### 2016-12-02 Code/1_Explore_Data_v2.Rmd
generalized code for new make_cod_plot fx

@@
                                                             age.groups = "Ages", 
-                                                            cause.of.death = "Cause.of.death")
+                                                            cause.of.death = "Cause.of.death", 
+                                                            sign.var = "sign", decomp.var = "C_xi", 
+                                                            decomp.var.prop = "C_xi_proportion")
 
@@
                                                             age.groups = "Ages", 
-                                                            cause.of.death = "Cause.of.death") 
+                                                            cause.of.death = "Cause.of.death", 
+                                                            sign.var = "sign", decomp.var = "C_xi", 
+                                                            decomp.var.prop = "C_xi_proportion") 
     } 

### 2016-12-02 Code/append_smoothed_results.Rmd
generalized code for new make_cod_plot fx

@@
                                                             age.groups = "Ages", 
-                                                            cause.of.death = "Cause.of.death")  
+                                                            cause.of.death = "Cause.of.death",
+                                                            sign.var = "sign", decomp.var = "C_xi", 
+                                                            decomp.var.prop = "C_xi_proportion")  
     } 

### 2016-12-14 Code/Prep SH HA data for our analysis.Rmd
prep SH HA data for our modeling

---
title: "Compare with SH HA paper"
author: "Corinne Riddell"
date: "December 13, 2016"
output: html_document
---

```{r}
load("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/SH_HA_data.Rdata")
names(dtotal)
names(dtotal) <- c("State", "Age", "Sex", "Race", "Year", "Crude_Rate", "Count", "Population", "COD")
```

```{r recode_variables}
library(dplyr)

dtotal <- dtotal %>%
  mutate(State2 = factor(State, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
                                                             "Colorado", "Connecticut", "Delaware", "Washington DC",
                                                             "Florida", "Georgia", "Hawaii","Idaho", "Illinois", 
                                                             "Indiana","Iowa", "Kansas", "Kentucky", "Louisiana", 
                                                             "Maine", "Maryland", "Massachusetts", "Michigan", 
                                                             "Minnesota", "Mississippi","Missouri", "Montana", 
                                                             "Nebraska", "Nevada", "New Hampshire", "New Jersey",
                                                             "New Mexico", "New York","North Carolina","North Dakota",
                                                             "Ohio", "Oklahoma","Oregon", "Pennsylvania", 
                                                             "Rhode Island", "South Carolina", "South Dakota", 
                                                             "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
                                                             "Washington", "West Virginia", "Wisconsin", "Wyoming")), 
         Age2 = factor(Age, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                         "15-19 years", "20-24 years", "25-29 years", "30-34 years",
                                                         "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
                                                         "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
                                                         "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
         Age3 = 0*(Age2 == "<1 year") + 1*(Age2 == "1-4 years") + 5*(Age2 == "5-9 years") +
                10*(Age2 == "10-14 years") + 15*(Age2 == "15-19 years") + 20*(Age2 == "20-24 years") + 
                25*(Age2 == "25-29 years") + 30*(Age2 == "30-34 years") + 35*(Age2 == "35-39 years") +
                40*(Age2 == "40-44 years") + 45*(Age2 == "45-49 years") + 50*(Age2 == "50-54 years") + 
                55*(Age2 == "55-59 years") + 60*(Age2 == "60-64 years") + 65*(Age2 == "65-69 years") +   
                70*(Age2 == "70-74 years") + 75*(Age2 == "75-79 years") + 80*(Age2 == "80-84 years") + 
                85*(Age2 == "85+ years"),
         nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years"),
         Sex2 = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
         Race2 = factor(Race, levels = c(0, 1), labels = c("White", "Black")),
         COD2 = factor(COD), 
         Year2 = factor(Year, levels = c(0:44), labels = as.character(1969:2013)),
         Year3 = as.numeric(as.character(Year2)),
         RaceSex = interaction(Race2, Sex2, sep = " "),
         StateYearSex = interaction(State2, Year2, Sex2),
         StateYearRaceSex = interaction(State2, Year2, Race2, Sex2))
```

```{r unknown_age_group}
#there is an unknown age group for each combo of state-year-race-sex-age:
table(dtotal$Age2)

#Count of death is always equal to 0 for unknown age groups
table(dtotal$Age, dtotal$Count==0)

#remove these data from the data we analyze:
dtotal <- dtotal[dtotal$Age2 != "Unknown", ]

save(dtotal, file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/SH_HA_data.Rdata")
```

### 2016-12-14 Manuscript/Manuscript.Rmd
manuscript in rmd

---
title: "Exploring how changes in age at death and cause of death have narrowed the black-white life expectancy gap over forty-five years in the United States."
output: html_notebook
---

##Riddell CA, Morrison KT, Kaufman JS, Harper S

###Abstract

###Introduction
The life expectancy of a population serves as a barometer for overall health. In 2014, life expectancy in the United States was 78.7 years, but hidden within this figure, are differences in life expectancy by race. Non-hispanic whites lived 3.6 years longer than non-hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white americans have experienced increases in life expectancy since X, the difference between their life expectancies has varied. During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, heart disease [Harper 2007]. The gap continued to narrow during the 2000s, but states differed in the magnitude of improvement that was made. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014]. This study seeks to better understand these state-level differences, by examining how changes in age at death and cause of death over time have worked to narrow or exacerbate the life expectancy gap.

###Methods

**The data**

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEER*Stat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia. The mortality data was originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data was derived by the NCHS using intercensal bridged-race population estimates. This study using data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using The International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were then grouped into 9X causes of death within the software, which we further aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes. This recoding is illustrated in eTable 1 in the eAppendix of the Supplement. When the number of deaths for a population is between 1 and 9, this data is suppressed by the software and produces a missing death count for that strata. However, death counts of exactly 0 or greater than 9 are known precisely.

**Statistical Methods**

The data produced two statistical issues that were addressed by the statistical methods we chose. First, our methodology needed to handle the missing death counts. Second, several states had small blacks populations which produced even smaller population sizes in each sex- and age-specific grouping. In these cases, the estimation of the age-specific mortality rates is imprecise, and produces abrupt changes in the estimation of mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

We used a bayesian model that handled the missing data and smoothing ....

- bayesian methodology to smooth over year
- bayesian methodology to sample from truncated poisson distribution
- produces smoothed death counts 
- sampled from posterior distribution

- data was then aggregated over causes of death to get the total death count and this data was used to calculation LE and to perform the decomposition by age.
- standard life table methods to calculate life expectancy
- gap in life expectancy in years was the difference between White and Black LE and was calculated for each year and sex
- Arriaga's decomposition method to calculate the contribution of age and COD to the gap in years
- Gives the contribution in years, can also calculation the proportional contribution.
- to compare changes over time on the state level - can take the difference (in years) of the two contributions. Can also get the proportional contribution by...

-Software:

###Supplemental Information

**eTable1: Cause of Death Recode**

| Broad cause of death     | Included causes of death                                                     |
|--------------------------|------------------------------------------------------------------------------|
| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer. |
| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries. | 
| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Sydrome and Nephrosis.  |
| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death. |

```{r}
plot(cars)
```

### 2016-12-14 shinyapp main/app.R
manuscripting

@@ -3,6 +3,7 @@ library(plotly)
 library(viridis)
 library(scales)
 library(shinythemes)
+library(dplyr)
 load(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
 source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
 

### 2016-12-14 Manuscript/Manuscript.Rmd

@@
 title: "Exploring how changes in age at death and cause of death have narrowed the black-white life expectancy gap over forty-five years in the United States."
-output: html_notebook
+output: html_document
 ---
@@
 This study using data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death.
-Cause of death was originally recorded using The International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013).
+Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013).
 These codes were then grouped into 9X causes of death within the software, which we further aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes.
@@
 
+- CR to add line about which states were not considered.
+- CR to add line about hispanic origin.
+
 **Statistical Methods**
@@
 The data produced two statistical issues that were addressed by the statistical methods we chose.
-First, our methodology needed to handle the missing death counts.
+First, our methodology needed to impute the missing death counts.
 Second, several states had small blacks populations which produced even smaller population sizes in each sex- and age-specific grouping.
@@
 
-We used a bayesian model that handled the missing data and smoothing ....
+KTM to add bayesian modeling paragraph
 
-- bayesian methodology to smooth over year
-- bayesian methodology to sample from truncated poisson distribution
-- produces smoothed death counts 
-- sampled from posterior distribution
+- type of bayesian model used and what was modeled Y vs X, separately for each state, race, and sex
+- how the smoothing happens over year
+- how the truncated poisson is involved
+- this methodology produces smoothed death counts 
+- sampled from posterior distribution with thinning
+- software: R software packages and their versions
 
-- data was then aggregated over causes of death to get the total death count and this data was used to calculation LE and to perform the decomposition by age.
-- standard life table methods to calculate life expectancy
-- gap in life expectancy in years was the difference between White and Black LE and was calculated for each year and sex
-- Arriaga's decomposition method to calculate the contribution of age and COD to the gap in years
-- Gives the contribution in years, can also calculation the proportional contribution.
-- to compare changes over time on the state level - can take the difference (in years) of the two contributions. Can also get the proportional contribution by...
+Thus, the model was run separately for every combination of state, race, and sex, and produced death counts that had been smoothed over time for every cause of death within each age group.
+To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref].
+Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex.
+The difference in life expectancies between blacks and white was calculated by subtracting the black life expectancy from the white life expectancy in every state, year, and sex.
 
--Software:
+We then performed a decomposition analysis using the Arriaga method, to decompose the black-white difference in life expectancy according to age group and cause of death [ref].
+The decomposition analysis partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher mortality rates for each age group.
+For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites.
+The age-specific decomposition is then further partitioned according to cause of death, producing the number of years of the gap that is attributable to each age-cause-of-death combination.
+For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old.
+We also calculated the marginal contribution of each cause of death by summing the cause-specific contributions across age groups.
+
+Finally, we calculated the change in the life expectancy gap over time within each state-, race-, and sex-specific stratum.
+The change in the life expectancy gap is calculated by substracting the gap that occurred in the later year from the gap that occurred in the earlier year.
+This change in gap can be decomposed by age by taking the difference in the age-specfic contributions across the two years, and can similarly be calculated by cause.
+
+All data analyses were performed using R version 3.3.1.
+The bayesian smoothing was performed using `JAGS` (version X.X), and the life expectnacy and decomposition analysis were performed using user-written functions found in the Supplement.
+
+###Results
+
+```{r initialize, echo=FALSE, warning=F, message=F}
+library(shiny)
+library(plotly)
+library(viridis)
+library(scales)
+library(shinythemes)
+library(dplyr)
+load(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+```
+
+```{r setup_figure1_and_2, echo=F}
+year1 <- 1969
+year2 <- 2013
+sex <- "Male"
+
+bw_dat <- BlackWhite %>% 
+  filter(Year3 %in% c(year1, year2), Sex2 == sex, is.na(WBgap.s) == F) %>%
+  select(State2, Year3, WBgap.s) %>%
+  arrange(Year3)
+
+bw_dat <- tidyr::spread(bw_dat, Year3, WBgap.s) 
+names(bw_dat)[2:3] <- c("first.gap","second.gap")  
+bw_dat <- bw_dat %>% mutate(gap.diff = first.gap - second.gap,
+                            State.gdiff.order = reorder(State2, first.gap),
+                            State.gdiff.num = as.numeric(State.gdiff.order)
+)
+
+cod.df <- data.frame(State2 = factor(), COD = factor(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
+age.df <- data.frame(State2 = factor(), Ages = factor(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
+
+for (state.i in levels(BlackWhite$State2)) {
+  first.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year1 & paired.ids$Sex2 == sex)
+  second.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year2 & paired.ids$Sex2 == sex)
+
+  if(is.null(list.cod.marginal.tables.smoothed[[first.index]][1]) == FALSE & 
+     is.null(list.cod.marginal.tables.smoothed[[second.index]][1]) == FALSE) {
+
+    temp2 <- data.frame(State2 = state.i,
+                        contribution.to.gap.change(type.of.decomp = "COD",
+                                                   list.cod.marginal.tables.smoothed[[first.index]],
+                                                   list.cod.marginal.tables.smoothed[[second.index]])
+    )
+
+    temp3 <- data.frame(State2 = state.i,
+                        contribution.to.gap.change(type.of.decomp = "Age",
+                                                   list.age.decomp.tables.smoothed[[first.index]],
+                                                   list.age.decomp.tables.smoothed[[second.index]])
+    )
+
+    cod.df <- rbind(cod.df, temp2)
+    age.df <- rbind(age.df, temp3)
+  }
+}
+
+##### COD ####
+cod.df <- make_dataset_cod_plot(cod.decomp.table = cod.df, age.groups = "State2", 
+                                 cause.of.death = "COD", sign.var = "narrowed_gap", 
+                                 decomp.var = "Contribution.to.change", decomp.var.prop = "Contrib.to.change.prop")
+
+cod.df <- merge(cod.df, bw_dat, by = "State2")
+
+cod.df <- cod.df %>% mutate(new.start = -start + first.gap, new.finish = -finish + first.gap)
+
+#### Age ####
+age.df <- make_dataset_cod_plot(cod.decomp.table = age.df, age.groups = "State2", 
+                                  cause.of.death = "Ages", sign.var = "narrowed_gap", 
+                                  decomp.var = "Contribution.to.change", decomp.var.prop = "Contrib.to.change.prop")
+
+age.df <- merge(age.df, bw_dat, by = "State2")
+
+age.df <- age.df %>% mutate(new.start = -start + first.gap, new.finish = -finish + first.gap)
+```
+
+**Figure 1: Decomposition of the change in the black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
+```{r figure1, fig.align = "center", fig.height=8, fig.width=8, echo=F}
+x.labels <- levels(factor(cod.df$State.gdiff.order))
+x.num.breaks <- length(levels(factor(cod.df$State.gdiff.order)))
+
+ggplotly(
+  ggplot(cod.df, aes(y = State.gdiff.num, x = new.start)) +
+    geom_rect(aes(xmin = new.start, ymin = State.gdiff.num - 0.45, 
+                  ymax = State.gdiff.num + 0.45, xmax = new.finish, fill = COD), color = "white") +
+    geom_segment(data = bw_dat, aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) +
+    geom_segment(data = bw_dat, aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks, labels = x.labels)) %>% 
+  layout(xaxis = list(title = "Life expectancy gap (years)"), yaxis = list(title = NA, autorange = "reversed"))
+
+```
+
+**Figure 2: Decomposition of the change in the black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
+```{r figure2, fig.align = "center", fig.height=8, fig.width=9, echo=F}
+ggplotly(                                     
+  ggplot(age.df, aes(y = State.gdiff.num, x = new.start)) +
+    geom_rect(aes(xmin = new.start, ymin = State.gdiff.num - 0.45, ymax = State.gdiff.num + 0.45, xmax = new.finish, fill = Ages), color = "white") + 
+    geom_segment(data = bw_dat, aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) + 
+    geom_segment(data = bw_dat, aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) + 
+    theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks, labels = x.labels) + scale_fill_viridis(discrete = T, direction = -1)) %>% 
+  layout(xaxis = list(title = "Life expectancy gap (years)"), yaxis = list(title = NA, autorange = "reversed")
+  )
+```
+
+###Discussion
 
@@
 
-```{r}
-plot(cars)
-```
+**Life table and decomposition methodology**
+
+**Life table and decomposition user-written functions**

### 2016-12-14 Manuscript/Manuscript.Rmd
updated ms

@@
 year2 <- 2013
-sex <- "Male"
 
 bw_dat <- BlackWhite %>% 
-  filter(Year3 %in% c(year1, year2), Sex2 == sex, is.na(WBgap.s) == F) %>%
-  select(State2, Year3, WBgap.s) %>%
+  filter(Year3 %in% c(year1, year2), is.na(WBgap.s) == F) %>%
+  select(State2, Sex2, Year3, WBgap.s) %>%
   arrange(Year3)
@@
 bw_dat <- tidyr::spread(bw_dat, Year3, WBgap.s) 
-names(bw_dat)[2:3] <- c("first.gap","second.gap")  
+names(bw_dat)[3:4] <- c("first.gap","second.gap")  
 bw_dat <- bw_dat %>% mutate(gap.diff = first.gap - second.gap,
@@
 
-cod.df <- data.frame(State2 = factor(), COD = factor(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
-age.df <- data.frame(State2 = factor(), Ages = factor(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
+cod.df <- data.frame(State2 = factor(), Sex2 = factor(), COD = factor(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
+age.df <- data.frame(State2 = factor(), Sex2 = factor(), Ages = factor(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
 
 for (state.i in levels(BlackWhite$State2)) {
-  first.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year1 & paired.ids$Sex2 == sex)
-  second.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year2 & paired.ids$Sex2 == sex)
+
+  for (sex.i in c("Male", "Female")) {
+  first.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year1 & paired.ids$Sex2 == sex.i)
+  second.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year2 & paired.ids$Sex2 == sex.i)
 
@@
 
-    temp2 <- data.frame(State2 = state.i,
+    temp2 <- data.frame(State2 = state.i, Sex2 = sex.i,
                         contribution.to.gap.change(type.of.decomp = "COD",
@@
 
-    temp3 <- data.frame(State2 = state.i,
+    temp3 <- data.frame(State2 = state.i, Sex2 = sex.i,
                         contribution.to.gap.change(type.of.decomp = "Age",
@@
   }
+}
 
 ##### COD ####
-cod.df <- make_dataset_cod_plot(cod.decomp.table = cod.df, age.groups = "State2", 
+cod.df.male <- make_dataset_cod_plot(cod.decomp.table = subset(cod.df, Sex2 == "Male"), age.groups = "State2", 
                                  cause.of.death = "COD", sign.var = "narrowed_gap", 
@@
 
-cod.df <- merge(cod.df, bw_dat, by = "State2")
+cod.df.female <- make_dataset_cod_plot(cod.decomp.table = subset(cod.df, Sex2 == "Female"), age.groups = "State2", 
+                                 cause.of.death = "COD", sign.var = "narrowed_gap", 
+                                 decomp.var = "Contribution.to.change", decomp.var.prop = "Contrib.to.change.prop")
+
+cod.df <- merge(rbind(cod.df.male, cod.df.female), bw_dat, by = c("State2", "Sex2"))
 
@@
 #### Age ####
-age.df <- make_dataset_cod_plot(cod.decomp.table = age.df, age.groups = "State2", 
+age.df.male <- make_dataset_cod_plot(cod.decomp.table = subset(age.df, Sex2 == "Male"), age.groups = "State2", 
+                                  cause.of.death = "Ages", sign.var = "narrowed_gap", 
+                                  decomp.var = "Contribution.to.change", decomp.var.prop = "Contrib.to.change.prop")
+age.df.female <- make_dataset_cod_plot(cod.decomp.table = subset(age.df, Sex2 == "Female"), age.groups = "State2", 
                                   cause.of.death = "Ages", sign.var = "narrowed_gap", 
@@
 
-age.df <- merge(age.df, bw_dat, by = "State2")
+age.df <- merge(rbind(age.df.male, age.df.female), bw_dat, by = c("State2", "Sex2"))
 
@@
 
-**Figure 1: Decomposition of the change in the black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
+Update the below text to also represent women:
+**Figure 1** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs.
+their gap in 2013 (solid dashed line), with the change in the gap being represented by the number of years between these lines.
+The decomposition (in years) of the change in the gap by cause of death is also shown for each state.
+In the US, X of Y years of the change in gap is attributable to injuries, implying relatively more favourable *changes* to injury-related mortality rates in blacks vs.
+whites.
+The next large contributor to reducing the overall gap was "All other causes", which largely represents infant mortality and maternal mortality in our analysis (CR to confirm this).
+Blacks also saw relatively favourable changes in death related to communicable diseases, wher X of Y years of the change in the gap is attributable to changes in communicable disease-related mortality.
+Overall, X of Y years of the gap is attributable to cardiovarscular disease, but the story is very different by state.
+For example, California and Washington both underwent large unfavourable changes to CVD-related mortality among blacks relative to whites, where these conditions served to limit the change in the gap in California (which would have been much larger had CVD-related mortality rates had changed more favourably) and led to a much larger gap in the District of Columbia in 2013 vs.
+1969, where X of Y years of the increase in the gap is attributable to CVD-related mortality.
+
+**Figure 1: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
 ```{r figure1, fig.align = "center", fig.height=8, fig.width=8, echo=F}
@@
 ggplotly(
-  ggplot(cod.df, aes(y = State.gdiff.num, x = new.start)) +
+  ggplot(subset(cod.df, Sex2 == "Male"), aes(y = State.gdiff.num, x = new.start)) +
     geom_rect(aes(xmin = new.start, ymin = State.gdiff.num - 0.45, 
                   ymax = State.gdiff.num + 0.45, xmax = new.finish, fill = COD), color = "white") +
-    geom_segment(data = bw_dat, aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) +
-    geom_segment(data = bw_dat, aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) +
+    geom_segment(data = subset(bw_dat, Sex2 == "Male"), aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) +
+    geom_segment(data = subset(bw_dat, Sex2 == "Male"), aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks, labels = x.labels)) %>% 
+  layout(xaxis = list(title = "Life expectancy gap (years)"), yaxis = list(title = NA, autorange = "reversed"))
+
+```
+**This figure depicts how the gap in life expectancy for Males changed between 1969 (dashed vertical line) and 2013 (solid vertical line).
+Causes to the left of the dashed line narrowed the gap over time, whereas causes to the right exacerbated it.**
+
+**Figure 2: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
+```{r figure2, fig.align = "center", fig.height=8, fig.width=8, echo=F}
+x.labels <- levels(factor(cod.df$State.gdiff.order))
+x.num.breaks <- length(levels(factor(cod.df$State.gdiff.order)))
+
+ggplotly(
+  ggplot(subset(cod.df, Sex2 == "Female"), aes(y = State.gdiff.num, x = new.start)) +
+    geom_rect(aes(xmin = new.start, ymin = State.gdiff.num - 0.45, 
+                  ymax = State.gdiff.num + 0.45, xmax = new.finish, fill = COD), color = "white") +
+    geom_segment(data = subset(bw_dat, Sex2 == "Female"), aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) +
+    geom_segment(data = subset(bw_dat, Sex2 == "Female"), aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) +
     theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks, labels = x.labels)) %>% 
@@
 
-**Figure 2: Decomposition of the change in the black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
-```{r figure2, fig.align = "center", fig.height=8, fig.width=9, echo=F}
+**Figure 3: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
+```{r figure3, fig.align = "center", fig.height=8, fig.width=9, echo=F}
 ggplotly(                                     
-  ggplot(age.df, aes(y = State.gdiff.num, x = new.start)) +
+  ggplot(subset(age.df, Sex2 == "Male"), aes(y = State.gdiff.num, x = new.start)) +
     geom_rect(aes(xmin = new.start, ymin = State.gdiff.num - 0.45, ymax = State.gdiff.num + 0.45, xmax = new.finish, fill = Ages), color = "white") + 
-    geom_segment(data = bw_dat, aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) + 
-    geom_segment(data = bw_dat, aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) + 
+    geom_segment(data = subset(bw_dat, Sex2 == "Male"), aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) + 
+    geom_segment(data = subset(bw_dat, Sex2 == "Male"), aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) + 
+    theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks, labels = x.labels) + scale_fill_viridis(discrete = T, direction = -1)) %>% 
+  layout(xaxis = list(title = "Life expectancy gap (years)"), yaxis = list(title = NA, autorange = "reversed")
+  )
+```
+
+**Figure 4: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
+```{r figure4, fig.align = "center", fig.height=8, fig.width=9, echo=F}
+ggplotly(                                     
+  ggplot(subset(age.df, Sex2 == "Female"), aes(y = State.gdiff.num, x = new.start)) +
+    geom_rect(aes(xmin = new.start, ymin = State.gdiff.num - 0.45, ymax = State.gdiff.num + 0.45, xmax = new.finish, fill = Ages), color = "white") + 
+    geom_segment(data = subset(bw_dat, Sex2 == "Female"), aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) + 
+    geom_segment(data = subset(bw_dat, Sex2 == "Female"), aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) + 
     theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks, labels = x.labels) + scale_fill_viridis(discrete = T, direction = -1)) %>% 

### 2016-12-15 Code/SH HA run model.R
fixing bayes functions

@@ -1,0 +1,100 @@
+
+setwd("~/BlackWhiteMortalityGap/Code/bayesian_smoothing")
+source('bayes_smoothing_src.R') 
+load("~/SH_HA_data.Rdata")
+
+
+excluded_states <- unique(dtotal$State2[dtotal$Population < 5]) 
+states <- unique(dtotal$State2)
+included_states <- setdiff(states, excluded_states) 
+
+dtotal.alabama <- subset(dtotal, State2 == "Alabama")
+cod_list <- c('Total')
+#cod_list <- c('Cancers')
+
+
+dtotal$COD3 <- "Cancers"
+dtotal$COD4 <- factor(dtotal$COD3)
+
+colname_sex = 'Sex2'
+colname_race = 'Race2'
+colname_pop = 'Population'
+colname_state = 'State2'
+colname_cod = 'COD'
+#colname_cod = 'COD4'
+colname_year = 'Year'
+colname_deaths = 'Count'
+colname_agebins = 'Age' 
+
+##########################################################################################
+# Kathryn shooting troubles  
+##########################################################################################
+
+ds_Al = dtotal.alabama[, c(colname_state, colname_pop, colname_deaths, colname_cod, 
+                       colname_year, colname_sex, colname_agebins, colname_race)]
+
+ds_T = dtotal[, c(colname_state, colname_pop, colname_deaths, colname_cod, 
+                           colname_year, colname_sex, colname_agebins, colname_race)]
+
+head(ds_Al)
+head(ds_T)
+
+tail(ds_Al)
+tail(ds_T)
+
+table(is.na(ds_Al$Count))
+table(is.na(ds_T$Count))
+
+setwd("~/black_white_mortality_project/SH_HA_compare")
+
+length(ds_Al$State2)
+
+states = 'Alabama'
+run_smoothing_models_mulitple_states(ds_T, states, cod_list, colname_state, colname_pop, colname_deaths, 
+                                     colname_cod, colname_year, colname_sex, colname_agebins, colname_race) 
+#Error in node binned.id_w[72015]
+# Well there should only be 1520 rows when subset by one state, so this is clearly not subsetting properly! And CR's manual subset fixes that issue.
+
+run_smoothing_models_mulitple_states(ds_Al, states, cod_list, colname_state, colname_pop, colname_deaths, 
+                                     colname_cod, colname_year, colname_sex, colname_agebins, colname_race) 
+str(ds_Al)
+str(ds_T)
+
[…]

### 2016-12-15 Code/bayesian_smoothing/bayes_smoothing_run.R

@@ -19,7 +19,7 @@ included_states = setdiff(states, excluded_states)
 cod_list = c('Injuries', 'Cardiovascular', 'Cancers', 'Communicable', 
              'Non-communicable', 'All other causes')
 
-#load('~/black_white_mortality_project/main_datasets.Rdata') # change for SERVER 
+load('~/black_white_mortality_project/main_datasets.Rdata') # change for SERVER 
 #load('/Users/kathryn/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata') # change for LOCAL 
 
 colname_sex = 'Sex2'
@@ -32,6 +32,9 @@ colname_deaths = 'Count'
 colname_agebins = 'Age' 
 
 
+mclapply(X = "Alabama", FUN=run_smoothing_models_mulitple_states, mc.cores=1, df=dat.clean, 
+                     cod_list, colname_state, colname_pop, colname_deaths, colname_cod, 
+                     colname_year, colname_sex, colname_agebins, colname_race) 
 
 #mclapply(X=included_states, FUN=run_smoothing_models_mulitple_states, mc.cores=1, df=dat.clean, 
 #         cod_list, colname_state, colname_pop, colname_deaths, colname_cod, 

### 2016-12-15 Manuscript/Manuscript.Rmd
added figures for females

--- 2016-12-14 Manuscript/Manuscript.Rmd
@@
 
-```{r setup_figure1_and_2, echo=F}
+```{r setup_figure1_and_3_males, echo=F}
 year1 <- 1969
@@
 for (state.i in levels(BlackWhite$State2)) {
+
   first.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year1 & paired.ids$Sex2 == sex)
@@
 
-    temp2 <- data.frame(State2 = state.i,
+    temp2 <- data.frame(State2 = state.i, Sex2 = sex,
                         contribution.to.gap.change(type.of.decomp = "COD",
@@
 
-    temp3 <- data.frame(State2 = state.i,
+    temp3 <- data.frame(State2 = state.i, Sex2 = sex,
                         contribution.to.gap.change(type.of.decomp = "Age",
@@
   }
+
 }
@@
 ##### COD ####
-cod.df <- make_dataset_cod_plot(cod.decomp.table = cod.df, age.groups = "State2", 
+cod.df<- make_dataset_cod_plot(cod.decomp.table = cod.df, age.groups = "State2", 
                                  cause.of.death = "COD", sign.var = "narrowed_gap", 
@@
 
-cod.df <- merge(cod.df, bw_dat, by = "State2")
+cod.df <- merge(cod.df, bw_dat, by = c("State2"))
 
@@
 
-**Figure 1: Decomposition of the change in the black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
+```{r setup_figure1_and_3_females, echo=F}
+sex <- "Female"
+
+bw_dat.f <- BlackWhite %>% 
+  filter(Year3 %in% c(year1, year2), Sex2 == sex, is.na(WBgap.s) == F) %>%
+  select(State2, Year3, WBgap.s) %>%
+  arrange(Year3)
+
+bw_dat.f <- tidyr::spread(bw_dat.f, Year3, WBgap.s) 
+names(bw_dat.f)[2:3] <- c("first.gap","second.gap")  
+bw_dat.f <- bw_dat.f %>% mutate(gap.diff = first.gap - second.gap,
+                            State.gdiff.order = reorder(State2, first.gap),
+                            State.gdiff.num = as.numeric(State.gdiff.order)
+)
+
+cod.df.f <- data.frame(State2 = factor(), COD = factor(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
+age.df.f <- data.frame(State2 = factor(), Ages = factor(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
+
+for (state.i in levels(BlackWhite$State2)) {
+
+  first.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year1 & paired.ids$Sex2 == sex)
+  second.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year2 & paired.ids$Sex2 == sex)
+
+  if(is.null(list.cod.marginal.tables.smoothed[[first.index]][1]) == FALSE & 
+     is.null(list.cod.marginal.tables.smoothed[[second.index]][1]) == FALSE) {
+
+    temp2 <- data.frame(State2 = state.i, Sex2 = sex,
+                        contribution.to.gap.change(type.of.decomp = "COD",
+                                                   list.cod.marginal.tables.smoothed[[first.index]],
+                                                   list.cod.marginal.tables.smoothed[[second.index]])
+    )
+
+    temp3 <- data.frame(State2 = state.i, Sex2 = sex,
+                        contribution.to.gap.change(type.of.decomp = "Age",
+                                                   list.age.decomp.tables.smoothed[[first.index]],
+                                                   list.age.decomp.tables.smoothed[[second.index]])
+    )
+
+    cod.df.f <- rbind(cod.df.f, temp2)
+    age.df.f <- rbind(age.df.f, temp3)
+  }
+
+}
+
+##### COD ####
+cod.df.f <- make_dataset_cod_plot(cod.decomp.table = cod.df.f, age.groups = "State2", 
+                                 cause.of.death = "COD", sign.var = "narrowed_gap", 
+                                 decomp.var = "Contribution.to.change", decomp.var.prop = "Contrib.to.change.prop")
+
+cod.df.f <- merge(cod.df.f, bw_dat.f, by = c("State2"))
+
+cod.df.f <- cod.df.f %>% mutate(new.start = -start + first.gap, new.finish = -finish + first.gap)
+
+#### Age ####
+age.df.f <- make_dataset_cod_plot(cod.decomp.table = age.df.f, age.groups = "State2", 
+                                  cause.of.death = "Ages", sign.var = "narrowed_gap", 
+                                  decomp.var = "Contribution.to.change", decomp.var.prop = "Contrib.to.change.prop")
+
+age.df.f <- merge(age.df.f, bw_dat.f, by = "State2")
+
+age.df.f <- age.df.f %>% mutate(new.start = -start + first.gap, new.finish = -finish + first.gap)
+```
+
+Update the below text to also represent women:
+**Figure 1** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs.
+their gap in 2013 (solid dashed line), with the change in the gap being represented by the number of years between these lines.
+The decomposition (in years) of the change in the gap by cause of death is also shown for each state.
+In the US, X of Y years of the change in gap is attributable to injuries, implying relatively more favourable *changes* to injury-related mortality rates in blacks vs.
+whites.
+The next large contributor to reducing the overall gap was "All other causes", which largely represents infant mortality and maternal mortality in our analysis (CR to confirm this).
+Blacks also saw relatively favourable changes in death related to communicable diseases, wher X of Y years of the change in the gap is attributable to changes in communicable disease-related mortality.
+Overall, X of Y years of the gap is attributable to cardiovarscular disease, but the story is very different by state.
+For example, California and Washington both underwent large unfavourable changes to CVD-related mortality among blacks relative to whites, where these conditions served to limit the change in the gap in California (which would have been much larger had CVD-related mortality rates had changed more favourably) and led to a much larger gap in the District of Columbia in 2013 vs.
+1969, where X of Y years of the increase in the gap is attributable to CVD-related mortality.
+
+**Figure 1: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
 ```{r figure1, fig.align = "center", fig.height=8, fig.width=8, echo=F}
@@
 ```
+**This figure depicts how the gap in life expectancy for Males changed between 1969 (dashed vertical line) and 2013 (solid vertical line).
+Causes to the left of the dashed line narrowed the gap over time, whereas causes to the right exacerbated it.**
 
-**Figure 2: Decomposition of the change in the black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
-```{r figure2, fig.align = "center", fig.height=8, fig.width=9, echo=F}
+**Figure 2: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
+```{r figure2, fig.align = "center", fig.height=8, fig.width=8, echo=F}
+x.labels.f <- levels(factor(cod.df.f$State.gdiff.order))
+x.num.breaks.f <- length(levels(factor(cod.df.f$State.gdiff.order)))
+
+ggplotly(
+  ggplot(cod.df.f, aes(y = State.gdiff.num, x = new.start)) +
+    geom_rect(aes(xmin = new.start, ymin = State.gdiff.num - 0.45, 
+                  ymax = State.gdiff.num + 0.45, xmax = new.finish, fill = COD), color = "white") +
+    geom_segment(data = bw_dat.f, aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) +
+    geom_segment(data = bw_dat.f, aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks.f, labels = x.labels.f)) %>% 
+  layout(xaxis = list(title = "Life expectancy gap (years)"), yaxis = list(title = NA, autorange = "reversed"))
+
+```
+
+**Figure 3: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
+```{r figure3, fig.align = "center", fig.height=8, fig.width=9, echo=F}
 ggplotly(                                     
@@
     theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks, labels = x.labels) + scale_fill_viridis(discrete = T, direction = -1)) %>% 
+  layout(xaxis = list(title = "Life expectancy gap (years)"), yaxis = list(title = NA, autorange = "reversed")
+  )
+```
+
+**Figure 4: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
+```{r figure4, fig.align = "center", fig.height=8, fig.width=9, echo=F}
+ggplotly(                                     
+  ggplot(age.df.f, aes(y = State.gdiff.num, x = new.start)) +
+    geom_rect(aes(xmin = new.start, ymin = State.gdiff.num - 0.45, ymax = State.gdiff.num + 0.45, xmax = new.finish, fill = Ages), color = "white") + 
+    geom_segment(data = bw_dat.f, aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) + 
+    geom_segment(data = bw_dat.f, aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) + 
+    theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks.f, labels = x.labels.f) + scale_fill_viridis(discrete = T, direction = -1)) %>% 
   layout(xaxis = list(title = "Life expectancy gap (years)"), yaxis = list(title = NA, autorange = "reversed")

### 2016-12-16 Code/append_smoothed_results.Rmd
LE for overall US

@@
 
+```{r calculate_overall_LE}
+dat.smoothed.aggregatedCOD$YearRaceSex <- interaction(dat.smoothed.aggregatedCOD$Year3, dat.smoothed.aggregatedCOD$Race2, 
+                                                      dat.smoothed.aggregatedCOD$Sex2)
+
+dat.smoothed.US <- dat.smoothed.aggregatedCOD %>% 
+  ungroup() %>% group_by(Year2, RaceSex, Age2) %>%
+  arrange(YearRaceSex, Age2) %>%
+  summarise(US_smoothed_deaths = sum(total_smoothed_deaths),
+            US_population = sum(Population), 
+            nx = first(nx),
+            Age3 = first(Age3),
+            Year3 = first(Year3),
+            YearRaceSex = first(YearRaceSex),
+            Sex2 = first(Sex2),
+            Race2 = first(Race2))
+
+US.stratum.id <- unique(dat.smoothed.US$YearRaceSex)
+
+US.life.tables <- vector(mode = "list", length = length(US.stratum.id))
+list.index <- 1
+
+#runs in a few seconds
+for (stratum.i in US.stratum.id) {
+
+  US.life.tables[[list.index]] <- life.table(data = subset(dat.smoothed.US, YearRaceSex == stratum.i), 
+                                    num.ages.in.group = "nx", 
+                                    death.counts = "US_smoothed_deaths", 
+                                    population.counts = "US_population")
+  list.index <- list.index + 1
+}
+```
+
 Add the life expectancy at birth to the dat2 data file:

### 2016-12-20 Code/append_smoothed_results.Rmd
Added census region to BW dataset

@@
 rm(BlackWhite3, BlackSmooth, WhiteSmooth)
+
+BlackWhite$Census_Region <- "South"
+BlackWhite$Census_Region[BlackWhite$State2 %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
+                                                 "Nebraska", "North Dakota", "South Dakota", 
+                                                 "Illinois", "Indiana", "Michigan", "Ohio", 
+                                                 "Wisconsin")] <- "Midwest"
+
+BlackWhite$Census_Region[BlackWhite$State2 %in% c("Connecticut", "Maine", "Massachusetts", 
+                                                 "New Hampshire", "Rhode Island", "Vermont",
+                                                 "New Jersey", "New York", "Pennsylvania")] <- "Northeast"
+
+BlackWhite$Census_Region[BlackWhite$State2 %in% c("Arizona", "Colorado", "Idaho", "Montana", 
+                                                 "Nevada", "New Mexico", "Utah", "Wyoming",
+                                                 "Alaska", "California", "Hawaii", "Oregon",
+                                                 "Washington")] <- "West"
 ```

### 2016-12-20 Manuscript/Manuscript.Rmd
Added census region to BW dataset

@@
 
+**Figure 5: Trends in the black-white life expectancy gap for males between 1969 and 2013 by US state**
+```{r figure5, fig.align="center", fig.height=8, fig.width=8, echo=F}
+ggplotly(ggplot(subset(BlackWhiteSmooth, Sex2 == "Male"), aes(y=WBgap.s, x = Year3)) + geom_line(aes(col = State2)))
+```
+
+**Figure 6: Trends in the black-white life expectancy gap for females between 1969 and 2013 by US state**
+```{r figure5, fig.align="center", fig.height=8, fig.width=8, echo=F}
+ggplotly(ggplot(subset(BlackWhiteSmooth, Sex2 == "Female"), aes(y=WBgap.s, x = Year3)) + geom_line(aes(col = State2)))
+```
+
 ###Discussion

### 2016-12-20 Manuscript/Manuscript.Rmd

--- 2016-12-15 Manuscript/Manuscript.Rmd
@@
 
+**Figure 5: Trends in the black-white life expectancy gap for males between 1969 and 2013 by US state**
+```{r figure5, fig.align="center", fig.height=8, fig.width=8, echo=F}
+ggplotly(ggplot(subset(BlackWhite, Sex2 == "Male"), aes(y=WBgap.s, x = Year3)) + geom_line(aes(col = State2))
+         + facet_wrap(~Census_Region) +
+           ylab("Black-white life expectancy gap (years)") +
+           xlab("Year") #+
+           #geom_text(data = subset(BlackWhite, Year3 == 2010), aes(label = State2), check_overlap = T)
+         )
+```
+
+**Figure 6: Trends in the black-white life expectancy gap for females between 1969 and 2013 by US state**
+```{r figure5, fig.align="center", fig.height=8, fig.width=8, echo=F}
+ggplotly(ggplot(subset(BlackWhite, Sex2 == "Female"), aes(y=WBgap.s, x = Year3)) + geom_line(aes(col = State2))
+         + facet_wrap(~Census_Region))
+```
+
 ###Discussion

### 2016-12-20 Manuscript/Manuscript.Rmd

--- 2016-12-15 Manuscript/Manuscript.Rmd
@@
 
+**Figure 5: Trends in the black-white life expectancy gap for males between 1969 and 2013 by US state**
+
+```{r figure5, fig.align="center", fig.height=8, fig.width=8, echo=F}
+ggplotly(ggplot(subset(BlackWhite, Sex2 == "Male"), aes(y=WBgap.s, x = Year3)) + geom_line(aes(col = State2))
+         + facet_wrap(~Census_Region) +
+           ylab("Black-white life expectancy gap (years)") +
+           xlab("Year") #+
+           #geom_text(data = subset(BlackWhite, Year3 == 2010), aes(label = State2), check_overlap = T)
+         )
+```
+
+**Figure 6: Trends in the black-white life expectancy gap for females between 1969 and 2013 by US state**
+
+```{r figure6, fig.align="center", fig.height=8, fig.width=8, echo=F}
+ggplotly(ggplot(subset(BlackWhite, Sex2 == "Female"), aes(y=WBgap.s, x = Year3)) + geom_line(aes(col = State2))
+         + facet_wrap(~Census_Region))
+```
+
 ###Discussion

### 2016-12-20 Manuscript/Manuscript.Rmd
added notes from meeting with SH

@@
 This study seeks to better understand these state-level differences, by examining how changes in age at death and cause of death over time have worked to narrow or exacerbate the life expectancy gap.
+
+**Background notes from meeting with SH on Dec 20**
+-CVD declining starting 1968 - see paper Ruel Stallones - cited in SH JAMA paper - history of decline in CVD.
+-Richard cooper also written about BW diff in CVD with epi perspective
+-for men there is stagnation from 1969 to late 1980s and then increase in gap with HIV and homicides
+-massive crime wave in 1980s, black LE went down from 1984 to 1987 - period when both HIV and homicide went up -- the whole story during that time was HIV and homicide
+-fundamentally interested in the qu is there a common story across most states or are there important differences in what happened in NY vs CA vs TX - can we say anything about what we know about state-specific policies, for example CA has strong anti-smoking legislation.
+Does this help us understand why the gap in Wisconsin was increasing?
+-is there a lot of heterogeneity or is there not?
+- same something about the proportionate contribution by COD and by age - e.g., changes in injuries accounting for X% for closing the gap in NY but in AL it was Y%.
+-heterogeneity between states in the contribution of the causes to the decline in the gap - maybe need to break up into at least two time periods with split at 1990.
+-this is where the new figures come in which on the y-axis have the contribution to LE gap (in years or %) vs year on the x axis, and geom_line for each COD (do one COD at a time), and individual line for each state (may facet by census region).
+-then have a view of how the different COD have contributed to the decline in the gap over time.
 
@@
 
+**Figures 7 and 8 to be added: Trends in the contribution of CVD to the black-white life expectancy gap (in years) by US state**
+-make one for males and one for females
+-separately for each COD group
+-included CIs
+
 ###Discussion

### 2017-01-04 Manuscript/Manuscript.Rmd
add state abbrevs

@@
            ylab("Black-white life expectancy gap (years)") +
-           xlab("Year") #+
-           #geom_text(data = subset(BlackWhite, Year3 == 2010), aes(label = State2), check_overlap = T)
+           xlab("Year") +
+           geom_text(data = subset(BlackWhite, Year3 == 2013 & Sex2 == "Male"), aes(label = stabbrs), check_overlap = T) #+
+           #scale_x_continuous(limits = c(1969, 2020))
          ) 
@@
 ggplotly(ggplot(subset(BlackWhite, Sex2 == "Female"), aes(y=WBgap.s, x = Year3)) + geom_line(aes(col = State2))
-         + facet_wrap(~Census_Region))
+         + facet_wrap(~Census_Region) +
+           ylab("Black-white life expectancy gap (years)") +
+           xlab("Year") +
+           geom_text(data = subset(BlackWhite, Year3 == 2013 & Sex2 == "Female"), aes(label = stabbrs), check_overlap = T))
 ```

### 2017-01-04 Code/1_Explore_Data_v2.Rmd

@@
 dat.clean <- dat[dat$Age2 != "Unknown", ]
-```
-
-```{r addvar_State_abbreviation}
-dat.clean <- dat.clean %>% mutate(stabbrs = factor(State, levels = c(0:50), 
-                                      labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
-                                                 "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", 
-                                                 "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", 
-                                                 "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", 
-                                                 "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
-                                                 "VT", "VA", "WA", "WV", "WI", "WY")))
-
 ```

### 2017-01-04 Code/append_smoothed_results.Rmd

@@
 rm(BlackWhite3, BlackSmooth, WhiteSmooth)
+```
 
+```{r add_Census_Region}
 BlackWhite$Census_Region <- "South"
@@
                                                  "Washington")] <- "West"
+```
+
+```{r addvar_State_abbreviation}
+BlackWhite <- BlackWhite %>% mutate(stabbrs = factor(State2,  
+                                      labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
+                                                 "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", 
+                                                 "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", 
+                                                 "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", 
+                                                 "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
+                                                 "VT", "VA", "WA", "WV", "WI", "WY")))
+
 ```

### 2017-01-05 Code/append_smoothed_results.Rmd

@@
 list.cod.marginal.tables.smoothed <- vector(mode = "list", length = dim(paired.ids)[1])
+State2 <- NULL
+Year3 <- NULL
+Sex2 <- NULL
 
@@
        list.cod.decomp.tables.smoothed[[i]][1] != "Encountered age group with no deaths") { 
-       list.cod.marginal.tables.smoothed[[i]] <-  list.cod.decomp.tables.smoothed[[i]] %>%
-                                                  dplyr::group_by(Cause.of.death) %>% 
-                                                  dplyr::summarise(C_x_COD = sum(C_xi)) %>%
-                                                  dplyr::mutate(C_x_COD_proportion = C_x_COD/sum(C_x_COD),
-                                                         adds_to_gap = ifelse(C_x_COD > 0, T, F))
+       # list.cod.marginal.tables.smoothed[[i]] <-  list.cod.decomp.tables.smoothed[[i]] %>%
+       #                                            dplyr::group_by(Cause.of.death) %>%
+       #                                            dplyr::summarise(C_x_COD = sum(C_xi)) %>%
+       #                                            dplyr::mutate(C_x_COD_proportion = C_x_COD/sum(C_x_COD),
+       #                                                   adds_to_gap = ifelse(C_x_COD > 0, T, F))
 
+       Sex2 <- c(Sex2, rep(as.character(paired.ids$Sex2[i]), 6))
+       State2 <- c(State2, rep(as.character(paired.ids$State2[i]), 6))
+       Year3 <- c(Year3, rep(paired.ids$Year3[i], 6))
     }
 }
+```
+
+```{r create_dataset_COD_contributions}
+cod.contributions <- do.call(rbind, list.cod.marginal.tables.smoothed)
+cod.contributions$Year3 <- Year3
+cod.contributions$State2 <- State2
+cod.contributions$Sex2 <- Sex2
+rm(Year3, State2, Sex2)
+
+cod.contributions <- cod.contributions %>% mutate(stabbrs = factor(State2,  
+                                      labels = c("AL", "AZ", "AR", "CA", "CO", "CT", "DE", 
+                                                 "FL", "GA", "IL", "IN", "IA", "KS", "KY", 
+                                                 "LA", "MD", "MA", "MI", "MN", "MS", "MO", 
+                                                 "NE", "NV", "NJ", "NM", "NY", "NC", "OH", 
+                                                 "OK", "PA", "RI", "SC", "TN", "TX", "UT", 
+                                                 "VA", "WA", "DC", "WV", "WI")))
+table(cod.contributions$State2, cod.contributions$stabbrs)
+
+cod.contributions$Census_Region <- "South"
+cod.contributions$Census_Region[cod.contributions$State2 %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
+                                                 "Nebraska", "North Dakota", "South Dakota", 
+                                                 "Illinois", "Indiana", "Michigan", "Ohio", 
+                                                 "Wisconsin")] <- "Midwest"
+
+cod.contributions$Census_Region[cod.contributions$State2 %in% c("Connecticut", "Maine", "Massachusetts", 
+                                                 "New Hampshire", "Rhode Island", "Vermont",
+                                                 "New Jersey", "New York", "Pennsylvania")] <- "Northeast"
+
+cod.contributions$Census_Region[cod.contributions$State2 %in% c("Arizona", "Colorado", "Idaho", "Montana", 
+                                                 "Nevada", "New Mexico", "Utah", "Wyoming",
+                                                 "Alaska", "California", "Hawaii", "Oregon",
+                                                 "Washington")] <- "West"
+
 ```

### 2017-01-05 Manuscript/Manuscript.Rmd

--- 2016-12-20 Manuscript/Manuscript.Rmd
@@
 
-```{r figure5, fig.align="center", fig.height=8, fig.width=8, echo=F}
+```{r figure5, fig.align="center", fig.height=8, fig.width=9, echo=F}
 ggplotly(ggplot(subset(BlackWhite, Sex2 == "Male"), aes(y=WBgap.s, x = Year3)) + geom_line(aes(col = State2))
-         + facet_wrap(~Census_Region) +
+         + facet_wrap( ~ Census_Region) +
            ylab("Black-white life expectancy gap (years)") +
-           xlab("Year") #+
-           #geom_text(data = subset(BlackWhite, Year3 == 2010), aes(label = State2), check_overlap = T)
+           xlab("Year") +
+           geom_text(data = subset(BlackWhite, Year3 == 2013 & Sex2 == "Male"), aes(label = stabbrs), check_overlap = T, size = 2.5) 
          ) 
@@
 
-```{r figure6, fig.align="center", fig.height=8, fig.width=8, echo=F}
+```{r figure6, fig.align="center", fig.height=8, fig.width=9, echo=F}
 ggplotly(ggplot(subset(BlackWhite, Sex2 == "Female"), aes(y=WBgap.s, x = Year3)) + geom_line(aes(col = State2))
-         + facet_wrap(~Census_Region))
+         + facet_wrap(~Census_Region) +
+           ylab("Black-white life expectancy gap (years)") +
+           xlab("Year") +
+           geom_text(data = subset(BlackWhite, Year3 == 2013 & Sex2 == "Female"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
 ```
 
-**Figures 7 and 8 to be added: Trends in the contribution of CVD to the black-white life expectancy gap (in years) by US state**
--make one for males and one for females
--separately for each COD group
--included CIs
+**Figure 7a: Contribution of cancer to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
+
+```{figure7a, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cancers"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Cancers"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+
+#something is wrong with the percent contribution calculated for RI around 2008 ... hmmph
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cancers"), aes(x = Year3, y = C_x_COD_proportion)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Cancers"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 7b: Contribution of cardiovascular disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
+
+```{figure7b, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cardiovascular"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Cardiovascular"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+
+#need to figure out what is weird about this one in NM in 2002 -- something is very odd. 
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cardiovascular"), aes(x = Year3, y = C_x_COD_proportion)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Cardiovascular"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 7c: Contribution of communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
+
+```{figure7c, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Communicable"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Communicable"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 7d: Contribution of non-communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
+
+```{figure7d, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Non-communicable"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Non-communicable"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 7e: Contribution of injuries to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
+
+```{figure7e, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Injuries"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Injuries"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 7f: Contribution of all other causes to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
+
+```{figure7f, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "All other causes"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "All other causes"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 8a: Contribution of cancer to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
+
+```{figure8a, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Cancers"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Female" & Cause.of.death == "Cancers"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 8b: Contribution of cardiovascular disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
+
+```{figure8b, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Cardiovascular"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Female" & Cause.of.death == "Cardiovascular"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 8c: Contribution of communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
+
+```{figure8c, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Communicable"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Female" & Cause.of.death == "Communicable"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 8d: Contribution of non-communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
+
+```{figure8d, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Non-communicable"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Female" & Cause.of.death == "Non-Communicable"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+-add CIs when I have them
 

### 2017-01-05 Code/append_smoothed_results.Rmd
added heatmaps

@@
        list.cod.decomp.tables.smoothed[[i]][1] != "Encountered age group with no deaths") { 
-       # list.cod.marginal.tables.smoothed[[i]] <-  list.cod.decomp.tables.smoothed[[i]] %>%
-       #                                            dplyr::group_by(Cause.of.death) %>%
-       #                                            dplyr::summarise(C_x_COD = sum(C_xi)) %>%
-       #                                            dplyr::mutate(C_x_COD_proportion = C_x_COD/sum(C_x_COD),
-       #                                                   adds_to_gap = ifelse(C_x_COD > 0, T, F))
+       list.cod.marginal.tables.smoothed[[i]] <-  list.cod.decomp.tables.smoothed[[i]] %>%
+                                                  dplyr::group_by(Cause.of.death) %>%
+                                                  dplyr::summarise(C_x_COD = sum(C_xi)) %>%
+                                                  dplyr::mutate(C_x_COD_proportion = C_x_COD/sum(C_x_COD),
+                                                         adds_to_gap = ifelse(C_x_COD > 0, T, F))
 
@@
      list.age.decomp.tables.smoothed, list.cod.decomp.tables.smoothed, list.cod.marginal.tables.smoothed,
-     non.zero.stratums, non.zero.stratums.smoothed,
+     non.zero.stratums, non.zero.stratums.smoothed, cod.contributions,
      file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")

### 2017-01-05 Manuscript/Manuscript.Rmd
added heatmaps

--- 2016-12-20 Manuscript/Manuscript.Rmd
@@
 
-```{r figure6, fig.align="center", fig.height=8, fig.width=8, echo=F}
+```{r figure6, fig.align="center", fig.height=8, fig.width=9, echo=F}
 ggplotly(ggplot(subset(BlackWhite, Sex2 == "Female"), aes(y=WBgap.s, x = Year3)) + geom_line(aes(col = State2))
-         + facet_wrap(~Census_Region))
+         + facet_wrap(~Census_Region) +
+           ylab("Black-white life expectancy gap (years)") +
+           xlab("Year") +
+           geom_text(data = subset(BlackWhite, Year3 == 2013 & Sex2 == "Female"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
 ```
 
-**Figures 7 and 8 to be added: Trends in the contribution of CVD to the black-white life expectancy gap (in years) by US state**
--make one for males and one for females
--separately for each COD group
--included CIs
+**Figure 7a: Contribution of cancer to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
+```{r figure7a, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cancers"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Cancers"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+
+    #theme_grey(base_size = 9) + labs(x = "", y = "") + 
+    #scale_x_discrete(expand = c(0, 0)) +
+    #scale_y_discrete(expand = c(0, 0)) +
+    #theme(legend.position = "none", axis.ticks = element_blank(), axis.text.x = element_text(size = 9*0.8, angle = 330, hjust = 0, colour = "grey50"))
+
+#something is wrong with the percent contribution calculated for RI around 2008 ... hmmph
+# ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cancers"), aes(x = Year3, y = C_x_COD_proportion)) + 
+#          geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+#          ylab("Contribution to LE Gap (years)") +
+#          xlab("Year") +
+#          geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Cancers"), 
+#          aes(label = stabbrs), check_overlap = T, size = 2.5))
+```
+
+**Figure 7b: Contribution of cardiovascular disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
+```{r figure7b, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cardiovascular"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Cardiovascular"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+
+#need to figure out what is weird about this one in NM in 2002 -- something is very odd. 
+# ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cardiovascular"), aes(x = Year3, y = C_x_COD_proportion)) + 
+#          geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+#          ylab("Contribution to LE Gap (years)") +
+#          xlab("Year") +
+#          geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Cardiovascular"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+# )
+```
+
+**Figure 7c: Contribution of communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
+
+```{r figure7c, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Communicable"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Communicable"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 7d: Contribution of non-communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
+
+```{r figure7d, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Non-communicable"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Non-communicable"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 7e: Contribution of injuries to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
+
+```{r figure7e, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Injuries"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Injuries"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 7f: Contribution of all other causes to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
+
+```{r figure7f, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "All other causes"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "All other causes"), aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 7g: Different way to portray the above data for males**
+
+```{r figure7g, fig.align="center", fig.height=8, fig.width=9, echo=F}
+for (cod in levels(cod.contributions$Cause.of.death)) {
+  print(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == cod), aes(x = Year3, y = State2)) + 
+          geom_tile(aes(fill = C_x_COD), col = "white") +
+          facet_wrap(~ Census_Region, scales = "free_y") + 
+          scale_fill_viridis() + theme_minimal() + 
+          ylab("Contribution to LE Gap (years)") +
+          xlab("Year") + ggtitle(cod))
+}
+```
+
+**Figure 8a: Contribution of cancer to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
+
+```{r figure8a, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Cancers"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Female" & Cause.of.death == "Cancers"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 8b: Contribution of cardiovascular disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
+
+```{r figure8b, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Cardiovascular"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Female" & Cause.of.death == "Cardiovascular"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 8c: Contribution of communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
+
+```{r figure8c, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Communicable"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Female" & Cause.of.death == "Communicable"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 8d: Contribution of non-communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
+
+```{r figure8d, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Non-communicable"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Female" & Cause.of.death == "Non-Communicable"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 8e: Contribution of injuries to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
+
+```{r figure8e, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Injuries"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Female" & Cause.of.death == "Injuries"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 8f: Contribution of all other causes to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
+
+```{r figure8f, fig.align="center", fig.height=8, fig.width=9, echo=F}
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "All other causes"), aes(x = Year3, y = C_x_COD)) + 
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Female" & Cause.of.death == "All other causes"), 
+         aes(label = stabbrs), check_overlap = T, size = 2.5)
+)
+```
+
+**Figure 8g: Different way to portray the above data for females**
+
+```{r figure8g, fig.align="center", fig.height=8, fig.width=9, echo=F}
+for (cod in levels(cod.contributions$Cause.of.death)) {
+  print(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == cod), aes(x = Year3, y = State2)) + 
+          geom_tile(aes(fill = C_x_COD), col = "white") +
+          facet_wrap(~ Census_Region, scales = "free_y") + 
+          scale_fill_viridis() + theme_minimal() + 
+          ylab("Contribution to LE Gap (years)") +
+          xlab("Year") + ggtitle(cod))
+}
+```
+
+-add CIs when I have them
 

### 2017-01-05 Code/append_smoothed_results.Rmd
blach

@@
                                                  "Washington")] <- "West"
+
+BlackWhite$Census_Region[BlackWhite$State2 %in% c("Delaware", "Florida", "Georgia", "Maryland", 
+                                                  "North Carolina", "South Carolina", 
+                                                  "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"
+
+BlackWhite$Census_Region[BlackWhite$State2 %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee", 
+                                                  "Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "West or East South Central"
+
 ```
@@
 
+cod.contributions$Census_Region[cod.contributions$State2 %in% c("Delaware", "Florida", "Georgia", "Maryland", 
+                                                                "North Carolina", "South Carolina", 
+                                                                "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"
+
+cod.contributions$Census_Region[cod.contributions$State2 %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee", 
+                                                                "Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "West or East South Central"
+
 ```

### 2017-01-05 Manuscript/Manuscript.Rmd
blach

@@
 
-#something is wrong with the percent contribution calculated for RI around 2008 ... hmmph
-# ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cancers"), aes(x = Year3, y = C_x_COD_proportion)) + 
-#          geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
-#          ylab("Contribution to LE Gap (years)") +
-#          xlab("Year") +
-#          geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Cancers"), 
-#          aes(label = stabbrs), check_overlap = T, size = 2.5))
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cancers"), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (years)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Cancers"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
+#the wonky percent contributions occur in the states where the LE gap crosses 0 years.
+
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cancers" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > -0.5), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (%)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2000 & Sex2 == "Male" & Cause.of.death == "Cancers"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
 ```
@@
 
-#need to figure out what is weird about this one in NM in 2002 -- something is very odd. 
-# ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cardiovascular"), aes(x = Year3, y = C_x_COD_proportion)) + 
-#          geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
-#          ylab("Contribution to LE Gap (years)") +
-#          xlab("Year") +
-#          geom_text(data = subset(cod.contributions, Year3 == 2013 & Sex2 == "Male" & Cause.of.death == "Cardiovascular"), aes(label = stabbrs), check_overlap = T, size = 2.5)
-# )
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Cardiovascular" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > -0.5), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (%)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2000 & Sex2 == "Male" & Cause.of.death == "Cardiovascular"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
 ```
@@
 )
+
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Communicable" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > -0.5), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (%)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2000 & Sex2 == "Male" & Cause.of.death == "Communicable"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
 ```

### 2017-01-06 Manuscript/Manuscript.Rmd
added final contribution plots

@@
 
-ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Communicable" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > -0.5), aes(x = Year3, y = C_x_COD_proportion)) +
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Communicable" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > 0), aes(x = Year3, y = C_x_COD_proportion)) +
          geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
@@
 )
+
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Non-communicable" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > 0), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (%)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2000 & Sex2 == "Male" & Cause.of.death == "Non-communicable"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
 ```
@@
 )
+
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "Injuries" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > 0), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (%)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2000 & Sex2 == "Male" & Cause.of.death == "Injuries"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
 ```
@@
 )
+
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Male" & Cause.of.death == "All other causes" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > 0), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (%)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2000 & Sex2 == "Male" & Cause.of.death == "All other causes"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
 ```
@@
 )
+
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Cancers" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > 0), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (%)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2000 & Sex2 == "Male" & Cause.of.death == "Cancers"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
 ```
@@
 )
+
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Cardiovascular" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > 0), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (%)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2000 & Sex2 == "Male" & Cause.of.death == "Cardiovascular"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
 ```
@@
 )
+
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Communicable" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > 0), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (%)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2000 & Sex2 == "Male" & Cause.of.death == "Communicable"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
 ```
@@
 )
+
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Non-communicable" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > 0), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (%)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2000 & Sex2 == "Male" & Cause.of.death == "Non-communicable"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
 ```
@@
 )
+
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "Injuries" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > -0.2), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (%)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2000 & Sex2 == "Male" & Cause.of.death == "Injuries"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
 ```
@@
 )
+
+ggplotly(ggplot(subset(cod.contributions, Sex2 == "Female" & Cause.of.death == "All other causes" & C_x_COD_proportion < 0.5 & C_x_COD_proportion > -0.2), aes(x = Year3, y = C_x_COD_proportion)) +
+         geom_line(aes(col = State2)) + facet_wrap(~ Census_Region) +
+         ylab("Contribution to LE Gap (%)") +
+         xlab("Year") +
+         geom_text(data = subset(cod.contributions, Year3 == 2000 & Sex2 == "Male" & Cause.of.death == "All other causes"),
+         aes(label = stabbrs), check_overlap = T, size = 2.5))
 ```

### 2017-01-09 Manuscript/Appendix.Rmd

---
title: "Appendix material"
author: "Kathryn Morrison"
date: "December 19, 2016"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Statistical model

There is a single estimated mortality rate for each sub-category of state, year, race, age, sex, and cause of death. We consider the true underlying mortality rate to be *latent* and we estimate it with the observed number of deaths per sub-category relative to the number of at-risk individuals, as described previously. The true risk of mortality cannot be zero and is unlikely to exceed 0.1, but these extreme values may occur when the at-risk population is small. To address the challenges in precisely measuring mortality rate with discrete numbers, we perform temporal smoothing using a Bayesian autoregressive approach.

We denote the true underlying mortality rate as $\lambda$, the observed number of deaths as $D$ and the at-risk population as $P$. For each given state, race, age, sex, and cause of death subgroup, we model the deaths per year as a Poisson random variable, with mean $\mu_t$ (the expected number of deaths per year) that is equal to the mortality rate multiplied by the population: $\lambda_t \times P_t =\frac{D_t}{P_t} \times P_t = D_t$.

$D_t \sim Poisson(\mu_t)$

$\mu_t = exp[log(\lambda_t) + log(P_t)]$

We log transform the expected value because the mortality rate tends to be heavily right skewed but approximately log-normal. This transformation allows us to model the log rate as autoregressive to smooth between years:

$log(\lambda_t) \sim N(\ log(\lambda_{t-1}),\ \tau)$

For the first year, a weakly informative prior is placed on the mortality rate:

$log(\lambda_t) \sim N(-4, \ 0.1)$

And the prior in the variance is also weakly informative:

$\tau \sim gamma(0.01, 0.01)$

This autoregressive approach to smoothing addresses the issues with small population sizes causing unrealistic mortality rates in some sub-categories. However, there was also the additional challenge of corsened data, where cells with counts between one and nine were binned into a single category. To estimate the mortality rate in these cells, we use a truncated Poisson bounded between 1 and an upper bound, where the bound is the smaller of 9 or the population size in that sub category.

### 2017-01-09 Manuscript/Manuscript.Rmd

@@
 
-The data produced two statistical issues that were addressed by the statistical methods we chose.
+The data produced two statistical issues that were addressed by our statistical methods.
 First, our methodology needed to impute the missing death counts.
-Second, several states had small blacks populations which produced even smaller population sizes in each sex- and age-specific grouping.
-In these cases, the estimation of the age-specific mortality rates is imprecise, and produces abrupt changes in the estimation of mortality rates over time that are unrealistic.
+Second, several states had small blacks populations which produced even smaller population sizes in each sex-age category.
+In these cases, the estimation of the sex-age mortality rates is imprecise, and produces abrupt changes in the estimation of mortality rates over time that are unrealistic.
 To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.
 
-KTM to add bayesian modeling paragraph
+To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts.
+Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population.
+The mortality rates over time within each state-age-sex-race category are modelled autoregressively, where the log mortality rate in the current year was modelled as a normal variable centred at the log mortality rate estimated from the previous year.
+Truncated Poisson regression was used to impute the supressed death counts, by specifying that the supressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9).
+These models were run separately for every combination of state, race, and sex, and produced death counts that had been smoothed over time for every cause of death within each age group.
 
-- type of bayesian model used and what was modeled Y vs X, separately for each state, race, and sex
-- how the smoothing happens over year
-- how the truncated poisson is involved
-- this methodology produces smoothed death counts 
-- sampled from posterior distribution with thinning
-- software: R software packages and their versions
-
-Thus, the model was run separately for every combination of state, race, and sex, and produced death counts that had been smoothed over time for every cause of death within each age group.
 To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref].
@@
 
-We then performed a decomposition analysis using the Arriaga method, to decompose the black-white difference in life expectancy according to age group and cause of death [ref].
-The decomposition analysis partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher mortality rates for each age group.
+To decompose the black-white difference in life expectancy, we performed a decomposition analysis using the Arriaga method [ref].
+First, the decomposition analysis partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks.
 For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites.
-The age-specific decomposition is then further partitioned according to cause of death, producing the number of years of the gap that is attributable to each age-cause-of-death combination.
+This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination.
 For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old.
-We also calculated the marginal contribution of each cause of death by summing the cause-specific contributions across age groups.
+We also summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference.
 
+**Will need to change the following**
 Finally, we calculated the change in the life expectancy gap over time within each state-, race-, and sex-specific stratum.
@@
 The bayesian smoothing was performed using `JAGS` (version X.X), and the life expectnacy and decomposition analysis were performed using user-written functions found in the Supplement.
+
+**Add in the stuff about precision**
+- sampled from posterior distribution with thinning
+- software: R software packages and their versions
 

### 2017-01-18 Code/1_Explore_Data.Rmd
cleaning up repo

@@
 library(tidyr)
+library(plotly)
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
 ```
@@
          Year2 = factor(Year, levels = c(0:44), labels = as.character(1969:2013)),
-         RaceSex = interaction(Race2, Sex2),
+         Year3 = as.numeric(as.character(Year2)),
+         RaceSex = interaction(Race2, Sex2, sep = " "),
+         StateYearSex = interaction(State2, Year2, Sex2),
          StateYearRaceSex = interaction(State2, Year2, Race2, Sex2))
@@
 
-#now the Counts should only equal NA when the true count is between 1 and 9.
-
 summary(dat.clean$Crude_Rate[dat.clean$Population == 0])
-table(dat.clean$Crude_Rate[dat.clean$Population==0], useNA = "always")
 #whenever the population is 0, the Crude_Rate is equal to NA.
 #need to think: should these be equal to 0, or NA? How will it affect the calculations?
+
+#States with at least one strata of 0
+na.pops <- dat.clean %>% 
+  select(State2, Population) %>% 
+  filter(Population == 0)
+
+unique(na.pops$State2)
+
+pops.le.20 <- dat.clean %>% select(State2, Population) %>% filter(Population <= 20)
+unique(pops.le.20$State2)
 ```
@@
 
-- Set all missing counts to 1, 5, and 9.
-- Set all missing counts to 1 for Whites and 9 for Blacks and vice versa.
+- Set all missing counts to min({1, 5, and 9}, population_size).
 
 ```{r fill_in_missing_counts}
-dat.clean <- dat.clean %>% mutate(Count_md1 = ifelse(is.na(Count)==T, 1, Count),
-                                  Count_md5 = ifelse(is.na(Count)==T, 5, Count),
-                                  Count_md9 = ifelse(is.na(Count)==T, 9, Count)
+dat.clean <- dat.clean %>% mutate(Count_md1 = ifelse(is.na(Count)==T, pmin(1, Population), Count),
+                                  Count_md5 = ifelse(is.na(Count)==T, pmin(5, Population), Count),
+                                  Count_md9 = ifelse(is.na(Count)==T, pmin(9, Population), Count)
                                   )
-
-dat.clean$Count_b9w1 <- dat.clean$Count
-dat.clean$Count_b9w1[is.na(dat.clean$Count) & dat.clean$Race2 == "White"] <- 1
-dat.clean$Count_b9w1[is.na(dat.clean$Count) & dat.clean$Race2 == "Black"] <- 9
-
-dat.clean$Count_b1w9 <- dat.clean$Count
-dat.clean$Count_b1w9[is.na(dat.clean$Count) & dat.clean$Race2 == "White"] <- 9
-dat.clean$Count_b1w9[is.na(dat.clean$Count) & dat.clean$Race2 == "Black"] <- 1
 
@@
 
-table(dat.clean$Count_grouped, dat.clean$Count_b9w1_grouped, dat.clean$Race2, useNA = "always")
-table(dat.clean$Count_grouped, dat.clean$Count_b1w9_grouped, dat.clean$Race2, useNA = "always")
-
-dat.clean <- dat.clean[ , !names(dat.clean) %in% c("Count_grouped", "Count1_grouped", "Count5_grouped", "Count9_grouped", 
-                                                   "Count_b9w1_grouped", "Count_b1w9_grouped")]
+dat.clean <- dat.clean[ , !names(dat.clean) %in% c("Count_grouped", "Count1_grouped", "Count5_grouped", "Count9_grouped")]
 ```
 
-To calculate life expectancy, we have to aggregate the data across the causes of death.
+```{r addvar_cod_proportions}
+dat.clean <- dat.clean %>% group_by(State2, RaceSex, Year3, Age3) %>%
+                           arrange(State2, RaceSex, Year3, Age3) %>%
+                           mutate(total_deaths1 = sum(Count_md1),
+                                  total_deaths5 = sum(Count_md5),
+                                  total_deaths9 = sum(Count_md9),
+                                  cod_prop_death1 = Count_md1/total_deaths1, 
+                                  cod_prop_death5 = Count_md5/total_deaths5,
+                                  cod_prop_death9 = Count_md9/total_deaths9)
+
+length(dat.clean$total_deaths5[dat.clean$total_deaths5 == 0])
+summary(dat.clean$cod_prop_death5[dat.clean$total_deaths5 == 0])
+#76998 strata with 0 deaths from any cause (12833 when divided by 6 -- the number of causes)
+#7% of strata have no deaths from any cause.
+```
+
+To calculate life expectancy, we have to aggregate the data across causes of death.
 
@@
   arrange(State2, RaceSex, Year2, Age2) %>%
-  summarise(total_deaths1 = sum(Count_md1), total_deaths5 = sum(Count_md5),
-            total_deaths9 = sum(Count_md9), 
-            total_deaths_b9 = sum(Count_b9w1), total_deaths_w9 = sum(Count_b1w9),
-            Population = first(Population), Race2 = first(Race2), Sex2 = first(Sex2), 
-            StateYearRaceSex = first(StateYearRaceSex))
+  summarise(Population = first(Population), 
+            Race2 = first(Race2), 
+            Sex2 = first(Sex2), 
+            StateYearRaceSex = first(StateYearRaceSex),
+            nx = first(nx),
+            Age3 = first(Age3),
+            Year3 = first(Year3),
+            total_deaths1 = first(total_deaths1),
+            total_deaths5 = first(total_deaths5),
+            total_deaths9 = first(total_deaths9),
+            cod_prop_death1 = first(cod_prop_death1),
+            cod_prop_death5 = first(cod_prop_death5),
+            cod_prop_death9 = first(cod_prop_death9))
+
+#return stratum ids with any age-specific population counts of 0.
+#unique(dat2$StateYearRaceSex[dat2$Population == 0])
+non.zero.stratums <- setdiff(levels(dat2$StateYearRaceSex), unique(dat2$StateYearRaceSex[dat2$Population == 0]))
 ```
 
-We implemented the calculations performed in the technical supplement found here [http://www.lho.org.uk/Download/Public/7656/1/tech_supp_3.pdf], also saved in our dropbox.
+Add the total population (across age brackets) to the dat2 dataset.
 
-Let's make sure we can do the calculation all the way through for a few of the groupings, including a grouping with a stratum with a zero population count.
+```{r create_var_pop_across_age}
+dat3 <- dat2 %>% ungroup() %>% 
+  group_by(StateYearRaceSex) %>%
+  summarise(Pop_across_age = sum(Population))
 
-```{r calc_life_expectancy1}
-
-dat2$numYrsInt <- 5
-dat2$numYrsInt[dat2$Age2 == "<1 year"] <- 1
-dat2$numYrsInt[dat2$Age2 == "1-4 years"] <- 4
-
-dat2 <- dat2 %>% mutate( 
-            Death_Rate1 = ifelse(Population!=0, total_deaths1/Population, NA),
-            Death_Rate5 = ifelse(Population!=0, total_deaths5/Population, NA),
-            Death_Rate9 = ifelse(Population!=0, total_deaths9/Population, NA),
-            Death_Rateb9 = ifelse(Population!=0, total_deaths_b9/Population, NA),
-            Death_Ratew9 = ifelse(Population!=0, total_deaths_w9/Population, NA),
-            Ave_Lived_by_Died = ifelse(Age2=="<1 year", 0.1, 0.5),
-            Prob_Dying1 = ifelse(Population!=0, 
-                                 numYrsInt*Death_Rate1/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Rate1), 
-                                 NA),
-            Prob_Dying5 = ifelse(Population!=0, 
-                                 numYrsInt*Death_Rate5/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Rate5), 
-                                 NA),
-            Prob_Dying9 = ifelse(Population!=0, 
-                                 numYrsInt*Death_Rate9/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Rate9), 
-                                 NA),
-            Prob_Dyingb9 = ifelse(Population!=0, 
-                                 numYrsInt*Death_Rateb9/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Rateb9), 
-                                 NA),
-            Prob_Dyingw9 = ifelse(Population!=0, 
-                                 numYrsInt*Death_Ratew9/(1 + numYrsInt*(1 - Ave_Lived_by_Died)*Death_Ratew9), 
-                                 NA),
-            Prob_Surv1 = ifelse(Population!=0, 1 - Prob_Dying1, NA),
-            Prob_Surv5 = ifelse(Population!=0, 1 - Prob_Dying5, NA),
-            Prob_Surv9 = ifelse(Population!=0, 1 - Prob_Dying9, NA),            
-            Prob_Survb9 = ifelse(Population!=0, 1 - Prob_Dyingb9, NA),
-            Prob_Survw9 = ifelse(Population!=0, 1 - Prob_Dyingw9, NA),
-            alive_at_start_int1 = ifelse(Age2 == "<1 year", 1000000, NA),
-            alive_at_start_int5 = ifelse(Age2 == "<1 year", 1000000, NA),
-            alive_at_start_int9 = ifelse(Age2 == "<1 year", 1000000, NA),
-            alive_at_start_intb9 = ifelse(Age2 == "<1 year", 1000000, NA),
-            alive_at_start_intw9 = ifelse(Age2 == "<1 year", 1000000, NA),
-            deaths_during_int1 = ifelse(Age2 == "<1 year", alive_at_start_int1*Prob_Dying1, NA),
-            deaths_during_int5 = ifelse(Age2 == "<1 year", alive_at_start_int5*Prob_Dying5, NA),
-            deaths_during_int9 = ifelse(Age2 == "<1 year", alive_at_start_int9*Prob_Dying9, NA),
-            deaths_during_intb9 = ifelse(Age2 == "<1 year", alive_at_start_intb9*Prob_Dyingb9, NA),
-            deaths_during_intw9 = ifelse(Age2 == "<1 year", alive_at_start_intw9*Prob_Dyingw9, NA))
+dat2 <- merge(dat2, dat3, by = "StateYearRaceSex")
+rm(dat3)
+dat2 <- dat2 %>% arrange(State2, Year3, Race2, Sex2, Age3)
 ```
 
-```{r first_five}
-levels(dat2$StateYearRaceSex)[1:5]
-summary(dat2$Population[dat2$StateYearRaceSex %in% levels(dat2$StateYearRaceSex)[1:5]])
+In those strata that do not have any age-specific population counts of 0, create a life table and store it in a list.
 
-timer1 <- system.time (
-  for(x in levels(dat2$StateYearRaceSex)[1:5]) {
-      for(row in 2:length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) { 
-        dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row - 1]
+```{r calculate_life_expectancy}
+#takes 10 mins for all three
 
-          if (row != length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) {
-              dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying1[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying5[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying9[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingb9[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingw9[dat2$StateYearRaceSex == x][row]
+list.life.tables <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
+#list.life.tables1 <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
+list.index <- 1
+
+system.time(
+for(stratum.id in non.zero.stratums) {
+#for(stratum.id in head(non.zero.stratums)) {
+
+  list.life.tables[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2),   
+  #list.life.tables1[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
+                                               num.ages.in.group = "nx", 
+                                               death.counts = "total_deaths1", 
+                                               population.counts = "Population")
+
+  list.life.tables5[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
+                                               num.ages.in.group = "nx", 
+                                               death.counts = "total_deaths5", 
+                                               population.counts = "Population")
+
+  list.life.tables9[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
+                                               num.ages.in.group = "nx", 
+                                               death.counts = "total_deaths9", 
+                                               population.counts = "Population")
+  list.index <- list.index + 1
+  print(list.index)
   }
-          else { dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] #everyone dies in the 85+ grouping
-                 dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]
-                 dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]
-                 dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]
-                 dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]
+)
+```
 
+Add the life expectancy at birth to the data file:
+
+```{r add_le_to_dataset}
+le <- le5 <- le9 <- rep(NA, length(non.zero.stratums))
+
+for (i in 1:length(non.zero.stratums)) {
+  le[i] <- list.life.tables[[i]]$e_x[1]
+  le5[i] <- list.life.tables5[[i]]$e_x[1]
+  le9[i] <- list.life.tables9[[i]]$e_x[1]
+}
+
+les <- data.frame(life.expectancy.birth = le, 
+                  life.expectancy.birth.5 = le5, 
+                  life.expectancy.birth.9 = le9, 
+                  StateYearRaceSex = non.zero.stratums)
+
+dat2 <- merge(dat2, les, by = "StateYearRaceSex", all.x =T)
+
+rm(les)
+
+ggplot(dat2 %>% filter(Age2 == "<1 year"), aes(x = life.expectancy.birth)) + 
+  geom_density(aes(fill="Impute 1"), alpha = 0.3) + 
+  geom_density(aes(x = life.expectancy.birth.5, fill = "Impute 5"), alpha = 0.3) +
+  geom_density(aes(x = life.expectancy.birth.9, fill="Impute 9"), alpha = 0.3) +
+  facet_grid(Sex2 ~ Race2)
+```
+
+Make a new dataset that contains information on the gap:
+
+```{r create_dataset_BlackWhite}
+#this dataset contains the black-white data within the same row (wide format) for each year-state-sex strata
+
+Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
+                  select(State2, Year3, Sex2, Pop_across_age, 
+                         life.expectancy.birth, life.expectancy.birth.5, 
+                         life.expectancy.birth.9) %>%
+                  rename(Pop_black = Pop_across_age,  
+                         le.birth.black = life.expectancy.birth, 
+                         le.birth.black5 = life.expectancy.birth.5, 
+                         le.birth.black9 = life.expectancy.birth.9)
+
+White <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
+                  select(State2, Year3, Sex2, Pop_across_age, 
+                         life.expectancy.birth, life.expectancy.birth.5, 
+                         life.expectancy.birth.9) %>%
+                  rename(Pop_white = Pop_across_age,  
+                         le.birth.white = life.expectancy.birth,
+                         le.birth.white5 = life.expectancy.birth.5,
+                         le.birth.white9 = life.expectancy.birth.9)
+
+BlackWhite <- merge(Black, White, by = c("State2", "Year3", "Sex2"))
+
+BlackWhite <- BlackWhite %>% mutate(WBgap = le.birth.white - le.birth.black,
+                                    WBgap5 = le.birth.white5 - le.birth.black5,
+                                    WBgap9 = le.birth.white9 - le.birth.black9)
+
+rm(Black, White)
+```
+
+Decompose the gap by age. That is, for each year-state-sex combination, compare the life table for blacks to whites and decompose the gap in life expectancy at birth by age.
+
+```{r create_paired_ids_dataset}
+#paired.ids will contain the index names that are sent to the age decomposition function in the next code chunk
+
+#first, create list of stratum.ids for each race-sex combination
+for (rs in unique(levels(dat.clean$RaceSex))) {
+  assign(paste0(gsub(" ", "", rs), ".stratum.ids"), 
+         dat.clean %>% 
+           ungroup() %>%
+           filter(RaceSex == rs) %>%
+           select(StateYearRaceSex, StateYearSex, Sex2, State2, Year3) %>%
+           group_by(StateYearRaceSex, StateYearSex) %>%
+           filter(row_number() == 1)
+         )
+
+}
+
+#then, merge the black-white lists
+Female.ids <- merge(WhiteFemale.stratum.ids, BlackFemale.stratum.ids[, -c(3:5)], 
+                    by = "StateYearSex", suffixes = c("White", "Black"))
+head(Female.ids)
+
+Male.ids <- merge(WhiteMale.stratum.ids, BlackMale.stratum.ids[, -c(3:5)],
+                  by = "StateYearSex", suffixes = c("White", "Black"))
+head(Male.ids)
+
+paired.ids <- rbind(Male.ids, Female.ids)
+```
+
+```{r calculate_age_decomp}
+#runs in about a minute
+
+list.age.decomp.tables <- list.age.decomp.tables5 <- list.age.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])
+
+for(i in 1:dim(paired.ids)[1]) {
+  white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
+  black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])
+
+  if(length(white.index) != 0 & length(black.index) != 0) {
+    list.age.decomp.tables[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]), 
+                                                 name.lt1 = "Black", 
+                                                 life.table2 = as.data.frame(list.life.tables[[white.index]]), 
+                                                 name.lt2 = "White", age.groups = "Age3")
+
+    list.age.decomp.tables5[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]), 
+                                                 name.lt1 = "Black", 
+                                                 life.table2 = as.data.frame(list.life.tables5[[white.index]]), 
+                                                 name.lt2 = "White", age.groups = "Age3")
+
+    list.age.decomp.tables9[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables9[[black.index]]), 
+                                                 name.lt1 = "Black", 
+                                                 life.table2 = as.data.frame(list.life.tables9[[white.index]]), 
+                                                 name.lt2 = "White", age.groups = "Age3")
+  } else {
+    list.age.decomp.tables[[i]] <- list.age.decomp.tables5[[i]] <- list.age.decomp.tables9[[i]] <-"Couldn't compare life tables"
   }
@@
 
+```
+
+```{r calculate_cod_decomp}
+#stopped here - need to figure this out
+#if the count of deaths == 0 for all causes of death within an age band then the COD proportion is NA
+#this messes up the cod decomposition. What to do?
+#example - no black people dying between the ages of 5-9 of anything. but have whites dying in that age group so their
+#COD proportions are okay.
+
+#takes 1.5 hours for all three
+list.cod.decomp.tables <- list.cod.decomp.tables5 <- list.cod.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])
+
+system.time(
+for(i in 1:dim(paired.ids)[1]) {
+  white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
+  black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])
+
+  if(length(white.index) != 0 & length(black.index) != 0) {
+    ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
+                ungroup() %>%
+                select(Age2, COD2, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
+                rename(prop.white1 = cod_prop_death1, prop.white5 = cod_prop_death5, prop.white9 = cod_prop_death9)
+
+    ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
+                ungroup() %>%
+                select(Age2, COD2, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
+                rename(prop.black1 = cod_prop_death1, prop.black5 = cod_prop_death5, prop.black9 = cod_prop_death9)
+    ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
+    ct.both <- ct.both %>% arrange(COD2, Age2)
+
+    if(any(is.na(ct.both$prop.black1)) == F & any(is.na(ct.both$prop.white1)) == F) { #ensures that there are no ages with 0 deaths - for these COD prop == 0 and can't compute
+
+        list.cod.decomp.tables[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
+                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
+                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
+                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white1, prop.black1), 
+                                                              age.colname.cod.table = "Age2", 
+                                                              COD.colname.cod.table = "COD2", 
+                                                              prop1.colname.cod.table = "prop.black1", 
+                                                              prop2.colname.cod.table = "prop.white1")
+
+        list.cod.decomp.tables5[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
+                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
+                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
+                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white5, prop.black5), 
+                                                              age.colname.cod.table = "Age2", 
+                                                              COD.colname.cod.table = "COD2", 
+                                                              prop1.colname.cod.table = "prop.black5", 
+                                                              prop2.colname.cod.table = "prop.white5")
+
+        list.cod.decomp.tables9[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
+                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
+                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
+                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white9, prop.black9), 
+                                                              age.colname.cod.table = "Age2", 
+                                                              COD.colname.cod.table = "COD2", 
+                                                              prop1.colname.cod.table = "prop.black9", 
+                                                              prop2.colname.cod.table = "prop.white9")
+
+    } else { list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Encountered age group with no deaths"
+
+    }
+  } else {
+
+    list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Couldn't compare life tables"
+  }
+
+  print(i)
 }
 )
-
-timer1 # 25 seconds for 5 -- guess 13 hrs for all the strata
 ```
 
-```{r calc_pop0}
-View(dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[2789], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
-                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1")])
-timer2 <- system.time (
-  for(x in levels(dat2$StateYearRaceSex)[2789]) {
-      for(row in 2:length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) { 
-        dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row - 1]
-        dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row - 1]
+```{r  update_decomp_for_plotting}
+#update the decomposition table for plotting purposes
+for(i in 1:dim(paired.ids)[1]) {
+    if(list.cod.decomp.tables[[i]][1] != "Couldn't compare life tables" & 
+       list.cod.decomp.tables[[i]][1] != "Encountered age group with no deaths") {
 
-          if (row != length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) {
-              dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying1[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying5[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dying9[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingb9[dat2$StateYearRaceSex == x][row]
-              dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]*dat2$Prob_Dyingw9[dat2$StateYearRaceSex == x][row]
+       list.cod.decomp.tables[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables[[i]],
+                                                            age.groups = "Ages", 
+                                                            cause.of.death = "Cause.of.death", 
+                                                            sign.var = "sign", decomp.var = "C_xi", 
+                                                            decomp.var.prop = "C_xi_proportion")
+
+       list.cod.decomp.tables5[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables5[[i]],
+                                                            age.groups = "Ages", 
+                                                            cause.of.death = "Cause.of.death", 
+                                                            sign.var = "sign", decomp.var = "C_xi", 
+                                                            decomp.var.prop = "C_xi_proportion")
+
+       list.cod.decomp.tables9[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables9[[i]],
+                                                            age.groups = "Ages", 
+                                                            cause.of.death = "Cause.of.death", 
+                                                            sign.var = "sign", decomp.var = "C_xi", 
+                                                            decomp.var.prop = "C_xi_proportion") 
     } 
-          else { dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] #everyone dies in the 85+ grouping
-                 dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row]
-                 dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row]
-                 dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row]
-                 dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row]
-
+  print(i)
 }  
-        }
-
-  }
-)
-
-timer2
 ```
 
-This takes `r time1[3]` seconds to run.
+```{r pretty_cod}
+ggplotly(ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
+           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
+           xlab("Contribution to life expectancy gap") +
+           ylab("Age group\n") + theme_minimal() +
+           geom_vline(xintercept = 0))
 
-```{r rest_of_life_expectancy_calc}
-dat2$num_PY_int1 <- (dat2$alive_at_start_int1 - dat2$deaths_during_int1)*dat2$numYrsInt + (dat2$deaths_during_int1*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
-dat2$num_PY_int5 <- (dat2$alive_at_start_int5 - dat2$deaths_during_int5)*dat2$numYrsInt + (dat2$deaths_during_int5*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
-dat2$num_PY_int9 <- (dat2$alive_at_start_int9 - dat2$deaths_during_int9)*dat2$numYrsInt + (dat2$deaths_during_int9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
-dat2$num_PY_intb9 <- (dat2$alive_at_start_intb9 - dat2$deaths_during_intb9)*dat2$numYrsInt + (dat2$deaths_during_intb9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
-dat2$num_PY_intw9 <- (dat2$alive_at_start_intw9 - dat2$deaths_during_intw9)*dat2$numYrsInt + (dat2$deaths_during_intw9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
-
-dat2$num_PY_int1[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_int1[dat2$Age2 == "85+ years"])/dat2$Death_Rate1[dat2$Age2 == "85+ years"]
-dat2$num_PY_int5[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_int5[dat2$Age2 == "85+ years"])/dat2$Death_Rate5[dat2$Age2 == "85+ years"]
-dat2$num_PY_int9[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_int9[dat2$Age2 == "85+ years"])/dat2$Death_Rate9[dat2$Age2 == "85+ years"]
-dat2$num_PY_intb9[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_intb9[dat2$Age2 == "85+ years"])/dat2$Death_Rateb9[dat2$Age2 == "85+ years"]
-dat2$num_PY_intw9[dat2$Age2 == "85+ years"] <- (dat2$alive_at_start_intw9[dat2$Age2 == "85+ years"])/dat2$Death_Ratew9[dat2$Age2 == "85+ years"]
-
-dat2 <- dat2 %>% group_by(StateYearRaceSex) %>% arrange(desc(Age2)) %>%
-                 mutate(tot_PY_after1 = cumsum(num_PY_int1), tot_PY_after5 = cumsum(num_PY_int5), tot_PY_after9 = cumsum(num_PY_int9),
-                        tot_PY_afterb9 = cumsum(num_PY_intb9),tot_PY_afterw9 = cumsum(num_PY_intw9),
-                        life_exp1 = tot_PY_after1/alive_at_start_int1, life_exp5 = tot_PY_after5/alive_at_start_int5, 
-                        life_exp9 = tot_PY_after9/alive_at_start_int9, 
-                        life_expb9 = tot_PY_afterb9/alive_at_start_intb9, life_expw9 = tot_PY_afterw9/alive_at_start_intw9 
-                )
-
-dat2 <- dat2 %>% group_by(StateYearRaceSex) %>% arrange(Age2)
+ggplotly(ggplot(list.cod.decomp.tables9[[1]], aes(y = Ages, x = start)) + 
+           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
+           xlab("Contribution to life expectancy gap") +
+           ylab("Age group\n") + theme_minimal() +
+           geom_vline(xintercept = 0))
 ```
 
-I compared the following with my calculations in the excel spreadsheet to check the calculations -- all good.
-```{r}
-dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[1], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
-                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1", 
-                                                                       "alive_at_start_int1", "deaths_during_int1", "num_PY_int1", "tot_PY_after1", "life_exp1")]
-
-dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[4], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
-                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1", 
-                                                                       "alive_at_start_int1", "deaths_during_int1", "num_PY_int1", "tot_PY_after1", "life_exp1")]
-
-dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[2789], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
-                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1", 
-                                                                       "alive_at_start_int1", "deaths_during_int1", "num_PY_int1", "tot_PY_after1", "life_exp1")]
-```
-
+###Initial graphs on missing data
 Notes on Figure 1 (below):
@@
 ```{r save_workspace}
-save.image(file = "/Users/corinneriddell/Dropbox/SRJ/Data/corinne/Sep5_workspace_bw_gap.Rdata")
+save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
 ```
+
+```{r save_specified_files}
+save(dat.clean, dat2, BlackWhite, BlackWhiteSmooth, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
+     list.age.decomp.tables5, list.cod.decomp.tables5, list.age.decomp.tables9, list.cod.decomp.tables9,
+     list.age.decomp.tables.smoothed, list.cod.decomp.tables.smoothed, list.cod.marginal.tables.smoothed,
+     non.zero.stratums, non.zero.stratums.smoothed, cod.contributions,
+     file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
+```

### 2017-01-18 Code/4_bayesSE.R
archiving

@@ -1,65 +1,0 @@
-
-library(rjags)
-library(R2jags) 
-library(devtools)
-library(jagstools)
-library(gtools)
-library(bayesboot)
-
-
- # Simulation to convince myself we can get post-hoc CIs for LE 
-
-# 1. Simulate data for a poisson glm 
-N = 100
-x = rnorm(N, mean=10, sd=1)
-mu = exp(1 + 0.2*x)
-y = rpois(N, lambda=mu) 
-plot(x,y)
-
-
-# 2. MCMC model 
-
-ds = list(x=x, y=y, N=N) 
-df = data.frame(x=x,y=y)
-
-head(ds)
-
-model = function() {  
-  for(i in 1:N) {
-    y[i] ~ dpois(mu[i])
-    log(mu[i]) <- beta0 + beta1*x[i] 
-  }
-  combo <- sum(mu[1:5]) / sum(mu[6:7])
-  beta0 ~ dnorm(0, 0.001)
-  beta1 ~ dnorm(0, 0.001)
-}
-
-params = c('mu', 'combo', 'beta0', 'beta1')
-nchains = 2
-
-jags_model = jags(data=ds, param=params, n.thin=1, n.chains=nchains, n.iter=2000, n.burnin=500, model.file=model) 
-
-class(jags_model)
-
-jags_model
-  
-#manual from posterior samples 
-
-p_mcmc = as.mcmc(jags_model) 
-p = data.frame(p_mcmc[[1]])   
-
-combo = (p$mu.1. + p$mu.2. + p$mu.3. + p$mu.4. +p$mu.5. ) / (p$mu.6. + p$mu.7.)
-
-#combo = log(pmu1*pmu2)*pmu3 + pmu4
-
-
-print(c(jagsresults(x=jags_model, params=c('combo'))[, '2.5%'],
-  jagsresults(x=jags_model, params=c('combo'))[, '97.5%']))
-  
-print(c(quantile(combo, 0.025),
-        quantile(combo, 0.975)))
-
-
-
-  
-

### 2017-01-18 Code/4_smoothing_post.R

@@ -1,62 +1,0 @@
-
-# Cleaning up the results imported from server 
-
-r1 = read.csv('/Users/kathryn/Documents/BlackWhiteMortalityGap/Code/results_femalesNov4.csv') 
-
-r2 = read.csv('/Users/kathryn/Documents/BlackWhiteMortalityGap/Code/results_malesNov4.csv')
-
-r1$X = NULL ; r2$X = NULL 
-
-r2$sex = rep('Male', length(r2$race)) 
-head(r2)
-
-r = rbind(r1, r2) 
-head(r) ; tail(r) 
-
-
-require(sqldf)
-
-r2 = sqldf("select * from r order by state, sex, race, age_bin, year, COD ") 
-head(r2) ; tail(r2)
-
-write.csv(r2, 'smoothed_results.csv') 
-
-head(dat.clean)
-
-plot(r2$deaths, r2$smoothed_deaths, pch=16, cex=0.1)
-
-
-# Quick figure to look at the results 
-
-# black males by age group & cod, alabama 
-
-ds1 = r2[r2$race=='Black' & r2$sex=='Male' & r2$state=='Alabama', ]
-head(ds1)
-
-ds2 = ds1[ ,c('deaths', 'smoothed_deaths', 'age_bin', 'year', 'cod')]
-
-ds3 = melt(ds2, id.var=c('year','age_bin', 'cod')) 
-head(ds3)
-
-
-g = ggplot(data=ds3) + geom_point(aes(x=year, y=value, col=variable)) + facet_grid(age_bin ~ cod, scales='free')
-
-g = ggpretty(g) 
-ggsave(g, filename='test.png', width=18, height=30) 
-
-
-
-
-ggpretty = function(g) { 
-  g = g + 
-    theme(legend.position="top", legend.title=element_blank(), 
-          panel.background = element_rect(fill = '#F9F9FE',color='white'),
-          axis.text.x = element_text(colour=c('black')),
-          axis.title.x = element_text(colour = 'black', size=8),
-          axis.title.y = element_text(colour = 'black', size=8, angle = 90, hjust = 0.5),
-          strip.background = element_rect(fill="#F1F3FD"))
-  return(g) 
-}
-
-
-

### 2017-01-18 Code/4_smoothing_time_fun.R

@@ -1,270 +1,0 @@
-
-
-subset_data = function(race, sex, cod, state) {
-  
-  selected_vars = c('Age', 'Sex2', 'Race2', 'COD2', 'Year', 'Count', 'Population', 'upper_bound')
-  
-  ds = dat.clean[(dat.clean$Race2==race & 
-                    dat.clean$Sex2==sex &
-                    dat.clean$State2==state &
-                    dat.clean$COD2==cod), selected_vars]
-  
-  df = data.frame(deaths=ds$Count, pop = ds$Population, age=(ds$Age+1), year=(ds$Year+1), upper_bound = ds$upper_bound) 
-  df$censored = ifelse(is.na(df$deaths),1,0)
-  
-  ds2 = df[order(df$censored), ]
-  
-  ds_jags = list(deaths = ds2$deaths, 
-                 lnpop = log(ds2$pop), 
-                 age.bin = ds2$age, 
-                 year = ds2$year, 
-                 upper_bound = ds2$upper_bound, 
-                 binned = is.na(ds2$deaths), 
-                 not.binned = !is.na(ds2$deaths), 
-                 n.binned = sum(is.na(ds2$deaths)), 
-                 n.not.binned = sum(!is.na(ds2$deaths)),
-                 n.not.binned.plus.1 = sum(!is.na(ds2$deaths)) + 1,
-                 n.rows = sum(is.na(ds2$deaths)) + sum(!is.na(ds2$deaths)),
-                 n.age.bins = length(unique(ds2$age)),
-                 n.years = length(unique(ds2$year)), 
-                 binned.id = ds2$censored) 
-  
-  return(ds_jags) 
-} 
-
-bwmort_smooth_time = function(ds_jags_bw) { 
-  
-  model = function() {  
-    
-    #BLACK 
-    for(i in 1:n.not.binned_b) {
-      deaths_b[i] ~ dpois(mu_b[i])
-      log(mu_b[i]) <- lnrate_b[age.bin_b[i], year_b[i]] + lnpop_b[i]
-    }
-    
-    for(i in n.not.binned.plus.1_b:n.rows_b) {
-      binned.id_b[i] ~ dinterval(deaths_b[i], c(0, upper_bound_b[i])) 
-      deaths_b[i] ~ dpois(mu_b[i])
[…]

### 2017-01-18 Code/4_smoothing_time_src.R

@@ -1,29 +1,0 @@
-
-library(rjags)
-library(R2jags) 
-library(devtools)
-#install_github(repo='johnbaums/jagstools') 
-library(jagstools)
-library(ggplot2)
-library(reshape2)
-library(parallel)
-library(gtools)
-library(bayesboot)
-library(coda)
-library(ggplot2)
-library(grid)
-library(gridExtra)
-
-source('~/BlackWhiteMortalityGap/Code/4_smoothing_time_fun.R') 
-source('~/BlackWhiteMortalityGap/Code/life_expectancy_functions.R') 
-
-load('~/black_white_mortality_project/main_datasets.Rdata') # change for SERVER 
-#load('/Users/kathryn/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata') # change for LOCAL 
-
-
-dat.clean$upper_bound = ifelse(dat.clean$Population<9, dat.clean$Population, 9) 
-states = unique(dat.clean$State2)
-ds = dat.clean
-
-
-

### 2017-01-18 Code/archived/4_smoothing_SE_fun.R
archiving

@@ -1,0 +1,50 @@
+
+
+extract_mcmc_dist = function(year, state, sex, n_post_samps) {
+  library(rjags)
+  library(R2jags) 
+  library(devtools)
+  library(jagstools)
+  library(gtools)
+  library(bayesboot)
+  library(coda)
+  Sex = ifelse(sex=='female', 'Female', 'Male') 
+  Year = year - 1968
+  name = paste0('results_',sex, '_', state, '.RData') 
+  load(name)
+  temp_df_list = list()
+  base_df = read.csv("base_df.csv")
+  cod_list = c('Injuries', 'Cardiovascular', 'Cancers', 'Communicable', 
+               'Non-communicable', 'All other causes')
+  temp_df = base_df[base_df$sex==Sex & base_df$state == state & base_df$year==Year, ]
+  n_agebins = length(unique(temp_df$age_bin))
+  
+  for(k in 1:n_post_samps) {
+    
+    temp_df$smooth_rate = rep(NA, length(temp_df$X))
+    
+    for(codi in 1:length(cod_list)) {
+      jags_modeli = results_female$jags_bw[[codi]]
+      p_mcmc = as.mcmc(jags_modeli) 
+      p = data.frame(p_mcmc[[1]])  
+      
+      
+      for(age_bini in 1:n_agebins) {
+        name_rate_w = paste0('lnrate_w.',age_bini, ".", Year, ".") 
+        name_rate_b = paste0('lnrate_b.',age_bini, ".", Year, ".") 
+        temp_df$smooth_rate[temp_df$age_bin==age_bini 
+                            & temp_df$race=='White' 
+                            & temp_df$cod==cod_list[codi]] = exp(p[name_rate_w])[[1]][k] 
+        
+        temp_df$smooth_rate[temp_df$age_bin==age_bini 
+                            & temp_df$race=='Black' 
+                            & temp_df$cod==cod_list[codi]] = exp(p[name_rate_b])[[1]][k] 
+      }
+    }
+    temp_df$smoothed_deaths = temp_df$smooth_rate * temp_df$population
+    temp_df_list[[k]] = temp_df
+    }
+    return(temp_df_list) 
+}
+
+

### 2017-01-18 Code/archived/4_smoothing_time_fun.R

@@ -1,0 +1,270 @@
+
+
+subset_data = function(race, sex, cod, state) {
+  
+  selected_vars = c('Age', 'Sex2', 'Race2', 'COD2', 'Year', 'Count', 'Population', 'upper_bound')
+  
+  ds = dat.clean[(dat.clean$Race2==race & 
+                    dat.clean$Sex2==sex &
+                    dat.clean$State2==state &
+                    dat.clean$COD2==cod), selected_vars]
+  
+  df = data.frame(deaths=ds$Count, pop = ds$Population, age=(ds$Age+1), year=(ds$Year+1), upper_bound = ds$upper_bound) 
+  df$censored = ifelse(is.na(df$deaths),1,0)
+  
+  ds2 = df[order(df$censored), ]
+  
+  ds_jags = list(deaths = ds2$deaths, 
+                 lnpop = log(ds2$pop), 
+                 age.bin = ds2$age, 
+                 year = ds2$year, 
+                 upper_bound = ds2$upper_bound, 
+                 binned = is.na(ds2$deaths), 
+                 not.binned = !is.na(ds2$deaths), 
+                 n.binned = sum(is.na(ds2$deaths)), 
+                 n.not.binned = sum(!is.na(ds2$deaths)),
+                 n.not.binned.plus.1 = sum(!is.na(ds2$deaths)) + 1,
+                 n.rows = sum(is.na(ds2$deaths)) + sum(!is.na(ds2$deaths)),
+                 n.age.bins = length(unique(ds2$age)),
+                 n.years = length(unique(ds2$year)), 
+                 binned.id = ds2$censored) 
+  
+  return(ds_jags) 
+} 
+
+bwmort_smooth_time = function(ds_jags_bw) { 
+  
+  model = function() {  
+    
+    #BLACK 
+    for(i in 1:n.not.binned_b) {
+      deaths_b[i] ~ dpois(mu_b[i])
+      log(mu_b[i]) <- lnrate_b[age.bin_b[i], year_b[i]] + lnpop_b[i]
+    }
+    
+    for(i in n.not.binned.plus.1_b:n.rows_b) {
+      binned.id_b[i] ~ dinterval(deaths_b[i], c(0, upper_bound_b[i])) 
+      deaths_b[i] ~ dpois(mu_b[i])
[…]

### 2017-01-18 Code/archived/4_smoothing_time_src.R

@@ -1,0 +1,29 @@
+
+library(rjags)
+library(R2jags) 
+library(devtools)
+#install_github(repo='johnbaums/jagstools') 
+library(jagstools)
+library(ggplot2)
+library(reshape2)
+library(parallel)
+library(gtools)
+library(bayesboot)
+library(coda)
+library(ggplot2)
+library(grid)
+library(gridExtra)
+
+source('~/BlackWhiteMortalityGap/Code/4_smoothing_time_fun.R') 
+source('~/BlackWhiteMortalityGap/Code/life_expectancy_functions.R') 
+
+load('~/black_white_mortality_project/main_datasets.Rdata') # change for SERVER 
+#load('/Users/kathryn/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata') # change for LOCAL 
+
+
+dat.clean$upper_bound = ifelse(dat.clean$Population<9, dat.clean$Population, 9) 
+states = unique(dat.clean$State2)
+ds = dat.clean
+
+
+

### 2017-01-18 Code/Run_model_on_SH_data/Prep SH HA data for our analysis.Rmd

---
title: "Compare with SH HA paper"
author: "Corinne Riddell"
date: "December 13, 2016"
output: html_document
---

```{r}
library(foreign)
library(dplyr)

dtotal <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/dataverse_files/deaths-total-nhisp.txt", header = F, sep="\t")

names(dtotal)
#no names affixed to dataset. But the names are stored in the data dictionary file found at:
#/Users/corinneriddell/Dropbox/BlackWhiteGap/dataverse_files/deaths-total-nhisp.dic

str(dtotal)
table(dtotal$V1) #50 states + DC
table(dtotal$V2) #20 age groups
table(dtotal$V3) #sex
table(dtotal$V4) #race
table(dtotal$V5) #20 years

names(dtotal) <- c("State", "Age", "Sex", "Race", "Year", "Crude_Rate", "Count", "Population") #see the data dictionary - path is above.
```

```{r recode_variables}
dtotal <- dtotal %>%
  mutate(State2 = factor(State, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
                                                             "Colorado", "Connecticut", "Delaware", "Washington DC",
                                                             "Florida", "Georgia", "Hawaii","Idaho", "Illinois", 
                                                             "Indiana","Iowa", "Kansas", "Kentucky", "Louisiana", 
                                                             "Maine", "Maryland", "Massachusetts", "Michigan", 
                                                             "Minnesota", "Mississippi","Missouri", "Montana", 
                                                             "Nebraska", "Nevada", "New Hampshire", "New Jersey",
                                                             "New Mexico", "New York","North Carolina","North Dakota",
                                                             "Ohio", "Oklahoma","Oregon", "Pennsylvania", 
                                                             "Rhode Island", "South Carolina", "South Dakota", 
                                                             "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
                                                             "Washington", "West Virginia", "Wisconsin", "Wyoming")), 
         Age2 = factor(Age, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                         "15-19 years", "20-24 years", "25-29 years", "30-34 years",
                                                         "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
                                                         "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
                                                         "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
         Age3 = 0*(Age2 == "<1 year") + 1*(Age2 == "1-4 years") + 5*(Age2 == "5-9 years") +
                10*(Age2 == "10-14 years") + 15*(Age2 == "15-19 years") + 20*(Age2 == "20-24 years") + 
                25*(Age2 == "25-29 years") + 30*(Age2 == "30-34 years") + 35*(Age2 == "35-39 years") +
                40*(Age2 == "40-44 years") + 45*(Age2 == "45-49 years") + 50*(Age2 == "50-54 years") + 
                55*(Age2 == "55-59 years") + 60*(Age2 == "60-64 years") + 65*(Age2 == "65-69 years") +   
                70*(Age2 == "70-74 years") + 75*(Age2 == "75-79 years") + 80*(Age2 == "80-84 years") + 
                85*(Age2 == "85+ years"),
         nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years"),
         Sex2 = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
         Race2 = factor(Race, levels = c(0, 1), labels = c("White", "Black")),
         Year2 = factor(Year, levels = c(0:19), labels = as.character(1990:2009)),
         Year3 = as.numeric(as.character(Year2)),
         RaceSex = interaction(Race2, Sex2, sep = " "),
         StateYearSex = interaction(State2, Year2, Sex2),
         StateYearRaceSex = interaction(State2, Year2, Race2, Sex2))
```

```{r unknown_age_group}
#there is an unknown age group for each combo of state-year-race-sex-age:
table(dtotal$Age2)

#Count of death is always equal to 0 for unknown age groups
table(dtotal$Age, dtotal$Count==0)

#remove these data from the data we analyze:
dtotal <- dtotal[dtotal$Age2 != "Unknown", ]

save(dtotal, file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/SH_HA_data.Rdata")
```

### 2017-01-18 Code/Run_model_on_SH_data/Prep SH HA data for our analysis.Rmd

@@
 library(dplyr)
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
 
-dtotal <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/dataverse_files/deaths-total-nhisp.txt", header = F, sep="\t")
+dtotal <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/dataverse_files/deaths-total.txt", header = F, sep="\t", stringsAsFactors = F, colClasses = c(rep("numeric", 8)),
+                     na.strings = ".")
 
@@
 #no names affixed to dataset. But the names are stored in the data dictionary file found at:
-#/Users/corinneriddell/Dropbox/BlackWhiteGap/dataverse_files/deaths-total-nhisp.dic
+#/Users/corinneriddell/Dropbox/BlackWhiteGap/dataverse_files/deaths-total.dic
 
@@
          StateYearSex = interaction(State2, Year2, Sex2),
-         StateYearRaceSex = interaction(State2, Year2, Race2, Sex2))
+         StateYearRaceSex = interaction(State2, Year2, Race2, Sex2),
+         Crude_Rate = Count/Population)
 ```
@@
 
-#Count of death is always equal to 0 for unknown age groups
+#no rows with the unknown age group (guess: was removed when pulling the data from CDC wonder)
 table(dtotal$Age, dtotal$Count==0)
@@
 
+excluded.states <- c("Idaho", "Montana", "North Dakota", "South Dakota", "Vermont")
+included.states <- setdiff(levels(dtotal$State2), excluded.states)
+dtotal <- dtotal %>% filter(State2 %in% included.states)
+
 save(dtotal, file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/SH_HA_data.Rdata")
 ```
+
+```{r}
+prop.table(table(is.na(dtotal$Count)))
+
+prop.table(table(is.na(dtotal$Count[dtotal$State2 == "New York"])))
+prop.table(table(is.na(dtotal$Count[dtotal$State2 == "Georgia"])))
+prop.table(table(is.na(dtotal$Count[dtotal$State2 == "Florida"])))
+
+test.le.fx <- function(data, state, sex, race, year) {
+  test.data <- data[data["State2"] == state & data["Sex2"] == sex & data["Race2"] == race & data["Year3"] == year, ]
+  lt <- life.table(data = test.data, num.ages.in.group = "nx",
+             death.counts = "Count", population.counts = "Population")
+  return(round(lt$e_x[1], 1))
+}
+
+test.le.fx(dtotal, "Florida", "Male", "White", 1990)
+test.le.fx(dtotal, "Florida", "Male", "White", 2009)
+
+test.le.fx(dtotal, "Georgia", "Male", "White", 1990)
+test.le.fx(dtotal, "Georgia", "Male", "White", 2009)
+
+test.le.fx(dtotal, "New York", "Male", "White", 1990)
+test.le.fx(dtotal, "New York", "Male", "White", 2009)
+```

### 2017-01-18 Code/Run_model_on_SH_data/Prep SH HA data for our analysis.Rmd
add COD col

@@
          StateYearRaceSex = interaction(State2, Year2, Race2, Sex2),
-         Crude_Rate = Count/Population)
+         Crude_Rate = Count/Population, 
+         COD = "Total")
 ```
@@
 dtotal <- dtotal[dtotal$Age2 != "Unknown", ]
+```
 
+```{r exclude_states}
 excluded.states <- c("Idaho", "Montana", "North Dakota", "South Dakota", "Vermont")

### 2017-01-19 Code/Run_model_on_SH_data/Prep SH HA data for our analysis.Rmd
added Alabama tests

@@
 prop.table(table(is.na(dtotal$Count[dtotal$State2 == "Florida"])))
+table(is.na(dtotal$Count[dtotal$State2 == "Alabama" & dtotal$Year3 == 1990]))
 
@@
 test.le.fx(dtotal, "New York", "Male", "White", 2009)
+
+test.le.fx(dtotal, "Alabama", "Male", "White", 1990)
+test.le.fx(dtotal, "Alabama", "Male", "White", 2009)
+
+test.le.fx(dtotal, "Alabama", "Male", "Black", 1990)
+#test.le.fx(dtotal, "Alabama", "Male", "Black", 2009)
+
 ```

### 2017-01-19 Code/3_Explore_Missingness.Rmd
cleaning up

--- 2017-01-04 Code/1_Explore_Data_v2.Rmd
@@
 ---
-title: "Calculating life expectancy and exploring missingness"
+title: "3_Explore_Missingness"
 author: "Corinne Riddell"
-date: "July 22, 2016"
+date: "1/19/2017"
 output: html_document
 ---
-
-```{r load_libs, echo=F, warning=F, message=F}
-library(plyr)
-library(dplyr)
-library(ggplot2)
-library(tidyr)
-library(plotly)
-source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
-```
-
-*About the dataset*
-The original dataset was collected by Sam using software from the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (known as the SEER*Stat Software) to access the National Vital Statistics System that is maintained by the National Center for Health Statistics.
-
-```{r read_data}
-dat <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/deaths-cause-6913.txt", header = F,sep="\t")
-names(dat) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
-```
-
-```{r recode_variables}
-dat <- dat %>%
-  mutate(State2 = factor(State, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
-                                                             "Colorado", "Connecticut", "Delaware", "Washington DC",
-                                                             "Florida", "Georgia", "Hawaii","Idaho", "Illinois", 
-                                                             "Indiana","Iowa", "Kansas", "Kentucky", "Louisiana", 
-                                                             "Maine", "Maryland", "Massachusetts", "Michigan", 
-                                                             "Minnesota", "Mississippi","Missouri", "Montana", 
-                                                             "Nebraska", "Nevada", "New Hampshire", "New Jersey",
-                                                             "New Mexico", "New York","North Carolina","North Dakota",
-                                                             "Ohio", "Oklahoma","Oregon", "Pennsylvania", 
-                                                             "Rhode Island", "South Carolina", "South Dakota", 
-                                                             "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
-                                                             "Washington", "West Virginia", "Wisconsin", "Wyoming")), 
-         Age2 = factor(Age, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
-                                                         "15-19 years", "20-24 years", "25-29 years", "30-34 years",
-                                                         "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
-                                                         "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
-                                                         "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
-         Age3 = 0*(Age2 == "<1 year") + 1*(Age2 == "1-4 years") + 5*(Age2 == "5-9 years") +
-                10*(Age2 == "10-14 years") + 15*(Age2 == "15-19 years") + 20*(Age2 == "20-24 years") + 
-                25*(Age2 == "25-29 years") + 30*(Age2 == "30-34 years") + 35*(Age2 == "35-39 years") +
-                40*(Age2 == "40-44 years") + 45*(Age2 == "45-49 years") + 50*(Age2 == "50-54 years") + 
-                55*(Age2 == "55-59 years") + 60*(Age2 == "60-64 years") + 65*(Age2 == "65-69 years") +   
-                70*(Age2 == "70-74 years") + 75*(Age2 == "75-79 years") + 80*(Age2 == "80-84 years") + 
-                85*(Age2 == "85+ years"),
-         nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years"),
-         Sex2 = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
-         Race2 = factor(Race, levels = c(0, 1), labels = c("White", "Black")),
-         COD2 = factor(COD, levels = c(0:5), labels = c("Cardiovascular", "Cancers", "Communicable", 
-                                                        "Non-communicable", "Injuries", "All other causes")), 
-         Year2 = factor(Year, levels = c(0:44), labels = as.character(1969:2013)),
-         Year3 = as.numeric(as.character(Year2)),
-         RaceSex = interaction(Race2, Sex2, sep = " "),
-         StateYearSex = interaction(State2, Year2, Sex2),
-         StateYearRaceSex = interaction(State2, Year2, Race2, Sex2))
-```
-
-```{r unknown_age_group}
-#there is an unknown age group for each combo of state-year-race-sex-age:
-table(dat$Age2)
-
-#Count of death is always equal to 0 for unknown age groups
-table(dat$Age, dat$Count==0)
-
-#remove these data from the data we analyze:
-dat.clean <- dat[dat$Age2 != "Unknown", ]
-```
-
-##Population counts of 0
-
-It is important to explore how many cells have population counts of zero because these break the Bayesian crank, and also currently breaks the life expectancy function.
-
-```{r explore_dataset}
-table(dat.clean$Population == 0, useNA = "always")
-#654 cells have Populations of 0.
-
-table(dat.clean$Count[dat.clean$Population == 0] == 0, useNA = "always")
-#Of these, all have Counts of 0, except for 26 with counts of NA
-
-#replace the NA Counts with 0 for the 26 cells with Populations of 0.
-dat.clean$Count[is.na(dat.clean$Count) == T & dat.clean$Population == 0] <- 0
-summary(dat.clean$Count[dat.clean$Population == 0])
-
-summary(dat.clean$Crude_Rate[dat.clean$Population == 0])
-#whenever the population is 0, the Crude_Rate is equal to NA.
-#need to think: should these be equal to 0, or NA? How will it affect the calculations?
-
-#States with at least one strata of 0
-na.pops <- dat.clean %>% 
-  select(State2, Population) %>% 
-  filter(Population == 0)
-
-unique(na.pops$State2)
-
-pops.le.20 <- dat.clean %>% select(State2, Population) %>% filter(Population <= 20)
-unique(pops.le.20$State2)
-```
-
-**Missingness of Count (of deaths)**
-If there are between 1 and 9 deaths in the strata, this information is 'coarsened' and coded as NA.
-Later in this document we investigate trends in supression/missingness to understand how it differs according to state, age, race, and sex.
-
-Since our goal is to compute the difference in life expectancy between Blacks and Whites by strata, we impute death counts at their logical extremes to explore this affects the results.
-
-- Set all missing counts to min({1, 5, and 9}, population_size).
-
-```{r fill_in_missing_counts}
-dat.clean <- dat.clean %>% mutate(Count_md1 = ifelse(is.na(Count)==T, pmin(1, Population), Count),
-                                  Count_md5 = ifelse(is.na(Count)==T, pmin(5, Population), Count),
-                                  Count_md9 = ifelse(is.na(Count)==T, pmin(9, Population), Count)
-                                  )
-
-#check recoding
-dat.clean$Count_grouped <- cut(dat.clean$Count, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
-dat.clean$Count1_grouped <- cut(dat.clean$Count_md1, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
-dat.clean$Count5_grouped <- cut(dat.clean$Count_md5, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
-dat.clean$Count9_grouped <- cut(dat.clean$Count_md9, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
-
-table(dat.clean$Count_grouped, dat.clean$Count1_grouped, useNA = "always")
-table(dat.clean$Count_grouped, dat.clean$Count5_grouped, useNA = "always")
-table(dat.clean$Count_grouped, dat.clean$Count9_grouped, useNA = "always")
-
-dat.clean <- dat.clean[ , !names(dat.clean) %in% c("Count_grouped", "Count1_grouped", "Count5_grouped", "Count9_grouped")]
-```
-
-```{r addvar_cod_proportions}
-dat.clean <- dat.clean %>% group_by(State2, RaceSex, Year3, Age3) %>%
-                           arrange(State2, RaceSex, Year3, Age3) %>%
-                           mutate(total_deaths1 = sum(Count_md1),
-                                  total_deaths5 = sum(Count_md5),
-                                  total_deaths9 = sum(Count_md9),
-                                  cod_prop_death1 = Count_md1/total_deaths1, 
-                                  cod_prop_death5 = Count_md5/total_deaths5,
-                                  cod_prop_death9 = Count_md9/total_deaths9)
-
-length(dat.clean$total_deaths5[dat.clean$total_deaths5 == 0])
-summary(dat.clean$cod_prop_death5[dat.clean$total_deaths5 == 0])
-#76998 strata with 0 deaths from any cause (12833 when divided by 6 -- the number of causes)
-#7% of strata have no deaths from any cause.
-```
-
-To calculate life expectancy, we have to aggregate the data across causes of death.
-
-```{r aggregate_over_COD}
-dat2 <- dat.clean %>% group_by(State2, Year2, RaceSex, Age2) %>%
-  arrange(State2, RaceSex, Year2, Age2) %>%
-  summarise(Population = first(Population), 
-            Race2 = first(Race2), 
-            Sex2 = first(Sex2), 
-            StateYearRaceSex = first(StateYearRaceSex),
-            nx = first(nx),
-            Age3 = first(Age3),
-            Year3 = first(Year3),
-            total_deaths1 = first(total_deaths1),
-            total_deaths5 = first(total_deaths5),
-            total_deaths9 = first(total_deaths9),
-            cod_prop_death1 = first(cod_prop_death1),
-            cod_prop_death5 = first(cod_prop_death5),
-            cod_prop_death9 = first(cod_prop_death9))
-
-#return stratum ids with any age-specific population counts of 0.
-#unique(dat2$StateYearRaceSex[dat2$Population == 0])
-non.zero.stratums <- setdiff(levels(dat2$StateYearRaceSex), unique(dat2$StateYearRaceSex[dat2$Population == 0]))
-```
-
-Add the total population (across age brackets) to the dat2 dataset.
-
-```{r create_var_pop_across_age}
-dat3 <- dat2 %>% ungroup() %>% 
-  group_by(StateYearRaceSex) %>%
-  summarise(Pop_across_age = sum(Population))
-
-dat2 <- merge(dat2, dat3, by = "StateYearRaceSex")
-rm(dat3)
-dat2 <- dat2 %>% arrange(State2, Year3, Race2, Sex2, Age3)
-```
-
-In those strata that do not have any age-specific population counts of 0, create a life table and store it in a list.
-
-```{r calculate_life_expectancy}
-#takes 10 mins for all three
-
-list.life.tables <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
-#list.life.tables1 <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
-list.index <- 1
-
-system.time(
-for(stratum.id in non.zero.stratums) {
-#for(stratum.id in head(non.zero.stratums)) {
-
-  list.life.tables[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2),   
-  #list.life.tables1[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
-                                               num.ages.in.group = "nx", 
-                                               death.counts = "total_deaths1", 
-                                               population.counts = "Population")
-
-  list.life.tables5[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
-                                               num.ages.in.group = "nx", 
-                                               death.counts = "total_deaths5", 
-                                               population.counts = "Population")
-
-  list.life.tables9[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
-                                               num.ages.in.group = "nx", 
-                                               death.counts = "total_deaths9", 
-                                               population.counts = "Population")
-  list.index <- list.index + 1
-  print(list.index)
-  }
-)
-```
-
-Add the life expectancy at birth to the data file:
-
-```{r add_le_to_dataset}
-le <- le5 <- le9 <- rep(NA, length(non.zero.stratums))
-
-for (i in 1:length(non.zero.stratums)) {
-  le[i] <- list.life.tables[[i]]$e_x[1]
-  le5[i] <- list.life.tables5[[i]]$e_x[1]
-  le9[i] <- list.life.tables9[[i]]$e_x[1]
-}
-
-les <- data.frame(life.expectancy.birth = le, 
-                  life.expectancy.birth.5 = le5, 
-                  life.expectancy.birth.9 = le9, 
-                  StateYearRaceSex = non.zero.stratums)
-
-dat2 <- merge(dat2, les, by = "StateYearRaceSex", all.x =T)
-
-rm(les)
-
-ggplot(dat2 %>% filter(Age2 == "<1 year"), aes(x = life.expectancy.birth)) + 
-  geom_density(aes(fill="Impute 1"), alpha = 0.3) + 
-  geom_density(aes(x = life.expectancy.birth.5, fill = "Impute 5"), alpha = 0.3) +
-  geom_density(aes(x = life.expectancy.birth.9, fill="Impute 9"), alpha = 0.3) +
-  facet_grid(Sex2 ~ Race2)
-```
-
-Make a new dataset that contains information on the gap:
-
-```{r create_dataset_BlackWhite}
-#this dataset contains the black-white data within the same row (wide format) for each year-state-sex strata
-
-Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
-                  select(State2, Year3, Sex2, Pop_across_age, 
-                         life.expectancy.birth, life.expectancy.birth.5, 
-                         life.expectancy.birth.9) %>%
-                  rename(Pop_black = Pop_across_age,  
-                         le.birth.black = life.expectancy.birth, 
-                         le.birth.black5 = life.expectancy.birth.5, 
-                         le.birth.black9 = life.expectancy.birth.9)
-
-White <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
-                  select(State2, Year3, Sex2, Pop_across_age, 
-                         life.expectancy.birth, life.expectancy.birth.5, 
-                         life.expectancy.birth.9) %>%
-                  rename(Pop_white = Pop_across_age,  
-                         le.birth.white = life.expectancy.birth,
-                         le.birth.white5 = life.expectancy.birth.5,
-                         le.birth.white9 = life.expectancy.birth.9)
-
-BlackWhite <- merge(Black, White, by = c("State2", "Year3", "Sex2"))
-
-BlackWhite <- BlackWhite %>% mutate(WBgap = le.birth.white - le.birth.black,
-                                    WBgap5 = le.birth.white5 - le.birth.black5,
-                                    WBgap9 = le.birth.white9 - le.birth.black9)
-
-rm(Black, White)
-```
-
-Decompose the gap by age. That is, for each year-state-sex combination, compare the life table for blacks to whites and decompose the gap in life expectancy at birth by age.
-
-```{r create_paired_ids_dataset}
-#paired.ids will contain the index names that are sent to the age decomposition function in the next code chunk
-
-#first, create list of stratum.ids for each race-sex combination
-for (rs in unique(levels(dat.clean$RaceSex))) {
-  assign(paste0(gsub(" ", "", rs), ".stratum.ids"), 
-         dat.clean %>% 
-           ungroup() %>%
-           filter(RaceSex == rs) %>%
-           select(StateYearRaceSex, StateYearSex, Sex2, State2, Year3) %>%
-           group_by(StateYearRaceSex, StateYearSex) %>%
-           filter(row_number() == 1)
-         )
-
-}
-
-#then, merge the black-white lists
-Female.ids <- merge(WhiteFemale.stratum.ids, BlackFemale.stratum.ids[, -c(3:5)], 
-                    by = "StateYearSex", suffixes = c("White", "Black"))
-head(Female.ids)
-
-Male.ids <- merge(WhiteMale.stratum.ids, BlackMale.stratum.ids[, -c(3:5)],
-                  by = "StateYearSex", suffixes = c("White", "Black"))
-head(Male.ids)
-
-paired.ids <- rbind(Male.ids, Female.ids)
-```
-
-```{r calculate_age_decomp}
-#runs in about a minute
-
-list.age.decomp.tables <- list.age.decomp.tables5 <- list.age.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])
-
-for(i in 1:dim(paired.ids)[1]) {
-  white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
-  black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])
-
-  if(length(white.index) != 0 & length(black.index) != 0) {
-    list.age.decomp.tables[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]), 
-                                                 name.lt1 = "Black", 
-                                                 life.table2 = as.data.frame(list.life.tables[[white.index]]), 
-                                                 name.lt2 = "White", age.groups = "Age3")
-
-    list.age.decomp.tables5[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]), 
-                                                 name.lt1 = "Black", 
-                                                 life.table2 = as.data.frame(list.life.tables5[[white.index]]), 
-                                                 name.lt2 = "White", age.groups = "Age3")
-
-    list.age.decomp.tables9[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables9[[black.index]]), 
-                                                 name.lt1 = "Black", 
-                                                 life.table2 = as.data.frame(list.life.tables9[[white.index]]), 
-                                                 name.lt2 = "White", age.groups = "Age3")
-  } else {
-    list.age.decomp.tables[[i]] <- list.age.decomp.tables5[[i]] <- list.age.decomp.tables9[[i]] <-"Couldn't compare life tables"
-  }
-}
-
-```
-
-```{r calculate_cod_decomp}
-#stopped here - need to figure this out
-#if the count of deaths == 0 for all causes of death within an age band then the COD proportion is NA
-#this messes up the cod decomposition. What to do?
-#example - no black people dying between the ages of 5-9 of anything. but have whites dying in that age group so their
-#COD proportions are okay.
-
-#takes 1.5 hours for all three
-list.cod.decomp.tables <- list.cod.decomp.tables5 <- list.cod.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])
-
-system.time(
-for(i in 1:dim(paired.ids)[1]) {
-  white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
-  black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])
-
-  if(length(white.index) != 0 & length(black.index) != 0) {
-    ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
-                ungroup() %>%
-                select(Age2, COD2, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
-                rename(prop.white1 = cod_prop_death1, prop.white5 = cod_prop_death5, prop.white9 = cod_prop_death9)
-
-    ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
-                ungroup() %>%
-                select(Age2, COD2, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
-                rename(prop.black1 = cod_prop_death1, prop.black5 = cod_prop_death5, prop.black9 = cod_prop_death9)
-    ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
-    ct.both <- ct.both %>% arrange(COD2, Age2)
-
-    if(any(is.na(ct.both$prop.black1)) == F & any(is.na(ct.both$prop.white1)) == F) { #ensures that there are no ages with 0 deaths - for these COD prop == 0 and can't compute
-
-        list.cod.decomp.tables[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
-                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
-                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
-                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white1, prop.black1), 
-                                                              age.colname.cod.table = "Age2", 
-                                                              COD.colname.cod.table = "COD2", 
-                                                              prop1.colname.cod.table = "prop.black1", 
-                                                              prop2.colname.cod.table = "prop.white1")
-
-        list.cod.decomp.tables5[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
-                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
-                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
-                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white5, prop.black5), 
-                                                              age.colname.cod.table = "Age2", 
-                                                              COD.colname.cod.table = "COD2", 
-                                                              prop1.colname.cod.table = "prop.black5", 
-                                                              prop2.colname.cod.table = "prop.white5")
-
-        list.cod.decomp.tables9[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
-                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
-                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
-                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white9, prop.black9), 
-                                                              age.colname.cod.table = "Age2", 
-                                                              COD.colname.cod.table = "COD2", 
-                                                              prop1.colname.cod.table = "prop.black9", 
-                                                              prop2.colname.cod.table = "prop.white9")
-
-    } else { list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Encountered age group with no deaths"
-
-    }
-  } else {
-
-    list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Couldn't compare life tables"
-  }
-
-  print(i)
-}
-)
-```
-
-```{r  update_decomp_for_plotting}
-#update the decomposition table for plotting purposes
-for(i in 1:dim(paired.ids)[1]) {
-    if(list.cod.decomp.tables[[i]][1] != "Couldn't compare life tables" & 
-       list.cod.decomp.tables[[i]][1] != "Encountered age group with no deaths") {
-
-       list.cod.decomp.tables[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables[[i]],
-                                                            age.groups = "Ages", 
-                                                            cause.of.death = "Cause.of.death", 
-                                                            sign.var = "sign", decomp.var = "C_xi", 
-                                                            decomp.var.prop = "C_xi_proportion")
-
-       list.cod.decomp.tables5[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables5[[i]],
-                                                            age.groups = "Ages", 
-                                                            cause.of.death = "Cause.of.death", 
-                                                            sign.var = "sign", decomp.var = "C_xi", 
-                                                            decomp.var.prop = "C_xi_proportion")
-
-       list.cod.decomp.tables9[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables9[[i]],
-                                                            age.groups = "Ages", 
-                                                            cause.of.death = "Cause.of.death", 
-                                                            sign.var = "sign", decomp.var = "C_xi", 
-                                                            decomp.var.prop = "C_xi_proportion") 
-    } 
-  print(i)
-}  
-```
-
-```{r pretty_cod}
-ggplotly(ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
-           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
-           xlab("Contribution to life expectancy gap") +
-           ylab("Age group\n") + theme_minimal() +
-           geom_vline(xintercept = 0))
-
-ggplotly(ggplot(list.cod.decomp.tables9[[1]], aes(y = Ages, x = start)) + 
-           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
-           xlab("Contribution to life expectancy gap") +
-           ylab("Age group\n") + theme_minimal() +
-           geom_vline(xintercept = 0))
-```
 
@@
 ```
-
-```{r save_workspace}
-save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
-```
-
-```{r save_specified_files}
-save(dat.clean, dat2, BlackWhite, BlackWhiteSmooth, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
-     list.age.decomp.tables5, list.cod.decomp.tables5, list.age.decomp.tables9, list.cod.decomp.tables9,
-     list.age.decomp.tables.smoothed, list.cod.decomp.tables.smoothed,
-     file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
-```

### 2017-01-20 Code/1_Explore_Data.Rmd

@@
 
-```{r load_libs, echo=F, warning=F, message=F}
-library(plyr)
-library(dplyr)
-library(ggplot2)
-library(tidyr)
-library(plotly)
-source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+```{r load_libraries, echo=F, warning=F, message=F}
+library(tidyverse)
 ```
@@
 
-```{r read_data}
+```{r read_raw_data}
 dat <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/deaths-cause-6913.txt", header = F,sep="\t")
 names(dat) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
+# names and order of variables in the dataset are noted in the .dic dictionary file found in the same dir as the dataset.
 ```
@@
 dat <- dat %>%
-  mutate(State2 = factor(State, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
+  mutate(State_factor = factor(State, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
                                                              "Colorado", "Connecticut", "Delaware", "Washington DC",
@@
                                                              "Washington", "West Virginia", "Wisconsin", "Wyoming")), 
-         Age2 = factor(Age, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
+         Age_factor = factor(Age, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                          "15-19 years", "20-24 years", "25-29 years", "30-34 years",
@@
                                                          "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
-         Age3 = 0*(Age2 == "<1 year") + 1*(Age2 == "1-4 years") + 5*(Age2 == "5-9 years") +
-                10*(Age2 == "10-14 years") + 15*(Age2 == "15-19 years") + 20*(Age2 == "20-24 years") + 
-                25*(Age2 == "25-29 years") + 30*(Age2 == "30-34 years") + 35*(Age2 == "35-39 years") +
-                40*(Age2 == "40-44 years") + 45*(Age2 == "45-49 years") + 50*(Age2 == "50-54 years") + 
-                55*(Age2 == "55-59 years") + 60*(Age2 == "60-64 years") + 65*(Age2 == "65-69 years") +   
-                70*(Age2 == "70-74 years") + 75*(Age2 == "75-79 years") + 80*(Age2 == "80-84 years") + 
-                85*(Age2 == "85+ years"),
-         nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years"),
-         Sex2 = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
-         Race2 = factor(Race, levels = c(0, 1), labels = c("White", "Black")),
-         COD2 = factor(COD, levels = c(0:5), labels = c("Cardiovascular", "Cancers", "Communicable", 
+         Age_minbin = 0*(Age_factor == "<1 year") + 1*(Age_factor == "1-4 years") + 5*(Age_factor == "5-9 years") +
+                10*(Age_factor == "10-14 years") + 15*(Age_factor == "15-19 years") + 20*(Age_factor == "20-24 years") + 
+                25*(Age_factor == "25-29 years") + 30*(Age_factor == "30-34 years") + 35*(Age_factor == "35-39 years") +
+                40*(Age_factor == "40-44 years") + 45*(Age_factor == "45-49 years") + 50*(Age_factor == "50-54 years") + 
+                55*(Age_factor == "55-59 years") + 60*(Age_factor == "60-64 years") + 65*(Age_factor == "65-69 years") +   
+                70*(Age_factor == "70-74 years") + 75*(Age_factor == "75-79 years") + 80*(Age_factor == "80-84 years") + 
+                85*(Age_factor == "85+ years"),
+         Age_bin_size = 1*(Age_factor == "<1 year") + 4*(Age_factor == "1-4 years") + 5*(Age_factor != "<1 year" & Age_factor != "1-4 years"),
+         Sex_factor = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
+         Race_factor = factor(Race, levels = c(0, 1), labels = c("White", "Black")),
+         COD_factor = factor(COD, levels = c(0:5), labels = c("Cardiovascular", "Cancers", "Communicable", 
                                                         "Non-communicable", "Injuries", "All other causes")), 
-         Year2 = factor(Year, levels = c(0:44), labels = as.character(1969:2013)),
-         Year3 = as.numeric(as.character(Year2)),
-         RaceSex = interaction(Race2, Sex2, sep = " "),
-         StateYearSex = interaction(State2, Year2, Sex2),
-         StateYearRaceSex = interaction(State2, Year2, Race2, Sex2))
+         Year_factor = factor(Year, levels = c(0:44), labels = as.character(1969:2013)),
+         Year_actual = as.numeric(as.character(Year_factor)),
+         RaceSex = interaction(Race_factor, Sex_factor, sep = " "),
+         StateYearSex = interaction(State_factor, Year_factor, Sex_factor),
+         StateYearRaceSex = interaction(State_factor, Year_factor, Race_factor, Sex_factor))
+
+head(dat)
 ```
 
-```{r unknown_age_group}
+```{r remove_unknown_age_group}
 #there is an unknown age group for each combo of state-year-race-sex-age:
-table(dat$Age2)
+table(dat$Age_factor)
 
 #Count of death is always equal to 0 for unknown age groups
-table(dat$Age, dat$Count==0)
+table(dat$Age_factor, dat$Count==0)
 
 #remove these data from the data we analyze:
-dat.clean <- dat[dat$Age2 != "Unknown", ]
+dat.clean <- dat[dat$Age_factor != "Unknown", ]
 ```
@@
 
-```{r explore_dataset}
+```{r recode_pops_of_zero}
 table(dat.clean$Population == 0, useNA = "always")
@@
 #need to think: should these be equal to 0, or NA? How will it affect the calculations?
+```
 
+```{r list_states_0_le20_popcounts}
 #States with at least one strata of 0
 na.pops <- dat.clean %>% 
-  select(State2, Population) %>% 
+  select(State_factor, Population) %>% 
   filter(Population == 0)
 
-unique(na.pops$State2)
+unique(na.pops$State_factor)
 
-pops.le.20 <- dat.clean %>% select(State2, Population) %>% filter(Population <= 20)
-unique(pops.le.20$State2)
+pops.le.20 <- dat.clean %>% select(State_factor, Population) %>% filter(Population <= 20)
+unique(pops.le.20$State_factor)
 ```
@@
 **Missingness of Count (of deaths)**
-If there are between 1 and 9 deaths in the strata, this information is 'coarsened' and coded as NA.
+If there are between 1 and 9 deaths in the strata, this information is 'coarsened' and the number of deaths (Count) is coded as NA.
 Later in this document we investigate trends in supression/missingness to understand how it differs according to state, age, race, and sex.
 
-Since our goal is to compute the difference in life expectancy between Blacks and Whites by strata, we impute death counts at their logical extremes to explore this affects the results.
-
-- Set all missing counts to min({1, 5, and 9}, population_size).
+Since our goal is to compute the difference in life expectancy between Blacks and Whites by strata, we impute death counts at their logical extremes to explore how this affects the results.
+Further, we use this rough data to create the shiny app while the Bayesian model is being developed/run.
 
@@
 ```{r addvar_cod_proportions} 
-dat.clean <- dat.clean %>% group_by(State2, RaceSex, Year3, Age3) %>%
-                           arrange(State2, RaceSex, Year3, Age3) %>%
+# Add the proportions of deaths due to each cause as a new variable.
+# This variable is used to decompose the life expectancy gap by cause of death.
+
+dat.clean <- dat.clean %>% group_by(State_factor, RaceSex, Year_actual, Age_minbin) %>%
+                           arrange(State_factor, RaceSex, Year_actual, Age_minbin) %>%
                            mutate(total_deaths1 = sum(Count_md1),
@@
 ```{r aggregate_over_COD}
-dat2 <- dat.clean %>% group_by(State2, Year2, RaceSex, Age2) %>%
-  arrange(State2, RaceSex, Year2, Age2) %>%
+dat.aggregated <- dat.clean %>% group_by(State_factor, Year_factor, RaceSex, Age_factor) %>%
+                                arrange(State_factor, RaceSex, Year_factor, Age_factor) %>%
                                 summarise(Population = first(Population), 
-            Race2 = first(Race2), 
-            Sex2 = first(Sex2), 
+                                          Race_factor = first(Race_factor), 
+                                          Sex_factor = first(Sex_factor), 
                                           StateYearRaceSex = first(StateYearRaceSex),
-            nx = first(nx),
-            Age3 = first(Age3),
-            Year3 = first(Year3),
+                                          Age_bin_size = first(Age_bin_size),
+                                          Age_minbin = first(Age_minbin),
+                                          Year_actual = first(Year_actual),
                                           total_deaths1 = first(total_deaths1),
@@
 
-#return stratum ids with any age-specific population counts of 0.
-#unique(dat2$StateYearRaceSex[dat2$Population == 0])
-non.zero.stratums <- setdiff(levels(dat2$StateYearRaceSex), unique(dat2$StateYearRaceSex[dat2$Population == 0]))
+#return "stratum ids" with any age-specific population counts of 0.
+non.zero.stratums <- setdiff(levels(dat.aggregated$StateYearRaceSex), 
+                             unique(dat.aggregated$StateYearRaceSex[dat.aggregated$Population == 0]))
 ```
 
-Add the total population (across age brackets) to the dat2 dataset.
+Add the total population (across age brackets) to the dat.aggregated dataset.
 
 ```{r create_var_pop_across_age}
-dat3 <- dat2 %>% ungroup() %>% 
+dat3 <- dat.aggregated %>% ungroup() %>% 
                            group_by(StateYearRaceSex) %>%
@@
 
-dat2 <- merge(dat2, dat3, by = "StateYearRaceSex")
+dat.aggregated <- merge(dat.aggregated, dat3, by = "StateYearRaceSex")
+
 rm(dat3)
-dat2 <- dat2 %>% arrange(State2, Year3, Race2, Sex2, Age3)
+
+dat.aggregated <- dat.aggregated %>% arrange(State_factor, Year_actual, Race_factor, Sex_factor, Age_minbin)
 ```
 
-In those strata that do not have any age-specific population counts of 0, create a life table and store it in a list.
-
-```{r calculate_life_expectancy}
-#takes 10 mins for all three
-
-list.life.tables <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
-#list.life.tables1 <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
-list.index <- 1
-
-system.time(
-for(stratum.id in non.zero.stratums) {
-#for(stratum.id in head(non.zero.stratums)) {
-
-  list.life.tables[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2),   
-  #list.life.tables1[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
-                                               num.ages.in.group = "nx", 
-                                               death.counts = "total_deaths1", 
-                                               population.counts = "Population")
-
-  list.life.tables5[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
-                                               num.ages.in.group = "nx", 
-                                               death.counts = "total_deaths5", 
-                                               population.counts = "Population")
-
-  list.life.tables9[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
-                                               num.ages.in.group = "nx", 
-                                               death.counts = "total_deaths9", 
-                                               population.counts = "Population")
-  list.index <- list.index + 1
-  print(list.index)
-  }
-)
-```
-
-Add the life expectancy at birth to the data file:
-
-```{r add_le_to_dataset}
-le <- le5 <- le9 <- rep(NA, length(non.zero.stratums))
-
-for (i in 1:length(non.zero.stratums)) {
-  le[i] <- list.life.tables[[i]]$e_x[1]
-  le5[i] <- list.life.tables5[[i]]$e_x[1]
-  le9[i] <- list.life.tables9[[i]]$e_x[1]
-}
-
-les <- data.frame(life.expectancy.birth = le, 
-                  life.expectancy.birth.5 = le5, 
-                  life.expectancy.birth.9 = le9, 
-                  StateYearRaceSex = non.zero.stratums)
-
-dat2 <- merge(dat2, les, by = "StateYearRaceSex", all.x =T)
-
-rm(les)
-
-ggplot(dat2 %>% filter(Age2 == "<1 year"), aes(x = life.expectancy.birth)) + 
-  geom_density(aes(fill="Impute 1"), alpha = 0.3) + 
-  geom_density(aes(x = life.expectancy.birth.5, fill = "Impute 5"), alpha = 0.3) +
-  geom_density(aes(x = life.expectancy.birth.9, fill="Impute 9"), alpha = 0.3) +
-  facet_grid(Sex2 ~ Race2)
-```
-
-Make a new dataset that contains information on the gap:
-
-```{r create_dataset_BlackWhite}
-#this dataset contains the black-white data within the same row (wide format) for each year-state-sex strata
-
-Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
-                  select(State2, Year3, Sex2, Pop_across_age, 
-                         life.expectancy.birth, life.expectancy.birth.5, 
-                         life.expectancy.birth.9) %>%
-                  rename(Pop_black = Pop_across_age,  
-                         le.birth.black = life.expectancy.birth, 
-                         le.birth.black5 = life.expectancy.birth.5, 
-                         le.birth.black9 = life.expectancy.birth.9)
-
-White <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
-                  select(State2, Year3, Sex2, Pop_across_age, 
-                         life.expectancy.birth, life.expectancy.birth.5, 
-                         life.expectancy.birth.9) %>%
-                  rename(Pop_white = Pop_across_age,  
-                         le.birth.white = life.expectancy.birth,
-                         le.birth.white5 = life.expectancy.birth.5,
-                         le.birth.white9 = life.expectancy.birth.9)
-
-BlackWhite <- merge(Black, White, by = c("State2", "Year3", "Sex2"))
-
-BlackWhite <- BlackWhite %>% mutate(WBgap = le.birth.white - le.birth.black,
-                                    WBgap5 = le.birth.white5 - le.birth.black5,
-                                    WBgap9 = le.birth.white9 - le.birth.black9)
-
-rm(Black, White)
-```
-
-Decompose the gap by age. That is, for each year-state-sex combination, compare the life table for blacks to whites and decompose the gap in life expectancy at birth by age.
-
-```{r create_paired_ids_dataset}
-#paired.ids will contain the index names that are sent to the age decomposition function in the next code chunk
-
-#first, create list of stratum.ids for each race-sex combination
-for (rs in unique(levels(dat.clean$RaceSex))) {
-  assign(paste0(gsub(" ", "", rs), ".stratum.ids"), 
-         dat.clean %>% 
-           ungroup() %>%
-           filter(RaceSex == rs) %>%
-           select(StateYearRaceSex, StateYearSex, Sex2, State2, Year3) %>%
-           group_by(StateYearRaceSex, StateYearSex) %>%
-           filter(row_number() == 1)
-         )
-
-}
-
-#then, merge the black-white lists
-Female.ids <- merge(WhiteFemale.stratum.ids, BlackFemale.stratum.ids[, -c(3:5)], 
-                    by = "StateYearSex", suffixes = c("White", "Black"))
-head(Female.ids)
-
-Male.ids <- merge(WhiteMale.stratum.ids, BlackMale.stratum.ids[, -c(3:5)],
-                  by = "StateYearSex", suffixes = c("White", "Black"))
-head(Male.ids)
-
-paired.ids <- rbind(Male.ids, Female.ids)
-```
-
-```{r calculate_age_decomp}
-#runs in about a minute
-
-list.age.decomp.tables <- list.age.decomp.tables5 <- list.age.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])
-
-for(i in 1:dim(paired.ids)[1]) {
-  white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
-  black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])
-
-  if(length(white.index) != 0 & length(black.index) != 0) {
-    list.age.decomp.tables[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]), 
-                                                 name.lt1 = "Black", 
-                                                 life.table2 = as.data.frame(list.life.tables[[white.index]]), 
-                                                 name.lt2 = "White", age.groups = "Age3")
-
-    list.age.decomp.tables5[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]), 
-                                                 name.lt1 = "Black", 
-                                                 life.table2 = as.data.frame(list.life.tables5[[white.index]]), 
-                                                 name.lt2 = "White", age.groups = "Age3")
-
-    list.age.decomp.tables9[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables9[[black.index]]), 
-                                                 name.lt1 = "Black", 
-                                                 life.table2 = as.data.frame(list.life.tables9[[white.index]]), 
-                                                 name.lt2 = "White", age.groups = "Age3")
-  } else {
-    list.age.decomp.tables[[i]] <- list.age.decomp.tables5[[i]] <- list.age.decomp.tables9[[i]] <-"Couldn't compare life tables"
-  }
-}
+```{r write_data_to_csvs}
+#save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
+write_csv(dat.clean, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.Rdata", col_names = T)
+write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", col_names = T)
 
@@
 
-```{r calculate_cod_decomp}
-#stopped here - need to figure this out
-#if the count of deaths == 0 for all causes of death within an age band then the COD proportion is NA
-#this messes up the cod decomposition. What to do?
-#example - no black people dying between the ages of 5-9 of anything. but have whites dying in that age group so their
-#COD proportions are okay.
-
-#takes 1.5 hours for all three
-list.cod.decomp.tables <- list.cod.decomp.tables5 <- list.cod.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])
-
-system.time(
-for(i in 1:dim(paired.ids)[1]) {
-  white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
-  black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])
-
-  if(length(white.index) != 0 & length(black.index) != 0) {
-    ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
-                ungroup() %>%
-                select(Age2, COD2, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
-                rename(prop.white1 = cod_prop_death1, prop.white5 = cod_prop_death5, prop.white9 = cod_prop_death9)
-
-    ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
-                ungroup() %>%
-                select(Age2, COD2, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
-                rename(prop.black1 = cod_prop_death1, prop.black5 = cod_prop_death5, prop.black9 = cod_prop_death9)
-    ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
-    ct.both <- ct.both %>% arrange(COD2, Age2)
-
-    if(any(is.na(ct.both$prop.black1)) == F & any(is.na(ct.both$prop.white1)) == F) { #ensures that there are no ages with 0 deaths - for these COD prop == 0 and can't compute
-
-        list.cod.decomp.tables[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
-                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
-                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
-                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white1, prop.black1), 
-                                                              age.colname.cod.table = "Age2", 
-                                                              COD.colname.cod.table = "COD2", 
-                                                              prop1.colname.cod.table = "prop.black1", 
-                                                              prop2.colname.cod.table = "prop.white1")
-
-        list.cod.decomp.tables5[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
-                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
-                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
-                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white5, prop.black5), 
-                                                              age.colname.cod.table = "Age2", 
-                                                              COD.colname.cod.table = "COD2", 
-                                                              prop1.colname.cod.table = "prop.black5", 
-                                                              prop2.colname.cod.table = "prop.white5")
-
-        list.cod.decomp.tables9[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
-                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
-                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
-                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white9, prop.black9), 
-                                                              age.colname.cod.table = "Age2", 
-                                                              COD.colname.cod.table = "COD2", 
-                                                              prop1.colname.cod.table = "prop.black9", 
-                                                              prop2.colname.cod.table = "prop.white9")
-
-    } else { list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Encountered age group with no deaths"
-
-    }
-  } else {
-
-    list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Couldn't compare life tables"
-  }
-
-  print(i)
-}
-)
+```{r save_specified_files}
+# save(dat.clean, dat2, BlackWhite, BlackWhiteSmooth, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
+#      list.age.decomp.tables5, list.cod.decomp.tables5, list.age.decomp.tables9, list.cod.decomp.tables9,
+#      list.age.decomp.tables.smoothed, list.cod.decomp.tables.smoothed, list.cod.marginal.tables.smoothed,
+#      non.zero.stratums, non.zero.stratums.smoothed, cod.contributions,
+#      file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
 ```
-
-```{r  update_decomp_for_plotting}
-#update the decomposition table for plotting purposes
-for(i in 1:dim(paired.ids)[1]) {
-    if(list.cod.decomp.tables[[i]][1] != "Couldn't compare life tables" & 
-       list.cod.decomp.tables[[i]][1] != "Encountered age group with no deaths") {
-
-       list.cod.decomp.tables[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables[[i]],
-                                                            age.groups = "Ages", 
-                                                            cause.of.death = "Cause.of.death", 
-                                                            sign.var = "sign", decomp.var = "C_xi", 
-                                                            decomp.var.prop = "C_xi_proportion")
-
-       list.cod.decomp.tables5[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables5[[i]],
-                                                            age.groups = "Ages", 
-                                                            cause.of.death = "Cause.of.death", 
-                                                            sign.var = "sign", decomp.var = "C_xi", 
-                                                            decomp.var.prop = "C_xi_proportion")
-
-       list.cod.decomp.tables9[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables9[[i]],
-                                                            age.groups = "Ages", 
-                                                            cause.of.death = "Cause.of.death", 
-                                                            sign.var = "sign", decomp.var = "C_xi", 
-                                                            decomp.var.prop = "C_xi_proportion") 
-    } 
-  print(i)
-}  
-```
-
-```{r pretty_cod}
-ggplotly(ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
-           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
-           xlab("Contribution to life expectancy gap") +
-           ylab("Age group\n") + theme_minimal() +
-           geom_vline(xintercept = 0))
-
-ggplotly(ggplot(list.cod.decomp.tables9[[1]], aes(y = Ages, x = start)) + 
-           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
-           xlab("Contribution to life expectancy gap") +
-           ylab("Age group\n") + theme_minimal() +
-           geom_vline(xintercept = 0))
-```
-
-###Initial graphs on missing data
-Notes on Figure 1 (below):
-
-- Figure 1 is sorted by size of the black population in the state in 2013.
-This is so you can see how missingness is a function of population size.
-We expect that states with less black people will have more missing data and more counts of exactly zero.
-Thus, states with the biggest populations (of blacks and whites) will "hold constant" the effect of population size on the amount of missing data.
-
-- The first panels you see in Figure 1 (below) are for the states with the smallest black populations.
-For whites, the amount of missing data increases in age until about 40 years old, and then decreases.
-For blacks, the amount of missing data is very low across age.
-This seemed strange to me at first, but then I remembered that the data is not missing if the count is equal to 0, which will be the case for many strata in these states with the lowest black population.
-
-- As you read left to right (as the black population size increases by state), the trends in missing data across age look more similar by race, although blacks appear to have more missing data in individuals < 30 years old for the states with the most black people.
-
-**Figure 1: Proportion missing by State (panel) and Age group (x-axis), ordered by increasing size of black population in 2013**
-```{r bystate, fig.width = 12, fig.height = 37, echo=F}
-bystate <- 
-  dat %>%
-  group_by(State2, Race2, Age2) %>%
-  summarise(prop.missing = mean(is.na(Count)))
-
-#recode a new count variable that "counts the missing data" so that these rows are used when calculating the % of data == 0.
-dat$Count2 <- dat$Count
-dat$Count2[is.na(dat$Count) == T] <- 1
-
-bystate2 <- 
-  dat %>%
-  group_by(State2, Race2, Age2) %>%
-  summarise(prop.zero = mean(Count2 == 0))
-
-bystate <- merge(bystate, bystate2, by=c("State2", "Race2", "Age2"))
-
-#calculate the size of the black population in 2013 by State
-bystate.bpop <- 
-  dat %>%
-  filter(COD2 == "Cancers" & Year2 == 2013 & Race2 == "Black") %>% #to stop repeats of the population size
-  group_by(State2) %>%
-  summarize(pop_by_race = sum(as.numeric(Population)))
-
-#calculate the size of the black population in 1969 by State
-bystate.bpop.1969 <- 
-  dat %>%
-  filter(COD2 == "Cancers" & Year2 == 1969 & Race2 == "Black") %>% #to stop repeats of the population size
-  group_by(State2) %>%
-  summarize(pop_by_race_1969 = sum(as.numeric(Population)))
-
-bystate <- merge(bystate, bystate.bpop, by = "State2", all.x = T)
-bystate <- merge(bystate, bystate.bpop.1969, by = "State2", all.x = T)
-bystate$State3 <- reorder(bystate$State2, bystate$pop_by_race)
-
-ggplot(bystate, aes(x = Age2, y = prop.missing)) + 
-  geom_point(aes(col=Race2)) + facet_wrap(~ State3, ncol = 4) +
-  theme_bw(base_size = 11) +
-  theme(legend.position="top") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  ylab("Proportion Missing (%)") +
-  xlab("Age group")
-
-ggplot(subset(bystate, Age2 == "<1 year"), aes(pop_by_race_1969, pop_by_race)) + geom_point() + geom_abline(intercept = 0, slope = 1)
-
-unique(bystate$State2[bystate$Age2 == "<1 year" & bystate$pop_by_race_1969 < 10000])
-```
-
-Notes on Figure 2 (below):
-
-- Figure 2 examines the proportion of counts equal to zero.
-The purpose of this examination is to visually confirm that there are higher proportions of zero counts in the states with the least amount of blacks.
-This is very clear when viewing the figure.
-
-- Question - any explantion for why the portion of zeroes stays relatively higher in blacks vs. whites in the states with the most blacks (within the younger age groups)?
-
-**Figure 2: Proportion zero by State (panel) and Age group (x-axis), ordered by increasing size of black population in 2013**
-
-```{r zero_by_state, fig.width = 12, fig.height = 37, echo=F}
-ggplot(bystate, aes(x = Age2, y = prop.zero)) + 
-  geom_point(aes(col=Race2)) + facet_wrap(~ State3, ncol = 4) +
-  theme_bw(base_size = 11) +
-  theme(legend.position="top") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  ylab("Proportion Zero (%)") +
-  xlab("Age group") 
-```
-
-###Other plots (less interesting/informative)
-
-**How is missingness related to age, race, and sex?**
-
-```{r missing_data, fig.width = 10, fig.height = 5, echo=F}
-#table(is.na(dat$Count))
-
-table1.whitemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "White Male"]), 
-                                           dat$Age2[dat$RaceSex == "White Male"]),2)*100,2)
-table1.blackmale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "Black Male"]), 
-                                           dat$Age2[dat$RaceSex == "Black Male"]),2)*100,2)
-table1.whitefemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "White Female"]), 
-                                           dat$Age2[dat$RaceSex == "White Female"]),2)*100,2)
-table1.blackfemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "Black Female"]), 
-                                           dat$Age2[dat$RaceSex == "Black Female"]),2)*100,2)
-
-miss1 <- data.frame("white.male" = table1.whitemale[2, ], 
-                    "black.male"= table1.blackmale[2, ],
-                    "white.female" = table1.whitefemale[2, ], 
-                    "black.female"= table1.blackfemale[2, ],
-                    "age" = names(table1.whitemale[2, ]))
-miss1$age <- factor(miss1$age, levels(miss1$age)[c(1, 2, 11, 3:10, 12:20)])
-
-miss1 <- gather(miss1, racesex, missingness, c(white.male, black.male, white.female, black.female))
-
-ggplot(miss1, aes(x = age, y = missingness)) + geom_point() + geom_point(aes(col=racesex)) +
-  theme_bw(base_size = 16) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  ylab("Proportion Missing (%)") +
-  xlab("Age group")
-```
-
-**Let's look at the proportion with 0's**
-
-```{r zero_data, fig.width = 10, fig.height = 5, echo=F}
-table2.whitemale <- round(prop.table(table(dat$Count[dat$RaceSex == "White Male"]==0, 
-                                       dat$Age2[dat$RaceSex == "White Male"]),2)*100,2)
-table2.blackmale <- round(prop.table(table(dat$Count[dat$RaceSex == "Black Male"]==0, 
-                                       dat$Age2[dat$RaceSex == "Black Male"]),2)*100,2)
-table2.whitefemale <- round(prop.table(table(dat$Count[dat$RaceSex == "White Female"]==0, 
-                                       dat$Age2[dat$RaceSex == "White Female"]),2)*100,2)
-table2.blackfemale <- round(prop.table(table(dat$Count[dat$RaceSex == "Black Female"]==0, 
-                                       dat$Age2[dat$RaceSex == "Black Female"]),2)*100,2)
-
-miss2 <- data.frame("whitemale" = table2.whitemale[2, ], 
-                    "blackmale"= table2.blackmale[2, ],
-                    "whitefemale" = table2.whitefemale[2, ], 
-                    "blackfemale"= table2.blackfemale[2, ],
-                    "age" = names(table2.blackmale[2, ]))
-
-miss2$age <- factor(miss2$age, levels(miss2$age)[c(1, 2, 11, 3:10, 12:20)])
-
-miss2 <- gather(miss2, racesex, zeroes, c(whitemale, blackmale, whitefemale, blackfemale))
-
-ggplot(miss2, aes(x = age, y = zeroes)) + geom_point() + geom_point(aes(col=racesex)) +
-  theme_bw(base_size = 16) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  ylab("Proportion Zero (%)") +
-  xlab("Age group")
-```
-
-**Missing or zero (which is equivalent to < 10)**
-
-```{r miss_or_zero_data, fig.width = 10, fig.height = 5, echo=F}
-dat$miss_or_zero <- F
-dat$miss_or_zero[dat$Count == 0 | is.na(dat$Count) == T] <- T
-
-table3.whitemale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "White Male"]==T, 
-                                      dat$Age2[dat$RaceSex == "White Male"]),2)*100,2)
-table3.blackmale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "Black Male"]==T, 
-                                       dat$Age2[dat$RaceSex == "Black Male"]),2)*100,2)
-table3.whitefemale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "White Female"]==T, 
-                                       dat$Age2[dat$RaceSex == "White Female"]),2)*100,2)
-table3.blackfemale <- round(prop.table(table(dat$miss_or_zero[dat$RaceSex == "Black Female"]==T, 
-                                       dat$Age2[dat$RaceSex == "Black Female"]),2)*100,2)
-
-miss3 <- data.frame("whitemale" = table3.whitemale[2, ], 
-                    "blackmale"= table3.blackmale[2, ],
-                    "whitefemale" = table3.whitefemale[2, ], 
-                    "blackfemale"= table3.blackfemale[2, ],
-                    "age" = names(table3.blackmale[2, ]))
-
-miss3$age <- factor(miss3$age, levels(miss3$age)[c(1, 2, 11, 3:10, 12:20)])
-
-miss3 <- gather(miss3, racesex, miss_or_zero, c(whitemale, blackmale, whitefemale, blackfemale))
-
-ggplot(miss3, aes(x = age, y = miss_or_zero)) + geom_point() + geom_point(aes(col=racesex)) +
-  theme_bw(base_size = 16) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  ylab("Proportion Missing or Zero (%)") +
-  xlab("Age group")
-```
-
-```{r explore_data2, fig.width = 10, fig.height = 5, echo=F, eval=F}
-
-summary(dat$Count)
-ggplot(subset(dat, Count>0), aes(x = Count)) + geom_histogram(binwidth=100)
-
-ggplot(subset(dat, State2=="Louisiana"), aes(x = Population, y = Count)) + 
-  geom_point(aes(col=RaceSex), alpha=0.2) +
-  facet_wrap(~COD2) + theme_bw()
-
-```
-
-```{r save_workspace}
-save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
-```
-
-```{r save_specified_files}
-save(dat.clean, dat2, BlackWhite, BlackWhiteSmooth, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
-     list.age.decomp.tables5, list.cod.decomp.tables5, list.age.decomp.tables9, list.cod.decomp.tables9,
-     list.age.decomp.tables.smoothed, list.cod.decomp.tables.smoothed, list.cod.marginal.tables.smoothed,
-     non.zero.stratums, non.zero.stratums.smoothed, cod.contributions,
-     file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
-```

### 2017-01-20 Code/2_Run_LE_and_Decomp_Calcs.Rmd

---
title: "Run Life expectancy and decomposition calculations"
author: "Corinne Riddell"
date: "1/19/2017"
output: html_document
---

Depends on: output datasets created in "1_Explore_Data.Rmd".

```{r load_libraries, echo=F, warning=F, message=F}
library(tidyverse)
library(plotly)
options(tibble.width = Inf) #display all columns of tibbles
```

```{r load_my_functions}
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

```{r load_data}
dat.aggregated <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", header = T)
dat.clean<- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", header = T)
```

In those strata that do not have any age-specific population counts of 0, create a life table and store it in a list.

```{r calculate_life_expectancy}
non.zero.stratums <- setdiff(levels(dat.aggregated$StateYearRaceSex), 
                             unique(dat.aggregated$StateYearRaceSex[dat.aggregated$Population == 0]))

#takes 13 mins

list.life.tables1 <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
list.index <- 1

system.time(
  for(stratum.id in non.zero.stratums) {

    list.life.tables1[[list.index]]$stratum.id <- non.zero.stratums[list.index]
    list.life.tables5[[list.index]]$stratum.id <- non.zero.stratums[list.index]
    list.life.tables9[[list.index]]$stratum.id <- non.zero.stratums[list.index]

    list.life.tables1[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, StateYearRaceSex == stratum.id) %>% arrange(Age_factor),
                                                 num.ages.in.group = "Age_bin_size",
                                                 death.counts = "total_deaths1",
                                                 population.counts = "Population")

    list.life.tables5[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, StateYearRaceSex == stratum.id) %>% arrange(Age_factor),
                                                  num.ages.in.group = "Age_bin_size",
                                                  death.counts = "total_deaths5",
                                                  population.counts = "Population")

    list.life.tables9[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, StateYearRaceSex == stratum.id) %>% arrange(Age_factor),
                                                  num.ages.in.group = "Age_bin_size",
                                                  death.counts = "total_deaths9",
                                                  population.counts = "Population")
    list.index <- list.index + 1
    print(list.index)
  }
)
```

Add the life expectancy at birth to the data file:

```{r add_le_to_dataset}
le1 <- map_dbl(list.life.tables1, ~ .x$life.table$e_x[1])
le5 <- map_dbl(list.life.tables5, ~ .x$life.table$e_x[1])
le9 <- map_dbl(list.life.tables9, ~ .x$life.table$e_x[1])

sid <- map_chr(list.life.tables1, ~ .x$stratum.id)

les <- tibble(le1, le5, le9, sid)
names(les) = c("life.expectancy.birth.1", "life.expectancy.birth.5", "life.expectancy.birth.9", "StateYearRaceSex")

dat.aggregated <- merge(dat.aggregated, les, by = "StateYearRaceSex", all.x =T)

rm(les, le1, le5, le9)
```

```{r densityplot_life_expectancy}
ggplot(dat.aggregated %>% dplyr::filter(Age_factor == "<1 year"), aes(x = life.expectancy.birth.1)) + 
  geom_density(aes(fill="Impute 1"), alpha = 0.3) +
  geom_density(aes(x = life.expectancy.birth.5, fill = "Impute 5"), alpha = 0.3) +
  geom_density(aes(x = life.expectancy.birth.9, fill="Impute 9"), alpha = 0.3) +
  facet_grid(Sex_factor ~ Race_factor)
```

Make a new dataset that contains information on the gap:

```{r create_dataset_BlackWhite}
#this dataset contains the black-white data within the same row (wide format) for each year-state-sex strata

Black <- dat.aggregated %>% filter(Race_factor == "Black" & Age_minbin == 0) %>% 
                            select(State_factor, Year_actual, Sex_factor, Pop_across_age, 
                                   life.expectancy.birth.1, life.expectancy.birth.5, 
                                   life.expectancy.birth.9) %>%
                            rename(Pop_black = Pop_across_age,  
                                   le.birth.black1 = life.expectancy.birth.1, 
                                   le.birth.black5 = life.expectancy.birth.5, 
                                   le.birth.black9 = life.expectancy.birth.9)

White <- dat.aggregated %>% filter(Race_factor == "White" & Age_minbin == 0) %>% 
                            select(State_factor, Year_actual, Sex_factor, Pop_across_age, 
                                   life.expectancy.birth.1, life.expectancy.birth.5, 
                                   life.expectancy.birth.9) %>%
                            rename(Pop_white = Pop_across_age,  
                                   le.birth.white1 = life.expectancy.birth.1,
                                   le.birth.white5 = life.expectancy.birth.5,
                                   le.birth.white9 = life.expectancy.birth.9)

BlackWhite <- merge(Black, White, by = c("State_factor", "Year_actual", "Sex_factor"))

BlackWhite <- BlackWhite %>% mutate(WBgap1 = le.birth.white1 - le.birth.black1,
                                    WBgap5 = le.birth.white5 - le.birth.black5,
                                    WBgap9 = le.birth.white9 - le.birth.black9)

rm(Black, White)
```

Decompose the gap by age. That is, for each year-state-sex combination, compare the life table for blacks to whites and decompose the gap in life expectancy at birth by age.

```{r create_paired_ids_dataset}
#paired.ids will contain the index names that are sent to the age decomposition function in the next code chunk.
#it pairs ids between strata for blacks and whites

dat.clean$StateYearSex <- interaction(dat.clean$State_factor, dat.clean$Year_factor, dat.clean$Sex_factor)

#first, create list of stratum.ids for each race-sex combination
for (rs in unique(levels(dat.clean$RaceSex))) {
  assign(paste0(gsub(" ", "", rs), ".stratum.ids"), 
         dat.clean %>% 
           ungroup() %>%
           filter(RaceSex == rs) %>%
           select(StateYearRaceSex, StateYearSex, Sex_factor, State_factor, Year_actual) %>%
           group_by(StateYearRaceSex, StateYearSex) %>%
           filter(row_number() == 1)
  )

}

#then, merge the black-white lists
Female.ids <- merge(WhiteFemale.stratum.ids, BlackFemale.stratum.ids[, -c(3:5)], 
                    by = "StateYearSex", suffixes = c("White", "Black"))
head(Female.ids)

Male.ids <- merge(WhiteMale.stratum.ids, BlackMale.stratum.ids[, -c(3:5)],
                  by = "StateYearSex", suffixes = c("White", "Black"))
head(Male.ids)

paired.ids <- rbind(Male.ids, Female.ids)

rm(WhiteFemale.stratum.ids, BlackFemale.stratum.ids, WhiteMale.stratum.ids, BlackMale.stratum.ids, Female.ids, Male.ids)
```

```{r calculate_age_decomp}
#runs in about a minute

list.age.decomp.tables1 <- list.age.decomp.tables5 <- list.age.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])

system.time(
  for(i in 1:dim(paired.ids)[1]) {
    white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
    black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])

    list.age.decomp.tables1[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
    list.age.decomp.tables5[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
    list.age.decomp.tables9[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]

    if(length(white.index) != 0 & length(black.index) != 0) {
      list.age.decomp.tables1[[i]]$decomp.table <- le_age_decomp(life.table1 = as.data.frame(list.life.tables1[[black.index]]$life.table), 
                                                                 name.lt1 = "Black", 
                                                                 life.table2 = as.data.frame(list.life.tables1[[white.index]]$life.table), 
                                                                 name.lt2 = "White", age.groups = "Age_factor")

      list.age.decomp.tables5[[i]]$decomp.table <- le_age_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]$life.table), 
                                                                 name.lt1 = "Black", 
                                                                 life.table2 = as.data.frame(list.life.tables5[[white.index]]$life.table), 
                                                                 name.lt2 = "White", age.groups = "Age_factor")

      list.age.decomp.tables9[[i]]$decomp.table <- le_age_decomp(life.table1 = as.data.frame(list.life.tables9[[black.index]]$life.table), 
                                                                 name.lt1 = "Black", 
                                                                 life.table2 = as.data.frame(list.life.tables9[[white.index]]$life.table), 
                                                                 name.lt2 = "White", age.groups = "Age_factor")
    } else {
      list.age.decomp.tables1[[i]]$decomp.table <- list.age.decomp.tables5[[i]]$decomp.table <- list.age.decomp.tables9[[i]]$decomp.table <-"Couldn't compare life tables"
    }
  }
)

```

```{r calculate_cod_decomp}
#stopped here - need to figure this out
#if the count of deaths == 0 for all causes of death within an age band then the COD proportion is NA
#this messes up the cod decomposition. What to do?
#example - no black people dying between the ages of 5-9 of anything. but have whites dying in that age group so their
#COD proportions are okay.

#takes 1.5 hours for all three
list.cod.decomp.tables1 <- list.cod.decomp.tables5 <- list.cod.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])

system.time(
  for(i in 1:dim(paired.ids)[1]) {
    white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
    black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])

    if(length(white.index) != 0 & length(black.index) != 0) {
      ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
        ungroup() %>%
        select(Age_factor, COD_factor, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
        rename(prop.white1 = cod_prop_death1, prop.white5 = cod_prop_death5, prop.white9 = cod_prop_death9)

      ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
        ungroup() %>%
        select(Age_factor, COD_factor, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
        rename(prop.black1 = cod_prop_death1, prop.black5 = cod_prop_death5, prop.black9 = cod_prop_death9)

      ct.both <- merge(ct.white, ct.black, by = c("Age_factor", "COD_factor")) 
      ct.both <- ct.both %>% arrange(COD_factor, Age_factor)

      if(any(is.na(ct.both$prop.black1)) == F & any(is.na(ct.both$prop.white1)) == F) { #ensures that there are no ages with 0 deaths - for these COD prop == 0 and can't compute

        list.cod.decomp.tables1[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables1[[black.index]]),
                                                              life.table2 = as.data.frame(list.life.tables1[[white.index]]), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables1[[i]]),
                                                              cod.table = ct.both %>% select(Age_factor, COD_factor, prop.white1, prop.black1), 
                                                              age.colname.cod.table = "Age_factor", 
                                                              COD.colname.cod.table = "COD_factor", 
                                                              prop1.colname.cod.table = "prop.black1", 
                                                              prop2.colname.cod.table = "prop.white1")

        list.cod.decomp.tables5[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]),
                                                              life.table2 = as.data.frame(list.life.tables5[[white.index]]), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables5[[i]]),
                                                              cod.table = ct.both %>% select(Age_factor, COD_factor, prop.white5, prop.black5), 
                                                              age.colname.cod.table = "Age_factor", 
                                                              COD.colname.cod.table = "COD_factor", 
                                                              prop1.colname.cod.table = "prop.black5", 
                                                              prop2.colname.cod.table = "prop.white5")

        list.cod.decomp.tables9[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]),
                                                              life.table2 = as.data.frame(list.life.tables5[[white.index]]), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables5[[i]]),
                                                              cod.table = ct.both %>% select(Age_factor, COD_factor, prop.white9, prop.black9), 
                                                              age.colname.cod.table = "Age_factor", 
                                                              COD.colname.cod.table = "COD_factor", 
                                                              prop1.colname.cod.table = "prop.black9", 
                                                              prop2.colname.cod.table = "prop.white9")

      } else { list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Encountered age group with no deaths"

      }
    } else {

      list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Couldn't compare life tables"
    }

    print(i)
  }
)
```

```{r  update_decomp_for_plotting}
#update the decomposition table for plotting purposes
for(i in 1:dim(paired.ids)[1]) {
    if(list.cod.decomp.tables[[i]][1] != "Couldn't compare life tables" & 
       list.cod.decomp.tables[[i]][1] != "Encountered age group with no deaths") {

       list.cod.decomp.tables[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables[[i]],
                                                            age.groups = "Ages", 
                                                            cause.of.death = "Cause.of.death", 
                                                            sign.var = "sign", decomp.var = "C_xi", 
                                                            decomp.var.prop = "C_xi_proportion")

       list.cod.decomp.tables5[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables5[[i]],
                                                            age.groups = "Ages", 
                                                            cause.of.death = "Cause.of.death", 
                                                            sign.var = "sign", decomp.var = "C_xi", 
                                                            decomp.var.prop = "C_xi_proportion")

       list.cod.decomp.tables9[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables9[[i]],
                                                            age.groups = "Ages", 
                                                            cause.of.death = "Cause.of.death", 
                                                            sign.var = "sign", decomp.var = "C_xi", 
                                                            decomp.var.prop = "C_xi_proportion") 
    } 
  print(i)
}  
```

```{r pretty_cod}
ggplotly(ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
           xlab("Contribution to life expectancy gap") +
           ylab("Age group\n") + theme_minimal() +
           geom_vline(xintercept = 0))

ggplotly(ggplot(list.cod.decomp.tables9[[1]], aes(y = Ages, x = start)) + 
           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
           xlab("Contribution to life expectancy gap") +
           ylab("Age group\n") + theme_minimal() +
           geom_vline(xintercept = 0))
```

### 2017-01-20 Code/1_Explore_Data.Rmd
cleaning up

@@
 #save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
-write_csv(dat.clean, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.Rdata", col_names = T)
-write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", col_names = T)
+write_csv(dat.clean, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.csv", col_names = T)
+write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.csv", col_names = T)
 

### 2017-01-20 Code/2_Run_LE_and_Decomp_Calcs.Rmd
cleaning up

@@
 ```{r load_data}
-dat.aggregated <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", header = T)
-dat.clean<- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", header = T)
+dat.aggregated <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.csv", header = T)
+dat.clean<- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.csv", header = T)
 ```
@@
 
-dat.clean$StateYearSex <- interaction(dat.clean$State_factor, dat.clean$Year_factor, dat.clean$Sex_factor)
-
 #first, create list of stratum.ids for each race-sex combination
@@
 ```
+
+```{r save_objects}
+write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated2.csv", col_names = T)
+write_csv(BlackWhite, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/BlackWhite.csv", col_names = T)
+
+save(paired.ids, non.zero.stratums,
+     list.age.decomp.tables1, list.age.decomp.tables5, list.age.decomp.tables9,
+     list.life.tables1, list.life.tables5, list.life.tables9,  
+     file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/calculated_data.Rdata")
+```
+
+#stopped here!
 
@@
 
-       list.cod.decomp.tables[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables[[i]],
+       list.cod.decomp.tables[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables1[[i]],
                                                             age.groups = "Ages", 

### 2017-01-20 Code/2_Run_LE_and_Decomp_Calcs.Rmd

@@
 ```{r load_data}
-dat.aggregated <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.csv", header = T)
-dat.clean<- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.csv", header = T)
+dat.aggregated <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", header = T)
+dat.clean<- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.Rdata", header = T)
 ```
@@
 
-#stopped here!
-
 ```{r calculate_cod_decomp}
@@
 
+    list.cod.decomp.tables1[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
+    list.cod.decomp.tables5[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
+    list.cod.decomp.tables9[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
+
     if(length(white.index) != 0 & length(black.index) != 0) {
@@
 
-      if(any(is.na(ct.both$prop.black1)) == F & any(is.na(ct.both$prop.white1)) == F) { #ensures that there are no ages with 0 deaths - for these COD prop == 0 and can't compute
+      if(!any(is.na(ct.both$prop.black1)) & !any(is.na(ct.both$prop.white1))) { #ensures that there are no ages with 0 deaths - for these COD prop == 0 and can't compute
 
-        list.cod.decomp.tables1[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables1[[black.index]]),
-                                                              life.table2 = as.data.frame(list.life.tables1[[white.index]]), 
-                                                              decomp.table = as.data.frame(list.age.decomp.tables1[[i]]),
+        list.cod.decomp.tables1[[i]]$decomp.table <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables1[[black.index]]$life.table),
+                                                              life.table2 = as.data.frame(list.life.tables1[[white.index]]$life.table), 
+                                                              decomp.table = as.data.frame(list.age.decomp.tables1[[i]]$decomp.table),
                                                               cod.table = ct.both %>% select(Age_factor, COD_factor, prop.white1, prop.black1), 
@@
 
-        list.cod.decomp.tables5[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]),
-                                                              life.table2 = as.data.frame(list.life.tables5[[white.index]]), 
-                                                              decomp.table = as.data.frame(list.age.decomp.tables5[[i]]),
+        list.cod.decomp.tables5[[i]]$decomp.table <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]$life.table),
+                                                              life.table2 = as.data.frame(list.life.tables5[[white.index]]$life.table), 
+                                                              decomp.table = as.data.frame(list.age.decomp.tables5[[i]]$decomp.table),
                                                               cod.table = ct.both %>% select(Age_factor, COD_factor, prop.white5, prop.black5), 
@@
 
-        list.cod.decomp.tables9[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]),
-                                                              life.table2 = as.data.frame(list.life.tables5[[white.index]]), 
-                                                              decomp.table = as.data.frame(list.age.decomp.tables5[[i]]),
+        list.cod.decomp.tables9[[i]]$decomp.table <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]$life.table),
+                                                              life.table2 = as.data.frame(list.life.tables5[[white.index]]$life.table), 
+                                                              decomp.table = as.data.frame(list.age.decomp.tables5[[i]]$decomp.table),
                                                               cod.table = ct.both %>% select(Age_factor, COD_factor, prop.white9, prop.black9), 
@@
 
-      } else { list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Encountered age group with no deaths"
+      } else { list.cod.decomp.tables1[[i]]$decomp.table <- list.cod.decomp.tables5[[i]]$decomp.table <- list.cod.decomp.tables9[[i]]$decomp.table <- "Encountered age group with no deaths"
 
@@
 
-      list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Couldn't compare life tables"
+      list.cod.decomp.tables1[[i]]$decomp.table <- list.cod.decomp.tables5[[i]]$decomp.table <- list.cod.decomp.tables9[[i]]$decomp.table <- "Couldn't compare life tables"
     }

### 2017-01-20 Code/Run_model_on_SH_data/Prep SH HA data for our analysis.Rmd
small change

@@
 test.le.fx <- function(data, state, sex, race, year) {
-  test.data <- data[data["State2"] == state & data["Sex2"] == sex & data["Race2"] == race & data["Year3"] == year, ]
+  test.data <- data[data["State2"] == state & 
+                      data["Sex2"] == sex & 
+                      data["Race2"] == race & 
+                      data["Year3"] == year, ]
+
   lt <- life.table(data = test.data, num.ages.in.group = "nx",
@@
 
-test.le.fx(dtotal, "Alabama", "Male", "White", 1990)
 test.le.fx(dtotal, "Alabama", "Male", "White", 2009)
 
+test.le.fx(dtotal, "Alabama", "Male", "White", 1990)
 test.le.fx(dtotal, "Alabama", "Male", "Black", 1990)

### 2017-01-20 Code/1_Explore_Data.Rmd
change var format

@@
 names(dat) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
+names(dat) <- c("state.n", "age.n", "sex.n", "race.n", "COD.n", "year.0", "crude.rate", "deaths", "population") 
 # names and order of variables in the dataset are noted in the .dic dictionary file found in the same dir as the dataset.
@@
 dat <- dat %>%
-  mutate(State_factor = factor(State, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
+  mutate(state = factor(state.n, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
                                                              "Colorado", "Connecticut", "Delaware", "Washington DC",
@@
                                                              "Washington", "West Virginia", "Wisconsin", "Wyoming")), 
-         Age_factor = factor(Age, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
+         age = factor(age.n, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                          "15-19 years", "20-24 years", "25-29 years", "30-34 years",
@@
                                                          "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
-         Age_minbin = 0*(Age_factor == "<1 year") + 1*(Age_factor == "1-4 years") + 5*(Age_factor == "5-9 years") +
-                10*(Age_factor == "10-14 years") + 15*(Age_factor == "15-19 years") + 20*(Age_factor == "20-24 years") + 
-                25*(Age_factor == "25-29 years") + 30*(Age_factor == "30-34 years") + 35*(Age_factor == "35-39 years") +
-                40*(Age_factor == "40-44 years") + 45*(Age_factor == "45-49 years") + 50*(Age_factor == "50-54 years") + 
-                55*(Age_factor == "55-59 years") + 60*(Age_factor == "60-64 years") + 65*(Age_factor == "65-69 years") +   
-                70*(Age_factor == "70-74 years") + 75*(Age_factor == "75-79 years") + 80*(Age_factor == "80-84 years") + 
-                85*(Age_factor == "85+ years"),
-         Age_bin_size = 1*(Age_factor == "<1 year") + 4*(Age_factor == "1-4 years") + 5*(Age_factor != "<1 year" & Age_factor != "1-4 years"),
-         Sex_factor = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
-         Race_factor = factor(Race, levels = c(0, 1), labels = c("White", "Black")),
-         COD_factor = factor(COD, levels = c(0:5), labels = c("Cardiovascular", "Cancers", "Communicable", 
+         age_minbin = 0*(age == "<1 year") + 1*(age == "1-4 years") + 5*(age == "5-9 years") +
+                10*(age == "10-14 years") + 15*(age == "15-19 years") + 20*(age == "20-24 years") + 
+                25*(age == "25-29 years") + 30*(age == "30-34 years") + 35*(age == "35-39 years") +
+                40*(age == "40-44 years") + 45*(age == "45-49 years") + 50*(age == "50-54 years") + 
+                55*(age == "55-59 years") + 60*(age == "60-64 years") + 65*(age == "65-69 years") +   
+                70*(age == "70-74 years") + 75*(age == "75-79 years") + 80*(age == "80-84 years") + 
+                85*(age == "85+ years"),
+         age_bin_size = 1*(age == "<1 year") + 4*(age == "1-4 years") + 5*(age != "<1 year" & age != "1-4 years"),
+         sex = factor(sex.n, levels = c(0, 1), labels = c("Male", "Female")),
+         race = factor(race.n, levels = c(0, 1), labels = c("White", "Black")),
+         COD = factor(COD.n, levels = c(0:5), labels = c("Cardiovascular", "Cancers", "Communicable", 
                                                         "Non-communicable", "Injuries", "All other causes")), 
-         Year_factor = factor(Year, levels = c(0:44), labels = as.character(1969:2013)),
-         Year_actual = as.numeric(as.character(Year_factor)),
-         RaceSex = interaction(Race_factor, Sex_factor, sep = " "),
-         StateYearSex = interaction(State_factor, Year_factor, Sex_factor),
-         StateYearRaceSex = interaction(State_factor, Year_factor, Race_factor, Sex_factor))
+         year = factor(year.0, levels = c(0:44), labels = as.character(1969:2013)),
+         year.n = as.numeric(as.character(year)),
+         race.sex = interaction(race, sex, sep = " "),
+         state.year.sex = interaction(state, year, sex),
+         state.year.race.sex = interaction(state, year, race, sex))
 
@@
 #there is an unknown age group for each combo of state-year-race-sex-age:
-table(dat$Age_factor)
+table(dat$age)
 
 #Count of death is always equal to 0 for unknown age groups
-table(dat$Age_factor, dat$Count==0)
+table(dat$age, dat$deaths==0)
 
 #remove these data from the data we analyze:
-dat.clean <- dat[dat$Age_factor != "Unknown", ]
+dat.clean <- dat[dat$age != "Unknown", ]
 ```
@@
 ```{r recode_pops_of_zero}
-table(dat.clean$Population == 0, useNA = "always")
+table(dat.clean$population == 0, useNA = "always")
 #654 cells have Populations of 0.
 
-table(dat.clean$Count[dat.clean$Population == 0] == 0, useNA = "always")
+table(dat.clean$deaths[dat.clean$population == 0] == 0, useNA = "always")
 #Of these, all have Counts of 0, except for 26 with counts of NA
@@
 #replace the NA Counts with 0 for the 26 cells with Populations of 0.
-dat.clean$Count[is.na(dat.clean$Count) == T & dat.clean$Population == 0] <- 0
-summary(dat.clean$Count[dat.clean$Population == 0])
+dat.clean$deaths[is.na(dat.clean$deaths) == T & dat.clean$population == 0] <- 0
+summary(dat.clean$deaths[dat.clean$population == 0])
 
-summary(dat.clean$Crude_Rate[dat.clean$Population == 0])
-#whenever the population is 0, the Crude_Rate is equal to NA.
+summary(dat.clean$crude.rate[dat.clean$population == 0])
+#whenever the population is 0, the crude.rate is equal to NA.
 #need to think: should these be equal to 0, or NA? How will it affect the calculations?
@@
 na.pops <- dat.clean %>% 
-  select(State_factor, Population) %>% 
-  filter(Population == 0)
+  select(state, population) %>% 
+  filter(population == 0)
 
-unique(na.pops$State_factor)
+unique(na.pops$state)
 
-pops.le.20 <- dat.clean %>% select(State_factor, Population) %>% filter(Population <= 20)
-unique(pops.le.20$State_factor)
+pops.le.20 <- dat.clean %>% select(state, population) %>% filter(population <= 20)
+unique(pops.le.20$state)
 ```
@@
 ```{r fill_in_missing_counts}
-dat.clean <- dat.clean %>% mutate(Count_md1 = ifelse(is.na(Count)==T, pmin(1, Population), Count),
-                                  Count_md5 = ifelse(is.na(Count)==T, pmin(5, Population), Count),
-                                  Count_md9 = ifelse(is.na(Count)==T, pmin(9, Population), Count)
+dat.clean <- dat.clean %>% mutate(deaths_md1 = ifelse(is.na(deaths)==T, pmin(1, population), deaths),
+                                  deaths_md5 = ifelse(is.na(deaths)==T, pmin(5, population), deaths),
+                                  deaths_md9 = ifelse(is.na(deaths)==T, pmin(9, population), deaths)
                                   )
@@
 #check recoding
-dat.clean$Count_grouped <- cut(dat.clean$Count, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
-dat.clean$Count1_grouped <- cut(dat.clean$Count_md1, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
-dat.clean$Count5_grouped <- cut(dat.clean$Count_md5, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
-dat.clean$Count9_grouped <- cut(dat.clean$Count_md9, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
+dat.clean$deaths_grouped <- cut(dat.clean$deaths, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
+dat.clean$deaths1_grouped <- cut(dat.clean$deaths_md1, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
+dat.clean$deaths5_grouped <- cut(dat.clean$deaths_md5, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
+dat.clean$deaths9_grouped <- cut(dat.clean$deaths_md9, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
 
-table(dat.clean$Count_grouped, dat.clean$Count1_grouped, useNA = "always")
-table(dat.clean$Count_grouped, dat.clean$Count5_grouped, useNA = "always")
-table(dat.clean$Count_grouped, dat.clean$Count9_grouped, useNA = "always")
+table(dat.clean$deaths_grouped, dat.clean$deaths1_grouped, useNA = "always")
+table(dat.clean$deaths_grouped, dat.clean$deaths5_grouped, useNA = "always")
+table(dat.clean$deaths_grouped, dat.clean$deaths9_grouped, useNA = "always")
 
-dat.clean <- dat.clean[ , !names(dat.clean) %in% c("Count_grouped", "Count1_grouped", "Count5_grouped", "Count9_grouped")]
+dat.clean <- dat.clean[ , !names(dat.clean) %in% c("deaths_grouped", "deaths1_grouped", "deaths5_grouped", "deaths9_grouped")]
 ```
@@
 
-dat.clean <- dat.clean %>% group_by(State_factor, RaceSex, Year_actual, Age_minbin) %>%
-                           arrange(State_factor, RaceSex, Year_actual, Age_minbin) %>%
-                           mutate(total_deaths1 = sum(Count_md1),
-                                  total_deaths5 = sum(Count_md5),
-                                  total_deaths9 = sum(Count_md9),
-                                  cod_prop_death1 = Count_md1/total_deaths1, 
-                                  cod_prop_death5 = Count_md5/total_deaths5,
-                                  cod_prop_death9 = Count_md9/total_deaths9)
+dat.clean <- dat.clean %>% group_by(state, race.sex, year.n, age_minbin) %>%
+                           arrange(state, race.sex, year.n, age_minbin) %>%
+                           mutate(total_deaths1 = sum(deaths_md1),
+                                  total_deaths5 = sum(deaths_md5),
+                                  total_deaths9 = sum(deaths_md9),
+                                  cod_prop_death1 = deaths_md1/total_deaths1, 
+                                  cod_prop_death5 = deaths_md5/total_deaths5,
+                                  cod_prop_death9 = deaths_md9/total_deaths9)
 
@@
 ```{r aggregate_over_COD}
-dat.aggregated <- dat.clean %>% group_by(State_factor, Year_factor, RaceSex, Age_factor) %>%
-                                arrange(State_factor, RaceSex, Year_factor, Age_factor) %>%
-                                summarise(Population = first(Population), 
-                                          Race_factor = first(Race_factor), 
-                                          Sex_factor = first(Sex_factor), 
-                                          StateYearRaceSex = first(StateYearRaceSex),
-                                          Age_bin_size = first(Age_bin_size),
-                                          Age_minbin = first(Age_minbin),
-                                          Year_actual = first(Year_actual),
+dat.aggregated <- dat.clean %>% group_by(state, year, race.sex, age) %>%
+                                arrange(state, race.sex, year, age) %>%
+                                summarise(population = first(population), 
+                                          race = first(race), 
+                                          sex = first(sex), 
+                                          state.year.race.sex = first(state.year.race.sex),
+                                          age_bin_size = first(age_bin_size),
+                                          age_minbin = first(age_minbin),
+                                          year.n = first(year.n),
                                           total_deaths1 = first(total_deaths1),
@@
 #return "stratum ids" with any age-specific population counts of 0.
-non.zero.stratums <- setdiff(levels(dat.aggregated$StateYearRaceSex), 
-                             unique(dat.aggregated$StateYearRaceSex[dat.aggregated$Population == 0]))
+non.zero.stratums <- setdiff(levels(dat.aggregated$state.year.race.sex), 
+                             unique(dat.aggregated$state.year.race.sex[dat.aggregated$population == 0]))
 ```
@@
 dat3 <- dat.aggregated %>% ungroup() %>% 
-                           group_by(StateYearRaceSex) %>%
-                           summarise(Pop_across_age = sum(Population))
+                           group_by(state.year.race.sex) %>%
+                           summarise(pop_across_age = sum(population))
 
-dat.aggregated <- merge(dat.aggregated, dat3, by = "StateYearRaceSex")
+dat.aggregated <- merge(dat.aggregated, dat3, by = "state.year.race.sex")
 
@@
 
-dat.aggregated <- dat.aggregated %>% arrange(State_factor, Year_actual, Race_factor, Sex_factor, Age_minbin)
+dat.aggregated <- dat.aggregated %>% arrange(state, year.n, race, sex, age_minbin)
 ```
@@
 #save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
-write_csv(dat.clean, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.csv", col_names = T)
-write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.csv", col_names = T)
-
+write_csv(dat.clean, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.Rdata", col_names = T)
+write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", col_names = T)
 ```

### 2017-01-20 Code/1_Explore_Data.Rmd
change vars

@@
 #save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
-write_csv(dat.clean, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.Rdata", col_names = T)
-write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", col_names = T)
+write_csv(dat.clean, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.csv", col_names = T)
+write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.csv", col_names = T)
 ```

### 2017-01-20 Code/2_Run_LE_and_Decomp_Calcs.Rmd

@@
 library(tidyverse)
-library(plotly)
 options(tibble.width = Inf) #display all columns of tibbles
@@
 ```{r load_data}
-dat.aggregated <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", header = T)
-dat.clean<- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.Rdata", header = T)
+dat.aggregated <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.csv", header = T)
+dat.clean<- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.csv", header = T)
 ```
@@
 ```{r calculate_life_expectancy}
-non.zero.stratums <- setdiff(levels(dat.aggregated$StateYearRaceSex), 
-                             unique(dat.aggregated$StateYearRaceSex[dat.aggregated$Population == 0]))
+non.zero.stratums <- setdiff(levels(dat.aggregated$state.year.race.sex), 
+                             unique(dat.aggregated$state.year.race.sex[dat.aggregated$population == 0]))
 
@@
 
-    list.life.tables1[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, StateYearRaceSex == stratum.id) %>% arrange(Age_factor),
-                                                 num.ages.in.group = "Age_bin_size",
+    list.life.tables1[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, state.year.race.sex == stratum.id) %>% arrange(age),
+                                                 num.ages.in.group = "age_bin_size",
                                                  death.counts = "total_deaths1",
-                                                 population.counts = "Population")
+                                                 population.counts = "population")
 
-    list.life.tables5[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, StateYearRaceSex == stratum.id) %>% arrange(Age_factor),
-                                                  num.ages.in.group = "Age_bin_size",
+    list.life.tables5[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, state.year.race.sex == stratum.id) %>% arrange(age),
+                                                  num.ages.in.group = "age_bin_size",
                                                   death.counts = "total_deaths5",
-                                                  population.counts = "Population")
+                                                  population.counts = "population")
 
-    list.life.tables9[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, StateYearRaceSex == stratum.id) %>% arrange(Age_factor),
-                                                  num.ages.in.group = "Age_bin_size",
+    list.life.tables9[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, state.year.race.sex == stratum.id) %>% arrange(age),
+                                                  num.ages.in.group = "age_bin_size",
                                                   death.counts = "total_deaths9",
-                                                  population.counts = "Population")
+                                                  population.counts = "population")
     list.index <- list.index + 1
@@
 les <- tibble(le1, le5, le9, sid)
-names(les) = c("life.expectancy.birth.1", "life.expectancy.birth.5", "life.expectancy.birth.9", "StateYearRaceSex")
+names(les) = c("life.expectancy.birth.1", "life.expectancy.birth.5", "life.expectancy.birth.9", "state.year.race.sex")
 
-dat.aggregated <- merge(dat.aggregated, les, by = "StateYearRaceSex", all.x =T)
+dat.aggregated <- merge(dat.aggregated, les, by = "state.year.race.sex", all.x =T)
 
@@
 ```{r densityplot_life_expectancy}
-ggplot(dat.aggregated %>% dplyr::filter(Age_factor == "<1 year"), aes(x = life.expectancy.birth.1)) + 
+ggplot(dat.aggregated %>% dplyr::filter(age == "<1 year"), aes(x = life.expectancy.birth.1)) + 
   geom_density(aes(fill="Impute 1"), alpha = 0.3) +
@@
   geom_density(aes(x = life.expectancy.birth.9, fill="Impute 9"), alpha = 0.3) +
-  facet_grid(Sex_factor ~ Race_factor)
+  facet_grid(sex ~ race)
 ```
@@
 
-Black <- dat.aggregated %>% filter(Race_factor == "Black" & Age_minbin == 0) %>% 
-                            select(State_factor, Year_actual, Sex_factor, Pop_across_age, 
+Black <- dat.aggregated %>% filter(race == "Black" & age_minbin == 0) %>% 
+                            select(state, year.n, sex, pop_across_age, 
                                    life.expectancy.birth.1, life.expectancy.birth.5, 
                                    life.expectancy.birth.9) %>%
-                            rename(Pop_black = Pop_across_age,  
+                            rename(pop_black = pop_across_age,  
                                    le.birth.black1 = life.expectancy.birth.1, 
@@
 
-White <- dat.aggregated %>% filter(Race_factor == "White" & Age_minbin == 0) %>% 
-                            select(State_factor, Year_actual, Sex_factor, Pop_across_age, 
+White <- dat.aggregated %>% filter(race == "White" & age_minbin == 0) %>% 
+                            select(state, year.n, sex, pop_across_age, 
                                    life.expectancy.birth.1, life.expectancy.birth.5, 
                                    life.expectancy.birth.9) %>%
-                            rename(Pop_white = Pop_across_age,  
+                            rename(pop_white = pop_across_age,  
                                    le.birth.white1 = life.expectancy.birth.1,
@@
 
-BlackWhite <- merge(Black, White, by = c("State_factor", "Year_actual", "Sex_factor"))
+BlackWhite <- merge(Black, White, by = c("state", "year.n", "sex"))
 
@@
 #first, create list of stratum.ids for each race-sex combination
-for (rs in unique(levels(dat.clean$RaceSex))) {
+for (rs in unique(levels(dat.clean$race.sex))) {
   assign(paste0(gsub(" ", "", rs), ".stratum.ids"), 
@@
            ungroup() %>%
-           filter(RaceSex == rs) %>%
-           select(StateYearRaceSex, StateYearSex, Sex_factor, State_factor, Year_actual) %>%
-           group_by(StateYearRaceSex, StateYearSex) %>%
+           filter(race.sex == rs) %>%
+           select(state.year.race.sex, state.year.sex, sex, state, year.n) %>%
+           group_by(state.year.race.sex, state.year.sex) %>%
            filter(row_number() == 1)
@@
 Female.ids <- merge(WhiteFemale.stratum.ids, BlackFemale.stratum.ids[, -c(3:5)], 
-                    by = "StateYearSex", suffixes = c("White", "Black"))
+                    by = "state.year.sex", suffixes = c(".white", ".black"))
 head(Female.ids)
@@
 Male.ids <- merge(WhiteMale.stratum.ids, BlackMale.stratum.ids[, -c(3:5)],
-                  by = "StateYearSex", suffixes = c("White", "Black"))
+                  by = "state.year.sex", suffixes = c(".white", ".black"))
 head(Male.ids)
@@
   for(i in 1:dim(paired.ids)[1]) {
-    white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
-    black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])
+    white.index <- which(non.zero.stratums == paired.ids[i, "state.year.race.sex.white"])
+    black.index <- which(non.zero.stratums == paired.ids[i, "state.year.race.sex.black"])
 
-    list.age.decomp.tables1[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
-    list.age.decomp.tables5[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
-    list.age.decomp.tables9[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
+    list.age.decomp.tables1[[i]]$state.year.sex <- paired.ids[i, "state.year.sex"]
+    list.age.decomp.tables5[[i]]$state.year.sex <- paired.ids[i, "state.year.sex"]
+    list.age.decomp.tables9[[i]]$state.year.sex <- paired.ids[i, "state.year.sex"]
 
@@
                                                                  life.table2 = as.data.frame(list.life.tables1[[white.index]]$life.table), 
-                                                                 name.lt2 = "White", age.groups = "Age_factor")
+                                                                 name.lt2 = "White", age.groups = "age")
 
@@
                                                                  life.table2 = as.data.frame(list.life.tables5[[white.index]]$life.table), 
-                                                                 name.lt2 = "White", age.groups = "Age_factor")
+                                                                 name.lt2 = "White", age.groups = "age")
 
@@
                                                                  life.table2 = as.data.frame(list.life.tables9[[white.index]]$life.table), 
-                                                                 name.lt2 = "White", age.groups = "Age_factor")
+                                                                 name.lt2 = "White", age.groups = "age")
     } else {
@@
   for(i in 1:dim(paired.ids)[1]) {
-    white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
-    black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])
+    white.index <- which(non.zero.stratums == paired.ids[i, "state.year.race.sex.white"])
+    black.index <- which(non.zero.stratums == paired.ids[i, "state.year.race.sex.black"])
 
-    list.cod.decomp.tables1[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
-    list.cod.decomp.tables5[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
-    list.cod.decomp.tables9[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
+    list.cod.decomp.tables1[[i]]$state.year.sex <- paired.ids[i, "state.year.sex"]
+    list.cod.decomp.tables5[[i]]$state.year.sex <- paired.ids[i, "state.year.sex"]
+    list.cod.decomp.tables9[[i]]$state.year.sex <- paired.ids[i, "state.year.sex"]
 
     if(length(white.index) != 0 & length(black.index) != 0) {
-      ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
+      ct.white <- subset(dat.clean, state.year.race.sex == paired.ids[i, "state.year.race.sex.white"]) %>% 
         ungroup() %>%
-        select(Age_factor, COD_factor, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
+        select(age, COD, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
         rename(prop.white1 = cod_prop_death1, prop.white5 = cod_prop_death5, prop.white9 = cod_prop_death9)
 
-      ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
+      ct.black <- subset(dat.clean, state.year.race.sex == paired.ids[i, "state.year.race.sex.black"]) %>% 
         ungroup() %>%
-        select(Age_factor, COD_factor, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
+        select(age, COD, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
         rename(prop.black1 = cod_prop_death1, prop.black5 = cod_prop_death5, prop.black9 = cod_prop_death9)
 
-      ct.both <- merge(ct.white, ct.black, by = c("Age_factor", "COD_factor")) 
-      ct.both <- ct.both %>% arrange(COD_factor, Age_factor)
+      ct.both <- merge(ct.white, ct.black, by = c("age", "COD")) 
+      ct.both <- ct.both %>% arrange(COD, age)
 
@@
                                                               decomp.table = as.data.frame(list.age.decomp.tables1[[i]]$decomp.table),
-                                                              cod.table = ct.both %>% select(Age_factor, COD_factor, prop.white1, prop.black1), 
-                                                              age.colname.cod.table = "Age_factor", 
-                                                              COD.colname.cod.table = "COD_factor", 
+                                                              cod.table = ct.both %>% select(age, COD, prop.white1, prop.black1), 
+                                                              age.colname.cod.table = "age", 
+                                                              COD.colname.cod.table = "COD", 
                                                               prop1.colname.cod.table = "prop.black1", 
@@
                                                               decomp.table = as.data.frame(list.age.decomp.tables5[[i]]$decomp.table),
-                                                              cod.table = ct.both %>% select(Age_factor, COD_factor, prop.white5, prop.black5), 
-                                                              age.colname.cod.table = "Age_factor", 
-                                                              COD.colname.cod.table = "COD_factor", 
+                                                              cod.table = ct.both %>% select(age, COD, prop.white5, prop.black5), 
+                                                              age.colname.cod.table = "age", 
+                                                              COD.colname.cod.table = "COD", 
                                                               prop1.colname.cod.table = "prop.black5", 
@@
                                                               decomp.table = as.data.frame(list.age.decomp.tables5[[i]]$decomp.table),
-                                                              cod.table = ct.both %>% select(Age_factor, COD_factor, prop.white9, prop.black9), 
+                                                              cod.table = ct.both %>% select(age, COD, prop.white9, prop.black9), 
                                                               age.colname.cod.table = "Age_factor", 
@@
 ```{r pretty_cod}
-ggplotly(ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
+ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
   geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
@@
   ylab("Age group\n") + theme_minimal() +
-           geom_vline(xintercept = 0))
+  geom_vline(xintercept = 0)
 
-ggplotly(ggplot(list.cod.decomp.tables9[[1]], aes(y = Ages, x = start)) + 
+ggplot(list.cod.decomp.tables9[[1]], aes(y = Ages, x = start)) + 
   geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
@@
   ylab("Age group\n") + theme_minimal() +
-           geom_vline(xintercept = 0))
+  geom_vline(xintercept = 0)
 ```

### 2017-01-20 Code/2_smoothing_models/bayes_smoothing_src.R
cleaning up modelling scripts

@@ -15,7 +15,10 @@ library(grid)
 library(gridExtra)
 
 
-
+ 
 source('~/BlackWhiteMortalityGap/Code/bayesian_smoothing/bayes_smoothing_functions.R') 
 #source('life_expectancy_functions.R') 
+
+load('~/SH_HA_data.Rdata')
+
 

### 2017-01-22 Code/3_Explore_Missingness.Rmd

@@
   dat %>%
-  group_by(State2, Race2, Age2) %>%
-  summarise(prop.missing = mean(is.na(Count)))
+  group_by(state, race, age) %>%
+  summarise(prop.missing = mean(is.na(deaths)))
 

### 2017-01-24 Code/3_validation/SH_examine_comparisons.Rmd
save results and clean up code

---
title: "Compare findings"
output: github_document
---

```{r load_libraries}
library(dplyr)
library(stringr)
library(ggplot2)
```

```{r read_data}
le.table <- read.csv("~/BlackWhiteMortalityGap/Code/3_validation/le_table.csv", header = T)
SH.dat <- read.csv("~/SH_HA_supp_app_data.csv", header = T)
```

```{r tidy_data}
SH.dat <- SH.dat %>% rename(state = State, 
                            le.black.original = le.black,
                            le.white.original = le.white,
                            le.gap.original = le.gap, 
                            le.gap.ci.original = le.gap.ci)

compare.le <- merge(le.table, SH.dat, by = c("state", "sex", "year"))

test <- strsplit(stringr::str_sub(compare.le$le.gap.ci.original, 2, -2), ",")
test2 <- matrix(unlist(test), ncol=2, byrow=TRUE)

compare.le$le.gap.lcl.original <- as.numeric(test2[ , 1])
compare.le$le.gap.ucl.original <- as.numeric(test2[ , 2])
rm(test, test2)

compare.le <- compare.le %>% mutate(le.gap.width.original = le.gap.ucl.original - le.gap.lcl.original,
                                    le.gap.width = le.gap.ucl - le.gap.lcl,
                                    small.black.state = state %in% c("Hawaii", "Alaska", "Maine", 
                                                                     "New Hampshire","Utah", "New Mexico")
                                    )

write.csv(compare.le, "~/BlackWhiteMortalityGap/Code/3_validation/compare_LE.csv")
```

**Figure 1: Comparison of life expectancy between the original model and our model**

```{r}
ggplot(subset(compare.le, state!= "Wyoming" & year != "difference"), 
       aes(x = le.white.original, y = le.white)) +
  geom_point(aes(x = le.black.original, y = le.black, col = "blacks")) + 
  geom_point(aes(col = "whites")) +
  facet_grid(year~sex ) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Life expectancy (years)") +
  xlab("Original life expectancy (years)") +
  geom_text(data = subset(compare.le, small.black.state == T & year != "difference"),
            aes(x = le.black.original, y = le.black, label = state), size = 2.5)
```

**Figure 2: Comparison of life expectancy change over time (1990 vs. 2009) between the original model and our model**

```{r}
ggplot(subset(compare.le, state!= "Wyoming" & year == "difference"), 
       aes(x = le.white.original, y = le.white)) +
  geom_point(aes(x = le.black.original, y = le.black, col = "blacks")) + 
  geom_point(aes(col = "whites")) +
  facet_grid(~sex ) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Life expectancy change (years)") +
  xlab("Original life expectancy change (years)") +
  geom_text(data = subset(compare.le, small.black.state == T & year == "difference"),
            aes(x = le.black.original, y = le.black, label = state), size = 2.5)

```

**Figure 3: Comparison of life expectancy gap between the original model and our model**

```{r}
ggplot(subset(compare.le, state!= "Wyoming" & year != "difference"), 
       aes(x = le.gap.original, y = le.gap)) +
  geom_point() +
  facet_grid(year~sex ) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Life expectancy gap (years)") +
  xlab("Original life expectancy gap (years)") +
  geom_text(data = subset(compare.le, small.black.state == T & year != "difference"),
            aes(x = le.gap.original, y = le.gap, label = state), size = 2.5)
```

**Figure 4: Comparison of change over time (1990 vs. 2009) in the life expectancy gap between the original model and our model**

```{r}
ggplot(subset(compare.le, state!= "Wyoming" & year == "difference"), 
       aes(x = le.gap.original, y = le.gap)) +
  geom_point() +
  facet_grid(.~sex ) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Change in life expectancy gap (years)") +
  xlab("Original change in life expectancy gap (years)") +
  geom_text(data = subset(compare.le, small.black.state == T & year == "difference"),
            aes(x = le.gap.original, y = le.gap, label = state), size = 2.5)
```

**Figure 5: Comparison of CI width for LE gap between the original model and our model**

```{r}
ggplot(subset(compare.le, state!= "Wyoming" & year != "difference"), 
       aes(x = le.gap.width.original, y = le.gap.width)) +
  geom_point() +
  facet_grid(year~sex ) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Width of CI for LE gap") +
  xlab("Original width of CI for LE gap") +
  geom_text(data = subset(compare.le, small.black.state == T & year != "difference"),
            aes(x = le.gap.width.original, y = le.gap.width, label = state), size = 2.5)
```

**Figure 6: Comparison of the change over time (1990 vs. 2009) in the CI width for LE gap between the original model and our model**

```{r}
ggplot(subset(compare.le, state!= "Wyoming" & year == "difference"), 
       aes(x = le.gap.width.original, y = le.gap.width)) +
  geom_point() +
  facet_grid(.~sex ) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Width of CI for LE gap") +
  xlab("Original width of CI for LE gap") +
  geom_text(data = subset(compare.le, small.black.state == T & year == "difference"),
            aes(x = le.gap.width.original, y = le.gap.width, label = state), size = 2.5)
```

### 2017-01-27 Code/3_validation/run-models.Rmd
alabama males results!

---
title: "Run models"
output: github_document
---

```{r load_libraries}
library(Rcpp)
library(dplyr)
library(purrr)
library(ggplot2)
library(stringr)
library(rjags)
library(R2jags) 
library(devtools)
library(jagstools)
library(reshape2)
library(parallel)
library(gtools)
library(bayesboot)
library(coda)
library(grid)
library(gridExtra)
```

```{r load_user_written_functions}
source("~/BlackWhiteMortalityGap/Code/2_smoothing_models/bayes_smoothing_functions.R")
source('~/BlackWhiteMortalityGap/Code/life_expectancy_functions.R') 
```

```{r load_tidied_data}
dat.clean = read.csv('~/dat_clean.csv')
```

```{r fit_models_alabama_males}
#run time: 15 minutes
#fit the bayesian models and put the 1000 posterior samples into a data frame
r.White <- run_analysis(dataset = dat.clean, state1 = "Alabama", sex1 = "Male", race1 = "White")
r.White$race <- "White"

r.Black <- run_analysis(dataset = dat.clean, state1 = "Alabama", sex1 = "Male", race1 = "Black")
r.Black$race <- "Black"
```

```{r prepare_data_for_le_functions}
#run time: about 2 minutes
r.All <- rbind(r.White, r.Black)
rm(r.White, r.Black)

r.All$smoothed_deaths <- r.All$smoothed_rate*r.All$population

r.All <- r.All %>% group_by(post.samp, state, race.sex, year.n, age_minbin) %>%
                           arrange(post.samp, state, race.sex, year.n, age_minbin) %>%
                           mutate(total_smoothed_deaths = sum(smoothed_deaths),
                                  cod_prop_smoothed_deaths = smoothed_deaths/total_smoothed_deaths)
```

```{r run_calculations}
#run time ~40 mins
#run the life table and decomposition calculations for each of the 1000 posterior samples

n <- length(levels(droplevels(r.All$state.year.sex)))

le.calculations = vector(mode = "list", length = n)
counter = 1
system.time(
  for (stratum_i in levels(droplevels(r.All$state.year.sex))){

    data_sub <- subset(r.All, state.year.sex == stratum_i)

    le.calculations[[counter]]$stratum.id <- stratum_i
    le.calculations[[counter]]$calcs <- by(data = data_sub,
                                           INDICES = list(data_sub$post.samp),
                                           FUN = function(x) life_expectancy_and_gap(data = x))

    counter <- counter + 1

    print(stratum_i)
    rm(data_sub)
  }
)
rm(counter)

```

```{r extract_life_expectancy_results}
BlackWhite <- data.frame(stratum.id = character(),
                         LE_white_lcl = numeric(), LE_white = numeric(), LE_white_ucl = numeric(),
                         LE_black_lcl = numeric(), LE_black = numeric(), LE_black_ucl = numeric(),
                         LE_wbgap_lcl = numeric(), LE_wbgap = numeric(), LE_wbgap_ucl = numeric(),
                         stringsAsFactors = F)

age_minbin <- le.calculations[[1]]$calcs[[1]]$age.decomp$Ages
cods <- le.calculations[[1]]$calcs[[1]]$cod.marginal$Cause.of.death

age.decomp.estimates <- data.frame(stratum.id = rep(levels(droplevels(r.All$state.year.sex)), 
                                                    each = length(age_minbin)),
                                   age_minbin = numeric(19*n),
                                   age_cont_yrs_lcl = numeric(19*n), 
                                   age_cont_yrs = numeric(19*n), 
                                   age_cont_yrs_ucl = numeric(19*n),
                                   age_cont_prop_lcl = numeric(19*n), 
                                   age_cont_prop = numeric(19*n), 
                                   age_cont_prop_ucl = numeric(19*n))

cod.marginal.estimates <- data.frame(stratum.id = rep(levels(droplevels(r.All$state.year.sex)), 
                                                    each = length(cods)),
                                   COD = character(6*n),
                                   COD_cont_yrs_lcl = numeric(6*n), 
                                   COD_cont_yrs = numeric(6*n), 
                                   COD_cont_yrs_ucl = numeric(6*n),
                                   COD_cont_prop_lcl = numeric(6*n), 
                                   COD_cont_prop = numeric(6*n), 
                                   COD_cont_prop_ucl = numeric(6*n), 
                                   stringsAsFactors = F)

age.cod.estimates <- data.frame(stratum.id = rep(levels(droplevels(r.All$state.year.sex)), 
                                                 each = length(cods)*length(age_minbin)),
                                age_minbin = numeric(19*6*n),   
                                COD = character(19*6*n),
                                age_COD_cont_yrs_lcl = numeric(19*6*n), 
                                age_COD_cont_yrs = numeric(19*6*n), 
                                age_COD_cont_yrs_ucl = numeric(19*6*n),
                                age_COD_cont_prop_lcl = numeric(19*6*n), 
                                age_COD_cont_prop = numeric(19*6*n), 
                                age_COD_cont_prop_ucl = numeric(19*6*n), 
                                stringsAsFactors = F)

for(i in 1:n) {

  BlackWhite[i, c("stratum.id")] <- le.calculations[[i]]$stratum.id

  #LIFE EXPECTANCY EXTRACT RESULTS
  BlackWhite[i , c("LE_white_lcl", "LE_white", "LE_white_ucl")] <- quantile(map_dbl(le.calculations[[i]]$calcs,
                                                                                    ~.x$le.df$LE_White), 
                                                                            c(0.025, 0.5, 0.975))
  BlackWhite[i , c("LE_black_lcl", "LE_black", "LE_black_ucl")] <- quantile(map_dbl(le.calculations[[i]]$calcs,
                                                                                    ~.x$le.df$LE_Black), 
                                                                            c(0.025, 0.5, 0.975))
  BlackWhite[i , c("LE_wbgap_lcl", "LE_wbgap", "LE_wbgap_ucl")] <- quantile(map_dbl(le.calculations[[i]]$calcs,
                                                                                    ~.x$le.df$LE_WBgap), 
                                                                            c(0.025, 0.5, 0.975))

  #AGE DECOMPOSITION EXTRACT RESULTS
  age.decomp.posteriors <- map_df(le.calculations[[i]]$calcs, ~ .x$age.decomp$C_x)
  age.decomp.posteriors$age_minbin <- age_minbin
  age.decomp.posteriors <- age.decomp.posteriors %>% tidyr::gather(`1`:`1000`, key = "post.samp", value = "C_x")

  age.decomp.estimates[age.decomp.estimates$stratum.id == le.calculations[[i]]$stratum.id, 
                       c("age_minbin", "age_cont_yrs_lcl", "age_cont_yrs", "age_cont_yrs_ucl")] <- 
    age.decomp.posteriors %>% 
    group_by(age_minbin) %>%
    summarise(lcl_C_x = quantile(C_x, 0.025),
              median_C_x = quantile(C_x, 0.5), 
              ucl_C_x = quantile(C_x, 0.975))

  rm(age.decomp.posteriors)

  age.decomp.posteriors2 <- map_df(le.calculations[[i]]$calcs, ~ .x$age.decomp$C_x_proportion)
  age.decomp.posteriors2$age_minbin <- age_minbin
  age.decomp.posteriors2 <- age.decomp.posteriors2 %>% tidyr::gather(`1`:`1000`, key = "post.samp", value = "C_x_proportion")

  age.decomp.estimates[age.decomp.estimates$stratum.id == le.calculations[[i]]$stratum.id, 
                       c("age_minbin", "age_cont_prop_lcl", "age_cont_prop", "age_cont_prop_ucl")] <- 
    age.decomp.posteriors2 %>% 
    group_by(age_minbin) %>%
    summarise(lcl_C_x_proportion = quantile(C_x_proportion, 0.025),
              median_C_x_proportion = quantile(C_x_proportion, 0.5), 
              ucl_C_x_proportion = quantile(C_x_proportion, 0.975))

  rm(age.decomp.posteriors2)

  #COD DECOMPOSITION EXTRACT RESULTS
  cod.marginal.posteriors <- map_df(le.calculations[[i]]$calcs, ~ .x$cod.marginal$C_x)
  cod.marginal.posteriors$COD <- as.character(cods)
  cod.marginal.posteriors <- cod.marginal.posteriors %>% tidyr::gather(`1`:`1000`, key = "post.samp", value = "C_x")

  cod.marginal.estimates[cod.marginal.estimates$stratum.id == le.calculations[[i]]$stratum.id, 
                       c("COD", "COD_cont_yrs_lcl", "COD_cont_yrs", "COD_cont_yrs_ucl")] <- 
    cod.marginal.posteriors %>% 
    group_by(COD) %>%
    summarise(lcl_C_x = quantile(C_x, 0.025),
              median_C_x = quantile(C_x, 0.5), 
              ucl_C_x = quantile(C_x, 0.975))

  rm(cod.marginal.posteriors)

  cod.marginal.posteriors2 <- map_df(le.calculations[[i]]$calcs, ~ .x$cod.marginal$C_x_proportion)
  cod.marginal.posteriors2$COD <- as.character(cods)
  cod.marginal.posteriors2 <- cod.marginal.posteriors2 %>% tidyr::gather(`1`:`1000`, key = "post.samp", 
                                                                         value = "C_x_proportion")

  cod.marginal.estimates[cod.marginal.estimates$stratum.id == le.calculations[[i]]$stratum.id,
                       c("COD", "COD_cont_prop_lcl", "COD_cont_prop", "COD_cont_prop_ucl")] <-
    cod.marginal.posteriors2 %>%
    group_by(COD) %>%
    summarise(lcl_C_x_proportion = quantile(C_x_proportion, 0.025),
              median_C_x_proportion = quantile(C_x_proportion, 0.5),
              ucl_C_x_proportion = quantile(C_x_proportion, 0.975))

  rm(cod.marginal.posteriors)

  #AGE-COD EXTRACT RESULTS
  age.cod.decomp.posteriors <- map_df(le.calculations[[i]]$calcs, ~ .x$cod.decomp$C_xi)
  age.cod.decomp.posteriors$age_minbin <- le.calculations[[i]]$calcs[[1]]$cod.decomp$Ages
  age.cod.decomp.posteriors$COD <- le.calculations[[i]]$calcs[[1]]$cod.decomp$Cause.of.death
  age.cod.decomp.posteriors <- age.cod.decomp.posteriors %>% tidyr::gather(`1`:`1000`, key = "post.samp", value = "C_xi")

  age.cod.estimates[age.cod.estimates$stratum.id == le.calculations[[i]]$stratum.id,
                       c("age_minbin", "COD", "age_COD_cont_yrs_lcl", "age_COD_cont_yrs", "age_COD_cont_yrs_ucl")] <-
    age.cod.decomp.posteriors %>%
    group_by(age_minbin, COD) %>%
    summarise(lcl_C_x = quantile(C_xi, 0.025),
              median_C_x = quantile(C_xi, 0.5),
              ucl_C_x = quantile(C_xi, 0.975))

  rm(age.cod.decomp.posteriors)

  age.cod.decomp.posteriors2 <- map_df(le.calculations[[i]]$calcs, ~ .x$cod.decomp$C_xi_proportion)
  age.cod.decomp.posteriors2$age_minbin <- le.calculations[[i]]$calcs[[1]]$cod.decomp$Ages
  age.cod.decomp.posteriors2$COD <- le.calculations[[i]]$calcs[[1]]$cod.decomp$Cause.of.death
  age.cod.decomp.posteriors2 <- age.cod.decomp.posteriors2 %>% tidyr::gather(`1`:`1000`, key = "post.samp", 
                                                                             value = "C_xi_proportion")

  age.cod.estimates[age.cod.estimates$stratum.id == le.calculations[[i]]$stratum.id,
                       c("age_minbin", "COD", "age_COD_cont_prop_lcl", "age_COD_cont_prop", "age_COD_cont_prop_ucl")] <-
    age.cod.decomp.posteriors2 %>%
    group_by(age_minbin, COD) %>%
    summarise(lcl_C_x_proportion = quantile(C_xi_proportion, 0.025),
              median_C_x_proportion = quantile(C_xi_proportion, 0.5),
              ucl_C_x_proportion = quantile(C_xi_proportion, 0.975))

  rm(age.cod.decomp.posteriors2)
}
```

```{r}
# load('~/SH_HA_data.Rdata')
# SH_sub <- dtotal %>% filter(state == "Alabama", sex == "Male", race == "Black", year == 1990)
# 
# SH_sub %>% select(age, age_minbin, deaths, population)
```

### 2017-01-27 Code/3_validation/run-models.Rmd
model for first 5 non0 states

@@
 ```{r fit_models_alabama_males}
-#run time: 15 minutes
+#run time: expect 2 hours
 #fit the bayesian models and put the 1000 posterior samples into a data frame
-r.White <- run_analysis(dataset = dat.clean, state1 = "Alabama", sex1 = "Male", race1 = "White")
+
+system.time(
+  for(sex_i in c("Male", "Female")){
+    for(state_i in c("Alabama", "Arizona", "Arkansas", "California")) {
+      assign(paste0("r.White.", state_i, sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "White"))
+      assign(paste0("r.Black.", state_i, sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "Black"))  
+    }
+  }
+)
+
+r.White <- rbind(r.White.Alabama.Male, r.White.Arizona.Male, r.White.Arkansas.Male, r.White.California.Male,
+                 r.White.Alabama.Female, r.White.Arizona.Female, r.White.Arkansas.Female, r.White.California.Female)
 r.White$race <- "White"
 
-r.Black <- run_analysis(dataset = dat.clean, state1 = "Alabama", sex1 = "Male", race1 = "Black")
+r.Black <- rbind(r.Black.Alabama.Male, r.Black.Arizona.Male, r.Black.Arkansas.Male, r.Black.California.Male,
+                 r.Black.Alabama.Female, r.Black.Arizona.Female, r.Black.Arkansas.Female, r.Black.California.Female)
 r.Black$race <- "Black"
@@
 ```{r prepare_data_for_le_functions}
-#run time: about 2 minutes
+#run time: used to be 2 minutes
 r.All <- rbind(r.White, r.Black)
@@
 
-r.All <- r.All %>% group_by(post.samp, state, race.sex, year.n, age_minbin) %>%
+system.time(r.All <- r.All %>% group_by(post.samp, state, race.sex, year.n, age_minbin) %>%
                            arrange(post.samp, state, race.sex, year.n, age_minbin) %>%
                            mutate(total_smoothed_deaths = sum(smoothed_deaths),
-                                  cod_prop_smoothed_deaths = smoothed_deaths/total_smoothed_deaths)
+                                  cod_prop_smoothed_deaths = smoothed_deaths/total_smoothed_deaths))
 ```
@@
 ```{r run_calculations}
-#run time ~40 mins
+#run time ~40 mins -- expect 5.5 hours
 #run the life table and decomposition calculations for each of the 1000 posterior samples
@@
 )
-rm(counter)
 
@@
 
+system.time(
   for(i in 1:n) {
@@
   }
+)
 ```

### 2017-01-30 Code/3_validation/run-models.Rmd
models running

@@
 library(gridExtra)
+library(parallel)
 ```
@@
 
+list_set2 <- list("Colorado", "Connecticut", "Delaware", "Washington DC", "Florida", "Georgia") #saved as results_set_2
+
 system.time(
   for(sex_i in c("Male", "Female")){
-    for(state_i in c("Alabama", "Arizona", "Arkansas", "California")) {
-      assign(paste0("r.White.", state_i, sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "White"))
-      assign(paste0("r.Black.", state_i, sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "Black"))  
+    #for(state_i in c("Alabama", "Arizona", "Arkansas", "California")) { #saved as results_set_1
+    for(state_i in c("Colorado", "Connecticut", "Delaware", "Washington DC")) { #saved as results_set_2
+      assign(paste0("r.White.", gsub(" ", "", state_i), ".", sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "White"))
+      assign(paste0("r.Black.", gsub(" ", "", state_i), ".", sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "Black"))
     }
@@
 
-r.White <- rbind(r.White.Alabama.Male, r.White.Arizona.Male, r.White.Arkansas.Male, r.White.California.Male,
-                 r.White.Alabama.Female, r.White.Arizona.Female, r.White.Arkansas.Female, r.White.California.Female)
+# r.White <- rbind(r.White.AlabamaMale, r.White.ArizonaMale, r.White.ArkansasMale, r.White.CaliforniaMale,
+#                  r.White.AlabamaFemale, r.White.ArizonaFemale, r.White.ArkansasFemale, r.White.CaliforniaFemale)
+r.White <- rbind(r.White.Colorado.Male, r.White.Connecticut.Male, r.White.Delaware.Male, r.White.WashingtonDC.Male,
+                 r.White.Colorado.Female, r.White.Connecticut.Female, r.White.Delaware.Female, r.White.WashingtonDC.Female)
 r.White$race <- "White"
 
-r.Black <- rbind(r.Black.Alabama.Male, r.Black.Arizona.Male, r.Black.Arkansas.Male, r.Black.California.Male,
-                 r.Black.Alabama.Female, r.Black.Arizona.Female, r.Black.Arkansas.Female, r.Black.California.Female)
+# r.Black <- rbind(r.Black.AlabamaMale, r.Black.ArizonaMale, r.Black.ArkansasMale, r.Black.CaliforniaMale,
+#                  r.Black.AlabamaFemale, r.Black.ArizonaFemale, r.Black.ArkansasFemale, r.Black.CaliforniaFemale)
+r.White <- rbind(r.Black.Colorado.Male, r.Black.Connecticut.Male, r.Black.Delaware.Male, r.Black.WashingtonDC.Male,
+                 r.Black.Colorado.Female, r.Black.Connecticut.Female, r.Black.Delaware.Female, r.Black.WashingtonDC.Female)
 r.Black$race <- "Black"
@@
 r.All$smoothed_deaths <- r.All$smoothed_rate*r.All$population
+table(r.All$smoothed_deaths == 0) #check this -- if no deaths in an age-band (i.e., no death for any COD, then can't run the COD calculations for that stratum.)
 
 system.time(r.All <- r.All %>% group_by(post.samp, state, race.sex, year.n, age_minbin) %>%
-                           arrange(post.samp, state, race.sex, year.n, age_minbin) %>%
-                           mutate(total_smoothed_deaths = sum(smoothed_deaths),
-                                  cod_prop_smoothed_deaths = smoothed_deaths/total_smoothed_deaths))
+                           arrange(post.samp, state, race.sex, year.n, age_minbin))
 ```
@@
 ```{r run_calculations}
-#run time ~40 mins -- expect 5.5 hours
+#run time: 8.5 hours for four states
 #run the life table and decomposition calculations for each of the 1000 posterior samples
@@
 
-```{r}
-# load('~/SH_HA_data.Rdata')
-# SH_sub <- dtotal %>% filter(state == "Alabama", sex == "Male", race == "Black", year == 1990)
-# 
-# SH_sub %>% select(age, age_minbin, deaths, population)
+```{r save_results}
+
+#SET 1: Alabama, Arizona, Arkansas, California
+save(r.All, le.calculations, file = "~/BW_results/results_set_1.Rdata") 
+write.csv(x = BlackWhite, file = "~/BW_results/BlackWhite_set_1.csv")
+write.csv(x = age.decomp.estimates, file = "~/BW_results/age_decomp_set_1.csv")
+write.csv(x = cod.marginal.estimates, file = "~/BW_results/cod_marginal_set_1.csv")
+write.csv(x = age.cod.estimates, file = "~/BW_results/age_cod_set_1.csv")
 ```

### 2017-01-30 Code/2_smoothing_models/bayes_smoothing_functions.R

@@ -128,4 +128,9 @@ clean_smoothing_results = function(data, jags_model) {
   return(merge(data$sub, smoothed.rates.1000, by = c("year.n", "age.n", "COD")))
 }
  
-
+run_analysis <- function(dataset, state1, sex1, race1) {
+  ds_sub = subset_data(ds = dataset, state = state1, sex = sex1, race = race1)
+  data.jags = jagsify_data(ds_sub)
+  jags.model = run_smoothing_model(data.jags)  
+  return(clean_smoothing_results(data = data.jags, jags_model = jags.model))
+}

### 2017-01-30 Code/2_smoothing_models/bayes_smoothing_src.R

@@ -12,7 +12,7 @@ library(bayesboot)
 library(coda)
 library(grid)
 library(gridExtra)
-library(dplyr)
+#library(dplyr)
 
 
 source("~/BlackWhiteMortalityGap/Code/2_smoothing_models/bayes_smoothing_functions.R")

### 2017-01-30 Code/life_expectancy_functions.R

@@ -208,9 +208,15 @@ life_expectancy_and_gap <- function(data) {
   age_decomp <- le_age_decomp(lt.black, "Black", lt.white, "White", "age_minbin")
   
   #calculate cod decomp
-  cod_decomp <- cause_of_death_decomp(lt.black, lt.white, age_decomp, ct.both, "age_minbin", "COD", "prop.black", "prop.white")
+  if(!any(is.na(ct.both$prop.black)) & !any(is.na(ct.both$prop.white))) { #ensures that there are no ages with 0 deaths - for these COD prop == 0 and can't compute
+    cod_decomp <- cause_of_death_decomp(lt.black, lt.white, age_decomp, ct.both, "age_minbin", "COD", "prop.black", "prop.white")
+    cod_marginal <- make.cod.marginal(cod_decomp)
+  } else {
+    cod_decomp <- NA
+    cod_marginal <- NA
+  }
   
-  cod_marginal <- make.cod.marginal(cod_decomp)
+  
   #returns a list
   return(list(le.df = data.frame("LE_Black" = lt.black$e_x[1],
                                  "LE_White" = lt.white$e_x[1],

### 2017-01-30 Code/3_validation/run-models.Rmd

@@
 #fit the bayesian models and put the 1000 posterior samples into a data frame
-
-list_set2 <- list("Colorado", "Connecticut", "Delaware", "Washington DC", "Florida", "Georgia") #saved as results_set_2
-
 system.time(
@@
 ```{r save_results}
+#SET 1: Alabama, Arizona, Arkansas, California
+# save(r.All, le.calculations, file = "~/BW_results/results_set_1.Rdata") 
+# write.csv(x = BlackWhite, file = "~/BW_results/BlackWhite_set_1.csv")
+# write.csv(x = age.decomp.estimates, file = "~/BW_results/age_decomp_set_1.csv")
+# write.csv(x = cod.marginal.estimates, file = "~/BW_results/cod_marginal_set_1.csv")
+# write.csv(x = age.cod.estimates, file = "~/BW_results/age_cod_set_1.csv")
 
-#SET 1: Alabama, Arizona, Arkansas, California
-save(r.All, le.calculations, file = "~/BW_results/results_set_1.Rdata") 
-write.csv(x = BlackWhite, file = "~/BW_results/BlackWhite_set_1.csv")
-write.csv(x = age.decomp.estimates, file = "~/BW_results/age_decomp_set_1.csv")
-write.csv(x = cod.marginal.estimates, file = "~/BW_results/cod_marginal_set_1.csv")
-write.csv(x = age.cod.estimates, file = "~/BW_results/age_cod_set_1.csv")
+#SET 2: Colorado, Connecticut, Delaware, Washington DC
+save(r.All, le.calculations, file = "~/BlackWhiteMortalityGap/Results/results_set_2.Rdata") 
+write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv")
+write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_2.csv")
+write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_2.csv")
+write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_2.csv")
 ```

### 2017-01-31 Code/run-models.Rmd
moved things and dup for local

--- 2017-01-30 Code/3_validation/run-models.Rmd
@@
 #SET 2: Colorado, Connecticut, Delaware, Washington DC
-save(r.All, le.calculations, file = "~/BlackWhiteMortalityGap/Results/results_set_2.Rdata") 
+save(r.All, le.calculations, file = "~/BW_results/results_set_2.Rdata") 
 write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv")

### 2017-01-31 Code/run-models-local.Rmd

--- 2017-01-30 Code/3_validation/run-models.Rmd
@@
 ---
-title: "Run models"
+title: "Run models - local version"
 output: github_document
@@
 library(gridExtra)
-library(parallel)
 ```
@@
 ```{r load_user_written_functions}
-source("~/BlackWhiteMortalityGap/Code/2_smoothing_models/bayes_smoothing_functions.R")
-source('~/BlackWhiteMortalityGap/Code/life_expectancy_functions.R') 
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/2_smoothing_models/bayes_smoothing_functions.R")
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
 ```
@@
 ```{r load_tidied_data}
-dat.clean = read.csv('~/dat_clean.csv') 
+dat.clean = read.csv('/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.csv') 
 ```
@@
 #fit the bayesian models and put the 1000 posterior samples into a data frame
-
-list_set2 <- list("Colorado", "Connecticut", "Delaware", "Washington DC", "Florida", "Georgia") #saved as results_set_2
-
 system.time(
+  for(state_i in c("Florida", "Georgia", "Illinois", "Indiana")) { #saved as results_set_3 
     for(sex_i in c("Male", "Female")){
-    #for(state_i in c("Alabama", "Arizona", "Arkansas", "California")) { #saved as results_set_1
-    for(state_i in c("Colorado", "Connecticut", "Delaware", "Washington DC")) { #saved as results_set_2
       assign(paste0("r.White.", gsub(" ", "", state_i), ".", sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "White"))
       assign(paste0("r.Black.", gsub(" ", "", state_i), ".", sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "Black"))
+      print(sex_i)
     }
+    print(state_i)
   }
@@
 
-# r.White <- rbind(r.White.AlabamaMale, r.White.ArizonaMale, r.White.ArkansasMale, r.White.CaliforniaMale,
-#                  r.White.AlabamaFemale, r.White.ArizonaFemale, r.White.ArkansasFemale, r.White.CaliforniaFemale)
-r.White <- rbind(r.White.Colorado.Male, r.White.Connecticut.Male, r.White.Delaware.Male, r.White.WashingtonDC.Male,
-                 r.White.Colorado.Female, r.White.Connecticut.Female, r.White.Delaware.Female, r.White.WashingtonDC.Female)
+r.White <- rbind(r.White.Florida.Male, r.White.Georgia.Male, r.White.Illinois.Male, r.White.Indiana.Male,
+                 r.White.Florida.Female, r.White.Georgia.Female, r.White.Illinois.Female, r.White.Indiana.Female)
 r.White$race <- "White"
 
-# r.Black <- rbind(r.Black.AlabamaMale, r.Black.ArizonaMale, r.Black.ArkansasMale, r.Black.CaliforniaMale,
-#                  r.Black.AlabamaFemale, r.Black.ArizonaFemale, r.Black.ArkansasFemale, r.Black.CaliforniaFemale)
-r.White <- rbind(r.Black.Colorado.Male, r.Black.Connecticut.Male, r.Black.Delaware.Male, r.Black.WashingtonDC.Male,
-                 r.Black.Colorado.Female, r.Black.Connecticut.Female, r.Black.Delaware.Female, r.Black.WashingtonDC.Female)
+r.White <- rbind(r.Black.Florida.Male, r.Black.Georgia.Male, r.Black.Illinois.Male, r.Black.Indiana.Male,
+                 r.Black.Florida.Female, r.Black.Georgia.Female, r.Black.Illinois.Female, r.Black.Indiana.Female)
 r.Black$race <- "Black"
@@
 ```{r save_results}
-
-#SET 1: Alabama, Arizona, Arkansas, California
-save(r.All, le.calculations, file = "~/BW_results/results_set_1.Rdata") 
-write.csv(x = BlackWhite, file = "~/BW_results/BlackWhite_set_1.csv")
-write.csv(x = age.decomp.estimates, file = "~/BW_results/age_decomp_set_1.csv")
-write.csv(x = cod.marginal.estimates, file = "~/BW_results/cod_marginal_set_1.csv")
-write.csv(x = age.cod.estimates, file = "~/BW_results/age_cod_set_1.csv")
+#SET 3: Florida, Georgia, Illinois, Indiana
+save(r.All, le.calculations, file = "/Users/corinneriddell/Documents/BW_results/results_set_3.Rdata") 
+write.csv(x = BlackWhite, file = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv")
+write.csv(x = age.decomp.estimates, file = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv")
+write.csv(x = cod.marginal.estimates, file = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv")
+write.csv(x = age.cod.estimates, file = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv")
 ```

### 2017-01-31 Manuscript/Manuscript using actual data.Rmd

---
title: "Untitled"
author: "Corinne Riddell"
date: "January 31, 2017"
output: html_document
---
```{r initialize, echo=FALSE, warning=F, message=F}
library(shiny)
library(plotly)
library(viridis)
library(scales)
library(shinythemes)
library(dplyr)
library(purrr)
library(stringr)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")

age_cod_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv")
age_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_1.csv")
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_1.csv")
cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_1.csv")
```

```{r tidy_data, echo=FALSE}

ncols <- str_split_fixed(as.character(BlackWhite_results$stratum.id), "\\.", 3)
BlackWhite_results$state <- as.factor(ncols[ , 1])
BlackWhite_results$year <- as.numeric(ncols[ , 2])
BlackWhite_results$sex <- as.factor(ncols[ , 3])

ncols <- str_split_fixed(as.character(cod_decomp_results$stratum.id), "\\.", 3)
cod_decomp_results$state <- as.factor(ncols[ , 1])
cod_decomp_results$year <- as.numeric(ncols[ , 2])
cod_decomp_results$sex <- as.factor(ncols[ , 3])
```

```{r add_Census_Region, echo=FALSE}
BlackWhite_results$Census_Region <- "South"
BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
                                                 "Nebraska", "North Dakota", "South Dakota", 
                                                 "Illinois", "Indiana", "Michigan", "Ohio", 
                                                 "Wisconsin")] <- "Midwest"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Connecticut", "Maine", "Massachusetts", 
                                                 "New Hampshire", "Rhode Island", "Vermont",
                                                 "New Jersey", "New York", "Pennsylvania")] <- "Northeast"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Arizona", "Colorado", "Idaho", "Montana", 
                                                 "Nevada", "New Mexico", "Utah", "Wyoming",
                                                 "Alaska", "California", "Hawaii", "Oregon",
                                                 "Washington")] <- "West"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Delaware", "Florida", "Georgia", "Maryland", 
                                                  "North Carolina", "South Carolina", 
                                                  "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee", 
                                                  "Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "South Central"

BlackWhite_results$Census_Region <- factor(BlackWhite_results$Census_Region, levels = c("West", "Midwest", "South Central", "South Atlantic", "Northeast"))
```

```{r addvar_State_abbreviation, echo=FALSE}
BlackWhite_results <- BlackWhite_results %>% mutate(stabbrs = factor(state,  
                                                                     labels = c("AL", "AZ", "AR", "CA")))
                                      # labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
                                      #            "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", 
                                      #            "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", 
                                      #            "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", 
                                      #            "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
                                      #            "VT", "VA", "WA", "WV", "WI", "WY")))

cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, stabbrs), cod_decomp_results, by = "stratum.id", all.y = T)
```

**Figure 5: Trends in life expectancy for males between 1969 and 2013 by US state**

```{r figure5, fig.align="center", fig.height=4, fig.width=9, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_white, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_line(aes(y = LE_black, col = state), lty=2) +
           geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = state), alpha=0.25) +           
           geom_ribbon(aes(ymin = LE_white_lcl, ymax = LE_white_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_wrap( ~ Census_Region) +
           ylab("Life expectancy (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), aes(label = stabbrs), check_overlap = T, size = 2.5) +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), aes(label = stabbrs, y = LE_black), check_overlap = T, size = 2.5))

```

**Figure 5: Trends in the black-white life expectancy gap for males between 1969 and 2013 by US state**

```{r figure5b, fig.align="center", fig.height=4, fig.width=9, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_wbgap, x = year)) + 
           geom_vline(aes(xintercept = 1983), lty = 2, lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), lty = 2, lwd = 0.5) +
           geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_wrap( ~ Census_Region) +
           ylab("Black-white life expectancy gap (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), aes(label = stabbrs), check_overlap = T, size = 2.5))

```

**Figure 6: Trends in life expectancy for females between 1969 and 2013 by US state**

```{r figure6a, fig.align="center", fig.height=4, fig.width=9, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_white, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_line(aes(y = LE_black, col = state), lty=2) +
           geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = state), alpha=0.25) +           
           geom_ribbon(aes(ymin = LE_white_lcl, ymax = LE_white_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_wrap( ~ Census_Region) +
           ylab("Life expectancy (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), aes(label = stabbrs), check_overlap = T, size = 2.5) +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), aes(label = stabbrs, y = LE_black), check_overlap = T, size = 2.5))

```
**Figure 6: Trends in the black-white life expectancy gap for females between 1969 and 2013 by US state**

```{r figure6, fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_wbgap, x = year)) + 
           geom_vline(aes(xintercept = 1983), lty = 2, lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), lty = 2, lwd = 0.5) +
           geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_wrap( ~ Census_Region) +
           ylab("Black-white life expectancy gap (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), aes(label = stabbrs), check_overlap = T, size = 2.5) 
) 
```

**Figure 7a: Contribution of cancer to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
```{r, echo=FALSE}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cancers"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cancers"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cancers"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cancers"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 7b: Contribution of cardiovascular disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
```{r, echo=FALSE}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cardiovascular"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cardiovascular"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 7c: Contribution of communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
```{r, echo=FALSE}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Communicable"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Communicable"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 7d: Contribution of non-communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**

```{r, echo=FALSE}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Non-communicable"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Non-communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Non-communicable"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Non-communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 7d: Contribution of injuries to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
```{r, echo=FALSE}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Injuries"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Injuries"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 7e: Contribution of all other causes to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
```{r, echo=FALSE}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "All other causes"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "All other causes"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "All other causes"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "All other causes"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8a: Contribution of cancer to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r, echo=FALSE}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cancers"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cancers"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cancers"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cancers"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8b: Contribution of cardiovascular disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r, echo=FALSE}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cardiovascular"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cardiovascular"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8c: Contribution of communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r, echo=FALSE}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Communicable"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Communicable"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8d: Contribution of non-communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r, echo=FALSE}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Non-communicable"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Non-communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Non-communicable"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Non-communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8e: Contribution of injuries to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r, echo=FALSE}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Injuries"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Injuries"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8f: Contribution of all other causes to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r, echo=FALSE}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "All other causes"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "All other causes"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "All other causes"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "All other causes"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

### 2017-02-01 Code/run-models.Rmd

@@
 
-```{r fit_models_alabama_males}
+```{r fit_models}
 #run time: expect 2 hours
@@
   for(sex_i in c("Male", "Female")){
+    print(sex_i)
     #for(state_i in c("Alabama", "Arizona", "Arkansas", "California")) { #saved as results_set_1
-    for(state_i in c("Colorado", "Connecticut", "Delaware", "Washington DC")) { #saved as results_set_2
+    #for(state_i in c("Colorado", "Connecticut", "Delaware", "Washington DC")) { #saved as results_set_2
+    for(state_i in c("Florida", "Georgia", "Illinois", "Indiana")) { #saved as results_set_3  
+      print(state_i)
       assign(paste0("r.White.", gsub(" ", "", state_i), ".", sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "White"))
@@
 #                  r.White.AlabamaFemale, r.White.ArizonaFemale, r.White.ArkansasFemale, r.White.CaliforniaFemale)
-r.White <- rbind(r.White.Colorado.Male, r.White.Connecticut.Male, r.White.Delaware.Male, r.White.WashingtonDC.Male,
-                 r.White.Colorado.Female, r.White.Connecticut.Female, r.White.Delaware.Female, r.White.WashingtonDC.Female)
+# r.White <- rbind(r.White.Colorado.Male, r.White.Connecticut.Male, r.White.Delaware.Male, r.White.WashingtonDC.Male,
+#                  r.White.Colorado.Female, r.White.Connecticut.Female, r.White.Delaware.Female, r.White.WashingtonDC.Female)
+r.White <- rbind(r.White.Florida.Male, r.White.Georgia.Male, r.White.Illinois.Male, r.White.Indiana.Male,
+                 r.White.Florida.Female, r.White.Georgia.Female, r.White.Illinois.Female, r.White.Indiana.Female)
 r.White$race <- "White"
@@
 #                  r.Black.AlabamaFemale, r.Black.ArizonaFemale, r.Black.ArkansasFemale, r.Black.CaliforniaFemale)
-r.White <- rbind(r.Black.Colorado.Male, r.Black.Connecticut.Male, r.Black.Delaware.Male, r.Black.WashingtonDC.Male,
-                 r.Black.Colorado.Female, r.Black.Connecticut.Female, r.Black.Delaware.Female, r.Black.WashingtonDC.Female)
+r.Black <- rbind(r.Black.Florida.Male, r.Black.Georgia.Male, r.Black.Illinois.Male, r.Black.Indiana.Male,
+                 r.Black.Florida.Female, r.Black.Georgia.Female, r.Black.Illinois.Female, r.Black.Indiana.Female)
 r.Black$race <- "Black"
@@
 le.calculations = vector(mode = "list", length = n)
+
 counter = 1
@@
 #SET 2: Colorado, Connecticut, Delaware, Washington DC
-save(r.All, le.calculations, file = "~/BW_results/results_set_2.Rdata") 
-write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv")
-write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_2.csv")
-write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_2.csv")
-write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_2.csv")
+# save(r.All, le.calculations, file = "~/BW_results/results_set_2.Rdata") 
+# write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv")
+# write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_2.csv")
+# write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_2.csv")
+# write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_2.csv")
+
+#SET 3: Florida, Georgia, Illinois, Indiana
+save(r.All, le.calculations, file = "~/BW_results/results_set_3.Rdata") 
+write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv")
+write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv")
+write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv")
+write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_3.csv")
 ```

### 2017-02-01 Manuscript/Manuscript using actual data.Rmd
updated with set 2 results

@@
 
-age_cod_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv")
-age_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_1.csv")
-BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_1.csv")
-cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_1.csv")
+age_cod_results <-rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"))
+
+age_cod_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"))
+
+age_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_1.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_2.csv"))
+
+BlackWhite_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_1.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv"))
+
+cod_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_1.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_2.csv"))
 ```
@@
 BlackWhite_results <- BlackWhite_results %>% mutate(stabbrs = factor(state,  
-                                                                     labels = c("AL", "AZ", "AR", "CA")))
+                                                                     labels = c("AL", "AZ", "AR", "CA",
+                                                                                "CO", "CT", "DE", "DC")))
                                       # labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",

### 2017-02-01 Code/life_expectancy_functions.R
added figures to manuscript

@@ -108,23 +108,41 @@ cause_of_death_decomp <- function(life.table1, life.table2, decomp.table,
   return(COD.decomp.table)
   }
 
-
-make_dataset_cod_plot <- function(cod.decomp.table, age.groups, cause.of.death, sign.var, decomp.var, decomp.var.prop) {
-  cod.decomp.table["group"] <- interaction(cod.decomp.table[[age.groups]], cod.decomp.table[[sign.var]], sep = ".")
-
-  cod.decomp.table <- cod.decomp.table[order(cod.decomp.table[[age.groups]], 
-                                             cod.decomp.table[[sign.var]], 
-                                             cod.decomp.table[[cause.of.death]]), ]
+#a general function that takes the decomposition table and adds variables to allow the user
+#to easily plot the decomposition in a pleasing manner
+make_pretty_decomp_plot <- function(decomp.table, strat.var.1, strat.var.2, strat.var.3, partition.bar.var, 
+                                    sign.var, decomp.var, decomp.var.prop, type.of.data = NA) {
+  if(is.na(type.of.data) == T | !(type.of.data %in% c("practice", "results"))){
+    return(print("Must supply type of data as either practice or results."))
+  }else if(type.of.data == "practice"){
+    decomp.table["group"] <- interaction(decomp.table[[strat.var.1]], 
+                                         decomp.table[[sign.var]], sep = ".")
+    
+    decomp.table <- decomp.table[order(decomp.table[[strat.var.1]], 
+                                       decomp.table[[sign.var]], 
+                                       decomp.table[[partition.bar.var]]), ]
+  }else if(type.of.data == "results"){
+    decomp.table["group"] <- interaction(decomp.table[[strat.var.1]], 
+                                         decomp.table[[strat.var.2]], 
+                                         decomp.table[[strat.var.3]], 
+                                         decomp.table[[sign.var]], sep = ".")
+    
+    decomp.table <- decomp.table[order(decomp.table[[strat.var.1]], 
+                                       decomp.table[[strat.var.2]], 
+                                       decomp.table[[strat.var.3]], 
+                                       decomp.table[[sign.var]], 
+                                       decomp.table[[partition.bar.var]]), ]    
+  }
   
[…]

### 2017-02-01 Manuscript/Manuscript using actual data.Rmd

@@
 
-ncols <- str_split_fixed(as.character(cod_decomp_results$stratum.id), "\\.", 3)
-cod_decomp_results$state <- as.factor(ncols[ , 1])
-cod_decomp_results$year <- as.numeric(ncols[ , 2])
-cod_decomp_results$sex <- as.factor(ncols[ , 3])
+age_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_decomp_results, by = "stratum.id")
+cod_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], cod_decomp_results, by = "stratum.id")
+age_cod_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_cod_results, by = "stratum.id")
+
+COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
+                "Non-communicable", "Injuries", "All other causes")
+cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
+age_cod_results$COD <- factor(age_cod_results$COD, COD.levels)
 ```
@@
 
-cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, stabbrs), cod_decomp_results, by = "stratum.id", all.y = T)
+cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, stabbrs), 
+                            cod_decomp_results, by = "stratum.id", all.y = T)
+```
+
+```{r prep_era_figure_males}
+year1 <- 1969
+year2 <- 1983
+year3 <- 1993
+year4 <- 2013
+
+bw_dat <- BlackWhite_results %>% 
+  filter(year %in% c(year1, year2, year3, year4), is.na(LE_wbgap) == F) %>%
+  select(state, year, LE_wbgap, sex) %>%
+  arrange(year)
+
+bw_dat <- tidyr::spread(bw_dat, year, LE_wbgap) 
+names(bw_dat)[3:6] <- c("first.gap","second.gap", "third.gap", "fourth.gap")  
+bw_dat <- bw_dat %>% mutate(era1.diff = first.gap - second.gap,
+                            era2.diff = second.gap - third.gap,
+                            era3.diff = third.gap - fourth.gap,
+                            state.gdiff.order.1 = reorder(state, first.gap),
+                            state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
+                            state.gdiff.order.2 = reorder(state, second.gap),
+                            state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
+                            state.gdiff.order.3 = reorder(state, third.gap),
+                            state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
+)
+
+cod.df <- data.frame(state = factor(), sex = factor(), era = character(), COD = factor(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
+age.df <- data.frame(state = factor(), sex = factor(), era = character(), age_minbin = numeric(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
+
+years <- c(1969, 1983, 1993, 2013)
+
+for (state.i in levels(BlackWhite_results$state)) {
+  for (sex.i in c("Male", "Female")) {
+    for (year.i in 1:3) {
+
+      table1 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(COD)
+      table2 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(COD)
+
+      table3 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(age_minbin)
+      table4 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(age_minbin)
+
+      temp1 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
+                          contribution.to.gap.change(type.of.decomp = "COD", 
+                                                     decomp.table1 = table1, 
+                                                     decomp.table2 = table2, 
+                                                     type.of.data = "results"))
+
+      temp2 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
+                          contribution.to.gap.change(type.of.decomp = "Age", 
+                                                     decomp.table1 = table3, 
+                                                     decomp.table2 = table4,
+                                                     type.of.data = "results"))
+
+      cod.df <- rbind(cod.df, temp1)
+      age.df <- rbind(age.df, temp2)
+    }
+  }
+}
+
+```
+
+```{r more_figure_prep}
+
+cod.df <- make_pretty_decomp_plot(decomp.table = cod.df, strat.var.1 = "state", 
+                        strat.var.2 = "sex", strat.var.3 = "era", 
+                        sign.var = "narrowed_gap",
+                        decomp.var = "Contribution.to.change",
+                        decomp.var.prop = "Contrib.to.change.prop", 
+                        partition.bar.var = "COD", type.of.data = "results")
+
+age.df <- make_pretty_decomp_plot(decomp.table = age.df, strat.var.1 = "state", 
+                        strat.var.2 = "sex", strat.var.3 = "era", 
+                        sign.var = "narrowed_gap",
+                        decomp.var = "Contribution.to.change",
+                        decomp.var.prop = "Contrib.to.change.prop", 
+                        partition.bar.var = "Ages", type.of.data = "results")
+
+```
+
+**Figure 1a: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
+
+```{r figure1, fig.align = "center", echo=F}
+
+cod.df2 <- merge(cod.df, bw_dat, by = c("state", "sex"))
+cod.df2 <- cod.df2 %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
+                              new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
+                              new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)
+
+y.labels <- levels(factor(cod.df2$state.gdiff.order.1))
+y.num.breaks <- length(levels(factor(cod.df2$state.gdiff.order.1)))
+
+ggplotly(
+  ggplot(subset(cod.df2, sex == "Male" & era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
+    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
+                  ymax = state.gdiff.num.1 + 0.45, 
+                  xmax = new.finish.1, fill = COD), 
+              color = "white") +
+    geom_segment(data = subset(bw_dat, sex == "Male"), 
+                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
+    geom_segment(data = subset(bw_dat, sex == "Male"), 
+                 aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>% 
+  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
+
+```
+
+**Figure 1b: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1983 and 1993 by US state**
+
+```{r}
+y.labels.2 <- levels(factor(cod.df2$state.gdiff.order.2))
+y.num.breaks.2 <- length(levels(factor(cod.df2$state.gdiff.order.2)))
+
+ggplotly(
+  ggplot(subset(cod.df2, sex == "Male" & era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
+    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
+                  ymax = state.gdiff.num.2 + 0.45, 
+                  xmax = new.finish.2, fill = COD), 
+              color = "white") +
+    geom_segment(data = subset(bw_dat, sex == "Male"), 
+                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
+    geom_segment(data = subset(bw_dat, sex == "Male"), 
+                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>% 
+  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
+```
+
+**Figure 1c: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
+
+```{r}
+y.labels.3 <- levels(factor(cod.df2$state.gdiff.order.3))
+y.num.breaks.3 <- length(levels(factor(cod.df2$state.gdiff.order.3)))
+
+ggplotly(
+  ggplot(subset(cod.df2, sex == "Male" & era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
+    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
+                  ymax = state.gdiff.num.3 + 0.45, 
+                  xmax = new.finish.3, fill = COD), 
+              color = "white") +
+    geom_segment(data = subset(bw_dat, sex == "Male"), 
+                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
+    geom_segment(data = subset(bw_dat, sex == "Male"), 
+                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>% 
+  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))
 ```

### 2017-02-02 Manuscript/Manuscript using actual data.Rmd

@@
 age_cod_results <-rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"))
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"))
 
 age_cod_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"))
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"))
 
 age_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_1.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_2.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_2.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv"))
 
 BlackWhite_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_1.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv"))
 
 cod_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_1.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_2.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_2.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv"))
 ```
@@
                                                                      labels = c("AL", "AZ", "AR", "CA",
-                                                                                "CO", "CT", "DE", "DC")))
+                                                                                "CO", "CT", "DE", "DC",
+                                                                                "FL", "GA", "IL", "IN")))
                                       # labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",

### 2017-02-02 Manuscript/Manuscript using actual data.Rmd
fixed order in plots

@@
 names(bw_dat)[3:6] <- c("first.gap","second.gap", "third.gap", "fourth.gap")  
-bw_dat <- bw_dat %>% mutate(era1.diff = first.gap - second.gap,
+bw_dat.males <- bw_dat %>% filter(sex == "Male") %>%
+  mutate(era1.diff = first.gap - second.gap,
          era2.diff = second.gap - third.gap,
@@
 
-cod.df <- data.frame(state = factor(), sex = factor(), era = character(), COD = factor(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
-age.df <- data.frame(state = factor(), sex = factor(), era = character(), age_minbin = numeric(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
+bw_dat.females <- bw_dat %>% filter(sex == "Female") %>%
+  mutate(era1.diff = first.gap - second.gap,
+         era2.diff = second.gap - third.gap,
+         era3.diff = third.gap - fourth.gap,
+         state.gdiff.order.1 = reorder(state, first.gap),
+         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
+         state.gdiff.order.2 = reorder(state, second.gap),
+         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
+         state.gdiff.order.3 = reorder(state, third.gap),
+         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
+  )
+
+cod.df <- data.frame(state = factor(), sex = factor(), era = character(), 
+                     COD = factor(), Contribution.to.change = numeric(), 
+                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())
+
+age.df <- data.frame(state = factor(), sex = factor(), era = character(), 
+                     age_minbin = numeric(), Contribution.to.change = numeric(), 
+                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())
 
@@
 
-cod.df2 <- merge(cod.df, bw_dat, by = c("state", "sex"))
-cod.df2 <- cod.df2 %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
+cod.df.males <- merge(subset(cod.df, sex == "Male"), bw_dat.males, by = c("state"))
+cod.df.males <- cod.df.males %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
                                         new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
@@
 
-y.labels <- levels(factor(cod.df2$state.gdiff.order.1))
-y.num.breaks <- length(levels(factor(cod.df2$state.gdiff.order.1)))
+y.labels <- levels(factor(cod.df.males$state.gdiff.order.1))
+y.num.breaks <- length(levels(factor(cod.df.males$state.gdiff.order.1)))
 
 ggplotly(
-  ggplot(subset(cod.df2, sex == "Male" & era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
+  ggplot(subset(cod.df.males, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
     geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
@@
               color = "white") +
-    geom_segment(data = subset(bw_dat, sex == "Male"), 
+    geom_segment(data = bw_dat.males,
                  aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
-    geom_segment(data = subset(bw_dat, sex == "Male"), 
+    geom_segment(data = bw_dat.males,
                  aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
@@
 ```{r}
-y.labels.2 <- levels(factor(cod.df2$state.gdiff.order.2))
-y.num.breaks.2 <- length(levels(factor(cod.df2$state.gdiff.order.2)))
+y.labels.2 <- levels(factor(cod.df.males$state.gdiff.order.2))
+y.num.breaks.2 <- length(levels(factor(cod.df.males$state.gdiff.order.2)))
 
 ggplotly(
-  ggplot(subset(cod.df2, sex == "Male" & era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
+  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
     geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
@@
               color = "white") +
-    geom_segment(data = subset(bw_dat, sex == "Male"), 
+    geom_segment(data = bw_dat.males, 
                  aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
-    geom_segment(data = subset(bw_dat, sex == "Male"), 
+    geom_segment(data = bw_dat.males, 
                  aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
@@
 ```{r}
-y.labels.3 <- levels(factor(cod.df2$state.gdiff.order.3))
-y.num.breaks.3 <- length(levels(factor(cod.df2$state.gdiff.order.3)))
+y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
+y.num.breaks.3 <- length(levels(factor(cod.df.males$state.gdiff.order.3)))
 
 ggplotly(
-  ggplot(subset(cod.df2, sex == "Male" & era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
+  ggplot(subset(cod.df.males, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
     geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
@@
               color = "white") +
-    geom_segment(data = subset(bw_dat, sex == "Male"), 
+    geom_segment(data = bw_dat.males, 
                  aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
-    geom_segment(data = subset(bw_dat, sex == "Male"), 
+    geom_segment(data = bw_dat.males, 
                  aes(x = third.gap, xend= third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +

### 2017-02-02 Manuscript/Manuscript using actual data.Rmd
added interpretation

@@
 
-```{r prep_era_figure_males}
+```{r prep_era_figure_males, echo = F}
 year1 <- 1969
@@
 
-```{r more_figure_prep}
+```{r more_figure_prep, echo = F}
 
@@
 
+What we see so far (based on these 12 states):
+
+- The gap increases for some states (DC, big increase, Cali, Connecticut, and small increase in 
+Colorado and Indiana). It decreases for the rest, with the largest decrease in gap seen in 
+Georgia (largely due to injuries), Florida (injuries), Illinois, and Alabama/
+
+- Injuries across the board contribute to narrowing the gap over time.
+
+- Communicable disease narrows (except arizona), but much smaller contribution than injuries.
+
+- All other causes narrows, except DC, again much smaller contribution than injuries.
+
+- CVD increase gap, except in Illinois.
+
+- Cancer increases the gap everywhere. Size of contribution to change is a little smaller than CVD.
+
+- Non-communicable increase the gap (except Colorado) but the contribution is tiny.
+
 **Figure 1b: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1983 and 1993 by US state**
 
-```{r}
+```{r, echo = F}
 y.labels.2 <- levels(factor(cod.df.males$state.gdiff.order.2))
@@
 
+- Be careful with the interpretation: the x-axis is much wider for this plot because of DC.
+
+- Except for Arizona, the life expectancy gap increases in each state, with the biggest driver being communicable disease (HIV/AIDS).
+Injuries also contribute to the gap widening over this time frame.
+Thus, compared to previous period, these two causes of death led to a widening rather than a narrowing of the gap.
+
+- CVD and cancer also contributed to widening the gap over time, which also occurred in the previous period.
+Also similar was that non-communicable disease doesn't have much of an effect on the gap changing, while all other causes contributes to narrowing the gap.
+
+- DC was especially hard hit, with injuries driving the increase in the gap from 10.5 years to 15.9 years in this 10-year period.
+
 **Figure 1c: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
 
-```{r}
+```{r, echo = F}
 y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
@@
 ```
+
+- After 1993, the gap narrowed for all states shown, and in most cases all of the causes contributed to the narrowing of the gap.
 
@@
 **Figure 7a: Contribution of cancer to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-```{r, echo=FALSE}
+```{r, fig.align="center", fig.height=5, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cancers"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 7b: Contribution of cardiovascular disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-```{r, echo=FALSE}
+```{r, fig.align="center", fig.height=5, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 7c: Contribution of communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-```{r, echo=FALSE}
+```{r, fig.align="center", fig.height=5, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Communicable"), aes(x = year, y = COD_cont_yrs)) + 
@@
 
-```{r, echo=FALSE}
+```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Non-communicable"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 7d: Contribution of injuries to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-```{r, echo=FALSE}
+```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Injuries"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 7e: Contribution of all other causes to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-```{r, echo=FALSE}
+```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "All other causes"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 8a: Contribution of cancer to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r, echo=FALSE}
+```{r, fig.align="center", fig.height=5, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cancers"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 8b: Contribution of cardiovascular disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r, echo=FALSE}
+```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 8c: Contribution of communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r, echo=FALSE}
+```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Communicable"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 8d: Contribution of non-communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r, echo=FALSE}
+```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Non-communicable"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 8e: Contribution of injuries to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r, echo=FALSE}
+```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Injuries"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 8f: Contribution of all other causes to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r, echo=FALSE}
+```{r, fig.align="center", fig.height=5, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "All other causes"), aes(x = year, y = COD_cont_yrs)) + 

### 2017-02-02 Manuscript/Manuscript using actual data.Rmd

@@
 
-ggplotly(
+males.era1 <- ggplotly(
   ggplot(subset(cod.df.males, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
@@
 
+males.era1
 ```
@@
 
-ggplotly(
-  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
-    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
-                  ymax = state.gdiff.num.2 + 0.45, 
+males.era2 <- ggplotly(
+  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.1, x = new.start.2)) +
+    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.1 - 0.45, 
+                  ymax = state.gdiff.num.1 + 0.45, 
                   xmax = new.finish.2, fill = COD), 
@@
     geom_segment(data = bw_dat.males, 
-                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
+                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
     geom_segment(data = bw_dat.males, 
-                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>% 
+                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>% 
   layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
+
+males.era2 
 ```
@@
 
-ggplotly(
-  ggplot(subset(cod.df.males, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
-    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
-                  ymax = state.gdiff.num.3 + 0.45, 
+males.era3 <- ggplotly(
+  ggplot(subset(cod.df.males, era == "1993-2013"), aes(y = state.gdiff.num.1, x = new.start.3)) +
+    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.1 - 0.45, 
+                  ymax = state.gdiff.num.1 + 0.45, 
                   xmax = new.finish.3, fill = COD), 
@@
     geom_segment(data = bw_dat.males, 
-                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
+                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
     geom_segment(data = bw_dat.males, 
-                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>% 
+                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>% 
   layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))
+
+males.era3
 ```
@@
 - After 1993, the gap narrowed for all states shown, and in most cases all of the causes contributed to the narrowing of the gap.
+
+```{r, fig.align="center", fig.height=5, fig.width=9, echo=F, eval = F}
+subplot(males.era1, males.era2, males.era3, shareY = T)
+```
+
+**Figure 1a: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
+
+```{r figure1, fig.align = "center", echo=F}
+
+cod.df.females <- merge(subset(cod.df, sex == "Female"), bw_dat.females, by = c("state"))
+cod.df.females <- cod.df.females %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
+                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
+                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)
+
+y.labels <- levels(factor(cod.df.females$state.gdiff.order.1))
+y.num.breaks <- length(levels(factor(cod.df.females$state.gdiff.order.1)))
+
+females.era1 <- ggplotly(
+  ggplot(subset(cod.df.females, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
+    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
+                  ymax = state.gdiff.num.1 + 0.45, 
+                  xmax = new.finish.1, fill = COD), 
+              color = "white") +
+    geom_segment(data = bw_dat.females,
+                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
+    geom_segment(data = bw_dat.females,
+                 aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
+  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
+
+females.era1
+```
+
+- The gap in life expectancy narrowed for all states shown, except for Arizona and small increase in DC.
+
+- CVD was the biggest contributor to the gap narrowing, but all other causes and communicable diseases led to narrowing in many states.
+Injuries and non-communicable disease also led to narrowing, but the contribution was small.
+Cancer prevented the gap from narrowing further in some states.
 

### 2017-02-02 Manuscript/Manuscript using actual data.Rmd
added all female plots

@@
 
+cod.df <- merge(cod.df, BlackWhite_results[BlackWhite_results$year == 1969 & BlackWhite_results$sex == "Male", 
+                                           c("state", "Census_Region", "stabbrs")], by = c("state"))
 ```
@@
               color = "white") +
-    geom_segment(data = bw_dat.males,
-                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
-    geom_segment(data = bw_dat.males,
-                 aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
+#    facet_wrap(~ Census_Region) +
+    geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
+        geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
+   # geom_segment(data = bw_dat.males,
+  #               aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
+  #  geom_segment(data = bw_dat.males,
+  #               aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
     theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
@@
 
-```{r, echo = F}
+```{r figure1b, echo = F}
 y.labels.2 <- levels(factor(cod.df.males$state.gdiff.order.2))
@@
 males.era2 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.1, x = new.start.2)) +
-    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.1 - 0.45, 
-                  ymax = state.gdiff.num.1 + 0.45, 
+  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
+    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
+                  ymax = state.gdiff.num.2 + 0.45, 
                   xmax = new.finish.2, fill = COD), 
@@
     geom_segment(data = bw_dat.males, 
-                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
+                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
     geom_segment(data = bw_dat.males, 
-                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>% 
+                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>% 
   layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
@@
 
-```{r, echo = F}
+```{r figure1c, echo = F}
 y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
@@
 males.era3 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "1993-2013"), aes(y = state.gdiff.num.1, x = new.start.3)) +
-    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.1 - 0.45, 
-                  ymax = state.gdiff.num.1 + 0.45, 
+  ggplot(subset(cod.df.males, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
+    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
+                  ymax = state.gdiff.num.3 + 0.45, 
                   xmax = new.finish.3, fill = COD), 
@@
     geom_segment(data = bw_dat.males, 
-                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
+                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
     geom_segment(data = bw_dat.males, 
-                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>% 
+                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>% 
   layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))
@@
 
-**Figure 1a: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
+**Figure 2a: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
 
-```{r figure1, fig.align = "center", echo=F}
+```{r figure2a, fig.align = "center", echo=F}
 
@@
 Cancer prevented the gap from narrowing further in some states.
+
+**Figure 2b: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1983 and 1993 by US state**
+
+```{r figure2b, fig.align = "center", echo=F}
+y.labels.2 <- levels(factor(cod.df.females$state.gdiff.order.2))
+y.num.breaks.2 <- length(levels(factor(cod.df.females$state.gdiff.order.2)))
+
+females.era2 <- ggplotly(
+  ggplot(subset(cod.df.females, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
+    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
+                  ymax = state.gdiff.num.2 + 0.45, 
+                  xmax = new.finish.2, fill = COD), 
+              color = "white") +
+    geom_segment(data = bw_dat.females,
+                 aes(x = third.gap, xend = third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
+    geom_segment(data = bw_dat.females,
+                 aes(x = second.gap, xend = second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>%
+  layout(xaxis = list(title = "Life expectancy gap (years): 1983-1993"), yaxis = list(title = NA, autorange = "reversed"))
+
+females.era2
+```
+
+- Recall that during this time frame, the gap for males widened, and the largest contributors across the board were communicable disease and injuries.
+For females, the story is less consistent across states -- for some states, the gap widened, and for others, it narrowed.
+
+- Injuries did play a large role at increasing the gap in Florida, DC, and Connecticut.
+CVD was also generally associated with the gap increasing, while 'all other causes' was associated with narrowing the gap.
+
+**Figure 2c: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
+
+```{r figure2c, fig.align = "center", echo=F}
+y.labels.3 <- levels(factor(cod.df.females$state.gdiff.order.3))
+y.num.breaks.3 <- length(levels(factor(cod.df.females$state.gdiff.order.3)))
+
+females.era3 <- ggplotly(
+  ggplot(subset(cod.df.females, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
+    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
+                  ymax = state.gdiff.num.3 + 0.45, 
+                  xmax = new.finish.3, fill = COD), 
+              color = "white") +
+    geom_segment(data = bw_dat.females,
+                 aes(x = fourth.gap, xend = fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
+    geom_segment(data = bw_dat.females,
+                 aes(x = third.gap, xend = third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>%
+  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))
+
+females.era3
+```
+
+- In-line with males, the life expectancy gap decreased during the last 20 year period, and generally speaking all the causes contributed to the gap narrowing over time.
+The biggest contributors were CVD and all other causes.
 

### 2017-02-03 Code/run-models.Rmd
set 4 results

@@
     #for(state_i in c("Colorado", "Connecticut", "Delaware", "Washington DC")) { #saved as results_set_2
-    for(state_i in c("Florida", "Georgia", "Illinois", "Indiana")) { #saved as results_set_3  
+    #for(state_i in c("Florida", "Georgia", "Illinois", "Indiana")) { #saved as results_set_3  
+    for(state_i in c("Iowa", "Kansas", "Kentucky", "Louisiana")) { #saved as results_set_4  
       print(state_i)
@@
 #                  r.White.Colorado.Female, r.White.Connecticut.Female, r.White.Delaware.Female, r.White.WashingtonDC.Female)
-r.White <- rbind(r.White.Florida.Male, r.White.Georgia.Male, r.White.Illinois.Male, r.White.Indiana.Male,
-                 r.White.Florida.Female, r.White.Georgia.Female, r.White.Illinois.Female, r.White.Indiana.Female)
+# r.White <- rbind(r.White.Florida.Male, r.White.Georgia.Male, r.White.Illinois.Male, r.White.Indiana.Male,
+#                  r.White.Florida.Female, r.White.Georgia.Female, r.White.Illinois.Female, r.White.Indiana.Female)
+r.White <- rbind(r.White.Iowa.Male, r.White.Kansas.Male, r.White.Kentucky.Male, r.White.Louisiana.Male,
+                 r.White.Iowa.Female, r.White.Kansas.Female, r.White.Kentucky.Female, r.White.Louisiana.Female)
 r.White$race <- "White"
@@
 #                  r.Black.AlabamaFemale, r.Black.ArizonaFemale, r.Black.ArkansasFemale, r.Black.CaliforniaFemale)
-r.Black <- rbind(r.Black.Florida.Male, r.Black.Georgia.Male, r.Black.Illinois.Male, r.Black.Indiana.Male,
-                 r.Black.Florida.Female, r.Black.Georgia.Female, r.Black.Illinois.Female, r.Black.Indiana.Female)
+r.Black <- rbind(r.Black.Iowa.Male, r.Black.Kansas.Male, r.Black.Kentucky.Male, r.Black.Louisiana.Male,
+                 r.Black.Iowa.Female, r.Black.Kansas.Female, r.Black.Kentucky.Female, r.Black.Louisiana.Female)
 r.Black$race <- "Black"
@@
 #SET 3: Florida, Georgia, Illinois, Indiana
-save(r.All, le.calculations, file = "~/BW_results/results_set_3.Rdata") 
-write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv")
-write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv")
-write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv")
-write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_3.csv")
+# save(r.All, le.calculations, file = "~/BW_results/results_set_3.Rdata") 
+# write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv")
+# write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv")
+# write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv")
+# write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_3.csv")
+
+#SET 4: Iowa, Kansas, Kentucky, Louisiana
+save(r.All, le.calculations, file = "~/BW_results/results_set_4.Rdata") 
+write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv")
+write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv")
+write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv")
+write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_4.csv")
+
 ```

### 2017-02-03 Manuscript/Manuscript using actual data.Rmd
changes to shiny app

@@
 ```
+
+**eFigure 1: Difference in life expectancy between black in white males, United States, 1969-2013**
+
+![Difference in life expectancy between black in white males, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Men.bmp)
+
+**eFigure 2: Difference in life expectancy between black in white females, United States, 1969-2013**
+
+![Difference in life expectancy between black in white females, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Women.bmp)

### 2017-02-03 Manuscript/Manuscript.Rmd

@@
 **Background notes from meeting with SH on Dec 20**
--CVD declining starting 1968 - see paper Ruel Stallones - cited in SH JAMA paper - history of decline in CVD.
--Richard cooper also written about BW diff in CVD with epi perspective
--for men there is stagnation from 1969 to late 1980s and then increase in gap with HIV and homicides
--massive crime wave in 1980s, black LE went down from 1984 to 1987 - period when both HIV and homicide went up -- the whole story during that time was HIV and homicide
--fundamentally interested in the qu is there a common story across most states or are there important differences in what happened in NY vs CA vs TX - can we say anything about what we know about state-specific policies, for example CA has strong anti-smoking legislation.
+- CVD declining starting 1968 - see paper Ruel Stallones - cited in SH JAMA paper - history of decline in CVD.
+
+- Richard cooper also written about BW diff in CVD with epi perspective
+
+- for men there is stagnation from 1969 to late 1980s and then increase in gap with HIV and homicides
+
+- massive crime wave in 1980s, black LE went down from 1984 to 1987 - period when both HIV and homicide went up -- the whole story during that time was HIV and homicide
+
+- fundamentally interested in the qu is there a common story across most states or are there important differences in what happened in NY vs CA vs TX - can we say anything about what we know about state-specific policies, for example CA has strong anti-smoking legislation.
 Does this help us understand why the gap in Wisconsin was increasing?
--is there a lot of heterogeneity or is there not?
+
+- is there a lot of heterogeneity or is there not?
 - same something about the proportionate contribution by COD and by age - e.g., changes in injuries accounting for X% for closing the gap in NY but in AL it was Y%.
--heterogeneity between states in the contribution of the causes to the decline in the gap - maybe need to break up into at least two time periods with split at 1990.
--this is where the new figures come in which on the y-axis have the contribution to LE gap (in years or %) vs year on the x axis, and geom_line for each COD (do one COD at a time), and individual line for each state (may facet by census region).
--then have a view of how the different COD have contributed to the decline in the gap over time.
+
+- heterogeneity between states in the contribution of the causes to the decline in the gap - maybe need to break up into at least two time periods with split at 1990.
+
+- this is where the new figures come in which on the y-axis have the contribution to LE gap (in years or %) vs year on the x axis, and geom_line for each COD (do one COD at a time), and individual line for each state (may facet by census region).
+
+- then have a view of how the different COD have contributed to the decline in the gap over time.
 
@@
 Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex.
-The difference in life expectancies between blacks and white was calculated by subtracting the black life expectancy from the white life expectancy in every state, year, and sex.
+The difference in life expectancies between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in every state, year, and sex.
 
-To decompose the black-white difference in life expectancy, we performed a decomposition analysis using the Arriaga method [ref].
+To decompose the year- and state- specific difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref].
 First, the decomposition analysis partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks.
@@
 We also summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference.
+
+In order to understand whether states fared similarly over time, we divided the time frame into three eras which roughly encapsulate an initial period of increases in life expectancy for both races and sexes (1969-1983), an intermediate period of decreases and stagnating life expectancy for blacks but not for whites (1983-1993) and a final period of increasing life expectancy for both races (1993-2013).
+While these eras represent different lengths of time, they best capture three distinct periods of changes to the difference in life expectancy: an intial narrowing of the gap, followed by a widening of the gap that diminished previous progress, and the most recent twenty years of sustained narrowing (Appendix Figures 1 and 2).
 
@@
 
+**eFigure 1: Difference in life expectancy between black in white males, United States, 1969-2013**
+
+![Difference in life expectancy between black in white males, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Men.bmp)
+
+**eFigure 2: Difference in life expectancy between black in white females, United States, 1969-2013**
+
+![Difference in life expectancy between black in white females, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Women.bmp)
+
 **Life table and decomposition methodology**

### 2017-02-03 Code/run-models.Rmd

@@
     #for(state_i in c("Florida", "Georgia", "Illinois", "Indiana")) { #saved as results_set_3  
-    for(state_i in c("Iowa", "Kansas", "Kentucky", "Louisiana")) { #saved as results_set_4  
+    #for(state_i in c("Iowa", "Kansas", "Kentucky", "Louisiana")) { #saved as results_set_4  
+     for(state_i in c("Maryland", "New York", "North Carolina", "Texas", "Virginia", "Wisconsin" )) { #saved as results_set_5  
       print(state_i)
@@
 #                  r.White.Florida.Female, r.White.Georgia.Female, r.White.Illinois.Female, r.White.Indiana.Female)
-r.White <- rbind(r.White.Iowa.Male, r.White.Kansas.Male, r.White.Kentucky.Male, r.White.Louisiana.Male,
-                 r.White.Iowa.Female, r.White.Kansas.Female, r.White.Kentucky.Female, r.White.Louisiana.Female)
+# r.White <- rbind(r.White.Iowa.Male, r.White.Kansas.Male, r.White.Kentucky.Male, r.White.Louisiana.Male,
+#                  r.White.Iowa.Female, r.White.Kansas.Female, r.White.Kentucky.Female, r.White.Louisiana.Female)
+r.White <- rbind(r.White.Maryland.Male, r.White.NewYork.Male, r.White.NorthCarolina.Male, r.White.Texas.Male, 
+                 r.White.Virginia.Male, r.White.Wisconsin.Male, 
+                 r.White.Maryland.Female, r.White.NewYork.Female, r.White.NorthCarolina.Female, r.White.Texas.Female, 
+                 r.White.Virginia.Female, r.White.Wisconsin.Female)
 r.White$race <- "White"
@@
 #                  r.Black.AlabamaFemale, r.Black.ArizonaFemale, r.Black.ArkansasFemale, r.Black.CaliforniaFemale)
-r.Black <- rbind(r.Black.Iowa.Male, r.Black.Kansas.Male, r.Black.Kentucky.Male, r.Black.Louisiana.Male,
-                 r.Black.Iowa.Female, r.Black.Kansas.Female, r.Black.Kentucky.Female, r.Black.Louisiana.Female)
+r.Black <- rbind(r.Black.Maryland.Male, r.Black.NewYork.Male, r.Black.NorthCarolina.Male, r.Black.Texas.Male, 
+                 r.Black.Virginia.Male, r.Black.Wisconsin.Male, 
+                 r.Black.Maryland.Female, r.Black.NewYork.Female, r.Black.NorthCarolina.Female, r.Black.Texas.Female, 
+                 r.Black.Virginia.Female, r.Black.Wisconsin.Female)
 r.Black$race <- "Black"
@@
 #SET 4: Iowa, Kansas, Kentucky, Louisiana
-save(r.All, le.calculations, file = "~/BW_results/results_set_4.Rdata") 
-write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv")
-write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv")
-write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv")
-write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_4.csv")
+# save(r.All, le.calculations, file = "~/BW_results/results_set_4.Rdata") 
+# write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv")
+# write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv")
+# write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv")
+# write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_4.csv")
 
+#SET 5: Maryland, New York, North Carolina, Texas, Virginia, Wisconsin
+save(r.All, le.calculations, file = "~/BW_results/results_set_5.Rdata") 
+write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv")
+write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv")
+write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv")
+write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_5.csv")
 ```

### 2017-02-03 Manuscript/Manuscript using actual data.Rmd

@@
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"))
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"),
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"))
 
@@
                          read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"))
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"))
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_2.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv"))
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv"))
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_2.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv"))
 ```
@@
                                                                      labels = c("AL", "AZ", "AR", "CA",
-                                                                                "CO", "CT", "DE", "DC",
-                                                                                "FL", "GA", "IL", "IN")))
+                                                                                "CO", "CT", "DE",
+                                                                                "FL", "GA", "IL", "IN",
+                                                                                "IA", "KS", "KY", "LA",
+                                                                                "DC")))
                                       # labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
@@
 
-```{r figure5, fig.align="center", fig.height=4, fig.width=9, echo=F}
+```{r figure5, fig.align="center", fig.height=4, fig.width=11, echo=F}
 ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_white, x = year)) + 
@@
            geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_line(aes(y = LE_black, col = state), lty=2) +
-           geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = state), alpha=0.25) +           
            geom_ribbon(aes(ymin = LE_white_lcl, ymax = LE_white_ucl, fill = state), alpha=0.25) +
@@
            xlab("Year") +
-           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), aes(label = stabbrs), check_overlap = T, size = 2.5) +
-           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), aes(label = stabbrs, y = LE_black), check_overlap = T, size = 2.5))
+           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), 
+                     aes(label = stabbrs), check_overlap = T, size = 2.5))
+
+ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_black, x = year)) + 
+           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
+           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
+           geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = state), alpha=0.25) +
+           geom_line(aes(col = state)) +
+           facet_wrap( ~ Census_Region) +
+           ylab("Life expectancy (years)") +
+           xlab("Year") +
+           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), 
+                     aes(label = stabbrs), check_overlap = T, size = 2.5))
 
@@
            xlab("Year") +
-           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), aes(label = stabbrs), check_overlap = T, size = 2.5) +
-           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), aes(label = stabbrs, y = LE_black), check_overlap = T, size = 2.5))
+           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), 
+                     aes(label = stabbrs), check_overlap = T, size = 2.5))
+
+ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_black, x = year)) + 
+           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
+           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
+           geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = state), alpha=0.25) +
+           geom_line(aes(col = state)) +
+           facet_wrap( ~ Census_Region) +
+           ylab("Life expectancy (years)") +
+           xlab("Year") +
+           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), 
+                     aes(label = stabbrs), check_overlap = T, size = 2.5))
 

### 2017-02-03 Manuscript/Manuscript using actual data.Rmd
added set 5 results

@@
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"))
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"))
 
@@
                          read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"))
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"))
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv"))
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv"))
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv"))
 ```
@@
                                                                                 "IA", "KS", "KY", "LA",
-                                                                                "DC")))
+                                                                                "MD", "NY", "NC", "TX",
+                                                                                "VA", "DC", "WI")))
+
                                       # labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",

### 2017-02-03 Manuscript/Manuscript_using_actual_data.Rmd
more text

---
title: "Exploring how changes in age at death and cause of death have narrowed the black-white life expectancy gap over forty-five years in the United States."
output: html_document
---

##Riddell CA, Morrison KT, Kaufman JS, Harper S

###Abstract

###Introduction
The life expectancy of a population serves as a barometer for overall health. In 2014, life expectancy in the United States was 78.7 years, but hidden within this figure, are differences in life expectancy by race. Non-hispanic whites lived 3.6 years longer than non-hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white americans have experienced increases in life expectancy since X, the difference between their life expectancies has varied. During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, heart disease [Harper 2007]. The gap continued to narrow during the 2000s, but states differed in the magnitude of improvement that was made. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014]. This study seeks to better understand these state-level differences, by examining how changes in age at death and cause of death over time have worked to narrow or exacerbate the life expectancy gap.

- fundamentally interested in the qu is there a common story across most states or are there important differences in what happened in NY vs CA vs TX - can we say anything about what we know about state-specific policies, for example CA has strong anti-smoking legislation. Does this help us understand why the gap in Wisconsin was increasing?

- is there a lot of heterogeneity or is there not? - same something about the proportionate contribution by COD and by age - e.g., changes in injuries accounting for X% for closing the gap in NY but in AL it was Y%.

###Methods

**The data**

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEER*Stat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia. The mortality data was originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data was derived by the NCHS using intercensal bridged-race population estimates. This study using data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were then grouped into 9X causes of death within the software, which we further aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes. This recoding is illustrated in eTable 1 in the eAppendix of the Supplement. When the number of deaths for a population is between 1 and 9, this data is suppressed by the software and produces a missing death count for that strata. However, death counts of exactly 0 or greater than 9 are known precisely.

- CR to add line about which states were not considered.

- CR to add line about hispanic origin.

**Statistical Methods**

The data produced two statistical issues that were addressed by our statistical methods. First, our methodology needed to impute the missing death counts. Second, several states had small blacks populations which produced even smaller population sizes in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and produces abrupt changes in the estimation of mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts. Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population. The mortality rates over time within each state-age-sex-race category are modelled autoregressively, where the log mortality rate in the current year was modelled as a normal variable centred at the log mortality rate estimated from the previous year. Truncated Poisson regression was used to impute the supressed death counts, by specifying that the supressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9). These models were run separately for every combination of state, race, and sex, and produced death counts that had been smoothed over time for every cause of death within each age group.

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancies between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in every state, year, and sex.

To decompose the year- and state- specific difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref]. First, the decomposition analysis partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks. For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites. This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination. For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old. We also summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference.

In order to understand whether states fared similarly over time, we divided the time frame into three eras which roughly encapsulate an initial period of increases in life expectancy for both races and sexes (1969-1983), an intermediate period of decreases and stagnating life expectancy for blacks but not for whites (1983-1993) and a final period of increasing life expectancy for both races (1993-2013). While these eras represent different lengths of time, they best capture three distinct periods of changes to the difference in life expectancy: an intial narrowing of the gap, followed by a widening of the gap that diminished previous progress, and the most recent twenty years of sustained narrowing (eAppendix Figures 1 and 2).

**Will need to change the following**
Finally, we calculated the change in the life expectancy gap over time within each state-, race-, and sex-specific stratum. The change in the life expectancy gap is calculated by substracting the gap that occurred in the later year from the gap that occurred in the earlier year. This change in gap can be decomposed by age by taking the difference in the age-specfic contributions across the two years, and can similarly be calculated by cause.

All data analyses were performed using R version 3.3.1. The bayesian smoothing was performed using `JAGS` (version X.X), and the life expectnacy and decomposition analysis were performed using user-written functions found in the Supplement.

**Add in the stuff about credible intervals**

- sampled from posterior distribution with thinning

###Results
```{r initialize, echo=FALSE, warning=F, message=F}
library(shiny)
library(plotly)
library(viridis)
library(scales)
library(shinythemes)
library(dplyr)
library(purrr)
library(stringr)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")

age_cod_results <-rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"))

age_cod_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"))

age_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_1.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_2.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv"))

BlackWhite_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_1.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv"))

cod_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_1.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_2.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv"))
```

```{r tidy_data, echo=FALSE}

ncols <- str_split_fixed(as.character(BlackWhite_results$stratum.id), "\\.", 3)
BlackWhite_results$state <- as.factor(ncols[ , 1])
BlackWhite_results$year <- as.numeric(ncols[ , 2])
BlackWhite_results$sex <- as.factor(ncols[ , 3])

age_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_decomp_results, by = "stratum.id")
cod_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], cod_decomp_results, by = "stratum.id")
age_cod_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_cod_results, by = "stratum.id")

COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
                "Non-communicable", "Injuries", "All other causes")
cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
age_cod_results$COD <- factor(age_cod_results$COD, COD.levels)
```

```{r add_Census_Region, echo=FALSE}
BlackWhite_results$Census_Region <- "South"
BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
                                                 "Nebraska", "North Dakota", "South Dakota", 
                                                 "Illinois", "Indiana", "Michigan", "Ohio", 
                                                 "Wisconsin")] <- "Midwest"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Connecticut", "Maine", "Massachusetts", 
                                                 "New Hampshire", "Rhode Island", "Vermont",
                                                 "New Jersey", "New York", "Pennsylvania")] <- "Northeast"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Arizona", "Colorado", "Idaho", "Montana", 
                                                 "Nevada", "New Mexico", "Utah", "Wyoming",
                                                 "Alaska", "California", "Hawaii", "Oregon",
                                                 "Washington")] <- "West"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Delaware", "Florida", "Georgia", "Maryland", 
                                                  "North Carolina", "South Carolina", 
                                                  "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee", 
                                                  "Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "South Central"

BlackWhite_results$Census_Region <- factor(BlackWhite_results$Census_Region, levels = c("West", "Midwest", "South Central", "South Atlantic", "Northeast"))
```

```{r addvar_State_abbreviation, echo=FALSE}
BlackWhite_results <- BlackWhite_results %>% mutate(stabbrs = factor(state,  
                                                                     labels = c("AL", "AZ", "AR", "CA",
                                                                                "CO", "CT", "DE",
                                                                                "FL", "GA", "IL", "IN",
                                                                                "IA", "KS", "KY", "LA",
                                                                                "MD", "NY", "NC", "TX",
                                                                                "VA", "DC", "WI")))

                                      # labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
                                      #            "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", 
                                      #            "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", 
                                      #            "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", 
                                      #            "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
                                      #            "VT", "VA", "WA", "WV", "WI", "WY")))

cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, stabbrs), 
                            cod_decomp_results, by = "stratum.id", all.y = T)
```

```{r prep_era_figure_males, echo = F}
year1 <- 1969
year2 <- 1983
year3 <- 1993
year4 <- 2013

bw_dat <- BlackWhite_results %>% 
  filter(year %in% c(year1, year2, year3, year4), is.na(LE_wbgap) == F) %>%
  select(state, year, LE_wbgap, sex) %>%
  arrange(year)

bw_dat <- tidyr::spread(bw_dat, year, LE_wbgap) 
names(bw_dat)[3:6] <- c("first.gap","second.gap", "third.gap", "fourth.gap")  
bw_dat.males <- bw_dat %>% filter(sex == "Male") %>%
  mutate(era1.diff = first.gap - second.gap,
         era2.diff = second.gap - third.gap,
         era3.diff = third.gap - fourth.gap,
         state.gdiff.order.1 = reorder(state, first.gap),
         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
         state.gdiff.order.2 = reorder(state, second.gap),
         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
         state.gdiff.order.3 = reorder(state, third.gap),
         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
  )

bw_dat.females <- bw_dat %>% filter(sex == "Female") %>%
  mutate(era1.diff = first.gap - second.gap,
         era2.diff = second.gap - third.gap,
         era3.diff = third.gap - fourth.gap,
         state.gdiff.order.1 = reorder(state, first.gap),
         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
         state.gdiff.order.2 = reorder(state, second.gap),
         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
         state.gdiff.order.3 = reorder(state, third.gap),
         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
  )

cod.df <- data.frame(state = factor(), sex = factor(), era = character(), 
                     COD = factor(), Contribution.to.change = numeric(), 
                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())

age.df <- data.frame(state = factor(), sex = factor(), era = character(), 
                     age_minbin = numeric(), Contribution.to.change = numeric(), 
                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())

years <- c(1969, 1983, 1993, 2013)

for (state.i in levels(BlackWhite_results$state)) {
  for (sex.i in c("Male", "Female")) {
    for (year.i in 1:3) {

      table1 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(COD)
      table2 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(COD)

      table3 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(age_minbin)
      table4 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(age_minbin)

      temp1 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
                          contribution.to.gap.change(type.of.decomp = "COD", 
                                                     decomp.table1 = table1, 
                                                     decomp.table2 = table2, 
                                                     type.of.data = "results"))

      temp2 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
                          contribution.to.gap.change(type.of.decomp = "Age", 
                                                     decomp.table1 = table3, 
                                                     decomp.table2 = table4,
                                                     type.of.data = "results"))

      cod.df <- rbind(cod.df, temp1)
      age.df <- rbind(age.df, temp2)
    }
  }
}

```

```{r more_figure_prep, echo = F}

cod.df <- make_pretty_decomp_plot(decomp.table = cod.df, strat.var.1 = "state", 
                        strat.var.2 = "sex", strat.var.3 = "era", 
                        sign.var = "narrowed_gap",
                        decomp.var = "Contribution.to.change",
                        decomp.var.prop = "Contrib.to.change.prop", 
                        partition.bar.var = "COD", type.of.data = "results")

age.df <- make_pretty_decomp_plot(decomp.table = age.df, strat.var.1 = "state", 
                        strat.var.2 = "sex", strat.var.3 = "era", 
                        sign.var = "narrowed_gap",
                        decomp.var = "Contribution.to.change",
                        decomp.var.prop = "Contrib.to.change.prop", 
                        partition.bar.var = "Ages", type.of.data = "results")

cod.df <- merge(cod.df, BlackWhite_results[BlackWhite_results$year == 1969 & BlackWhite_results$sex == "Male", 
                                           c("state", "Census_Region", "stabbrs")], by = c("state"))
```

*Changes to the differency in life expectancy in US males*

**Figure 1a** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs. their gap in 1983 (solid dashed line), with the change in the gap being represented by the number of years between these lines. The decomposition (in years) of the change in the gap by cause of death is also shown for each state.

```{r, echo = F}
sum1 <- BlackWhite_results %>% filter(year == 1969 & sex == "Male") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)

diff <- bw_dat.males %>% arrange(era1.diff)
```

What we see so far (based on these 22 states):

- The gap in 1969 ranges from `r round(sum1[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `.

- For most states, the gap decreases between 1969 and 1983, with the largest decrease found in `r diff[dim(diff)[1], ]$state` (change in gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)` (all other causes), 
`r diff[dim(diff)[1]-1, ]$state` (largely due to injuries), `r diff[dim(diff)[1]-2, ]$state` (injuries), `r diff[dim(diff)[1]-3, ]$state` (injuries), and `r diff[dim(diff)[1]-4, ]$state` (injuries).

- The gap increases for some states. Most notably in DC (CVD and other causes), Wisconsin (CVD), California (CVD). Also, for these states injuries do not contribute as much to narrowing the gap, so their gap increasing appears to be due to a combination of changes to CVD making blacks relatively worse off and less changes to injuries than experienced by other states.

- Injuries across the board contribute to narrowing the gap over time.

- Communicable disease narrows (except arizona), but much smaller contribution than injuries.

- Generally, all other causes narrows, again much smaller contribution than injuries.

- Generally, CVD increase gap.

- Cancer increases the gap everywhere. Size of contribution to change is a little smaller than CVD.

- Generally, non-communicable increase the gap but the contribution is tiny. Notable diff is in NYC and MD it decrease gap.

**Figure 1a: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**

```{r figure1, fig.align = "center", echo=F}

cod.df.males <- merge(subset(cod.df, sex == "Male"), bw_dat.males, by = c("state"))
cod.df.males <- cod.df.males %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)

y.labels <- levels(factor(cod.df.males$state.gdiff.order.1))
y.num.breaks <- length(levels(factor(cod.df.males$state.gdiff.order.1)))

males.era1 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = COD), 
              color = "white") +
#    facet_grid(.~ Census_Region) +
    geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
        geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
   # geom_segment(data = bw_dat.males,
  #               aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
  #  geom_segment(data = bw_dat.males,
  #               aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era1
```

```{r, echo =F}
sum2 <- BlackWhite_results %>% filter(year == 1983 & sex == "Male") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
```

**Figure 1b** displays each state's male life expectancy gap in 1983 (vertical dashed line) vs. their gap in 1993 (solid dashed line).

- Be careful with the interpretation: the x-axis is much wider for this plot because of DC. (Zoom into same x range to aid interpretation at first)

- The gap in 1983 ranges from `r round(sum2[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum2[1, ]$LE_wbgap_lcl, 1)`, `r round(sum2[1, ]$LE_wbgap_ucl, 1)`] in `r sum2[1, ]$state ` to `r round(sum2[dim(sum2)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum2[dim(sum2)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum2[dim(sum2)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum2[dim(sum2)[1], ]$state `.

- Except for Arizona and Kentucky, the life expectancy gap increases in each state, with the biggest driver being communicable disease (HIV/AIDS), especially in MD, NY, and CT. Injuries also contribute to the gap widening over this time frame, most notably in Wisconsin, MD, and LA. Thus, compared to previous period, these two causes of death led to a widening rather than a narrowing of the gap.

- Similar to the firts time frame, CVD and cancer also contributed to widening the gap over time. Also similar is the minimal effect of  non-communicable disease on the gap, while all other causes contributes to narrowing the gap, with greatest contribution to narrowing in NY, CT, and FL.

- DC was especially hard hit, with injuries driving the increase in the gap from 10.5 years to 15.9 years in this 10-year period.

**Figure 1b: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1983 and 1993 by US state**

```{r figure1b, echo = F}
y.labels.2 <- levels(factor(cod.df.males$state.gdiff.order.2))
y.num.breaks.2 <- length(levels(factor(cod.df.males$state.gdiff.order.2)))

males.era2 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
                  ymax = state.gdiff.num.2 + 0.45, 
                  xmax = new.finish.2, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era2 
```

```{r, echo =F}
sum3 <- BlackWhite_results %>% filter(year == 1993 & sex == "Male") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)

sum4 <- BlackWhite_results %>% filter(year == 2013 & sex == "Male") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
```

- The gap in 1993 ranges from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state `.

-By 2013, it ranges from `r round(sum4[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum4[dim(sum4)[1], ]$state `.

- After 1993, the gap narrowed for all states shown, and in most cases all of the causes contributed to the narrowing of the gap. Notice again that in Wisconsin and DC, changes to CVD mortality counteract the gap narrowing, while in Arizona changes to communicable disease counteract the narrowing.

**Figure 1c: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1993 and 2013 by US state**

```{r figure1c, echo = F}
y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
y.num.breaks.3 <- length(levels(factor(cod.df.males$state.gdiff.order.3)))

males.era3 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
                  ymax = state.gdiff.num.3 + 0.45, 
                  xmax = new.finish.3, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))

males.era3
```

```{r, fig.align="center", fig.height=5, fig.width=9, echo=F, eval = F}
subplot(males.era1, males.era2, males.era3, shareY = T)
```
** **
** **

```{r, echo =F}
sum5 <- BlackWhite_results %>% filter(year == 1969 & sex == "Female") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
```

*Changes to the differency in life expectancy in US females*

Quick summary

- Most states saw decreasing gaps in first era. Generally, all of the causes, except for cancer, were associated with changes that relatively benefitted blacks, with the biggest contributor to the narrowing gap being CVD and all other causes. However, in Wisconsin and CA, CVD led to a widening gap.

- During the second era, the gap stayed approx. the same or widened across most states. Delaware, Kentucky, Kansas, Alabama, were notable exceptions were the gap narrowed -- didn't see effect of communicable disease like other states on the gap increasing (NY, FL, DC, MD, CT, hardest hit).

- In women, changes in injury related deaths doesn't play a big role (except DC).

- CVD differential by state. Contributes to widening the gap in CA and DC, for example, and narrowing it in KS and Kentucky.

- All other causes narrows across states (except DC, Indiana). Cancer increases the gap by a bit.

- During the third era, in-line with males, the life expectancy gap decreased during the last 20 year period, and generally speaking all the causes contributed to the gap narrowing over time. The biggest contributors were CVD and all other causes.

**Figure 2a: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**

```{r figure2a, fig.align = "center", echo=F}

cod.df.females <- merge(subset(cod.df, sex == "Female"), bw_dat.females, by = c("state"))
cod.df.females <- cod.df.females %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)

y.labels <- levels(factor(cod.df.females$state.gdiff.order.1))
y.num.breaks <- length(levels(factor(cod.df.females$state.gdiff.order.1)))

females.era1 <- ggplotly(
  ggplot(subset(cod.df.females, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.females,
                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
    geom_segment(data = bw_dat.females,
                 aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

females.era1
```

**Figure 2b: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1983 and 1993 by US state**

```{r figure2b, fig.align = "center", echo=F}
y.labels.2 <- levels(factor(cod.df.females$state.gdiff.order.2))
y.num.breaks.2 <- length(levels(factor(cod.df.females$state.gdiff.order.2)))

females.era2 <- ggplotly(
  ggplot(subset(cod.df.females, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
                  ymax = state.gdiff.num.2 + 0.45, 
                  xmax = new.finish.2, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.females,
                 aes(x = third.gap, xend = third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
    geom_segment(data = bw_dat.females,
                 aes(x = second.gap, xend = second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1983-1993"), yaxis = list(title = NA, autorange = "reversed"))

females.era2
```

**Figure 2c: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1993 and 2013 by US state**

```{r figure2c, fig.align = "center", echo=F}
y.labels.3 <- levels(factor(cod.df.females$state.gdiff.order.3))
y.num.breaks.3 <- length(levels(factor(cod.df.females$state.gdiff.order.3)))

females.era3 <- ggplotly(
  ggplot(subset(cod.df.females, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
                  ymax = state.gdiff.num.3 + 0.45, 
                  xmax = new.finish.3, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.females,
                 aes(x = fourth.gap, xend = fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
    geom_segment(data = bw_dat.females,
                 aes(x = third.gap, xend = third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))

females.era3
```

**Figure 5a: Trends in life expectancy for white males between 1969 and 2013 by US state**

```{r figure5a, fig.align="center", fig.height=4, fig.width=11, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_white, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_ribbon(aes(ymin = LE_white_lcl, ymax = LE_white_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_grid(. ~ Census_Region) +
           ylab("Life expectancy (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5))

```

**Figure 5b: Trends in life expectancy for black males between 1969 and 2013 by US state**

```{r figure5b, fig.align="center", fig.height=4, fig.width=11, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_black, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_grid(. ~ Census_Region) +
           ylab("Life expectancy (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5))

```

**Figure 5c: Trends in the black-white life expectancy gap for males between 1969 and 2013 by US state**

```{r figure5c, fig.align="center", fig.height=4, fig.width=11, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_wbgap, x = year)) + 
           geom_vline(aes(xintercept = 1983), lty = 2, lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), lty = 2, lwd = 0.5) +
           geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_grid(. ~ Census_Region) +
           ylab("Black-white life expectancy gap (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), aes(label = stabbrs), check_overlap = T, size = 2.5))

```

**Figure 6a: Contribution of cancer to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
```{r figure6a, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cancers"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(.~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cancers"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 6a (ii): Contribution of cancer (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state**
```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cancers"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(.~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cancers"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 6b: Contribution of cardiovascular disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**

```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(.~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cardiovascular"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 6b (ii): Contribution of cardiovascular disease (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**

```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cardiovascular"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 6c: Contribution of communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**

```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Communicable"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 6c (ii): Contribution of communicable disease (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**

```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Communicable"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 6d: Contribution of non-communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**

```{r, fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Non-communicable"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Non-communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 6d (ii): Contribution of non-communicable disease (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**

```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Non-communicable"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Non-communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 6e: Contribution of injuries to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**

```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Injuries"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 6e (ii): Contribution of injuries (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**

```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Injuries"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 6f: Contribution of all other causes to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "All other causes"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "All other causes"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 6f (ii): Contribution of all other causes (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**

```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "All other causes"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "All other causes"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 7a: Trends in life expectancy for white females between 1969 and 2013 by US state**

```{r, fig.align="center", fig.height=4, fig.width=11, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_white, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_ribbon(aes(ymin = LE_white_lcl, ymax = LE_white_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_grid(. ~ Census_Region) +
           ylab("Life expectancy (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5))
```

**Figure 7b: Trends in life expectancy for black females between 1969 and 2013 by US state**

```{r, fig.align="center", fig.height=4, fig.width=11, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_black, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_grid(. ~ Census_Region) +
           ylab("Life expectancy (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5))

```

**Figure 7c: Trends in the black-white life expectancy gap for females between 1969 and 2013 by US state**

```{r figure6, fig.align="center", fig.height=4, fig.width=11, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_wbgap, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_grid(. ~ Census_Region) +
           ylab("Black-white life expectancy gap (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), aes(label = stabbrs), check_overlap = T, size = 2.5) 
) 
```

**Figure 8a: Contribution of cancer to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**

```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cancers"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cancers"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8a: Contribution of cancer (%) to the black-white life expectancy gap for females between 1969 and 2013 by US state**

```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cancers"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cancers"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8b: Contribution of cardiovascular disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r,  fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cardiovascular"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cardiovascular"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8c: Contribution of communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r,  fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Communicable"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Communicable"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8d: Contribution of non-communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r,  fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Non-communicable"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Non-communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Non-communicable"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Non-communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8e: Contribution of injuries to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r,  fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Injuries"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Injuries"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8f: Contribution of all other causes to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "All other causes"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "All other causes"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "All other causes"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "All other causes"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

###Supplemental Information

**eTable1: Cause of Death Recode**

| Broad cause of death     | Included causes of death                                                     |
|--------------------------|------------------------------------------------------------------------------|
| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer. |
| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries. | 
| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Sydrome and Nephrosis.  |
| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death. |

**eFigure 1: Difference in life expectancy between black in white males, United States, 1969-2013**

![Difference in life expectancy between black in white males, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Men.bmp)

**eFigure 2: Difference in life expectancy between black in white females, United States, 1969-2013**

![Difference in life expectancy between black in white females, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Women.bmp)

**Background notes from meeting with SH on Dec 20**

- CVD declining starting 1968 - see paper Ruel Stallones - cited in SH JAMA paper - history of decline in CVD.

- Richard cooper also written about BW diff in CVD with epi perspective

- for men there is stagnation from 1969 to late 1980s and then increase in gap with HIV and homicides

- massive crime wave in 1980s, black LE went down from 1984 to 1987 - period when both HIV and homicide went up -- the whole story during that time was HIV and homicide

- heterogeneity between states in the contribution of the causes to the decline in the gap - maybe need to break up into at least two time periods with split at 1990.

- this is where the new figures come in which on the y-axis have the contribution to LE gap (in years or %) vs year on the x axis, and geom_line for each COD (do one COD at a time), and individual line for each state (may facet by census region).

- then have a view of how the different COD have contributed to the decline in the gap over time.

### 2017-02-05 Code/run-models.Rmd
set 6 results

--- 2017-02-03 Code/run-models.Rmd
@@
     #for(state_i in c("Florida", "Georgia", "Illinois", "Indiana")) { #saved as results_set_3  
-    for(state_i in c("Iowa", "Kansas", "Kentucky", "Louisiana")) { #saved as results_set_4  
+    #for(state_i in c("Iowa", "Kansas", "Kentucky", "Louisiana")) { #saved as results_set_4  
+    # for(state_i in c("Maryland", "New York", "North Carolina", "Texas", "Virginia", "Wisconsin" )) { #saved as results_set_5  
+    for(state_i in c("Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Nebraska" )) { #saved as results_set_6  
       print(state_i)
@@
 #                  r.White.Florida.Female, r.White.Georgia.Female, r.White.Illinois.Female, r.White.Indiana.Female)
-r.White <- rbind(r.White.Iowa.Male, r.White.Kansas.Male, r.White.Kentucky.Male, r.White.Louisiana.Male,
-                 r.White.Iowa.Female, r.White.Kansas.Female, r.White.Kentucky.Female, r.White.Louisiana.Female)
+# r.White <- rbind(r.White.Iowa.Male, r.White.Kansas.Male, r.White.Kentucky.Male, r.White.Louisiana.Male,
+#                  r.White.Iowa.Female, r.White.Kansas.Female, r.White.Kentucky.Female, r.White.Louisiana.Female)
+# r.White <- rbind(r.White.Maryland.Male, r.White.NewYork.Male, r.White.NorthCarolina.Male, r.White.Texas.Male, 
+#                  r.White.Virginia.Male, r.White.Wisconsin.Male, 
+#                  r.White.Maryland.Female, r.White.NewYork.Female, r.White.NorthCarolina.Female, r.White.Texas.Female, 
+#                  r.White.Virginia.Female, r.White.Wisconsin.Female)
+r.White <- rbind(r.White.Massachusetts.Male, r.White.Michigan.Male, r.White.Minnesota.Male, r.White.Mississippi.Male, 
+                 r.White.Missouri.Male, r.White.Nebraska.Male, 
+                 r.White.Massachusetts.Female, r.White.Michigan.Female, r.White.Minnesota.Female, r.White.Mississippi.Female, 
+                 r.White.Missouri.Female, r.White.Nebraska.Female)
 r.White$race <- "White"
@@
 #                  r.Black.AlabamaFemale, r.Black.ArizonaFemale, r.Black.ArkansasFemale, r.Black.CaliforniaFemale)
-r.Black <- rbind(r.Black.Iowa.Male, r.Black.Kansas.Male, r.Black.Kentucky.Male, r.Black.Louisiana.Male,
-                 r.Black.Iowa.Female, r.Black.Kansas.Female, r.Black.Kentucky.Female, r.Black.Louisiana.Female)
+r.Black <- rbind(r.Black.Massachusetts.Male, r.Black.Michigan.Male, r.Black.Minnesota.Male, r.Black.Mississippi.Male, 
+                 r.Black.Missouri.Male, r.Black.Nebraska.Male, 
+                 r.Black.Massachusetts.Female, r.Black.Michigan.Female, r.Black.Minnesota.Female, r.Black.Mississippi.Female, 
+                 r.Black.Missouri.Female, r.Black.Nebraska.Female)
 r.Black$race <- "Black"
@@
 #SET 4: Iowa, Kansas, Kentucky, Louisiana
-save(r.All, le.calculations, file = "~/BW_results/results_set_4.Rdata") 
-write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv")
-write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv")
-write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv")
-write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_4.csv")
+# save(r.All, le.calculations, file = "~/BW_results/results_set_4.Rdata") 
+# write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv")
+# write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv")
+# write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv")
+# write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_4.csv")
 
+#SET 5: Maryland, New York, North Carolina, Texas, Virginia, Wisconsin
+# save(r.All, le.calculations, file = "~/BW_results/results_set_5.Rdata") 
+# write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv")
+# write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv")
+# write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv")
+# write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_5.csv")
+
+#SET 6: "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Nebraska"
+save(r.All, le.calculations, file = "~/BW_results/results_set_6.Rdata") 
+write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_6.csv")
+write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_6.csv")
+write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_6.csv")
+write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_6.csv")
 ```

### 2017-02-06 Code/run-models.Rmd

--- 2017-02-03 Code/run-models.Rmd
@@
     #for(state_i in c("Florida", "Georgia", "Illinois", "Indiana")) { #saved as results_set_3  
-    for(state_i in c("Iowa", "Kansas", "Kentucky", "Louisiana")) { #saved as results_set_4  
+    #for(state_i in c("Iowa", "Kansas", "Kentucky", "Louisiana")) { #saved as results_set_4  
+    # for(state_i in c("Maryland", "New York", "North Carolina", "Texas", "Virginia", "Wisconsin" )) { #saved as results_set_5  
+    #for(state_i in c("Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Nebraska" )) { #saved as results_set_6  
+    for(state_i in c("Nevada", "New Jersey", "New Mexico", "Ohio", "Oklahoma", "Oregon" )) { #saved as results_set_7  
       print(state_i)
@@
 #                  r.White.Florida.Female, r.White.Georgia.Female, r.White.Illinois.Female, r.White.Indiana.Female)
-r.White <- rbind(r.White.Iowa.Male, r.White.Kansas.Male, r.White.Kentucky.Male, r.White.Louisiana.Male,
-                 r.White.Iowa.Female, r.White.Kansas.Female, r.White.Kentucky.Female, r.White.Louisiana.Female)
+# r.White <- rbind(r.White.Iowa.Male, r.White.Kansas.Male, r.White.Kentucky.Male, r.White.Louisiana.Male,
+#                  r.White.Iowa.Female, r.White.Kansas.Female, r.White.Kentucky.Female, r.White.Louisiana.Female)
+# r.White <- rbind(r.White.Maryland.Male, r.White.NewYork.Male, r.White.NorthCarolina.Male, r.White.Texas.Male, 
+#                  r.White.Virginia.Male, r.White.Wisconsin.Male, 
+#                  r.White.Maryland.Female, r.White.NewYork.Female, r.White.NorthCarolina.Female, r.White.Texas.Female, 
+#                  r.White.Virginia.Female, r.White.Wisconsin.Female)
+
+r.White <- rbind(r.White.Nevada.Male, r.White.NewJersey.Male, r.White.NewMexico.Male, r.White.Ohio.Male, 
+                 r.White.Oklahoma.Male, r.White.Oregon.Male, 
+                 r.White.Nevada.Female, r.White.NewJersey.Female, r.White.NewMexico.Female, r.White.Ohio.Female, 
+                 r.White.Oklahoma.Female, r.White.Oregon.Female)
 r.White$race <- "White"
@@
 #                  r.Black.AlabamaFemale, r.Black.ArizonaFemale, r.Black.ArkansasFemale, r.Black.CaliforniaFemale)
-r.Black <- rbind(r.Black.Iowa.Male, r.Black.Kansas.Male, r.Black.Kentucky.Male, r.Black.Louisiana.Male,
-                 r.Black.Iowa.Female, r.Black.Kansas.Female, r.Black.Kentucky.Female, r.Black.Louisiana.Female)
+r.Black <- rbind(r.Black.Nevada.Male, r.Black.NewJersey.Male, r.Black.NewMexico.Male, r.Black.Ohio.Male, 
+                 r.Black.Oklahoma.Male, r.Black.Oregon.Male, 
+                 r.Black.Nevada.Female, r.Black.NewJersey.Female, r.Black.NewMexico.Female, r.Black.Ohio.Female, 
+                 r.Black.Oklahoma.Female, r.Black.Oregon.Female)
 r.Black$race <- "Black"
@@
 #SET 4: Iowa, Kansas, Kentucky, Louisiana
-save(r.All, le.calculations, file = "~/BW_results/results_set_4.Rdata") 
-write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv")
-write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv")
-write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv")
-write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_4.csv")
+# save(r.All, le.calculations, file = "~/BW_results/results_set_4.Rdata") 
+# write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv")
+# write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv")
+# write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv")
+# write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_4.csv")
 
+#SET 5: Maryland, New York, North Carolina, Texas, Virginia, Wisconsin
+# save(r.All, le.calculations, file = "~/BW_results/results_set_5.Rdata") 
+# write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv")
+# write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv")
+# write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv")
+# write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_5.csv")
+
+#SET 6: "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Nebraska"
+# save(r.All, le.calculations, file = "~/BW_results/results_set_6.Rdata") 
+# write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_6.csv")
+# write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_6.csv")
+# write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_6.csv")
+# write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_6.csv")
+
+#SET 7: "Nevada", "New Jersey", "New Mexico", "Ohio", "Oklahoma", "Oregon"
+save(r.All, le.calculations, file = "~/BW_results/results_set_7.Rdata") 
+write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_7.csv")
+write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_7.csv")
+write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_7.csv")
+write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_7.csv")
 ```

### 2017-02-06 Manuscript/Manuscript_using_actual_data.Rmd

@@
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"))
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"),
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"))
 
@@
                          read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"))
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"))
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_6.csv"))
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_6.csv"))
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_6.csv"))
 ```
@@
                                                                                 "IA", "KS", "KY", "LA",
-                                                                                "MD", "NY", "NC", "TX",
+                                                                                "MD", "MA", "MI", "MN", 
+                                                                                "MS", "MO", "NE", "NY", 
+                                                                                "NC", "TX",
                                                                                 "VA", "DC", "WI")))

### 2017-02-06 Manuscript/Manuscript_using_actual_data.Rmd
add census division

--- 2017-02-03 Manuscript/Manuscript_using_actual_data.Rmd
@@
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"))
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"),
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"),
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv"))
 
@@
                          read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"))
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv"))
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_6.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_7.csv"))
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_6.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_7.csv"))
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_6.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_7.csv"))
 ```
@@
 
+```{r add_Census_Division, echo = F}
+BlackWhite_results$Census_Division <- NA
+
+BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Connecticut", "Maine", "Massachusetts", 
+                                                                   "New Hampshire", "Rhode Island", "Vermont")] <- "New England"
+
+BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("New Jersey", "New York", "Pennsylvania")] <- "Mid-Atlantic"
+
+BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Illinois", "Indiana", "Michigan", "Ohio", 
+                                                                 "Wisconsin")] <- "East North Central"
+
+BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
+                                                                 "Nebraska", "North Dakota", "South Dakota")] <- "West North Central"
+
+BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Delaware", "Florida", "Georgia", "Maryland", 
+                                                                 "North Carolina", "South Carolina", 
+                                                                 "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"
+
+BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee")] <- "East South Central"
+
+BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "West South Central"
+
+BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Arizona", "Colorado", "Idaho", "Montana", 
+                                                                 "Nevada", "New Mexico", "Utah", "Wyoming")] <- "Mountain"
+
+BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Alaska", "California", "Hawaii", "Oregon","Washington")] <- "Pacific"
+
+```
+
 ```{r addvar_State_abbreviation, echo=FALSE}
@@
                                                                                 "IA", "KS", "KY", "LA",
-                                                                                "MD", "NY", "NC", "TX",
+                                                                                "MD", "MA", "MI", "MN", 
+                                                                                "MS", "MO", "NE", 
+                                                                                "NV", "NJ", "NM", 
+                                                                                "NY", 
+                                                                                "NC", "OH", "OK", "OR",
+                                                                                "TX",
                                                                                 "VA", "DC", "WI")))

### 2017-02-06 Manuscript/Manuscript_using_actual_data.Rmd
graphs by census division

--- 2017-02-03 Manuscript/Manuscript_using_actual_data.Rmd
@@
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"))
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"),
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"),
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv")#,
+                        #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
+                        )
 
@@
                          read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"))
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv")#,
+                         #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
+                         )
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_6.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_7.csv")#,
+                            #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_8.csv")
+                            )
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_6.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_7.csv")#,
+                            #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_8.csv")
+                            )
 
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv"))
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_6.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_7.csv")#,
+                            #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_8.csv")
+                            )
 ```
@@
 ```{r add_Census_Region, echo=FALSE}
+#note that the Census Region of the South is partitioned into two for better display.
+
 BlackWhite_results$Census_Region <- "South"
@@
 
+```{r add_Census_Division, ncol = 3, echo = F}
+BlackWhite_results$Census_Division <- NA
+
+BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Connecticut", "Maine", "Massachusetts", 
+                                                                   "New Hampshire", "Rhode Island", "Vermont")] <- "New England"
+
+BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("New Jersey", "New York", "Pennsylvania")] <- "Mid-Atlantic"
+
+BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Illinois", "Indiana", "Michigan", "Ohio", 
+                                                                 "Wisconsin")] <- "East North Central"
+
+BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
+                                                                 "Nebraska", "North Dakota", "South Dakota")] <- "West North Central"
+
+BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Delaware", "Florida", "Georgia", "Maryland", 
+                                                                 "North Carolina", "South Carolina", 
+                                                                 "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"
+
+BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee")] <- "East South Central"
+
+BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "West South Central"
+
+BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Arizona", "Colorado", "Idaho", "Montana", 
+                                                                 "Nevada", "New Mexico", "Utah", "Wyoming")] <- "Mountain"
+
+BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Alaska", "California", "Hawaii", "Oregon","Washington")] <- "Pacific"
+
+BlackWhite_results$Census_Division <- factor(BlackWhite_results$Census_Division, levels = c("New England", "Mid-Atlantic", "East North Central", "West North Central", "South Atlantic", "East South Central", "West South Central", "Mountain", "Pacific"))
+
+```
+
 ```{r addvar_State_abbreviation, echo=FALSE}
@@
                                                                      labels = c("AL", "AZ", "AR", "CA",
-                                                                                "CO", "CT", "DE",
-                                                                                "FL", "GA", "IL", "IN",
-                                                                                "IA", "KS", "KY", "LA",
-                                                                                "MD", "NY", "NC", "TX",
-                                                                                "VA", "DC", "WI")))
+                                                                                "CO", "CT", "DE", "FL",
+                                                                                "GA", "IL", "IN", "IA",
+                                                                                "KS", "KY", "LA", "MD",
+                                                                                "MA", "MI", "MN", "MS",
+                                                                                "MO", "NE", "NV", "NJ", 
+                                                                                "NM", "NY", "NC", "OH", 
+                                                                                "OK", "OR", "TX", "VA", 
+                                                                                "DC", "WI")))
+#set 8:
+#"PA", "RI", "SC", "TN", "WA", "WV"
 
@@
 
-*Changes to the differency in life expectancy in US males*
+*Changes to the difference in life expectancy in US males*
 
@@
 
-*Changes to the differency in life expectancy in US females*
+*Changes to the difference in life expectancy in US females*
 
@@
            geom_line(aes(col = state)) +
-           facet_grid(. ~ Census_Region) +
+           facet_wrap( ~ Census_Division, ncol = 3) +
            ylab("Life expectancy (years)") +
@@
            geom_line(aes(col = state)) +
-           facet_grid(. ~ Census_Region) +
+           facet_wrap( ~ Census_Division, ncol = 3) +
            ylab("Black-white life expectancy gap (years)") +
@@
            geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_grid(.~ Census_Region) +
+           geom_line(aes(col = state)) + facet_wrap(~ Census_Division, ncol = 3) +
            ylab("Contribution to LE Gap (years)") +
@@
            geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_grid(.~ Census_Region) +
+           geom_line(aes(col = state)) + facet_wrap(~ Census_Division, ncol = 3) +
            ylab("Contribution to LE Gap (%)") +
@@
            geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
+           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
            ylab("Contribution to LE Gap (%)") +
@@
            geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_grid(. ~ Census_Region) +
+           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
            ylab("Contribution to LE Gap (years)") +

### 2017-02-06 Manuscript/Manuscript_using_actual_data.Rmd
updated interpretation with new results

@@
 
-cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, stabbrs), 
+cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, Census_Division, stabbrs), 
                             cod_decomp_results, by = "stratum.id", all.y = T)
@@
 cod.df <- merge(cod.df, BlackWhite_results[BlackWhite_results$year == 1969 & BlackWhite_results$sex == "Male", 
-                                           c("state", "Census_Region", "stabbrs")], by = c("state"))
+                                           c("state", "Census_Region", "Census_Division", "stabbrs")], by = c("state"))
 ```
@@
 diff <- bw_dat.males %>% arrange(era1.diff)
-```
-
-What we see so far (based on these 22 states):
+
+max1 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1], ]$state & cod.df$era == "1969-1983"])
+mc1 <- cod.df$COD[cod.df$state == diff[dim(diff)[1], ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max1]
+
+max2 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-1, ]$state & cod.df$era == "1969-1983"])
+mc2 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-1, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max2]
+
+max3 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-2, ]$state & cod.df$era == "1969-1983"])
+mc3 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-2, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max3]
+
+max4 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-3, ]$state & cod.df$era == "1969-1983"])
+mc4 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-3, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max4]
+
+max5 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-4, ]$state & cod.df$era == "1969-1983"])
+mc5 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-4, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max5]
+```
+
+What we see so far (based on these `r dim(sum1)[1]` states):
 
@@
 
-- For most states, the gap decreases between 1969 and 1983, with the largest decrease found in `r diff[dim(diff)[1], ]$state` (change in gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)` (all other causes),
-`r diff[dim(diff)[1]-1, ]$state` (largely due to injuries), `r diff[dim(diff)[1]-2, ]$state` (injuries), `r diff[dim(diff)[1]-3, ]$state` (injuries), and `r diff[dim(diff)[1]-4, ]$state` (injuries).
+- For most states, the gap decreases between 1969 and 1983, with the largest decrease found in `r diff[dim(diff)[1], ]$state` (change in gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)`, `r round(max1,1)` of which is due to `r mc1`,
+`r diff[dim(diff)[1]-1, ]$state` (`r round(max2,2)` due to `r mc2`), `r diff[dim(diff)[1]-2, ]$state` (`r round(max3,1)` due to `r mc3`), `r diff[dim(diff)[1]-3, ]$state` (`r round(max4,1)` due to `r mc4`), and `r diff[dim(diff)[1]-4, ]$state` (`r round(max5,1)` due to `r mc5`).
 
 - The gap increases for some states.
-Most notably in DC (CVD and other causes), Wisconsin (CVD), California (CVD).
-Also, for these states injuries do not contribute as much to narrowing the gap, so their gap increasing appears to be due to a combination of changes to CVD making blacks relatively worse off and less changes to injuries than experienced by other states.
-
-- Injuries across the board contribute to narrowing the gap over time.
-
-- Communicable disease narrows (except arizona), but much smaller contribution than injuries.
+Most notably in DC (CVD and other causes), Minnesota (CVD), Wisconsin (CVD), California (CVD).
+Also, for these states injuries do not contribute as much to narrowing the gap (and in Minnesota, injuries widens the gap), so their gap increasing appears to be due to a combination of changes to CVD making blacks relatively worse off and less changes to injuries than experienced by other states.
+
+- Injuries across the board contribute to narrowing the gap over time, except for Minnesota.
+
+- Communicable disease narrows (except Arizona), but much smaller contribution than injuries.
 
@@
 
-- Generally, CVD increase gap.
-
-- Cancer increases the gap everywhere. Size of contribution to change is a little smaller than CVD.
-
-- Generally, non-communicable increase the gap but the contribution is tiny. Notable diff is in NYC and MD it decrease gap.
+- Generally, CVD increases gap.
+
+- Cancer increases the gap everywhere except NM. Size of contribution to change is a little smaller than CVD.
+
+- Generally, non-communicable disease increases the gap but the contribution is tiny. Notable differences in Massachusetts, New York, and Maryland where it decreases gap.
 
@@
 
-- Except for Arizona and Kentucky, the life expectancy gap increases in each state, with the biggest driver being communicable disease (HIV/AIDS), especially in MD, NY, and CT.
+- Except for New Mexico, Arizona, and Kentucky, the life expectancy gap increases in each state, with the biggest driver being communicable disease (HIV/AIDS), especially in MD, NY, NJ, and CT.
 Injuries also contribute to the gap widening over this time frame, most notably in Wisconsin, MD, and LA.
@@
 
-- Similar to the firts time frame, CVD and cancer also contributed to widening the gap over time.
-Also similar is the minimal effect of  non-communicable disease on the gap, while all other causes contributes to narrowing the gap, with greatest contribution to narrowing in NY, CT, and FL.
-
-- DC was especially hard hit, with injuries driving the increase in the gap from 10.5 years to 15.9 years in this 10-year period.
+- Similar to the firts time frame, CVD and cancer also contribute to widening the gap over time.
+Also similar is the minimal effect of  non-communicable disease on the gap, while all other causes contributes to narrowing the gap, with greatest contribution to narrowing in NY, CT, Mississippi, and FL.
+
+- DC was especially hard hit, with injuries driving the increase in the gap from 10.5 years to 15.9 years in this 10-year period, an increase of X, Y of which can be attributed to injuries.
 
@@
 
-- The gap in 1993 ranges from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state `.
-
--By 2013, it ranges from `r round(sum4[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum4[dim(sum4)[1], ]$state `.
+- The gap in 1993 ranges from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state ` (or `r round(sum3[dim(sum3)[1]-1, ]$LE_wbgap, 1)` when DC is excluded).
+
+- By 2013, it ranges from `r round(sum4[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum4[dim(sum4)[1], ]$state ` (or `r round(sum4[dim(sum4)[1]-1, ]$LE_wbgap, 1)` when DC is excluded).
 
@@
 - Most states saw decreasing gaps in first era.
-Generally, all of the causes, except for cancer, were associated with changes that relatively benefitted blacks, with the biggest contributor to the narrowing gap being CVD and all other causes.
-However, in Wisconsin and CA, CVD led to a widening gap.
+
+- However, increases were found in Arizona, Oregon, Minnesota, and Wisconsin, largely due to cardiovascular disease. Thus, the effect of CVD on the LE varied by state over this time period in women.
+
+- Otherwise, most of the other causes contributed to the gap narrowing, although cancer was associated with narrowing in some states (MS, NM, Oregon, for example) and increases in others (OH, Missouri, Kentucky).
+
+- The biggest contributor to the narrowing gap being CVD (in most states, although heterogeniety) and all other causes.
 
@@
 
-- CVD differential by state. Contributes to widening the gap in CA and DC, for example, and narrowing it in KS and Kentucky.
-
-- All other causes narrows across states (except DC, Indiana). Cancer increases the gap by a bit.
+- CVD differential by state. Contributes to widening the gap in CA, Mississippi, and DC, for example, and narrowing it in Kansas, Delaware, and Kentucky.
+
+- All other causes narrows across states (except DC, Indiana). Cancer increases the gap by a bit (except Oregon and Minnesota).
 
@@
 The biggest contributors were CVD and all other causes.
+
+- In a few states, Cancer contributed to widening the gap (NM, Oregon, Arizona, DC).
 

### 2017-02-07 Code/run-models.Rmd
set 8 results

@@
     #for(state_i in c("Iowa", "Kansas", "Kentucky", "Louisiana")) { #saved as results_set_4  
-    # for(state_i in c("Maryland", "New York", "North Carolina", "Texas", "Virginia", "Wisconsin" )) { #saved as results_set_5  
+    #for(state_i in c("Maryland", "New York", "North Carolina", "Texas", "Virginia", "Wisconsin" )) { #saved as results_set_5  
     #for(state_i in c("Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Nebraska" )) { #saved as results_set_6  
-    for(state_i in c("Nevada", "New Jersey", "New Mexico", "Ohio", "Oklahoma", "Oregon" )) { #saved as results_set_7  
+    #for(state_i in c("Nevada", "New Jersey", "New Mexico", "Ohio", "Oklahoma", "Oregon" )) { #saved as results_set_7  
+    for(state_i in c("Pennsylvania", "Rhode Island", "South Carolina", "Tennessee", "Washington", "West Virginia" )) { #saved as results_set_8  
       print(state_i)
@@
 #                  r.White.Virginia.Female, r.White.Wisconsin.Female)
+# r.White <- rbind(r.White.Nevada.Male, r.White.NewJersey.Male, r.White.NewMexico.Male, r.White.Ohio.Male, 
+#                  r.White.Oklahoma.Male, r.White.Oregon.Male, 
+#                  r.White.Nevada.Female, r.White.NewJersey.Female, r.White.NewMexico.Female, r.White.Ohio.Female, 
+#                  r.White.Oklahoma.Female, r.White.Oregon.Female)
 
-r.White <- rbind(r.White.Nevada.Male, r.White.NewJersey.Male, r.White.NewMexico.Male, r.White.Ohio.Male, 
-                 r.White.Oklahoma.Male, r.White.Oregon.Male, 
-                 r.White.Nevada.Female, r.White.NewJersey.Female, r.White.NewMexico.Female, r.White.Ohio.Female, 
-                 r.White.Oklahoma.Female, r.White.Oregon.Female)
+r.White <- rbind(r.White.Pennsylvania.Male, r.White.RhodeIsland.Male, r.White.SouthCarolina.Male, r.White.Tennessee.Male, 
+                 r.White.Washington.Male, r.White.WestVirginia.Male, 
+                 r.White.Pennsylvania.Female, r.White.RhodeIsland.Female, r.White.SouthCarolina.Female, r.White.Tennessee.Female, 
+                 r.White.Washington.Female, r.White.WestVirginia.Female)
 r.White$race <- "White"
@@
 #                  r.Black.AlabamaFemale, r.Black.ArizonaFemale, r.Black.ArkansasFemale, r.Black.CaliforniaFemale)
-r.Black <- rbind(r.Black.Nevada.Male, r.Black.NewJersey.Male, r.Black.NewMexico.Male, r.Black.Ohio.Male, 
-                 r.Black.Oklahoma.Male, r.Black.Oregon.Male, 
-                 r.Black.Nevada.Female, r.Black.NewJersey.Female, r.Black.NewMexico.Female, r.Black.Ohio.Female, 
-                 r.Black.Oklahoma.Female, r.Black.Oregon.Female)
+r.Black <- rbind(r.Black.Pennsylvania.Male, r.Black.RhodeIsland.Male, r.Black.SouthCarolina.Male, r.Black.Tennessee.Male, 
+                 r.Black.Washington.Male, r.Black.WestVirginia.Male, 
+                 r.Black.Pennsylvania.Female, r.Black.RhodeIsland.Female, r.Black.SouthCarolina.Female, r.Black.Tennessee.Female, 
+                 r.Black.Washington.Female, r.Black.WestVirginia.Female)
 r.Black$race <- "Black"
@@
 #SET 7: "Nevada", "New Jersey", "New Mexico", "Ohio", "Oklahoma", "Oregon"
-save(r.All, le.calculations, file = "~/BW_results/results_set_7.Rdata") 
-write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_7.csv")
-write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_7.csv")
-write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_7.csv")
-write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_7.csv")
+# save(r.All, le.calculations, file = "~/BW_results/results_set_7.Rdata") 
+# write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_7.csv")
+# write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_7.csv")
+# write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_7.csv")
+# write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_7.csv")
+
+#SET 8: Pennsylvania, Rhode Island, South Carolina, Tennessee, Washington, West Virginia
+save(r.All, le.calculations, file = "~/BW_results/results_set_8.Rdata") 
+write.csv(x = BlackWhite, file = "~/BlackWhiteMortalityGap/Results/BlackWhite_set_8.csv")
+write.csv(x = age.decomp.estimates, file = "~/BlackWhiteMortalityGap/Results/age_decomp_set_8.csv")
+write.csv(x = cod.marginal.estimates, file = "~/BlackWhiteMortalityGap/Results/cod_marginal_set_8.csv")
+write.csv(x = age.cod.estimates, file = "~/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
 ```

### 2017-02-07 Manuscript/Manuscript_using_actual_data.Rmd
forget what i did

@@
 **Figure 5a: Trends in life expectancy for white males between 1969 and 2013 by US state**
-
-```{r figure5a, fig.align="center", fig.height=4, fig.width=11, echo=F}
+```{r jitter, echo=F}
+BlackWhite_results$jitter_year <- sample(x= c(2008:2013), length(BlackWhite_results$year), replace = T)
+```
+
+```{r figure5a, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_white, x = year)) + 
@@
 
-```{r figure5b, fig.align="center", fig.height=4, fig.width=11, echo=F}
+```{r figure5b, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_black, x = year)) + 
@@
 
-```{r figure5c, fig.align="center", fig.height=4, fig.width=11, echo=F}
+```{r figure5c, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_wbgap, x = year)) + 
@@
 **Figure 6a: Contribution of cancer to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-```{r figure6a, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r figure6a, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cancers"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 6a (ii): Contribution of cancer (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state**
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cancers"), aes(x = year, y = COD_cont_prop)) + 
@@
 
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_yrs)) + 
@@
 
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_prop)) + 
@@
 
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Communicable"), aes(x = year, y = COD_cont_yrs)) + 
@@
 
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Communicable"), aes(x = year, y = COD_cont_prop)) + 
@@
 
-```{r, fig.align="center", fig.height=5, fig.width=9, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=9, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Non-communicable"), aes(x = year, y = COD_cont_yrs)) + 
@@
 
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Non-communicable"), aes(x = year, y = COD_cont_prop)) + 
@@
 
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Injuries"), aes(x = year, y = COD_cont_yrs)) + 
@@
 
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Injuries"), aes(x = year, y = COD_cont_prop)) + 
@@
 **Figure 6f: Contribution of all other causes to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "All other causes"), aes(x = year, y = COD_cont_yrs)) + 
@@
 
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "All other causes"), aes(x = year, y = COD_cont_prop)) + 
@@
 
-```{r, fig.align="center", fig.height=4, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_white, x = year)) + 
@@
 
-```{r, fig.align="center", fig.height=4, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_black, x = year)) + 
@@
 
-```{r figure6, fig.align="center", fig.height=4, fig.width=11, echo=F}
+```{r figure6, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_wbgap, x = year)) + 
@@
 
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cancers"), aes(x = year, y = COD_cont_yrs)) + 
@@
 
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cancers"), aes(x = year, y = COD_cont_prop)) + 
@@
 **Figure 8b: Contribution of cardiovascular disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r,  fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r,  fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 8c: Contribution of communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r,  fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r,  fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Communicable"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 8d: Contribution of non-communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r,  fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r,  fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Non-communicable"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 8e: Contribution of injuries to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r,  fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r,  fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Injuries"), aes(x = year, y = COD_cont_yrs)) + 
@@
 **Figure 8f: Contribution of all other causes to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r, fig.align="center", fig.height=5, fig.width=11, echo=F}
+```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
 ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "All other causes"), aes(x = year, y = COD_cont_yrs)) + 

### 2017-02-07 Manuscript/Manuscript_using_actual_data.Rmd

@@
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv")#,
-                        #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv"),
+                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
                         )
@@
                          read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv")#,
-                         #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
                          )
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_6.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_7.csv")#,
-                            #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_8.csv")
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_7.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_8.csv")
                             )
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_6.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_7.csv")#,
-                            #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_8.csv")
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_7.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_8.csv")
                             )
@@
                             read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_6.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_7.csv")#,
-                            #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_8.csv")
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_7.csv"),
+                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_8.csv")
                             )
@@
                                                                                 "NM", "NY", "NC", "OH", 
-                                                                                "OK", "OR", "TX", "VA", 
-                                                                                "DC", "WI")))
-#set 8:
-#"PA", "RI", "SC", "TN", "WA", "WV"
+                                                                                "OK", "OR", "PA", "RI", 
+                                                                                "SC", "TN", "TX", "VA", 
+                                                                                "WA", "DC", "WV", "WI")))
 
@@
 - For most states, the gap decreases between 1969 and 1983, with the largest decrease found in `r diff[dim(diff)[1], ]$state` (change in gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)`, `r round(max1,1)` of which is due to `r mc1`,
-`r diff[dim(diff)[1]-1, ]$state` (`r round(max2,2)` due to `r mc2`), `r diff[dim(diff)[1]-2, ]$state` (`r round(max3,1)` due to `r mc3`), `r diff[dim(diff)[1]-3, ]$state` (`r round(max4,1)` due to `r mc4`), and `r diff[dim(diff)[1]-4, ]$state` (`r round(max5,1)` due to `r mc5`).
+`r diff[dim(diff)[1]-1, ]$state` (change of `r round(diff[dim(diff)[1]-1, ]$era1.diff, 1)`, `r round(max2,2)` due to `r mc2`), `r diff[dim(diff)[1]-2, ]$state` (change of `r round(diff[dim(diff)[1]-2, ]$era1.diff, 1)`, `r round(max3,1)` due to `r mc3`), `r diff[dim(diff)[1]-3, ]$state` (change of `r round(diff[dim(diff)[1]-3, ]$era1.diff, 1)`, `r round(max4,1)` due to `r mc4`), and `r diff[dim(diff)[1]-4, ]$state` (change of `r round(diff[dim(diff)[1]-4, ]$era1.diff, 1)`, `r round(max5,1)` due to `r mc5`).
 
@@
 
-- Injuries across the board contribute to narrowing the gap over time, except for Minnesota.
-
-- Communicable disease narrows (except Arizona), but much smaller contribution than injuries.
+- Injuries across the board contribute to narrowing the gap over time, except for Minnesota and Nebraska.
+
+- Communicable disease narrows (except Arizona and Minnesota), but much smaller contribution than injuries.
 
 - Generally, all other causes narrows, again much smaller contribution than injuries.
+In some states its contribution is larger to the narrowing -- namely, New York, Missouria, Pennsylvania, Rhode Island, and Mississippi.
 
@@
 
-- Cancer increases the gap everywhere except NM. Size of contribution to change is a little smaller than CVD.
-
-- Generally, non-communicable disease increases the gap but the contribution is tiny. Notable differences in Massachusetts, New York, and Maryland where it decreases gap.
+- Cancer increases the gap everywhere except New Mexico. Size of contribution to change is often smaller than contribution of CVD, but not always the case.
+
+- Generally, non-communicable disease increases the gap but the contribution is tiny. Notable differences in Massachusetts, New York, Oregon, and Maryland where it decreases gap.
 
@@
 
-- Except for New Mexico, Arizona, and Kentucky, the life expectancy gap increases in each state, with the biggest driver being communicable disease (HIV/AIDS), especially in MD, NY, NJ, and CT.
+- Except for New Mexico, Arizona, and Kentucky, West Virginia, and Oregon, the life expectancy gap increases in each state, with the biggest driver being communicable disease (HIV/AIDS), especially in MD, NY, NJ, and CT.
 Injuries also contribute to the gap widening over this time frame, most notably in Wisconsin, MD, and LA.
@@
 - Similar to the firts time frame, CVD and cancer also contribute to widening the gap over time.
-Also similar is the minimal effect of  non-communicable disease on the gap, while all other causes contributes to narrowing the gap, with greatest contribution to narrowing in NY, CT, Mississippi, and FL.
+The magnitude of the contribution of CVD is quite variable by state, with Mississippi, Minnesota, and Pennsylvania seeing the largest absolute contributions.
+
+- Also similar is the minimal effect of  non-communicable disease on the gap, while all other causes contributes to narrowing the gap, with greatest contribution to narrowing in NY, Pennsylvania, CT, Mississippi, and FL.
 
@@
 
-- However, increases were found in Arizona, Oregon, Minnesota, and Wisconsin, largely due to cardiovascular disease. Thus, the effect of CVD on the LE varied by state over this time period in women.
-
-- Otherwise, most of the other causes contributed to the gap narrowing, although cancer was associated with narrowing in some states (MS, NM, Oregon, for example) and increases in others (OH, Missouri, Kentucky).
+- However, increases were found in Arizona, Oregon, Minnesota, Washington, and Wisconsin, largely due to cardiovascular disease.
+As you can see, in some states CVD contributed to excacerbating the gap while in most others it makes a large contribution to narrowing the gap.
+
+- Otherwise, most of the other causes contributed to the gap narrowing, although cancer was associated with narrowing in some states (MS, NM, Oregon, for example) and increases in others (Pennsylvania, Ohio, Missouri, Kentucky).
 
@@
 the same or widened across most states.
-Delaware, Kentucky, Kansas, Alabama, were notable exceptions were the gap narrowed -- didn't see effect of communicable disease like other states on the gap increasing (NY, FL, DC, MD, CT, hardest hit).
+Delaware, Kentucky, Oregon, Rhode Island, West Virginia, Alabama, were notable exceptions were the gap narrowed -- didn't see effect of communicable disease like other states on the gap increasing (NY, FL, DC, MD, CT, hardest hit).
 
@@
 
-- All other causes narrows across states (except DC, Indiana). Cancer increases the gap by a bit (except Oregon and Minnesota).
-
-- During the third era, in-line with males, the life expectancy gap decreased during the last 20 year period, and generally speaking all the causes contributed to the gap narrowing over time.
+- All other causes narrows across states (except DC, Indiana). Cancer increases the gap by a bit (except Rhode Island, Oregon, and Minnesota).
+
+- During the third era, in-line with males, the life expectancy gap decreased during the last 20 year period (except DC, Wisconsin), and generally speaking all of the causes contributed to the gap narrowing over time.
 The biggest contributors were CVD and all other causes.
 
-- In a few states, Cancer contributed to widening the gap (NM, Oregon, Arizona, DC).
+- In a few states, cancer contributed to widening the gap (NM, Oregon, Arizona, DC, Wisconsin).
 

### 2017-02-07 Code/calculate_std_mort_rates.Rmd

---
title: "Untitled"
output: html_document
---

```{r load_libraries}
library(dplyr)
library(purrr)
library(ggplot2)
library(stringr)
library(reshape2)
library(grid)
library(gridExtra)
```

```{r}
load("~/BW_results/results_set_1.Rdata")
r.All.set1 <- r.All %>% filter(post.samp %in% c(1,2))

for(i in 2:8){
  print(i)
  load(paste0("~/BW_results/results_set_", i, ".Rdata"))
  assign(paste0("r.All.set", i), r.All %>% filter(post.samp %in% c(1,2)))
  rm(r.All)
}
```

```{r load_std_pop}
std.pop <- read.csv("~/BlackWhiteMortalityGap/Data/US_Standard_Population.csv")

std.pop <- std.pop %>% select(Age, US_Standard_2000) %>% filter(Age != "Total")

std.pop <- std.pop %>% mutate(age.weight = US_Standard_2000/1000000)
```

### 2017-02-07 Manuscript/Manuscript_COD_decomp.Rmd
worked on text

--- 2017-02-07 Manuscript/Manuscript_using_actual_data.Rmd
@@
 ###Introduction
-The life expectancy of a population serves as a barometer for overall health.
-In 2014, life expectancy in the United States was 78.7 years, but hidden within this figure, are differences in life expectancy by race.
-Non-hispanic whites lived 3.6 years longer than non-hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf].
-While both black and white americans have experienced increases in life expectancy since X, the difference between their life expectancies has varied.
+The life expectancy of a population is a barometer for overall health.
+In 2014, life expectancy in the United States was 78.7 years, but hidden within this figure are differences by race.
+Non-Hispanic whites lived 3.6 years longer than Non-Hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf].
+While both black and white Americans have experienced increases in life expectancy since X, the difference between their life expectancies has varied.
 During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, heart disease [Harper 2007].
@@
 For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014].
-This study seeks to better understand these state-level differences, by examining how changes in age at death and cause of death over time have worked to narrow or exacerbate the life expectancy gap.
-
-- fundamentally interested in the qu is there a common story across most states or are there important differences in what happened in NY vs CA vs TX - can we say anything about what we know about state-specific policies, for example CA has strong anti-smoking legislation.
-Does this help us understand why the gap in Wisconsin was increasing?
-
-- is there a lot of heterogeneity or is there not?
-- same something about the proportionate contribution by COD and by age - e.g., changes in injuries accounting for X% for closing the gap in NY but in AL it was Y%.
+This study seeks to better understand these state-level differences, by examining how changes in the cause of death [*and age at death(?)*] over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013.
+Thus, common reasons for the decreasing life expectancy gap can be identified, and anomalous states can be identified.
 
@@
 
-We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEER*Stat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia.
+We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEER*Stat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.
 The mortality data was originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.
@@
 Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013).
-These codes were then grouped into 9X causes of death within the software, which we further aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes.
+These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes.
 This recoding is illustrated in eTable 1 in the eAppendix of the Supplement.
@@
 
-- CR to add line about which states were not considered.
-
-- CR to add line about hispanic origin.
+For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming.
+These states contained less than 0.4% of the blacks in the United States in 2010 [ref].
+We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since the quality of the coding of ethnicity varied by state over the time period included [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf].
 
@@
 The data produced two statistical issues that were addressed by our statistical methods.
-First, our methodology needed to impute the missing death counts.
+First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9).
 Second, several states had small blacks populations which produced even smaller population sizes in each sex-age category.
@@
 Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex.
-The difference in life expectancies between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in every state, year, and sex.
+The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in every state, year, and sex.
 
-To decompose the year- and state- specific difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref].
+To decompose the difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref].
 First, the decomposition analysis partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks.
@@
 For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old.
-We also summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference.
+We summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference.
 
@@
 
-**Will need to change the following**
-Finally, we calculated the change in the life expectancy gap over time within each state-, race-, and sex-specific stratum.
-The change in the life expectancy gap is calculated by substracting the gap that occurred in the later year from the gap that occurred in the earlier year.
-This change in gap can be decomposed by age by taking the difference in the age-specfic contributions across the two years, and can similarly be calculated by cause.
+For each era, we calculated the degree that the life expectancy gap has narrowed or widened over the era and decomposed this *change* to the *difference* in the life expectnacy between blacks and whites according to cause of death.
+To do so, we simply subtracted the contribution (in years) that each cause of death category had made to the difference in life expectancy at the beginning of the era from that made at the beginning of the era.
+An example of the calculation is found in the eAppendix.
+
+*To re-write*: To calculate 95% credible intervals, we sampled 1000 times from the posterior distribution to create 1000 samples for each strata.
+We performed all calculated on each sample and took the median of the estimates as the estimate and the 2.5 and 97.5th qunatiles as the lower and upper bound for the 95% CI.
+Anything about thinning?
 
 All data analyses were performed using R version 3.3.1.
-The bayesian smoothing was performed using `JAGS` (version X.X), and the life expectnacy and decomposition analysis were performed using user-written functions found in the Supplement.
-
-**Add in the stuff about credible intervals**
-
-- sampled from posterior distribution with thinning
+The bayesian smoothing was performed using `JAGS` (version X.X), and the life expectancy and decomposition analysis were performed using user-written functions found in the Supplement.
 
 ###Results
+
 ```{r initialize, echo=FALSE, warning=F, message=F}
@@
 
-**Figure 5a: Trends in life expectancy for white males between 1969 and 2013 by US state**
-```{r jitter, echo=F}
-BlackWhite_results$jitter_year <- sample(x= c(2008:2013), length(BlackWhite_results$year), replace = T)
-```
-
-```{r figure5a, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_white, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_ribbon(aes(ymin = LE_white_lcl, ymax = LE_white_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Life expectancy (years)") +
-           xlab("Year") +
-           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
-
-```
-
-**Figure 5b: Trends in life expectancy for black males between 1969 and 2013 by US state**
-
-```{r figure5b, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_black, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Life expectancy (years)") +
-           xlab("Year") +
-           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
-
-```
-
-**Figure 5c: Trends in the black-white life expectancy gap for males between 1969 and 2013 by US state**
-
-```{r figure5c, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_wbgap, x = year)) + 
-           geom_vline(aes(xintercept = 1983), lty = 2, lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), lty = 2, lwd = 0.5) +
-           geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Black-white life expectancy gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), aes(label = stabbrs), check_overlap = T, size = 2.5))
-
-```
-
-**Figure 6a: Contribution of cancer to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-```{r figure6a, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cancers"), aes(x = year, y = COD_cont_yrs)) + 
-           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap(~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cancers"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 6a (ii): Contribution of cancer (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state**
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cancers"), aes(x = year, y = COD_cont_prop)) + 
-           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap(~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (%)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cancers"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 6b: Contribution of cardiovascular disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_yrs)) + 
-           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap(~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cardiovascular"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 6b (ii): Contribution of cardiovascular disease (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_prop)) + 
-           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (%)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cardiovascular"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 6c: Contribution of communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Communicable"), aes(x = year, y = COD_cont_yrs)) + 
-           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Communicable"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 6c (ii): Contribution of communicable disease (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Communicable"), aes(x = year, y = COD_cont_prop)) + 
-           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (%)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Communicable"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 6d: Contribution of non-communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-
-```{r, fig.align="center", fig.height=9, fig.width=9, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Non-communicable"), aes(x = year, y = COD_cont_yrs)) + 
-           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Non-communicable"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 6d (ii): Contribution of non-communicable disease (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Non-communicable"), aes(x = year, y = COD_cont_prop)) + 
-           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (%)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Non-communicable"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 6e: Contribution of injuries to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Injuries"), aes(x = year, y = COD_cont_yrs)) + 
-           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 6e (ii): Contribution of injuries (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Injuries"), aes(x = year, y = COD_cont_prop)) + 
-           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (%)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 6f: Contribution of all other causes to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "All other causes"), aes(x = year, y = COD_cont_yrs)) + 
-           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "All other causes"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 6f (ii): Contribution of all other causes (%) to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
-
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "All other causes"), aes(x = year, y = COD_cont_prop)) + 
-           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (%)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "All other causes"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 7a: Trends in life expectancy for white females between 1969 and 2013 by US state**
-
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_white, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_ribbon(aes(ymin = LE_white_lcl, ymax = LE_white_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Life expectancy (years)") +
-           xlab("Year") +
-           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
-```
-
-**Figure 7b: Trends in life expectancy for black females between 1969 and 2013 by US state**
-
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_black, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Life expectancy (years)") +
-           xlab("Year") +
-           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
-
-```
-
-**Figure 7c: Trends in the black-white life expectancy gap for females between 1969 and 2013 by US state**
-
-```{r figure6, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_wbgap, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Black-white life expectancy gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), aes(label = stabbrs), check_overlap = T, size = 2.5) 
-) 
-```
-
-**Figure 8a: Contribution of cancer to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cancers"), aes(x = year, y = COD_cont_yrs)) + 
-           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cancers"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 8a: Contribution of cancer (%) to the black-white life expectancy gap for females between 1969 and 2013 by US state**
-
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cancers"), aes(x = year, y = COD_cont_prop)) + 
-           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (%)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cancers"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 8b: Contribution of cardiovascular disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r,  fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_yrs)) + 
-           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cardiovascular"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_prop)) + 
-           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (%)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cardiovascular"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 8c: Contribution of communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r,  fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Communicable"), aes(x = year, y = COD_cont_yrs)) + 
-           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Communicable"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Communicable"), aes(x = year, y = COD_cont_prop)) + 
-           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (%)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Communicable"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 8d: Contribution of non-communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r,  fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Non-communicable"), aes(x = year, y = COD_cont_yrs)) + 
-           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Non-communicable"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Non-communicable"), aes(x = year, y = COD_cont_prop)) + 
-           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (%)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Non-communicable"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 8e: Contribution of injuries to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r,  fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Injuries"), aes(x = year, y = COD_cont_yrs)) + 
-           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Injuries"), aes(x = year, y = COD_cont_prop)) + 
-           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (%)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
-**Figure 8f: Contribution of all other causes to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
-```{r, fig.align="center", fig.height=9, fig.width=11, echo=F}
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "All other causes"), aes(x = year, y = COD_cont_yrs)) + 
-           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (years)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "All other causes"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-
-ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "All other causes"), aes(x = year, y = COD_cont_prop)) + 
-           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
-           geom_line(aes(col = state)) + facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Contribution to LE Gap (%)") +
-           xlab("Year") +
-           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "All other causes"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5)
-)
-```
-
 ###Supplemental Information
@@
 
+**eExample 1: Decomposing the change to the difference in life expectancy over time by cause of death: an example**
+
+To create.
+
 **Background notes from meeting with SH on Dec 20**

### 2017-02-07 Manuscript/Manuscript_trends.Rmd

--- 2017-02-07 Manuscript/Manuscript_using_actual_data.Rmd
@@
 ---
-title: "Exploring how changes in age at death and cause of death have narrowed the black-white life expectancy gap over forty-five years in the United States."
+title: "Untitled"
+author: "Corinne Riddell"
+date: "February 7, 2017"
 output: html_document
 ---
-
-##Riddell CA, Morrison KT, Kaufman JS, Harper S
-
-###Abstract
-
-###Introduction
-The life expectancy of a population serves as a barometer for overall health.
-In 2014, life expectancy in the United States was 78.7 years, but hidden within this figure, are differences in life expectancy by race.
-Non-hispanic whites lived 3.6 years longer than non-hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf].
-While both black and white americans have experienced increases in life expectancy since X, the difference between their life expectancies has varied.
-During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, heart disease [Harper 2007].
-The gap continued to narrow during the 2000s, but states differed in the magnitude of improvement that was made.
-For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014].
-This study seeks to better understand these state-level differences, by examining how changes in age at death and cause of death over time have worked to narrow or exacerbate the life expectancy gap.
-
-- fundamentally interested in the qu is there a common story across most states or are there important differences in what happened in NY vs CA vs TX - can we say anything about what we know about state-specific policies, for example CA has strong anti-smoking legislation.
-Does this help us understand why the gap in Wisconsin was increasing?
-
-- is there a lot of heterogeneity or is there not?
-- same something about the proportionate contribution by COD and by age - e.g., changes in injuries accounting for X% for closing the gap in NY but in AL it was Y%.
-
-###Methods
-
-**The data**
-
-We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEER*Stat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia.
-The mortality data was originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.
-The population data was derived by the NCHS using intercensal bridged-race population estimates.
-This study using data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death.
-Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013).
-These codes were then grouped into 9X causes of death within the software, which we further aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes.
-This recoding is illustrated in eTable 1 in the eAppendix of the Supplement.
-When the number of deaths for a population is between 1 and 9, this data is suppressed by the software and produces a missing death count for that strata.
-However, death counts of exactly 0 or greater than 9 are known precisely.
-
-- CR to add line about which states were not considered.
-
-- CR to add line about hispanic origin.
-
-**Statistical Methods**
-
-The data produced two statistical issues that were addressed by our statistical methods.
-First, our methodology needed to impute the missing death counts.
-Second, several states had small blacks populations which produced even smaller population sizes in each sex-age category.
-In these cases, the estimation of the sex-age mortality rates is imprecise, and produces abrupt changes in the estimation of mortality rates over time that are unrealistic.
-To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.
-
-To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts.
-Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population.
-The mortality rates over time within each state-age-sex-race category are modelled autoregressively, where the log mortality rate in the current year was modelled as a normal variable centred at the log mortality rate estimated from the previous year.
-Truncated Poisson regression was used to impute the supressed death counts, by specifying that the supressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9).
-These models were run separately for every combination of state, race, and sex, and produced death counts that had been smoothed over time for every cause of death within each age group.
-
-To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref].
-Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex.
-The difference in life expectancies between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in every state, year, and sex.
-
-To decompose the year- and state- specific difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref].
-First, the decomposition analysis partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks.
-For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites.
-This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination.
-For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old.
-We also summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference.
-
-In order to understand whether states fared similarly over time, we divided the time frame into three eras which roughly encapsulate an initial period of increases in life expectancy for both races and sexes (1969-1983), an intermediate period of decreases and stagnating life expectancy for blacks but not for whites (1983-1993) and a final period of increasing life expectancy for both races (1993-2013).
-While these eras represent different lengths of time, they best capture three distinct periods of changes to the difference in life expectancy: an intial narrowing of the gap, followed by a widening of the gap that diminished previous progress, and the most recent twenty years of sustained narrowing (eAppendix Figures 1 and 2).
-
-**Will need to change the following**
-Finally, we calculated the change in the life expectancy gap over time within each state-, race-, and sex-specific stratum.
-The change in the life expectancy gap is calculated by substracting the gap that occurred in the later year from the gap that occurred in the earlier year.
-This change in gap can be decomposed by age by taking the difference in the age-specfic contributions across the two years, and can similarly be calculated by cause.
-
-All data analyses were performed using R version 3.3.1.
-The bayesian smoothing was performed using `JAGS` (version X.X), and the life expectnacy and decomposition analysis were performed using user-written functions found in the Supplement.
-
-**Add in the stuff about credible intervals**
-
-- sampled from posterior distribution with thinning
-
-###Results
-```{r initialize, echo=FALSE, warning=F, message=F}
-library(shiny)
-library(plotly)
-library(viridis)
-library(scales)
-library(shinythemes)
-library(dplyr)
-library(purrr)
-library(stringr)
-source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
-
-age_cod_results <-rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv")#,
-                        #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
-                        )
-
-age_cod_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv")#,
-                         #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
-                         )
-
-age_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_1.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_2.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_6.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_7.csv")#,
-                            #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_8.csv")
-                            )
-
-BlackWhite_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_1.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_6.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_7.csv")#,
-                            #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_8.csv")
-                            )
-
-cod_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_1.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_2.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_6.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_7.csv")#,
-                            #read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_8.csv")
-                            )
-```
-
-```{r tidy_data, echo=FALSE}
-
-ncols <- str_split_fixed(as.character(BlackWhite_results$stratum.id), "\\.", 3)
-BlackWhite_results$state <- as.factor(ncols[ , 1])
-BlackWhite_results$year <- as.numeric(ncols[ , 2])
-BlackWhite_results$sex <- as.factor(ncols[ , 3])
-
-age_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_decomp_results, by = "stratum.id")
-cod_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], cod_decomp_results, by = "stratum.id")
-age_cod_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_cod_results, by = "stratum.id")
-
-COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
-                "Non-communicable", "Injuries", "All other causes")
-cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
-age_cod_results$COD <- factor(age_cod_results$COD, COD.levels)
-```
-
-```{r add_Census_Region, echo=FALSE}
-#note that the Census Region of the South is partitioned into two for better display.
-
-BlackWhite_results$Census_Region <- "South"
-BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
-                                                 "Nebraska", "North Dakota", "South Dakota", 
-                                                 "Illinois", "Indiana", "Michigan", "Ohio", 
-                                                 "Wisconsin")] <- "Midwest"
-
-BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Connecticut", "Maine", "Massachusetts", 
-                                                 "New Hampshire", "Rhode Island", "Vermont",
-                                                 "New Jersey", "New York", "Pennsylvania")] <- "Northeast"
-
-BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Arizona", "Colorado", "Idaho", "Montana", 
-                                                 "Nevada", "New Mexico", "Utah", "Wyoming",
-                                                 "Alaska", "California", "Hawaii", "Oregon",
-                                                 "Washington")] <- "West"
-
-BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Delaware", "Florida", "Georgia", "Maryland", 
-                                                  "North Carolina", "South Carolina", 
-                                                  "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"
-
-BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee", 
-                                                  "Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "South Central"
-
-BlackWhite_results$Census_Region <- factor(BlackWhite_results$Census_Region, levels = c("West", "Midwest", "South Central", "South Atlantic", "Northeast"))
-```
-
-```{r add_Census_Division, ncol = 3, echo = F}
-BlackWhite_results$Census_Division <- NA
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Connecticut", "Maine", "Massachusetts", 
-                                                                   "New Hampshire", "Rhode Island", "Vermont")] <- "New England"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("New Jersey", "New York", "Pennsylvania")] <- "Mid-Atlantic"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Illinois", "Indiana", "Michigan", "Ohio", 
-                                                                 "Wisconsin")] <- "East North Central"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
-                                                                 "Nebraska", "North Dakota", "South Dakota")] <- "West North Central"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Delaware", "Florida", "Georgia", "Maryland", 
-                                                                 "North Carolina", "South Carolina", 
-                                                                 "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee")] <- "East South Central"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "West South Central"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Arizona", "Colorado", "Idaho", "Montana", 
-                                                                 "Nevada", "New Mexico", "Utah", "Wyoming")] <- "Mountain"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Alaska", "California", "Hawaii", "Oregon","Washington")] <- "Pacific"
-
-BlackWhite_results$Census_Division <- factor(BlackWhite_results$Census_Division, levels = c("New England", "Mid-Atlantic", "East North Central", "West North Central", "South Atlantic", "East South Central", "West South Central", "Mountain", "Pacific"))
-
-```
-
-```{r addvar_State_abbreviation, echo=FALSE}
-BlackWhite_results <- BlackWhite_results %>% mutate(stabbrs = factor(state,  
-                                                                     labels = c("AL", "AZ", "AR", "CA",
-                                                                                "CO", "CT", "DE", "FL",
-                                                                                "GA", "IL", "IN", "IA",
-                                                                                "KS", "KY", "LA", "MD",
-                                                                                "MA", "MI", "MN", "MS",
-                                                                                "MO", "NE", "NV", "NJ", 
-                                                                                "NM", "NY", "NC", "OH", 
-                                                                                "OK", "OR", "TX", "VA", 
-                                                                                "DC", "WI")))
-#set 8:
-#"PA", "RI", "SC", "TN", "WA", "WV"
-
-                                      # labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
-                                      #            "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", 
-                                      #            "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", 
-                                      #            "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", 
-                                      #            "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
-                                      #            "VT", "VA", "WA", "WV", "WI", "WY")))
-
-cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, Census_Division, stabbrs), 
-                            cod_decomp_results, by = "stratum.id", all.y = T)
-```
-
-```{r prep_era_figure_males, echo = F}
-year1 <- 1969
-year2 <- 1983
-year3 <- 1993
-year4 <- 2013
-
-bw_dat <- BlackWhite_results %>% 
-  filter(year %in% c(year1, year2, year3, year4), is.na(LE_wbgap) == F) %>%
-  select(state, year, LE_wbgap, sex) %>%
-  arrange(year)
-
-bw_dat <- tidyr::spread(bw_dat, year, LE_wbgap) 
-names(bw_dat)[3:6] <- c("first.gap","second.gap", "third.gap", "fourth.gap")  
-bw_dat.males <- bw_dat %>% filter(sex == "Male") %>%
-  mutate(era1.diff = first.gap - second.gap,
-         era2.diff = second.gap - third.gap,
-         era3.diff = third.gap - fourth.gap,
-         state.gdiff.order.1 = reorder(state, first.gap),
-         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
-         state.gdiff.order.2 = reorder(state, second.gap),
-         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
-         state.gdiff.order.3 = reorder(state, third.gap),
-         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
-  )
-
-bw_dat.females <- bw_dat %>% filter(sex == "Female") %>%
-  mutate(era1.diff = first.gap - second.gap,
-         era2.diff = second.gap - third.gap,
-         era3.diff = third.gap - fourth.gap,
-         state.gdiff.order.1 = reorder(state, first.gap),
-         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
-         state.gdiff.order.2 = reorder(state, second.gap),
-         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
-         state.gdiff.order.3 = reorder(state, third.gap),
-         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
-  )
-
-cod.df <- data.frame(state = factor(), sex = factor(), era = character(), 
-                     COD = factor(), Contribution.to.change = numeric(), 
-                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())
-
-age.df <- data.frame(state = factor(), sex = factor(), era = character(), 
-                     age_minbin = numeric(), Contribution.to.change = numeric(), 
-                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())
-
-years <- c(1969, 1983, 1993, 2013)
-
-for (state.i in levels(BlackWhite_results$state)) {
-  for (sex.i in c("Male", "Female")) {
-    for (year.i in 1:3) {
-
-      table1 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(COD)
-      table2 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(COD)
-
-      table3 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(age_minbin)
-      table4 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(age_minbin)
-
-      temp1 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
-                          contribution.to.gap.change(type.of.decomp = "COD", 
-                                                     decomp.table1 = table1, 
-                                                     decomp.table2 = table2, 
-                                                     type.of.data = "results"))
-
-      temp2 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
-                          contribution.to.gap.change(type.of.decomp = "Age", 
-                                                     decomp.table1 = table3, 
-                                                     decomp.table2 = table4,
-                                                     type.of.data = "results"))
-
-      cod.df <- rbind(cod.df, temp1)
-      age.df <- rbind(age.df, temp2)
-    }
-  }
-}
-
-```
-
-```{r more_figure_prep, echo = F}
-
-cod.df <- make_pretty_decomp_plot(decomp.table = cod.df, strat.var.1 = "state", 
-                        strat.var.2 = "sex", strat.var.3 = "era", 
-                        sign.var = "narrowed_gap",
-                        decomp.var = "Contribution.to.change",
-                        decomp.var.prop = "Contrib.to.change.prop", 
-                        partition.bar.var = "COD", type.of.data = "results")
-
-age.df <- make_pretty_decomp_plot(decomp.table = age.df, strat.var.1 = "state", 
-                        strat.var.2 = "sex", strat.var.3 = "era", 
-                        sign.var = "narrowed_gap",
-                        decomp.var = "Contribution.to.change",
-                        decomp.var.prop = "Contrib.to.change.prop", 
-                        partition.bar.var = "Ages", type.of.data = "results")
-
-cod.df <- merge(cod.df, BlackWhite_results[BlackWhite_results$year == 1969 & BlackWhite_results$sex == "Male", 
-                                           c("state", "Census_Region", "Census_Division", "stabbrs")], by = c("state"))
-```
-
-*Changes to the difference in life expectancy in US males*
-
-**Figure 1a** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs.
-their gap in 1983 (solid dashed line), with the change in the gap being represented by the number of years between these lines.
-The decomposition (in years) of the change in the gap by cause of death is also shown for each state.
-
-```{r, echo = F}
-sum1 <- BlackWhite_results %>% filter(year == 1969 & sex == "Male") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
-
-diff <- bw_dat.males %>% arrange(era1.diff)
-
-max1 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1], ]$state & cod.df$era == "1969-1983"])
-mc1 <- cod.df$COD[cod.df$state == diff[dim(diff)[1], ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max1]
-
-max2 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-1, ]$state & cod.df$era == "1969-1983"])
-mc2 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-1, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max2]
-
-max3 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-2, ]$state & cod.df$era == "1969-1983"])
-mc3 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-2, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max3]
-
-max4 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-3, ]$state & cod.df$era == "1969-1983"])
-mc4 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-3, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max4]
-
-max5 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-4, ]$state & cod.df$era == "1969-1983"])
-mc5 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-4, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max5]
-```
-
-What we see so far (based on these `r dim(sum1)[1]` states):
-
-- The gap in 1969 ranges from `r round(sum1[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `.
-
-- For most states, the gap decreases between 1969 and 1983, with the largest decrease found in `r diff[dim(diff)[1], ]$state` (change in gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)`, `r round(max1,1)` of which is due to `r mc1`,
-`r diff[dim(diff)[1]-1, ]$state` (`r round(max2,2)` due to `r mc2`), `r diff[dim(diff)[1]-2, ]$state` (`r round(max3,1)` due to `r mc3`), `r diff[dim(diff)[1]-3, ]$state` (`r round(max4,1)` due to `r mc4`), and `r diff[dim(diff)[1]-4, ]$state` (`r round(max5,1)` due to `r mc5`).
-
-- The gap increases for some states.
-Most notably in DC (CVD and other causes), Minnesota (CVD), Wisconsin (CVD), California (CVD).
-Also, for these states injuries do not contribute as much to narrowing the gap (and in Minnesota, injuries widens the gap), so their gap increasing appears to be due to a combination of changes to CVD making blacks relatively worse off and less changes to injuries than experienced by other states.
-
-- Injuries across the board contribute to narrowing the gap over time, except for Minnesota.
-
-- Communicable disease narrows (except Arizona), but much smaller contribution than injuries.
-
-- Generally, all other causes narrows, again much smaller contribution than injuries.
-
-- Generally, CVD increases gap.
-
-- Cancer increases the gap everywhere except NM. Size of contribution to change is a little smaller than CVD.
-
-- Generally, non-communicable disease increases the gap but the contribution is tiny. Notable differences in Massachusetts, New York, and Maryland where it decreases gap.
-
-**Figure 1a: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
-
-```{r figure1, fig.align = "center", echo=F}
-
-cod.df.males <- merge(subset(cod.df, sex == "Male"), bw_dat.males, by = c("state"))
-cod.df.males <- cod.df.males %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
-                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
-                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)
-
-y.labels <- levels(factor(cod.df.males$state.gdiff.order.1))
-y.num.breaks <- length(levels(factor(cod.df.males$state.gdiff.order.1)))
-
-males.era1 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
-    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
-                  ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.1, fill = COD), 
-              color = "white") +
-#    facet_grid(.~ Census_Region) +
-    geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
-        geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
-   # geom_segment(data = bw_dat.males,
-  #               aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
-  #  geom_segment(data = bw_dat.males,
-  #               aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
-  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
-
-males.era1
-```
-
-```{r, echo =F}
-sum2 <- BlackWhite_results %>% filter(year == 1983 & sex == "Male") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
-```
-
-**Figure 1b** displays each state's male life expectancy gap in 1983 (vertical dashed line) vs. their gap in 1993 (solid dashed line).
-
-- Be careful with the interpretation: the x-axis is much wider for this plot because of DC. (Zoom into same x range to aid interpretation at first)
-
-- The gap in 1983 ranges from `r round(sum2[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum2[1, ]$LE_wbgap_lcl, 1)`, `r round(sum2[1, ]$LE_wbgap_ucl, 1)`] in `r sum2[1, ]$state ` to `r round(sum2[dim(sum2)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum2[dim(sum2)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum2[dim(sum2)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum2[dim(sum2)[1], ]$state `.
-
-- Except for New Mexico, Arizona, and Kentucky, the life expectancy gap increases in each state, with the biggest driver being communicable disease (HIV/AIDS), especially in MD, NY, NJ, and CT.
-Injuries also contribute to the gap widening over this time frame, most notably in Wisconsin, MD, and LA.
-Thus, compared to previous period, these two causes of death led to a widening rather than a narrowing of the gap.
-
-- Similar to the firts time frame, CVD and cancer also contribute to widening the gap over time.
-Also similar is the minimal effect of  non-communicable disease on the gap, while all other causes contributes to narrowing the gap, with greatest contribution to narrowing in NY, CT, Mississippi, and FL.
-
-- DC was especially hard hit, with injuries driving the increase in the gap from 10.5 years to 15.9 years in this 10-year period, an increase of X, Y of which can be attributed to injuries.
-
-**Figure 1b: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1983 and 1993 by US state**
-
-```{r figure1b, echo = F}
-y.labels.2 <- levels(factor(cod.df.males$state.gdiff.order.2))
-y.num.breaks.2 <- length(levels(factor(cod.df.males$state.gdiff.order.2)))
-
-males.era2 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
-    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
-                  ymax = state.gdiff.num.2 + 0.45, 
-                  xmax = new.finish.2, fill = COD), 
-              color = "white") +
-    geom_segment(data = bw_dat.males, 
-                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
-    geom_segment(data = bw_dat.males, 
-                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>% 
-  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
-
-males.era2 
-```
-
-```{r, echo =F}
-sum3 <- BlackWhite_results %>% filter(year == 1993 & sex == "Male") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
-
-sum4 <- BlackWhite_results %>% filter(year == 2013 & sex == "Male") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
-```
-
-- The gap in 1993 ranges from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state ` (or `r round(sum3[dim(sum3)[1]-1, ]$LE_wbgap, 1)` when DC is excluded).
-
-- By 2013, it ranges from `r round(sum4[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum4[dim(sum4)[1], ]$state ` (or `r round(sum4[dim(sum4)[1]-1, ]$LE_wbgap, 1)` when DC is excluded).
-
-- After 1993, the gap narrowed for all states shown, and in most cases all of the causes contributed to the narrowing of the gap.
-Notice again that in Wisconsin and DC, changes to CVD mortality counteract the gap narrowing, while in Arizona changes to communicable disease counteract the narrowing.
-
-**Figure 1c: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
-
-```{r figure1c, echo = F}
-y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
-y.num.breaks.3 <- length(levels(factor(cod.df.males$state.gdiff.order.3)))
-
-males.era3 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
-    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
-                  ymax = state.gdiff.num.3 + 0.45, 
-                  xmax = new.finish.3, fill = COD), 
-              color = "white") +
-    geom_segment(data = bw_dat.males, 
-                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
-    geom_segment(data = bw_dat.males, 
-                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>% 
-  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))
-
-males.era3
-```
-
-```{r, fig.align="center", fig.height=5, fig.width=9, echo=F, eval = F}
-subplot(males.era1, males.era2, males.era3, shareY = T)
-```
-** **
-** **
-
-```{r, echo =F}
-sum5 <- BlackWhite_results %>% filter(year == 1969 & sex == "Female") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
-```
-
-*Changes to the difference in life expectancy in US females*
-
-Quick summary
-
-- Most states saw decreasing gaps in first era.
-
-- However, increases were found in Arizona, Oregon, Minnesota, and Wisconsin, largely due to cardiovascular disease. Thus, the effect of CVD on the LE varied by state over this time period in women.
-
-- Otherwise, most of the other causes contributed to the gap narrowing, although cancer was associated with narrowing in some states (MS, NM, Oregon, for example) and increases in others (OH, Missouri, Kentucky).
-
-- The biggest contributor to the narrowing gap being CVD (in most states, although heterogeniety) and all other causes.
-
-- During the second era, the gap stayed approx.
-the same or widened across most states.
-Delaware, Kentucky, Kansas, Alabama, were notable exceptions were the gap narrowed -- didn't see effect of communicable disease like other states on the gap increasing (NY, FL, DC, MD, CT, hardest hit).
-
-- In women, changes in injury related deaths doesn't play a big role (except DC).
-
-- CVD differential by state. Contributes to widening the gap in CA, Mississippi, and DC, for example, and narrowing it in Kansas, Delaware, and Kentucky.
-
-- All other causes narrows across states (except DC, Indiana). Cancer increases the gap by a bit (except Oregon and Minnesota).
-
-- During the third era, in-line with males, the life expectancy gap decreased during the last 20 year period, and generally speaking all the causes contributed to the gap narrowing over time.
-The biggest contributors were CVD and all other causes.
-
-- In a few states, Cancer contributed to widening the gap (NM, Oregon, Arizona, DC).
-
-**Figure 2a: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
-
-```{r figure2a, fig.align = "center", echo=F}
-
-cod.df.females <- merge(subset(cod.df, sex == "Female"), bw_dat.females, by = c("state"))
-cod.df.females <- cod.df.females %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
-                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
-                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)
-
-y.labels <- levels(factor(cod.df.females$state.gdiff.order.1))
-y.num.breaks <- length(levels(factor(cod.df.females$state.gdiff.order.1)))
-
-females.era1 <- ggplotly(
-  ggplot(subset(cod.df.females, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
-    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
-                  ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.1, fill = COD), 
-              color = "white") +
-    geom_segment(data = bw_dat.females,
-                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
-    geom_segment(data = bw_dat.females,
-                 aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
-  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
-
-females.era1
-```
-
-**Figure 2b: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1983 and 1993 by US state**
-
-```{r figure2b, fig.align = "center", echo=F}
-y.labels.2 <- levels(factor(cod.df.females$state.gdiff.order.2))
-y.num.breaks.2 <- length(levels(factor(cod.df.females$state.gdiff.order.2)))
-
-females.era2 <- ggplotly(
-  ggplot(subset(cod.df.females, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
-    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
-                  ymax = state.gdiff.num.2 + 0.45, 
-                  xmax = new.finish.2, fill = COD), 
-              color = "white") +
-    geom_segment(data = bw_dat.females,
-                 aes(x = third.gap, xend = third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
-    geom_segment(data = bw_dat.females,
-                 aes(x = second.gap, xend = second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>%
-  layout(xaxis = list(title = "Life expectancy gap (years): 1983-1993"), yaxis = list(title = NA, autorange = "reversed"))
-
-females.era2
-```
-
-**Figure 2c: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
-
-```{r figure2c, fig.align = "center", echo=F}
-y.labels.3 <- levels(factor(cod.df.females$state.gdiff.order.3))
-y.num.breaks.3 <- length(levels(factor(cod.df.females$state.gdiff.order.3)))
-
-females.era3 <- ggplotly(
-  ggplot(subset(cod.df.females, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
-    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
-                  ymax = state.gdiff.num.3 + 0.45, 
-                  xmax = new.finish.3, fill = COD), 
-              color = "white") +
-    geom_segment(data = bw_dat.females,
-                 aes(x = fourth.gap, xend = fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
-    geom_segment(data = bw_dat.females,
-                 aes(x = third.gap, xend = third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>%
-  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))
-
-females.era3
-```
 
@@
 ```
-
-###Supplemental Information
-
-**eTable1: Cause of Death Recode**
-
-| Broad cause of death     | Included causes of death                                                     |
-|--------------------------|------------------------------------------------------------------------------|
-| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer.
-|
-| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries.
-|
-| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
-| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Sydrome and Nephrosis.
-|
-| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
-| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death.
-|
-
-**eFigure 1: Difference in life expectancy between black in white males, United States, 1969-2013**
-
-![Difference in life expectancy between black in white males, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Men.bmp)
-
-**eFigure 2: Difference in life expectancy between black in white females, United States, 1969-2013**
-
-![Difference in life expectancy between black in white females, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Women.bmp)
-
-**Background notes from meeting with SH on Dec 20**
-
-- CVD declining starting 1968 - see paper Ruel Stallones - cited in SH JAMA paper - history of decline in CVD.
-
-- Richard cooper also written about BW diff in CVD with epi perspective
-
-- for men there is stagnation from 1969 to late 1980s and then increase in gap with HIV and homicides
-
-- massive crime wave in 1980s, black LE went down from 1984 to 1987 - period when both HIV and homicide went up -- the whole story during that time was HIV and homicide
-
-- heterogeneity between states in the contribution of the causes to the decline in the gap - maybe need to break up into at least two time periods with split at 1990.
-
-- this is where the new figures come in which on the y-axis have the contribution to LE gap (in years or %) vs year on the x axis, and geom_line for each COD (do one COD at a time), and individual line for each state (may facet by census region).
-
-- then have a view of how the different COD have contributed to the decline in the gap over time.

### 2017-02-07 Code/calculate_std_mort_rates.Rmd

@@
 }
+
+r.All.post1.2 <- rbind(r.All.set1, r.All.set2, r.All.set3, r.All.set4, 
+                      r.All.set5, r.All.set6, r.All.set7, r.All.set8)
 ```
@@
 std.pop <- std.pop %>% mutate(age.weight = US_Standard_2000/1000000)
+
+std.pop <- std.pop %>% mutate(age_minbin = 0*(Age == "00 years") + 1*(Age == "01-04 years") + 5*(Age == "05-09 years") +
+                     10*(Age == "10-14 years") + 15*(Age == "15-19 years") + 20*(Age == "20-24 years") +
+                     25*(Age == "25-29 years") + 30*(Age == "30-34 years") + 35*(Age == "35-39 years") +
+                     40*(Age == "40-44 years") + 45*(Age == "45-49 years") + 50*(Age == "50-54 years") +
+                     55*(Age == "55-59 years") + 60*(Age == "60-64 years") + 65*(Age == "65-69 years") +
+                     70*(Age == "70-74 years") + 75*(Age == "75-79 years") + 80*(Age == "80-84 years") +
+                     85*(Age == "85+ years"))
 ```
+
+```{r merge}
+r.All.post1.2 <- merge(r.All.post1.2, std.pop, by = "age_minbin")
+
+mortality.rates <- r.All.post1.2 %>% 
+  group_by(post.samp, state.n, year.n, race.n, sex.n, COD) %>% 
+  arrange(age_minbin) %>%
+  summarise(age.std.mortality.rate = sum(age.weight*smoothed_rate),
+            race = first(race),
+            sex = first(sex),
+            state = first(state))
+
+ggplot(subset(mortality.rates, post.samp == 1 & state == "Alabama"), aes(y = age.std.mortality.rate, x = year.n)) + 
+  geom_line(aes(col = COD)) + facet_grid(race ~ sex)
+
+write.csv(r.All.post1.2, "~/BlackWhiteMortalityGap/Data/states_combined_post_samp_1_2.csv")
+write.csv(mortality.rates, "~/BlackWhiteMortalityGap/Data/mortality_rates.csv")
+```

### 2017-02-08 Code/life_expectancy_functions.R
added fxs for census vars

@@ -496,3 +496,63 @@ investigate_convergence_rest <- function(mcmc_dist, graph_file_name, min, max) {
   plots <- plot_estimate_CI_vs_posterior(posterior.df, posterior.df.age, posterior.df.cod, posterior.df.codage, graph_file_name, min, max)
   return(list(mcmc_dist, saved_bayes, posterior.df, posterior.df.age, posterior.df.cod, posterior.df.codage, plots))
 }
+
+add_Census_Division <- function(data, state) {
+  data$Census_Division <- NA
+  
+  data$Census_Division[data$state %in% c("Connecticut", "Maine", "Massachusetts", 
+                                                                     "New Hampshire", "Rhode Island", "Vermont")] <- "New England"
+  
+  data$Census_Division[data$state %in% c("New Jersey", "New York", "Pennsylvania")] <- "Mid-Atlantic"
+  
+  data$Census_Division[data$state %in% c("Illinois", "Indiana", "Michigan", "Ohio", 
+                                                                     "Wisconsin")] <- "East North Central"
+  
+  data$Census_Division[data$state %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
+                                                                     "Nebraska", "North Dakota", "South Dakota")] <- "West North Central"
+  
+  data$Census_Division[data$state %in% c("Delaware", "Florida", "Georgia", "Maryland", 
+                                                                     "North Carolina", "South Carolina", 
+                                                                     "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"
+  
+  data$Census_Division[data$state %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee")] <- "East South Central"
+  
+  data$Census_Division[data$state %in% c("Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "West South Central"       
+  
+  data$Census_Division[data$state %in% c("Arizona", "Colorado", "Idaho", "Montana", 
+                                                                     "Nevada", "New Mexico", "Utah", "Wyoming")] <- "Mountain"
+  
+  data$Census_Division[data$state %in% c("Alaska", "California", "Hawaii", "Oregon","Washington")] <- "Pacific"
+  
+  data$Census_Division <- factor(data$Census_Division, levels = c("New England", "Mid-Atlantic", "East North Central", "West North Central", "South Atlantic", "East South Central", "West South Central", "Mountain", "Pacific"))
+  
+  return(data)
+}
+
+add_Census_Region <- function(data, state) {
[…]

### 2017-02-08 Manuscript/Manuscript_COD_decomp.Rmd
refactored cxode

@@
                 "Non-communicable", "Injuries", "All other causes")
+
 cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
@@
 
-```{r add_Census_Region, echo=FALSE}
-#note that the Census Region of the South is partitioned into two for better display.
-
-BlackWhite_results$Census_Region <- "South"
-BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
-                                                 "Nebraska", "North Dakota", "South Dakota", 
-                                                 "Illinois", "Indiana", "Michigan", "Ohio", 
-                                                 "Wisconsin")] <- "Midwest"
-
-BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Connecticut", "Maine", "Massachusetts", 
-                                                 "New Hampshire", "Rhode Island", "Vermont",
-                                                 "New Jersey", "New York", "Pennsylvania")] <- "Northeast"
-
-BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Arizona", "Colorado", "Idaho", "Montana", 
-                                                 "Nevada", "New Mexico", "Utah", "Wyoming",
-                                                 "Alaska", "California", "Hawaii", "Oregon",
-                                                 "Washington")] <- "West"
-
-BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Delaware", "Florida", "Georgia", "Maryland", 
-                                                  "North Carolina", "South Carolina", 
-                                                  "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"
-
-BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee", 
-                                                  "Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "South Central"
-
-BlackWhite_results$Census_Region <- factor(BlackWhite_results$Census_Region, levels = c("West", "Midwest", "South Central", "South Atlantic", "Northeast"))
-```
-
-```{r add_Census_Division, ncol = 3, echo = F}
-BlackWhite_results$Census_Division <- NA
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Connecticut", "Maine", "Massachusetts", 
-                                                                   "New Hampshire", "Rhode Island", "Vermont")] <- "New England"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("New Jersey", "New York", "Pennsylvania")] <- "Mid-Atlantic"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Illinois", "Indiana", "Michigan", "Ohio", 
-                                                                 "Wisconsin")] <- "East North Central"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
-                                                                 "Nebraska", "North Dakota", "South Dakota")] <- "West North Central"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Delaware", "Florida", "Georgia", "Maryland", 
-                                                                 "North Carolina", "South Carolina", 
-                                                                 "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee")] <- "East South Central"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "West South Central"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Arizona", "Colorado", "Idaho", "Montana", 
-                                                                 "Nevada", "New Mexico", "Utah", "Wyoming")] <- "Mountain"
-
-BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Alaska", "California", "Hawaii", "Oregon","Washington")] <- "Pacific"
-
-BlackWhite_results$Census_Division <- factor(BlackWhite_results$Census_Division, levels = c("New England", "Mid-Atlantic", "East North Central", "West North Central", "South Atlantic", "East South Central", "West South Central", "Mountain", "Pacific"))
-
-```
-
-```{r addvar_State_abbreviation, echo=FALSE}
-BlackWhite_results <- BlackWhite_results %>% mutate(stabbrs = factor(state,  
-                                                                     labels = c("AL", "AZ", "AR", "CA",
-                                                                                "CO", "CT", "DE", "FL",
-                                                                                "GA", "IL", "IN", "IA",
-                                                                                "KS", "KY", "LA", "MD",
-                                                                                "MA", "MI", "MN", "MS",
-                                                                                "MO", "NE", "NV", "NJ", 
-                                                                                "NM", "NY", "NC", "OH", 
-                                                                                "OK", "OR", "PA", "RI", 
-                                                                                "SC", "TN", "TX", "VA", 
-                                                                                "WA", "DC", "WV", "WI")))
-
-                                      # labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
-                                      #            "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", 
-                                      #            "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", 
-                                      #            "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", 
-                                      #            "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
-                                      #            "VT", "VA", "WA", "WV", "WI", "WY")))
+```{r add_Census_Variables, echo = F}
+BlackWhite_results <- add_Census_Region(BlackWhite_results, state)
+BlackWhite_results <- add_Census_Division(BlackWhite_results, state)
+BlackWhite_results$stabbrs <- state.abb[match(BlackWhite_results$state, state.name)]
 

### 2017-02-08 Code/calculate_std_mort_rates.Rmd

@@
 library(gridExtra)
+
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+
 ```
 
-```{r}
+```{r read_in_posterior_results_samples_1_and_2}
+#this take a long time (a few hours!)
 load("~/BW_results/results_set_1.Rdata")
@@
 
-```{r merge}
+```{r merge_and_calculated_age_stdized_mortality}
 r.All.post1.2 <- merge(r.All.post1.2, std.pop, by = "age_minbin")
 
+#returns dataset with new variable "age.std.mortality.rate" 
 mortality.rates <- r.All.post1.2 %>% 
@@
   arrange(age_minbin) %>%
-  summarise(age.std.mortality.rate = sum(age.weight*smoothed_rate),
+  summarise(age.std.mortality.rate = sum(age.weight*smoothed_rate), #the sum is across age bins
             race = first(race),
@@
 
-ggplot(subset(mortality.rates, post.samp == 1 & state == "Alabama"), aes(y = age.std.mortality.rate, x = year.n)) + 
-  geom_line(aes(col = COD)) + facet_grid(race ~ sex)
+```
 
-write.csv(r.All.post1.2, "~/BlackWhiteMortalityGap/Data/states_combined_post_samp_1_2.csv")
+```{r tidy_mortality.rates}
+COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
+                "Non-communicable", "Injuries", "All other causes")
+
+mortality.rates$COD <- factor(mortality.rates$COD, COD.levels)
+
+mortality.rates <- add_Census_Division(mortality.rates, state)
+mortality.rates <- add_Census_Region(mortality.rates, state)
+mortality.rates$stabbrs <- state.abb[match(mortality.rates$state, state.name)]
+mortality.rates$year <- mortality.rates$year.n + 1968
+mortality.rates$rate.per.100k <- mortality.rates$age.std.mortality.rate*100000
+```
+
+```{r save_mortality_dataset}
+#write.csv(r.All.post1.2, "~/BlackWhiteMortalityGap/Data/states_combined_post_samp_1_2.csv")
 write.csv(mortality.rates, "~/BlackWhiteMortalityGap/Data/mortality_rates.csv")

### 2017-02-08 Code/plot_age_standardardized_mortality_trends.Rmd

---
title: "Untitled"
author: "Corinne Riddell"
date: "2/8/2017"
output: html_document
---

```{r load_libraries, echo=FALSE, warning=F}
library(dplyr)
library(purrr)
library(ggplot2)
library(stringr)
library(reshape2)
library(grid)
library(gridExtra)
library(plotly)
```

```{r, echo=F}
mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates.csv")
```

**Age-adjusted injury-related mortality (per 100,000) in white males, United States, 1969-2013**

```{r, echo=F, fig.width=11, fig.height=9}
ggplotly(ggplot(subset(mortality.rates, sex == "Male" & race == "White" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_line(aes(col = state, lty = race)) +
           facet_wrap( ~ Census_Division, ncol = 3) +
           ylab("Age-standardized mortality rate (per 100,000)") +
           xlab("Year") +
           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Male" & race == "White" & post.samp == 1 & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5))

ggplotly(ggplot(subset(mortality.rates, sex == "Male" & race == "White" & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_line(aes(col = state, lty = factor(post.samp))) +
           facet_wrap( ~ Census_Division, ncol = 3) +
           ylab("Age-standardized mortality rate (per 100,000)") +
           xlab("Year") +
           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Male" & race == "White" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5))
```

**Age-adjusted injury-related mortality (per 100,000) in black males, United States, 1969-2013**

```{r, echo=F, fig.width=11, fig.height=9}
ggplotly(ggplot(subset(mortality.rates, sex == "Male" & race == "Black" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_line(aes(col = state, lty = race)) +
           facet_wrap( ~ Census_Division, ncol = 3) +
           ylab("Age-standardized mortality rate (per 100,000)") +
           xlab("Year") +
           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Male" & race == "Black" & post.samp == 1 & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5))

ggplotly(ggplot(subset(mortality.rates, sex == "Male" & race == "Black" & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_line(aes(col = state, lty = factor(post.samp))) +
           facet_wrap( ~ Census_Division, ncol = 3) +
           ylab("Age-standardized mortality rate (per 100,000)") +
           xlab("Year") +
           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Male" & race == "Black" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5))
```

**Age-adjusted injury-related mortality (per 100,000) in white females, United States, 1969-2013**

```{r, echo=F, fig.width=11, fig.height=9}
ggplotly(ggplot(subset(mortality.rates, sex == "Female" & race == "White" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_line(aes(col = state, lty = race)) +
           facet_wrap( ~ Census_Division, ncol = 3) +
           ylab("Age-standardized mortality rate (per 100,000)") +
           xlab("Year") +
           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Female" & race == "White" & post.samp == 1 & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5))
```

**Age-adjusted injury-related mortality (per 100,000) in black females, United States, 1969-2013**

```{r, echo=F, fig.width=11, fig.height=9}
ggplotly(ggplot(subset(mortality.rates, sex == "Female" & race == "Black" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_line(aes(col = state, lty = race)) +
           facet_wrap( ~ Census_Division, ncol = 3) +
           ylab("Age-standardized mortality rate (per 100,000)") +
           xlab("Year") +
           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Female" & race == "Black" & post.samp == 1 & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5))
```

### 2017-02-08 Code/plot_age_standardardized_mortality_trends.Rmd
small changes

@@
 
-```{r load_libraries, echo=FALSE, warning=F}
+```{r load_libraries, echo=FALSE, warning=F, message=F}
 library(dplyr)
@@
 
-**Age-adjusted injury-related mortality (per 100,000) in white males, United States, 1969-2013**
+**MALES**
+
+**Age-adjusted cancer-related mortality (per 100,000) in males, United States, 1969-2013**
 
 ```{r, echo=F, fig.width=11, fig.height=9}
-ggplotly(ggplot(subset(mortality.rates, sex == "Male" & race == "White" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
+ggplotly(ggplot(subset(mortality.rates, sex == "Male" & post.samp == 1 & COD == "Cancers"), aes(y=rate.per.100k, x = year)) + 
            geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
@@
            geom_line(aes(col = state, lty = race)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
+           facet_wrap( ~ state, ncol = 5) +
            ylab("Age-standardized mortality rate (per 100,000)") +
-           xlab("Year") +
-           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Male" & race == "White" & post.samp == 1 & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
-
-ggplotly(ggplot(subset(mortality.rates, sex == "Male" & race == "White" & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_line(aes(col = state, lty = factor(post.samp))) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Age-standardized mortality rate (per 100,000)") +
-           xlab("Year") +
-           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Male" & race == "White" & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
+           xlab("Year"))
 ```
 
-**Age-adjusted injury-related mortality (per 100,000) in black males, United States, 1969-2013**
+**Age-adjusted cardiovascular-related mortality (per 100,000) in males, United States, 1969-2013**
 
 ```{r, echo=F, fig.width=11, fig.height=9}
-ggplotly(ggplot(subset(mortality.rates, sex == "Male" & race == "Black" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
+ggplotly(ggplot(subset(mortality.rates, sex == "Male" & post.samp == 1 & COD == "Cardiovascular"), aes(y=rate.per.100k, x = year)) + 
            geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
@@
            geom_line(aes(col = state, lty = race)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
+           facet_wrap( ~ state, ncol = 5) +
            ylab("Age-standardized mortality rate (per 100,000)") +
-           xlab("Year") +
-           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Male" & race == "Black" & post.samp == 1 & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
-
-ggplotly(ggplot(subset(mortality.rates, sex == "Male" & race == "Black" & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_line(aes(col = state, lty = factor(post.samp))) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Age-standardized mortality rate (per 100,000)") +
-           xlab("Year") +
-           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Male" & race == "Black" & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
+           xlab("Year"))
 ```
 
-**Age-adjusted injury-related mortality (per 100,000) in white females, United States, 1969-2013**
+**Age-adjusted communicable disease-related mortality (per 100,000) in males, United States, 1969-2013**
 
 ```{r, echo=F, fig.width=11, fig.height=9}
-ggplotly(ggplot(subset(mortality.rates, sex == "Female" & race == "White" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
+ggplotly(ggplot(subset(mortality.rates, sex == "Male" & post.samp == 1 & COD == "Communicable"), aes(y=rate.per.100k, x = year)) + 
            geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
@@
            geom_line(aes(col = state, lty = race)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
+           facet_wrap( ~ state, ncol = 5) +
            ylab("Age-standardized mortality rate (per 100,000)") +
-           xlab("Year") +
-           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Female" & race == "White" & post.samp == 1 & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
+           xlab("Year"))
 ```
 
-**Age-adjusted injury-related mortality (per 100,000) in black females, United States, 1969-2013**
+**Age-adjusted injury-related mortality (per 100,000) in males, United States, 1969-2013**
 
 ```{r, echo=F, fig.width=11, fig.height=9}
-ggplotly(ggplot(subset(mortality.rates, sex == "Female" & race == "Black" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
+ggplotly(ggplot(subset(mortality.rates, sex == "Male" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
            geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
@@
            geom_line(aes(col = state, lty = race)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
+           facet_wrap( ~ state, ncol = 5) +
            ylab("Age-standardized mortality rate (per 100,000)") +
-           xlab("Year") +
-           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Female" & race == "Black" & post.samp == 1 & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
+           xlab("Year"))
 ```
+
+**Age-adjusted non-communicable disease-related mortality (per 100,000) in males, United States, 1969-2013**
+
+```{r, echo=F, fig.width=11, fig.height=9}
+ggplotly(ggplot(subset(mortality.rates, sex == "Male" & post.samp == 1 & COD == "Non-communicable"), aes(y=rate.per.100k, x = year)) + 
+           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
+           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
+           geom_line(aes(col = state, lty = race)) +
+           facet_wrap( ~ state, ncol = 5) +
+           ylab("Age-standardized mortality rate (per 100,000)") +
+           xlab("Year"))
+```
+
+**Age-adjusted mortality (per 100,000) in males for all other causes, United States, 1969-2013**
+
+```{r, echo=F, fig.width=11, fig.height=9}
+ggplotly(ggplot(subset(mortality.rates, sex == "Male" & post.samp == 1 & COD == "All other causes"), aes(y=rate.per.100k, x = year)) + 
+           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
+           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
+           geom_line(aes(col = state, lty = race)) +
+           facet_wrap( ~ state, ncol = 5) +
+           ylab("Age-standardized mortality rate (per 100,000)") +
+           xlab("Year"))
+```
+
+**FEMALES**
+
+**Age-adjusted cancer-related mortality (per 100,000) in females, United States, 1969-2013**
+
+```{r, echo=F, fig.width=11, fig.height=9}
+ggplotly(ggplot(subset(mortality.rates, sex == "Female" & post.samp == 1 & COD == "Cancers"), aes(y=rate.per.100k, x = year)) + 
+           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
+           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
+           geom_line(aes(col = state, lty = race)) +
+           facet_wrap( ~ state, ncol = 5) +
+           ylab("Age-standardized mortality rate (per 100,000)") +
+           xlab("Year"))
+```
+
+**Age-adjusted cardiovascular-related mortality (per 100,000) in females, United States, 1969-2013**
+
+```{r, echo=F, fig.width=11, fig.height=9}
+ggplotly(ggplot(subset(mortality.rates, sex == "Female" & post.samp == 1 & COD == "Cardiovascular"), aes(y=rate.per.100k, x = year)) + 
+           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
+           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
+           geom_line(aes(col = state, lty = race)) +
+           facet_wrap( ~ state, ncol = 5) +
+           ylab("Age-standardized mortality rate (per 100,000)") +
+           xlab("Year"))
+```
+
+**Age-adjusted communicable disease-related mortality (per 100,000) in females, United States, 1969-2013**
+
+```{r, echo=F, fig.width=11, fig.height=9}
+ggplotly(ggplot(subset(mortality.rates, sex == "Female" & post.samp == 1 & COD == "Communicable"), aes(y=rate.per.100k, x = year)) + 
+           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
+           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
+           geom_line(aes(col = state, lty = race)) +
+           facet_wrap( ~ state, ncol = 5) +
+           ylab("Age-standardized mortality rate (per 100,000)") +
+           xlab("Year"))
+```
+
+**Age-adjusted injury-related mortality (per 100,000) in females, United States, 1969-2013**
+
+```{r, echo=F, fig.width=11, fig.height=9}
+ggplotly(ggplot(subset(mortality.rates, sex == "Female" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
+           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
+           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
+           geom_line(aes(col = state, lty = race)) +
+           facet_wrap( ~ state, ncol = 5) +
+           ylab("Age-standardized mortality rate (per 100,000)") +
+           xlab("Year"))
+```
+
+**Age-adjusted non-communicable disease-related mortality (per 100,000) in females, United States, 1969-2013**
+
+```{r, echo=F, fig.width=11, fig.height=9}
+ggplotly(ggplot(subset(mortality.rates, sex == "Female" & post.samp == 1 & COD == "Non-communicable"), aes(y=rate.per.100k, x = year)) + 
+           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
+           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
+           geom_line(aes(col = state, lty = race)) +
+           facet_wrap( ~ state, ncol = 5) +
+           ylab("Age-standardized mortality rate (per 100,000)") +
+           xlab("Year"))
+```
+
+**Age-adjusted mortality (per 100,000) in females for all other causes, United States, 1969-2013**
+
+```{r, echo=F, fig.width=11, fig.height=9}
+ggplotly(ggplot(subset(mortality.rates, sex == "Female" & post.samp == 1 & COD == "All other causes"), aes(y=rate.per.100k, x = year)) + 
+           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
+           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
+           geom_line(aes(col = state, lty = race)) +
+           facet_wrap( ~ state, ncol = 5) +
+           ylab("Age-standardized mortality rate (per 100,000)") +
+           xlab("Year"))
+```

### 2017-02-09 Manuscript/Manuscript_COD_decomp.Rmd
fig 1a writeup

@@
 cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
-age_cod_results$COD <- factor(age_cod_results$COD, COD.levels)
+age_cod_results$COD <- factor(age_cod_results$COD, levels = 1:6, 
+                               labels = c("All other causes", "Cancers", "Cardiovascular", "Communicable",
+                                          "Injuries", "Non-communicable"))
 ```
@@
 
-*Changes to the difference in life expectancy in US males*
+*Changes to the difference in life expectancy in US males, 1969-1983*
 
@@
 their gap in 1983 (solid dashed line), with the change in the gap being represented by the number of years between these lines.
-The decomposition (in years) of the change in the gap by cause of death is also shown for each state.
+This gap is decomposed by cause of death, where causes to the left of the dashed line contributed to narrowing the gap and causes to the right contributed to widening the gap.
+The length of these coloured bars represent the number of years attributable to the cause.
+
+For example, in the **Figure 1a**, New York's gap changed from 8.8 years in 1969 to 6.28 in 1983.
+Causes to the left of the dashed line -- All other causes (pink), injuries (blue), non-communicable diseases (teal), and communicable diseases (green) -- each contributed to narrowing the gap, while cardiovascular disease (red) and cancer (yellow) contributed to widening the gap.
+Since the contribution in years of the leftward causes is larger than the contribution of the rightward causes, the net change is a smaller life expectancy gap in 1983, where all other causes (the longest bar) made the largest contribution.
 
@@
 
-max1 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1], ]$state & cod.df$era == "1969-1983"])
-mc1 <- cod.df$COD[cod.df$state == diff[dim(diff)[1], ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max1]
+for(i in 1:5) {
+  j <- i-1
 
-max2 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-1, ]$state & cod.df$era == "1969-1983"])
-mc2 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-1, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max2]
+  max.value <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-j, ]$state & 
+                                            cod.df$era == "1969-1983" &
+                                            cod.df$sex == "Male" ])
 
-max3 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-2, ]$state & cod.df$era == "1969-1983"])
-mc3 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-2, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max3]
+  max.cause <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-j, ]$state &
+                    cod.df$sex == "Male" &
+                    cod.df$era == "1969-1983" & 
+                    cod.df$Contribution.to.change == max.value]
 
-max4 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-3, ]$state & cod.df$era == "1969-1983"])
-mc4 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-3, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max4]
+  min.value <- min(cod.df$Contribution.to.change[cod.df$state == diff[i, ]$state & 
+                                            cod.df$era == "1969-1983" &
+                                            cod.df$sex == "Male" ])
 
-max5 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-4, ]$state & cod.df$era == "1969-1983"])
-mc5 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-4, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max5]
+  min.cause <- cod.df$COD[cod.df$state == diff[i, ]$state &
+                    cod.df$sex == "Male" &
+                    cod.df$era == "1969-1983" & 
+                    cod.df$Contribution.to.change == min.value]
+
+  assign(paste0("max", i), max.value)
+  assign(paste0("max.cause", i), max.cause)
+  assign(paste0("min", i), min.value)
+  assign(paste0("min.cause", i), min.cause)
+}
 ```
 
-What we see so far (based on these `r dim(sum1)[1]` states):
+In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `.
+Injuries was the largest contributor towards narrowing the gap during this time period, as both races underwent decreased rates of injury-related mortality, but blacks experienced larger decreases to their rates, especially in states with relatively higher rates in 1969 (**eFigure X: Age-adjusted injury related mortality**).
+Minnesota and Nebraska are exceptions, where both states had relatively low rates of injuries among blacks in 1969 and smaller gaps.
+All other causes also contributed to narrower gaps in most states, but generally made a smaller contribution than injuries.
+In New York, all other causes made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks [eFigure showing change in age-cod decomp for New York].
+Communicable diseases modestly contributed to narrowing the gap, except in Arizona and Minnesota.
+Generally, cardivascular disease contributed towards increasing the gap, and reflected slower declines for blacks in cardiovascular mortality than for whites.
+Cancer generally contributed towards increasing the gap, and largely reflected increases to cancer-related mortality for both races, with blacks undergoing steeper changes.
+In most states, non-comunicable diseases contributed a small amount towards increasing the gap, although it decreased the gap among four states, most notably in New York.
 
-- The gap in 1969 ranges from `r round(sum1[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `.
+For `r dim(bw_dat.males[bw_dat.males$era1.diff > 0,])[1]` states, the gap decreased between 1969 and 1983, with the largest decreases in `r diff[dim(diff)[1], ]$state` (decrease in the gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)`, `r round(max1,1)` years, of which is due to `r max.cause1`), `r diff[dim(diff)[1]-1, ]$state` (change of `r round(diff[dim(diff)[1]-1, ]$era1.diff, 1)`, `r round(max2,1)` due to `r max.cause2`), `r diff[dim(diff)[1]-2, ]$state` (change of `r round(diff[dim(diff)[1]-2, ]$era1.diff, 1)`, `r round(max3,1)` due to `r max.cause3`), `r diff[dim(diff)[1]-3, ]$state` (change of `r round(diff[dim(diff)[1]-3, ]$era1.diff, 1)`, `r round(max4,1)` due to `r max.cause4`), and `r diff[dim(diff)[1]-4, ]$state` (change of `r round(diff[dim(diff)[1]-4, ]$era1.diff, 1)`, `r round(max5,1)` due to `r max.cause5`).
 
-- For most states, the gap decreases between 1969 and 1983, with the largest decrease found in `r diff[dim(diff)[1], ]$state` (change in gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)`, `r round(max1,1)` of which is due to `r mc1`,
-`r diff[dim(diff)[1]-1, ]$state` (change of `r round(diff[dim(diff)[1]-1, ]$era1.diff, 1)`, `r round(max2,2)` due to `r mc2`), `r diff[dim(diff)[1]-2, ]$state` (change of `r round(diff[dim(diff)[1]-2, ]$era1.diff, 1)`, `r round(max3,1)` due to `r mc3`), `r diff[dim(diff)[1]-3, ]$state` (change of `r round(diff[dim(diff)[1]-3, ]$era1.diff, 1)`, `r round(max4,1)` due to `r mc4`), and `r diff[dim(diff)[1]-4, ]$state` (change of `r round(diff[dim(diff)[1]-4, ]$era1.diff, 1)`, `r round(max5,1)` due to `r mc5`).
-
-- The gap increases for some states.
-Most notably in DC (CVD and other causes), Minnesota (CVD), Wisconsin (CVD), California (CVD).
-Also, for these states injuries do not contribute as much to narrowing the gap (and in Minnesota, injuries widens the gap), so their gap increasing appears to be due to a combination of changes to CVD making blacks relatively worse off and less changes to injuries than experienced by other states.
-
-- Injuries across the board contribute to narrowing the gap over time, except for Minnesota and Nebraska.
-
-- Communicable disease narrows (except Arizona and Minnesota), but much smaller contribution than injuries.
-
-- Generally, all other causes narrows, again much smaller contribution than injuries.
-In some states its contribution is larger to the narrowing -- namely, New York, Missouria, Pennsylvania, Rhode Island, and Mississippi.
-
-- Generally, CVD increases gap.
-
-- Cancer increases the gap everywhere except New Mexico. Size of contribution to change is often smaller than contribution of CVD, but not always the case.
-
-- Generally, non-communicable disease increases the gap but the contribution is tiny. Notable differences in Massachusetts, New York, Oregon, and Maryland where it decreases gap.
+The gap increased for `r dim(bw_dat.males[bw_dat.males$era1.diff < 0,])[1]` states.
+Most notably in `r diff$state[1]` (increase in gap of `r round(abs(diff$era1.diff[1]), 1)`, `r abs(round(min1,1))` due to `r min.cause1`), `r diff$state[2]` (increase in gap of `r round(abs(diff$era1.diff[2]), 1)`, `r abs(round(min2,1))` due to `r min.cause2`), `r diff$state[3]` (increase in gap of `r round(abs(diff$era1.diff[3]), 1)`, `r abs(round(min3,1))` due to `r min.cause3`), `r diff$state[4]` (increase in gap of `r round(abs(diff$era1.diff[4]), 1)`, `r abs(round(min4,1))` due to `r min.cause4`).
+For these states, injuries do not contribute as much to narrowing the gap as in other states (and in Minnesota, injuries widens the gap), so these gap increases appear to be due to a combination of changes to cardiovascular-related mortality making blacks relatively worse off and less changes to injuries than experienced by other states.
+[*These states had lower rates of injury-related mortality for blacks in 1969, and underwent more modest decreases in this rate than many other states that started the era with much higher injury rates for blacks.*]
 

### 2017-02-09 Manuscript/Manuscript_COD_decomp.Rmd
more writing

@@
 This study seeks to better understand these state-level differences, by examining how changes in the cause of death [*and age at death(?)*] over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013.
-Thus, common reasons for the decreasing life expectancy gap can be identified, and anomalous states can be identified.
+Thus, common reasons for the decreasing life expectancy gap can be identified, and anomalous states can be examined.
 
@@
 source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
-
-age_cod_results <-rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv"),
-                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
-                        )
-
-age_cod_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv"),
-                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
-                         )
-
-age_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_1.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_2.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_6.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_7.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_8.csv")
-                            )
-
-BlackWhite_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_1.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_6.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_7.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_8.csv")
-                            )
-
-cod_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_1.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_2.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_6.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_7.csv"),
-                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_8.csv")
-                            )
 ```
 
-```{r tidy_data, echo=FALSE}
-
-ncols <- str_split_fixed(as.character(BlackWhite_results$stratum.id), "\\.", 3)
-BlackWhite_results$state <- as.factor(ncols[ , 1])
-BlackWhite_results$year <- as.numeric(ncols[ , 2])
-BlackWhite_results$sex <- as.factor(ncols[ , 3])
-
-age_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_decomp_results, by = "stratum.id")
-cod_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], cod_decomp_results, by = "stratum.id")
-age_cod_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_cod_results, by = "stratum.id")
+```{r read_data, echo=F}
+age_cod_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results.csv")
+age_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_results.csv")
+cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
+BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
 
@@
 cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
-age_cod_results$COD <- factor(age_cod_results$COD, levels = 1:6, 
-                               labels = c("All other causes", "Cancers", "Cardiovascular", "Communicable",
-                                          "Injuries", "Non-communicable"))
-```
+age_cod_results$COD <- factor(age_cod_results$COD, levels(age_cod_results$COD)[c(3, 2, 4, 5, 6, 1)])
 
-```{r add_Census_Variables, echo = F}
-BlackWhite_results <- add_Census_Region(BlackWhite_results, state)
-BlackWhite_results <- add_Census_Division(BlackWhite_results, state)
-BlackWhite_results$stabbrs <- state.abb[match(BlackWhite_results$state, state.name)]
+age_decomp_results$age <- factor(age_decomp_results$age, levels = levels(age_decomp_results$age)[c(1, 2, 11, 3:10, 12:19)])
 
-cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, Census_Division, stabbrs), 
-                            cod_decomp_results, by = "stratum.id", all.y = T)
 ```
@@
 
-For example, in the **Figure 1a**, New York's gap changed from 8.8 years in 1969 to 6.28 in 1983.
+For example, in the **Figure 1a**, New York's gap changed from 8.8 years in 1969 to 6.3 in 1983.
 Causes to the left of the dashed line -- All other causes (pink), injuries (blue), non-communicable diseases (teal), and communicable diseases (green) -- each contributed to narrowing the gap, while cardiovascular disease (red) and cancer (yellow) contributed to widening the gap.
@@
 In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `.
-Injuries was the largest contributor towards narrowing the gap during this time period, as both races underwent decreased rates of injury-related mortality, but blacks experienced larger decreases to their rates, especially in states with relatively higher rates in 1969 (**eFigure X: Age-adjusted injury related mortality**).
+Injuries was the largest contributor towards narrowing the gap during this time period, as both races underwent decreased rates of injury-related mortality, but blacks experienced larger decreases to their rates, especially in states with relatively higher rates in 1969 (**eAppendix B: Age-adjusted mortality by causes of death**).
 Minnesota and Nebraska are exceptions, where both states had relatively low rates of injuries among blacks in 1969 and smaller gaps.
 All other causes also contributed to narrower gaps in most states, but generally made a smaller contribution than injuries.
-In New York, all other causes made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks [eFigure showing change in age-cod decomp for New York].
+In New York, all other causes made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks (**eFigure 3**).
 Communicable diseases modestly contributed to narrowing the gap, except in Arizona and Minnesota.
-Generally, cardivascular disease contributed towards increasing the gap, and reflected slower declines for blacks in cardiovascular mortality than for whites.
-Cancer generally contributed towards increasing the gap, and largely reflected increases to cancer-related mortality for both races, with blacks undergoing steeper changes.
+Generally, cardivascular disease contributed towards increasing the gap, and reflects slower declines for blacks in cardiovascular mortality than for whites.
+Cancer generally contributed towards increasing the gap, and largely reflects increased cancer-related mortality for both races, with blacks undergoing steeper changes.
 In most states, non-comunicable diseases contributed a small amount towards increasing the gap, although it decreased the gap among four states, most notably in New York.
@@
 For these states, injuries do not contribute as much to narrowing the gap as in other states (and in Minnesota, injuries widens the gap), so these gap increases appear to be due to a combination of changes to cardiovascular-related mortality making blacks relatively worse off and less changes to injuries than experienced by other states.
-[*These states had lower rates of injury-related mortality for blacks in 1969, and underwent more modest decreases in this rate than many other states that started the era with much higher injury rates for blacks.*]
 
@@
 
-```{r figure1, fig.align = "center", echo=F}
+```{r figure1, fig.align = "center", echo=F, fig.height=9}
 
@@
 
+**eFigure 3: Comparison of the contribution of age and cause of death to the difference in life expectancy between blacks and whites in New York, 1969 vs. 1983**
+
+```{r ny_1969_1983_age_cod_decomp, echo = F, fig.width = 9}
+ny.dat <- age_cod_results %>% filter(state == "New York" & year %in% c(1969, 1983) & sex == "Male") %>% 
+                                     mutate(sign.var = sign(age_COD_cont_yrs))
+
+ny.1969 <- make_pretty_decomp_plot(decomp.table = subset(ny.dat, year == 1969), sign.var = "sign.var",
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
+                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+ny.1983 <- make_pretty_decomp_plot(decomp.table = subset(ny.dat, year == 1983), sign.var = "sign.var",
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
+                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+
+ggplotly(ggplot(rbind(ny.1969, ny.1983),
+                aes(y = age, x = start)) + 
+           geom_segment(aes(xend = finish, col = COD, yend = age), lwd = 3) +
+           xlab(paste0("Contribution to life expectancy gap in years")) +
+           ylab("Age group\n") + theme_minimal() +
+           geom_vline(xintercept = 0) + 
+           facet_wrap(~ year))
+
+```
+
 **Background notes from meeting with SH on Dec 20**

### 2017-02-09 Manuscript/Manuscript_COD_decomp.Rmd
age factor levels

@@
 age_decomp_results$age <- factor(age_decomp_results$age, levels = levels(age_decomp_results$age)[c(1, 2, 11, 3:10, 12:19)])
-
+age_cod_results$age <- factor(age_cod_results$age, levels = levels(age_cod_results$age)[c(1, 2, 11, 3:10, 12:19)])                                                             
 ```

### 2017-02-09 Code/combine_results.Rmd

---
title: "Combine_results"
author: "Corinne Riddell"
date: "2/9/2017"
output: html_document
---

**Combine results across the sets**
The analysis was run in 8 sets. In this file we combine the results and tidy the data for graph and presentation.

```{r initialize}
library(stringr)
library(dplyr)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

```{r combine_results}
age_cod_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv")
age_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_1.csv")
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_1.csv")
cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_1.csv")

for(i in 2:8){
  age_cod_results <- rbind(age_cod_results, 
                           read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_",
                                           i, ".csv")))
  age_decomp_results <- rbind(age_decomp_results, 
                              read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_",
                                              i, ".csv")))
  BlackWhite_results <- rbind(BlackWhite_results, 
                              read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_",
                                              i, ".csv")))
  cod_decomp_results <- rbind(cod_decomp_results, 
                              read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_",
                                              i, ".csv")))
}
```

```{r tidy_data, echo=FALSE}
ncols <- str_split_fixed(as.character(BlackWhite_results$stratum.id), "\\.", 3)
BlackWhite_results$state <- as.factor(ncols[ , 1])
BlackWhite_results$year <- as.numeric(ncols[ , 2])
BlackWhite_results$sex <- as.factor(ncols[ , 3])

age_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_decomp_results, by = "stratum.id")
cod_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], cod_decomp_results, by = "stratum.id")
age_cod_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_cod_results, by = "stratum.id")

COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
                "Non-communicable", "Injuries", "All other causes")

cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
age_cod_results$COD <- factor(age_cod_results$COD, levels = 1:6, 
                               labels = c("All other causes", "Cancers", "Cardiovascular", "Communicable",
                                          "Injuries", "Non-communicable"))

age_cod_results$age <- factor(age_cod_results$age_minbin, levels = c(0, 1, seq(5, 85, by = 5)),
                                                                     labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                                                "15-19 years", "20-24 years", "25-29 years", "30-34 years",
                                                                                "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
                                                                                "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
                                                                                "75-79 years", "80-84 years","85+ years"), ordered = T)

age_decomp_results$age <- factor(age_decomp_results$age_minbin, levels = c(0, 1, seq(5, 85, by = 5)),
                                                                     labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                                                "15-19 years", "20-24 years", "25-29 years", "30-34 years",
                                                                                "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
                                                                                "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
                                                                                "75-79 years", "80-84 years","85+ years"), ordered = T)
```

```{r add_Census_Variables, echo = F}
BlackWhite_results <- add_Census_Region(BlackWhite_results, state)
BlackWhite_results <- add_Census_Division(BlackWhite_results, state)
BlackWhite_results$stabbrs <- state.abb[match(BlackWhite_results$state, state.name)]

cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, Census_Division, stabbrs), 
                            cod_decomp_results, by = "stratum.id", all.y = T)
```

```{r save_combined_datasets, echo = F}
write.csv(age_cod_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results.csv")
write.csv(age_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_results.csv")
write.csv(cod_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
write.csv(BlackWhite_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
```

### 2017-02-09 Manuscript/Manuscript_COD_decomp.Rmd

@@
 ---
-title: "Exploring how changes in age at death and cause of death have narrowed the black-white life expectancy gap over forty-five years in the United States."
-output: html_document
+title: Exploring how changes in age at death and cause of death have narrowed the
+  black-white life expectancy gap over forty-five years in the United States.
+output:
+  html_document: default
+  word_document: default
+always_allow_html: yes
 ---
@@
 
-```{r figure1, fig.align = "center", echo=F, fig.height=9}
+```{r figure1, fig.align = "center", echo=F, fig.height=9, fig.width=7}
 
@@
   select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
+
+  min.value.DC <- min(cod.df$Contribution.to.change[cod.df$state == "Washington DC" & 
+                                            cod.df$era == "1983-1993" &
+                                            cod.df$sex == "Male" ])
+
+  min.cause.DC <- cod.df$COD[cod.df$state == "Washington DC" &
+                    cod.df$sex == "Male" &
+                    cod.df$era == "1983-1993" & 
+                    cod.df$Contribution.to.change == min.value.DC]
 ```
 
-**Figure 1b** displays each state's male life expectancy gap in 1983 (vertical dashed line) vs. their gap in 1993 (solid dashed line).
+**Figure 1b** displays each state's male life expectancy gap in 1983 (vertical dashed line) vs.
+their gap in 1993 (solid dashed line).[Be careful with the interpretation: the x-axis is much wider for this plot because of DC.]
 
-- Be careful with the interpretation: the x-axis is much wider for this plot because of DC. (Zoom into same x range to aid interpretation at first)
-
-- The gap in 1983 ranges from `r round(sum2[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum2[1, ]$LE_wbgap_lcl, 1)`, `r round(sum2[1, ]$LE_wbgap_ucl, 1)`] in `r sum2[1, ]$state ` to `r round(sum2[dim(sum2)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum2[dim(sum2)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum2[dim(sum2)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum2[dim(sum2)[1], ]$state `.
-
-- Except for New Mexico, Arizona, and Kentucky, West Virginia, and Oregon, the life expectancy gap increases in each state, with the biggest driver being communicable disease (HIV/AIDS), especially in MD, NY, NJ, and CT.
-Injuries also contribute to the gap widening over this time frame, most notably in Wisconsin, MD, and LA.
-Thus, compared to previous period, these two causes of death led to a widening rather than a narrowing of the gap.
-
-- Similar to the firts time frame, CVD and cancer also contribute to widening the gap over time.
-The magnitude of the contribution of CVD is quite variable by state, with Mississippi, Minnesota, and Pennsylvania seeing the largest absolute contributions.
-
-- Also similar is the minimal effect of  non-communicable disease on the gap, while all other causes contributes to narrowing the gap, with greatest contribution to narrowing in NY, Pennsylvania, CT, Mississippi, and FL.
-
-- DC was especially hard hit, with injuries driving the increase in the gap from 10.5 years to 15.9 years in this 10-year period, an increase of X, Y of which can be attributed to injuries.
+During the second time period, from 1983 to 1993, the life expectancy gap for males in `r dim(bw_dat.males[bw_dat.males$era2.diff < 0, ])[1]` of the analysed `r dim(bw_dat.males)[1]` widened.
+For states such as New York, Maryland, New Jersey, and Connecticut, the biggest driver of the widening gap was communicable disease, reflecting the heightened risk of dying from HIV/AIDS in black males during the height of the AIDS epidemic.
+While both blacks and whites experienced a dramatic increase in mortality related to communicable diseases, blacks' rates were even higher.
+Injuries also contributed to the gap widening, most notably in Wisconsin, Maryland, and Louisiana.
+For many states, injury-related mortality increased for blacks but not for whites during this time period.
+Compared to the previous period, these two causes of death led to widwer rather than narrowed gaps.
+Cancer and cardiovascular disease contributed towards wider gaps, although the magnitude of the contribution of cardiovascular disease is quite hetergenous by state, with Mississippi, Minnesota, and Pennsylvania seeing the largest absolute contributions.
+All other causes contributed to a narrower gap, especially in New York, Pennsylvania, Connecticut, Mississippi, and Florida [note: This wasn't related to infant mortality improvements among blacks in NY, PA, but partially related to IMR improvements in CT, MS, and FL].
+Most notably, the Disrict of Columbia had the highest gap to begin and underwent the largest increase to their gap, which increased from 10.5 years to 15.9 years over this 10-year period, reflecting an increase of `r abs(round(bw_dat.males$era2.diff[bw_dat.males$sex == "Male" & bw_dat.males$state == "Washington DC"], 1))` years, of which `r round(min.value.DC, 1)` was due to `r min.cause.DC`.
 
@@
 
-```{r figure1b, echo = F}
+```{r figure1b, echo = F, fig.height=9, fig.width=11}
+# - The gap in 1983 ranges from `r round(sum2[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum2[1, ]$LE_wbgap_lcl, 1)`, `r round(sum2[1, ]$LE_wbgap_ucl, 1)`] in `r sum2[1, ]$state ` to `r round(sum2[dim(sum2)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum2[dim(sum2)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum2[dim(sum2)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum2[dim(sum2)[1], ]$state `.
+
 y.labels.2 <- levels(factor(cod.df.males$state.gdiff.order.2))
@@
 ```{r ny_1969_1983_age_cod_decomp, echo = F, fig.width = 9}
-ny.dat <- age_cod_results %>% filter(state == "New York" & year %in% c(1969, 1983) & sex == "Male") %>% 
+ny.dat <- age_cod_results %>% filter(state == "New York" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                      mutate(sign.var = sign(age_COD_cont_yrs))
@@
                                   decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+ny.1993 <- make_pretty_decomp_plot(decomp.table = subset(ny.dat, year == 1993), sign.var = "sign.var",
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
+                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
 
-ggplotly(ggplot(rbind(ny.1969, ny.1983),
+ggplotly(ggplot(rbind(ny.1969, ny.1983, ny.1993),
                 aes(y = age, x = start)) + 
@@
 
-**Background notes from meeting with SH on Dec 20**
+**Pennsylvania**
+```{r pa_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
+pa.dat <- age_cod_results %>% filter(state == "Pennsylvania" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
+                                     mutate(sign.var = sign(age_COD_cont_yrs))
+
+pa.1993 <- make_pretty_decomp_plot(decomp.table = subset(pa.dat, year == 1993), sign.var = "sign.var",
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
+                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+pa.1983 <- make_pretty_decomp_plot(decomp.table = subset(pa.dat, year == 1983), sign.var = "sign.var",
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
+                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+
+ggplotly(ggplot(rbind(pa.1993, pa.1983),
+                aes(y = age, x = start)) + 
+           geom_segment(aes(xend = finish, col = COD, yend = age), lwd = 3) +
+           xlab(paste0("Contribution to life expectancy gap in years")) +
+           ylab("Age group\n") + theme_minimal() +
+           geom_vline(xintercept = 0) + 
+           facet_wrap(~ year)) 
+```
+
+**Connecticut**
+```{r ct_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
+ct.dat <- age_cod_results %>% filter(state == "Connecticut" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
+                                     mutate(sign.var = sign(age_COD_cont_yrs))
+
+ct.1993 <- make_pretty_decomp_plot(decomp.table = subset(ct.dat, year == 1993), sign.var = "sign.var",
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
+                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+ct.1983 <- make_pretty_decomp_plot(decomp.table = subset(ct.dat, year == 1983), sign.var = "sign.var",
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
+                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+
+ggplotly(ggplot(rbind(ct.1993, ct.1983),
+                aes(y = age, x = start)) + 
+           geom_segment(aes(xend = finish, col = COD, yend = age), lwd = 3) +
+           xlab(paste0("Contribution to life expectancy gap in years")) +
+           ylab("Age group\n") + theme_minimal() +
+           geom_vline(xintercept = 0) + 
+           facet_wrap(~ year)) 
+```
+
+**Mississippi**
+```{r ms_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
+ms.dat <- age_cod_results %>% filter(state == "Mississippi" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
+                                     mutate(sign.var = sign(age_COD_cont_yrs))
+
+ms.1993 <- make_pretty_decomp_plot(decomp.table = subset(ms.dat, year == 1993), sign.var = "sign.var",
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
+                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+ms.1983 <- make_pretty_decomp_plot(decomp.table = subset(ms.dat, year == 1983), sign.var = "sign.var",
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
+                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+
+ggplotly(ggplot(rbind(ms.1993, ms.1983),
+                aes(y = age, x = start)) + 
+           geom_segment(aes(xend = finish, col = COD, yend = age), lwd = 3) +
+           xlab(paste0("Contribution to life expectancy gap in years")) +
+           ylab("Age group\n") + theme_minimal() +
+           geom_vline(xintercept = 0) + 
+           facet_wrap(~ year)) 
+```
+
+**Florida**
+```{r fl_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
+fl.dat <- age_cod_results %>% filter(state == "Florida" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
+                                     mutate(sign.var = sign(age_COD_cont_yrs))
+
+fl.1993 <- make_pretty_decomp_plot(decomp.table = subset(fl.dat, year == 1993), sign.var = "sign.var",
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
+                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+fl.1983 <- make_pretty_decomp_plot(decomp.table = subset(fl.dat, year == 1983), sign.var = "sign.var",
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
+                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+
+ggplotly(ggplot(rbind(fl.1993, fl.1983),
+                aes(y = age, x = start)) + 
+           geom_segment(aes(xend = finish, col = COD, yend = age), lwd = 3) +
+           xlab(paste0("Contribution to life expectancy gap in years")) +
+           ylab("Age group\n") + theme_minimal() +
+           geom_vline(xintercept = 0) + 
+           facet_wrap(~ year)) 
+```
+
+**Things for corinne to review**
 
@@
 - Richard cooper also written about BW diff in CVD with epi perspective
-
-- for men there is stagnation from 1969 to late 1980s and then increase in gap with HIV and homicides
-
-- massive crime wave in 1980s, black LE went down from 1984 to 1987 - period when both HIV and homicide went up -- the whole story during that time was HIV and homicide
-
-- heterogeneity between states in the contribution of the causes to the decline in the gap - maybe need to break up into at least two time periods with split at 1990.
-
-- this is where the new figures come in which on the y-axis have the contribution to LE gap (in years or %) vs year on the x axis, and geom_line for each COD (do one COD at a time), and individual line for each state (may facet by census region).
-
-- then have a view of how the different COD have contributed to the decline in the gap over time.

### 2017-02-09 Manuscript/Manuscript_COD_decomp.Rmd
finished text for males

@@
 output:
+  word_document: default
   html_document: default
-  word_document: default
 always_allow_html: yes
@@
 }
+
+cod.df.males <- merge(subset(cod.df, sex == "Male"), bw_dat.males, by = c("state"))
+cod.df.males <- cod.df.males %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
+                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
+                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)
+
+```
+
+```{r, echo =F}
+sum2 <- BlackWhite_results %>% filter(year == 1983 & sex == "Male") %>% 
+  arrange(LE_wbgap) %>% 
+  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
+
+  min.value.DC <- min(cod.df$Contribution.to.change[cod.df$state == "Washington DC" & 
+                                            cod.df$era == "1983-1993" &
+                                            cod.df$sex == "Male" ])
+
+  min.cause.DC <- cod.df$COD[cod.df$state == "Washington DC" &
+                    cod.df$sex == "Male" &
+                    cod.df$era == "1983-1993" & 
+                    cod.df$Contribution.to.change == min.value.DC]
+```
+
+```{r, echo =F}
+sum3 <- BlackWhite_results %>% filter(year == 1993 & sex == "Male") %>% 
+  arrange(LE_wbgap) %>% 
+  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
+
+sum4 <- BlackWhite_results %>% filter(year == 2013 & sex == "Male") %>% 
+  arrange(LE_wbgap) %>% 
+  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
 ```
@@
 
+During the second time period, from 1983 to 1993, the life expectancy gap for males in `r dim(bw_dat.males[bw_dat.males$era2.diff < 0, ])[1]` of the analysed `r dim(bw_dat.males)[1]` widened (**Figure 1b**).
+For states such as New York, Maryland, New Jersey, and Connecticut, the biggest driver of the widening gap was communicable disease, reflecting the heightened risk of dying from HIV/AIDS in black males during the height of the AIDS epidemic.
+While both blacks and whites experienced a dramatic increase in mortality related to communicable diseases, blacks' rates were even higher.
+Injuries also contributed to the gap widening, most notably in Wisconsin, Maryland, and Louisiana.
+For many states, injury-related mortality increased for blacks but not for whites during this time period.
+Compared to the previous period, these two causes of death led to widwer rather than narrowed gaps.
+Cancer and cardiovascular disease contributed towards wider gaps, although the magnitude of the contribution of cardiovascular disease is quite hetergenous by state, with Mississippi, Minnesota, and Pennsylvania seeing the largest absolute contributions.
+All other causes contributed to a narrower gap, especially in New York, Pennsylvania, Connecticut, Mississippi, and Florida [note: This wasn't related to infant mortality improvements among blacks in NY, PA, but partially related to IMR improvements in CT, MS, and FL].
+Most notably, the Disrict of Columbia had the highest gap to begin and underwent the largest increase to their gap, which increased from 10.5 years to 15.9 years over this 10-year period, reflecting an increase of `r abs(round(bw_dat.males$era2.diff[bw_dat.males$sex == "Male" & bw_dat.males$state == "Washington DC"], 1))` years, of which `r round(min.value.DC, 1)` was due to `r min.cause.DC`.
+
+The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC.
+In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state ` (or `r round(sum3[dim(sum3)[1]-1, ]$LE_wbgap, 1)` in `r sum3[dim(sum3)[1]-1, ]$state ` when DC is excluded).
+By 2013, it ranged from `r round(sum4[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum4[dim(sum4)[1], ]$state ` (or `r round(sum4[dim(sum4)[1]-1, ]$LE_wbgap, 1)` in `r sum4[dim(sum4)[1]-1, ]$state` when DC is excluded).
+During this time period all of the causes contributed to narrowing the gap for most states.
+In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardivascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.
+
 **Figure 1a: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
@@
 ```{r figure1, fig.align = "center", echo=F, fig.height=9, fig.width=7}
-
-cod.df.males <- merge(subset(cod.df, sex == "Male"), bw_dat.males, by = c("state"))
-cod.df.males <- cod.df.males %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
-                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
-                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)
-
 y.labels <- levels(factor(cod.df.males$state.gdiff.order.1))
@@
 ```
-
-```{r, echo =F}
-sum2 <- BlackWhite_results %>% filter(year == 1983 & sex == "Male") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
-
-  min.value.DC <- min(cod.df$Contribution.to.change[cod.df$state == "Washington DC" & 
-                                            cod.df$era == "1983-1993" &
-                                            cod.df$sex == "Male" ])
-
-  min.cause.DC <- cod.df$COD[cod.df$state == "Washington DC" &
-                    cod.df$sex == "Male" &
-                    cod.df$era == "1983-1993" & 
-                    cod.df$Contribution.to.change == min.value.DC]
-```
-
-**Figure 1b** displays each state's male life expectancy gap in 1983 (vertical dashed line) vs.
-their gap in 1993 (solid dashed line).[Be careful with the interpretation: the x-axis is much wider for this plot because of DC.]
-
-During the second time period, from 1983 to 1993, the life expectancy gap for males in `r dim(bw_dat.males[bw_dat.males$era2.diff < 0, ])[1]` of the analysed `r dim(bw_dat.males)[1]` widened.
-For states such as New York, Maryland, New Jersey, and Connecticut, the biggest driver of the widening gap was communicable disease, reflecting the heightened risk of dying from HIV/AIDS in black males during the height of the AIDS epidemic.
-While both blacks and whites experienced a dramatic increase in mortality related to communicable diseases, blacks' rates were even higher.
-Injuries also contributed to the gap widening, most notably in Wisconsin, Maryland, and Louisiana.
-For many states, injury-related mortality increased for blacks but not for whites during this time period.
-Compared to the previous period, these two causes of death led to widwer rather than narrowed gaps.
-Cancer and cardiovascular disease contributed towards wider gaps, although the magnitude of the contribution of cardiovascular disease is quite hetergenous by state, with Mississippi, Minnesota, and Pennsylvania seeing the largest absolute contributions.
-All other causes contributed to a narrower gap, especially in New York, Pennsylvania, Connecticut, Mississippi, and Florida [note: This wasn't related to infant mortality improvements among blacks in NY, PA, but partially related to IMR improvements in CT, MS, and FL].
-Most notably, the Disrict of Columbia had the highest gap to begin and underwent the largest increase to their gap, which increased from 10.5 years to 15.9 years over this 10-year period, reflecting an increase of `r abs(round(bw_dat.males$era2.diff[bw_dat.males$sex == "Male" & bw_dat.males$state == "Washington DC"], 1))` years, of which `r round(min.value.DC, 1)` was due to `r min.cause.DC`.
 
@@
 
-```{r, echo =F}
-sum3 <- BlackWhite_results %>% filter(year == 1993 & sex == "Male") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
-
-sum4 <- BlackWhite_results %>% filter(year == 2013 & sex == "Male") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
-```
-
-- The gap in 1993 ranges from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state ` (or `r round(sum3[dim(sum3)[1]-1, ]$LE_wbgap, 1)` when DC is excluded).
-
-- By 2013, it ranges from `r round(sum4[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum4[dim(sum4)[1], ]$state ` (or `r round(sum4[dim(sum4)[1]-1, ]$LE_wbgap, 1)` when DC is excluded).
-
-- After 1993, the gap narrowed for all states shown, and in most cases all of the causes contributed to the narrowing of the gap.
-Notice again that in Wisconsin and DC, changes to CVD mortality counteract the gap narrowing, while in Arizona changes to communicable disease counteract the narrowing.
-
 **Figure 1c: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
 
-```{r figure1c, echo = F}
+```{r figure1c, echo = F, fig.height=9, fig.width=11}
 y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))

### 2017-02-10 Manuscript/Manuscript_COD_decomp.Rmd
wrote half of abstract

@@
 ---
-title: Exploring how changes in age at death and cause of death have narrowed the
+title: Exploring how changes in cause of death have narrowed the
   black-white life expectancy gap over forty-five years in the United States.
 output:
+  html_document: default
   word_document: default
-  html_document: default
 always_allow_html: yes
@@
 ###Abstract
+
+**Background**
+
+The difference in life expectancy between black and whites in the US has decreased over time.
+States have fared differently in their progress towards eliminating the life expectancy gap, with states such as New York undergoing greater reductions to the gap compared to states such as California and Texas.
+The objective of this study is to compare how changes to cause-specific mortality between 1969 and 2013 have resulted in relatively narrow life expectancy gaps in some states and  wide gaps in other states.
+
+**Methods and Findings**
+Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes according to year (1969-2013), state, age, sex, race (black or white), and six cause of death categories (Cancer, cardiovascular disease, communicable disease, non-communicable disease, injuries, and all other causes).
+We excluded 10 states with the smallest black populations, but included the District of Columbia.
+A bayesian autoregressive model was used to model smoothed mortality rates over time within each strata, and a truncated Poisson model was used to imputed death counts that were suppressed due to small cell sizes.
+We calculated life expectancy for blacks and whites within each strata using traditional life table methods, and decomposed the difference in life expectancy by cause of death using Arriaga's method.
+
+[Add findings]
+
+**Conclusions**
+[Add conclusion]
 
@@
 For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014].
-This study seeks to better understand these state-level differences, by examining how changes in the cause of death [*and age at death(?)*] over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013.
+This study seeks to better understand these state-level differences, by examining how changes in the cause of death over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013.
 Thus, common reasons for the decreasing life expectancy gap can be identified, and anomalous states can be examined.
@@
 
-The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC.
+The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Figure 1c**).
 In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state ` (or `r round(sum3[dim(sum3)[1]-1, ]$LE_wbgap, 1)` in `r sum3[dim(sum3)[1]-1, ]$state ` when DC is excluded).

### 2017-02-10 Code/plot_age_standardardized_mortality_trends.Rmd
update fig count

@@
 
-**Age-adjusted cancer-related mortality (per 100,000) in males, United States, 1969-2013**
+**Fig S3: Age-adjusted cancer-related mortality (per 100,000) in males, United States, 1969-2013**
 
@@
 
-**Age-adjusted cardiovascular-related mortality (per 100,000) in males, United States, 1969-2013**
+**Fig S4: Age-adjusted cardiovascular-related mortality (per 100,000) in males, United States, 1969-2013**
 
@@
 
-**Age-adjusted communicable disease-related mortality (per 100,000) in males, United States, 1969-2013**
+**Fig S5: Age-adjusted communicable disease-related mortality (per 100,000) in males, United States, 1969-2013**
 
@@
 
-**Age-adjusted injury-related mortality (per 100,000) in males, United States, 1969-2013**
+**Fig S6: Age-adjusted injury-related mortality (per 100,000) in males, United States, 1969-2013**
 
@@
 
-**Age-adjusted non-communicable disease-related mortality (per 100,000) in males, United States, 1969-2013**
+**Fig S7: Age-adjusted non-communicable disease-related mortality (per 100,000) in males, United States, 1969-2013**
 
@@
 
-**Age-adjusted mortality (per 100,000) in males for all other causes, United States, 1969-2013**
+**Fig S8: Age-adjusted mortality (per 100,000) in males for all other causes, United States, 1969-2013**
 
@@
 
-**Age-adjusted cancer-related mortality (per 100,000) in females, United States, 1969-2013**
+**Fig S9: Age-adjusted cancer-related mortality (per 100,000) in females, United States, 1969-2013**
 
@@
 
-**Age-adjusted cardiovascular-related mortality (per 100,000) in females, United States, 1969-2013**
+**Fig S10: Age-adjusted cardiovascular-related mortality (per 100,000) in females, United States, 1969-2013**
 
@@
 
-**Age-adjusted communicable disease-related mortality (per 100,000) in females, United States, 1969-2013**
+**Fig S11: Age-adjusted communicable disease-related mortality (per 100,000) in females, United States, 1969-2013**
 
@@
 
-**Age-adjusted injury-related mortality (per 100,000) in females, United States, 1969-2013**
+**Fig S12: Age-adjusted injury-related mortality (per 100,000) in females, United States, 1969-2013**
 
@@
 
-**Age-adjusted non-communicable disease-related mortality (per 100,000) in females, United States, 1969-2013**
+**Fig S13: Age-adjusted non-communicable disease-related mortality (per 100,000) in females, United States, 1969-2013**
 
@@
 
-**Age-adjusted mortality (per 100,000) in females for all other causes, United States, 1969-2013**
+**Fig S14: Age-adjusted mortality (per 100,000) in females for all other causes, United States, 1969-2013**
 

### 2017-02-10 Manuscript/Manuscript_COD_decomp.Rmd

@@
 
-##Riddell CA, Morrison KT, Kaufman JS, Harper S
+Corinne A Riddell, PhD
+
+Kathryn T Morrison, MSc
+
+Jay S Kaufman, PhD
+
+Sam Harper, PhD
+
+Department of Epidemiology, Biostatistics and Occupational Health, McGill University, Montreal, Canada.
+
+Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada.
+Email: <corinne.riddell@mail.mcgill.ca>
 
@@
 
-We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEER*Stat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.
+We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.
 The mortality data was originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.
@@
 Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013).
-These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes.
-This recoding is illustrated in eTable 1 in the eAppendix of the Supplement.
+These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**).
 When the number of deaths for a population is between 1 and 9, this data is suppressed by the software and produces a missing death count for that strata.
@@
 In order to understand whether states fared similarly over time, we divided the time frame into three eras which roughly encapsulate an initial period of increases in life expectancy for both races and sexes (1969-1983), an intermediate period of decreases and stagnating life expectancy for blacks but not for whites (1983-1993) and a final period of increasing life expectancy for both races (1993-2013).
-While these eras represent different lengths of time, they best capture three distinct periods of changes to the difference in life expectancy: an intial narrowing of the gap, followed by a widening of the gap that diminished previous progress, and the most recent twenty years of sustained narrowing (eAppendix Figures 1 and 2).
+While these eras represent different lengths of time, they best capture three distinct periods of changes to the difference in life expectancy: an intial narrowing of the gap, followed by a widening of the gap that diminished previous progress, and the most recent twenty years of sustained narrowing (**S1 Fig** and **S2 Fig**).
 
@@
 
-**Figure 1a** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs.
+**Fig 1** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs.
 their gap in 1983 (solid dashed line), with the change in the gap being represented by the number of years between these lines.
@@
 
-For example, in the **Figure 1a**, New York's gap changed from 8.8 years in 1969 to 6.3 in 1983.
+For example, in the **Fig 1**, New York's gap changed from 8.8 years in 1969 to 6.3 in 1983.
 Causes to the left of the dashed line -- All other causes (pink), injuries (blue), non-communicable diseases (teal), and communicable diseases (green) -- each contributed to narrowing the gap, while cardiovascular disease (red) and cancer (yellow) contributed to widening the gap.
@@
 In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `.
-Injuries was the largest contributor towards narrowing the gap during this time period, as both races underwent decreased rates of injury-related mortality, but blacks experienced larger decreases to their rates, especially in states with relatively higher rates in 1969 (**eAppendix B: Age-adjusted mortality by causes of death**).
+Injuries was the largest contributor towards narrowing the gap during this time period, as both races underwent decreased rates of injury-related mortality, but blacks experienced larger decreases to their rates, especially in states with relatively higher rates in 1969 (**S3-S14 Figs**).
 Minnesota and Nebraska are exceptions, where both states had relatively low rates of injuries among blacks in 1969 and smaller gaps.
 All other causes also contributed to narrower gaps in most states, but generally made a smaller contribution than injuries.
-In New York, all other causes made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks (**eFigure 3**).
+In New York, all other causes made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks (**S15 Fig**).
 Communicable diseases modestly contributed to narrowing the gap, except in Arizona and Minnesota.
@@
 
-During the second time period, from 1983 to 1993, the life expectancy gap for males in `r dim(bw_dat.males[bw_dat.males$era2.diff < 0, ])[1]` of the analysed `r dim(bw_dat.males)[1]` widened (**Figure 1b**).
+During the second time period, from 1983 to 1993, the life expectancy gap for males in `r dim(bw_dat.males[bw_dat.males$era2.diff < 0, ])[1]` of the analysed `r dim(bw_dat.males)[1]` widened (**Fig 2**).
 For states such as New York, Maryland, New Jersey, and Connecticut, the biggest driver of the widening gap was communicable disease, reflecting the heightened risk of dying from HIV/AIDS in black males during the height of the AIDS epidemic.
@@
 
-The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Figure 1c**).
+The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Fig 3**).
 In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state ` (or `r round(sum3[dim(sum3)[1]-1, ]$LE_wbgap, 1)` in `r sum3[dim(sum3)[1]-1, ]$state ` when DC is excluded).
@@
 
-**Figure 1a: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
+**Fig 1: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
 
@@
 
-**Figure 1b: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1983 and 1993 by US state**
+**Fig 2: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1983 and 1993 by US state**
 
-```{r figure1b, echo = F, fig.height=9, fig.width=11}
+```{r figure2, echo = F, fig.height=9, fig.width=11}
 # - The gap in 1983 ranges from `r round(sum2[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum2[1, ]$LE_wbgap_lcl, 1)`, `r round(sum2[1, ]$LE_wbgap_ucl, 1)`] in `r sum2[1, ]$state ` to `r round(sum2[dim(sum2)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum2[dim(sum2)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum2[dim(sum2)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum2[dim(sum2)[1], ]$state `.
@@
 
-**Figure 1c: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
+**Fig 3: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
 
-```{r figure1c, echo = F, fig.height=9, fig.width=11}
+```{r figure3, echo = F, fig.height=9, fig.width=11}
 y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
@@
 ```{r, fig.align="center", fig.height=5, fig.width=9, echo=F, eval = F}
-subplot(males.era1, males.era2, males.era3, shareY = T)
+males.era2.2 <- ggplotly(
+  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.1, x = new.start.2)) +
+    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.1 - 0.45, 
+                  ymax = state.gdiff.num.1 + 0.45, 
+                  xmax = new.finish.2, fill = COD), 
+              color = "white") +
+    geom_segment(data = bw_dat.males, 
+                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
+    geom_segment(data = bw_dat.males, 
+                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>% 
+  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
+
+males.era2.2
+
+subplot(males.era1, males.era2.2, shareY = T)
 ```
@@
 
-**Figure 2a: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
+**Fig 4: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
 
-```{r figure2a, fig.align = "center", echo=F}
+```{r figure4, fig.align = "center", echo=F}
 
@@
 
-**Figure 2b: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1983 and 1993 by US state**
+**Fig 5: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1983 and 1993 by US state**
 
@@
 
-**Figure 2c: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
+**Fig 6: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
 
-```{r figure2c, fig.align = "center", echo=F}
+```{r fig6, fig.align = "center", echo=F}
 y.labels.3 <- levels(factor(cod.df.females$state.gdiff.order.3))
@@
 
-###Supplemental Information
+###Supporting Information
 
-**eTable1: Cause of Death Recode**
+**S1 Table: Cause of death groupings**
 
@@
 
-**eFigure 1: Difference in life expectancy between black in white males, United States, 1969-2013**
+**S1 Text: Decomposing the change to the difference in life expectancy over time by cause of death: an example**
+To create.
+
+**S1 Fig: Difference in life expectancy between black in white males, United States, 1969-2013**
 
@@
 
-**eFigure 2: Difference in life expectancy between black in white females, United States, 1969-2013**
+**S2 Fig: Difference in life expectancy between black in white females, United States, 1969-2013**
 
@@
 
-**eExample 1: Decomposing the change to the difference in life expectancy over time by cause of death: an example**
+**S3-S14 Fig: Trends in age-standardized mortality by state, race, and sex, 1969-2013 (currently included in other file)**
 
-To create.
-
-**eFigure 3: Comparison of the contribution of age and cause of death to the difference in life expectancy between blacks and whites in New York, 1969 vs. 1983**
+**S15 Fig: Comparison of the contribution of age and cause of death to the difference in life expectancy between blacks and whites in New York, 1969 vs. 1983**
 

### 2017-02-10 Manuscript/Manuscript_COD_decomp.Rmd
more changes

@@
 
-**Background**
-
+**Background:**
 The difference in life expectancy between black and whites in the US has decreased over time.
@@
 
-**Methods and Findings**
+**Methods and Findings:**
 Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes according to year (1969-2013), state, age, sex, race (black or white), and six cause of death categories (Cancer, cardiovascular disease, communicable disease, non-communicable disease, injuries, and all other causes).
@@
 
-**Conclusions**
+**Conclusions:**
 [Add conclusion]
@@
 
-For each era, we calculated the degree that the life expectancy gap has narrowed or widened over the era and decomposed this *change* to the *difference* in the life expectnacy between blacks and whites according to cause of death.
+For each era, we calculated the degree that the life expectancy gap has narrowed or widened over the era and decomposed this *change* to the *difference* in the life expectancy between blacks and whites according to cause of death.
 To do so, we simply subtracted the contribution (in years) that each cause of death category had made to the difference in life expectancy at the beginning of the era from that made at the beginning of the era.
-An example of the calculation is found in the eAppendix.
+An example of the calculation is found in **S1 Text**.
 
@@
               color = "white") +
-#    facet_grid(.~ Census_Region) +
     geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
         geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
-   # geom_segment(data = bw_dat.males,
-  #               aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
-  #  geom_segment(data = bw_dat.males,
-  #               aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
+    theme_minimal() + 
+    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
+    scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
   layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
@@
                  aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>% 
+    theme_minimal() + 
+    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
+    scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2), minor_breaks = NULL) %>% 
   layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
@@
                  aes(x = third.gap, xend= third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>% 
+    theme_minimal() + 
+    scale_x_continuous(breaks = seq(0, 16, by = 2), labels = seq(0, 16, by = 2)) +
+    scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>% 
   layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))
@@
 **S1 Text: Decomposing the change to the difference in life expectancy over time by cause of death: an example**
+
 To create.
+1. First show the data for NY in 1993 and 2013 for black and white males (smoothed deaths, population, for each age strata).
+
+2. Create life tables for blacks and whites in the two years (4 tables)
+
+3. Calculate the LE gap for each year (two numbers)
+
+4. Decompose by age the BW gap within year (two decomps)
+
+5. Decompose by age and COD the BW gap within year (two decomps)
+
+6. Marginal COD decomp within year (two decomps)
+
+7. Subtract the COD decomp from year 1 from year 2. (one change in decomp)
 

### 2017-02-14 Manuscript/Manuscript_COD_decomp.Rmd
more changes
first set of JK revisions

--- 2017-02-10 Manuscript/Manuscript_COD_decomp.Rmd
@@
 title: Exploring how changes in cause of death have narrowed the
-  black-white life expectancy gap over forty-five years in the United States.
+  black-white life expectancy gap in the United States.
 output:
@@
 
-Kathryn T Morrison, MSc
+Kathryn T Morrison, PhD
 
@@
 
-The difference in life expectancy between black and whites in the US has decreased over time.
-States have fared differently in their progress towards eliminating the life expectancy gap, with states such as New York undergoing greater reductions to the gap compared to states such as California and Texas.
-The objective of this study is to compare how changes to cause-specific mortality between 1969 and 2013 have resulted in relatively narrow life expectancy gaps in some states and  wide gaps in other states.
+The overall difference in life expectancy between black and whites in the US has generally decreased over time.
+States have fared differently in their progress towards eliminating the life expectancy gap, with states such as New York experiencing greater reductions to the gap compared to states such as California and Texas.
+The objective of this study is to compare how changes to cause-specific mortality between 1969 and 2013 have resulted in narrower life expectancy gaps in some states and  wider gaps in other states.
 
@@
 We excluded 10 states with the smallest black populations, but included the District of Columbia.
-A bayesian autoregressive model was used to model smoothed mortality rates over time within each strata, and a truncated Poisson model was used to imputed death counts that were suppressed due to small cell sizes.
-We calculated life expectancy for blacks and whites within each strata using traditional life table methods, and decomposed the difference in life expectancy by cause of death using Arriaga's method.
+A Bayesian autoregressive model was used to model smoothed mortality rates over time within each stratum, and a truncated Poisson model was used to impute death counts that were suppressed due to small cell sizes.
+We calculated life expectancy for blacks and whites within each stratum using traditional life table methods, and decomposed the difference in life expectancy by cause of death using Arriaga's method.
 
@@
 Non-Hispanic whites lived 3.6 years longer than Non-Hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf].
-While both black and white Americans have experienced increases in life expectancy since X, the difference between their life expectancies has varied.
-During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, heart disease [Harper 2007].
-The gap continued to narrow during the 2000s, but states differed in the magnitude of improvement that was made.
+While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.
+During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [Harper 2007].
+The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement.
 For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014].
 This study seeks to better understand these state-level differences, by examining how changes in the cause of death over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013.
-Thus, common reasons for the decreasing life expectancy gap can be identified, and anomalous states can be examined.
+Common reasons for the decreasing life expectancy gap can thereby be identified, and anomalous states can be examined.
+Ideally, one would like to identify policies or social conditions in highly successful states that help to explain their unusual performance.
 
@@
 We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.
-The mortality data was originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.
-The population data was derived by the NCHS using intercensal bridged-race population estimates.
-This study using data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death.
+The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.
+The population data were derived by the NCHS using intercensal bridged-race population estimates.
+This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death.
 Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013).
 These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**).
-When the number of deaths for a population is between 1 and 9, this data is suppressed by the software and produces a missing death count for that strata.
-However, death counts of exactly 0 or greater than 9 are known precisely.
+When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.
 
 For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming.
-These states contained less than 0.4% of the blacks in the United States in 2010 [ref].
+These states contained less than 0.4% of the black population in the United States in 2010 [ref].
 We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since the quality of the coding of ethnicity varied by state over the time period included [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf].
@@
 
-The data produced two statistical issues that were addressed by our statistical methods.
+The data structure produced two statistical challenges that were addressed by our quantitative approach.
 First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9).
-Second, several states had small blacks populations which produced even smaller population sizes in each sex-age category.
-In these cases, the estimation of the sex-age mortality rates is imprecise, and produces abrupt changes in the estimation of mortality rates over time that are unrealistic.
+Second, several states had small black populations which produced sparse counts in each sex-age category.
+In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic.
 To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.
@@
 Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population.
-The mortality rates over time within each state-age-sex-race category are modelled autoregressively, where the log mortality rate in the current year was modelled as a normal variable centred at the log mortality rate estimated from the previous year.
-Truncated Poisson regression was used to impute the supressed death counts, by specifying that the supressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9).
-These models were run separately for every combination of state, race, and sex, and produced death counts that had been smoothed over time for every cause of death within each age group.
+The mortality rates over time within each state-age-sex-race category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year.
+Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9).
+These models were run separately for every combination of state, race, and sex, and produced death counts that were smoothed over time for every cause of death within each age group.
 
@@
 Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex.
-The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in every state, year, and sex.
+The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.
 
-To decompose the difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref].
-First, the decomposition analysis partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks.
+To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref].
+First, the model partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks.
 For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites.
@@
 For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old.
-We summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference.
+We summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference, for each state and for each sex.
 
@@
 
-For each era, we calculated the degree that the life expectancy gap has narrowed or widened over the era and decomposed this *change* to the *difference* in the life expectnacy between blacks and whites according to cause of death.
+For each era, we calculated the degree that the life expectancy gap has narrowed or widened and decomposed this *change* to the *difference* in the life expectancy between blacks and whites according to cause of death.
 To do so, we simply subtracted the contribution (in years) that each cause of death category had made to the difference in life expectancy at the beginning of the era from that made at the beginning of the era.
@@
 
-*To re-write*: To calculate 95% credible intervals, we sampled 1000 times from the posterior distribution to create 1000 samples for each strata.
-We performed all calculated on each sample and took the median of the estimates as the estimate and the 2.5 and 97.5th qunatiles as the lower and upper bound for the 95% CI.
+*To re-write*: To calculate 95% credible intervals, we sampled 1000 times from the posterior distribution to create 1000 samples for each stratum.
+We performed all calculations on each sample and took the median of the estimates as the estimate and the 2.5 and 97.5th quantiles as the lower and upper bound for the 95% CI.
 Anything about thinning?
@@
 All data analyses were performed using R version 3.3.1.
-The bayesian smoothing was performed using `JAGS` (version X.X), and the life expectancy and decomposition analysis were performed using user-written functions found in the Supplement.
+The Bayesian smoothing was performed using `JAGS` (version X.X), and the life expectancy and decomposition analysis were performed using user-written functions found in the Supplement.
 
@@
 **Fig 1** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs.
-their gap in 1983 (solid dashed line), with the change in the gap being represented by the number of years between these lines.
+their gap in 1983 (solid dashed line), with the change in the gap being represented by the number of years between these lines along the x-axis.
 This gap is decomposed by cause of death, where causes to the left of the dashed line contributed to narrowing the gap and causes to the right contributed to widening the gap.
-The length of these coloured bars represent the number of years attributable to the cause.
+The length of these coloured bars represent the number of years attributable to the cause along the x-axis.
 
@@
 
-In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `.
-Injuries was the largest contributor towards narrowing the gap during this time period, as both races underwent decreased rates of injury-related mortality, but blacks experienced larger decreases to their rates, especially in states with relatively higher rates in 1969 (**S3-S14 Figs**).
-Minnesota and Nebraska are exceptions, where both states had relatively low rates of injuries among blacks in 1969 and smaller gaps.
-All other causes also contributed to narrower gaps in most states, but generally made a smaller contribution than injuries.
-In New York, all other causes made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks (**S15 Fig**).
-Communicable diseases modestly contributed to narrowing the gap, except in Arizona and Minnesota.
-Generally, cardivascular disease contributed towards increasing the gap, and reflects slower declines for blacks in cardiovascular mortality than for whites.
-Cancer generally contributed towards increasing the gap, and largely reflects increased cancer-related mortality for both races, with blacks undergoing steeper changes.
-In most states, non-comunicable diseases contributed a small amount towards increasing the gap, although it decreased the gap among four states, most notably in New York.
+In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `.
+Injury was the largest contributor towards narrowing the gap during this time period, as both race groups experienced decreased rates of injury-related mortality, but blacks had larger rates decreases, especially in states with relatively higher rates in 1969 (**S3-S14 Figs**).
+A few states, including Minnesota and Nebraska, had relatively low rates of injuries among blacks in 1969 and smaller gaps.
+The residual category also contributed to narrower gaps in most states, but generally made a smaller contribution than injuries.
+In New York, the residual category made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks (**S15 Fig**).
+Communicable diseases contributed modestly to narrowing the gap, except in Arizona and Minnesota.
+Generally, cardiovascular disease contributed towards increasing the gap, and reflects slower declines for blacks in cardiovascular mortality than for whites.
+Cancer contributed towards increasing the gap in most states, and largely reflects increased cancer-related mortality for both races during this period, with blacks undergoing steeper changes.
+In most states, non-communicable diseases contributed a small amount towards increasing the gap, although it decreased the gap among four states, most notably in New York.
 
-For `r dim(bw_dat.males[bw_dat.males$era1.diff > 0,])[1]` states, the gap decreased between 1969 and 1983, with the largest decreases in `r diff[dim(diff)[1], ]$state` (decrease in the gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)`, `r round(max1,1)` years, of which is due to `r max.cause1`), `r diff[dim(diff)[1]-1, ]$state` (change of `r round(diff[dim(diff)[1]-1, ]$era1.diff, 1)`, `r round(max2,1)` due to `r max.cause2`), `r diff[dim(diff)[1]-2, ]$state` (change of `r round(diff[dim(diff)[1]-2, ]$era1.diff, 1)`, `r round(max3,1)` due to `r max.cause3`), `r diff[dim(diff)[1]-3, ]$state` (change of `r round(diff[dim(diff)[1]-3, ]$era1.diff, 1)`, `r round(max4,1)` due to `r max.cause4`), and `r diff[dim(diff)[1]-4, ]$state` (change of `r round(diff[dim(diff)[1]-4, ]$era1.diff, 1)`, `r round(max5,1)` due to `r max.cause5`).
+For `r dim(bw_dat.males[bw_dat.males$era1.diff > 0,])[1]` of the `r dim(bw_dat.males)[1]` states, the gap decreased between 1969 and 1983, with the largest decreases in `r diff[dim(diff)[1], ]$state` (decrease in the gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)` years, `r round(max1,1)` years, of which is due to `r max.cause1`), `r diff[dim(diff)[1]-1, ]$state` (change of `r round(diff[dim(diff)[1]-1, ]$era1.diff, 1)` years, `r round(max2,1)` due to `r max.cause2`), `r diff[dim(diff)[1]-2, ]$state` (change of `r round(diff[dim(diff)[1]-2, ]$era1.diff, 1)` years, `r round(max3,1)` due to `r max.cause3`), `r diff[dim(diff)[1]-3, ]$state` (change of `r round(diff[dim(diff)[1]-3, ]$era1.diff, 1)` years, `r round(max4,1)` due to `r max.cause4`), and `r diff[dim(diff)[1]-4, ]$state` (change of `r round(diff[dim(diff)[1]-4, ]$era1.diff, 1)` years, `r round(max5,1)` due to `r max.cause5`).
 
 The gap increased for `r dim(bw_dat.males[bw_dat.males$era1.diff < 0,])[1]` states.
-Most notably in `r diff$state[1]` (increase in gap of `r round(abs(diff$era1.diff[1]), 1)`, `r abs(round(min1,1))` due to `r min.cause1`), `r diff$state[2]` (increase in gap of `r round(abs(diff$era1.diff[2]), 1)`, `r abs(round(min2,1))` due to `r min.cause2`), `r diff$state[3]` (increase in gap of `r round(abs(diff$era1.diff[3]), 1)`, `r abs(round(min3,1))` due to `r min.cause3`), `r diff$state[4]` (increase in gap of `r round(abs(diff$era1.diff[4]), 1)`, `r abs(round(min4,1))` due to `r min.cause4`).
-For these states, injuries do not contribute as much to narrowing the gap as in other states (and in Minnesota, injuries widens the gap), so these gap increases appear to be due to a combination of changes to cardiovascular-related mortality making blacks relatively worse off and less changes to injuries than experienced by other states.
+Most notably in `r diff$state[1]` (increase in gap of `r round(abs(diff$era1.diff[1]), 1)` years, `r abs(round(min1,1))` due to `r min.cause1`), `r diff$state[2]` (increase in gap of `r round(abs(diff$era1.diff[2]), 1)` years, `r abs(round(min2,1))` due to `r min.cause2`), `r diff$state[3]` (increase in gap of `r round(abs(diff$era1.diff[3]), 1)` years, `r abs(round(min3,1))` due to `r min.cause3`), `r diff$state[4]` (increase in gap of `r round(abs(diff$era1.diff[4]), 1)` years, `r abs(round(min4,1))` due to `r min.cause4`).
+For these states, injuries do not contribute as much to narrowing the gap as in other states (and in Minnesota, injuries widened the gap), so these disparity increases appear to be due to a combination of changes to cardiovascular-related mortality making blacks relatively worse off and less changes to injuries than experienced by other states.
 
-During the second time period, from 1983 to 1993, the life expectancy gap for males in `r dim(bw_dat.males[bw_dat.males$era2.diff < 0, ])[1]` of the analysed `r dim(bw_dat.males)[1]` widened (**Fig 2**).
-For states such as New York, Maryland, New Jersey, and Connecticut, the biggest driver of the widening gap was communicable disease, reflecting the heightened risk of dying from HIV/AIDS in black males during the height of the AIDS epidemic.
+During the second time period, from 1983 to 1993, the life expectancy gap for males widened in `r dim(bw_dat.males[bw_dat.males$era2.diff < 0, ])[1]` of the `r dim(bw_dat.males)[1]` states (**Fig 2**).
+For states such as New York, Maryland, New Jersey, and Connecticut, the biggest driver of the widening gap was communicable disease, reflecting the heightened burden from HIV/AIDS in black males during the height of the AIDS epidemic.
 While both blacks and whites experienced a dramatic increase in mortality related to communicable diseases, blacks' rates were even higher.
@@
 Cancer and cardiovascular disease contributed towards wider gaps, although the magnitude of the contribution of cardiovascular disease is quite hetergenous by state, with Mississippi, Minnesota, and Pennsylvania seeing the largest absolute contributions.
-All other causes contributed to a narrower gap, especially in New York, Pennsylvania, Connecticut, Mississippi, and Florida [note: This wasn't related to infant mortality improvements among blacks in NY, PA, but partially related to IMR improvements in CT, MS, and FL].
-Most notably, the Disrict of Columbia had the highest gap to begin and underwent the largest increase to their gap, which increased from 10.5 years to 15.9 years over this 10-year period, reflecting an increase of `r abs(round(bw_dat.males$era2.diff[bw_dat.males$sex == "Male" & bw_dat.males$state == "Washington DC"], 1))` years, of which `r round(min.value.DC, 1)` was due to `r min.cause.DC`.
+The residual category contributed to a narrower gap, especially in New York, Pennsylvania, Connecticut, Mississippi, and Florida [note: This wasn't related to infant mortality improvements among blacks in NY, PA, but partially related to IMR improvements in CT, MS, and FL].
+Most notably, the District of Columbia had the highest gap to begin and underwent the largest increase to their gap, which increased from 10.5 years to 15.9 years over this 10-year period, reflecting an increase in the gap of `r abs(round(bw_dat.males$era2.diff[bw_dat.males$sex == "Male" & bw_dat.males$state == "Washington DC"], 1))` years, of which `r round(min.value.DC, 1)` was due to `r min.cause.DC`.
 
 The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Fig 3**).
-In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state ` (or `r round(sum3[dim(sum3)[1]-1, ]$LE_wbgap, 1)` in `r sum3[dim(sum3)[1]-1, ]$state ` when DC is excluded).
-By 2013, it ranged from `r round(sum4[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum4[dim(sum4)[1], ]$state ` (or `r round(sum4[dim(sum4)[1]-1, ]$LE_wbgap, 1)` in `r sum4[dim(sum4)[1]-1, ]$state` when DC is excluded).
+In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` years [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum3[dim(sum3)[1], ]$state `.
+By 2013, the gap ranged from `r round(sum4[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum4[dim(sum4)[1], ]$state `.
 During this time period all of the causes contributed to narrowing the gap for most states.
-In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardivascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.
+In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardiovascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.
 
@@
 
-- However, increases were found in Arizona, Oregon, Minnesota, Washington, and Wisconsin, largely due to cardiovascular disease.
+- However, increases were found in Arizona, Oregon, Minnesota, Washington, and Wisconsin, largely due to CVD.
 As you can see, in some states CVD contributed to excacerbating the gap while in most others it makes a large contribution to narrowing the gap.

### 2017-02-14 Manuscript/Manuscript_COD_decomp.Rmd

@@
   black-white life expectancy gap in the United States.
+
 output:
+  word_document: default
   html_document: default
-  word_document: default
 always_allow_html: yes
@@
 To do so, we simply subtracted the contribution (in years) that each cause of death category had made to the difference in life expectancy at the beginning of the era from that made at the beginning of the era.
-An example of the calculation is found in the eAppendix.
+An example of the calculation is found in **S1 Text**.
 

### 2017-02-14 Code/calculate_std_mort_rates.Rmd
update std mort plots

@@
 ```
+
+```{r calculate_difference_in_mortality}
+mortality.rates.white <- mortality.rates %>% filter(race == "White") %>% 
+  select(sex, state, year, COD, post.samp, age.std.mortality.rate, rate.per.100k) %>%
+  rename(age.std.mortality.rate.white = age.std.mortality.rate,
+         rate.per.100k.white = rate.per.100k)
+
+mortality.rates.black <- mortality.rates %>% filter(race == "Black") %>% 
+  rename(age.std.mortality.rate.black = age.std.mortality.rate,
+         rate.per.100k.black = rate.per.100k)
+
+mortality.rates.wide <- merge(mortality.rates.black, mortality.rates.white, by = c("post.samp", "sex", "state", "year", "COD"))
+
+mortality.rates.wide <- mortality.rates.wide %>% mutate(rate.difference = rate.per.100k.black - rate.per.100k.white)
+
+write.csv(mortality.rates.wide, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_wide.csv")
+```

### 2017-02-14 Code/plot_age_standardardized_mortality_trends.Rmd

--- 2017-02-08 Code/plot_age_standardardized_mortality_trends.Rmd
@@
 mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates.csv")
+mortality.rates.wide <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_wide.csv")
 ```
 
-**Age-adjusted injury-related mortality (per 100,000) in white males, United States, 1969-2013**
-
-```{r, echo=F, fig.width=11, fig.height=9}
-ggplotly(ggplot(subset(mortality.rates, sex == "Male" & race == "White" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
+```{r, echo=F, fig.width=11}
+for(sex_i in c("Male", "Female")) {
+  for(cod in levels(mortality.rates$COD)) {
+    for(region in levels(mortality.rates$Census_Region)) {
+      data.sub <- subset(mortality.rates, sex == sex_i & post.samp == 1 & COD == cod  & Census_Region == region)
+      a <- ggplot(data.sub, 
+                      aes(y=rate.per.100k, x = year)) + 
+                 geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
+                 geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
                  geom_line(aes(col = state, lty = race)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
+                 facet_wrap( ~ state, ncol = 5) +
                  ylab("Age-standardized mortality rate (per 100,000)") +
                  xlab("Year") + 
-           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Male" & race == "White" & post.samp == 1 & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
-
-ggplotly(ggplot(subset(mortality.rates, sex == "Male" & race == "White" & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_line(aes(col = state, lty = factor(post.samp))) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Age-standardized mortality rate (per 100,000)") +
+                 ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod, " (Census Region: ", region, ")"))
+      print(a)
+    }
+    for(region in levels(mortality.rates$Census_Region)) {    
+          b <- ggplot(subset(mortality.rates.wide, sex == sex_i & post.samp == 1 & COD == cod & Census_Region == region), 
+                      aes(y=rate.difference, x = year)) + 
+                 geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
+                 geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
+                 geom_hline(aes(yintercept = 0), col = "#636363") +
+                 geom_line(aes(col = state)) +
+                 facet_wrap( ~ state, ncol = 5) +
+                 ylab("Black-white difference (per 100,000)") +
                  xlab("Year") + 
-           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Male" & race == "White" & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
+                 ggtitle(paste0("Black-white difference in age-adjusted mortality (per 100,000) in ", 
+                                sex_i, " for ", cod, " (Census Region: ", region, ")"))
+          print(b)
+    }
+  }
+}
 ```
-
-**Age-adjusted injury-related mortality (per 100,000) in black males, United States, 1969-2013**
-
-```{r, echo=F, fig.width=11, fig.height=9}
-ggplotly(ggplot(subset(mortality.rates, sex == "Male" & race == "Black" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_line(aes(col = state, lty = race)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Age-standardized mortality rate (per 100,000)") +
-           xlab("Year") +
-           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Male" & race == "Black" & post.samp == 1 & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
-
-ggplotly(ggplot(subset(mortality.rates, sex == "Male" & race == "Black" & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_line(aes(col = state, lty = factor(post.samp))) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Age-standardized mortality rate (per 100,000)") +
-           xlab("Year") +
-           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Male" & race == "Black" & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
-```
-
-**Age-adjusted injury-related mortality (per 100,000) in white females, United States, 1969-2013**
-
-```{r, echo=F, fig.width=11, fig.height=9}
-ggplotly(ggplot(subset(mortality.rates, sex == "Female" & race == "White" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_line(aes(col = state, lty = race)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Age-standardized mortality rate (per 100,000)") +
-           xlab("Year") +
-           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Female" & race == "White" & post.samp == 1 & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
-```
-
-**Age-adjusted injury-related mortality (per 100,000) in black females, United States, 1969-2013**
-
-```{r, echo=F, fig.width=11, fig.height=9}
-ggplotly(ggplot(subset(mortality.rates, sex == "Female" & race == "Black" & post.samp == 1 & COD == "Injuries"), aes(y=rate.per.100k, x = year)) + 
-           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
-           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
-           geom_line(aes(col = state, lty = race)) +
-           facet_wrap( ~ Census_Division, ncol = 3) +
-           ylab("Age-standardized mortality rate (per 100,000)") +
-           xlab("Year") +
-           geom_text(data = subset(mortality.rates, year == 2013 & sex == "Female" & race == "Black" & post.samp == 1 & COD == "Injuries"), 
-                     aes(label = stabbrs), check_overlap = T, size = 2.5))
-```

### 2017-02-14 Code/plot_age_standardardized_mortality_trends.Rmd
make plots bettah

@@
 mortality.rates.wide <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_wide.csv")
+
+mortality.rates <- mortality.rates %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
+mortality.rates.wide <- mortality.rates.wide %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
 ```
 
-```{r, echo=F, fig.width=11}
+```{r, echo=F, fig.width=11, fig.height=11}
+for(sex_i in c("Male", "Female")) {
+  for(cod in levels(mortality.rates$COD)) {
+    data.sub <- subset(mortality.rates, sex == sex_i & post.samp == 1 & COD == cod)
+    a <- ggplot(data.sub, 
+                aes(y=rate.per.100k, x = year)) + 
+      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
+      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
+      geom_line(aes(col = Census_Division, lty = race)) +
+      facet_wrap( ~ state.reorder, ncol = 5) +
+      ylab("Age-standardized mortality rate (per 100,000)") +
+      xlab("Year") + 
+      ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))
+
+    print(a)
+
+    b <- ggplot(subset(mortality.rates.wide, sex == sex_i & post.samp == 1 & COD == cod), 
+                aes(y=rate.difference, x = year)) + 
+      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
+      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
+      geom_hline(aes(yintercept = 0), col = "#636363") +
+      geom_line(aes(col = Census_Division)) +
+      facet_wrap( ~ state.reorder, ncol = 5) +
+      ylab("Black-white difference (per 100,000)") +
+      xlab("Year") + 
+      ggtitle(paste0("Black-white difference in age-adjusted mortality (per 100,000) in ", 
+                     sex_i, " for ", cod))
+    print(b)
+
+  }
+}
+```
+
+```{r, echo=F, fig.width=11, eval = F}
 for(sex_i in c("Male", "Female")) {

### 2017-02-15 Code/plot_age_standardardized_mortality_trends.Rmd
fixed weird facet display issue

@@
       data.sub <- subset(mortality.rates, sex == sex_i & post.samp == 1 & COD == cod  & Census_Region == region)
-      a <- ggplot(data.sub, 
+      a <- ggplotly(ggplot(data.sub, 
                       aes(y=rate.per.100k, x = year)) + 
@@
                  geom_line(aes(col = state, lty = race)) +
-                 facet_wrap( ~ state, ncol = 5) +
+                 facet_wrap( ~ state, scales = "free_x") +
                  ylab("Age-standardized mortality rate (per 100,000)") +
                  xlab("Year") + 
-                 ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod, " (Census Region: ", region, ")"))
+                 ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod, " (Census Region: ", region, ")")))
       print(a)
@@
     for(region in levels(mortality.rates$Census_Region)) {    
-          b <- ggplot(subset(mortality.rates.wide, sex == sex_i & post.samp == 1 & COD == cod & Census_Region == region), 
+          b <- ggplotly(ggplot(subset(mortality.rates.wide, sex == sex_i & post.samp == 1 & COD == cod & Census_Region == region), 
                       aes(y=rate.difference, x = year)) + 
@@
                  geom_line(aes(col = state)) +
-                 facet_wrap( ~ state, ncol = 5) +
+                 facet_wrap( ~ state, scales = "free_x") +
                  ylab("Black-white difference (per 100,000)") +
@@
                  ggtitle(paste0("Black-white difference in age-adjusted mortality (per 100,000) in ", 
-                                sex_i, " for ", cod, " (Census Region: ", region, ")"))
+                                sex_i, " for ", cod, " (Census Region: ", region, ")")))
           print(b)

### 2017-02-15 Code/calculate_std_mort_rates.Rmd
set 1 agestd rates

--- 2017-02-07 Code/calculate_std_mort_rates.Rmd
@@
 library(gridExtra)
-```
 
-```{r}
-load("~/BW_results/results_set_1.Rdata")
-r.All.set1 <- r.All %>% filter(post.samp %in% c(1,2))
+source("~/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
 
-for(i in 2:8){
-  print(i)
-  load(paste0("~/BW_results/results_set_", i, ".Rdata"))
-  assign(paste0("r.All.set", i), r.All %>% filter(post.samp %in% c(1,2)))
-  rm(r.All)
-}
-
-r.All.post1.2 <- rbind(r.All.set1, r.All.set2, r.All.set3, r.All.set4, 
-                      r.All.set5, r.All.set6, r.All.set7, r.All.set8)
 ```
@@
 
-```{r merge}
-r.All.post1.2 <- merge(r.All.post1.2, std.pop, by = "age_minbin")
+```{r calculate_age_std_mort_rates_and_bw_diff}
+#this take a long time (a few hours!)
 
-mortality.rates <- r.All.post1.2 %>% 
+system.time(
+  for(i in 1:8){
+    print(i)
+    load(paste0("~/BW_results/results_set_", i, ".Rdata"))
+    r.All <- merge(r.All, std.pop, by = "age_minbin")
+
+    #returns dataset with new variable "age.std.mortality.rate" 
+    mortality.rates <- r.All %>% 
       group_by(post.samp, state.n, year.n, race.n, sex.n, COD) %>% 
       arrange(age_minbin) %>%
-  summarise(age.std.mortality.rate = sum(age.weight*smoothed_rate),
+      summarise(age.std.mortality.rate = sum(age.weight*smoothed_rate), #the sum is across age bins
                 race = first(race),
                 sex = first(sex),
-            state = first(state))
+                state = first(state),
+                year = first(year))
 
-ggplot(subset(mortality.rates, post.samp == 1 & state == "Alabama"), aes(y = age.std.mortality.rate, x = year.n)) + 
-  geom_line(aes(col = COD)) + facet_grid(race ~ sex)
+    mortality.rates.summary <- mortality.rates %>% group_by(state.n, year.n, race.n, sex.n, COD) %>%
+      summarise(race = first(race),
+                sex = first(sex),
+                state = first(state),
+                year = first(year),
+                age.std.mortality.rate_LCL = quantile(age.std.mortality.rate, 0.025), #summarize across posterior samples (post.samp)
+                age.std.mortality.rate_med = quantile(age.std.mortality.rate, 0.5),
+                age.std.mortality.rate_UCL = quantile(age.std.mortality.rate, 0.0975))
 
-write.csv(r.All.post1.2, "~/BlackWhiteMortalityGap/Data/states_combined_post_samp_1_2.csv")
-write.csv(mortality.rates, "~/BlackWhiteMortalityGap/Data/mortality_rates.csv")
+    #tidy mortality rates summary
+    COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
+                    "Non-communicable", "Injuries", "All other causes")
+
+    mortality.rates.summary$COD <- factor(mortality.rates.summary$COD, COD.levels)
+
+    mortality.rates.summary <- add_Census_Division(mortality.rates.summary, state)
+    mortality.rates.summary <- add_Census_Region(mortality.rates.summary, state)
+    mortality.rates.summary$stabbrs <- state.abb[match(mortality.rates.summary$state, state.name)]
+    mortality.rates.summary$rate.per.100k <- mortality.rates.summary$age.std.mortality.rate_med*100000
+    mortality.rates.summary <- mortality.rates.summary %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
+
+    #compute black-white difference in mortality
+    mortality.rates.white <- mortality.rates %>% filter(race == "White") %>% 
+      ungroup() %>%
+      select(sex, state, year, COD, post.samp, age.std.mortality.rate) %>%
+      rename(age.std.mortality.rate.white = age.std.mortality.rate)
+
+    mortality.rates.black <- mortality.rates %>% filter(race == "Black") %>% 
+      rename(age.std.mortality.rate.black = age.std.mortality.rate)
+
+    mortality.rates.wide <- merge(mortality.rates.black, mortality.rates.white, by = c("post.samp", "sex", "state", "year", "COD"))
+
+    mortality.rates.wide <- mortality.rates.wide %>% mutate(rate.difference = age.std.mortality.rate.black - age.std.mortality.rate.white)
+
+    mortality.rates.difference <- mortality.rates.wide %>% 
+      group_by(state.n, year.n, sex.n, COD) %>%
+      summarise(rate.difference_LCL = quantile(rate.difference, 0.025), #summarize across posterior samples (post.samp)
+                rate.difference_med = quantile(rate.difference, 0.5),
+                rate.difference_UCL = quantile(rate.difference, 0.0975),
+                sex = first(sex),
+                state = first(state),
+                year = first(year))
+
+    mortality.rates.difference %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
+
+    write.csv(mortality.rates.summary, paste0("~/BlackWhiteMortalityGap/Data/mortality_rates_", i, ".csv"))
+    write.csv(mortality.rates.difference, paste0("~/BlackWhiteMortalityGap/Data/mortality_rates_diff_", i, ".csv"))
+  }
+)
 ```

### 2017-02-15 Code/plot_age_standardardized_mortality_trends.Rmd
set 1 agestd rates

--- 2017-02-14 Code/plot_age_standardardized_mortality_trends.Rmd
@@
 mortality.rates.wide <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_wide.csv")
-
-mortality.rates <- mortality.rates %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
-mortality.rates.wide <- mortality.rates.wide %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
 ```
@@
 ```
-
-```{r, echo=F, fig.width=11, eval = F}
-for(sex_i in c("Male", "Female")) {
-  for(cod in levels(mortality.rates$COD)) {
-    for(region in levels(mortality.rates$Census_Region)) {
-      data.sub <- subset(mortality.rates, sex == sex_i & post.samp == 1 & COD == cod  & Census_Region == region)
-      a <- ggplot(data.sub, 
-                      aes(y=rate.per.100k, x = year)) + 
-                 geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
-                 geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
-                 geom_line(aes(col = state, lty = race)) +
-                 facet_wrap( ~ state, ncol = 5) +
-                 ylab("Age-standardized mortality rate (per 100,000)") +
-                 xlab("Year") + 
-                 ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod, " (Census Region: ", region, ")"))
-      print(a)
-    }
-    for(region in levels(mortality.rates$Census_Region)) {    
-          b <- ggplot(subset(mortality.rates.wide, sex == sex_i & post.samp == 1 & COD == cod & Census_Region == region), 
-                      aes(y=rate.difference, x = year)) + 
-                 geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
-                 geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
-                 geom_hline(aes(yintercept = 0), col = "#636363") +
-                 geom_line(aes(col = state)) +
-                 facet_wrap( ~ state, ncol = 5) +
-                 ylab("Black-white difference (per 100,000)") +
-                 xlab("Year") + 
-                 ggtitle(paste0("Black-white difference in age-adjusted mortality (per 100,000) in ", 
-                                sex_i, " for ", cod, " (Census Region: ", region, ")"))
-          print(b)
-    }
-  }
-}
-```

### 2017-02-15 Manuscript/Manuscript_COD_decomp.Rmd
update std mort plots

@@
 ---
-title: Exploring how changes in cause of death have narrowed the
-  black-white life expectancy gap in the United States.
+title: Exploring how changes in cause of death have narrowed the black-white life expectancy gap in the United States.
 
@@
 The life expectancy of a population is a barometer for overall health.
-In 2014, life expectancy in the United States was 78.7 years, but hidden within this figure are differences by race.
-Non-Hispanic whites lived 3.6 years longer than Non-Hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf].
-While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.
+In 2014, life expectancy in the United States was 78.8 years, but hidden within this figure are differences by race.
+Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf].
+While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.
 During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [Harper 2007].
@@
 
-```{r prep_era_figure_males, echo = F}
+```{r prep_era_figure, echo = F}
 year1 <- 1969
@@
 
+For females, the life expectancy gap narrowed in `r dim(bw_dat.females[bw_dat.females$era1.diff > 0, ])[1]` of `r dim(bw_dat.females)[1]` states between 1969 and 1983.
+In `r length(cod.df$narrowed_gap[cod.df$sex == "Female" & cod.df$era == "1969-1983" & cod.df$COD == "Cardiovascular" & cod.df$narrowed_gap == T])` states, CVD made a substantial contribution towards narrowing the gap (for example, Virginia, Kentucky, South Carolina), while in others, its contribution was towards widening the gap (CI may be very important here -- the states with very large contributions also have very few blacks...).
+
 Quick summary

### 2017-02-15 Code/calculate_std_mort_rates.Rmd

@@
                 year = first(year),
-                age.std.mortality.rate_LCL = quantile(age.std.mortality.rate, 0.025), #summarize across posterior samples (post.samp)
+                age.std.mortality.rate_LCL = quantile(age.std.mortality.rate, 0.025), #summarise across posterior samples (post.samp)
                 age.std.mortality.rate_med = quantile(age.std.mortality.rate, 0.5),
-                age.std.mortality.rate_UCL = quantile(age.std.mortality.rate, 0.0975))
+                age.std.mortality.rate_UCL = quantile(age.std.mortality.rate, 0.975))
 
@@
                 rate.difference_med = quantile(rate.difference, 0.5),
-                rate.difference_UCL = quantile(rate.difference, 0.0975),
+                rate.difference_UCL = quantile(rate.difference, 0.975),
                 sex = first(sex),

### 2017-02-15 Code/calculate_std_mort_rates.Rmd

@@
 system.time(
-  for(i in 1:8){
+  for(i in 2:8){
     print(i)
@@
     mortality.rates.summary$rate.per.100k <- mortality.rates.summary$age.std.mortality.rate_med*100000
-    mortality.rates.summary <- mortality.rates.summary %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
+    mortality.rates.summary <- mortality.rates.summary %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)),
+                                                                  age.std.mortality.rate_LCL = age.std.mortality.rate_LCL*100000,
+                                                                  age.std.mortality.rate_med = age.std.mortality.rate_med*100000,
+                                                                  age.std.mortality.rate_UCL = age.std.mortality.rate_UCL*100000)
 
@@
 
-    mortality.rates.difference %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
+    #tidy mortality rates summary
+    mortality.rates.difference$COD <- factor(mortality.rates.difference$COD, COD.levels)
+    mortality.rates.difference <- add_Census_Division(mortality.rates.difference, state)
+    mortality.rates.difference <- add_Census_Region(mortality.rates.difference, state)
+    mortality.rates.difference$stabbrs <- state.abb[match(mortality.rates.difference$state, state.name)]
+    mortality.rates.difference <- mortality.rates.difference %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)),
+                                                                        rate.difference_LCL = rate.difference_LCL*100000,
+                                                                        rate.difference_med = rate.difference_med*100000,
+                                                                        rate.difference_UCL = rate.difference_UCL*100000)
 

### 2017-02-15 Code/plot_age_standardardized_mortality_trends.Rmd
use findings with CIs

@@
 ```{r, echo=F}
-mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates.csv")
-mortality.rates.wide <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_wide.csv")
+# mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates.csv")
+# mortality.rates.wide <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_wide.csv")
+
+mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_1.csv")
+mortality.rates.wide <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_diff_1.csv")
 ```
@@
   for(cod in levels(mortality.rates$COD)) {
-    data.sub <- subset(mortality.rates, sex == sex_i & post.samp == 1 & COD == cod)
+    data.sub <- subset(mortality.rates, sex == sex_i & COD == cod)
     a <- ggplot(data.sub, 
-                aes(y=rate.per.100k, x = year)) + 
+                aes(y = age.std.mortality.rate_med*100000, x = year)) + 
       geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
       geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
+      geom_ribbon(aes(ymin = age.std.mortality.rate_LCL*100000, ymax = age.std.mortality.rate_UCL*100000, col = race)) +
       geom_line(aes(col = Census_Division, lty = race)) +

### 2017-02-15 Code/plot_age_standardardized_mortality_trends.Rmd

@@
 mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_1.csv")
-mortality.rates.wide <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_diff_1.csv")
+mortality.rates.diff <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_diff_1.csv")
 ```
@@
     a <- ggplot(data.sub, 
-                aes(y = age.std.mortality.rate_med*100000, x = year)) + 
+                aes(y = age.std.mortality.rate_med, x = year)) + 
       geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
       geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
-      geom_ribbon(aes(ymin = age.std.mortality.rate_LCL*100000, ymax = age.std.mortality.rate_UCL*100000, col = race)) +
+      geom_ribbon(aes(ymin = age.std.mortality.rate_LCL, ymax = age.std.mortality.rate_UCL, fill = Census_Division, group = race), col = NA, alpha = 0.5) +
       geom_line(aes(col = Census_Division, lty = race)) +
@@
       xlab("Year") + 
-      ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))
+      ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))# + theme_minimal()
 
@@
 
-    b <- ggplot(subset(mortality.rates.wide, sex == sex_i & post.samp == 1 & COD == cod), 
-                aes(y=rate.difference, x = year)) + 
+    b <- ggplot(subset(mortality.rates.diff, sex == sex_i & COD == cod), 
+                aes(y=rate.difference_med, x = year)) + 
       geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
       geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
+      geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL, fill = Census_Division), col = NA, alpha = 0.5) +
       geom_hline(aes(yintercept = 0), col = "#636363") +

### 2017-02-16 Code/plot_age_standardardized_mortality_trends.Rmd

@@
 
-mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_1.csv")
-mortality.rates.diff <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_diff_1.csv")
+mortality.rates <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_1.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_2.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_3.csv"),
+                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_4.csv"))
+
+mortality.rates.diff <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_diff_1.csv"),
+                              read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_diff_2.csv"),
+                              read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_diff_3.csv"),
+                              read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_diff_4.csv"))
+
+ mortality.rates <- mortality.rates %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
+
+ mortality.rates$state.reorder2 <- factor(mortality.rates$state, 
+                          levels = c("e1", "e2", "e3", "e4", "e5", "e6", "e7", "e8", "e9", "e10", "Maine",
+                                     "e11", "e12", "e13", "e14", "e15", "e16", "e17", "e18", "e19", "Vermont", "New Hampshire",
+                                     "Washington", "Idaho", "Montana", "North Dakota", "Minnesota", "Illinois", "Wisconsin", "Michigan", "New York", "Massachusettes", "Rhode Island",
+                                     "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", "e20",
+                                     "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Washington DC", "e21",
+                                     "e22", "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Delaware", "e23", "e24",
+                                     "e25", "e26", "e27", "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", "e28", "e29",
+                                     "e30", "e31", "e32", "e33", "Texas", "e34", "e35", "e36", "e37", "Florida"))
+
+ data.sub <- subset(mortality.rates, sex == sex_i & COD == cod)
+ test <- ggplot(data.sub, 
+                aes(y = age.std.mortality.rate_med, x = year)) + 
+      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
+      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
+      geom_ribbon(aes(ymin = age.std.mortality.rate_LCL, ymax = age.std.mortality.rate_UCL, group = race, fill = Census_Division), col = NA, alpha = 0.5) +
+      geom_line(aes(col = Census_Division, lty = race)) +
+      facet_wrap( ~ state.reorder2, ncol = 11, drop = F) +
+      ylab("Age-standardized mortality rate (per 100,000)") +
+      xlab("Year") + 
+      ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))# + theme_minimal()
+ test
+
+ library(gridExtra)
+g <- ggplotGrob(test)
+## remove empty panels
+g$grobs[names(g$grobs) %in% c("panel-1-1", "panel-1-2")] <- NULL
+## remove them from the layout
+g$layout <- g$layout[!(g$layout$name %in% c("panel-1-1", "panel-1-2")),]
+## move axis closer to panel
+#g$layout[g$layout$name == "axis_b-9", c("t", "b")] = c(9,9)
+
+grid.newpage()
+grid.draw(g)
 ```

### 2017-02-16 Code/plot_age_standardardized_mortality_trends.Rmd

@@
 
- data.sub <- subset(mortality.rates, sex == sex_i & COD == cod)
+ data.sub <- subset(mortality.rates, sex == "Male" & COD == "Injuries")
  test <- ggplot(data.sub, 
@@
       ylab("Age-standardized mortality rate (per 100,000)") +
-      xlab("Year") + 
-      ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))# + theme_minimal()
+      xlab("Year") #+ 
+     # ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))# + theme_minimal()
  test
 
- library(gridExtra)
-g <- ggplotGrob(test)
-## remove empty panels
-g$grobs[names(g$grobs) %in% c("panel-1-1", "panel-1-2")] <- NULL
-## remove them from the layout
-g$layout <- g$layout[!(g$layout$name %in% c("panel-1-1", "panel-1-2")),]
-## move axis closer to panel
-#g$layout[g$layout$name == "axis_b-9", c("t", "b")] = c(9,9)
-
-grid.newpage()
-grid.draw(g)
+#  library(gridExtra)
+# g <- ggplotGrob(test)
+# ## remove empty panels
+# g$grobs[names(g$grobs) %in% c("panel-1-1", "panel-1-2")] <- NULL
+# ## remove them from the layout
+# g$layout <- g$layout[!(g$layout$name %in% c("panel-1-1", "panel-1-2")),]
+# ## move axis closer to panel
+# #g$layout[g$layout$name == "axis_b-9", c("t", "b")] = c(9,9)
+# 
+# grid.newpage()
+# grid.draw(g)
 ```

### 2017-02-21 Code/life_expectancy_functions.R
added mort rates and change into run models

@@ -182,7 +182,7 @@ contribution.to.gap.change <- function(type.of.decomp, decomp.table1, decomp.tab
       return(print("Must supply type of data as either practice or results."))
     }else if(type.of.data == "practice"){
       contribution_data <- data.frame("COD" = decomp.table1[["Cause.of.death"]], 
-                                      "Contribution.to.change" = decomp.table1[["C_x_COD"]] - decomp.table2[["C_x_COD"]])      
+                                      "Contribution.to.change" = decomp.table1[["C_x"]] - decomp.table2[["C_x"]])      
     }else if(type.of.data == "results"){
       contribution_data <- data.frame("COD" = decomp.table1[["COD"]], 
                                       "Contribution.to.change" = decomp.table1[["COD_cont_yrs"]] - decomp.table2[["COD_cont_yrs"]])

### 2017-02-21 Code/run-models.R

@@ -1,0 +1,516 @@
+
+#load_libraries
+library(Rcpp)
+library(dplyr)
+library(purrr)
+library(ggplot2)
+library(stringr)
+library(rjags)
+library(R2jags) 
+library(devtools)
+library(jagstools)
+library(reshape2)
+library(parallel)
+library(gtools)
+library(bayesboot)
+library(coda)
+library(grid)
+library(gridExtra)
+library(parallel)
+
+#load_user_written_functions
+source("~/BlackWhiteMortalityGap/Code/2_smoothing_models/bayes_smoothing_functions.R")
+source('~/BlackWhiteMortalityGap/Code/life_expectancy_functions.R') 
+
+#load_tidied_data
+dat.clean = read.csv('~/dat_clean.csv') 
+
+#load US standard population
+std.pop <- read.csv("~/BlackWhiteMortalityGap/Data/US_Standard_Population.csv")
+
+std.pop <- std.pop %>% select(Age, US_Standard_2000) %>% filter(Age != "Total")
+
+std.pop <- std.pop %>% mutate(age.weight = US_Standard_2000/1000000)
+
+std.pop <- std.pop %>% mutate(age_minbin = 0*(Age == "00 years") + 1*(Age == "01-04 years") + 5*(Age == "05-09 years") +
+                                10*(Age == "10-14 years") + 15*(Age == "15-19 years") + 20*(Age == "20-24 years") +
+                                25*(Age == "25-29 years") + 30*(Age == "30-34 years") + 35*(Age == "35-39 years") +
+                                40*(Age == "40-44 years") + 45*(Age == "45-49 years") + 50*(Age == "50-54 years") +
+                                55*(Age == "55-59 years") + 60*(Age == "60-64 years") + 65*(Age == "65-69 years") +
+                                70*(Age == "70-74 years") + 75*(Age == "75-79 years") + 80*(Age == "80-84 years") +
+                                85*(Age == "85+ years")) %>% select(US_Standard_2000, age.weight, age_minbin)
+
+#state_i <- "Maryland"
+
+entire.analysis <- function(state_i) {
+  
+  #fit_models
+  #run time: 32 mins
+  #fit the bayesian models and put the 1000 posterior samples into a data frame
+  for(sex_i in c("Male", "Female")){
+    print(sex_i)
[…]

### 2017-02-21 Code/plot_age_standardardized_mortality_trends.Rmd

--- 2017-02-15 Code/plot_age_standardardized_mortality_trends.Rmd
@@
 ---
-title: "Untitled"
+title: "Appendix: Trends in age-adjusted mortality rates for blacks and whites by cause and sex, 1969-2013, United States"
 author: "Corinne Riddell"
@@
 
+```{r}
+included.states <- c("Alabama", "Arizona", "Arkansas", "California", "Colorado", "Connecticut",
+"Delaware", "Florida", "Georgia", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
+"Louisiana", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi",   
+"Missouri", "Nebraska", "Nevada", "New Jersey", "New Mexico", "New York",
+"North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island",
+"South Carolina",  "Tennessee", "Texas", "Virginia", "Washington", "Washington DC",
+"West Virginia", "Wisconsin")
+
+completed.states <- included.states[17:24]
+```
+
 ```{r, echo=F}
-# mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates.csv")
-# mortality.rates.wide <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_wide.csv")
 
-mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_1.csv")
-mortality.rates.diff <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_diff_1.csv")
+mortality.rates <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_", completed.states[1], ".csv"))
+mortality.rates.diff <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_", completed.states[1], ".csv"))
+
+for(state_i in 1:length(completed.states)){
+  mortality.rates <- rbind(mortality.rates, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_", completed.states[state_i], ".csv")))
+
+    mortality.rates.diff <- rbind(mortality.rates.diff, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_", completed.states[state_i], ".csv")))
+}
+
+ mortality.rates <- mortality.rates %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
+ mortality.rates.diff <- mortality.rates.diff %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
+# 
+#  mortality.rates$state.reorder2 <- factor(mortality.rates$state, 
+#                           levels = c("e1", "e2", "e3", "e4", "e5", "e6", "e7", "e8", "e9", "e10", "Maine",
+#                                      "e11", "e12", "e13", "e14", "e15", "e16", "e17", "e18", "e19", "Vermont", "New Hampshire",
+#                                      "Washington", "Idaho", "Montana", "North Dakota", "Minnesota", "Illinois", "Wisconsin", "Michigan", "New York", "Massachusettes", "Rhode Island",
+#                                      "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", "e20",
+#                                      "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Washington DC", "e21",
+#                                      "e22", "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Delaware", "e23", "e24",
+#                                      "e25", "e26", "e27", "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", "e28", "e29",
+#                                      "e30", "e31", "e32", "e33", "Texas", "e34", "e35", "e36", "e37", "Florida"))   
+#     
+#  data.sub <- subset(mortality.rates, sex == "Male" & COD == "Injuries")
+#  test <- ggplot(data.sub, 
+#                 aes(y = age.std.mortality.rate_med, x = year)) + 
+#       geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
+#       geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
+#       geom_ribbon(aes(ymin = age.std.mortality.rate_LCL, ymax = age.std.mortality.rate_UCL, group = race, fill = Census_Division), col = NA, alpha = 0.5) +
+#       geom_line(aes(col = Census_Division, lty = race)) +
+#       facet_wrap( ~ state.reorder2, ncol = 11, drop = F) +
+#       ylab("Age-standardized mortality rate (per 100,000)") +
+#       xlab("Year") #+ 
+#      # ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))# + theme_minimal()
+#  test
+
+#  library(gridExtra)
+# g <- ggplotGrob(test)
+# ## remove empty panels
+# g$grobs[names(g$grobs) %in% c("panel-1-1", "panel-1-2")] <- NULL
+# ## remove them from the layout
+# g$layout <- g$layout[!(g$layout$name %in% c("panel-1-1", "panel-1-2")),]
+# ## move axis closer to panel
+# #g$layout[g$layout$name == "axis_b-9", c("t", "b")] = c(9,9)
+# 
+# grid.newpage()
+# grid.draw(g)
 ```
@@
 ```{r, echo=F, fig.width=11, fig.height=11}
+
+sex_i <- "Male"
+cod <- "Injuries"
+
 for(sex_i in c("Male", "Female")) {
@@
     a <- ggplot(data.sub, 
-                aes(y = age.std.mortality.rate_med, x = year)) + 
+                aes(y = rate.per.100k_med, x = year)) + 
       geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
       geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
-      geom_ribbon(aes(ymin = age.std.mortality.rate_LCL, ymax = age.std.mortality.rate_UCL, fill = Census_Division, group = race), col = NA, alpha = 0.5) +
+      geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, fill = Census_Division, group = race), col = NA, alpha = 0.5) +
       geom_line(aes(col = Census_Division, lty = race)) +
+      # geom_line(aes(y = rate.per.100k_mean, lty = race)) +
       facet_wrap( ~ state.reorder, ncol = 5) +
@@
       geom_hline(aes(yintercept = 0), col = "#636363") +
+      #geom_line(aes(y = rate.difference_mean)) +
       geom_line(aes(col = Census_Division)) +

### 2017-02-21 Code/plot_age_standardardized_mortality_trends.Rmd
cleaning up

@@
 library(plotly)
-```
+library(forcats)
 
-```{r}
-included.states <- c("Alabama", "Arizona", "Arkansas", "California", "Colorado", "Connecticut",
-"Delaware", "Florida", "Georgia", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
-"Louisiana", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi",   
-"Missouri", "Nebraska", "Nevada", "New Jersey", "New Mexico", "New York",
-"North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island",
-"South Carolina",  "Tennessee", "Texas", "Virginia", "Washington", "Washington DC",
-"West Virginia", "Wisconsin")
+mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_combined.csv")
+mortality.rates.diff <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_combined.csv")
 
-completed.states <- included.states[17:24]
+mortality.rates <- mortality.rates %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
+mortality.rates.diff <- mortality.rates.diff %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
 ```
@@
 ```{r, echo=F}
-
-mortality.rates <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_", completed.states[1], ".csv"))
-mortality.rates.diff <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_", completed.states[1], ".csv"))
-
-for(state_i in 1:length(completed.states)){
-  mortality.rates <- rbind(mortality.rates, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_", completed.states[state_i], ".csv")))
-
-    mortality.rates.diff <- rbind(mortality.rates.diff, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_", completed.states[state_i], ".csv")))
-}
-
- mortality.rates <- mortality.rates %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
- mortality.rates.diff <- mortality.rates.diff %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
 # 

### 2017-02-21 Code/combine_results.Rmd

@@
 
+```{r}
+included.states <- c("Alabama", "Arizona", "Arkansas", "California", "Colorado", "Connecticut",
+"Delaware", "Florida", "Georgia", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
+"Louisiana", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi",   
+"Missouri", "Nebraska", "Nevada", "New Jersey", "New Mexico", "New York",
+"North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island",
+"South Carolina",  "Tennessee", "Texas", "Virginia", "Washington", "Washington DC",
+"West Virginia", "Wisconsin")
+
+completed.states <- included.states[17:24]
+```
+
 ```{r combine_results}
-age_cod_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv")
-age_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_1.csv")
-BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_1.csv")
-cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_1.csv")
+age_cod_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_", completed.states[1], ".csv"))
+age_decomp_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_", completed.states[1], ".csv"))
+BlackWhite_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_", completed.states[1], ".csv"))
+cod_decomp_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_", completed.states[1], ".csv"))
+cod_change_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_change_", completed.states[1], ".csv"))
 
-for(i in 2:8){
-  age_cod_results <- rbind(age_cod_results, 
-                           read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_",
-                                           i, ".csv")))
-  age_decomp_results <- rbind(age_decomp_results, 
-                              read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_",
-                                              i, ".csv")))
-  BlackWhite_results <- rbind(BlackWhite_results, 
-                              read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_",
-                                              i, ".csv")))
-  cod_decomp_results <- rbind(cod_decomp_results, 
-                              read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_",
-                                              i, ".csv")))
+mortality.rates <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_", completed.states[1], ".csv"))
+mortality.rates.diff <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_", completed.states[1], ".csv"))
+
+for(state_i in 1:length(completed.states)){
+
+    age_cod_results <- rbind(age_cod_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_", completed.states[state_i], ".csv")))
+
+    age_decomp_results <- rbind(age_decomp_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_", completed.states[state_i], ".csv")))
+
+    BlackWhite_results <- rbind(BlackWhite_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_", completed.states[state_i], ".csv")))
+
+    cod_decomp_results <- rbind(cod_decomp_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_", completed.states[state_i], ".csv")))
+
+    cod_change_results <- rbind(cod_change_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_change_", completed.states[state_i], ".csv")))
+
+  mortality.rates <- rbind(mortality.rates, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_", completed.states[state_i], ".csv")))
+
+    mortality.rates.diff <- rbind(mortality.rates.diff, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_", completed.states[state_i], ".csv")))
 }
+
+mortality.rates <- mortality.rates %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
+mortality.rates.diff <- mortality.rates.diff %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
 ```
@@
 write.csv(BlackWhite_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
+
+write.csv(mortality.rates, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_combined.csv")
+write.csv(mortality.rates.diff, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_combined.csv")
 ```

### 2017-02-22 Code/combine_results.Rmd

@@
 title: "Combine_results"
-author: "Corinne Riddell"
 date: "2/9/2017"
@@
 
-completed.states <- included.states[17:24]
+completed.states <- included.states[c(1:4, 17:24)]
 ```
@@
 
-for(state_i in 1:length(completed.states)){
+for(state_i in 2:length(completed.states)){
 
@@
 
-COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
-                "Non-communicable", "Injuries", "All other causes")
+# COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
+#                 "Non-communicable", "Injuries", "All other causes")
 
-cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
+#cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
 age_cod_results$COD <- factor(age_cod_results$COD, levels = 1:6, 
@@
 write.csv(BlackWhite_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
+write.csv(cod_change_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_change_results.csv")
 

### 2017-02-22 Manuscript/Manuscript_COD_decomp.Rmd

@@
 cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
+cod_change_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_change_results.csv")
 BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
 
-COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
-                "Non-communicable", "Injuries", "All other causes")
-
-cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
 age_cod_results$COD <- factor(age_cod_results$COD, levels(age_cod_results$COD)[c(3, 2, 4, 5, 6, 1)])
-
 age_decomp_results$age <- factor(age_decomp_results$age, levels = levels(age_decomp_results$age)[c(1, 2, 11, 3:10, 12:19)])
 age_cod_results$age <- factor(age_cod_results$age, levels = levels(age_cod_results$age)[c(1, 2, 11, 3:10, 12:19)])
+
+cod_change_results$Cause.of.death <- factor(cod_change_results$Cause.of.death, levels(cod_change_results$Cause.of.death)[c(3, 2, 4, 5, 6, 1)])
 ```
@@
 bw_dat <- BlackWhite_results %>% 
-  filter(year %in% c(year1, year2, year3, year4), is.na(LE_wbgap) == F) %>%
-  select(state, year, LE_wbgap, sex) %>%
+  filter(year %in% c(year1, year2, year3, year4), is.na(LE_wbgap_med) == F) %>%
+  select(state, year, LE_wbgap_med, sex) %>%
   arrange(year)
 
-bw_dat <- tidyr::spread(bw_dat, year, LE_wbgap) 
+bw_dat <- tidyr::spread(bw_dat, year, LE_wbgap_med) 
 names(bw_dat)[3:6] <- c("first.gap","second.gap", "third.gap", "fourth.gap")  
@@
 
-cod.df <- data.frame(state = factor(), sex = factor(), era = character(), 
-                     COD = factor(), Contribution.to.change = numeric(), 
-                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())
-
-age.df <- data.frame(state = factor(), sex = factor(), era = character(), 
-                     age_minbin = numeric(), Contribution.to.change = numeric(), 
-                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())
-
-years <- c(1969, 1983, 1993, 2013)
-
-for (state.i in levels(BlackWhite_results$state)) {
-  for (sex.i in c("Male", "Female")) {
-    for (year.i in 1:3) {
-
-      table1 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(COD)
-      table2 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(COD)
-
-      table3 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(age_minbin)
-      table4 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(age_minbin)
-
-      temp1 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
-                          contribution.to.gap.change(type.of.decomp = "COD", 
-                                                     decomp.table1 = table1, 
-                                                     decomp.table2 = table2, 
-                                                     type.of.data = "results"))
-
-      temp2 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
-                          contribution.to.gap.change(type.of.decomp = "Age", 
-                                                     decomp.table1 = table3, 
-                                                     decomp.table2 = table4,
-                                                     type.of.data = "results"))
-
-      cod.df <- rbind(cod.df, temp1)
-      age.df <- rbind(age.df, temp2)
-    }
-  }
-}
+# cod.df <- data.frame(state = factor(), sex = factor(), era = character(), 
+#                      COD = factor(), Contribution.to.change = numeric(), 
+#                      Contrib.to.change.prop = numeric(), narrowed_gap = logical())
+# 
+# age.df <- data.frame(state = factor(), sex = factor(), era = character(), 
+#                      age_minbin = numeric(), Contribution.to.change = numeric(), 
+#                      Contrib.to.change.prop = numeric(), narrowed_gap = logical())
+# 
+# years <- c(1969, 1983, 1993, 2013)
+# 
+# for (state.i in levels(BlackWhite_results$state)) {
+#   for (sex.i in c("Male", "Female")) {
+#     for (year.i in 1:3) {
+#       
+#       table1 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(COD)
+#       table2 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(COD)
+#       
+#       table3 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(age_minbin)
+#       table4 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(age_minbin)
+#       
+#       temp1 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
+#                           contribution.to.gap.change(type.of.decomp = "COD", 
+#                                                      decomp.table1 = table1, 
+#                                                      decomp.table2 = table2, 
+#                                                      type.of.data = "results"))
+#       
+#       temp2 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
+#                           contribution.to.gap.change(type.of.decomp = "Age", 
+#                                                      decomp.table1 = table3, 
+#                                                      decomp.table2 = table4,
+#                                                      type.of.data = "results"))
+#       
+#       cod.df <- rbind(cod.df, temp1)
+#       age.df <- rbind(age.df, temp2)
+#     }
+#   }
+# }
 
@@
 ```{r more_figure_prep, echo = F}
+cod.df <- cod_change_results %>% select(state, sex, Cause.of.death, Cx_era1_diff_mean, Cx_era2_diff_mean, Cx_era3_diff_mean)
+cod.df <- tidyr::gather(cod.df, Cx_era1_diff_mean, Cx_era2_diff_mean, Cx_era3_diff_mean, key = "era", value = "Cx_diff_mean")
+ncols <- str_split_fixed(as.character(cod.df$era), "_", 4)
+cod.df$era <- as.factor(ncols[ , 2])
+
+cod.df2 <- cod_change_results %>% select(state, sex, Cause.of.death, Cx_p_era1_diff_mean, Cx_p_era2_diff_mean, Cx_p_era3_diff_mean)
+cod.df2 <- tidyr::gather(cod.df2, Cx_p_era1_diff_mean, Cx_p_era2_diff_mean, Cx_p_era3_diff_mean, key = "era", value = "Cx_p_diff_mean")
+ncols <- str_split_fixed(as.character(cod.df2$era), "_", 5)
+cod.df2$era <- as.factor(ncols[ , 3])
+
+cod.df <- merge(cod.df, cod.df2, by = c("state", "sex", "Cause.of.death", "era"))
+
+cod.df$narrowed_gap <- T
+cod.df$narrowed_gap[cod.df$Cx_diff_mean < 0 ] <- F
 
 cod.df <- make_pretty_decomp_plot(decomp.table = cod.df, strat.var.1 = "state",
-                        strat.var.2 = "sex", strat.var.3 = "era", 
-                        sign.var = "narrowed_gap",
-                        decomp.var = "Contribution.to.change",
-                        decomp.var.prop = "Contrib.to.change.prop", 
-                        partition.bar.var = "COD", type.of.data = "results")
-
-age.df <- make_pretty_decomp_plot(decomp.table = age.df, strat.var.1 = "state", 
-                        strat.var.2 = "sex", strat.var.3 = "era", 
-                        sign.var = "narrowed_gap",
-                        decomp.var = "Contribution.to.change",
-                        decomp.var.prop = "Contrib.to.change.prop", 
-                        partition.bar.var = "Ages", type.of.data = "results")
+                                  strat.var.2 = "sex", strat.var.3 = "era", sign.var = "narrowed_gap",
+                                  decomp.var = "Cx_diff_mean",
+                                  decomp.var.prop = "Cx_p_diff_mean",
+                                  partition.bar.var = "Cause.of.death",
+                                  type.of.data = "results")
 
@@
 sum1 <- BlackWhite_results %>% filter(year == 1969 & sex == "Male") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
+  arrange(LE_wbgap_mean) %>% 
+  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)
 
@@
 sum2 <- BlackWhite_results %>% filter(year == 1983 & sex == "Male") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
+  arrange(LE_wbgap_mean) %>% 
+  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)
 
@@
 sum3 <- BlackWhite_results %>% filter(year == 1993 & sex == "Male") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
+  arrange(LE_wbgap_mean) %>% 
+  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)
 
 sum4 <- BlackWhite_results %>% filter(year == 2013 & sex == "Male") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
+  arrange(LE_wbgap_mean) %>% 
+  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)
 ```
@@
 males.era1 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
+  ggplot(subset(cod.df.males, era == "era1"), aes(y = state.gdiff.num.1, x = new.start.1)) +
     geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                   ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.1, fill = COD), 
+                  xmax = new.finish.1, fill = Cause.of.death), 
               color = "white") +
-    geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
-        geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
+    geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
+        geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
     theme_minimal() + 

### 2017-02-22 Manuscript/Manuscript_COD_decomp.Rmd
incorporated new findings

@@
 
-age_cod_results$COD <- factor(age_cod_results$COD, levels(age_cod_results$COD)[c(3, 2, 4, 5, 6, 1)])
+age_cod_results$COD <- factor(age_cod_results$COD, levels(age_cod_results$COD)[c(3, 2, 4, 6, 5, 1)])
 age_decomp_results$age <- factor(age_decomp_results$age, levels = levels(age_decomp_results$age)[c(1, 2, 11, 3:10, 12:19)])
@@
 
-cod_change_results$Cause.of.death <- factor(cod_change_results$Cause.of.death, levels(cod_change_results$Cause.of.death)[c(3, 2, 4, 5, 6, 1)])
+cod_change_results$Cause.of.death <- factor(cod_change_results$Cause.of.death, levels(cod_change_results$Cause.of.death)[c(3, 2, 4, 6, 5, 1)])
 ```
@@
 bw_dat <- BlackWhite_results %>% 
-  filter(year %in% c(year1, year2, year3, year4), is.na(LE_wbgap_med) == F) %>%
-  select(state, year, LE_wbgap_med, sex) %>%
+  filter(year %in% c(year1, year2, year3, year4), is.na(LE_wbgap_mean) == F) %>%
+  select(state, year, LE_wbgap_mean, sex) %>%
   arrange(year)
 
-bw_dat <- tidyr::spread(bw_dat, year, LE_wbgap_med) 
+bw_dat <- tidyr::spread(bw_dat, year, LE_wbgap_mean) 
 names(bw_dat)[3:6] <- c("first.gap","second.gap", "third.gap", "fourth.gap")  

### 2017-02-22 Code/run-models.R
move tidy std pop

@@ -1,4 +1,7 @@
 
+
+entire.analysis <- function(state_i) {
+  
 #load_libraries
 library(Rcpp)
 library(dplyr)
@@ -26,24 +29,8 @@ source('~/BlackWhiteMortalityGap/Code/life_expectancy_functions.R')
 dat.clean = read.csv('~/dat_clean.csv') 
 
 #load US standard population
-std.pop <- read.csv("~/BlackWhiteMortalityGap/Data/US_Standard_Population.csv")
+std.pop <- read.csv("~/BlackWhiteMortalityGap/Data/US_Standard_Population_for_model.csv") %>% select(US_Standard_2000, age.weight, age_minbin)
 
-std.pop <- std.pop %>% select(Age, US_Standard_2000) %>% filter(Age != "Total")
-
-std.pop <- std.pop %>% mutate(age.weight = US_Standard_2000/1000000)
-
-std.pop <- std.pop %>% mutate(age_minbin = 0*(Age == "00 years") + 1*(Age == "01-04 years") + 5*(Age == "05-09 years") +
-                                10*(Age == "10-14 years") + 15*(Age == "15-19 years") + 20*(Age == "20-24 years") +
-                                25*(Age == "25-29 years") + 30*(Age == "30-34 years") + 35*(Age == "35-39 years") +
-                                40*(Age == "40-44 years") + 45*(Age == "45-49 years") + 50*(Age == "50-54 years") +
-                                55*(Age == "55-59 years") + 60*(Age == "60-64 years") + 65*(Age == "65-69 years") +
-                                70*(Age == "70-74 years") + 75*(Age == "75-79 years") + 80*(Age == "80-84 years") +
-                                85*(Age == "85+ years")) %>% select(US_Standard_2000, age.weight, age_minbin)
-
-#state_i <- "Maryland"
-
-entire.analysis <- function(state_i) {
-  
   #fit_models
   #run time: 32 mins
   #fit the bayesian models and put the 1000 posterior samples into a data frame
@@ -149,7 +136,7 @@ entire.analysis <- function(state_i) {
     rm(data_sub)
   }
   
-  save(r.All, file = paste0("~/BW_results/r_All_", state_i, ".csv")) 
+  save(r.All, file = paste0("~/BW_results/r_All_", state_i, ".Rdata")) 
   levels.state.year.sex <- levels(droplevels(r.All$state.year.sex))
   rm(r.All)
   
[…]

### 2017-02-22 Manuscript/Manuscript_COD_decomp.Rmd
updated code to reflect changes

@@
 males.era2 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
+  ggplot(subset(cod.df.males, era == "era2"), aes(y = state.gdiff.num.2, x = new.start.2)) +
     geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
                   ymax = state.gdiff.num.2 + 0.45, 
-                  xmax = new.finish.2, fill = COD), 
+                  xmax = new.finish.2, fill = Cause.of.death), 
               color = "white") +
@@
 males.era3 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
+  ggplot(subset(cod.df.males, era == "era3"), aes(y = state.gdiff.num.3, x = new.start.3)) +
     geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
                   ymax = state.gdiff.num.3 + 0.45, 
-                  xmax = new.finish.3, fill = COD), 
+                  xmax = new.finish.3, fill = Cause.of.death), 
               color = "white") +
@@
 males.era2.2 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.1, x = new.start.2)) +
+  ggplot(subset(cod.df.males, era == "era2"), aes(y = state.gdiff.num.1, x = new.start.2)) +
     geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.1 - 0.45, 
                   ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.2, fill = COD), 
+                  xmax = new.finish.2, fill = Cause.of.death), 
               color = "white") +
@@
 females.era1 <- ggplotly(
-  ggplot(subset(cod.df.females, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
+  ggplot(subset(cod.df.females, era == "era1"), aes(y = state.gdiff.num.1, x = new.start.1)) +
     geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                   ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.1, fill = COD), 
+                  xmax = new.finish.1, fill = Cause.of.death), 
               color = "white") +
@@
 females.era2 <- ggplotly(
-  ggplot(subset(cod.df.females, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
+  ggplot(subset(cod.df.females, era == "era2"), aes(y = state.gdiff.num.2, x = new.start.2)) +
     geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
                   ymax = state.gdiff.num.2 + 0.45, 
-                  xmax = new.finish.2, fill = COD), 
+                  xmax = new.finish.2, fill = Cause.of.death), 
               color = "white") +
@@
 females.era3 <- ggplotly(
-  ggplot(subset(cod.df.females, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
+  ggplot(subset(cod.df.females, era == "era3"), aes(y = state.gdiff.num.3, x = new.start.3)) +
     geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
                   ymax = state.gdiff.num.3 + 0.45, 
-                  xmax = new.finish.3, fill = COD), 
+                  xmax = new.finish.3, fill = Cause.of.death), 
               color = "white") +
@@
 
-**S15 Fig: Comparison of the contribution of age and cause of death to the difference in life expectancy between blacks and whites in New York, 1969 vs. 1983**
+**S15 Fig: Comparison of the contribution of age and cause of death to the difference in life expectancy between blacks and whites males in New York, 1969 vs. 1983**
 
@@
 ny.1969 <- make_pretty_decomp_plot(decomp.table = subset(ny.dat, year == 1969), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
-                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
+                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
 ny.1983 <- make_pretty_decomp_plot(decomp.table = subset(ny.dat, year == 1983), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
-                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
+                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
 ny.1993 <- make_pretty_decomp_plot(decomp.table = subset(ny.dat, year == 1993), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
-                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
+                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
 

### 2017-02-23 Manuscript/Manuscript_COD_decomp.Rmd
updating ms

--- 2017-02-22 Manuscript/Manuscript_COD_decomp.Rmd
@@
   )
-
-# cod.df <- data.frame(state = factor(), sex = factor(), era = character(), 
-#                      COD = factor(), Contribution.to.change = numeric(), 
-#                      Contrib.to.change.prop = numeric(), narrowed_gap = logical())
-# 
-# age.df <- data.frame(state = factor(), sex = factor(), era = character(), 
-#                      age_minbin = numeric(), Contribution.to.change = numeric(), 
-#                      Contrib.to.change.prop = numeric(), narrowed_gap = logical())
-# 
-# years <- c(1969, 1983, 1993, 2013)
-# 
-# for (state.i in levels(BlackWhite_results$state)) {
-#   for (sex.i in c("Male", "Female")) {
-#     for (year.i in 1:3) {
-#       
-#       table1 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(COD)
-#       table2 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(COD)
-#       
-#       table3 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(age_minbin)
-#       table4 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(age_minbin)
-#       
-#       temp1 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
-#                           contribution.to.gap.change(type.of.decomp = "COD", 
-#                                                      decomp.table1 = table1, 
-#                                                      decomp.table2 = table2, 
-#                                                      type.of.data = "results"))
-#       
-#       temp2 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
-#                           contribution.to.gap.change(type.of.decomp = "Age", 
-#                                                      decomp.table1 = table3, 
-#                                                      decomp.table2 = table4,
-#                                                      type.of.data = "results"))
-#       
-#       cod.df <- rbind(cod.df, temp1)
-#       age.df <- rbind(age.df, temp2)
-#     }
-#   }
-# }
-
 ```
@@
 ```{r more_figure_prep, echo = F}
+widths <- cod_change_results %>% mutate(tot_Cx_era1_CI_width = tot_Cx_era1_diff_ucl - tot_Cx_era1_diff_lcl,
+                                        tot_Cx_era2_CI_width = tot_Cx_era2_diff_ucl - tot_Cx_era2_diff_lcl,
+                                        tot_Cx_era3_CI_width = tot_Cx_era3_diff_ucl - tot_Cx_era3_diff_lcl) %>%
+  filter(Cause.of.death == "Injuries") %>%
+  select(state, sex, tot_Cx_era1_CI_width, tot_Cx_era2_CI_width, tot_Cx_era3_CI_width)
+
+widths <- tidyr::gather(widths, tot_Cx_era1_CI_width, tot_Cx_era2_CI_width, tot_Cx_era3_CI_width, key = "era", value = "tot_Cx_CI_width")
+ncols <- str_split_fixed(as.character(widths$era), "_", 5)
+widths$era <- as.factor(ncols[ , 3])
+
 cod.df <- cod_change_results %>% select(state, sex, Cause.of.death, Cx_era1_diff_mean, Cx_era2_diff_mean, Cx_era3_diff_mean)
@@
 cod.df <- merge(cod.df, cod.df2, by = c("state", "sex", "Cause.of.death", "era"))
+
+cod.df <- merge(cod.df, widths, by = c("state", "sex", "era"))
 
@@
 
-In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `.
+In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap_mean, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `.
 Injury was the largest contributor towards narrowing the gap during this time period, as both race groups experienced decreased rates of injury-related mortality, but blacks had larger rates decreases, especially in states with relatively higher rates in 1969 (**S3-S14 Figs**).
@@
 The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Fig 3**).
-In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` years [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum3[dim(sum3)[1], ]$state `.
+In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum3[dim(sum3)[1], ]$state `.
 By 2013, the gap ranged from `r round(sum4[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum4[dim(sum4)[1], ]$state `.
@@
                   xmax = new.finish.1, fill = Cause.of.death), 
-              color = "white") +
+              color = "white") + # + scale_alpha(range = c(0.64, 4.04) 
     geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
@@
 males.era1
+
+plot <-   ggplot(subset(cod.df.males, era == "era1"), aes(y = state.gdiff.num.1, x = new.start.1)) +
+    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
+                  ymax = state.gdiff.num.1 + 0.45, 
+                  xmax = new.finish.1, fill = Cause.of.death), #, alpha = tot_Cx_CI_width
+              color = "white") + # + scale_alpha(range = c(0.64, 4.04) 
+    geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
+        geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
+    theme_minimal() + 
+    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
+    scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)
 ```
@@
 
-![Difference in life expectancy between black in white males, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Men.bmp)
+![Difference in life expectancy between black and white males, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Men.bmp)
 
@@
 
-![Difference in life expectancy between black in white females, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Women.bmp)
+![Difference in life expectancy between black and white females, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Women.bmp)
 
@@
 
-```{r ny_1969_1983_age_cod_decomp, echo = F, fig.width = 9}
+```{r ny_1969_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
 ny.dat <- age_cod_results %>% filter(state == "New York" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
@@
 **Pennsylvania**
-```{r pa_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
+```{r pa_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
 pa.dat <- age_cod_results %>% filter(state == "Pennsylvania" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
@@
 **Connecticut**
-```{r ct_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
+```{r ct_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
 ct.dat <- age_cod_results %>% filter(state == "Connecticut" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
@@
 **Mississippi**
-```{r ms_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
+```{r ms_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
 ms.dat <- age_cod_results %>% filter(state == "Mississippi" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
@@
 **Florida**
-```{r fl_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
+```{r fl_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
 fl.dat <- age_cod_results %>% filter(state == "Florida" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
@@
 
+**S2 Table: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
+
+```{r}
+cod_change_results2 <- data.frame(cod_change_results[c(3, 40, 41)], apply(cod_change_results[4:39], 2, function(x) x*(-1)))
+
+cod_change_results2 <- cod_change_results2 %>% 
+  mutate(total_era1_diff = paste0(round(tot_Cx_era1_diff_mean, 2), " [", round(tot_Cx_era1_diff_lcl, 2), ",", round(tot_Cx_era1_diff_ucl, 2), "]"),
+         total_era2_diff = paste0(round(tot_Cx_era2_diff_mean, 2), " [", round(tot_Cx_era2_diff_lcl, 2), ",", round(tot_Cx_era2_diff_ucl, 2), "]"),
+         total_era3_diff = paste0(round(tot_Cx_era1_diff_mean, 2), " [", round(tot_Cx_era3_diff_lcl, 2), ",", round(tot_Cx_era3_diff_ucl, 2), "]"),
+         contribution_era1_diff = paste0(round(Cx_era1_diff_mean, 2)," [", round(Cx_era1_diff_lcl, 2), ",", round(Cx_era1_diff_ucl, 2), "]"),
+         contribution_era2_diff = paste0(round(Cx_era2_diff_mean, 2)," [", round(Cx_era2_diff_lcl, 2), ",", round(Cx_era2_diff_ucl, 2), "]"),
+         contribution_era3_diff = paste0(round(Cx_era3_diff_mean, 2)," [", round(Cx_era3_diff_lcl, 2), ",", round(Cx_era3_diff_ucl, 2), "]")
+         ) %>%
+  select(sex, state, Cause.of.death, total_era1_diff, contribution_era1_diff, total_era2_diff, contribution_era2_diff, total_era3_diff, contribution_era3_diff) %>%
+  rename(Sex = sex, State = state, 
+         `Change from 1969 to 1983, years` = total_era1_diff,
+         `Change from 1983 to 1993, years` = total_era2_diff,
+         `Change from 1993 to 2013, years` = total_era3_diff,
+         `Contribution to gap change ('69-'83), years` = contribution_era1_diff,
+         `Contribution to gap change ('83-'93), years` = contribution_era2_diff,
+         `Contribution to gap change ('93-'03), years` = contribution_era3_diff)
+
+data.table::data.table(cod_change_results2 %>% filter(Sex == "Male"))
+```
+
+**S3 Table: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
+
+```{r}
+data.table::data.table(cod_change_results2 %>% filter(Sex == "Female"))
+```
+
 **Things for corinne to review**

### 2017-02-23 Code/run-models.R
three more

@@ -1,13 +1,14 @@
 
-included.states <- c("Alabama", "Arizona", "Arkansas", "California", "Colorado", "Connecticut",
-"Delaware", "Florida", "Georgia", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
-"Louisiana", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi",   
-"Missouri", "Nebraska", "Nevada", "New Jersey", "New Mexico", "New York",
-"North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island",
-"South Carolina",  "Tennessee", "Texas", "Virginia", "Washington", "Washington DC",
-"West Virginia", "Wisconsin") 
-
-for(i in c(10, 12:16, 25)){
+for(i in c(12, 13, 15, 16, 25)){
+  
+  included.states <- c("Alabama", "Arizona", "Arkansas", "California", "Colorado", "Connecticut",
+                       "Delaware", "Florida", "Georgia", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
+                       "Louisiana", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi",   
+                       "Missouri", "Nebraska", "Nevada", "New Jersey", "New Mexico", "New York",
+                       "North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island",
+                       "South Carolina",  "Tennessee", "Texas", "Virginia", "Washington", "Washington DC",
+                       "West Virginia", "Wisconsin") 
+  
   source("~/BlackWhiteMortalityGap/Code/entire_analysis.R")
   system.time(entire.analysis(state_i = included.states[i]))
   rm(list = ls(all = TRUE))

### 2017-02-23 Manuscript/Appendix-agestd-mort-rates.Rmd

---
title: "Appendix: Trends in age-adjusted mortality rates for blacks and whites by cause and sex, 1969-2013, United States"
output: html_document
---

Corinne A Riddell, PhD

Department of Epidemiology, Biostatistics and Occupational Health, McGill University, Montreal, Canada.

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>

```{r load_libraries, echo=FALSE, warning=F, message=F}
library(dplyr)
library(purrr)
library(ggplot2)
library(stringr)
library(reshape2)
library(grid)
library(gridExtra)
library(plotly)
library(forcats)

mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_combined.csv")
mortality.rates.diff <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_combined.csv")

mortality.rates <- mortality.rates %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division))) %>% rename(Race = race)
mortality.rates.diff <- mortality.rates.diff %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division))) 
```

```{r refactor-for-us-map-plot, echo=F}

# Create unique blank strip labels for empty facets
bl = sapply(1:37, function(n) paste(rep(" ",n),collapse=""))

mortality.rates$state.reorder2 <- factor(mortality.rates$state,
                                        levels = c(bl[1:10], "Maine",
                                                   bl[11:19], "Vermont", "New Hampshire",
                                                   "Washington", "Idaho", "Montana", "North Dakota", "Minnesota", "Illinois", "Wisconsin", "Michigan", "New York", "Massachusetts", "Rhode Island",
                                                   "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", bl[20],
                                                   "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Washington DC", bl[21],
                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Delaware", bl[23:24],
                                                   bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
                                                   bl[30:33], "Texas", bl[34:37], "Florida"))

mortality.rates.diff$state.reorder2 <- factor(mortality.rates.diff$state,
                                        levels = c(bl[1:10], "Maine",
                                                   bl[11:19], "Vermont", "New Hampshire",
                                                   "Washington", "Idaho", "Montana", "North Dakota", "Minnesota", "Illinois", "Wisconsin", "Michigan", "New York", "Massachusetts", "Rhode Island",
                                                   "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", bl[20],
                                                   "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Washington DC", bl[21],
                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Delaware", bl[23:24],
                                                   bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
                                                   bl[30:33], "Texas", bl[34:37], "Florida"))
```

```{r plot-arrange-as-us-map, echo=F, fig.width=18.5, fig.height=14}

# sex_i <- "Male"
# cod <- "Injuries"

for(sex_i in c("Male", "Female")) {
  for(cod in levels(mortality.rates$COD)) {

# for(sex_i in c("Male")) {
#   for(cod in levels(mortality.rates$COD)[1]) {
    data.sub <- subset(mortality.rates, sex == sex_i & COD == cod)
    map.plot1 <- ggplot(data.sub,
                        aes(y = rate.per.100k_mean, x = year)) +
      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
      geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, group = Race), fill = "grey", col = NA) +
      geom_line(aes(lty = Race)) +
      facet_wrap( ~ state.reorder2, ncol = 11, drop = F, strip.position="bottom") +
      theme_classic(base_size = 20) +
      theme(axis.text.x = element_blank(),
            strip.background=element_blank(),
            axis.line=element_blank(),
            axis.ticks=element_blank()) +
      #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      ylab("Age-standardized mortality rate (per 100,000)") +
      xlab("Year (1969-2013)") +
      ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))# + theme_minimal()

    print(map.plot1)

        map.plot2 <- ggplot(subset(mortality.rates.diff, sex == sex_i & COD == cod),
                        aes(y = rate.difference_mean, x = year)) +
      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
      geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL), fill = "grey", col = NA) +
      geom_line() +
      geom_hline(aes(yintercept = 0), col = "#636363") +
      facet_wrap( ~ state.reorder2, ncol = 11, drop = F, strip.position="bottom") +
      theme_classic(base_size = 20) +
      theme(axis.text.x = element_blank(),
            strip.background=element_blank(),
            axis.line=element_blank(),
            axis.ticks=element_blank()) +
      #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      ylab("Black-white difference (per 100,000)") +
      xlab("Year (1969-2013)") +
      ggtitle(paste0("Black-white difference in age-adjusted mortality (per 100,000) in ", 
                     sex_i, " for ", cod))

    print(map.plot2)

  }
}
```

```{r, echo=F, fig.width=11, fig.height=11, eval=F}

sex_i <- "Male"
cod <- "Injuries"

for(sex_i in c("Male", "Female")) {
  for(cod in levels(mortality.rates$COD)) {

# for(sex_i in c("Male")) {
#   for(cod in levels(mortality.rates$COD)[1]) {
    data.sub <- subset(mortality.rates, sex == sex_i & COD == cod)
    a <- ggplot(data.sub, 
                aes(y = rate.per.100k_mean, x = year)) + 
      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
      geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, fill = Census_Division, group = Race), col = NA, alpha = 0.5) +
      geom_line(aes(col = Census_Division, lty = Race)) +
      # geom_line(aes(y = rate.per.100k_med, lty = Race)) +
      facet_wrap( ~ state.reorder, ncol = 5) +
      ylab("Age-standardized mortality rate (per 100,000)") +
      xlab("Year") + 
      ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))# + theme_minimal()

    print(a)

    b <- ggplot(subset(mortality.rates.diff, sex == sex_i & COD == cod), 
                aes(y=rate.difference_mean, x = year)) + 
      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
      geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL, fill = Census_Division), col = NA, alpha = 0.5) +
      geom_hline(aes(yintercept = 0), col = "#636363") +
      #geom_line(aes(y = rate.difference_med)) +
      geom_line(aes(col = Census_Division)) +
      facet_wrap( ~ state.reorder, ncol = 5) +
      ylab("Black-white difference (per 100,000)") +
      xlab("Year") + 
      ggtitle(paste0("Black-white difference in age-adjusted mortality (per 100,000) in ", 
                     sex_i, " for ", cod))
    print(b)

  }
}
```

### 2017-02-23 Manuscript/Manuscript_COD_decomp.Rmd
new results

@@
 
-*To re-write*: To calculate 95% credible intervals, we sampled 1000 times from the posterior distribution to create 1000 samples for each stratum.
-We performed all calculations on each sample and took the median of the estimates as the estimate and the 2.5 and 97.5th quantiles as the lower and upper bound for the 95% CI.
-Anything about thinning?
+*ktm to read over*: The bayesian model was fit using two markov chains of length 10,000.
+One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration (to reduce autocorrelation).
+These 1,000 iterations approximate the shape of the posterior distribution, and we used the mean as the estimate of the log mortality rate, and the 2.5th and 97.5th percentiles as the lower and upper bound of the 95% credible interval (CI).
 
 All data analyses were performed using R version 3.3.1.
-The Bayesian smoothing was performed using `JAGS` (version X.X), and the life expectancy and decomposition analysis were performed using user-written functions found in the Supplement.
+The Bayesian smoothing was performed using `rjags` (version 4-6) and `R2jags` (version 0.5-7), and the life expectancy and decomposition analysis were performed using user-written functions found in the Supplement.
 

### 2017-02-23 Code/combine_results.Rmd
updating ms

@@
 
-completed.states <- included.states[c(1:4, 17:24)]
+completed.states <- included.states[c(1:11, 14, 17:24, 26:27, 33:34)]
 ```
@@
 
-# COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
-#                 "Non-communicable", "Injuries", "All other causes")
-
-#cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
 age_cod_results$COD <- factor(age_cod_results$COD, levels = 1:6, 

### 2017-02-23 Manuscript/Manuscript_COD_decomp.Rmd

@@
                                            c("state", "Census_Region", "Census_Division", "stabbrs")], by = c("state"))
+cod.df.males <- merge(subset(cod.df, sex == "Male"), bw_dat.males, by = c("state"))
+cod.df.males <- cod.df.males %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
+                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
+                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)
+
+cod.df.males$tot_Cx_CI_width2 <- -(cod.df.males$tot_Cx_CI_width - min(cod.df.males$tot_Cx_CI_width))/(max(cod.df.males$tot_Cx_CI_width) - min(cod.df.males$tot_Cx_CI_width)) + 1
+cod.df.males$width.factor <- as.factor(cut(cod.df.males$tot_Cx_CI_width2, breaks = c(-0.01, 0.25, 0.50, 0.75, 1)))
+
 ```
@@
 }
-
-cod.df.males <- merge(subset(cod.df, sex == "Male"), bw_dat.males, by = c("state"))
-cod.df.males <- cod.df.males %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
-                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
-                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)
-
 ```
@@
 **Fig 1: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
-
 ```{r figure1, fig.align = "center", echo=F, fig.height=9, fig.width=7}
@@
                   ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.1, fill = Cause.of.death), 
-              color = "white") + # + scale_alpha(range = c(0.64, 4.04) 
+                  xmax = new.finish.1, fill = Cause.of.death, alpha = width.factor), 
+              color = "white") +
     geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
@@
 males.era1
-
-plot <-   ggplot(subset(cod.df.males, era == "era1"), aes(y = state.gdiff.num.1, x = new.start.1)) +
-    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
-                  ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.1, fill = Cause.of.death), #, alpha = tot_Cx_CI_width
-              color = "white") + # + scale_alpha(range = c(0.64, 4.04) 
-    geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
-        geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
-    theme_minimal() + 
-    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
-    scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)
 ```
@@
                   ymax = state.gdiff.num.2 + 0.45, 
-                  xmax = new.finish.2, fill = Cause.of.death), 
+                  xmax = new.finish.2, fill = Cause.of.death, alpha = width.factor), 
               color = "white") +
@@
                   ymax = state.gdiff.num.3 + 0.45, 
-                  xmax = new.finish.3, fill = Cause.of.death), 
+                  xmax = new.finish.3, fill = Cause.of.death, alpha = width.factor), 
               color = "white") +
@@
                   ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.2, fill = Cause.of.death), 
+                  xmax = new.finish.2, fill = Cause.of.death, alpha = width.factor), 
               color = "white") +
@@
 
-subplot(males.era1, males.era2.2, shareY = T)
+males.era3.2 <- ggplotly(
+  ggplot(subset(cod.df.males, era == "era3"), aes(y = state.gdiff.num.1, x = new.start.3)) +
+    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.1 - 0.45, 
+                  ymax = state.gdiff.num.1 + 0.45, 
+                  xmax = new.finish.3, fill = Cause.of.death, alpha = width.factor), 
+              color = "white") +
+    geom_segment(data = bw_dat.males, 
+                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
+    geom_segment(data = bw_dat.males, 
+                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
+    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>% 
+  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
+
+males.era3.2
+
+subplot(males.era1, males.era2.2, males.era3.2, shareY = T)
 ```
@@
 cod.df.females <- merge(subset(cod.df, sex == "Female"), bw_dat.females, by = c("state"))
+cod.df.females$tot_Cx_CI_width2 <- -(cod.df.females$tot_Cx_CI_width - min(cod.df.females$tot_Cx_CI_width))/(max(cod.df.females$tot_Cx_CI_width) - min(cod.df.females$tot_Cx_CI_width)) + 1
+cod.df.females$width.factor <- as.factor(cut(cod.df.females$tot_Cx_CI_width2, breaks = c(-0.01, 0.25, 0.50, 0.75, 1)))
+
 cod.df.females <- cod.df.females %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
@@
                   ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.1, fill = Cause.of.death), 
+                  xmax = new.finish.1, fill = Cause.of.death, alpha = width.factor), 
               color = "white") +
@@
 | Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
-| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Sydrome and Nephrosis.
+| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Syndrome and Nephrosis.
 |

### 2017-02-24 Manuscript/Appendix-agestd-mort-rates.Rmd

@@
 # Create unique blank strip labels for empty facets
-bl = sapply(1:37, function(n) paste(rep(" ",n),collapse=""))
+bl = sapply(1:38, function(n) paste(rep(" ",n),collapse=""))
 
@@
                                                    bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
-                                                   bl[30:33], "Texas", bl[34:37], "Florida"))
+                                                   bl[30:33], "Texas", bl[34:38], "Florida"))
 
@@
                                                    bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
-                                                   bl[30:33], "Texas", bl[34:37], "Florida"))
+                                                   bl[30:33], "Texas", bl[34:38], "Florida"))
 ```
 
-```{r plot-arrange-as-us-map, echo=F, fig.width=18.5, fig.height=14}
+```{r plot-arrange-as-us-map, echo=F, fig.width=19.5, fig.height=14, results="asis", message = F}
+roundUp <- function(x) 10^ceiling(log10(x))
 
 # sex_i <- "Male"
-# cod <- "Injuries"
+# cod <- "Cardiovascular"
 
 for(sex_i in c("Male", "Female")) {
-  for(cod in levels(mortality.rates$COD)) {
-
+  for(cod in levels(mortality.rates$COD)[c(3, 2, 4, 6, 5, 1)]) {
 # for(sex_i in c("Male")) {
 #   for(cod in levels(mortality.rates$COD)[1]) {
+    cat("  \n### Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod, "  \n")
+
     data.sub <- subset(mortality.rates, sex == sex_i & COD == cod)
+    lim.y <- max(data.sub$rate.per.100k_mean)
     map.plot1 <- ggplot(data.sub,
@@
       geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
-      geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, group = Race), fill = "grey", col = NA) +
-      geom_line(aes(lty = Race)) +
+      geom_line(aes(lty = Race, col = Race)) +
+      geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, group = Race, fill = Race), col = NA, alpha = 0.5) +
+      scale_color_manual(values = c("#2166ac", "#b2182b")) +
+      scale_fill_manual(values = c("#67a9cf", "#ef8a62")) +
+      scale_y_continuous(limits = c(0, lim.y)) + 
       facet_wrap( ~ state.reorder2, ncol = 11, drop = F, strip.position="bottom") +
@@
       ylab("Age-standardized mortality rate (per 100,000)") +
-      xlab("Year (1969-2013)") +
-      ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))# + theme_minimal()
-
+      xlab("Year (1969-2013)")
     print(map.plot1)
 
-        map.plot2 <- ggplot(subset(mortality.rates.diff, sex == sex_i & COD == cod),
+    cat("  \n")
+
+    cat("  \n### Black-white difference in age-adjusted mortality (per 100,000) in ", 
+        sex_i, " for ", cod, "  \n")
+
+    data.sub2 <- subset(mortality.rates.diff, sex == sex_i & COD == cod)
+    lim.y2 <- max(data.sub2$rate.difference_mean)
+
+    map.plot2 <- ggplot(data.sub2,
                         aes(y = rate.difference_mean, x = year)) +
@@
       geom_hline(aes(yintercept = 0), col = "#636363") +
+      scale_y_continuous(limits = c(0, lim.y)) + 
       facet_wrap( ~ state.reorder2, ncol = 11, drop = F, strip.position="bottom") +
@@
       ylab("Black-white difference (per 100,000)") +
-      xlab("Year (1969-2013)") +
-      ggtitle(paste0("Black-white difference in age-adjusted mortality (per 100,000) in ", 
-                     sex_i, " for ", cod))
+      xlab("Year (1969-2013)")
 
     print(map.plot2)
-
+    cat("  \n")
+    cat("  \n")
   }
+  cat("  \n")
 }

### 2017-02-24 Manuscript/Manuscript_COD_decomp.Rmd

@@
 ---
-title: Exploring how changes in cause of death have narrowed the black-white life expectancy gap in the United States.
-
+title: Exploring how changes in cause of death have narrowed the black-white life
+  expectancy gap in the United States.
 output:
+  html_document: default
   word_document: default
-  html_document: default
 always_allow_html: yes
@@
 library(stringr)
+library(DT)
 source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
@@
 
-  max.value <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-j, ]$state & 
-                                            cod.df$era == "1969-1983" &
+  max.value <- max(cod.df$Cx_diff_mean[cod.df$state == diff[dim(diff)[1]-j, ]$state & 
+                                            cod.df$era == "era1" &
                                             cod.df$sex == "Male" ])
 
-  max.cause <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-j, ]$state &
+  max.cause <- cod.df$Cause.of.death[cod.df$state == diff[dim(diff)[1]-j, ]$state &
                     cod.df$sex == "Male" &
-                    cod.df$era == "1969-1983" & 
-                    cod.df$Contribution.to.change == max.value]
+                    cod.df$era == "era1" & 
+                    cod.df$Cx_diff_mean == max.value]
 
-  min.value <- min(cod.df$Contribution.to.change[cod.df$state == diff[i, ]$state & 
-                                            cod.df$era == "1969-1983" &
+  min.value <- min(cod.df$Cx_diff_mean[cod.df$state == diff[i, ]$state & 
+                                            cod.df$era == "era1" &
                                             cod.df$sex == "Male" ])
 
-  min.cause <- cod.df$COD[cod.df$state == diff[i, ]$state &
+  min.cause <- cod.df$Cause.of.death[cod.df$state == diff[i, ]$state &
                     cod.df$sex == "Male" &
-                    cod.df$era == "1969-1983" & 
-                    cod.df$Contribution.to.change == min.value]
+                    cod.df$era == "era1" & 
+                    cod.df$Cx_diff_mean == min.value]
 
@@
 
-  min.value.DC <- min(cod.df$Contribution.to.change[cod.df$state == "Washington DC" & 
-                                            cod.df$era == "1983-1993" &
+  min.value.DC <- min(cod.df$Cx_diff_mean[cod.df$state == "Washington DC" & 
+                                            cod.df$era == "era2" &
                                             cod.df$sex == "Male" ])
 
-  min.cause.DC <- cod.df$COD[cod.df$state == "Washington DC" &
+  min.cause.DC <- cod.df$Cause.of.death[cod.df$state == "Washington DC" &
                     cod.df$sex == "Male" &
-                    cod.df$era == "1983-1993" & 
-                    cod.df$Contribution.to.change == min.value.DC]
+                    cod.df$era == "era2" & 
+                    cod.df$Cx_diff_mean == min.value.DC]
 ```
@@
 In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum3[dim(sum3)[1], ]$state `.
-By 2013, the gap ranged from `r round(sum4[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum4[dim(sum4)[1], ]$state `.
+By 2013, the gap ranged from `r round(sum4[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum4[dim(sum4)[1], ]$state `.
 During this time period all of the causes contributed to narrowing the gap for most states.
@@
 
-```{r figure2, echo = F, fig.height=9, fig.width=11}
+```{r figure2, echo = F, fig.height=9, fig.width=10}
 # - The gap in 1983 ranges from `r round(sum2[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum2[1, ]$LE_wbgap_lcl, 1)`, `r round(sum2[1, ]$LE_wbgap_ucl, 1)`] in `r sum2[1, ]$state ` to `r round(sum2[dim(sum2)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum2[dim(sum2)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum2[dim(sum2)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum2[dim(sum2)[1], ]$state `.
@@
 
-```{r figure3, echo = F, fig.height=9, fig.width=11}
+```{r figure3, echo = F, fig.height = 9, fig.width = 11}
 y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
@@
 
-```{r, fig.align="center", fig.height=5, fig.width=9, echo=F, eval = F}
+**All the above together as one figure**
+
+```{r, fig.align="center", fig.height=9, fig.width=15, echo=F, eval = T}
 males.era2.2 <- ggplotly(
@@
   layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
-
-males.era2.2
 
@@
 
-males.era3.2
-
 subplot(males.era1, males.era2.2, males.era3.2, shareY = T)
@@
 sum5 <- BlackWhite_results %>% filter(year == 1969 & sex == "Female") %>% 
-  arrange(LE_wbgap) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
+  arrange(LE_wbgap_mean) %>% 
+  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)
 ```
@@
 
-```{r figure4, fig.align = "center", echo=F}
+```{r figure2a, fig.align = "center", echo=F, fig.height = 9, fig.width = 7.5}
 
@@
 
-```{r figure2b, fig.align = "center", echo=F}
+```{r figure2b, fig.align = "center", echo=F, fig.height = 9, fig.width = 7.5}
 y.labels.2 <- levels(factor(cod.df.females$state.gdiff.order.2))
@@
 
-```{r fig6, fig.align = "center", echo=F}
+```{r fig6, fig.align = "center", echo=F, fig.height = 9, fig.width = 9}
 y.labels.3 <- levels(factor(cod.df.females$state.gdiff.order.3))
@@
 
-```{r ny_1969_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
+```{r ny_1969_1983_age_cod_decomp, echo = F, fig.width = 9}
 ny.dat <- age_cod_results %>% filter(state == "New York" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
-                                     mutate(sign.var = sign(age_COD_cont_yrs))
+                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))
 
@@
 **Pennsylvania**
-```{r pa_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
+```{r pa_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
 pa.dat <- age_cod_results %>% filter(state == "Pennsylvania" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
-                                     mutate(sign.var = sign(age_COD_cont_yrs))
+                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))
 
 pa.1993 <- make_pretty_decomp_plot(decomp.table = subset(pa.dat, year == 1993), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
-                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
+                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
 pa.1983 <- make_pretty_decomp_plot(decomp.table = subset(pa.dat, year == 1983), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
-                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
+                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
 
@@
 **Connecticut**
-```{r ct_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
+```{r ct_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
 ct.dat <- age_cod_results %>% filter(state == "Connecticut" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
-                                     mutate(sign.var = sign(age_COD_cont_yrs))
+                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))
 
 ct.1993 <- make_pretty_decomp_plot(decomp.table = subset(ct.dat, year == 1993), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
-                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
+                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
 ct.1983 <- make_pretty_decomp_plot(decomp.table = subset(ct.dat, year == 1983), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
-                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
+                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
 
@@
 **Mississippi**
-```{r ms_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
+```{r ms_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
 ms.dat <- age_cod_results %>% filter(state == "Mississippi" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
-                                     mutate(sign.var = sign(age_COD_cont_yrs))
+                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))
 
 ms.1993 <- make_pretty_decomp_plot(decomp.table = subset(ms.dat, year == 1993), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
-                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
+                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
 ms.1983 <- make_pretty_decomp_plot(decomp.table = subset(ms.dat, year == 1983), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
-                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
+                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
 
@@
 **Florida**
-```{r fl_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
+```{r fl_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
 fl.dat <- age_cod_results %>% filter(state == "Florida" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
-                                     mutate(sign.var = sign(age_COD_cont_yrs))
+                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))
 
 fl.1993 <- make_pretty_decomp_plot(decomp.table = subset(fl.dat, year == 1993), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
-                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
+                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
 fl.1983 <- make_pretty_decomp_plot(decomp.table = subset(fl.dat, year == 1983), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
-                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
+                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
+                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
 
@@
 
-```{r}
+```{r, echo=F}
 cod_change_results2 <- data.frame(cod_change_results[c(3, 40, 41)], apply(cod_change_results[4:39], 2, function(x) x*(-1)))
@@
 
-data.table::data.table(cod_change_results2 %>% filter(Sex == "Male"))
+DT::datatable(cod_change_results2 %>% filter(Sex == "Male"))
 ```
@@
 
-```{r}
-data.table::data.table(cod_change_results2 %>% filter(Sex == "Female"))
+```{r, echo=F}
+DT::datatable(cod_change_results2 %>% filter(Sex == "Female"))
 ```

### 2017-02-28 Manuscript/Manuscript_COD_decomp.Rmd
reformat S2 table

@@
 
+Sam Harper, PhD
+
 Jay S Kaufman, PhD
-
-Sam Harper, PhD
 
@@
 **Methods and Findings**
+
 Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes according to year (1969-2013), state, age, sex, race (black or white), and six cause of death categories (Cancer, cardiovascular disease, communicable disease, non-communicable disease, injuries, and all other causes).
@@
 
-**All the above together as one figure**
-
-```{r, fig.align="center", fig.height=9, fig.width=15, echo=F, eval = T}
+```{r, fig.align="center", fig.height=9, fig.width=15, echo=F, eval = F}
 males.era2.2 <- ggplotly(
@@
 
-**S2 Table: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
+**S2 Table: Decomposition of the change in the black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
 
@@
          ) %>%
-  select(sex, state, Cause.of.death, total_era1_diff, contribution_era1_diff, total_era2_diff, contribution_era2_diff, total_era3_diff, contribution_era3_diff) %>%
-  rename(Sex = sex, State = state, 
-         `Change from 1969 to 1983, years` = total_era1_diff,
-         `Change from 1983 to 1993, years` = total_era2_diff,
-         `Change from 1993 to 2013, years` = total_era3_diff,
-         `Contribution to gap change ('69-'83), years` = contribution_era1_diff,
-         `Contribution to gap change ('83-'93), years` = contribution_era2_diff,
-         `Contribution to gap change ('93-'03), years` = contribution_era3_diff)
+  select(sex, state, Cause.of.death, total_era1_diff, contribution_era1_diff, total_era2_diff, contribution_era2_diff, total_era3_diff, contribution_era3_diff)
 
-DT::datatable(cod_change_results2 %>% filter(Sex == "Male"))
-```
+totals <- cod_change_results2 %>% 
+  filter(Cause.of.death == "All other causes") %>% #to get one row per sex-state stratum
+  select(sex, state, total_era1_diff, total_era2_diff, total_era3_diff) %>% 
+  rename(contribution_era1_diff = total_era1_diff, 
+         contribution_era2_diff = total_era2_diff,
+         contribution_era3_diff = total_era3_diff) %>%
+  mutate(Cause.of.death = "Total change, years")
 
-**S3 Table: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
+cod_change_results2 <- cod_change_results2 %>% 
+  select(sex, state, Cause.of.death, contribution_era1_diff, contribution_era2_diff, contribution_era3_diff)
 
-```{r, echo=F}
-DT::datatable(cod_change_results2 %>% filter(Sex == "Female"))
+dim(cod_change_results2)
+cod_change_results2 <- rbind(cod_change_results2, totals)
+dim(cod_change_results2)
+
+cod_change_results2$Cause.of.death <- factor(cod_change_results2$Cause.of.death, levels = c("Total change, years", "Cardiovascular", "Cancers",
+                                                                             "Communicable", "Non-communicable", "Injuries",
+                                                                             "All other causes"))
+
+cod_change_results_era1 <- cod_change_results2 %>% select(sex, state, Cause.of.death, contribution_era1_diff) %>% tidyr::spread(sex, contribution_era1_diff) %>%
+  rename(Female1 = Female, Male1 = Male)
+
+cod_change_results_era2 <- cod_change_results2 %>% select(sex, state, Cause.of.death, contribution_era2_diff) %>% tidyr::spread(sex, contribution_era2_diff) %>%
+  rename(Female2 = Female, Male2 = Male)
+
+cod_change_results_era3 <- cod_change_results2 %>% select(sex, state, Cause.of.death, contribution_era3_diff) %>% tidyr::spread(sex, contribution_era3_diff) %>%
+  rename(Female3 = Female, Male3 = Male)
+
+cod_change_results3 <- merge(cod_change_results_era1, cod_change_results_era2, by = c("state", "Cause.of.death"))
+cod_change_results3 <- merge(cod_change_results3, cod_change_results_era3, by = c("state", "Cause.of.death"))
+
+cod_change_results3 <- cod_change_results3 %>% arrange(state, Cause.of.death) %>%
+  rename(State = state, `Cause of death` = Cause.of.death)
+
+sketch = htmltools::withTags(table(
+  class = 'display',
+  thead(
+    tr(
+      th(rowspan = 3, 'State'),
+      th(rowspan = 3, 'Cause of death'),
+      th(colspan = 6, 'Contribution to change in gap, years')
+    ),
+    tr(
+      #lapply(c('Era1', 'Era2', 'Era3'), th)
+      th(colspan = 2, '1969-1983'),
+      th(colspan = 2, '1983-1993'),
+      th(colspan = 2, '1993-2013')
+    ),
+    tr(
+      lapply(rep(c('Female', 'Male'), 3), th)
+    )
+  )
+))
+print(sketch)
+
+DT::datatable(cod_change_results3, container = sketch, rownames = F)
 ```

### 2017-02-28 Manuscript/Manuscript_COD_decomp.Rmd
added cross sectional decomp 2013

@@
 age_cod_results$COD <- factor(age_cod_results$COD, levels(age_cod_results$COD)[c(3, 2, 4, 6, 5, 1)])
+cod_decomp_results$COD <- factor(cod_decomp_results$COD, levels(cod_decomp_results$COD)[c(3, 2, 4, 6, 5, 1)])
 age_decomp_results$age <- factor(age_decomp_results$age, levels = levels(age_decomp_results$age)[c(1, 2, 11, 3:10, 12:19)])
@@
 
+**Fig 7: Decomposition of the male black-white life expectancy gap by cause of death in 2013 by US state**
+
+```{r cross-sectional-decomp-males}
+cod.2013.males <- cod_decomp_results %>% filter(year == 2013 & sex == "Male") %>% 
+  mutate(widened_gap = sign(COD_cont_yrs_mean), state.n = as.numeric(state))
+
+y.labels.2013 <- levels(factor(cod.2013.males$state))
+y.num.breaks.2013 <- length(levels(factor(cod.2013.males$state)))
+
+cod.2013.males <- make_pretty_decomp_plot(decomp.table = cod.2013.males, 
+                                          strat.var.1 = "state",
+                                          sign.var = "widened_gap",
+                                          decomp.var = "COD_cont_yrs_mean",
+                                          decomp.var.prop = "COD_cont_prop_mean",
+                                          partition.bar.var = "COD",
+                                          type.of.data = "practice")
+
+males.2013 <- ggplotly(
+  ggplot(cod.2013.males, aes(y = state.n, x = start)) +
+    geom_rect(aes(xmin = start, ymin = state.n - 0.45, 
+                  ymax = state.n + 0.45, 
+                  xmax = finish, fill = COD), 
+              color = "white") +
+     scale_y_continuous(breaks = 1:y.num.breaks.2013, labels = y.labels.2013) +
+    theme_minimal())  %>% layout(xaxis = list(title = "Contribution to gap, years"), yaxis = list(title = NA, autorange = "reversed"))
+
+males.2013
+```
+
+**Fig 8: Decomposition of the female black-white life expectancy gap by cause of death in 2013 by US state**
+
+```{r cross-sectional-decomp-females}
+cod.2013.females <- cod_decomp_results %>% filter(year == 2013 & sex == "Female") %>% 
+  mutate(widened_gap = sign(COD_cont_yrs_mean), state.n = as.numeric(state))
+
+y.labels.2013 <- levels(factor(cod.2013.females$state))
+y.num.breaks.2013 <- length(levels(factor(cod.2013.females$state)))
+
+cod.2013.females <- make_pretty_decomp_plot(decomp.table = cod.2013.females, 
+                                          strat.var.1 = "state",
+                                          sign.var = "widened_gap",
+                                          decomp.var = "COD_cont_yrs_mean",
+                                          decomp.var.prop = "COD_cont_prop_mean",
+                                          partition.bar.var = "COD",
+                                          type.of.data = "practice")
+
+females.2013 <- ggplotly(
+  ggplot(cod.2013.females, aes(y = state.n, x = start)) +
+    geom_rect(aes(xmin = start, ymin = state.n - 0.45, 
+                  ymax = state.n + 0.45, 
+                  xmax = finish, fill = COD), 
+              color = "white") +
+     scale_y_continuous(breaks = 1:y.num.breaks.2013, labels = y.labels.2013) +
+     #scale_x_continuous(breaks = seq(-2, 16, by = 1), labels = seq(-2, 16, by = 1)) +
+    theme_minimal())  %>% layout(xaxis = list(title = "Contribution to gap, years"), yaxis = list(title = NA, autorange = "reversed"))
+
+#ggplot(data = cod.2013.females, aes(x=state, y=COD_cont_yrs_mean, fill=COD)) + geom_bar(stat = "identity", col = "white") + coord_flip() + theme_minimal()
+
+# p <- plot_ly(cod.2013.females, x = ~state, y = ~COD_cont_yrs_mean, type = 'bar', name = 'blah') %>%
+#   layout(yaxis = list(title = 'Count'), barmode = 'stack')
+
+# 
+# info <- plotly_build(females.2013)
+# str(info)
+# names(info$x$data[[1]])
+```
+
+**Figure 9: Distribution of the contribution of each cause of death to the life expectancy gap in males, 2013**
+
+```{r}
+ggplot(cod.2013.males, aes(COD_cont_yrs_mean)) + 
+  geom_histogram(aes(fill = COD), col = "white") + 
+  theme_minimal() + 
+  geom_vline(xintercept = 0) +
+  facet_wrap(~COD) + 
+  ylab("Number of states") + guides(fill = F) +
+  scale_y_continuous(breaks = c(seq(from = 0, to = 15, by = 5)), minor_breaks = c(seq(from = 0, to = 20, by = 5))) +
+  xlab("Contribution to gap in 2013 (years)") + scale_x_continuous(breaks = c(-1:4), minor_breaks = c(-1:4))
+```
+
+**Figure 10: Distribution of the contribution of each cause of death to the life expectancy gap in females, 2013**
+
+```{r}
+# cod.2013.females$CI.width <- cod.2013.females$COD_cont_yrs_ucl - cod.2013.females$COD_cont_yrs_lcl
+# ggplot(cod.2013.females, aes(y = COD_cont_yrs_mean, x = CI.width)) + 
+#   geom_point(aes(col = COD), alpha = 0.8) + geom_text(aes(label = stabbrs), hjust = 1.5, vjust = 0.1, size = 3) +
+#   facet_wrap(~COD, scales = "free_x") + 
+#   geom_abline(intercept = 0 , slope = 0) +
+#   ylab("Contribution to gap in 2013 (years)") + xlab("Width of credible interval")
+
+ggplot(cod.2013.females, aes(COD_cont_yrs_mean)) + 
+  geom_histogram(aes(fill = COD), col = "white") + 
+  theme_minimal() + 
+  geom_vline(xintercept = 0) +
+  facet_wrap(~COD) + 
+  ylab("Number of states") + guides(fill = F) +
+  scale_y_continuous(breaks = c(seq(from = 0, to = 15, by = 5)), minor_breaks = c(seq(from = 0, to = 20, by = 5))) + 
+  xlab("Contribution to gap in 2013 (years)") + scale_x_continuous(breaks = c(-1:4), minor_breaks = c(-1:4))
+```
+
 ###Supporting Information

### 2017-02-28 Code/combine_results.Rmd
updated with all states

@@
 
-completed.states <- included.states[c(1:11, 14, 17:24, 26:27, 33:34)]
+completed.states <- included.states
 ```
@@
 BlackWhite_results$stabbrs <- state.abb[match(BlackWhite_results$state, state.name)]
+BlackWhite_results$stabbrs[BlackWhite_results$state == "Washington DC"] <- "DC"
 

### 2017-02-28 Manuscript/Appendix-agestd-mort-rates.Rmd

@@
 # Create unique blank strip labels for empty facets
-bl = sapply(1:38, function(n) paste(rep(" ",n),collapse=""))
+bl = sapply(1:37, function(n) paste(rep(" ",n),collapse=""))
 
@@
                                                    "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", bl[20],
-                                                   "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Washington DC", bl[21],
-                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Delaware", bl[23:24],
+                                                   "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Delaware", bl[21],
+                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Washington DC", bl[23:24],
                                                    bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
-                                                   bl[30:33], "Texas", bl[34:38], "Florida"))
+                                                   bl[30:33], "Texas", bl[34:37], "Florida"))
 
@@
                                                    "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", bl[20],
-                                                   "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Washington DC", bl[21],
-                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Delaware", bl[23:24],
+                                                   "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Delaware", bl[21],
+                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Washington DC", bl[23:24],
                                                    bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
-                                                   bl[30:33], "Texas", bl[34:38], "Florida"))
+                                                   bl[30:33], "Texas", bl[34:37], "Florida"))
 ```

### 2017-02-28 Manuscript/Cover-Letter-plos.Rmd

---
output: html_document
bibliography: bibliography.bib
---

Corinne Riddell, PhD
Department of Epidemiology, Biostatistics & Occupational Health, 
McGill University, 
1020 Pine Avenue West
Montreal, QC
H2T 2E3
phone: 1-438-863-0402. email: <corinne.riddell@mail.mcgill.ca>

February 28^th^, 2017

Chief Editor, Larry Peiperl
*PLOS Medicine*

Dear Editors:

This presubmission inquiry is for our manuscript entitled, "Exploring how changes in cause of death have narrowed the black-white life expectancy gap in the United States".

The measure of life expectancy serves as a barometer for the health of a nation. In 2014, life expectancy in the United States was 78.8 years, but hidden within this figure are differences by race. Whites lived 3.4 years longer than blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both white and black Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied. During the 1980s, the life expectancy gap in the U.S. widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [@harper2007trends]. The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [@harper2014trends]. To-date, there are no published studies that have explored how changes in causes of death over time have contributed to wider life expectancy gaps in some states and narrower gaps in others. Thus, our objective was to **explore how changes in the cause of death over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013.**

**Insert what we found:**

We hope that you agree that our paper is suitable for submission to *PLOS Medicine*. Thank you for your consideration and we look forward to your response.

Sincerely,

Corinne Riddell, PhD
Department of Epidemiology, Biostatistics & Occupational Health, 
McGill University, 
email: <corinne.riddell@mail.mcgill.ca>

**References**

### 2017-03-10 Code/examining_heterogeneity.Rmd

---
title: ""
author: "Corinne Riddell"
date: "March 9, 2017"
output: html_document
---

###Examining heterogeneity

We had discussed wanting to have a quantifiable method for identifying those states that follow different trends from the rest. My first go at doing this involved taking the results of the COD decompositions over time for each state by sex and creating a linear model of the COD decompositions (Y) as a function of year and fixed effects for each state. I modelled year using a linear spline with 3 interior knots spaced equally. While this sounds rough, you'll see that it adequately captures the trends for most states. We can also consider placing the knots at pre-specified years (say 1983, 1993). The first plot illustrates the observed COD contribution (teal) vs. the estimated contribution (pink) and is ordered by how close these trends overlap.

I then calculated the residual (observed COD contribution - estimated COD contribution) at each year and the next plot shows the residuals vs. year. The sum of the absolute residuals is displayed for each state, such that the states with the highest absolute residuals are most different from the average trend. I admit that this is a rough way to do this -- it considered all states as having equal value (we could inverse weight by length of the CI for a start), but I think it allows us to capture which states are different from the average trend and speak to those. I also like that it doesn't abritrarily create a cut-off that classifies states as "the same" and "different".

```{r setup, echo = F, warning = FALSE, message=F}
library(splines)
library(broom)
library(tidyverse)
library(plotly)

cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
```

```{r modeling}
#run 12 stratified models. bs() created the b-spline basis
models <- cod_decomp_results %>% 
  group_by(COD, sex) %>%
  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state, data = .))

#calculated predicted values for all models
dfPred = broom::augment(models, fit)
dfPred$year <- 1969:2013

#merge the predictions with original dataset
cod_decomp_results <- merge(cod_decomp_results, 
                            dfPred %>% select(COD, sex, year, state, .fitted, .resid), 
                            by = c("COD", "sex", "state", "year"))

#calculated the sum of the residuals for each state-sex-cod strata (summed over years)
cod_decomp_results <- cod_decomp_results %>% group_by(state, sex, COD) %>%
  mutate(sum.residual = sum(abs(.resid)))
```

```{r plotting_fx, echo = F}
plot.resid.trends <- function(dataset, COD1, sex1) {
  subset.data <- subset(dataset, COD == COD1 & sex == sex1)
  subset.data$order <- reorder(subset.data$state, subset.data$sum.residual, first)

  a <- ggplot(subset.data,
         aes(x = year, y = COD_cont_yrs_mean)) + geom_line(aes(col = "Observed")) + 
    facet_wrap( ~ order) + 
    geom_line(aes(y = .fitted, col = "Estimated")) +
    xlab(" ") + ylab("observed vs. estimated contribution (years)") + 
    ggtitle(paste0("Observed vs. estimated in ", COD1, " in ", sex1)) +
    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

  b <- ggplot(subset.data,
              aes(x = year, y = .resid)) + geom_line(aes(col = "Residual")) + 
    facet_wrap( ~ order) +
    xlab(" ") + ylab("Residual") +
    geom_hline(yintercept = 0, col = "black") +
      geom_text(data = subset(subset.data, year == 2010), 
                aes(y = 1, label = round(sum.residual, 1)), size = 2.5) + 
    ggtitle(paste0("Residuals vs. year in ", COD1, " in ", sex1)) +
    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

    return(list(plot1 = a, plot2 = b))
}
```

```{r create-the-plots, echo = F, message = F}
for(COD_i in levels(cod_decomp_results$COD)) {
  for(sex_i in levels(cod_decomp_results$sex)) {
    assign(paste0(COD_i, ".", sex_i, ".plot"), plot.resid.trends(dataset = cod_decomp_results, 
                                                                         COD1 = COD_i, sex1 = sex_i))
  }
}
```

```{r print-em, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cardiovascular.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Cardiovascular.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
```

```{r print-em-2, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Cardiovascular.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
```

```{r print-em-3, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Cancers.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
```

```{r print-em-4, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Cancers.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
```

```{r print-3, fig.width = 9, fig.height=9, echo = F}
ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Communicable.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Communicable.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`Non-communicable.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`Non-communicable.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Injuries.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Injuries.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(`All other causes.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`All other causes.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(`All other causes.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`All other causes.Female.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
```

### 2017-03-10 Code/combine_results.Rmd
small change

@@
 
+```{r prep_decomp_for_plotting, echo = F}
+cod_decomp_results$widened_gap <- sign(cod_decomp_results$COD_cont_yrs_mean)
+cod_decomp_results <- make_pretty_decomp_plot(decomp.table = cod_decomp_results, strat.var.1 = "stratum.id", 
+                                   partition.bar.var = "COD", sign.var = "widened_gap",
+                                   decomp.var = "COD_cont_yrs_mean", decomp.var.prop = "COD_cont_prop_mean",
+                                   type.of.data = "practice")
+
+```
+
 ```{r save_combined_datasets, echo = F}

### 2017-03-10 Manuscript/Appendix-agestd-mort-rates.Rmd
cant remember

@@
 ```{r plot-arrange-as-us-map, echo=F, fig.width=19.5, fig.height=14, results="asis", message = F}
-roundUp <- function(x) 10^ceiling(log10(x))
-
 # sex_i <- "Male"
@@
       geom_hline(aes(yintercept = 0), col = "#636363") +
-      scale_y_continuous(limits = c(0, lim.y)) + 
+      scale_y_continuous(limits = c(0, lim.y2)) + 
       facet_wrap( ~ state.reorder2, ncol = 11, drop = F, strip.position="bottom") +

### 2017-03-10 shiny_app/app.R
updated shiny app with newest results

@@ -1,0 +1,417 @@
+library(shiny)
+library(plotly)
+library(viridis)
+library(scales)
+library(shinythemes)
+library(dplyr)
+
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+
+age_cod_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results.csv")
+age_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_results.csv")
+cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
+cod_change_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_change_results.csv")
+BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
+dat.aggregated <- read.csv('/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.csv') 
+
+age_cod_results$COD <- factor(age_cod_results$COD, levels(age_cod_results$COD)[c(3, 2, 4, 6, 5, 1)])
+cod_decomp_results$COD <- factor(cod_decomp_results$COD, levels(cod_decomp_results$COD)[c(3, 2, 4, 6, 5, 1)])
+age_decomp_results$age <- factor(age_decomp_results$age, levels = levels(age_decomp_results$age)[c(1, 2, 11, 3:10, 12:19)])
+age_cod_results$age <- factor(age_cod_results$age, levels = levels(age_cod_results$age)[c(1, 2, 11, 3:10, 12:19)])                                    
+
+cod_change_results$Cause.of.death <- factor(cod_change_results$Cause.of.death, levels(cod_change_results$Cause.of.death)[c(3, 2, 4, 6, 5, 1)])
+
+cod_marginal_results <- cod_decomp_results %>% 
+  group_by(stratum.id) %>% 
+  summarise(state = first(state), sex= first(sex), year = first(year), total_Cx = sum(COD_cont_yrs_mean))
+
+#cod_marginal_results_female <- cod_marginal_results %>% filter(sex == "Female") %>% mutate(reorder(state, total_Cx))
+
[…]

### 2017-03-10 Code/combine_results.Rmd

@@
 
+age_decomp_results$widened_gap <- sign(age_decomp_results$age_cont_yrs_mean)
+age_decomp_results <- make_pretty_decomp_plot(decomp.table = age_decomp_results, strat.var.1 = "stratum.id", 
+                                   partition.bar.var = "age_minbin", sign.var = "widened_gap",
+                                   decomp.var = "age_cont_yrs_mean", decomp.var.prop = "age_cont_prop_mean",
+                                   type.of.data = "practice")
+```
+
+```{r split-age-cod-data-for-save}
+age_cod_results_female <- age_cod_results %>% filter(sex == "Female")
+age_cod_results_male <- age_cod_results %>% filter(sex == "Male")
 ```
@@
 ```{r save_combined_datasets, echo = F}
-write.csv(age_cod_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results.csv")
+write.csv(age_cod_results_female, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results_female.csv")
+write.csv(age_cod_results_male, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results_male.csv")
 write.csv(age_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_results.csv")

### 2017-03-28 Code/examining_heterogeneity.Rmd
heterogeneity analysis

@@
 cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
+
+cod_decomp_results$COD_cont_yrs_width <- cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl
+cod_decomp_results$COD_cont_yrs_SE1 <- (cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_mean)/1.96
+cod_decomp_results$COD_cont_yrs_SE2 <- -(cod_decomp_results$COD_cont_yrs_lcl - cod_decomp_results$COD_cont_yrs_mean)/1.96
+cod_decomp_results$SE.diff <- cod_decomp_results$COD_cont_yrs_SE1 - cod_decomp_results$COD_cont_yrs_SE2
+summary(cod_decomp_results$SE.diff)
+
+# cod_decomp_results.means <- cod_decomp_results %>% 
+#   mutate(COD_cont_yrs_width = COD_cont_yrs_ucl - COD_cont_yrs_lcl) %>%
+#   group_by(state, sex, COD) %>%
+#   summarise(mean_COD_cont_yrs_width = mean(COD_cont_yrs_width))
+# 
+# mod1 <- lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state, data = subset.data)
+# 
+# mod2 <- lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state, weights = 1/(COD_cont_yrs_SE1)^2, 
+#            data = subset.data)
 ```
@@
   group_by(COD, sex) %>%
-  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state, data = .))
+  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state, 
+              weights = 1/(COD_cont_yrs_SE1)^2, 
+              data = .))
 
@@
 #calculated the sum of the residuals for each state-sex-cod strata (summed over years)
-cod_decomp_results <- cod_decomp_results %>% group_by(state, sex, COD) %>%
+cod_decomp_results <- cod_decomp_results %>% 
+  group_by(state, sex, COD) %>%
   mutate(sum.residual = sum(abs(.resid)))
@@
   a <- ggplot(subset.data,
-         aes(x = year, y = COD_cont_yrs_mean)) + geom_line(aes(col = "Observed")) + 
+         aes(x = year, y = COD_cont_yrs_mean)) + 
+    #geom_line(aes(col = "Observed")) + 
+    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#b2e2e2") +
+    #scale_fill_manual(values = c("#b2e2e2")) +
     facet_wrap( ~ order) + 
-    geom_line(aes(y = .fitted, col = "Estimated")) +
+    geom_line(aes(y = .fitted), col = "#006d2c") +
+    #scale_color_manual(values = c("#006d2c")) +
     xlab(" ") + ylab("observed vs. estimated contribution (years)") + 
@@
               aes(x = year, y = .resid)) + geom_line(aes(col = "Residual")) + 
+    scale_color_manual(values = c("#006d2c")) +    
     facet_wrap( ~ order) +

### 2017-03-28 Code/combine_results.Rmd
forgot

@@
                                    type.of.data = "practice")
+
+age_cod_results$widened_gap <- sign(age_cod_results$age_COD_cont_yrs_mean)
+age_cod_results <- make_pretty_decomp_plot(decomp.table = age_cod_results, strat.var.1 = "stratum.id",
+                                           strat.var.2 = "state", strat.var.3 = "age_minbin",
+                                           partition.bar.var = "COD", sign.var = "widened_gap",
+                                           decomp.var = "age_COD_cont_yrs_mean", decomp.var.prop = "age_COD_prop_mean",
+                                           type.of.data = "practice")
+
+#trying to get this to work 
+age_cod_results_test <- make_pretty_decomp_plot(decomp.table = subset(age_cod_results, stratum.id == "Alabama.1969.Female"), 
+                                                strat.var.1 = "stratum.id",
+                                           strat.var.2 = "state", strat.var.3 = "age_minbin",
+                                           partition.bar.var = "COD", sign.var = "widened_gap",
+                                           decomp.var = "age_COD_cont_yrs_mean", decomp.var.prop = "age_COD_cont_prop_mean",
+                                           type.of.data = "results")
+
+ggplot(data = age_cod_results_test, aes(y = age, x = start)) + geom_bar(fill = COD)
+
+age_cod_results_test$age.n <- as.numeric(age_cod_results_test$age) #need to fix the 5 year old when convert to nummeric
+    ggplotly(ggplot(age_cod_results_test,
+                    aes(y = age.n, x = start)) + 
+               geom_rect(aes(xmin = start, xmax = finish, ymin = age.n - 0.45, ymax = age.n + 0.45, fill = COD), col = "white") +
+               #xlab(paste0("Contribution to life expectancy gap ", xaxis.title())) +
+               #ylab("Age group\n") +
+               theme_minimal() +
+               geom_vline(xintercept = 0)  
+               #facet_wrap(~ year) +
+               #ggtitle(paste0(input$selected_sex, "s in ", input$state)) 
+             ) 
 ```

### 2017-03-29 Manuscript/Appendix-agestd-mort-rates.Rmd

@@
 
-```{r plot-arrange-as-us-map, echo=F, fig.width=19.5, fig.height=14, results="asis", message = F}
-# sex_i <- "Male"
-# cod <- "Cardiovascular"
+```{r plotting_fx, echo = F}
+plot.mortality.trends <- function(dataset, dataset2, COD1, sex1) {
+  subset.data <- subset(dataset, COD == COD1 & sex == sex1)
+  #lim.y <- c(min(subset.data$rate.per.100k_lcl), max(subset.data$rate.per.100k_ucl))
 
-for(sex_i in c("Male", "Female")) {
-  for(cod in levels(mortality.rates$COD)[c(3, 2, 4, 6, 5, 1)]) {
-# for(sex_i in c("Male")) {
-#   for(cod in levels(mortality.rates$COD)[1]) {
-    cat("  \n### Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod, "  \n")
+  subset.data2 <- subset(dataset2, sex == sex1 & COD == COD1)
+  #lim.y2 <- c(min(subset.data2$rate.difference_LCL), max(subset.data2$rate.difference_UCL))
 
-    data.sub <- subset(mortality.rates, sex == sex_i & COD == cod)
-    lim.y <- max(data.sub$rate.per.100k_mean)
-    map.plot1 <- ggplot(data.sub,
-                        aes(y = rate.per.100k_mean, x = year)) +
-      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
-      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
-      geom_line(aes(lty = Race, col = Race)) +
-      geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, group = Race, fill = Race), col = NA, alpha = 0.5) +
-      scale_color_manual(values = c("#2166ac", "#b2182b")) +
-      scale_fill_manual(values = c("#67a9cf", "#ef8a62")) +
-      scale_y_continuous(limits = c(0, lim.y)) + 
-      facet_wrap( ~ state.reorder2, ncol = 11, drop = F, strip.position="bottom") +
-      theme_classic(base_size = 20) +
+  a <- ggplot(subset.data, aes(x = year, y = rate.per.100k_mean)) + 
+              geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, fill = Race), col = NA, alpha = 0.5) +
+              geom_line(aes(col = Race)) +
+              facet_wrap( ~ state.reorder2, ncol = 11, drop = F) +
+              xlab(" ") + 
+              ylab("Age-standardized mortality rate (per 100,000)") + 
+              geom_hline(yintercept = 0, col = "black", lwd = 0.5) +
+              #scale_y_continuous(limits = lim.y) + 
+              ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex1, " for ", COD1)) +
+              theme_classic(base_size = 10) +
               theme(axis.text.x = element_blank(),
@@
                     axis.line=element_blank(),
-            axis.ticks=element_blank()) +
-      #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-      ylab("Age-standardized mortality rate (per 100,000)") +
-      xlab("Year (1969-2013)")
-    print(map.plot1)
+                    axis.ticks=element_blank())
 
-    cat("  \n")
-
-    cat("  \n### Black-white difference in age-adjusted mortality (per 100,000) in ", 
-        sex_i, " for ", cod, "  \n")
-
-    data.sub2 <- subset(mortality.rates.diff, sex == sex_i & COD == cod)
-    lim.y2 <- max(data.sub2$rate.difference_mean)
-
-    map.plot2 <- ggplot(data.sub2,
-                        aes(y = rate.difference_mean, x = year)) +
-      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
-      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
-      geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL), fill = "grey", col = NA) +
+  b <- ggplot(subset.data2, aes(x = year, y = rate.difference_mean)) + 
               geom_line() +
-      geom_hline(aes(yintercept = 0), col = "#636363") +
-      scale_y_continuous(limits = c(0, lim.y2)) + 
-      facet_wrap( ~ state.reorder2, ncol = 11, drop = F, strip.position="bottom") +
-      theme_classic(base_size = 20) +
-      theme(axis.text.x = element_blank(),
+              geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL), col = NA, alpha = 0.5) +
+              facet_wrap( ~ state.reorder2, ncol = 11, drop = F) +
+              xlab(" ") + 
+              ylab("Excess mortality among blacks (per 100,000)") +
+              #scale_y_continuous(breaks = 0) + #limits = lim.y2
+              geom_hline(yintercept = 0, col = "black", lwd = 0.5) +
+              #geom_vline(xintercept = 1969, col = "black", lwd = 0.5) +
+              ggtitle(paste0("Excess mortality among blacks (per 100,000) in ", sex1, " for ", COD1)) +
+              theme_classic(base_size = 10) +
+              theme(panel.grid.minor = element_blank(),
+                    axis.text.x = element_blank(),
                     strip.background=element_blank(),
                     axis.line=element_blank(),
-            axis.ticks=element_blank()) +
-      #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-      ylab("Black-white difference (per 100,000)") +
-      xlab("Year (1969-2013)")
+                    axis.ticks=element_blank())
 
-    print(map.plot2)
-    cat("  \n")
-    cat("  \n")
-  }
-  cat("  \n")
+    return(list(plot1 = a, plot2 = b))
 }
@@
 
-```{r, echo=F, fig.width=11, fig.height=11, eval=F}
-
-sex_i <- "Male"
-cod <- "Injuries"
-
-for(sex_i in c("Male", "Female")) {
-  for(cod in levels(mortality.rates$COD)) {
-
-# for(sex_i in c("Male")) {
-#   for(cod in levels(mortality.rates$COD)[1]) {
-    data.sub <- subset(mortality.rates, sex == sex_i & COD == cod)
-    a <- ggplot(data.sub, 
-                aes(y = rate.per.100k_mean, x = year)) + 
-      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
-      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
-      geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, fill = Census_Division, group = Race), col = NA, alpha = 0.5) +
-      geom_line(aes(col = Census_Division, lty = Race)) +
-      # geom_line(aes(y = rate.per.100k_med, lty = Race)) +
-      facet_wrap( ~ state.reorder, ncol = 5) +
-      ylab("Age-standardized mortality rate (per 100,000)") +
-      xlab("Year") + 
-      ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))# + theme_minimal()
-
-    print(a)
-
-    b <- ggplot(subset(mortality.rates.diff, sex == sex_i & COD == cod), 
-                aes(y=rate.difference_mean, x = year)) + 
-      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
-      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
-      geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL, fill = Census_Division), col = NA, alpha = 0.5) +
-      geom_hline(aes(yintercept = 0), col = "#636363") +
-      #geom_line(aes(y = rate.difference_med)) +
-      geom_line(aes(col = Census_Division)) +
-      facet_wrap( ~ state.reorder, ncol = 5) +
-      ylab("Black-white difference (per 100,000)") +
-      xlab("Year") + 
-      ggtitle(paste0("Black-white difference in age-adjusted mortality (per 100,000) in ", 
-                     sex_i, " for ", cod))
-    print(b)
-
+```{r create-the-plots, echo = F, message = F}
+for(COD_i in levels(mortality.rates$COD)) {
+  for(sex_i in levels(mortality.rates$sex)) {
+    assign(paste0(COD_i, ".", sex_i, ".mort.plot"), plot.mortality.trends(dataset = mortality.rates, dataset2 = mortality.rates.diff,
+                                                                         COD1 = COD_i, sex1 = sex_i))
   }
@@
 ```
+
+```{r print-em, fig.width = 9, fig.height = 9, echo = F}
+ggplotly(Cardiovascular.Male.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
+ggplotly(Cardiovascular.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+```
+
+```{r print-em-2, fig.width = 9, fig.height = 9, echo = F}
+ggplotly(Cardiovascular.Female.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
+ggplotly(Cardiovascular.Female.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+```
+
+```{r print-em-3, fig.width = 9, fig.height = 9, echo = F}
+ggplotly(Cancers.Male.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
+ggplotly(Cancers.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+```
+
+```{r print-em-4, fig.width = 9, fig.height = 9, echo = F}
+ggplotly(Cancers.Female.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
+ggplotly(Cancers.Female.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+```
+
+```{r print-3, fig.width = 9, fig.height=9, echo = F}
+ggplotly(Communicable.Male.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
+ggplotly(Communicable.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+
+ggplotly(Communicable.Female.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
+ggplotly(Communicable.Female.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+
+ggplotly(`Non-communicable.Male.mort.plot`[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
+ggplotly(`Non-communicable.Male.mort.plot`[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+
+ggplotly(`Non-communicable.Female.mort.plot`[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
+ggplotly(`Non-communicable.Male.mort.plot`[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+
+ggplotly(Injuries.Male.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
+ggplotly(Injuries.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+
+ggplotly(Injuries.Female.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
+ggplotly(Injuries.Female.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+
+ggplotly(`All other causes.Male.mort.plot`[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
+ggplotly(`All other causes.Male.mort.plot`[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+
+ggplotly(`All other causes.Female.mort.plot`[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
+ggplotly(`All other causes.Female.mort.plot`[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+```

### 2017-03-29 Manuscript/Manuscript_COD_Trends.Rmd
new manuscript

--- 2017-02-24 Manuscript/Manuscript_COD_decomp.Rmd
@@
 ---
-title: Exploring how changes in cause of death have narrowed the black-white life
-  expectancy gap in the United States.
+title: 
 output:
@@
 
-The overall difference in life expectancy between black and whites in the US has generally decreased over time.
-States have fared differently in their progress towards eliminating the life expectancy gap, with states such as New York experiencing greater reductions to the gap compared to states such as California and Texas.
-The objective of this study is to compare how changes to cause-specific mortality between 1969 and 2013 have resulted in narrower life expectancy gaps in some states and  wider gaps in other states.
-
 **Methods and Findings**
-Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes according to year (1969-2013), state, age, sex, race (black or white), and six cause of death categories (Cancer, cardiovascular disease, communicable disease, non-communicable disease, injuries, and all other causes).
-We excluded 10 states with the smallest black populations, but included the District of Columbia.
-A Bayesian autoregressive model was used to model smoothed mortality rates over time within each stratum, and a truncated Poisson model was used to impute death counts that were suppressed due to small cell sizes.
-We calculated life expectancy for blacks and whites within each stratum using traditional life table methods, and decomposed the difference in life expectancy by cause of death using Arriaga's method.
-
-[Add findings]
 
 **Conclusions:**
-[Add conclusion]
 
@@
 While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.
-During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [Harper 2007].
+
+During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [@harper2007trends].
 The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement.
-For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014].
-This study seeks to better understand these state-level differences, by examining how changes in the cause of death over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013.
+For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [@harper2014trends].
+
+The over-arching objective of this study was to describe the general pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and differences between the sexes.
+To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for males and females.
 Common reasons for the decreasing life expectancy gap can thereby be identified, and anomalous states can be examined.
-Ideally, one would like to identify policies or social conditions in highly successful states that help to explain their unusual performance.
+Ideally, this study may facilitate the identification of one policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.
 
@@
 **The data**
-
 We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.
@@
 **Statistical Methods**
-
 The data structure produced two statistical challenges that were addressed by our quantitative approach.
@@
 Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9).
-These models were run separately for every combination of state, race, and sex, and produced death counts that were smoothed over time for every cause of death within each age group.
+
+These models were run separately for every combination of state, race, and sex.
+They were fit using two markov chains of length 10,000.
+One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration (to reduce autocorrelation).
+Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition.
+These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.
 
@@
 We summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference, for each state and for each sex.
+Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 45 state-level trends and their credible intervals.
 
-In order to understand whether states fared similarly over time, we divided the time frame into three eras which roughly encapsulate an initial period of increases in life expectancy for both races and sexes (1969-1983), an intermediate period of decreases and stagnating life expectancy for blacks but not for whites (1983-1993) and a final period of increasing life expectancy for both races (1993-2013).
-While these eras represent different lengths of time, they best capture three distinct periods of changes to the difference in life expectancy: an intial narrowing of the gap, followed by a widening of the gap that diminished previous progress, and the most recent twenty years of sustained narrowing (**S1 Fig** and **S2 Fig**).
+To describe the general pattern in these trend lines, we regressed each state's mean contribution against a linear spline for calendar time that was estimated with five degrees of freedom.
+This model also contained fixed effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year.
+We then calculated residuals for each year by substracting the meta-estimated contribution from the smoothed-observed estimated contribution and summed these residuals over time for each state to rank states according to how well the model predicted their pattern.
+Plots of these residuals over time were used to identify divergent states.
 
-For each era, we calculated the degree that the life expectancy gap has narrowed or widened and decomposed this *change* to the *difference* in the life expectancy between blacks and whites according to cause of death.
-To do so, we simply subtracted the contribution (in years) that each cause of death category had made to the difference in life expectancy at the beginning of the era from that made at the beginning of the era.
-An example of the calculation is found in **S1 Text**.
-
-*ktm to read over*: The bayesian model was fit using two markov chains of length 10,000.
-One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration (to reduce autocorrelation).
-These 1,000 iterations approximate the shape of the posterior distribution, and we used the mean as the estimate of the log mortality rate, and the 2.5th and 97.5th percentiles as the lower and upper bound of the 95% credible interval (CI).
-
-All data analyses were performed using R version 3.3.1.
-The Bayesian smoothing was performed using `rjags` (version 4-6) and `R2jags` (version 0.5-7), and the life expectancy and decomposition analysis were performed using user-written functions found in the Supplement.
+**Need to add information on the trends in mortality**
 
@@
 
-```{r initialize, echo=FALSE, warning=F, message=F}
-library(shiny)
-library(plotly)
-library(viridis)
-library(scales)
-library(shinythemes)
-library(dplyr)
-library(purrr)
-library(stringr)
-library(DT)
-source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
-```
-
-```{r read_data, echo=F}
-age_cod_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results.csv")
-age_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_results.csv")
-cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
-cod_change_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_change_results.csv")
-BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
-
-age_cod_results$COD <- factor(age_cod_results$COD, levels(age_cod_results$COD)[c(3, 2, 4, 6, 5, 1)])
-age_decomp_results$age <- factor(age_decomp_results$age, levels = levels(age_decomp_results$age)[c(1, 2, 11, 3:10, 12:19)])
-age_cod_results$age <- factor(age_cod_results$age, levels = levels(age_cod_results$age)[c(1, 2, 11, 3:10, 12:19)])
-
-cod_change_results$Cause.of.death <- factor(cod_change_results$Cause.of.death, levels(cod_change_results$Cause.of.death)[c(3, 2, 4, 6, 5, 1)])
-```
-
-```{r prep_era_figure, echo = F}
-year1 <- 1969
-year2 <- 1983
-year3 <- 1993
-year4 <- 2013
-
-bw_dat <- BlackWhite_results %>% 
-  filter(year %in% c(year1, year2, year3, year4), is.na(LE_wbgap_mean) == F) %>%
-  select(state, year, LE_wbgap_mean, sex) %>%
-  arrange(year)
-
-bw_dat <- tidyr::spread(bw_dat, year, LE_wbgap_mean) 
-names(bw_dat)[3:6] <- c("first.gap","second.gap", "third.gap", "fourth.gap")  
-bw_dat.males <- bw_dat %>% filter(sex == "Male") %>%
-  mutate(era1.diff = first.gap - second.gap,
-         era2.diff = second.gap - third.gap,
-         era3.diff = third.gap - fourth.gap,
-         state.gdiff.order.1 = reorder(state, first.gap),
-         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
-         state.gdiff.order.2 = reorder(state, second.gap),
-         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
-         state.gdiff.order.3 = reorder(state, third.gap),
-         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
-  )
-
-bw_dat.females <- bw_dat %>% filter(sex == "Female") %>%
-  mutate(era1.diff = first.gap - second.gap,
-         era2.diff = second.gap - third.gap,
-         era3.diff = third.gap - fourth.gap,
-         state.gdiff.order.1 = reorder(state, first.gap),
-         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
-         state.gdiff.order.2 = reorder(state, second.gap),
-         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
-         state.gdiff.order.3 = reorder(state, third.gap),
-         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
-  )
-```
-
-```{r more_figure_prep, echo = F}
-widths <- cod_change_results %>% mutate(tot_Cx_era1_CI_width = tot_Cx_era1_diff_ucl - tot_Cx_era1_diff_lcl,
-                                        tot_Cx_era2_CI_width = tot_Cx_era2_diff_ucl - tot_Cx_era2_diff_lcl,
-                                        tot_Cx_era3_CI_width = tot_Cx_era3_diff_ucl - tot_Cx_era3_diff_lcl) %>%
-  filter(Cause.of.death == "Injuries") %>%
-  select(state, sex, tot_Cx_era1_CI_width, tot_Cx_era2_CI_width, tot_Cx_era3_CI_width)
-
-widths <- tidyr::gather(widths, tot_Cx_era1_CI_width, tot_Cx_era2_CI_width, tot_Cx_era3_CI_width, key = "era", value = "tot_Cx_CI_width")
-ncols <- str_split_fixed(as.character(widths$era), "_", 5)
-widths$era <- as.factor(ncols[ , 3])
-
-cod.df <- cod_change_results %>% select(state, sex, Cause.of.death, Cx_era1_diff_mean, Cx_era2_diff_mean, Cx_era3_diff_mean)
-cod.df <- tidyr::gather(cod.df, Cx_era1_diff_mean, Cx_era2_diff_mean, Cx_era3_diff_mean, key = "era", value = "Cx_diff_mean")
-ncols <- str_split_fixed(as.character(cod.df$era), "_", 4)
-cod.df$era <- as.factor(ncols[ , 2])
-
-cod.df2 <- cod_change_results %>% select(state, sex, Cause.of.death, Cx_p_era1_diff_mean, Cx_p_era2_diff_mean, Cx_p_era3_diff_mean)
-cod.df2 <- tidyr::gather(cod.df2, Cx_p_era1_diff_mean, Cx_p_era2_diff_mean, Cx_p_era3_diff_mean, key = "era", value = "Cx_p_diff_mean")
-ncols <- str_split_fixed(as.character(cod.df2$era), "_", 5)
-cod.df2$era <- as.factor(ncols[ , 3])
-
-cod.df <- merge(cod.df, cod.df2, by = c("state", "sex", "Cause.of.death", "era"))
-
-cod.df <- merge(cod.df, widths, by = c("state", "sex", "era"))
-
-cod.df$narrowed_gap <- T
-cod.df$narrowed_gap[cod.df$Cx_diff_mean < 0 ] <- F
-
-cod.df <- make_pretty_decomp_plot(decomp.table = cod.df, strat.var.1 = "state",
-                                  strat.var.2 = "sex", strat.var.3 = "era", sign.var = "narrowed_gap",
-                                  decomp.var = "Cx_diff_mean",
-                                  decomp.var.prop = "Cx_p_diff_mean",
-                                  partition.bar.var = "Cause.of.death",
-                                  type.of.data = "results")
-
-cod.df <- merge(cod.df, BlackWhite_results[BlackWhite_results$year == 1969 & BlackWhite_results$sex == "Male", 
-                                           c("state", "Census_Region", "Census_Division", "stabbrs")], by = c("state"))
-cod.df.males <- merge(subset(cod.df, sex == "Male"), bw_dat.males, by = c("state"))
-cod.df.males <- cod.df.males %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
-                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
-                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)
-
-cod.df.males$tot_Cx_CI_width2 <- -(cod.df.males$tot_Cx_CI_width - min(cod.df.males$tot_Cx_CI_width))/(max(cod.df.males$tot_Cx_CI_width) - min(cod.df.males$tot_Cx_CI_width)) + 1
-cod.df.males$width.factor <- as.factor(cut(cod.df.males$tot_Cx_CI_width2, breaks = c(-0.01, 0.25, 0.50, 0.75, 1)))
-
-```
-
-*Changes to the difference in life expectancy in US males, 1969-1983*
-
-**Fig 1** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs.
-their gap in 1983 (solid dashed line), with the change in the gap being represented by the number of years between these lines along the x-axis.
-This gap is decomposed by cause of death, where causes to the left of the dashed line contributed to narrowing the gap and causes to the right contributed to widening the gap.
-The length of these coloured bars represent the number of years attributable to the cause along the x-axis.
-
-For example, in the **Fig 1**, New York's gap changed from 8.8 years in 1969 to 6.3 in 1983.
-Causes to the left of the dashed line -- All other causes (pink), injuries (blue), non-communicable diseases (teal), and communicable diseases (green) -- each contributed to narrowing the gap, while cardiovascular disease (red) and cancer (yellow) contributed to widening the gap.
-Since the contribution in years of the leftward causes is larger than the contribution of the rightward causes, the net change is a smaller life expectancy gap in 1983, where all other causes (the longest bar) made the largest contribution.
-
-```{r, echo = F}
-sum1 <- BlackWhite_results %>% filter(year == 1969 & sex == "Male") %>% 
-  arrange(LE_wbgap_mean) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)
-
-diff <- bw_dat.males %>% arrange(era1.diff)
-
-for(i in 1:5) {
-  j <- i-1
-
-  max.value <- max(cod.df$Cx_diff_mean[cod.df$state == diff[dim(diff)[1]-j, ]$state & 
-                                            cod.df$era == "era1" &
-                                            cod.df$sex == "Male" ])
-
-  max.cause <- cod.df$Cause.of.death[cod.df$state == diff[dim(diff)[1]-j, ]$state &
-                    cod.df$sex == "Male" &
-                    cod.df$era == "era1" & 
-                    cod.df$Cx_diff_mean == max.value]
-
-  min.value <- min(cod.df$Cx_diff_mean[cod.df$state == diff[i, ]$state & 
-                                            cod.df$era == "era1" &
-                                            cod.df$sex == "Male" ])
-
-  min.cause <- cod.df$Cause.of.death[cod.df$state == diff[i, ]$state &
-                    cod.df$sex == "Male" &
-                    cod.df$era == "era1" & 
-                    cod.df$Cx_diff_mean == min.value]
-
-  assign(paste0("max", i), max.value)
-  assign(paste0("max.cause", i), max.cause)
-  assign(paste0("min", i), min.value)
-  assign(paste0("min.cause", i), min.cause)
-}
-```
-
-```{r, echo =F}
-sum2 <- BlackWhite_results %>% filter(year == 1983 & sex == "Male") %>% 
-  arrange(LE_wbgap_mean) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)
-
-  min.value.DC <- min(cod.df$Cx_diff_mean[cod.df$state == "Washington DC" & 
-                                            cod.df$era == "era2" &
-                                            cod.df$sex == "Male" ])
-
-  min.cause.DC <- cod.df$Cause.of.death[cod.df$state == "Washington DC" &
-                    cod.df$sex == "Male" &
-                    cod.df$era == "era2" & 
-                    cod.df$Cx_diff_mean == min.value.DC]
-```
-
-```{r, echo =F}
-sum3 <- BlackWhite_results %>% filter(year == 1993 & sex == "Male") %>% 
-  arrange(LE_wbgap_mean) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)
-
-sum4 <- BlackWhite_results %>% filter(year == 2013 & sex == "Male") %>% 
-  arrange(LE_wbgap_mean) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)
-```
-
-In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap_mean, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `.
-Injury was the largest contributor towards narrowing the gap during this time period, as both race groups experienced decreased rates of injury-related mortality, but blacks had larger rates decreases, especially in states with relatively higher rates in 1969 (**S3-S14 Figs**).
-A few states, including Minnesota and Nebraska, had relatively low rates of injuries among blacks in 1969 and smaller gaps.
-The residual category also contributed to narrower gaps in most states, but generally made a smaller contribution than injuries.
-In New York, the residual category made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks (**S15 Fig**).
-Communicable diseases contributed modestly to narrowing the gap, except in Arizona and Minnesota.
-Generally, cardiovascular disease contributed towards increasing the gap, and reflects slower declines for blacks in cardiovascular mortality than for whites.
-Cancer contributed towards increasing the gap in most states, and largely reflects increased cancer-related mortality for both races during this period, with blacks undergoing steeper changes.
-In most states, non-communicable diseases contributed a small amount towards increasing the gap, although it decreased the gap among four states, most notably in New York.
-
-For `r dim(bw_dat.males[bw_dat.males$era1.diff > 0,])[1]` of the `r dim(bw_dat.males)[1]` states, the gap decreased between 1969 and 1983, with the largest decreases in `r diff[dim(diff)[1], ]$state` (decrease in the gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)` years, `r round(max1,1)` years, of which is due to `r max.cause1`), `r diff[dim(diff)[1]-1, ]$state` (change of `r round(diff[dim(diff)[1]-1, ]$era1.diff, 1)` years, `r round(max2,1)` due to `r max.cause2`), `r diff[dim(diff)[1]-2, ]$state` (change of `r round(diff[dim(diff)[1]-2, ]$era1.diff, 1)` years, `r round(max3,1)` due to `r max.cause3`), `r diff[dim(diff)[1]-3, ]$state` (change of `r round(diff[dim(diff)[1]-3, ]$era1.diff, 1)` years, `r round(max4,1)` due to `r max.cause4`), and `r diff[dim(diff)[1]-4, ]$state` (change of `r round(diff[dim(diff)[1]-4, ]$era1.diff, 1)` years, `r round(max5,1)` due to `r max.cause5`).
-
-The gap increased for `r dim(bw_dat.males[bw_dat.males$era1.diff < 0,])[1]` states.
-Most notably in `r diff$state[1]` (increase in gap of `r round(abs(diff$era1.diff[1]), 1)` years, `r abs(round(min1,1))` due to `r min.cause1`), `r diff$state[2]` (increase in gap of `r round(abs(diff$era1.diff[2]), 1)` years, `r abs(round(min2,1))` due to `r min.cause2`), `r diff$state[3]` (increase in gap of `r round(abs(diff$era1.diff[3]), 1)` years, `r abs(round(min3,1))` due to `r min.cause3`), `r diff$state[4]` (increase in gap of `r round(abs(diff$era1.diff[4]), 1)` years, `r abs(round(min4,1))` due to `r min.cause4`).
-For these states, injuries do not contribute as much to narrowing the gap as in other states (and in Minnesota, injuries widened the gap), so these disparity increases appear to be due to a combination of changes to cardiovascular-related mortality making blacks relatively worse off and less changes to injuries than experienced by other states.
-
-During the second time period, from 1983 to 1993, the life expectancy gap for males widened in `r dim(bw_dat.males[bw_dat.males$era2.diff < 0, ])[1]` of the `r dim(bw_dat.males)[1]` states (**Fig 2**).
-For states such as New York, Maryland, New Jersey, and Connecticut, the biggest driver of the widening gap was communicable disease, reflecting the heightened burden from HIV/AIDS in black males during the height of the AIDS epidemic.
-While both blacks and whites experienced a dramatic increase in mortality related to communicable diseases, blacks' rates were even higher.
-Injuries also contributed to the gap widening, most notably in Wisconsin, Maryland, and Louisiana.
-For many states, injury-related mortality increased for blacks but not for whites during this time period.
-Compared to the previous period, these two causes of death led to widwer rather than narrowed gaps.
-Cancer and cardiovascular disease contributed towards wider gaps, although the magnitude of the contribution of cardiovascular disease is quite hetergenous by state, with Mississippi, Minnesota, and Pennsylvania seeing the largest absolute contributions.
-The residual category contributed to a narrower gap, especially in New York, Pennsylvania, Connecticut, Mississippi, and Florida [note: This wasn't related to infant mortality improvements among blacks in NY, PA, but partially related to IMR improvements in CT, MS, and FL].
-Most notably, the District of Columbia had the highest gap to begin and underwent the largest increase to their gap, which increased from 10.5 years to 15.9 years over this 10-year period, reflecting an increase in the gap of `r abs(round(bw_dat.males$era2.diff[bw_dat.males$sex == "Male" & bw_dat.males$state == "Washington DC"], 1))` years, of which `r round(min.value.DC, 1)` was due to `r min.cause.DC`.
-
-The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Fig 3**).
-In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum3[dim(sum3)[1], ]$state `.
-By 2013, the gap ranged from `r round(sum4[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum4[dim(sum4)[1], ]$state `.
-During this time period all of the causes contributed to narrowing the gap for most states.
-In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardiovascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.
-
-**Fig 1: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
-```{r figure1, fig.align = "center", echo=F, fig.height=9, fig.width=7}
-y.labels <- levels(factor(cod.df.males$state.gdiff.order.1))
-y.num.breaks <- length(levels(factor(cod.df.males$state.gdiff.order.1)))
-
-males.era1 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "era1"), aes(y = state.gdiff.num.1, x = new.start.1)) +
-    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
-                  ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.1, fill = Cause.of.death, alpha = width.factor), 
-              color = "white") +
-    geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
-        geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
-    theme_minimal() + 
-    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
-    scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
-  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
-
-males.era1
-```
-
-**Fig 2: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1983 and 1993 by US state**
-
-```{r figure2, echo = F, fig.height=9, fig.width=10}
-# - The gap in 1983 ranges from `r round(sum2[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum2[1, ]$LE_wbgap_lcl, 1)`, `r round(sum2[1, ]$LE_wbgap_ucl, 1)`] in `r sum2[1, ]$state ` to `r round(sum2[dim(sum2)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum2[dim(sum2)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum2[dim(sum2)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum2[dim(sum2)[1], ]$state `.
-
-y.labels.2 <- levels(factor(cod.df.males$state.gdiff.order.2))
-y.num.breaks.2 <- length(levels(factor(cod.df.males$state.gdiff.order.2)))
-
-males.era2 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "era2"), aes(y = state.gdiff.num.2, x = new.start.2)) +
-    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
-                  ymax = state.gdiff.num.2 + 0.45, 
-                  xmax = new.finish.2, fill = Cause.of.death, alpha = width.factor), 
-              color = "white") +
-    geom_segment(data = bw_dat.males, 
-                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
-    geom_segment(data = bw_dat.males, 
-                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
-    theme_minimal() + 
-    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
-    scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2), minor_breaks = NULL) %>% 
-  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
-
-males.era2 
-```
-
-**Fig 3: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
-
-```{r figure3, echo = F, fig.height = 9, fig.width = 11}
-y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
-y.num.breaks.3 <- length(levels(factor(cod.df.males$state.gdiff.order.3)))
-
-males.era3 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "era3"), aes(y = state.gdiff.num.3, x = new.start.3)) +
-    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
-                  ymax = state.gdiff.num.3 + 0.45, 
-                  xmax = new.finish.3, fill = Cause.of.death, alpha = width.factor), 
-              color = "white") +
-    geom_segment(data = bw_dat.males, 
-                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
-    geom_segment(data = bw_dat.males, 
-                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
-    theme_minimal() + 
-    scale_x_continuous(breaks = seq(0, 16, by = 2), labels = seq(0, 16, by = 2)) +
-    scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>% 
-  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))
-
-males.era3
-```
-
-**All the above together as one figure**
-
-```{r, fig.align="center", fig.height=9, fig.width=15, echo=F, eval = T}
-males.era2.2 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "era2"), aes(y = state.gdiff.num.1, x = new.start.2)) +
-    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.1 - 0.45, 
-                  ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.2, fill = Cause.of.death, alpha = width.factor), 
-              color = "white") +
-    geom_segment(data = bw_dat.males, 
-                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
-    geom_segment(data = bw_dat.males, 
-                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>% 
-  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
-
-males.era3.2 <- ggplotly(
-  ggplot(subset(cod.df.males, era == "era3"), aes(y = state.gdiff.num.1, x = new.start.3)) +
-    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.1 - 0.45, 
-                  ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.3, fill = Cause.of.death, alpha = width.factor), 
-              color = "white") +
-    geom_segment(data = bw_dat.males, 
-                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
-    geom_segment(data = bw_dat.males, 
-                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>% 
-  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
-
-subplot(males.era1, males.era2.2, males.era3.2, shareY = T)
-```
-** **
-** **
-
-```{r, echo =F}
-sum5 <- BlackWhite_results %>% filter(year == 1969 & sex == "Female") %>% 
-  arrange(LE_wbgap_mean) %>% 
-  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)
-```
-
-*Changes to the difference in life expectancy in US females*
-
-For females, the life expectancy gap narrowed in `r dim(bw_dat.females[bw_dat.females$era1.diff > 0, ])[1]` of `r dim(bw_dat.females)[1]` states between 1969 and 1983.
-In `r length(cod.df$narrowed_gap[cod.df$sex == "Female" & cod.df$era == "1969-1983" & cod.df$COD == "Cardiovascular" & cod.df$narrowed_gap == T])` states, CVD made a substantial contribution towards narrowing the gap (for example, Virginia, Kentucky, South Carolina), while in others, its contribution was towards widening the gap (CI may be very important here -- the states with very large contributions also have very few blacks...).
-
-Quick summary
-
-- Most states saw decreasing gaps in first era.
-
-- However, increases were found in Arizona, Oregon, Minnesota, Washington, and Wisconsin, largely due to CVD.
-As you can see, in some states CVD contributed to excacerbating the gap while in most others it makes a large contribution to narrowing the gap.
-
-- Otherwise, most of the other causes contributed to the gap narrowing, although cancer was associated with narrowing in some states (MS, NM, Oregon, for example) and increases in others (Pennsylvania, Ohio, Missouri, Kentucky).
-
-- The biggest contributor to the narrowing gap being CVD (in most states, although heterogeniety) and all other causes.
-
-- During the second era, the gap stayed approx.
-the same or widened across most states.
-Delaware, Kentucky, Oregon, Rhode Island, West Virginia, Alabama, were notable exceptions were the gap narrowed -- didn't see effect of communicable disease like other states on the gap increasing (NY, FL, DC, MD, CT, hardest hit).
-
-- In women, changes in injury related deaths doesn't play a big role (except DC).
-
-- CVD differential by state. Contributes to widening the gap in CA, Mississippi, and DC, for example, and narrowing it in Kansas, Delaware, and Kentucky.
-
-- All other causes narrows across states (except DC, Indiana). Cancer increases the gap by a bit (except Rhode Island, Oregon, and Minnesota).
-
-- During the third era, in-line with males, the life expectancy gap decreased during the last 20 year period (except DC, Wisconsin), and generally speaking all of the causes contributed to the gap narrowing over time.
-The biggest contributors were CVD and all other causes.
-
-- In a few states, cancer contributed to widening the gap (NM, Oregon, Arizona, DC, Wisconsin).
-
-**Fig 4: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
-
-```{r figure2a, fig.align = "center", echo=F, fig.height = 9, fig.width = 7.5}
-
-cod.df.females <- merge(subset(cod.df, sex == "Female"), bw_dat.females, by = c("state"))
-cod.df.females$tot_Cx_CI_width2 <- -(cod.df.females$tot_Cx_CI_width - min(cod.df.females$tot_Cx_CI_width))/(max(cod.df.females$tot_Cx_CI_width) - min(cod.df.females$tot_Cx_CI_width)) + 1
-cod.df.females$width.factor <- as.factor(cut(cod.df.females$tot_Cx_CI_width2, breaks = c(-0.01, 0.25, 0.50, 0.75, 1)))
-
-cod.df.females <- cod.df.females %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
-                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
-                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)
-
-y.labels <- levels(factor(cod.df.females$state.gdiff.order.1))
-y.num.breaks <- length(levels(factor(cod.df.females$state.gdiff.order.1)))
-
-females.era1 <- ggplotly(
-  ggplot(subset(cod.df.females, era == "era1"), aes(y = state.gdiff.num.1, x = new.start.1)) +
-    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
-                  ymax = state.gdiff.num.1 + 0.45, 
-                  xmax = new.finish.1, fill = Cause.of.death, alpha = width.factor), 
-              color = "white") +
-    geom_segment(data = bw_dat.females,
-                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
-    geom_segment(data = bw_dat.females,
-                 aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
-  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))
-
-females.era1
-```
-
-**Fig 5: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1983 and 1993 by US state**
-
-```{r figure2b, fig.align = "center", echo=F, fig.height = 9, fig.width = 7.5}
-y.labels.2 <- levels(factor(cod.df.females$state.gdiff.order.2))
-y.num.breaks.2 <- length(levels(factor(cod.df.females$state.gdiff.order.2)))
-
-females.era2 <- ggplotly(
-  ggplot(subset(cod.df.females, era == "era2"), aes(y = state.gdiff.num.2, x = new.start.2)) +
-    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
-                  ymax = state.gdiff.num.2 + 0.45, 
-                  xmax = new.finish.2, fill = Cause.of.death, alpha = width.factor), 
-              color = "white") +
-    geom_segment(data = bw_dat.females,
-                 aes(x = third.gap, xend = third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
-    geom_segment(data = bw_dat.females,
-                 aes(x = second.gap, xend = second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>%
-  layout(xaxis = list(title = "Life expectancy gap (years): 1983-1993"), yaxis = list(title = NA, autorange = "reversed"))
-
-females.era2
-```
-
-**Fig 6: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
-
-```{r fig6, fig.align = "center", echo=F, fig.height = 9, fig.width = 9}
-y.labels.3 <- levels(factor(cod.df.females$state.gdiff.order.3))
-y.num.breaks.3 <- length(levels(factor(cod.df.females$state.gdiff.order.3)))
-
-females.era3 <- ggplotly(
-  ggplot(subset(cod.df.females, era == "era3"), aes(y = state.gdiff.num.3, x = new.start.3)) +
-    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
-                  ymax = state.gdiff.num.3 + 0.45, 
-                  xmax = new.finish.3, fill = Cause.of.death, alpha = width.factor), 
-              color = "white") +
-    geom_segment(data = bw_dat.females,
-                 aes(x = fourth.gap, xend = fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
-    geom_segment(data = bw_dat.females,
-                 aes(x = third.gap, xend = third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
-    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>%
-  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))
-
-females.era3
-```
-
-###Supporting Information
-
-**S1 Table: Cause of death groupings**
-
-| Broad cause of death     | Included causes of death                                                     |
-|--------------------------|------------------------------------------------------------------------------|
-| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer.
-|
-| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries.
-|
-| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
-| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Syndrome and Nephrosis.
-|
-| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
-| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death.
-|
-
-**S1 Text: Decomposing the change to the difference in life expectancy over time by cause of death: an example**
-
-To create.
-1. First show the data for NY in 1993 and 2013 for black and white males (smoothed deaths, population, for each age strata).
-
-2. Create life tables for blacks and whites in the two years (4 tables)
-
-3. Calculate the LE gap for each year (two numbers)
-
-4. Decompose by age the BW gap within year (two decomps)
-
-5. Decompose by age and COD the BW gap within year (two decomps)
-
-6. Marginal COD decomp within year (two decomps)
-
-7. Subtract the COD decomp from year 1 from year 2. (one change in decomp)
-
-**S1 Fig: Difference in life expectancy between black in white males, United States, 1969-2013**
-
-![Difference in life expectancy between black and white males, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Men.bmp)
-
-**S2 Fig: Difference in life expectancy between black in white females, United States, 1969-2013**
-
-![Difference in life expectancy between black and white females, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Women.bmp)
-
-**S3-S14 Fig: Trends in age-standardized mortality by state, race, and sex, 1969-2013 (currently included in other file)**
-
-**S15 Fig: Comparison of the contribution of age and cause of death to the difference in life expectancy between blacks and whites males in New York, 1969 vs. 1983**
-
-```{r ny_1969_1983_age_cod_decomp, echo = F, fig.width = 9}
-ny.dat <- age_cod_results %>% filter(state == "New York" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
-                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))
-
-ny.1969 <- make_pretty_decomp_plot(decomp.table = subset(ny.dat, year == 1969), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
-                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
-ny.1983 <- make_pretty_decomp_plot(decomp.table = subset(ny.dat, year == 1983), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
-                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
-ny.1993 <- make_pretty_decomp_plot(decomp.table = subset(ny.dat, year == 1993), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
-                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
-
-ggplotly(ggplot(rbind(ny.1969, ny.1983, ny.1993),
-                aes(y = age, x = start)) + 
-           geom_segment(aes(xend = finish, col = COD, yend = age), lwd = 3) +
-           xlab(paste0("Contribution to life expectancy gap in years")) +
-           ylab("Age group\n") + theme_minimal() +
-           geom_vline(xintercept = 0) + 
-           facet_wrap(~ year))
-
-```
-
-**Pennsylvania**
-```{r pa_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
-pa.dat <- age_cod_results %>% filter(state == "Pennsylvania" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
-                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))
-
-pa.1993 <- make_pretty_decomp_plot(decomp.table = subset(pa.dat, year == 1993), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
-                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
-pa.1983 <- make_pretty_decomp_plot(decomp.table = subset(pa.dat, year == 1983), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
-                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
-
-ggplotly(ggplot(rbind(pa.1993, pa.1983),
-                aes(y = age, x = start)) + 
-           geom_segment(aes(xend = finish, col = COD, yend = age), lwd = 3) +
-           xlab(paste0("Contribution to life expectancy gap in years")) +
-           ylab("Age group\n") + theme_minimal() +
-           geom_vline(xintercept = 0) + 
-           facet_wrap(~ year)) 
-```
-
-**Connecticut**
-```{r ct_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
-ct.dat <- age_cod_results %>% filter(state == "Connecticut" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
-                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))
-
-ct.1993 <- make_pretty_decomp_plot(decomp.table = subset(ct.dat, year == 1993), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
-                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
-ct.1983 <- make_pretty_decomp_plot(decomp.table = subset(ct.dat, year == 1983), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
-                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
-
-ggplotly(ggplot(rbind(ct.1993, ct.1983),
-                aes(y = age, x = start)) + 
-           geom_segment(aes(xend = finish, col = COD, yend = age), lwd = 3) +
-           xlab(paste0("Contribution to life expectancy gap in years")) +
-           ylab("Age group\n") + theme_minimal() +
-           geom_vline(xintercept = 0) + 
-           facet_wrap(~ year)) 
-```
-
-**Mississippi**
-```{r ms_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
-ms.dat <- age_cod_results %>% filter(state == "Mississippi" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
-                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))
-
-ms.1993 <- make_pretty_decomp_plot(decomp.table = subset(ms.dat, year == 1993), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
-                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
-ms.1983 <- make_pretty_decomp_plot(decomp.table = subset(ms.dat, year == 1983), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
-                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
-
-ggplotly(ggplot(rbind(ms.1993, ms.1983),
-                aes(y = age, x = start)) + 
-           geom_segment(aes(xend = finish, col = COD, yend = age), lwd = 3) +
-           xlab(paste0("Contribution to life expectancy gap in years")) +
-           ylab("Age group\n") + theme_minimal() +
-           geom_vline(xintercept = 0) + 
-           facet_wrap(~ year)) 
-```
-
-**Florida**
-```{r fl_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
-fl.dat <- age_cod_results %>% filter(state == "Florida" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
-                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))
-
-fl.1993 <- make_pretty_decomp_plot(decomp.table = subset(fl.dat, year == 1993), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
-                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
-fl.1983 <- make_pretty_decomp_plot(decomp.table = subset(fl.dat, year == 1983), sign.var = "sign.var",
-                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
-                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
-
-ggplotly(ggplot(rbind(fl.1993, fl.1983),
-                aes(y = age, x = start)) + 
-           geom_segment(aes(xend = finish, col = COD, yend = age), lwd = 3) +
-           xlab(paste0("Contribution to life expectancy gap in years")) +
-           ylab("Age group\n") + theme_minimal() +
-           geom_vline(xintercept = 0) + 
-           facet_wrap(~ year)) 
-```
-
-**S2 Table: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
-
-```{r, echo=F}
-cod_change_results2 <- data.frame(cod_change_results[c(3, 40, 41)], apply(cod_change_results[4:39], 2, function(x) x*(-1)))
-
-cod_change_results2 <- cod_change_results2 %>% 
-  mutate(total_era1_diff = paste0(round(tot_Cx_era1_diff_mean, 2), " [", round(tot_Cx_era1_diff_lcl, 2), ",", round(tot_Cx_era1_diff_ucl, 2), "]"),
-         total_era2_diff = paste0(round(tot_Cx_era2_diff_mean, 2), " [", round(tot_Cx_era2_diff_lcl, 2), ",", round(tot_Cx_era2_diff_ucl, 2), "]"),
-         total_era3_diff = paste0(round(tot_Cx_era1_diff_mean, 2), " [", round(tot_Cx_era3_diff_lcl, 2), ",", round(tot_Cx_era3_diff_ucl, 2), "]"),
-         contribution_era1_diff = paste0(round(Cx_era1_diff_mean, 2)," [", round(Cx_era1_diff_lcl, 2), ",", round(Cx_era1_diff_ucl, 2), "]"),
-         contribution_era2_diff = paste0(round(Cx_era2_diff_mean, 2)," [", round(Cx_era2_diff_lcl, 2), ",", round(Cx_era2_diff_ucl, 2), "]"),
-         contribution_era3_diff = paste0(round(Cx_era3_diff_mean, 2)," [", round(Cx_era3_diff_lcl, 2), ",", round(Cx_era3_diff_ucl, 2), "]")
-         ) %>%
-  select(sex, state, Cause.of.death, total_era1_diff, contribution_era1_diff, total_era2_diff, contribution_era2_diff, total_era3_diff, contribution_era3_diff) %>%
-  rename(Sex = sex, State = state, 
-         `Change from 1969 to 1983, years` = total_era1_diff,
-         `Change from 1983 to 1993, years` = total_era2_diff,
-         `Change from 1993 to 2013, years` = total_era3_diff,
-         `Contribution to gap change ('69-'83), years` = contribution_era1_diff,
-         `Contribution to gap change ('83-'93), years` = contribution_era2_diff,
-         `Contribution to gap change ('93-'03), years` = contribution_era3_diff)
-
-DT::datatable(cod_change_results2 %>% filter(Sex == "Male"))
-```
-
-**S3 Table: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
-
-```{r, echo=F}
-DT::datatable(cod_change_results2 %>% filter(Sex == "Female"))
-```
-
-**Things for corinne to review**
-
-- CVD declining starting 1968 - see paper Ruel Stallones - cited in SH JAMA paper - history of decline in CVD.
-
-- Richard cooper also written about BW diff in CVD with epi perspective
+###Bibliography

### 2017-03-30 shiny_app/app.R
hmm

@@ -402,7 +402,7 @@ map.contribution <- reactive({
   geom_ribbon(aes(ymin = y1_lcl, ymax = y1_ucl, fill = Census_Division)) +
   geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
   facet_wrap(~state.reorder, ncol = 11, drop = F) +
-  theme_classic(base_size = 15) +
+  theme_classic(base_size = 10) +
   theme(axis.text.x = element_blank(),
         strip.background=element_blank(),
         axis.line=element_blank(),

### 2017-03-30 Manuscript/Appendix-agestd-mort-rates.Rmd
add hoverinfo comments

@@
 ggplotly(Cardiovascular.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+
+#manipulating hover info:
+#p <- ggplotly(Cardiovascular.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+#p$x$data[[1]]$hoverinfo = "none" #if want no hover info for a particular trace
+#p$x$data[[1]]$text #otherwise, modified what is stored here to display how you want it to look. 
 ```

### 2017-03-30 Code/life_expectancy_functions.R
added state map function

@@ -559,3 +559,21 @@ add_Census_Region <- function(data, state) {
   return(data)
 }
 
+#This function adds the variable "state.map.order" to the dataframe to be used to plot the facets to approximate the shape of the continental US
+#DOES NOT INCLUDE ALASKA AND HAWAII!!
+reorder.as.map <- function(dataset, state.var) {
+  # Create unique blank strip labels for empty facets
+  bl = sapply(1:37, function(n) paste(rep(" ",n),collapse=""))
+  
+  dataset["state.map.order"] <- factor(dataset[[state.var]], 
+                                       levels = c(bl[1:10], "Maine",
+                                                  bl[11:19], "Vermont", "New Hampshire",
+                                                  "Washington", "Idaho", "Montana", "North Dakota", "Minnesota", "Illinois", "Wisconsin", "Michigan", "New York", "Massachusetts", "Rhode Island",
+                                                  "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", bl[20],
+                                                  "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Delaware", bl[21],
+                                                  bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Washington DC", bl[23:24],
+                                                  bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
+                                                  bl[30:33], "Texas", bl[34:37], "Florida")) 
+  
+  return(dataset)
+}

### 2017-03-30 Manuscript/Manuscript_COD_Trends.Rmd

@@
 
-**The data**
+**Description of the mortality data**
 We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.
@@
 
-**Statistical Methods**
+**Statistical challenges to be addressed**
 The data structure produced two statistical challenges that were addressed by our quantitative approach.
@@
 
+**Estimation of smoothed mortality rates using a Bayesian autoregressive model**
 To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts.
@@
 
+**Decomposition of the life expectancy gap by cause of death**
 To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref].
@@
 
+**Description of the trend in each cause of death's contribution to the gap and identification of divergent states**
 To describe the general pattern in these trend lines, we regressed each state's mean contribution against a linear spline for calendar time that was estimated with five degrees of freedom.
@@
 
-**Need to add information on the trends in mortality**
+*Need to add information on the trends in mortality* -- where?
 
 ###Results
+```{r initialize, echo=FALSE, warning=F, message=F}
+library(shiny)
+library(plotly)
+library(viridis)
+library(scales)
+library(shinythemes)
+library(dplyr)
+library(purrr)
+library(stringr)
+library(DT)
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+```
+
+```{r load_data, echo=F}
+BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
+BlackWhite_results <- reorder.as.map(BlackWhite_results, "state")
+```
+
+```{r life-expectancy-trends, echo=F}
+LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
+  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
+  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
+  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 10) +
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Life expectancy gap (years)") +
+  xlab("Year (1969-2013)")
+
+LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
+  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
+  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
+  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 10) +
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Life expectancy gap (years)") +
+  xlab("Year (1969-2013)")
+```
+
+```{r plot_both, eval=F}
+#Informative: male and females at the same time
+ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
+  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
+  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
+  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 10) +
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Life expectancy gap (years)") +
+  xlab("Year (1969-2013)"))
+```
+
+**Figures 1 and 2** display trends in the life expectancy gap for males and females, respectively.
+For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males.
+
+**Figures 2 - 14** display trends in the contribution of each cause of death to the life expectancy gap for males and females, in years.
+The green band illustrates the credible interval for the smoothed contribution estimated for each state, and illustrates the heterogeneity across states in their contributions trends.
+These curves are overlaid with the meta-estimated common trend, shown with a red line, and the plots are organized according to the closeness of the [fit] (*need another word*) of the red curve to the green band.
+Thus, increasingly divergent states are shown further down the display.
+We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).
+
+**Cardiovascular disease**
+
+Cardiovascular disease is the most common cause of death for both men and women of both races.
+Mortality from cardiovascular disease decreased markedly since 1969 for both sexes and races (**eFigures 1 and 2**).
+However, white males underwent relatively larger mortality decreases than black males, resulting in a modest increase to the contribution of cardiovascular disease to the life expectancy gap for most states (**Figure 2**).
+In particular, Pennsylvania, Michigan, Wisconsin, and Washington DC experienced relatively larger increases to their CVD contributions to the gap for males, while some states such as Illinois and Kentucky exhibited relatively larger early contributions in the 1970s.
+
+The situation was reversed for females, with black females experiencing relatively larger mortality decreases over time compared to white females.
+Thus, the contribution of CVD to the life expectancy gap decreased by more than a year in several states.
+Of note, some states experienced a period of early increasing contributions, most notably, Arizona, Washington, and Oregon, that was not captured by the general pattern.
+Furthermore, Wisconsin and Washington DC both exhibit increasing contributions over time, as mortality gains among white females outpaced those among black females.
+
+**Cancer**
+
+**Figure 1: State-level trends in the difference in life expectancy between blacks and whites in males, United States, 1969-2013**
+
+```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11}
+ggplotly(LE_trend_plot_males)
+```
+
+**Figure 2: State-level trends in the difference in life expectancy between blacks and whites in females, United States, 1969-2013**
+
+```{r plot_LE_trends_females, echo=F, fig.height=9, fig.width=11}
+ggplotly(LE_trend_plot_females)
+```
+
+###Discussion
+
+###Appendix
+
+```{r, echo = F}
+#cant figure out how to do this yet (embed other Rmd file inside this one)
+knitr::knit('/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/Appendix-agestd-mort-rates.Rmd', tangle = T)
+source('/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/Appendix-agestd-mort-rates.R')
+```
 

### 2017-03-30 Manuscript/Manuscript_COD_Trends.Rmd
more writing

--- 2017-03-29 Manuscript/Manuscript_COD_Trends.Rmd
@@
 ---
-title: 
+title: "null"
+bibliography: bibliography.bib
 output:
+  word_document: default
   html_document: default
-  word_document: default
 always_allow_html: yes
@@
 ###Introduction
+
 The life expectancy of a population is a barometer for overall health.
@@
 
-**The data**
+**Description of the mortality data**
+
 We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.
@@
 
-**Statistical Methods**
+**Statistical challenges to be addressed**
+
 The data structure produced two statistical challenges that were addressed by our quantitative approach.
@@
 To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.
+
+**Estimation of smoothed mortality rates using a Bayesian autoregressive model**
 
@@
 These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.
+
+**Decomposition of the life expectancy gap by cause of death**
 
@@
 
+**Description of the trend in each cause of death's contribution to the gap and identification of divergent states**
+
 To describe the general pattern in these trend lines, we regressed each state's mean contribution against a linear spline for calendar time that was estimated with five degrees of freedom.
@@
 
-**Need to add information on the trends in mortality**
+*Need to add information on the trends in mortality* -- where?
 
 ###Results
+```{r initialize, echo=FALSE, warning=F, message=F}
+library(plotly)
+library(viridis)
+library(scales)
+library(dplyr)
+library(purrr)
+library(stringr)
+library(DT)
+library(splines)
+library(broom)
+library(tidyverse)
+
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+```
+
+```{r load_data, echo=F}
+BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
+BlackWhite_results <- reorder.as.map(BlackWhite_results, "state")
+
+cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
+cod_decomp_results$COD_cont_yrs_width <- cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl
+cod_decomp_results$COD_cont_yrs_SE <- (cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl)/3.92
+```
+
+```{r life-expectancy-trends, echo=F}
+LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
+  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
+  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
+  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 10) +
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Life expectancy gap (years)") +
+  xlab("Year (1969-2013)")
+
+LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
+  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
+  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
+  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 10) +
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Life expectancy gap (years)") +
+  xlab("Year (1969-2013)")
+```
+
+```{r plot_both, eval=F, echo=F}
+#Informative: male and females at the same time
+ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
+  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
+  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
+  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 10) +
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Life expectancy gap (years)") +
+  xlab("Year (1969-2013)"))
+```
+
+```{r meta_regression_of_cod_contribution, echo = F}
+#run 12 stratified models. bs() created the b-spline basis
+models <- cod_decomp_results %>% 
+  group_by(COD, sex) %>%
+  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state, 
+              weights = 1/(COD_cont_yrs_SE)^2, 
+              data = .))
+
+#calculated predicted values for all models
+dfPred = broom::augment(models, fit)
+dfPred$year <- 1969:2013
+
+#merge the predictions with original dataset
+cod_decomp_results <- merge(cod_decomp_results, 
+                            dfPred %>% select(COD, sex, year, state, .fitted, .resid), 
+                            by = c("COD", "sex", "state", "year"))
+
+#calculated the sum of the residuals for each state-sex-cod strata (summed over years)
+cod_decomp_results <- cod_decomp_results %>% 
+  group_by(state, sex, COD) %>%
+  mutate(sum.residual = sum(abs(.resid)))
+```
+
+```{r plotting_resid_fx, echo = F}
+plot.resid.trends <- function(dataset, COD1, sex1) {
+  subset.data <- subset(dataset, COD == COD1 & sex == sex1)
+  subset.data$order <- reorder(subset.data$state, subset.data$sum.residual, first)
+
+  a <- ggplot(subset.data,
+         aes(x = year, y = COD_cont_yrs_mean)) + 
+    #geom_line(aes(col = "Observed")) + 
+    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9") +
+    geom_hline(yintercept = 0, col = "black") +
+    #scale_fill_manual(values = c("#b2e2e2")) +
+    facet_wrap( ~ order) + 
+    geom_line(aes(y = .fitted), col = "#d7191c") +
+    #scale_color_manual(values = c("#006d2c")) +
+    xlab(" ") + ylab("estimated smoothed contribution vs. common trend (years)") + 
+    ggtitle(paste0("Estimated smoothed contribution vs. meta-estimated common trend in ", COD1, " in ", sex1)) +
+    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+  b <- ggplot(subset.data,
+              aes(x = year, y = .resid)) + geom_line(aes(col = "Residual")) + 
+    scale_color_manual(values = c("#d7191c")) +    
+    facet_wrap( ~ order) +
+    xlab(" ") + ylab("Residual") +
+    geom_hline(yintercept = 0, col = "black") +
+      geom_text(data = subset(subset.data, year == 2010), 
+                aes(y = 1, label = round(sum.residual, 1)), size = 2.5) + 
+    ggtitle(paste0("Residuals vs. year in ", COD1, " in ", sex1)) +
+    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+    return(list(plot1 = a, plot2 = b))
+}
+```
+
+```{r create-the-resid-plots, echo = F, message = F}
+for(COD_i in levels(cod_decomp_results$COD)) {
+  for(sex_i in levels(cod_decomp_results$sex)) {
+    assign(paste0(COD_i, ".", sex_i, ".plot"), plot.resid.trends(dataset = cod_decomp_results, 
+                                                                         COD1 = COD_i, sex1 = sex_i))
+  }
+}
+```
+
+**Figures 1 and 2** display trends in the life expectancy gap for males and females, respectively.
+For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males.
+
+**Figures 2 - 14** display trends in the contribution of each cause of death to the life expectancy gap for males and females, in years.
+The blue band illustrates the credible interval for the smoothed contribution estimated for each state, and illustrates the heterogeneity across states in their contributions trends.
+These curves are overlaid with the meta-estimated common trend, shown with a red line, and the plots are organized according to the closeness of the [fit] (*need another word*) of the red curve to the blue band.
+Thus, increasingly divergent states are shown further down the display.
+We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).
+
+**Cardiovascular disease**
+
+Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races.
+Mortality from CVD decreased markedly since 1969 for both sexes and races (**eFigures 1 and 2**).
+However, white males underwent relatively larger mortality decreases than black males, resulting in a modest increase to the contribution of CVD to the life expectancy gap for most states (**Figure 3**).
+In particular, Pennsylvania, Michigan, Wisconsin, and Washington DC experienced relatively larger increases to their CVD contributions to the gap for males, while some states such as Illinois and Kentucky exhibited relatively larger early contributions in the 1970s.
+
+The situation was reversed for females, with black females experiencing relatively larger mortality decreases over time compared to white females.
+Thus, the contribution of CVD to the life expectancy gap decreased by more than a year in several states (**Figure 4**).
+Of note, some states experienced a period of early increasing contributions, most notably, Arizona, Washington, and Oregon, that was not captured by the general pattern.
+Furthermore, Wisconsin and Washington DC both exhibit increasing contributions over time, as mortality gains among white females outpaced those among black females.
+
+**Cancer**
+
+For all races and sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**).
+Among males, this inverted U-shape pattern was accentuated among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**).
+Notable exceptions to this general pattern include Washington DC and Mississippi, states that underwent increasing cancer contributions over time, and New Mexico and Rhode Island, exhibiting decreasing contributions over time (although imprecisely estimated).
+
+For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern is similar between blacks and whites and most states do not exhibit the accentuated period of increase in mortality that was estimated for black males.
+Thus, the contribution of cancer to the life expectancy gap is generally flat over time (**Figure 6**).
+Washington DC deviates from this pattern, and exhibited increasing cancer contributions to the gap, while Rhode Island exhibited the reverse.
+Most notably, some states exhibited no inequality in cancer mortality (**eFigures 3 and 4**), including New York, and Connecticut (and for males, these states recently eliminated the black-white difference in cancer mortality).
+Furthermore, Massachusetts and (imprecisely) Rhode Island exhibit lower cancer rates among black women and men compared to whites in 2013.
+
+****
+
+**Figure 1: State-level trends in the difference in life expectancy between blacks and whites in males, United States, 1969-2013**
+
+```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11}
+ggplotly(LE_trend_plot_males)
+```
+
+**Figure 2: State-level trends in the difference in life expectancy between blacks and whites in females, United States, 1969-2013**
+
+```{r plot_LE_trends_females, echo=F, fig.height=9, fig.width=11}
+ggplotly(LE_trend_plot_females)
+```
+
+**Figure 3: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in males**
+
+```{r print-plot, fig.width = 9, fig.height = 9, echo = F}
+ggplotly(Cardiovascular.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Cardiovascular.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+```
+
+**Figure 4: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in females**
+
+```{r print-plot-2, fig.width = 9, fig.height = 9, echo = F}
+ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Cardiovascular.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+```
+
+**Figure 5: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in males**
+
+```{r print-plot-3, fig.width = 9, fig.height = 9, echo = F}
+ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Cancers.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+```
+
+**Figure 6: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in females**
+
+```{r print-plot-4, fig.width = 9, fig.height = 9, echo = F}
+ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Cancers.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+```
+
+```{r print-plot-5, fig.width = 9, fig.height=9, echo = F, eval = F}
+ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Communicable.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+
+ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Communicable.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+
+ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`Non-communicable.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
+
+ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`Non-communicable.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
+
+ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Injuries.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+
+ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Injuries.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+
+ggplotly(`All other causes.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`All other causes.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
+
+ggplotly(`All other causes.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`All other causes.Female.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
+```
+
+###Discussion
+
+###Appendix
 

### 2017-03-30 Manuscript/Appendix-agestd-mort-rates.Rmd

@@
 library(forcats)
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
 
@@
 mortality.rates.diff <- mortality.rates.diff %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
-```
 
-```{r refactor-for-us-map-plot, echo=F}
-
-# Create unique blank strip labels for empty facets
-bl = sapply(1:37, function(n) paste(rep(" ",n),collapse=""))
-
-mortality.rates$state.reorder2 <- factor(mortality.rates$state,
-                                        levels = c(bl[1:10], "Maine",
-                                                   bl[11:19], "Vermont", "New Hampshire",
-                                                   "Washington", "Idaho", "Montana", "North Dakota", "Minnesota", "Illinois", "Wisconsin", "Michigan", "New York", "Massachusetts", "Rhode Island",
-                                                   "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", bl[20],
-                                                   "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Delaware", bl[21],
-                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Washington DC", bl[23:24],
-                                                   bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
-                                                   bl[30:33], "Texas", bl[34:37], "Florida"))
-
-mortality.rates.diff$state.reorder2 <- factor(mortality.rates.diff$state,
-                                        levels = c(bl[1:10], "Maine",
-                                                   bl[11:19], "Vermont", "New Hampshire",
-                                                   "Washington", "Idaho", "Montana", "North Dakota", "Minnesota", "Illinois", "Wisconsin", "Michigan", "New York", "Massachusetts", "Rhode Island",
-                                                   "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", bl[20],
-                                                   "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Delaware", bl[21],
-                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Washington DC", bl[23:24],
-                                                   bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
-                                                   bl[30:33], "Texas", bl[34:37], "Florida"))
+mortality.rates <- reorder.as.map(mortality.rates, "state")
+mortality.rates.diff <- reorder.as.map(mortality.rates.diff, "state")
 ```
@@
               geom_line(aes(col = Race)) +
-              facet_wrap( ~ state.reorder2, ncol = 11, drop = F) +
+              facet_wrap( ~ state.map.order, ncol = 11, drop = F) +
               xlab(" ") + 
@@
               geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL), col = NA, alpha = 0.5) +
-              facet_wrap( ~ state.reorder2, ncol = 11, drop = F) +
+              facet_wrap( ~ state.map.order, ncol = 11, drop = F) +
               xlab(" ") + 
@@
 
+**eFigure 1**
+
 ```{r print-em, fig.width = 9, fig.height = 9, echo = F}
@@
 
+**eFigure 2**
+
 ```{r print-em-2, fig.width = 9, fig.height = 9, echo = F}
@@
 
+**eFigure 3**
+
 ```{r print-em-3, fig.width = 9, fig.height = 9, echo = F}
@@
 ```
+
+**eFigure 4**
 

### 2017-03-31 Manuscript/Manuscript_COD_Trends.Rmd
Update Manuscript_COD_Trends.Rmd
Hey Corinne-
I have not made a full pass through this yet, but tried to just give a couple of comments on the Intro and Results. Generally speaking, I think this the right way to go in terms of describing the results--report on the general pattern among all states but identify some key outliers of interest, particularly if they are regional in nature, which could be linked to other changes in risk factors, etc.

@@
 
-The over-arching objective of this study was to describe the general pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and differences between the sexes.
+The over-arching objective of this study was to describe the pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences.
 To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for males and females.
-Common reasons for the decreasing life expectancy gap can thereby be identified, and anomalous states can be examined.
+This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states.
 Ideally, this study may facilitate the identification of one policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.
@@
 For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males.
+Don't know whether you want to note it here or not, but I find it is interesting to see that the gap is now lower in many US southern (or southeastern) states than in the Midwest, suggesting that progress has actually been somewhat greater in the South than elsewhere (particularly the upper Midwest and especially for women).
+I think this is interesting and somewhat counter to the narrative that health indicators are worse in the South.
+I also wonder how this might relate to the Murray/IHME work on US counties with "declining" life expectancya among white women.
 
 **Figures 2 - 14** display trends in the contribution of each cause of death to the life expectancy gap for males and females, in years.
+[I wonder whether some readers might be confused by the changes in the ordering of the plots from the geographic layout in F1/2 to the rest of the plots].
 The blue band illustrates the credible interval for the smoothed contribution estimated for each state, and illustrates the heterogeneity across states in their contributions trends.
@@
 Thus, the contribution of cancer to the life expectancy gap is generally flat over time (**Figure 6**).
+[You could potentially stop here to save space and just include the comment about there being little inequality in cancer mortality ].
 Washington DC deviates from this pattern, and exhibited increasing cancer contributions to the gap, while Rhode Island exhibited the reverse.

### 2017-03-31 Manuscript/Manuscript_COD_Trends.Rmd
manu changes

--- 2017-03-30 Manuscript/Manuscript_COD_Trends.Rmd
@@
 ---
-title: "null"
+title: 'null'
 bibliography: bibliography.bib
 output:
+  html_document: default
   word_document: default
-  html_document: default
 always_allow_html: yes
@@
   group_by(COD, sex) %>%
-  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state, 
+  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5), # + state, #remove the state FE
               weights = 1/(COD_cont_yrs_SE)^2, 
@@
 
+**Non-communicable disease**
+
+Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**Figure X**).
+Recently, there has been a slight decrease in mortality rates, especially among black males.
+These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states.
+As such, the contribution of non-communicable disease to the mortality gap was general flat among men with a recent decrease, and exhibited a gradual decrease over time among women.
+
+**Injury**
+
+Injury is one of the least common causes of death.
+Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime in the 1990s resulted in heightened mortality from injuries among black males at that time (**Figure X**).
+For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**Figure X**).
+Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, especially among females.
+Thus, the contribution of injuries to the mortality gap has generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures X and X**).
+Among males, Wisconsin and Washington DC diverged the most from the general pattern, as both states experienced steep increases in the contribution of injuries during the 1990s that was not captured by the general trend.
+Washington DC also exhibited this pattern among females.
+
+**Communicable disease**
+
+Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communiable disease at the height of the HIV/AIDS epidemic before anti-retroviral therapy was available.
+This spike was greater among males, especially blacks, and within states with larger urban centers.
+Thus, the contribution of communicable disease to the life expectancy gap exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics.
+
 ****
@@
 
-```{r print-plot-5, fig.width = 9, fig.height=9, echo = F, eval = F}
+**Figure 7: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the overall pattern in males**
+
+```{r print-plot-5, fig.width = 9, fig.height = 9, echo = F}
+ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`Non-communicable.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
+```
+
+**Figure 8: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the overall pattern in females**
+
+```{r print-plot-6, fig.width = 9, fig.height = 9, echo = F}
+ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`Non-communicable.Female.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
+```
+
+**Figure 9: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the overall pattern in males**
+
+```{r print-plot-7, fig.width = 9, fig.height = 9, echo = F}
+ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Injuries.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+```
+
+**Figure 10: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the overall pattern in females**
+
+```{r print-plot-8, fig.width = 9, fig.height = 9, echo = F}
+ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Injuries.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+```
+
+**Figure 11: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the overall pattern in males**
+
+```{r print-plot-9, fig.width = 9, fig.height=9, echo = F}
 ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
 ggplotly(Communicable.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+```
 
+**Figure 12: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the overall pattern in females**
+
+```{r print-plot-10, fig.width = 9, fig.height=9, echo = F}
 ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
 ggplotly(Communicable.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
-
-ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`Non-communicable.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
-
-ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`Non-communicable.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
-
-ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Injuries.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
-
-ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Injuries.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
-
-ggplotly(`All other causes.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`All other causes.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
-
-ggplotly(`All other causes.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`All other causes.Female.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
 ```
 
+**Figure 13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males**
+**Figure 14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males**
+
 ###Discussion
+
+Limitations
+-does not account for migration. Washington DC - have a lot of migration of high SES white folks coming in, and perhaps also relatively high SES black folks moving out
+-does not account for hispanic ethinicity - demographic shifts in FL for example of black and white hispanics which will pull LE and disparities
 

### 2017-04-03 Manuscript/Manuscript_COD_Trends.Rmd

@@
 
-To describe the general pattern in these trend lines, we regressed each state's mean contribution against a linear spline for calendar time that was estimated with five degrees of freedom.
-This model also contained fixed effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year.
-We then calculated residuals for each year by substracting the meta-estimated contribution from the smoothed-observed estimated contribution and summed these residuals over time for each state to rank states according to how well the model predicted their pattern.
-Plots of these residuals over time were used to identify divergent states.
+To describe the general pattern in these trend lines we used random effects meta regression.
+Each state's mean contribution was regressed against a linear spline for calendar time that was estimated with five degrees of freedom.
+The model included random effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year.
+To identify divergent states, we compared the state-specific trends (estimated separately for each state using the Bayesian model) to the general pattern estimated using meta regression.
 
@@
 cod_decomp_results$COD_cont_yrs_SE <- (cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl)/3.92
+
+cod_decomp_results <- reorder.as.map(cod_decomp_results, "state")
 ```
@@
 ```{r meta_regression_of_cod_contribution, echo = F}
+#model option 1: FE analysis
+
 #run 12 stratified models. bs() created the b-spline basis
@@
   group_by(COD, sex) %>%
-  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5), # + state, #remove the state FE
+  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state,
               weights = 1/(COD_cont_yrs_SE)^2, 
@@
 #calculated predicted values for all models
-dfPred = broom::augment(models, fit)
+dfPred <- broom::augment(models, fit)
 dfPred$year <- 1969:2013
@@
                             dfPred %>% select(COD, sex, year, state, .fitted, .resid), 
-                            by = c("COD", "sex", "state", "year"))
+                            by = c("COD", "sex", "state", "year")) %>%
+                      rename(fitted.FE = .fitted, resid.FE = .resid)
 
@@
   group_by(state, sex, COD) %>%
-  mutate(sum.residual = sum(abs(.resid)))
+  mutate(sum.residual = sum(abs(resid.FE)))
+```
+
+```{r meta_RE_analysis, echo = F}
+library(lme4)
+
+#run 12 stratified models. bs() created the b-spline basis
+RE.models <- cod_decomp_results %>% 
+  group_by(COD, sex) %>%
+  do(fit = lmer(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + (1|state),
+              weights = 1/(COD_cont_yrs_SE)^2, 
+              data = .))
+
+#calculated predicted values for all models
+RE.pred <- broom::augment(RE.models, fit)
+RE.pred$year <- 1969:2013
+
+RE.models$RE.estimates <- purrr::map(RE.models$fit, ~ ranef(.x))
+
+sex <- rep(c(rep("Female", 40), rep("Male", 40)), 6)
+COD <- c(rep("All other causes", 80), rep("Cancers", 80), rep("Cardiovascular", 80),
+         rep("Communicable", 80), rep("Injuries", 80), rep("Non-communicable", 80))
+RE.estimates <- c(RE.models$RE.estimates[[1]]$state$`(Intercept)`, RE.models$RE.estimates[[2]]$state$`(Intercept)`, 
+                  RE.models$RE.estimates[[3]]$state$`(Intercept)`, RE.models$RE.estimates[[4]]$state$`(Intercept)`, 
+                  RE.models$RE.estimates[[5]]$state$`(Intercept)`, RE.models$RE.estimates[[6]]$state$`(Intercept)`, 
+                  RE.models$RE.estimates[[7]]$state$`(Intercept)`, RE.models$RE.estimates[[8]]$state$`(Intercept)`, 
+                  RE.models$RE.estimates[[9]]$state$`(Intercept)`, RE.models$RE.estimates[[10]]$state$`(Intercept)`,
+                  RE.models$RE.estimates[[11]]$state$`(Intercept)`, RE.models$RE.estimates[[12]]$state$`(Intercept)`)
+state <- rep(levels(cod_decomp_results$state), 12)
+REs <- data.frame(sex, COD, RE.estimates, state)
+
+cod_decomp_results <- merge(cod_decomp_results,
+                            REs, by = c("COD", "sex", "state"),
+                            all.x = T)
+
+#merge the predictions with original dataset
+cod_decomp_results <- merge(cod_decomp_results, 
+                             RE.pred %>% select(COD, sex, state, year, 
+                                                bs.year..degree...1..df...5.,
+                                                .fitted, .resid, .fixed, .mu), 
+                             by = c("COD", "sex", "state", "year"))
+
+cod_decomp_results <- cod_decomp_results %>% rename(fitted.RE = .fitted, fixed.from.RE = .fixed)
+
+# test1 <- lmer(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + (1|state), weights = 1/(COD_cont_yrs_SE)^2, data = subset(cod_decomp_results, COD == "Cardiovascular" & sex == "Male"))
+# 
+# pred.general.trend <- predict(test1, re.form = NA)
+# pred.with.REs <- predict(test1)
+# 
+# plot(y=pred.general.trend[1:45], x=1969:2013)
 ```
@@
          aes(x = year, y = COD_cont_yrs_mean)) + 
+    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
     #geom_line(aes(col = "Observed")) + 
     geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9") +
-    geom_hline(yintercept = 0, col = "black") +
+    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
     #scale_fill_manual(values = c("#b2e2e2")) +
-    facet_wrap( ~ order) + 
-    geom_line(aes(y = .fitted), col = "#d7191c") +
-    #scale_color_manual(values = c("#006d2c")) +
-    xlab(" ") + ylab("estimated smoothed contribution vs. common trend (years)") + 
+    facet_wrap( ~ state.map.order, ncol = 11, drop = F) + 
+    geom_line(aes(y = fitted.FE), col = "#d7191c") +
     ggtitle(paste0("Estimated smoothed contribution vs. meta-estimated common trend in ", COD1, " in ", sex1)) +
-    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
+    theme_classic(base_size = 10) + 
+    theme(axis.text.x = element_text(angle = 45, hjust = 1),
+          strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+    ylab("estimated smoothed contribution vs. common trend (years)") +
+    xlab("Year (1969-2013)")
 
   b <- ggplot(subset.data,
-              aes(x = year, y = .resid)) + geom_line(aes(col = "Residual")) + 
+              aes(x = year, y = resid.FE)) + geom_line(aes(col = "Residual")) + 
     scale_color_manual(values = c("#d7191c")) +    
@@
 
-    return(list(plot1 = a, plot2 = b))
+  c <- ggplot(subset.data,
+              aes(x = year, y = COD_cont_yrs_mean)) + 
+    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
+
+    #geom_line(aes(col = "Observed")) + 
+    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9") +
+    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
+    #scale_fill_manual(values = c("#b2e2e2")) +
+    facet_wrap( ~ state.map.order, ncol = 11, drop = F) + 
+    geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
+    #geom_line(aes(y = fitted.RE), col = "#d7191c", lty = 3) +
+    ggtitle(paste0("Estimated smoothed contribution vs. meta-estimated common trend in ", COD1, " in ", sex1)) +
+    theme_classic(base_size = 10) + 
+    theme(axis.text.x = element_blank(),
+          strip.background=element_blank(),
+          axis.line=element_blank(),
+          axis.ticks=element_blank()) +
+    ylab("estimated smoothed contribution vs. common trend (years)") +
+    xlab("Year (1969-2013)")
+
+  return(list(plot1 = a, plot2 = b, plot3 = c))
 }
@@
 }
+```
+
+```{r refine_plots, eval = F, echo = F}
+#need tp add the "grid lines post-hoc" because they depend on the max height of the plot
+
+    geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey") +
+    geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey") +
+    geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey") +
+    geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey") +
+
+#turn hover off for grid lines and axis lines
+refine.patterns <- function(plot) {
+  for(i in 1:320){
+  if(plot$x$data[[i]]$line$color == "rgba(190,190,190,1)"){
+    plot$x$data[[i]]$line$width <- 0.7559055
+    plot$x$data[[i]]$hoverinfo <- "none"
+  }
+    if(plot$x$data[[i]]$line$color == "rgba(0, 0, 0, 1)"){
+    plot$x$data[[i]]$hoverinfo <- "none"
+  }
+}
+}
+
 ```
@@
 ```{r print-plot, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cardiovascular.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cardiovascular.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Cardiovascular.Male.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
 ```
@@
 ```{r print-plot-2, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cardiovascular.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+#ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Cardiovascular.Female.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
 ```
@@
 ```{r print-plot-3, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cancers.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+#ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Cancers.Male.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
 ```
@@
 ```{r print-plot-4, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cancers.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+#ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Cancers.Female.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
 ```
@@
 ```{r print-plot-5, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`Non-communicable.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
+#ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`Non-communicable.Male.plot`[[3]]) %>% layout(xaxis = list(title = "Year"))
 ```
@@
 ```{r print-plot-6, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`Non-communicable.Female.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
+#ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`Non-communicable.Female.plot`[[3]]) %>% layout(xaxis = list(title = "Year"))
 ```
@@
 ```{r print-plot-7, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Injuries.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+#ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Injuries.Male.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
 ```
@@
 ```{r print-plot-8, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Injuries.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+#ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Injuries.Female.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
 ```
@@
 ```{r print-plot-9, fig.width = 9, fig.height=9, echo = F}
-ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Communicable.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+#ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Communicable.Male.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
 ```
@@
 ```{r print-plot-10, fig.width = 9, fig.height=9, echo = F}
-ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Communicable.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
+#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Communicable.Female.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
 ```
@@
 **Figure 13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males**
+
+```{r print-plot-11, fig.width = 9, fig.height=9, echo = F}
+#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`All other causes.Male.plot`[[3]]) %>% layout(xaxis = list(title = "Year"))
+```
+
 **Figure 14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males**
+
+```{r print-plot-12, fig.width = 9, fig.height=9, echo = F}
+#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`All other causes.Female.plot`[[3]]) %>% layout(xaxis = list(title = "Year"))
+```
 

### 2017-04-04 Manuscript/Manuscript_COD_Trends.Rmd
chchchanges

@@
 ---
-title: 'null'
+title: 'Historical trends in the contribution of six major causes of death to the black-white life expectancy gap by US state'
 bibliography: bibliography.bib
@@
               aes(x = year, y = COD_cont_yrs_mean)) + 
+    # geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey") +
+    # geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey") +
+    # geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey") +
+    # geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey") +
     geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
@@
     #geom_line(aes(y = fitted.RE), col = "#d7191c", lty = 3) +
-    ggtitle(paste0("Estimated smoothed contribution vs. meta-estimated common trend in ", COD1, " in ", sex1)) +
+    ggtitle(paste0("Estimated smoothed contribution of", COD1, " disease to the black-white life expectancy gap (blue 95% credible interval) vs.
+the meta-estimated general pattern (red line) in ", sex1, " 1969-2013")) +
     theme_classic(base_size = 10) + 
@@
           axis.ticks=element_blank()) +
-    ylab("estimated smoothed contribution vs. common trend (years)") +
+    ylab("Contribution to the life expectancy gap (years)") +
     xlab("Year (1969-2013)")
@@
 
-    geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey") +
-    geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey") +
-    geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey") +
-    geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey") +
-
 #turn hover off for grid lines and axis lines
@@
   }
+  return(plot)
 }
 
+#refine.patterns(Cardiovascular.Male.plot[[3]])
 ```
@@
 For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males.
-Don't know whether you want to note it here or not, but I find it is interesting to see that the gap is now lower in many US southern (or southeastern) states than in the Midwest, suggesting that progress has actually been somewhat greater in the South than elsewhere (particularly the upper Midwest and especially for women).
+*Don't know whether you want to note it here or not, but I find it is interesting to see that the gap is now lower in many US southern (or southeastern) states than in the Midwest, suggesting that progress has actually been somewhat greater in the South than elsewhere (particularly the upper Midwest and especially for women).
 I think this is interesting and somewhat counter to the narrative that health indicators are worse in the South.
-I also wonder how this might relate to the Murray/IHME work on US counties with "declining" life expectancya among white women.
+I also wonder how this might relate to the Murray/IHME work on US counties with "declining" life expectancy among white women.*
 
 **Figures 2 - 14** display trends in the contribution of each cause of death to the life expectancy gap for males and females, in years.
-[I wonder whether some readers might be confused by the changes in the ordering of the plots from the geographic layout in F1/2 to the rest of the plots].
-The blue band illustrates the credible interval for the smoothed contribution estimated for each state, and illustrates the heterogeneity across states in their contributions trends.
-These curves are overlaid with the meta-estimated common trend, shown with a red line, and the plots are organized according to the closeness of the [fit] (*need another word*) of the red curve to the blue band.
-Thus, increasingly divergent states are shown further down the display.
+The blue band illustrates the credible interval for the smoothed contribution estimated for each state, and illustrates the heterogeneity across states in their contribution trends.
+These curves are overlaid with the meta-estimated common trend, shown with a red line.
 We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).
@@
 ```{r print-plot, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cardiovascular.Male.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Cardiovascular.Male.plot[[3]]) 
 ```
@@
 #ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cardiovascular.Female.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Cardiovascular.Female.plot[[3]]) 
 ```
@@
 #ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cancers.Male.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Cancers.Male.plot[[3]]) 
 ```
@@
 #ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cancers.Female.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Cancers.Female.plot[[3]]) 
 ```
@@
 #ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`Non-communicable.Male.plot`[[3]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`Non-communicable.Male.plot`[[3]]) 
 ```
@@
 #ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`Non-communicable.Female.plot`[[3]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`Non-communicable.Female.plot`[[3]]) 
 ```
@@
 #ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Injuries.Male.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Injuries.Male.plot[[3]])
 ```
@@
 #ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Injuries.Female.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Injuries.Female.plot[[3]]) 
 ```
@@
 #ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Communicable.Male.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Communicable.Male.plot[[3]]) 
 ```
@@
 #ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Communicable.Female.plot[[3]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(Communicable.Female.plot[[3]]) 
 ```
@@
 #ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`All other causes.Male.plot`[[3]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`All other causes.Male.plot`[[3]])
 ```
@@
 #ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`All other causes.Female.plot`[[3]]) %>% layout(xaxis = list(title = "Year"))
+ggplotly(`All other causes.Female.plot`[[3]])
 ```

### 2017-04-05 Manuscript/Manuscript_COD_Trends.Rmd
formatting

@@
 BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
-BlackWhite_results <- reorder.as.map(BlackWhite_results, "state")
+BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")
 
@@
 
-cod_decomp_results <- reorder.as.map(cod_decomp_results, "state")
+cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
 ```
@@
   #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
-  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
+  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
   theme_classic(base_size = 10) +
@@
   geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
-  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
+  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
   theme_classic(base_size = 10) +
@@
     #scale_fill_manual(values = c("#b2e2e2")) +
-    facet_wrap( ~ state.map.order, ncol = 11, drop = F) + 
+    facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
     geom_line(aes(y = fitted.FE), col = "#d7191c") +
@@
               aes(x = year, y = COD_cont_yrs_mean)) + 
-    # geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey") +
-    # geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey") +
-    # geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey") +
-    # geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey") +
+    geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey") +
+    geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey") +
+    geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey") +
+    geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey") +
     geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
-
-    #geom_line(aes(col = "Observed")) + 
-    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9") +
+    geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
+    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
+    geom_line(col = "#08519c") + 
     #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
     #scale_fill_manual(values = c("#b2e2e2")) +
-    facet_wrap( ~ state.map.order, ncol = 11, drop = F) + 
-    geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
+    facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
     #geom_line(aes(y = fitted.RE), col = "#d7191c", lty = 3) +
-    ggtitle(paste0("Estimated smoothed contribution of", COD1, " disease to the black-white life expectancy gap (blue 95% credible interval) vs.
+    ggtitle(paste0("Estimated smoothed contribution of ", COD1, " disease to the black-white life expectancy gap (blue 95% credible interval) vs.
 the meta-estimated general pattern (red line) in ", sex1, " 1969-2013")) +

### 2017-04-05 Code/life_expectancy_functions.R

@@ -561,7 +561,7 @@ add_Census_Region <- function(data, state) {
 
 #This function adds the variable "state.map.order" to the dataframe to be used to plot the facets to approximate the shape of the continental US
 #DOES NOT INCLUDE ALASKA AND HAWAII!!
-reorder.as.map <- function(dataset, state.var) {
+reorder.as.map <- function(dataset, state.var, state.abbrev.var) {
   # Create unique blank strip labels for empty facets
   bl = sapply(1:37, function(n) paste(rep(" ",n),collapse=""))
   
@@ -574,6 +574,16 @@ reorder.as.map <- function(dataset, state.var) {
                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Washington DC", bl[23:24],
                                                   bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
                                                   bl[30:33], "Texas", bl[34:37], "Florida")) 
+  
+  dataset["stabbrs.map.order"] <- factor(dataset[[state.abbrev.var]], 
+                                       levels = c(bl[1:10], "ME",
+                                                  bl[11:19], "VT", "NH",
+                                                  "WA", "ID", "MT", "ND", "MN", "IL", "WI", "MI", "NY", "MA", "RI",
+                                                  "OR", "NV", "WY", "SD", "IA", "IN", "OH", "PA", "NJ", "CT", bl[20],
+                                                  "CA", "UT", "CO", "NE", "MO", "KY", "WV", "VA", "MD", "DE", bl[21],
+                                                  bl[22], "AZ", "NM", "KS", "AR", "TN", "NC", "SC", "DC", bl[23:24],
+                                                  bl[25:27], "OK", "LA", "MS", "AL", "GA", bl[28:29],
+                                                  bl[30:33], "TX", bl[34:37], "FL")) 
   
   return(dataset)
 }

### 2017-04-06 Manuscript/Appendix-agestd-mort-rates.Rmd
use state abbrevs

@@
 
-mortality.rates <- reorder.as.map(mortality.rates, "state")
-mortality.rates.diff <- reorder.as.map(mortality.rates.diff, "state")
+#DC state abbreviation wasn't accurately defined. Adding it here.
+levels(mortality.rates$stabbrs) <- c(levels(mortality.rates$stabbrs), "DC")
+levels(mortality.rates.diff$stabbrs) <- c(levels(mortality.rates.diff$stabbrs), "DC")
+mortality.rates$stabbrs[mortality.rates$state == "Washington DC"] <- "DC"
+mortality.rates.diff$stabbrs[mortality.rates.diff$state == "Washington DC"] <- "DC"
+
+mortality.rates <- reorder.as.map(mortality.rates, "state", "stabbrs")
+mortality.rates.diff <- reorder.as.map(mortality.rates.diff, "state", "stabbrs")
 ```
@@
               geom_line(aes(col = Race)) +
-              facet_wrap( ~ state.map.order, ncol = 11, drop = F) +
+              facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
               xlab(" ") + 
               ylab("Age-standardized mortality rate (per 100,000)") + 
-              geom_hline(yintercept = 0, col = "black", lwd = 0.5) +
-              #scale_y_continuous(limits = lim.y) + 
+              geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +  
               ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex1, " for ", COD1)) +
@@
   b <- ggplot(subset.data2, aes(x = year, y = rate.difference_mean)) +
+           # geom_hline(aes(yintercept = -300), lwd = 0.5, col = "grey") +
+           # geom_hline(aes(yintercept = -200), lwd = 0.5, col = "grey") +
+           # geom_hline(aes(yintercept = -100), lwd = 0.5, col = "grey") +
+           # geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey") +
+           # geom_hline(aes(yintercept = 200), lwd = 0.5, col = "grey") +
+           # geom_hline(aes(yintercept = 100), lwd = 0.5, col = "grey") +
               geom_line() +
               geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL), col = NA, alpha = 0.5) +
-              facet_wrap( ~ state.map.order, ncol = 11, drop = F) +
+              facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
               xlab(" ") + 
               ylab("Excess mortality among blacks (per 100,000)") +
-              #scale_y_continuous(breaks = 0) + #limits = lim.y2
-              geom_hline(yintercept = 0, col = "black", lwd = 0.5) +
-              #geom_vline(xintercept = 1969, col = "black", lwd = 0.5) +
+              geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +  
               ggtitle(paste0("Excess mortality among blacks (per 100,000) in ", sex1, " for ", COD1)) +
@@
 ```{r print-em, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cardiovascular.Male.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
-ggplotly(Cardiovascular.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+ggplotly(Cardiovascular.Male.mort.plot[[1]]) %>% 
+  layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F)
+
+ggplotly(Cardiovascular.Male.mort.plot[[2]]) %>% 
+  layout(xaxis = list(title = "Year (1969-2013)"))
 
@@
 ```{r print-em-2, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cardiovascular.Female.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
-ggplotly(Cardiovascular.Female.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+ggplotly(Cardiovascular.Female.mort.plot[[1]]) %>% 
+  layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F)
+
+ggplotly(Cardiovascular.Female.mort.plot[[2]]) %>% 
+  layout(xaxis = list(title = "Year (1969-2013)"))
 ```

### 2017-04-06 Code/1_Explore_Data.Rmd
add RID and partition data

@@
 head(dat)
+
+dat$stabbrs <- state.abb[match(dat$state, state.name)]
+unique(dat$state[is.na(dat$stabbrs)==T]) #Washington DC not defined
+
+dat$stabbrs[dat$state == "Washington DC"] <- "DC"
+
+dat$stabbrs <- factor(dat$stabbrs)
+
+dat$RID <- paste(dat$stabbrs, dat$year, substr(dat$sex, 1, 1),
+                 substr(dat$race, 1, 1), dat$age.n, dat$COD, sep = ".")
 ```
@@
 write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.csv", col_names = T)
+
+excluded.states <- c("Alaska", "Hawaii", "Idaho", "Maine", "Montana", 
+                     "New Hampshire", "North Dakota", "South Dakota", "Utah", "Vermont", "Wyoming")
+
+included.states <- setdiff(unique(levels(dat.clean$state)), excluded.states)
+
+for(i in 1:length(included.states)) { 
+  name = paste0('dat_clean_', included.states[i], '.csv')
+  ds_temp = dat.clean[dat.clean$state == included.states[i], ]
+  write.csv(ds_temp, file=name)
+}
+rm(ds_temp)
 ```

### 2017-04-06 Code/1_Explore_Data.Rmd
moved data

@@
   ds_temp = dat.clean[dat.clean$state == included.states[i], ]
-  write.csv(ds_temp, file=name)
+  write.csv(ds_temp, file = paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/", name))
 }

### 2017-04-06 Code/1_Explore_Data.Rmd

@@
 
-dat$RID <- paste(dat$stabbrs, dat$year, substr(dat$sex, 1, 1),
-                 substr(dat$race, 1, 1), dat$age.n, dat$COD, sep = ".")
+dat$COD.abbrev <- "CAR"
+dat$COD.abbrev[dat$COD == "Cancers"] <- "CAN"
+dat$COD.abbrev[dat$COD == "Communicable"] <- "CDS"
+dat$COD.abbrev[dat$COD == "Non-communicable"] <- "NDS"
+dat$COD.abbrev[dat$COD == "Injuries"] <- "INJ"
+dat$COD.abbrev[dat$COD == "All other causes"] <- "AOC"
+
+dat$RID <- paste0(dat$stabbrs, dat$year, substr(dat$sex, 1, 1),
+                 substr(dat$race, 1, 1), dat$age.n, dat$COD.abbrev)
 ```

### 2017-04-06 Code/2_smoothing_models/bayes_smoothing_functions.R
updates

@@ -1,11 +1,6 @@
 
-# User-defined subsets 
-subset_data = function(ds, sex, race) {
-  ds_sub = ds[ds$sex==sex & ds$race==race, ]
-  return(ds_sub)
-}
 
-# Prepare data for JAGS and put into list-by-COD format 
+# prepare data for JAGS and put into list-by-COD format 
 jagsify_data = function(ds_sub) { #subset by COD, put in lists 
   
   cods = unique(ds_sub$COD)
@@ -16,19 +11,20 @@ jagsify_data = function(ds_sub) { #subset by COD, put in lists
   ds_sub$year.n = ds_sub$year.n - (min(ds_sub$year.n) - 1)
   ds_sub$upper_bound = ifelse(ds_sub$population < 9, ds_sub$population, 9) 
   ds_sub$censored = ifelse(is.na(ds_sub$deaths), 1, 0)
+  ds2_save = list()
   
   for(i in 1:n.cods) {
     
     ds2 = ds_sub[ds_sub$COD == cods[i], ]
-    ds2 = ds2[order(ds2$censored), ]  # check on this 
+    ds2_save[[i]] = ds2
+    
+    ds2 = ds2[order(ds2$censored), ]  
     
     ds_jagsified[[i]] = list(deaths = ds2$deaths, 
                    lnpop = log(ds2$population), 
                    age.bin = ds2$age.n, 
                    year = ds2$year.n, 
                    upper_bound = ds2$upper_bound, 
-                   #binned = is.na(ds2$deaths), 
-                   #not.binned = !is.na(ds2$deaths), 
                    n.binned = sum(is.na(ds2$deaths)), 
                    n.not.binned = sum(!is.na(ds2$deaths)),
                    n.rows = sum(is.na(ds2$deaths)) + sum(!is.na(ds2$deaths)),
@@ -36,14 +32,14 @@ jagsify_data = function(ds_sub) { #subset by COD, put in lists
                    n.years = length(unique(ds2$year)), 
                    binned.id = ds2$censored,
                    RID = ds2$RID) 
-                   #cod = cods[i]) 
+    
+                   
   }
-  
-  ds_sub_onlyRID = ds_sub[ ,c('RID', 'deaths')] 
-  return(list(jagsified = ds_jagsified, sub = ds_sub_onlyRID))
+  return(list(jagsified = ds_jagsified, ds_ordered = ds2_save))
 }
 
-jags_smoothing_model = function(ds_jagsified_bycod) { 
+# the jags model 
[…]

### 2017-04-06 Code/entire_analysis.R

@@ -1,7 +1,17 @@
 
+#
+# entire_analysis.R: All functions wrapped together for full analysis by state  
+#               
+#
+# Written by Kathryn Morrison & Corinne Riddell, 2017 
+#
+#
+# kt.morrison@mail.mcgill.ca | corinne.riddell@mail.mcgill.ca 
+# Last revised: 06-04-2017 
 
-entire.analysis <- function(state_i) {
-  
+        # 0. --------------------- PREAMBLE --------------------- 
+prepare_environment = function(state_i)  {
+        
   #load_libraries
   library(Rcpp)
   library(dplyr)
@@ -20,158 +30,136 @@ entire.analysis <- function(state_i) {
   library(grid)
   library(gridExtra)
   library(parallel)
+  library(data.table)
+  library(dtplyr)
+  library(stringi)
+
+name <- paste0('~/BlackWhiteMortalityGap/Data/data_by_state/dat_clean_', state_i, '.csv')
+dat.clean <- read.csv(name) 
+return(dat.clean)
+}
   
-  # LOCAL -- edited -- 
-  #load_user_written_functions
-  source('/Users/kathryn/Documents/BlackWhiteMortalityGap/Code/2_smoothing_models/bayes_smoothing_functions.R')
-  source('/Users/kathryn/Documents/BlackWhiteMortalityGap/Code/life_expectancy_functions.R') 
+        # 1. --------------------- BAYES SMOOTHING --------------------- 
+
+bayes_smoothing_bystate = function(dat.clean, n.cods, seed) {        
   
-  # SERVER 
-  #load_user_written_functions
-  #source("~/BlackWhiteMortalityGap/Code/2_smoothing_models/bayes_smoothing_functions.R")
-  #source('~/BlackWhiteMortalityGap/Code/life_expectancy_functions.R') 
-  
-  #load_tidied_data -- for state_i --
-  
-  # state_i = 'Alabama'
-  name = paste0('/Users/kathryn/data_by_state/dat_clean_', state_i, '.csv')
-  dat.clean = read.csv(name) 
-  
-  #load US standard population
-  std.pop <- read.csv("~/BlackWhiteMortalityGap/Data/US_Standard_Population_for_model.csv") %>% select(US_Standard_2000, age.weight, age_minbin)
-  
-  #fit_models
-  #run time: 32 mins
-  #fit the bayesian models and put the 1000 posterior samples into a data frame
-  
-  n.cods = 6 
[…]

### 2017-04-06 Code/entire_analysis_run.R

@@ -1,0 +1,35 @@
+
+source("~/BlackWhiteMortalityGap/Code/entire_analysis.R")
+source('~/BlackWhiteMortalityGap/Code/life_expectancy_functions.R') 
+source("~/BlackWhiteMortalityGap/Code/2_smoothing_models/bayes_smoothing_functions.R")
+
+state_i = 'Alabama'
+n.cods = 6 
+
+dat.clean <- prepare_environment(state_i)
+r.All <- bayes_smoothing_bystate(dat.clean=dat.clean, n.cods = n.cods, seed=12345)
+age_stdz_mortality_bystate(dat.clean, state_i, r.All) 
+le.calculations <- life_expectancy_and_decomp(dat.clean, state_i, r.All)
+
+systime1 <- system.time(dat.clean <- prepare_environment(state_i)) # 1.923
+systime2 <- system.time(r.All <- bayes_smoothing_bystate(dat.clean = dat.clean, n.cods = n.cods, seed = 12345))   #736.76 seconds
+systime3 <- system.time(age_stdz_mortality_bystate(dat.clean, state_i, r.All)) # 84.821 
+systime4 <- system.time(le.calculations <- life_expectancy_and_decomp(dat.clean, state_i, r.All))  # 106.5 mins 
+life_expectancy_and_decomp_pt2(le.calculations, r.All)
+
+
+load('le_calculations_savedseed12345_A.RData')
+le.calculationsA = le.calculations 
+
+load('le_calculations_savedseed12345_B.RData')
+le.calculationsB = le.calculations 
+
+
+# 2-ish hours...still a lot but it's something 
+
+
+le.calculationsA[[5]]$calcs$`1`$le.df$LE_White
+le.calculationsB[[5]]$calcs$`1`$le.df$LE_White
+
+le.calculationsA[[5]]$calcs$`1`$le.df$LE_Black
+le.calculationsB[[5]]$calcs$`1`$le.df$LE_Black

### 2017-04-06 Code/life_expectancy_functions.R

@@ -6,7 +6,7 @@
 
 life.table <- function(data, sortable.age, num.ages.in.group, death.counts, population.counts, ave.prop.le1.lived = 0.09, return = "appended"){
 
-  data <- dplyr::arrange_(data, sortable.age)
+  data <- dplyr::arrange_(data, sortable.age) 
   
   data["R_x"] <- data[death.counts]/data[population.counts] #mortality rates
   data["a_x"] <- 0.5 #average proportion of interval lived
@@ -78,6 +78,10 @@ le_age_decomp <- function(life.table1, name.lt1 = 1, life.table2, name.lt2 = 2, 
   return(decomp.table)
 }
 
+# cause_of_death_decomp
+# life.table1 = lt.black ; life.table2 = lt.white ; decomp.table = age_decomp
+# cod.table = ct.both ;   age.colname.cod.table = "age_minbin" ;  COD.colname.cod.table = "COD"
+# prop1.colname.cod.table = "prop.black" ; prop2.colname.cod.table = "prop.white"
 
 #cause of death table must be organized in a very specific way
 #it needs to be sorted by COD, and then by age (within COD)
@@ -100,7 +104,7 @@ cause_of_death_decomp <- function(life.table1, life.table2, decomp.table,
                                  "C_xi" = C_xi)
   
   COD.decomp.table["C_xi_proportion"] <- COD.decomp.table["C_xi"]/sum(COD.decomp.table["C_xi"])
-  COD.decomp.table["sign"] <- ifelse(COD.decomp.table["C_xi"] < 0, "negative", "positive")
+  COD.decomp.table["sign"] <- sign(COD.decomp.table["C_xi"])
   
   names(COD.decomp.table)[2] <- prop1.colname.cod.table
   names(COD.decomp.table)[3] <- prop2.colname.cod.table
@@ -108,8 +112,7 @@ cause_of_death_decomp <- function(life.table1, life.table2, decomp.table,
   return(COD.decomp.table)
   }
 
-#a general function that takes the decomposition table and adds variables to allow the user
-#to easily plot the decomposition in a pleasing manner
+
 make_pretty_decomp_plot <- function(decomp.table, strat.var.1, strat.var.2, strat.var.3, partition.bar.var, 
                                     sign.var, decomp.var, decomp.var.prop, type.of.data = NA) {
[…]

### 2017-04-06 Code/optim_explore.R

@@ -1,48 +1,127 @@
+# ---- How are JAGS data ordered / what order are the results returned in?
 
-dat.clean = read.csv('~/dat_clean.csv') 
+#write.csv(ds2, 'test_data_mcmcoptim.csv', row.names=FALSE)
+ds = read.csv('test_data_mcmcoptim.csv')
+ds2 = ds[order(ds$censored), ]  
 
-included.states = c("Alabama", "Arizona", "Arkansas", "California", "Colorado", "Connecticut",
-                     "Delaware", "Florida", "Georgia", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
-                     "Louisiana", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi",   
-                     "Missouri", "Nebraska", "Nevada", "New Jersey", "New Mexico", "New York",
-                     "North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island",
-                     "South Carolina",  "Tennessee", "Texas", "Virginia", "Washington", "Washington DC",
-                     "West Virginia", "Wisconsin")  
+cbind(ds$censored, ds2$censored)
+    
+ds_jagsified = list(deaths = ds2$deaths, 
+                             lnpop = log(ds2$population), 
+                             age.bin = ds2$age.n, 
+                             year = ds2$year.n, 
+                             upper_bound = ds2$upper_bound, 
+                             n.binned = sum(is.na(ds2$deaths)), 
+                             n.not.binned = sum(!is.na(ds2$deaths)),
+                             n.rows = sum(is.na(ds2$deaths)) + sum(!is.na(ds2$deaths)),
+                             n.age.bins = length(unique(ds2$age.n)),
+                             n.years = length(unique(ds2$year)), 
+                             binned.id = ds2$censored,
+                             RID = ds2$RID) 
 
-state_abbrev = as.character(quote(c(AL, AZ, AR, CA, CO, CT, DE, FL, GA, IL, IN, IA, KS, KY, LA, 
[…]

### 2017-04-07 Code/combined-meta-regression.Rmd

--- 2017-04-04 Manuscript/Manuscript_COD_Trends.Rmd
@@
 ---
-title: 'Historical trends in the contribution of six major causes of death to the black-white life expectancy gap by US state'
-bibliography: bibliography.bib
-output:
-  html_document: default
-  word_document: default
-always_allow_html: yes
+title: "meta-analysis"
+author: "Corinne Riddell"
+date: "4/7/2017"
+output: html_document
 ---
 
-Corinne A Riddell, PhD
+In this file, we perform the meta-regression.
+This regression brings together the mean estimates and their 95% CIs from the state-specific models to create a smoothed line of the general pattern (across the US) of the contribution of each cause of death to the black-white life expectancy gap, separately for males and females.
 
-Kathryn T Morrison, PhD
+In the paper, we perform a random-effects meta regression, that includes random intercepts for each state and plot the general pattern (when the RE estimate is equal to 0).
+In this file we also include code for the fixed effect estimation, for interested readers.
+This model directly estimates vertical shifts to the mean curve for each state, which may be of interest to some readers.
 
-Sam Harper, PhD
-
-Jay S Kaufman, PhD
-
-Department of Epidemiology, Biostatistics and Occupational Health, McGill University, Montreal, Canada.
-
-Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada.
-Email: <corinne.riddell@mail.mcgill.ca>
-
-###Abstract
-
-**Background**
-
-**Methods and Findings**
-
-**Conclusions:**
-
-###Introduction
-
-The life expectancy of a population is a barometer for overall health.
-In 2014, life expectancy in the United States was 78.8 years, but hidden within this figure are differences by race.
-Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf].
-While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.
-
-During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [@harper2007trends].
-The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement.
-For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [@harper2014trends].
-
-The over-arching objective of this study was to describe the pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences.
-To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for males and females.
-This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states.
-Ideally, this study may facilitate the identification of one policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.
-
-###Methods
-
-**Description of the mortality data**
-
-We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.
-The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.
-The population data were derived by the NCHS using intercensal bridged-race population estimates.
-This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death.
-Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013).
-These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**).
-When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.
-
-For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming.
-These states contained less than 0.4% of the black population in the United States in 2010 [ref].
-We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since the quality of the coding of ethnicity varied by state over the time period included [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf].
-
-**Statistical challenges to be addressed**
-
-The data structure produced two statistical challenges that were addressed by our quantitative approach.
-First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9).
-Second, several states had small black populations which produced sparse counts in each sex-age category.
-In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic.
-To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.
-
-**Estimation of smoothed mortality rates using a Bayesian autoregressive model**
-
-To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts.
-Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population.
-The mortality rates over time within each state-age-sex-race category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year.
-Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9).
-
-These models were run separately for every combination of state, race, and sex.
-They were fit using two markov chains of length 10,000.
-One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration (to reduce autocorrelation).
-Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition.
-These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.
-
-**Decomposition of the life expectancy gap by cause of death**
-
-To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref].
-Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex.
-The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.
-
-To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref].
-First, the model partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks.
-For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites.
-This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination.
-For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old.
-We summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference, for each state and for each sex.
-Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 45 state-level trends and their credible intervals.
-
-**Description of the trend in each cause of death's contribution to the gap and identification of divergent states**
-
-To describe the general pattern in these trend lines we used random effects meta regression.
-Each state's mean contribution was regressed against a linear spline for calendar time that was estimated with five degrees of freedom.
-The model included random effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year.
-To identify divergent states, we compared the state-specific trends (estimated separately for each state using the Bayesian model) to the general pattern estimated using meta regression.
-
-*Need to add information on the trends in mortality* -- where?
-
-###Results
-```{r initialize, echo=FALSE, warning=F, message=F}
-library(plotly)
-library(viridis)
-library(scales)
-library(dplyr)
-library(purrr)
-library(stringr)
-library(DT)
+```{r initialize, warning=F, message=F}
+library(lme4)
 library(splines)
@@
 
-```{r load_data, echo=F}
-BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
-BlackWhite_results <- reorder.as.map(BlackWhite_results, "state")
-
+```{r load-data}
 cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
@@
 cod_decomp_results$COD_cont_yrs_SE <- (cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl)/3.92
-
-cod_decomp_results <- reorder.as.map(cod_decomp_results, "state")
 ```
 
-```{r life-expectancy-trends, echo=F}
-LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
-  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
-  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
-  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 10) +
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Life expectancy gap (years)") +
-  xlab("Year (1969-2013)")
-
-LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
-  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
-  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
-  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 10) +
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Life expectancy gap (years)") +
-  xlab("Year (1969-2013)")
-```
-
-```{r plot_both, eval=F, echo=F}
-#Informative: male and females at the same time
-ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
-  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
-  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
-  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 10) +
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Life expectancy gap (years)") +
-  xlab("Year (1969-2013)"))
-```
-
-```{r meta_regression_of_cod_contribution, echo = F}
-#model option 1: FE analysis
+```{r FE_meta_regression_of_cod_contribution, echo = F}
+#Fixed effects meta analysis (fixed effects for state)
 
@@
 
-```{r meta_RE_analysis, echo = F}
-library(lme4)
+```{r RE_meta_regression_of_cod_contribution, echo = F}
+#Random effects meta analysis (fixed effects for state)
 
@@
 cod_decomp_results <- merge(cod_decomp_results, 
-                             RE.pred %>% select(COD, sex, state, year, 
-                                                bs.year..degree...1..df...5.,
-                                                .fitted, .resid, .fixed, .mu), 
+                             RE.pred %>% select(COD, sex, state, year, .fitted, .resid, .fixed), 
                              by = c("COD", "sex", "state", "year"))
 
-cod_decomp_results <- cod_decomp_results %>% rename(fitted.RE = .fitted, fixed.from.RE = .fixed)
-
-# test1 <- lmer(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + (1|state), weights = 1/(COD_cont_yrs_SE)^2, data = subset(cod_decomp_results, COD == "Cardiovascular" & sex == "Male"))
-# 
-# pred.general.trend <- predict(test1, re.form = NA)
-# pred.with.REs <- predict(test1)
-# 
-# plot(y=pred.general.trend[1:45], x=1969:2013)
+cod_decomp_results <- cod_decomp_results %>% rename(fitted.RE = .fitted, fixed.from.RE = .fixed, resid.RE = .resid)
 ```
 
-```{r plotting_resid_fx, echo = F}
-plot.resid.trends <- function(dataset, COD1, sex1) {
-  subset.data <- subset(dataset, COD == COD1 & sex == sex1)
-  subset.data$order <- reorder(subset.data$state, subset.data$sum.residual, first)
-
-  a <- ggplot(subset.data,
-         aes(x = year, y = COD_cont_yrs_mean)) + 
-    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
-    #geom_line(aes(col = "Observed")) + 
-    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9") +
-    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
-    #scale_fill_manual(values = c("#b2e2e2")) +
-    facet_wrap( ~ state.map.order, ncol = 11, drop = F) + 
-    geom_line(aes(y = fitted.FE), col = "#d7191c") +
-    ggtitle(paste0("Estimated smoothed contribution vs. meta-estimated common trend in ", COD1, " in ", sex1)) +
-    theme_classic(base_size = 10) + 
-    theme(axis.text.x = element_text(angle = 45, hjust = 1),
-          strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-    ylab("estimated smoothed contribution vs. common trend (years)") +
-    xlab("Year (1969-2013)")
-
-  b <- ggplot(subset.data,
-              aes(x = year, y = resid.FE)) + geom_line(aes(col = "Residual")) + 
-    scale_color_manual(values = c("#d7191c")) +    
-    facet_wrap( ~ order) +
-    xlab(" ") + ylab("Residual") +
-    geom_hline(yintercept = 0, col = "black") +
-      geom_text(data = subset(subset.data, year == 2010), 
-                aes(y = 1, label = round(sum.residual, 1)), size = 2.5) + 
-    ggtitle(paste0("Residuals vs. year in ", COD1, " in ", sex1)) +
-    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-  c <- ggplot(subset.data,
-              aes(x = year, y = COD_cont_yrs_mean)) + 
-    # geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey") +
-    # geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey") +
-    # geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey") +
-    # geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey") +
-    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
-
-    #geom_line(aes(col = "Observed")) + 
-    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9") +
-    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
-    #scale_fill_manual(values = c("#b2e2e2")) +
-    facet_wrap( ~ state.map.order, ncol = 11, drop = F) + 
-    geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
-    #geom_line(aes(y = fitted.RE), col = "#d7191c", lty = 3) +
-    ggtitle(paste0("Estimated smoothed contribution of", COD1, " disease to the black-white life expectancy gap (blue 95% credible interval) vs.
-the meta-estimated general pattern (red line) in ", sex1, " 1969-2013")) +
-    theme_classic(base_size = 10) + 
-    theme(axis.text.x = element_blank(),
-          strip.background=element_blank(),
-          axis.line=element_blank(),
-          axis.ticks=element_blank()) +
-    ylab("Contribution to the life expectancy gap (years)") +
-    xlab("Year (1969-2013)")
-
-  return(list(plot1 = a, plot2 = b, plot3 = c))
-}
+```{r save_data}
+write.csv(cod_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
 ```
-
-```{r create-the-resid-plots, echo = F, message = F}
-for(COD_i in levels(cod_decomp_results$COD)) {
-  for(sex_i in levels(cod_decomp_results$sex)) {
-    assign(paste0(COD_i, ".", sex_i, ".plot"), plot.resid.trends(dataset = cod_decomp_results, 
-                                                                         COD1 = COD_i, sex1 = sex_i))
-  }
-}
-```
-
-```{r refine_plots, eval = F, echo = F}
-#need tp add the "grid lines post-hoc" because they depend on the max height of the plot
-
-#turn hover off for grid lines and axis lines
-refine.patterns <- function(plot) {
-  for(i in 1:320){
-  if(plot$x$data[[i]]$line$color == "rgba(190,190,190,1)"){
-    plot$x$data[[i]]$line$width <- 0.7559055
-    plot$x$data[[i]]$hoverinfo <- "none"
-  }
-    if(plot$x$data[[i]]$line$color == "rgba(0, 0, 0, 1)"){
-    plot$x$data[[i]]$hoverinfo <- "none"
-  }
-  }
-  return(plot)
-}
-
-#refine.patterns(Cardiovascular.Male.plot[[3]])
-```
-
-**Figures 1 and 2** display trends in the life expectancy gap for males and females, respectively.
-For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males.
-*Don't know whether you want to note it here or not, but I find it is interesting to see that the gap is now lower in many US southern (or southeastern) states than in the Midwest, suggesting that progress has actually been somewhat greater in the South than elsewhere (particularly the upper Midwest and especially for women).
-I think this is interesting and somewhat counter to the narrative that health indicators are worse in the South.
-I also wonder how this might relate to the Murray/IHME work on US counties with "declining" life expectancy among white women.*
-
-**Figures 2 - 14** display trends in the contribution of each cause of death to the life expectancy gap for males and females, in years.
-The blue band illustrates the credible interval for the smoothed contribution estimated for each state, and illustrates the heterogeneity across states in their contribution trends.
-These curves are overlaid with the meta-estimated common trend, shown with a red line.
-We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).
-
-**Cardiovascular disease**
-
-Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races.
-Mortality from CVD decreased markedly since 1969 for both sexes and races (**eFigures 1 and 2**).
-However, white males underwent relatively larger mortality decreases than black males, resulting in a modest increase to the contribution of CVD to the life expectancy gap for most states (**Figure 3**).
-In particular, Pennsylvania, Michigan, Wisconsin, and Washington DC experienced relatively larger increases to their CVD contributions to the gap for males, while some states such as Illinois and Kentucky exhibited relatively larger early contributions in the 1970s.
-
-The situation was reversed for females, with black females experiencing relatively larger mortality decreases over time compared to white females.
-Thus, the contribution of CVD to the life expectancy gap decreased by more than a year in several states (**Figure 4**).
-Of note, some states experienced a period of early increasing contributions, most notably, Arizona, Washington, and Oregon, that was not captured by the general pattern.
-Furthermore, Wisconsin and Washington DC both exhibit increasing contributions over time, as mortality gains among white females outpaced those among black females.
-
-**Cancer**
-
-For all races and sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**).
-Among males, this inverted U-shape pattern was accentuated among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**).
-Notable exceptions to this general pattern include Washington DC and Mississippi, states that underwent increasing cancer contributions over time, and New Mexico and Rhode Island, exhibiting decreasing contributions over time (although imprecisely estimated).
-
-For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern is similar between blacks and whites and most states do not exhibit the accentuated period of increase in mortality that was estimated for black males.
-Thus, the contribution of cancer to the life expectancy gap is generally flat over time (**Figure 6**).
-[You could potentially stop here to save space and just include the comment about there being little inequality in cancer mortality ].
-Washington DC deviates from this pattern, and exhibited increasing cancer contributions to the gap, while Rhode Island exhibited the reverse.
-Most notably, some states exhibited no inequality in cancer mortality (**eFigures 3 and 4**), including New York, and Connecticut (and for males, these states recently eliminated the black-white difference in cancer mortality).
-Furthermore, Massachusetts and (imprecisely) Rhode Island exhibit lower cancer rates among black women and men compared to whites in 2013.
-
-**Non-communicable disease**
-
-Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**Figure X**).
-Recently, there has been a slight decrease in mortality rates, especially among black males.
-These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states.
-As such, the contribution of non-communicable disease to the mortality gap was general flat among men with a recent decrease, and exhibited a gradual decrease over time among women.
-
-**Injury**
-
-Injury is one of the least common causes of death.
-Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime in the 1990s resulted in heightened mortality from injuries among black males at that time (**Figure X**).
-For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**Figure X**).
-Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, especially among females.
-Thus, the contribution of injuries to the mortality gap has generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures X and X**).
-Among males, Wisconsin and Washington DC diverged the most from the general pattern, as both states experienced steep increases in the contribution of injuries during the 1990s that was not captured by the general trend.
-Washington DC also exhibited this pattern among females.
-
-**Communicable disease**
-
-Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communiable disease at the height of the HIV/AIDS epidemic before anti-retroviral therapy was available.
-This spike was greater among males, especially blacks, and within states with larger urban centers.
-Thus, the contribution of communicable disease to the life expectancy gap exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics.
-
-****
-
-**Figure 1: State-level trends in the difference in life expectancy between blacks and whites in males, United States, 1969-2013**
-
-```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11}
-ggplotly(LE_trend_plot_males)
-```
-
-**Figure 2: State-level trends in the difference in life expectancy between blacks and whites in females, United States, 1969-2013**
-
-```{r plot_LE_trends_females, echo=F, fig.height=9, fig.width=11}
-ggplotly(LE_trend_plot_females)
-```
-
-**Figure 3: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in males**
-
-```{r print-plot, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cardiovascular.Male.plot[[3]]) 
-```
-
-**Figure 4: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in females**
-
-```{r print-plot-2, fig.width = 9, fig.height = 9, echo = F}
-#ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cardiovascular.Female.plot[[3]]) 
-```
-
-**Figure 5: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in males**
-
-```{r print-plot-3, fig.width = 9, fig.height = 9, echo = F}
-#ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cancers.Male.plot[[3]]) 
-```
-
-**Figure 6: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in females**
-
-```{r print-plot-4, fig.width = 9, fig.height = 9, echo = F}
-#ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cancers.Female.plot[[3]]) 
-```
-
-**Figure 7: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the overall pattern in males**
-
-```{r print-plot-5, fig.width = 9, fig.height = 9, echo = F}
-#ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`Non-communicable.Male.plot`[[3]]) 
-```
-
-**Figure 8: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the overall pattern in females**
-
-```{r print-plot-6, fig.width = 9, fig.height = 9, echo = F}
-#ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`Non-communicable.Female.plot`[[3]]) 
-```
-
-**Figure 9: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the overall pattern in males**
-
-```{r print-plot-7, fig.width = 9, fig.height = 9, echo = F}
-#ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Injuries.Male.plot[[3]])
-```
-
-**Figure 10: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the overall pattern in females**
-
-```{r print-plot-8, fig.width = 9, fig.height = 9, echo = F}
-#ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Injuries.Female.plot[[3]]) 
-```
-
-**Figure 11: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the overall pattern in males**
-
-```{r print-plot-9, fig.width = 9, fig.height=9, echo = F}
-#ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Communicable.Male.plot[[3]]) 
-```
-
-**Figure 12: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the overall pattern in females**
-
-```{r print-plot-10, fig.width = 9, fig.height=9, echo = F}
-#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Communicable.Female.plot[[3]]) 
-```
-
-**Figure 13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males**
-
-```{r print-plot-11, fig.width = 9, fig.height=9, echo = F}
-#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`All other causes.Male.plot`[[3]])
-```
-
-**Figure 14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males**
-
-```{r print-plot-12, fig.width = 9, fig.height=9, echo = F}
-#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`All other causes.Female.plot`[[3]])
-```
-
-###Discussion
-
-Limitations
--does not account for migration. Washington DC - have a lot of migration of high SES white folks coming in, and perhaps also relatively high SES black folks moving out
--does not account for hispanic ethinicity - demographic shifts in FL for example of black and white hispanics which will pull LE and disparities
-
-###Appendix
-
-###Bibliography

### 2017-04-07 Manuscript/Manuscript_COD_Trends.Rmd

@@
 
+```{r initialize, echo=FALSE, warning=F, message=F}
+library(plotly)
+library(viridis)
+library(scales)
+library(dplyr)
+library(purrr)
+library(stringr)
+library(DT)
+library(splines)
+library(broom)
+library(tidyverse)
+
+source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+```
+
+```{r load_data, echo=F}
+BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
+BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")
+
+cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
+cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
+```
+
 ###Abstract
@@
 
+The overall difference in life expectancy between blacks and whites in the US has decreased over time.
+However, states have fared differently in their progress towards eliminating the black-white life expectancy gap.
+Our objective is to describe the pattern of contributions of each of six major causes of deaths to the sex-specific black-white life expectancy gaps over the last half-century, and identify divergent states.
+
 **Methods and Findings**
+```{r summarize_results_for_abstract, echo = F}
+cvd.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
+cvd.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
+cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
+cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
 
-**Conclusions:**
+ncd.male.average <- cod_decomp_results %>% filter(sex == "Male" & COD == "Non-communicable" & state == "Alabama") %>% 
+  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
+ncd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
+ncd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
+
+can.male.max <- cod_decomp_results %>% filter(sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
+  select(fixed.from.RE) %>% summarize(max = round(max(fixed.from.RE), 1))
+can.female.average <- cod_decomp_results %>% filter(sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
+  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
+
+inj.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
+inj.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
+inj.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
+inj.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
+```
+
+Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, sex, race, 19 age groups, and six major causes of death.
+A Bayesian autoregressive model was used to estimate smoothed mortality rates over time and Arriaga’s method was used to decompose the difference in life expectancy.
+Random effects meta-analysis was used to generate an overall pattern across the states, and was contrasted with the state-specific patterns to identify divergent states.
+
+Although mortality from cardiovascular disease has decreased over time for both racial groups, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r cvd.female.2013` years).
+Non-communicable diseases’ contribution to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in 1988) and a stable contribution for women (approximately `r can.female.average` years).
+Both men and women exhibited a decreased contribution for injuries (men: `r inj.male.1969` to `r inj.male.2013` years, women: `r inj.female.1969` to `r inj.female.2013` years), with the contribution becoming negative for women.
+Several states did not follow these general trends.
+
+**Conclusions**
+
+The United States has shown improvement in life expectancy for both blacks and whites since 1969.
+For males, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality, and point to the cause-specific interventions necessary to decrease the black-white gap further.
+In women, a crossover in injury-related mortality led to a narrower gap that was realized partially by increasing injury-related mortality among whites.
+This work emphasizes the importance of understanding how the direction of changes in mortality by race results in changes to the life expectancy gap.
 
@@
 ###Results
-```{r initialize, echo=FALSE, warning=F, message=F}
-library(plotly)
-library(viridis)
-library(scales)
-library(dplyr)
-library(purrr)
-library(stringr)
-library(DT)
-library(splines)
-library(broom)
-library(tidyverse)
-
-source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
-```
-
-```{r load_data, echo=F}
-BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
-BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")
-
-cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
-cod_decomp_results$COD_cont_yrs_width <- cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl
-cod_decomp_results$COD_cont_yrs_SE <- (cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl)/3.92
-
-cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
-```
-
 ```{r life-expectancy-trends, echo=F}
@@
 
-```{r meta_regression_of_cod_contribution, echo = F}
-#model option 1: FE analysis
-
-#run 12 stratified models. bs() created the b-spline basis
-models <- cod_decomp_results %>% 
-  group_by(COD, sex) %>%
-  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state,
-              weights = 1/(COD_cont_yrs_SE)^2, 
-              data = .))
-
-#calculated predicted values for all models
-dfPred <- broom::augment(models, fit)
-dfPred$year <- 1969:2013
-
-#merge the predictions with original dataset
-cod_decomp_results <- merge(cod_decomp_results, 
-                            dfPred %>% select(COD, sex, year, state, .fitted, .resid), 
-                            by = c("COD", "sex", "state", "year")) %>%
-                      rename(fitted.FE = .fitted, resid.FE = .resid)
-
-#calculated the sum of the residuals for each state-sex-cod strata (summed over years)
-cod_decomp_results <- cod_decomp_results %>% 
-  group_by(state, sex, COD) %>%
-  mutate(sum.residual = sum(abs(resid.FE)))
-```
-
-```{r meta_RE_analysis, echo = F}
-library(lme4)
-
-#run 12 stratified models. bs() created the b-spline basis
-RE.models <- cod_decomp_results %>% 
-  group_by(COD, sex) %>%
-  do(fit = lmer(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + (1|state),
-              weights = 1/(COD_cont_yrs_SE)^2, 
-              data = .))
-
-#calculated predicted values for all models
-RE.pred <- broom::augment(RE.models, fit)
-RE.pred$year <- 1969:2013
-
-RE.models$RE.estimates <- purrr::map(RE.models$fit, ~ ranef(.x))
-
-sex <- rep(c(rep("Female", 40), rep("Male", 40)), 6)
-COD <- c(rep("All other causes", 80), rep("Cancers", 80), rep("Cardiovascular", 80),
-         rep("Communicable", 80), rep("Injuries", 80), rep("Non-communicable", 80))
-RE.estimates <- c(RE.models$RE.estimates[[1]]$state$`(Intercept)`, RE.models$RE.estimates[[2]]$state$`(Intercept)`, 
-                  RE.models$RE.estimates[[3]]$state$`(Intercept)`, RE.models$RE.estimates[[4]]$state$`(Intercept)`, 
-                  RE.models$RE.estimates[[5]]$state$`(Intercept)`, RE.models$RE.estimates[[6]]$state$`(Intercept)`, 
-                  RE.models$RE.estimates[[7]]$state$`(Intercept)`, RE.models$RE.estimates[[8]]$state$`(Intercept)`, 
-                  RE.models$RE.estimates[[9]]$state$`(Intercept)`, RE.models$RE.estimates[[10]]$state$`(Intercept)`,
-                  RE.models$RE.estimates[[11]]$state$`(Intercept)`, RE.models$RE.estimates[[12]]$state$`(Intercept)`)
-state <- rep(levels(cod_decomp_results$state), 12)
-REs <- data.frame(sex, COD, RE.estimates, state)
-
-cod_decomp_results <- merge(cod_decomp_results,
-                            REs, by = c("COD", "sex", "state"),
-                            all.x = T)
-
-#merge the predictions with original dataset
-cod_decomp_results <- merge(cod_decomp_results, 
-                             RE.pred %>% select(COD, sex, state, year, 
-                                                bs.year..degree...1..df...5.,
-                                                .fitted, .resid, .fixed, .mu), 
-                             by = c("COD", "sex", "state", "year"))
-
-cod_decomp_results <- cod_decomp_results %>% rename(fitted.RE = .fitted, fixed.from.RE = .fixed)
-
-# test1 <- lmer(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + (1|state), weights = 1/(COD_cont_yrs_SE)^2, data = subset(cod_decomp_results, COD == "Cardiovascular" & sex == "Male"))
-# 
-# pred.general.trend <- predict(test1, re.form = NA)
-# pred.with.REs <- predict(test1)
-# 
-# plot(y=pred.general.trend[1:45], x=1969:2013)
-```
-
 ```{r plotting_resid_fx, echo = F}
@@
               aes(x = year, y = COD_cont_yrs_mean)) + 
-    geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey") +
-    geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey") +
-    geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey") +
-    geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey") +
     geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
@@
 **Figures 2 - 14** display trends in the contribution of each cause of death to the life expectancy gap for males and females, in years.
-The blue band illustrates the credible interval for the smoothed contribution estimated for each state, and illustrates the heterogeneity across states in their contribution trends.
-These curves are overlaid with the meta-estimated common trend, shown with a red line.
+The blue line illustrates the mean of the smoothed contribution estimated for each state-year, and illustrates the heterogeneity across states in their contribution trends.
+These lines are overlaid with the meta-estimated common trend, shown with a red line.
 We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).
@@
 ```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11}
-ggplotly(LE_trend_plot_males)
+LE_trend_plot_males +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)
+
 ```
@@
 ```{r plot_LE_trends_females, echo=F, fig.height=9, fig.width=11}
-ggplotly(LE_trend_plot_females)
+LE_trend_plot_females +
+  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5)
+
 ```
@@
 ```{r print-plot, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cardiovascular.Male.plot[[3]]) 
+Cardiovascular.Male.plot[[3]] +
+           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)
+
 ```
@@
 #ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cardiovascular.Female.plot[[3]]) 
+Cardiovascular.Female.plot[[3]]  +
+           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)
+
 ```
@@
 #ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cancers.Male.plot[[3]]) 
+Cancers.Male.plot[[3]] +
+           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
+
 ```
@@
 #ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Cancers.Female.plot[[3]]) 
+Cancers.Female.plot[[3]] +
+           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
+
 ```
@@
 #ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`Non-communicable.Male.plot`[[3]]) 
+`Non-communicable.Male.plot`[[3]] +
+           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)
+
 ```
@@
 #ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`Non-communicable.Female.plot`[[3]]) 
+`Non-communicable.Female.plot`[[3]] +
+           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)
+
 ```

### 2017-04-07 Manuscript/Manuscript_COD_Trends.Rmd
AOC results

@@
 
+**All other causes**
+
+Mortality from all other causes generally decreased earlier in the time period, especially among blacks.
+White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, while black men have continued to experience decreasing rates within several states.
+As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and reached equality in New York, Massachusettes, and Rhode Island for both genders (**Figures 13 and 14**).
+The most disparate states include Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.
+
 ****
@@
 
-**Figure 14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males**
+**Figure 14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in females**
 

### 2017-04-10 Code/combine_results.Rmd
started lit review

@@
 
-```{r combine_results}
-age_cod_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_", completed.states[1], ".csv"))
-age_decomp_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_", completed.states[1], ".csv"))
-BlackWhite_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_", completed.states[1], ".csv"))
-cod_decomp_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_", completed.states[1], ".csv"))
-cod_change_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_change_", completed.states[1], ".csv"))
+```{r combine_results_seed123}
+age_cod_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_", completed.states[1], "_123.csv"))
+age_decomp_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_decomp_", completed.states[1], "_123.csv"))
+BlackWhite_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_", completed.states[1], "_123.csv"))
+cod_decomp_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_marginal_", completed.states[1], "_123.csv"))
+cod_change_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_change_", completed.states[1], "_123.csv"))
 
-mortality.rates <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_", completed.states[1], ".csv"))
-mortality.rates.diff <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_", completed.states[1], ".csv"))
+mortality.rates <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/mortality_rates_", completed.states[1], "_123.csv"))
+mortality.rates.diff <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/mortality_rates_diff_", completed.states[1], "_123.csv"))
 
@@
 
-    age_cod_results <- rbind(age_cod_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_", completed.states[state_i], ".csv")))
+    age_cod_results <- rbind(age_cod_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_", completed.states[state_i], "_123.csv")))
 
-    age_decomp_results <- rbind(age_decomp_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_", completed.states[state_i], ".csv")))
+    age_decomp_results <- rbind(age_decomp_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_decomp_", completed.states[state_i], "_123.csv")))
 
-    BlackWhite_results <- rbind(BlackWhite_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_", completed.states[state_i], ".csv")))
+    BlackWhite_results <- rbind(BlackWhite_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_", completed.states[state_i], "_123.csv")))
 
-    cod_decomp_results <- rbind(cod_decomp_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_", completed.states[state_i], ".csv")))
+    cod_decomp_results <- rbind(cod_decomp_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_marginal_", completed.states[state_i], "_123.csv")))
 
-    cod_change_results <- rbind(cod_change_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_change_", completed.states[state_i], ".csv")))
+    cod_change_results <- rbind(cod_change_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_change_", completed.states[state_i], "_123.csv")))
 
-  mortality.rates <- rbind(mortality.rates, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_", completed.states[state_i], ".csv")))
+  mortality.rates <- rbind(mortality.rates, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/mortality_rates_", completed.states[state_i], "_123.csv")))
 
-    mortality.rates.diff <- rbind(mortality.rates.diff, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_", completed.states[state_i], ".csv")))
+    mortality.rates.diff <- rbind(mortality.rates.diff, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/mortality_rates_diff_", completed.states[state_i], "_123.csv")))
 }
@@
 ```{r save_combined_datasets, echo = F}
-write.csv(age_cod_results_female, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results_female.csv")
-write.csv(age_cod_results_male, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results_male.csv")
-write.csv(age_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_results.csv")
-write.csv(cod_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
-write.csv(BlackWhite_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
-write.csv(cod_change_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_change_results.csv")
+write.csv(age_cod_results_female, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_female.csv")
+write.csv(age_cod_results_male, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_male.csv")
+write.csv(age_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_decomp_results.csv")
+write.csv(cod_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
+write.csv(BlackWhite_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
+write.csv(cod_change_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_change_results.csv")
 
-write.csv(mortality.rates, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_combined.csv")
-write.csv(mortality.rates.diff, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_combined.csv")
+write.csv(mortality.rates, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/mortality_rates_combined.csv")
+write.csv(mortality.rates.diff, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/mortality_rates_diff_combined.csv")
 ```

### 2017-04-10 Code/combined-meta-regression.Rmd

@@
 ```{r load-data}
-cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
+cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
 cod_decomp_results$COD_cont_yrs_width <- cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl
@@
 ```{r save_data}
-write.csv(cod_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
+write.csv(cod_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
 ```

### 2017-04-10 Manuscript/Manuscript_COD_Trends.Rmd

@@
 ```{r load_data, echo=F}
-BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
+BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
 BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")
 
-cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
+cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
 cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
@@
 The life expectancy of a population is a barometer for overall health.
-In 2014, life expectancy in the United States was 78.8 years, but hidden within this figure are differences by race.
+In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethinicity.
 Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf].
@@
 
-The over-arching objective of this study was to describe the pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences.
+The over-arching objective of this study is to describe the pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences.
 To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for males and females.
 This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states.
-Ideally, this study may facilitate the identification of one policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.
+This study may therefore facilitate the identification of policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.
 
@@
 To describe the general pattern in these trend lines we used random effects meta regression.
-Each state's mean contribution was regressed against a linear spline for calendar time that was estimated with five degrees of freedom.
+Each state's mean contribution was regressed against a linear spline for calendar time that was estimated using four internal knots.
 The model included random effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year.
@@
 
-**Figures 2 - 14** display trends in the contribution of each cause of death to the life expectancy gap for males and females, in years.
+**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for males and females, in years.
 The blue line illustrates the mean of the smoothed contribution estimated for each state-year, and illustrates the heterogeneity across states in their contribution trends.
@@
 Thus, the contribution of CVD to the life expectancy gap decreased by more than a year in several states (**Figure 4**).
-Of note, some states experienced a period of early increasing contributions, most notably, Arizona, Washington, and Oregon, that was not captured by the general pattern.
-Furthermore, Wisconsin and Washington DC both exhibit increasing contributions over time, as mortality gains among white females outpaced those among black females.
+Of note, some western states experienced a period of early increasing contributions that was not captured by the general pattern.
+Furthermore, Wisconsin, Michigan, and Washington DC both exhibit increasing contributions over time, as mortality gains among white females outpaced those among black females.
 
@@
 
-For all races and sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**).
+For both races and both sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**).
 Among males, this inverted U-shape pattern was accentuated among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**).
-Notable exceptions to this general pattern include Washington DC and Mississippi, states that underwent increasing cancer contributions over time, and New Mexico and Rhode Island, exhibiting decreasing contributions over time (although imprecisely estimated).
+Notable exceptions to this general pattern include Washington DC and Wisconsin, states that had higher initial contributions and underwent increasing cancer contributions over time, and New York and Massechusetts, exhibiting sharper decreases in cancer contributions.
 
-For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern is similar between blacks and whites and most states do not exhibit the accentuated period of increase in mortality that was estimated for black males.
-Thus, the contribution of cancer to the life expectancy gap is generally flat over time (**Figure 6**).
+For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern in mortality is similar between blacks and whites resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**).
 [You could potentially stop here to save space and just include the comment about there being little inequality in cancer mortality ].
-Washington DC deviates from this pattern, and exhibited increasing cancer contributions to the gap, while Rhode Island exhibited the reverse.
-Most notably, some states exhibited no inequality in cancer mortality (**eFigures 3 and 4**), including New York, and Connecticut (and for males, these states recently eliminated the black-white difference in cancer mortality).
-Furthermore, Massachusetts and (imprecisely) Rhode Island exhibit lower cancer rates among black women and men compared to whites in 2013.
+Washington DC deviates by exhibiting increasing cancer contributions to the gap, while New York, Connecticut, exhibit no inequality in cancer mortality, while Massachusetts exhibits lower cancer rates among blacks (**eFigure 4**).
 
@@
 
+Lit review
+
+Life Expectancy generally
+The spotlight on racial disparities in the US has intensified.
+Several recent papers examining US mortality by region, cause of death, or within population subgroups have received considerable attention while ignoring important black-white disparities.2, 3, 4 These disparities remain profound, and no previous paper has examined cause-specific contributions by state over an extended time period.
+Our analysis provides an important contribution to the literature by highlighting a high degree of heterogeneity in state-specific performance over time in eliminating these inequalities.
+from above:
+2. Chetty R, Stepner M, Abraham S, Lin S, Scuderi B, Turner N, Bergeron A, Cutler D. The association between income and life expectancy in the United States, 2001-2014. JAMA. 2016. 315(16):1750-1766.
+
+3.
+Dwyer-Lindgren L, Bertozzi-Villa A, Stubbs RW, Morozoff C, Kutz MJ, Huynh C, Barber RM, et al.
+US County-Level Trends in Mortality Rates for Major Causes of Death, 1980-2014.
+JAMA.
+2016.
+316(22):2385-2401.
+
+4. Case A, Deaton A. Rising morbidity and mortality in midlife among white non-Hispanic Americans in the 21st century. Proc Natl Acad Sci. 2015. 112(49):15078-83.
+
+Cancer
+
+Oliver. CEBP. 2008. Recent trends in BW disparities in Cancer Mortality. Note: higher mort rates in their paper likely because they restrict to >35yo.
+-BW diff in lung cancer mort mimics trends in overall mort rate in males. In females, no diff in lung cancer (although minor difference in other cancer related to smoking)
+-increasing disparity is seen in colorectal cancer in males between B and W
+-hump also seen in prostate cancer, especially among blacks, but falling mort rates since mid 1990s.
+-among females, increasing disparity in mort rates from colorectal cancer and breast cancer since 1975 (beginning of their study)
+-not sure what explains the inverted-U in cancer mort in our data among females - this CEBP people doesn't seem to shed light on this.
+
+Screening - when is it proven effective vs. when does it affect the recorded COD? Corinne to review that literature.
+-breast cancer: kills mostly among those 70+ yo.
+
+-trends in incidence may be related to overdx (prostate cancer, for ex) - for prostate cancer don't know whether the decline is due to improvements in treatment or due to screening -- could be both or just one
+
+https://nccd.cdc.gov/uscs/toptencancers.aspx#White
+-black women have higher rates of breast cancer than white women (28/100000 vs 20 for mortality)
+-blacks have higher rates of prostate cancer than whites (39.1 vs 18)
+
+Smoking and lung cancer
+see: https://www.washingtonpost.com/news/to-your-health/wp/2015/11/12/smoking-among-u-s-adults-has-fallen-to-historic-lows-these-7-charts-show-who-still-lights-up-the-most/?utm_term=.f2b261e63acf -- uses CDC data for citation
+-smoking rates have fallen steadily over the course of the study time period, but homogeneity in smoking still exists. used to be 42.4% in 1965 and down to 16.8% in 2014.
+-by race: nearly same for blacks and whites - 17.5% in blacks vs. 18.2% in whites, but 11.2% in hispanics (which are mixed into BW designation in our study)
+-by geography: highest in the midwest (20.7%), then south (17%), NE (15.3%), then west (13%), within heterogeneity within state (
+Dwyer-Lindgren  L, Mokdad  AH, Srebotnjak  T, Flaxman  AD, Hansen  GM, Murray  CJ.  Cigarette smoking prevalence in US counties: 1996-2012.  Popul Health Metr. 2014;12(1):5.)
+
+Obesity and...
+-high obesity in the south - ref 9
+-women and blacks have highest rates of obesity - ref 8
+
+Opiods
+https://www.cdc.gov/nchs/products/databriefs/db273.htm - Although rates increased 63% for non-Hispanic black persons (from 7.5 per 100,000 in 1999 to 12.2 in 2015) and 43% for Hispanic persons (from 5.4 in 1999 to 7.7 in 2015), non-Hispanic white persons experienced the greatest increase between 1999 and 2015.
+The age-adjusted rate of drug overdose deaths among non-Hispanic white persons increased 240% from 6.2 per 100,000 in 1999 to 21.1 in 2015.
+
+Centers for Disease Control and Prevention.
+Vital signs: overdoses of prescription opioid pain relievers and other drugs among women—United States, 1999–2010.
+MMWR Morb Mortal Wkly Rep.
+2013;62(26):537–542: opioid-related mortality increased 415% among women and 265% among men between 1999 and 2010.38
+
+Paulozzi LJ, Xi Y.
+Recent changes in drug poisoning mortality in the United States by urban-rural status and by drug type.
+Pharmacoepidemiol Drug Saf.
+2008;17(10):997–1005.: in 1999, large central metropolitan areas had the highest opioid-related mortality rates, and noncore areas had the lowest rates; by 2004, noncore areas had the highest rates and had seen the largest relative increase during that time period.50
+
+Paulozzi LJ, Ryan GW.
+Opioid analgesics and rates of fatal drug poisoning in the United States.
+Am J Prev Med.
+2006;31(6):506–511: drug poisoning mortality rates varied 8-fold between states, as did opioid consumption (methadone, 13-fold; oxycodone, 7-fold; overall, 4-fold).17
+
+Centers for Disease Control and Prevention.
+Vital signs: overdoses of prescription opioid pain relievers—United States, 1999–2008.
+MMWR Morb Mortal Wkly Rep.
+2011;60(43):1487–1492.: Another study showed that prescription opioid sales ranged from 3.7 (Illinois) to 12.6 (Florida) kilograms per 10,000 population, and opioid-related mortality rates ranged from 5.5 (Nebraska) to 27 (New Mexico) per 100,000 population.2
+
+Strengths
+
 Limitations
 -does not account for migration. Washington DC - have a lot of migration of high SES white folks coming in, and perhaps also relatively high SES black folks moving out
+-see the migration literature sent by Sam
 -does not account for hispanic ethinicity - demographic shifts in FL for example of black and white hispanics which will pull LE and disparities

### 2017-04-11 Manuscript/Manuscript_COD_Trends.Rmd
more lit

@@
 ---
-title: 'Historical trends in the contribution of six major causes of death to the black-white life expectancy gap by US state'
+title: Historical trends in the contribution of six major causes of death to the black-white
+  life expectancy gap by US state
 bibliography: bibliography.bib
 output:
+  word_document: default
   html_document: default
-  word_document: default
 always_allow_html: yes
@@
 For both races and both sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**).
-Among males, this inverted U-shape pattern was accentuated among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**).
+Among males, the peak mortality rate was much higher among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**).
 Notable exceptions to this general pattern include Washington DC and Wisconsin, states that had higher initial contributions and underwent increasing cancer contributions over time, and New York and Massechusetts, exhibiting sharper decreases in cancer contributions.
@@
 
+BW disparities (period-based studies)
+- "higher death rates due to heart disease, cancer, homicide (men), diabetes (women), and perinatal conditions" - kochanek et al 2013, harper et al 2007; see also Orsi 2010
+- harper 2012; 
+- Macinko and Elo (2009): decline in disparity among working-aged people since late 1980s.
+
 Life Expectancy generally
@@
 4. Case A, Deaton A. Rising morbidity and mortality in midlife among white non-Hispanic Americans in the 21st century. Proc Natl Acad Sci. 2015. 112(49):15078-83.
+
+CVD
+Why did NY and Mass do so well at decreasing the disparity? Why did Wisconsin and Michigan do poorly?
 
@@
 Dwyer-Lindgren  L, Mokdad  AH, Srebotnjak  T, Flaxman  AD, Hansen  GM, Murray  CJ.  Cigarette smoking prevalence in US counties: 1996-2012.  Popul Health Metr. 2014;12(1):5.)
+
+-Delancey et al (2008) identified more rapid decreases in smoking related cancers among black males since 1990 as contributing to the decrease in the overall gap.
 
@@
 
+Policy reflections
+-CVD still contributes the most to the currently LE gap.
+Firebaugh et al (2014) say that policy needs to focus on treatment once CVD is realized, not prevention ("The incidence component is negligible, indicating that blacks and whites are equally likely to succumb to heart disease.
+Heart disease nevertheless account for one quarter of the gap in 2015 because black victims tends to be younger than white victims.
+Iterventions should focus on identifying heart disease and equalizing treatment.
+Same goes for Cancers, non-comm diseases.
+Contrast with HIV/AIDS, prevention is key (since diff is in "incidence", not age at death with comm disease)"
+  - can we see this in our age-cod breakdown?/are these conclusions supported there - something about treatment vs. prevention?
+
+-should the focus be prevention of specific condictions/diseases, treatment of those conditions/diseases, or more distal --- SES differences - edu, employment, income (Hummer Du Bois review argues we need to focus on SES distal exposures, while Firebaugh 2014 differentiates whhen to focus on treatmnet of condition vs.
+prevention of the condition)
+
+-medicaid coverage varies among states.
+MS and NY have longer black LE than general trend and also have expanded medicare coverage (Bharmal 2012 citing the henry kaiser family foundation).
+CA NJ in middle for LE and coverage, AL TX lower for both -- so there is this observed ecological association
+
 Strengths
@@
 -see the migration literature sent by Sam
--does not account for hispanic ethinicity - demographic shifts in FL for example of black and white hispanics which will pull LE and disparities
+-does not account for hispanic ethinicity - demographic shifts in FL, CA, TX for example of black and white hispanics which will pull LE and disparities
+-high level of mexican immigration since 1965
+
+Problems specific to DC
+-migration's affect on LE over the course of the time period
+
+-migration - has been on the decline since the 1980s (Molloy 2011, 2017).
+migration happens more in people with bachelor's degrees in their early twenties (and declines thereafter) but is at a bit of a higher rate throughout lifetime than those without degree (Kaplan 2017)
+-roughly 1.5% of the popn moves between census regions per year, and 1.3% move within states in a census region.
+
+helpful: https://www.nytimes.com/2014/08/16/upshot/mapping-migration-in-the-united-states-since-1900.html
+-"western expansion", movement of souther blacks to northern states, and migration out of rust belt.
+-DC - influx in immigrants since 1980s, huge migration from other US states (NY, MD then Virginia), but also huge outflux to MD and Virginia. Should almost consider these three places as one.
+-lots of migraiton into MD
+-NY - 24% are immigrants
+-Texas has influx of immigrants to 17%
+-lots of migration into virginia, as well as immigration
+-Florida is the state most known for in-migration, both from states (Northeast particularly NY) and from outside the US
+-Alabama gets migrants from other southern states
+-Connecticut - influx in those immigrating from outside US since 1990s
 

### 2017-04-12 Manuscript/Manuscript_COD_Trends.Rmd

@@
 However, white males underwent relatively larger mortality decreases than black males, resulting in a modest increase to the contribution of CVD to the life expectancy gap for most states (**Figure 3**).
-In particular, Pennsylvania, Michigan, Wisconsin, and Washington DC experienced relatively larger increases to their CVD contributions to the gap for males, while some states such as Illinois and Kentucky exhibited relatively larger early contributions in the 1970s.
-
-The situation was reversed for females, with black females experiencing relatively larger mortality decreases over time compared to white females.
-Thus, the contribution of CVD to the life expectancy gap decreased by more than a year in several states (**Figure 4**).
-Of note, some western states experienced a period of early increasing contributions that was not captured by the general pattern.
-Furthermore, Wisconsin, Michigan, and Washington DC both exhibit increasing contributions over time, as mortality gains among white females outpaced those among black females.
+For females, black women experienced relatively larger mortality decreases over time compared to white females, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**).
+For both genders, Michigan, Wisconsin, California, and DC exhibit recent contributions that are relatively larger than the general pattern, while Massachusettes and New York exhibit persistently lower levels of CVD contributions.
 
@@
 Among males, the peak mortality rate was much higher among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**).
-Notable exceptions to this general pattern include Washington DC and Wisconsin, states that had higher initial contributions and underwent increasing cancer contributions over time, and New York and Massechusetts, exhibiting sharper decreases in cancer contributions.
-
 For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern in mortality is similar between blacks and whites resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**).
-[You could potentially stop here to save space and just include the comment about there being little inequality in cancer mortality ].
-Washington DC deviates by exhibiting increasing cancer contributions to the gap, while New York, Connecticut, exhibit no inequality in cancer mortality, while Massachusetts exhibits lower cancer rates among blacks (**eFigure 4**).
+For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among males, but not females), while New York and Massachusettes underwent large decreases in the amount that cancer contributed to the life expectancy gap.
 
@@
 
-Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**Figure X**).
+Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**eFigure X**).
 Recently, there has been a slight decrease in mortality rates, especially among black males.
-These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states.
-As such, the contribution of non-communicable disease to the mortality gap was general flat among men with a recent decrease, and exhibited a gradual decrease over time among women.
+These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure Y**).
+As such, the contribution of non-communicable disease to the mortality gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**).
+As with previous causes, New York and Massachusettes exhibit decreasing contributions, and Nevada has also made substantial progress.
+Furthermore, while DC's mortality rates decreased, they still have the highest contribution of non-communicable disease to the gap.
 
@@
 Injury is one of the least common causes of death.
-Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime in the 1990s resulted in heightened mortality from injuries among black males at that time (**Figure X**).
+Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**Figure X**).
 For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**Figure X**).
-Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, especially among females.
-Thus, the contribution of injuries to the mortality gap has generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures X and X**).
-Among males, Wisconsin and Washington DC diverged the most from the general pattern, as both states experienced steep increases in the contribution of injuries during the 1990s that was not captured by the general trend.
-Washington DC also exhibited this pattern among females.
+Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white females demonstrating sharper increases.
+Thus, the contribution of injuries to the mortality gap has generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures 9 and 10**).
+Among males, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions.
+Among females, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.
 
@@
 This spike was greater among males, especially blacks, and within states with larger urban centers.
-Thus, the contribution of communicable disease to the life expectancy gap exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics.
+Thus, the contribution of communicable disease to the life expectancy gap exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**).
+DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among females.
 
@@
 
+**Summary of findings**
+
+- In this paper, we examined state-level trends in six major CODs' contributions to the black-white LE gap.
+- While we found that many state-level trends followed the same pattern of COD-specific contributions, several states were divergent.
+Furthermore, different patterns emerged for males compared to females.
+-
+
+*Literature review*
+
+*Strengths and weaknesses*
+
+*Conclusions*
+
+- DC exhibits the highest contributions for many causes, but they have the greatest gap and a widening gap - so more room for each cause to contribute.
+Goes to say that all causes are contributing to the gap (it isn't being driven by one cause in particular)
+
 Lit review
@@
 ###Appendix
+**S1 Table: Cause of death groupings**
+
+| Broad cause of death     | Included causes of death                                                     |
+|--------------------------|------------------------------------------------------------------------------|
+| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer.
+|
+| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries.
+|
+| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
+| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Syndrome and Nephrosis.
+|
+| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
+| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death.
+|
 

### 2017-04-13 Code/old_3_Explore_Missingness.Rmd
renamed files

---
title: "3_Explore_Missingness"
author: "Corinne Riddell"
date: "1/19/2017"
output: github_document
---

```{r load_data}
library(tidyverse)
library(forcats)
dat.clean<- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.csv", header = T)
```

###Initial graphs on missing data
Notes on Figure 1 (below):

- Figure 1 is sorted by size of the black population in the state in 2013. This is so you can see how missingness is a function of population size. We expect that states with less black people will have more missing data and more counts of exactly zero. Thus, states with the biggest populations (of blacks and whites) will "hold constant" the effect of population size on the amount of missing data.

- The first panels you see in Figure 1 (below) are for the states with the smallest black populations. For whites, the amount of missing data increases in age until about 40 years old, and then decreases. For blacks, the amount of missing data is very low across age. This seemed strange to me at first, but then I remembered that the data is not missing if the count is equal to 0, which will be the case for many strata in these states with the lowest black population.

- As you read left to right (as the black population size increases by state), the trends in missing data across age look more similar by race, although blacks appear to have more missing data in individuals < 30 years old for the states with the most black people.

**Figure 1: Proportion missing by State (panel) and Age group (x-axis), ordered by increasing size of black population in 2013**
```{r bystate, fig.width = 12, fig.height = 37, echo=F}
bystate <- 
  dat.clean %>%
  group_by(state, race, age) %>%
  summarise(prop.death.missing = mean(is.na(deaths)),
            age_minbin = first(age_minbin))

#recode a new count variable that "counts the missing data" so that these rows are used when calculating the % of data == 0.
dat.clean$death.na1 <- dat.clean$deaths
dat.clean$death.na1[is.na(dat.clean$deaths) == T] <- 1

bystate2 <- 
  dat.clean %>%
  group_by(state, race, age) %>%
  summarise(prop.death.zero = mean(death.na1 == 0))

bystate <- merge(bystate, bystate2, by=c("state", "race", "age"))
rm(bystate2)

#calculate the size of the black population in 2013 by State
bystate.bpop.2013 <- 
  dat.clean %>%
  filter(COD == "Cancers" & year == 2013 & race == "Black") %>% #to stop repeats of the population size
  group_by(state) %>%
  summarize(pop_by_race_2013 = sum(as.numeric(population)))

#calculate the size of the black population in 1969 by State
bystate.bpop.1969 <- 
  dat.clean %>%
  filter(COD == "Cancers" & year == 1969 & race == "Black") %>% #to stop repeats of the population size
  group_by(state) %>%
  summarize(pop_by_race_1969 = sum(as.numeric(population)))

bystate <- merge(bystate, bystate.bpop.2013, by = "state", all.x = T)
bystate <- merge(bystate, bystate.bpop.1969, by = "state", all.x = T)
bystate$state_reorder <- reorder(bystate$state, bystate$pop_by_race_2013)

ggplot(bystate, aes(x = fct_reorder(age, age_minbin), y = prop.death.missing)) + 
  geom_point(aes(col=race)) + 
  facet_wrap(~ state_reorder, ncol = 4) +
  theme_bw(base_size = 11) +
  theme(legend.position="top") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Missing (%)") +
  xlab("Age group")
```

**Examine the growth of the black population between 1969 and 2013**

```{r}
ggplot(subset(bystate, age == "<1 year"), aes(pop_by_race_1969, pop_by_race_2013)) + 
  geom_point() + 
  geom_abline(intercept = 0, slope = 1) + 
  xlab("Black population in 1969") + 
  ylab("Black population in 2013")

unique(bystate$state[bystate$age == "<1 year" & bystate$pop_by_race_1969 < 10000])
```

Notes on Figure 2 (below):

- Figure 2 examines the proportion of counts equal to zero. The purpose of this examination is to visually confirm that there are higher proportions of zero counts in the states with the least amount of blacks. This is very clear when viewing the figure.

- Question - any explantion for why the portion of zeroes stays relatively higher in blacks vs. whites in the states with the most blacks (within the younger age groups)?

**Figure 2: Proportion zero by State (panel) and Age group (x-axis), ordered by increasing size of black population in 2013**

```{r zero_by_state, fig.width = 12, fig.height = 37, echo=F}
ggplot(bystate, aes(x = fct_reorder(age, age_minbin), y = prop.death.zero)) + 
  geom_point(aes(col = race)) + facet_wrap(~ state_reorder, ncol = 4) +
  theme_bw(base_size = 11) +
  theme(legend.position = "top") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion zero death count (%)") +
  xlab("Age group") 
```

###Other plots (less interesting/informative)

**How is missingness related to age, race, and sex?**

```{r missing_data, fig.width = 10, fig.height = 5, echo=F}
table1.whitemale <- round(prop.table(table(is.na(dat.clean$deaths[dat.clean$race.sex == "White Male"]), 
                                           dat.clean$age[dat.clean$race.sex == "White Male"]),2)*100,2)

table1.blackmale <- round(prop.table(table(is.na(dat.clean$deaths[dat.clean$race.sex == "Black Male"]), 
                                           dat.clean$age[dat.clean$race.sex == "Black Male"]),2)*100,2)

table1.whitefemale <- round(prop.table(table(is.na(dat.clean$deaths[dat.clean$race.sex == "White Female"]), 
                                           dat.clean$age[dat.clean$race.sex == "White Female"]),2)*100,2)

table1.blackfemale <- round(prop.table(table(is.na(dat.clean$deaths[dat.clean$race.sex == "Black Female"]), 
                                           dat.clean$age[dat.clean$race.sex == "Black Female"]),2)*100,2)

miss1 <- data.frame("white.male" = table1.whitemale[2, ], 
                    "black.male"= table1.blackmale[2, ],
                    "white.female" = table1.whitefemale[2, ], 
                    "black.female"= table1.blackfemale[2, ],
                    "age" = names(table1.whitemale[2, ]))

miss1$age <- factor(miss1$age, levels(miss1$age)[c(1, 2, 11, 3:10, 12:20)])

miss1 <- gather(miss1, race.sex, missingness, c(white.male, black.male, white.female, black.female))

ggplot(miss1, aes(x = age, y = missingness)) + geom_point() + geom_point(aes(col=race.sex)) +
  theme_bw(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Missing (%)") +
  xlab("Age group")
```

**Let's look at the proportion with 0's**

```{r zero_data, fig.width = 10, fig.height = 5, echo=F}
table2.whitemale <- round(prop.table(table(dat.clean$deaths[dat.clean$race.sex == "White Male"]==0, 
                                       dat.clean$age[dat.clean$race.sex == "White Male"]),2)*100,2)

table2.blackmale <- round(prop.table(table(dat.clean$deaths[dat.clean$race.sex == "Black Male"]==0, 
                                       dat.clean$age[dat.clean$race.sex == "Black Male"]),2)*100,2)

table2.whitefemale <- round(prop.table(table(dat.clean$deaths[dat.clean$race.sex == "White Female"]==0, 
                                       dat.clean$age[dat.clean$race.sex == "White Female"]),2)*100,2)

table2.blackfemale <- round(prop.table(table(dat.clean$deaths[dat.clean$race.sex == "Black Female"]==0, 
                                       dat.clean$age[dat.clean$race.sex == "Black Female"]),2)*100,2)

miss2 <- data.frame("whitemale" = table2.whitemale[2, ], 
                    "blackmale"= table2.blackmale[2, ],
                    "whitefemale" = table2.whitefemale[2, ], 
                    "blackfemale"= table2.blackfemale[2, ],
                    "age" = names(table2.blackmale[2, ]))

miss2$age <- factor(miss2$age, levels(miss2$age)[c(1, 2, 11, 3:10, 12:20)])

miss2 <- gather(miss2, racesex, zeroes, c(whitemale, blackmale, whitefemale, blackfemale))

ggplot(miss2, aes(x = age, y = zeroes)) + geom_point() + geom_point(aes(col=racesex)) +
  theme_bw(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Zero (%)") +
  xlab("Age group")
```

**Missing or zero (which is equivalent to < 10)**

```{r miss_or_zero_data, fig.width = 10, fig.height = 5, echo=F}
dat.clean$miss_or_zero <- F
dat.clean$miss_or_zero[dat.clean$deaths == 0 | is.na(dat.clean$deaths) == T] <- T

table3.whitemale <- round(prop.table(table(dat.clean$miss_or_zero[dat.clean$race.sex == "White Male"]==T, 
                                      dat.clean$age[dat.clean$race.sex == "White Male"]),2)*100,2)

table3.blackmale <- round(prop.table(table(dat.clean$miss_or_zero[dat.clean$race.sex == "Black Male"]==T, 
                                       dat.clean$age[dat.clean$race.sex == "Black Male"]),2)*100,2)

table3.whitefemale <- round(prop.table(table(dat.clean$miss_or_zero[dat.clean$race.sex == "White Female"]==T, 
                                       dat.clean$age[dat.clean$race.sex == "White Female"]),2)*100,2)

table3.blackfemale <- round(prop.table(table(dat.clean$miss_or_zero[dat.clean$race.sex == "Black Female"]==T, 
                                       dat.clean$age[dat.clean$race.sex == "Black Female"]),2)*100,2)

miss3 <- data.frame("whitemale" = table3.whitemale[2, ], 
                    "blackmale"= table3.blackmale[2, ],
                    "whitefemale" = table3.whitefemale[2, ], 
                    "blackfemale"= table3.blackfemale[2, ],
                    "age" = names(table3.blackmale[2, ]))

miss3$age <- factor(miss3$age, levels(miss3$age)[c(1, 2, 11, 3:10, 12:20)])

miss3 <- gather(miss3, racesex, miss_or_zero, c(whitemale, blackmale, whitefemale, blackfemale))

ggplot(miss3, aes(x = age, y = miss_or_zero)) + geom_point() + geom_point(aes(col=racesex)) +
  theme_bw(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Missing or Zero (%)") +
  xlab("Age group")
```

```{r explore_data2, fig.width = 10, fig.height = 5, echo=F, eval=F}
summary(dat.clean$deaths)

ggplot(subset(dat.clean, deaths>0), aes(x = deaths)) + 
  geom_histogram(binwidth=100)

ggplot(subset(dat.clean, state=="Louisiana"), aes(x = population, y = deaths)) + 
  geom_point(aes(col=race.sex), alpha=0.2) +
  facet_wrap( ~ COD) + theme_bw()

```

### 2017-04-13 Manuscript/Manuscript_COD_Trends.Rmd
words

@@
 output:
+  html_document: default
   word_document: default
-  html_document: default
 always_allow_html: yes
@@
 
+We also calculated age-standardized mortality rates for year-state-race-sex stratum, to aid the interpretation of the decomposition results.
+We used the US standard population from the year 2000 [ref].
+
 **Description of the trend in each cause of death's contribution to the gap and identification of divergent states**
 
-To describe the general pattern in these trend lines we used random effects meta regression.
+We used random effects meta regression to describe the general pattern in these trend lines.
 Each state's mean contribution was regressed against a linear spline for calendar time that was estimated using four internal knots.
@@
 To identify divergent states, we compared the state-specific trends (estimated separately for each state using the Bayesian model) to the general pattern estimated using meta regression.
-
-*Need to add information on the trends in mortality* -- where?
 
@@
 Injury is one of the least common causes of death.
-Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**Figure X**).
-For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**Figure X**).
+Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure X**).
+For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure X**).
 Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white females demonstrating sharper increases.
@@
 - In this paper, we examined state-level trends in six major CODs' contributions to the black-white LE gap.
+
 - While we found that many state-level trends followed the same pattern of COD-specific contributions, several states were divergent.
-Furthermore, different patterns emerged for males compared to females.
--
+For every causes of death except communicable disease, Massachusettes and New York exhibited steep declines to the contribution of each cause to the life expectancy gap, implying that decreasing inequality across a broad range of causes has results in eliminating the life expectancy gap in those states.
+
+- On the other hand, DC underwent an increasing life expectancy gap over time and has the largest life expectancy difference of all the states. This analysis
 

### 2017-04-13 Manuscript/Manuscript_COD_Trends.Rmd
stacked line charts

@@
 cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
+
+cod_decomp_results$COD_rev <- factor(cod_decomp_results$COD, levels = c("Cardiovascular", "Cancers", "All other causes",
+                                                                        "Communicable", "Non-communicable",  "Injuries"))
 ```
@@
 }
+```
+
+**Men**
+
+```{r stacked_trends_male, echo=F, fig.height=9, fig.width=11}
+ggplotly(ggplot(subset(cod_decomp_results, sex == "Male"), aes(x = year, y = COD_cont_yrs_mean)) + 
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
+  geom_area(aes(fill = COD_rev), stat = "identity") +
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (%)") +
+  xlab("Year (1969-2013)"))
+
+ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
+  geom_area(aes(fill = COD_rev), stat = "identity") +
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (%)") +
+  xlab("Year (2009-2013)"))
+```
+
+**Women**
+
+```{r stacked_trends_female, echo=F, fig.height=9, fig.width=11}
+ggplotly(ggplot(subset(cod_decomp_results, sex == "Female"), aes(x = year, y = COD_cont_yrs_mean)) + 
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
+  geom_area(aes(fill = COD_rev), stat = "identity") +
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (%)") +
+  xlab("Year (1969-2013)"))
+
+ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
+  geom_area(aes(fill = COD_rev), stat = "identity") +
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (%)") +
+  xlab("Year (2009-2013)"))
 ```

### 2017-04-13 Manuscript/Manuscript_COD_Trends.Rmd
moar

--- 2017-04-13 Manuscript/Manuscript_COD_Trends.Rmd
@@
 }
+```
+
+**Men**
+
+```{r stacked_trends_male, echo=F, fig.height=9, fig.width=11}
+plot1 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male"), aes(x = year, y = COD_cont_yrs_mean)) + 
+  geom_area(aes(fill = COD_rev), stat = "identity") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
+  geom_line(data = subset(BlackWhite_results, sex == "Male"), aes(x = year, y = LE_wbgap_mean)) +
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (%)") +
+  xlab("Year (1969-2013)"))
+
+for(i in 1:400){
+  if(plot1$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
+    plot1$x$data[[i]]$line$width <- 1
+  }
+}
+
+plot1
+
+plot2 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
+  geom_area(aes(fill = COD_rev), stat = "identity") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
+  geom_line(data = subset(BlackWhite_results, sex == "Male" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (%)") +
+  xlab("Year (2009-2013)"))
+
+for(i in 1:400){
+  if(plot2$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
+    plot2$x$data[[i]]$line$width <- 1
+  }
+}
+
+plot2
+```
+
+**Women**
+
+```{r stacked_trends_female, echo=F, fig.height=9, fig.width=11}
+plot3 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female"), aes(x = year, y = COD_cont_yrs_mean)) + 
+                      geom_area(aes(fill = COD_rev), stat = "identity") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
+  geom_line(data = subset(BlackWhite_results, sex == "Female"), aes(x = year, y = LE_wbgap_mean)) +
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (%)") +
+  xlab("Year (1969-2013)"))
+
+for(i in 1:400){
+  if(plot3$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
+    plot3$x$data[[i]]$line$width <- 1
+  }
+}
+
+plot3
+
+plot4 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
+                      geom_area(aes(fill = COD_rev), stat = "identity") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
+  geom_line(data = subset(BlackWhite_results, sex == "Female" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (%)") +
+  xlab("Year (2009-2013)"))
+
+for(i in 1:400){
+  if(plot4$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
+    plot4$x$data[[i]]$line$width <- 1
+  }
+}
+
+plot4
 ```
@@
 
-- While we found that many state-level trends followed the same pattern of COD-specific contributions, several states were divergent.
-For every causes of death except communicable disease, Massachusettes and New York exhibited steep declines to the contribution of each cause to the life expectancy gap, implying that decreasing inequality across a broad range of causes has results in eliminating the life expectancy gap in those states.
+- While we found that many state-level trends followed the same pattern of COD-specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result.
+New York and Massachusettes have virtually eliminated the black-white life expectancy gap for both men and women.
+While these states showed contribution decreases for all major causes of death except communicable disease, both states began with only a minimal contribution of CVD to the mortality gap.
 
-- On the other hand, DC underwent an increasing life expectancy gap over time and has the largest life expectancy difference of all the states. This analysis
+- How do these results compare with states with large black populations: Georgia, Florida, Texas, and California?
+In all of these states, deaths during the first year of life are relatively higher in blacks and in most cases, deaths from CVD during mid-life is also higher among blacks.
+California and Florida black young men are dying more of injuries, and black men and women in Florida are dying more of communicable disease.
+
+- These findings can be contrasted with those of Firebaugh et al (2014).
+These authors distinguish between inequality caused by a difference in CVD mortality incidence vs.
+that caused by a difference in age-at-death and find that while blacks and whites are equally likely to die of heart disease, blacks are dying at younger ages.
+[qu: how can the equally likely to die of Heart disease part be true when we have the age std mort rates showing otherwise?
+[*See page 27 of firebaugh - highlighted section - in dropbbox papers directory*].
+
+*Notes of Firebaugh 2014 paper:*
+-CVD still contributes the most to the current LE gap.
+Firebaugh et al (2014) say that policy needs to focus on treatment once CVD is realized, not prevention ("The incidence component is negligible, indicating that blacks and whites are equally likely to succumb to heart disease.
+Heart disease nevertheless account for one quarter of the gap in 2015 because black victims tends to be younger than white victims.
+Interventions should focus on identifying heart disease and equalizing treatment.
+Same goes for Cancers, non-comm diseases.
+Contrast with HIV/AIDS, prevention is key (since diff is in "incidence", not age at death with comm disease)"
+[*Does this reasoning make sense to you?
+I'm not sure I follow why a diff in age at death would imply focusing on treatment rather than prevention -- what if blacks get CVD at younger ages than whites?
+Doesn't focusing on equalizing treatment assume that blacks and whites have the same timing of onset?*]
 
@@
 
+BW disparities (period-based studies)
+- "higher death rates due to heart disease, cancer, homicide (men), diabetes (women), and perinatal conditions" - kochanek et al 2013, harper et al 2007; see also Orsi 2010
+- harper 2012; 
+- Macinko and Elo (2009): decline in disparity among working-aged people since late 1980s.
+
 *Strengths and weaknesses*
+
+We used data over 45 years, across two race/ethnicities, and for all age groups.
+By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year.
+This allowed us to recognize when the estimated contributions for a particular cause were relatively low *because they started low* vs.
+low *because they trended that way*.
+Furthermore, our analysis was centred on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
+This is especially important in the light of recent studies that have recieved criticism for narrowly focusing on a specific age which may present a distorted reality [Case & Deaton, Minton IJE 2016, ].
+Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.
+[Methodologically, using the Bayesian autoregressive model, allowed us to estimate mortality rates in strata that were sparsely populated...*something about methods?*]
+
+Our analysis has two main limitations.
+Firstly, we did not account for hispanic ethinicity, implying that both black and white populations in our study include some portion of hispanics that is known to differ over time and by state.
+We decided against excluding hispanics from this study because the indicator for ethnicity has been noted to have varying accuracy across states and during the earlier years of data.
+Thus, states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy and the contribution that are affected.
+However, we are somewhat reassured by a previous sensitivity analysis, where we found that accounting for hispanic ethinicity did not have a substantial effect on the results [ref Harper 2012].
+Secondly, our analysis did not account for the migration of individuals between states.
+Thus, states with black or white populations that have shifted over time will be affected.
+This limits the interpretation of results especially for Washington DC, as it has undergone substantial demographic shifts since X.
+
+Problems specific to DC
+-migration's effect on LE over the course of the time period
+
+-migration - has been on the decline since the 1980s (Molloy 2011, 2017).
+migration happens more in people with bachelor's degrees in their early twenties (and declines thereafter) but is at a bit of a higher rate throughout lifetime than those without degree (Kaplan 2017)
+-roughly 1.5% of the popn moves between census regions per year, and 1.3% move within states in a census region.
+
+helpful: https://www.nytimes.com/2014/08/16/upshot/mapping-migration-in-the-united-states-since-1900.html
+-"western expansion", movement of souther blacks to northern states, and migration out of rust belt.
+-DC - influx in immigrants since 1980s, huge migration from other US states (NY, MD then Virginia), but also huge outflux to MD and Virginia. Should almost consider these three places as one.
+-lots of migraiton into MD
+-NY - 24% are immigrants
+-Texas has influx of immigrants to 17%
+-lots of migration into virginia, as well as immigration
+-Florida is the state most known for in-migration, both from states (Northeast particularly NY) and from outside the US
+-Alabama gets migrants from other southern states
+-Connecticut - influx in those immigrating from outside US since 1990s
 
@@
 Goes to say that all causes are contributing to the gap (it isn't being driven by one cause in particular)
-
-Lit review
-
-BW disparities (period-based studies)
-- "higher death rates due to heart disease, cancer, homicide (men), diabetes (women), and perinatal conditions" - kochanek et al 2013, harper et al 2007; see also Orsi 2010
-- harper 2012; 
-- Macinko and Elo (2009): decline in disparity among working-aged people since late 1980s.
 
@@
 4. Case A, Deaton A. Rising morbidity and mortality in midlife among white non-Hispanic Americans in the 21st century. Proc Natl Acad Sci. 2015. 112(49):15078-83.
-
-CVD
-Why did NY and Mass do so well at decreasing the disparity? Why did Wisconsin and Michigan do poorly?
 
@@
 
-Policy reflections
--CVD still contributes the most to the currently LE gap.
-Firebaugh et al (2014) say that policy needs to focus on treatment once CVD is realized, not prevention ("The incidence component is negligible, indicating that blacks and whites are equally likely to succumb to heart disease.
-Heart disease nevertheless account for one quarter of the gap in 2015 because black victims tends to be younger than white victims.
-Iterventions should focus on identifying heart disease and equalizing treatment.
-Same goes for Cancers, non-comm diseases.
-Contrast with HIV/AIDS, prevention is key (since diff is in "incidence", not age at death with comm disease)"
-  - can we see this in our age-cod breakdown?/are these conclusions supported there - something about treatment vs. prevention?
-
 -should the focus be prevention of specific condictions/diseases, treatment of those conditions/diseases, or more distal --- SES differences - edu, employment, income (Hummer Du Bois review argues we need to focus on SES distal exposures, while Firebaugh 2014 differentiates whhen to focus on treatmnet of condition vs.
@@
 CA NJ in middle for LE and coverage, AL TX lower for both -- so there is this observed ecological association
-
-Strengths
-
-Limitations
--does not account for migration. Washington DC - have a lot of migration of high SES white folks coming in, and perhaps also relatively high SES black folks moving out
--see the migration literature sent by Sam
--does not account for hispanic ethinicity - demographic shifts in FL, CA, TX for example of black and white hispanics which will pull LE and disparities
--high level of mexican immigration since 1965
-
-Problems specific to DC
--migration's affect on LE over the course of the time period
-
--migration - has been on the decline since the 1980s (Molloy 2011, 2017).
-migration happens more in people with bachelor's degrees in their early twenties (and declines thereafter) but is at a bit of a higher rate throughout lifetime than those without degree (Kaplan 2017)
--roughly 1.5% of the popn moves between census regions per year, and 1.3% move within states in a census region.
-
-helpful: https://www.nytimes.com/2014/08/16/upshot/mapping-migration-in-the-united-states-since-1900.html
--"western expansion", movement of souther blacks to northern states, and migration out of rust belt.
--DC - influx in immigrants since 1980s, huge migration from other US states (NY, MD then Virginia), but also huge outflux to MD and Virginia. Should almost consider these three places as one.
--lots of migraiton into MD
--NY - 24% are immigrants
--Texas has influx of immigrants to 17%
--lots of migration into virginia, as well as immigration
--Florida is the state most known for in-migration, both from states (Northeast particularly NY) and from outside the US
--Alabama gets migrants from other southern states
--Connecticut - influx in those immigrating from outside US since 1990s
 

### 2017-04-14 Manuscript/Manuscript_COD_Trends.Rmd
migration

@@
 This allowed us to recognize when the estimated contributions for a particular cause were relatively low *because they started low* vs.
-low *because they trended that way*.
+low *because they had decreased over time*.
 Furthermore, our analysis was centred on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
-This is especially important in the light of recent studies that have recieved criticism for narrowly focusing on a specific age which may present a distorted reality [Case & Deaton, Minton IJE 2016, ].
+This is especially important in light of recent studies that have recieved criticism for narrowly focusing on a specific age group, which may present a distorted reality [Case & Deaton, Minton IJE 2016, ].
 Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.
@@
 Firstly, we did not account for hispanic ethinicity, implying that both black and white populations in our study include some portion of hispanics that is known to differ over time and by state.
-We decided against excluding hispanics from this study because the indicator for ethnicity has been noted to have varying accuracy across states and during the earlier years of data.
-Thus, states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy and the contribution that are affected.
-However, we are somewhat reassured by a previous sensitivity analysis, where we found that accounting for hispanic ethinicity did not have a substantial effect on the results [ref Harper 2012].
-Secondly, our analysis did not account for the migration of individuals between states.
-Thus, states with black or white populations that have shifted over time will be affected.
-This limits the interpretation of results especially for Washington DC, as it has undergone substantial demographic shifts since X.
+We decided against excluding or adjusting for hispanic ethnicity in this study because the indicator for ethnicity has been noted to have varying accuracy across states and during the earlier time period [ref].
+Thus, states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had hispanics been excluded [ref].
+In a previous sensitivity analysis, we found that accounting for hispanic ethinicity did not have a substantial effect on the results [ref Harper 2012], providing some evidence that not accounting for hispanic ethnicity should not substantially affect our findings.
 
-Problems specific to DC
--migration's effect on LE over the course of the time period
-
--migration - has been on the decline since the 1980s (Molloy 2011, 2017).
-migration happens more in people with bachelor's degrees in their early twenties (and declines thereafter) but is at a bit of a higher rate throughout lifetime than those without degree (Kaplan 2017)
--roughly 1.5% of the popn moves between census regions per year, and 1.3% move within states in a census region.
+Secondly, our analysis does not account for immigration, or migration between states.
+Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [ref Preston 2014].
+As we are examine the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white.
+Migration between states can also affect life expectancy.
+Migration is particularly important for the District of Columbia, because of it's small geographical bounds (it is easier to be considered migration when regions are smaller - how to say this?), and substantial in-migration from New York, in- and out-migration to contiguous Maryland and Virginia.
+Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are imigrating to DC.
+These changes would both serve to increase the racial gap in DC.
 
 helpful: https://www.nytimes.com/2014/08/16/upshot/mapping-migration-in-the-united-states-since-1900.html
--"western expansion", movement of souther blacks to northern states, and migration out of rust belt.
+-movement of southern blacks to northern states, and migration out of rust belt.
 -DC - influx in immigrants since 1980s, huge migration from other US states (NY, MD then Virginia), but also huge outflux to MD and Virginia. Should almost consider these three places as one.
@@
 -Connecticut - influx in those immigrating from outside US since 1990s
+
+Individuals who migrate are more likely to have completed a bachelor's degree than those who do not, suggesting that migration may contribute to heightened life expectancy.
+
+Thus, states with black or white populations that have shifted over time will be affected.
+This limits the interpretation of results especially for Washington DC, as it has undergone substantial demographic shifts since X.
+-foreign borns tend to have higher LE, so increase in immigration can lead to gains in LE in either blacks or whites, or both, depending on the country of origin - better because of migrant selectivity (select on traits related to good health) and better health (less likely to smoke or be obese or binge drink [ref], and return migration when they become sick)
+
+Problems specific to DC
+-migration's effect on LE over the course of the time period
+
+-migration - has been on the decline since the 1980s (Molloy 2011, 2017).
+migration happens more in people with bachelor's degrees in their early twenties (and declines thereafter) but is at a bit of a higher rate throughout lifetime than those without degree (Kaplan 2017)
+-roughly 1.5% of the popn moves between census regions per year, and 1.3% move within states in a census region.
 

### 2017-04-16 Code/3_combine_results.Rmd
move reorder as map to step3

--- 2017-04-10 Code/combine_results.Rmd
@@
 BlackWhite_results <- add_Census_Division(BlackWhite_results, state)
+BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")
+
 BlackWhite_results$stabbrs <- state.abb[match(BlackWhite_results$state, state.name)]
@@
                             cod_decomp_results, by = "stratum.id", all.y = T)
+cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
+
 ```

### 2017-04-17 ResearchDay/Riddell_ResearchDay.Rmd
research day

---
title: An interactive exploration of how changes in cause of death have narrowed the
  black-white life expectancy gap in the United States
author: "Corinne Riddell"
date: "April 16, 2017"
output: ioslides_presentation
---

```{r initialize, echo=FALSE, warning=F, message=F}
library(plotly)
library(viridis)
library(scales)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
library(splines)
library(broom)
library(tidyverse)

source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

```{r load_data, echo=F}
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
```

##Life expectancy

```{r life-expectancy-trends, echo=F}
LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")

LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")
```

##Life expectancy gap in men, 1969-2013

```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11, out.height="80%"}
LE_trend_plot_males +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)

```

##Life expectancy gap in women, 1969-2013

```{r plot_LE_trends_females, echo=F, fig.height=9, fig.width=11}
LE_trend_plot_females +
  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5)

```

---

```{r plot_both, echo=F}
#Informative: male and females at the same time
ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)"))
```

##Acknowledgements

```{r, echo=F}
library(png)
library(jpeg)
library(grid)
library(gridExtra)
img1 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/k_morrison.png")), interpolate = FALSE)
img2 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/s_harper2.png")), interpolate = FALSE)
img3 <-  rasterGrob(as.raster(readJPEG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/j_kaufman.jpg")), interpolate = FALSE)
grid.arrange(img1,
             img2,
             img3,
             ncol = 3)
```

----

```{r, echo=F, out.width="15%", dpi=72, fig.cap="Kathryn Morrison, PhD"}
knitr::include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/k_morrison.png")
```

```{r, echo=F, out.width="15%", dpi=72, fig.cap="Sam Harper, PhD"}
knitr::include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/s_harper.jpeg")
```

```{r, echo=F, out.width="15%", dpi=72, fig.cap="Jay Kaufman, PhD"}
knitr::include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/j_kaufman.jpg")
```

Brainstorming
=================

when do we use viz the most in academics?
-For figures in tables - that is, to show your main findings?

When else should we use vis?
-EDA (at the very beginning) - visualize relationships in the data, and you'll be more likely to catch coding errors you didn't even know you made
- you'll catch errors in the data itself (measurement problems), or funny things about how the data was coded
-Assumption checking (analysis stage)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2017-04-17 Manuscript/Manuscript_COD_Trends.Rmd

@@
 BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
-BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")
 
 cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
-cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
 
@@
 
+Read: racial disparitiy in where people from DC move:
+-https://ggwash.org/view/38326/theres-a-big-racial-disparity-in-where-people-born-in-dc-end-up-moving-or-not-moving
+
 *Conclusions*

### 2017-04-17 Manuscript/Manuscript_COD_Trends.Rmd

@@
 BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
-
+BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")
 cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
-
+cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
 cod_decomp_results$COD_rev <- factor(cod_decomp_results$COD, levels = c("Cardiovascular", "Cancers", "All other causes",

### 2017-04-17 ResearchDay/Riddell_ResearchDay.Rmd

@@
 
+library(png)
+library(jpeg)
+library(grid)
+library(gridExtra)
+library(knitr)
+
 source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
@@
 BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
+BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")
+
 cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
+cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
 ```
@@
 
-```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11, out.height="80%"}
-LE_trend_plot_males +
+```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11, out.height="90%", out.width = "90%"}
+LE_trend_plot_males <- LE_trend_plot_males + 
   geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)
+
+refine.patterns <- function(plot) {
+  for(i in 1:240){
+  if(plot$x$data[[i]]$line$color == "rgba(190,190,190,1)"){
+    plot$x$data[[i]]$line$width <- 0.7559055
+    plot$x$data[[i]]$hoverinfo <- "none"
+  }
+    if(plot$x$data[[i]]$line$color == "rgba(0, 0, 0, 1)"){
+    plot$x$data[[i]]$hoverinfo <- "none"
+  }
+  }
+  return(plot)
+}
+
+LE_trend_plot_males2 <- refine.patterns(ggplotly(LE_trend_plot_males))
+
+LE_trend_plot_males2
 
@@
 ```{r, echo=F}
-library(png)
-library(jpeg)
-library(grid)
-library(gridExtra)
 img1 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/k_morrison.png")), interpolate = FALSE)
@@
 ```{r, echo=F, out.width="15%", dpi=72, fig.cap="Kathryn Morrison, PhD"}
-knitr::include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/k_morrison.png")
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/k_morrison.png")
 ```
@@
 ```{r, echo=F, out.width="15%", dpi=72, fig.cap="Sam Harper, PhD"}
-knitr::include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/s_harper.jpeg")
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/s_harper.jpeg")
 ```
@@
 ```{r, echo=F, out.width="15%", dpi=72, fig.cap="Jay Kaufman, PhD"}
-knitr::include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/j_kaufman.jpg")
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/j_kaufman.jpg")
 ```

### 2017-04-25 Manuscript/Manuscript_COD_Trends.Rmd
changes

@@
 ###Discussion 
+- the next is to investigate potential state level policy and impact on underlying risk factors
+- our analysis highlittghs a few specific causes in a few specific period
+-injuries his cvd, not so much on cancer - because there is more homo. across contribution of cancer
+- limitations - major groupings but they have lots of heatergeneity of causes within the grouping
+- cont of HIV increases very dramatically, but the pace of decline differed - the places where there is a lag in the decline was the more rural places
 

### 2017-04-25 shiny_app/app.R

@@ -130,12 +130,12 @@ ui1 <- fluidPage(theme = shinytheme("cosmo"),
                                 
                                 conditionalPanel(condition = "input.tab == 'state.dashboard'",
                                                  selectInput(inputId = "state", label = "State:", 
-                                                             choices = levels(BlackWhite_results$state)))#,
-                                # absolutePanel(bottom = 0, left = 0, draggable = T,
-                                #   conditionalPanel(condition = "input.tab == 'state.dashboard'",
-                                #                    img(src = "COD.legend.png"))
-                                #   
-                                # )
+                                                             choices = levels(BlackWhite_results$state))),
+                                absolutePanel(bottom = 0, left = 0, draggable = T,
+                                  conditionalPanel(condition = "input.tab == 'state.dashboard'",
+                                                   img(src = "COD.legend.png"))
+
+                                )
                                 ),
                  
                    
@@ -807,10 +807,11 @@ output$age_cod1 <- renderPlot({
     ylab("Contribution to the life expectancy gap (years)") +
     xlab("") + ggtitle(paste0("Males, ", input$years_LEgap[1])) + 
     scale_y_continuous(limits = bounds())  + 
-    theme(legend.position = c(0.8, 0.8), 
-          legend.text=element_text(size=16),
-          legend.title = element_blank(),
-          axis.text = element_text(size=16), axis.title = element_text(size=16)) +
+    theme(#legend.position = c(0.8, 0.8), 
+          #legend.text=element_text(size=12),
+          #legend.title = element_blank(),
+      legend.position = "none",    
+      axis.text = element_text(size=16), axis.title = element_text(size=16)) +
     geom_hline(yintercept = 0, lwd = 1.5)
 
   plot2 <- ggplot(data = subset(age_cod_results, sex == "Male" & year == input$years_LEgap[2] & 

### 2017-04-25 Manuscript/Manuscript_COD_Trends.Rmd

--- 2017-04-17 Manuscript/Manuscript_COD_Trends.Rmd
@@
 The United States has shown improvement in life expectancy for both blacks and whites since 1969.
-For males, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality, and point to the cause-specific interventions necessary to decrease the black-white gap further.
+For males, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivaled by an increasing gap in CVD-related mortality, and point to the cause-specific interventions necessary to decrease the black-white gap further.
 In women, a crossover in injury-related mortality led to a narrower gap that was realized partially by increasing injury-related mortality among whites.
@@
 The life expectancy of a population is a barometer for overall health.
-In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethinicity.
+In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethnicity.
 Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf].
@@
 These models were run separately for every combination of state, race, and sex.
-They were fit using two markov chains of length 10,000.
+They were fit using two Markov chains of length 10,000.
 One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration (to reduce autocorrelation).
@@
 
-We used random effects meta regression to describe the general pattern in these trend lines.
-Each state's mean contribution was regressed against a linear spline for calendar time that was estimated using four internal knots.
-The model included random effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year.
-To identify divergent states, we compared the state-specific trends (estimated separately for each state using the Bayesian model) to the general pattern estimated using meta regression.
+To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimate from the previous step as a function of a state random effect and a linear spline for year (estimated using four internal knots).
+The model was inversely-weighted by the standard error of the mean estimate in each state-year.
+To identify divergent states, we compared the state-specific trends estimate in the previous step to the national pattern estimate by setting the random effect to zero.
 
@@
 For females, black women experienced relatively larger mortality decreases over time compared to white females, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**).
-For both genders, Michigan, Wisconsin, California, and DC exhibit recent contributions that are relatively larger than the general pattern, while Massachusettes and New York exhibit persistently lower levels of CVD contributions.
+For both genders, Michigan, Wisconsin, California, and DC exhibit recent contributions that are relatively larger than the general pattern, while Massachusetts and New York exhibit persistently lower levels of CVD contributions.
 
@@
 For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern in mortality is similar between blacks and whites resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**).
-For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among males, but not females), while New York and Massachusettes underwent large decreases in the amount that cancer contributed to the life expectancy gap.
+For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among males, but not females), while New York and Massachusetts underwent large decreases in the amount that cancer contributed to the life expectancy gap.
 
@@
 As such, the contribution of non-communicable disease to the mortality gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**).
-As with previous causes, New York and Massachusettes exhibit decreasing contributions, and Nevada has also made substantial progress.
+As with previous causes, New York and Massachusetts exhibit decreasing contributions, and Nevada has also made substantial progress.
 Furthermore, while DC's mortality rates decreased, they still have the highest contribution of non-communicable disease to the gap.
@@
 
-Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communiable disease at the height of the HIV/AIDS epidemic before anti-retroviral therapy was available.
+Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available.
 This spike was greater among males, especially blacks, and within states with larger urban centers.
@@
 White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, while black men have continued to experience decreasing rates within several states.
-As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and reached equality in New York, Massachusettes, and Rhode Island for both genders (**Figures 13 and 14**).
+As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and reached equality in New York, Massachusetts, and Rhode Island for both genders (**Figures 13 and 14**).
 The most disparate states include Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.
@@
 ###Discussion 
+- the next is to investigate potential state level policy and impact on underlying risk factors
+- our analysis highlights a few specific causes in a few specific period
+-injuries his CVD, not so much on cancer - because there is more homo. across contribution of cancer
+- limitations - major groupings but they have lots of heterogeneity of causes within the grouping
+- cont of HIV increases very dramatically, but the pace of decline differed - the places where there is a lag in the decline was the more rural places
 
@@
 - While we found that many state-level trends followed the same pattern of COD-specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result.
-New York and Massachusettes have virtually eliminated the black-white life expectancy gap for both men and women.
+New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women.
 While these states showed contribution decreases for all major causes of death except communicable disease, both states began with only a minimal contribution of CVD to the mortality gap.
@@
 low *because they had decreased over time*.
-Furthermore, our analysis was centred on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
-This is especially important in light of recent studies that have recieved criticism for narrowly focusing on a specific age group, which may present a distorted reality [Case & Deaton, Minton IJE 2016, ].
+Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
+This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group, which may present a distorted reality [Case & Deaton, Minton IJE 2016, ].
 Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.
@@
 Our analysis has two main limitations.
-Firstly, we did not account for hispanic ethinicity, implying that both black and white populations in our study include some portion of hispanics that is known to differ over time and by state.
-We decided against excluding or adjusting for hispanic ethnicity in this study because the indicator for ethnicity has been noted to have varying accuracy across states and during the earlier time period [ref].
-Thus, states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had hispanics been excluded [ref].
-In a previous sensitivity analysis, we found that accounting for hispanic ethinicity did not have a substantial effect on the results [ref Harper 2012], providing some evidence that not accounting for hispanic ethnicity should not substantially affect our findings.
+Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state.
+We decided against excluding or adjusting for Hispanic ethnicity in this study because the indicator for ethnicity has been noted to have varying accuracy across states and during the earlier time period [ref].
+Thus, states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded [ref].
+In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results [ref Harper 2012], providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings.
 
@@
 Migration is particularly important for the District of Columbia, because of it's small geographical bounds (it is easier to be considered migration when regions are smaller - how to say this?), and substantial in-migration from New York, in- and out-migration to contiguous Maryland and Virginia.
-Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are imigrating to DC.
+Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
 These changes would both serve to increase the racial gap in DC.
@@
 -movement of southern blacks to northern states, and migration out of rust belt.
--DC - influx in immigrants since 1980s, huge migration from other US states (NY, MD then Virginia), but also huge outflux to MD and Virginia. Should almost consider these three places as one.
--lots of migraiton into MD
+-DC - influx in immigrants since 1980s, huge migration from other US states (NY, MD then Virginia), but also huge out flux to MD and Virginia. Should almost consider these three places as one.
+-lots of migration into MD
 -NY - 24% are immigrants
 -Texas has influx of immigrants to 17%
--lots of migration into virginia, as well as immigration
+-lots of migration into Virginia, as well as immigration
 -Florida is the state most known for in-migration, both from states (Northeast particularly NY) and from outside the US
@@
 
-Read: racial disparitiy in where people from DC move:
+Read: racial disparity in where people from DC move:
 -https://ggwash.org/view/38326/theres-a-big-racial-disparity-in-where-people-born-in-dc-end-up-moving-or-not-moving

### 2017-04-25 Manuscript/Manuscript_COD_Trends.Rmd

@@
 In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethnicity.
-Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf].
+Whites lived 3.4 years longer than Blacks [ref nchs_LE].
 While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.
@@
 
-We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.
-The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.
+We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013[ref: seer stat].
+The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System [ref: nchs data].
 The population data were derived by the NCHS using intercensal bridged-race population estimates.
@@
 For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming.
-These states contained less than 0.4% of the black population in the United States in 2010 [ref].
-We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since the quality of the coding of ethnicity varied by state over the time period included [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf].
+These states contained less than 0.4% of the black population in the United States in 2010 [ref Census_map].
+We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since the quality of the coding of ethnicity varied by state over the time period included [ref cdc_hisp].
 
@@
 
-To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref].
+To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [@selvin_text; @preston_demography_text].
 Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex.
@@
 #ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Injuries.Male.plot[[3]])
+#ggplotly(Injuries.Male.plot[[3]])
+
+`Injuries.Male.plot`[[3]] +
+           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) + 
+           geom_hline(aes(yintercept = 6), lwd = 0.5, col = "grey", alpha = 0.5)
+
 ```
@@
 #ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Injuries.Female.plot[[3]]) 
+#ggplotly(Injuries.Female.plot[[3]])
+
+`Injuries.Female.plot`[[3]] +
+           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)
 ```
@@
 #ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Communicable.Male.plot[[3]]) 
+#ggplotly(Communicable.Male.plot[[3]])
+
+`Communicable.Male.plot`[[3]] +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) 
 ```
@@
 #ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(Communicable.Female.plot[[3]]) 
+#ggplotly(Communicable.Female.plot[[3]])
+
+`Communicable.Female.plot`[[3]] +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
 ```
@@
 #ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`All other causes.Male.plot`[[3]])
+#ggplotly(`All other causes.Male.plot`[[3]])
+
+`All other causes.Male.plot`[[3]] +
+  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) 
 ```
@@
 #ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-ggplotly(`All other causes.Female.plot`[[3]])
+#ggplotly(`All other causes.Female.plot`[[3]])
+
+`All other causes.Female.plot`[[3]] +
+  geom_hline(aes(yintercept = -2), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)
 ```
@@
 ###Discussion
-- the next is to investigate potential state level policy and impact on underlying risk factors
-- our analysis highlights a few specific causes in a few specific period
--injuries his CVD, not so much on cancer - because there is more homo. across contribution of cancer
-- limitations - major groupings but they have lots of heterogeneity of causes within the grouping
-- cont of HIV increases very dramatically, but the pace of decline differed - the places where there is a lag in the decline was the more rural places
 
@@
 
-- In this paper, we examined state-level trends in six major CODs' contributions to the black-white LE gap.
+In this paper, we examined state-level trends in six major CODs' contributions to the black-white life expectancy gap.
+While we found that many states mimicked the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result.
+We found that cardiovascular disease contributes the most to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
+Injuries continue to play a role in contributing to the current gap for males in several states, especially in the Midwest and DC, while contributing to a decreased gap among females in the South.
+Illinois, Michigan and Wisconsin also show relatively high contributions from cancer.
+On the other hand, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women.
+However, both states began with only a minimal contribution of CVD and cancer to the mortality gap.
+The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contribution of more than two years.
+Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing minimal difference and point to the importance of viewing the most recent estimates in light of their earlier values.
 
-- While we found that many state-level trends followed the same pattern of COD-specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result.
-New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women.
-While these states showed contribution decreases for all major causes of death except communicable disease, both states began with only a minimal contribution of CVD to the mortality gap.
+**Literature review**
 
-- How do these results compare with states with large black populations: Georgia, Florida, Texas, and California?
-In all of these states, deaths during the first year of life are relatively higher in blacks and in most cases, deaths from CVD during mid-life is also higher among blacks.
-California and Florida black young men are dying more of injuries, and black men and women in Florida are dying more of communicable disease.
-
-- These findings can be contrasted with those of Firebaugh et al (2014).
-These authors distinguish between inequality caused by a difference in CVD mortality incidence vs.
-that caused by a difference in age-at-death and find that while blacks and whites are equally likely to die of heart disease, blacks are dying at younger ages.
-[qu: how can the equally likely to die of Heart disease part be true when we have the age std mort rates showing otherwise?
-[*See page 27 of firebaugh - highlighted section - in dropbbox papers directory*].
-
-*Notes of Firebaugh 2014 paper:*
--CVD still contributes the most to the current LE gap.
-Firebaugh et al (2014) say that policy needs to focus on treatment once CVD is realized, not prevention ("The incidence component is negligible, indicating that blacks and whites are equally likely to succumb to heart disease.
-Heart disease nevertheless account for one quarter of the gap in 2015 because black victims tends to be younger than white victims.
-Interventions should focus on identifying heart disease and equalizing treatment.
-Same goes for Cancers, non-comm diseases.
-Contrast with HIV/AIDS, prevention is key (since diff is in "incidence", not age at death with comm disease)"
-[*Does this reasoning make sense to you?
-I'm not sure I follow why a diff in age at death would imply focusing on treatment rather than prevention -- what if blacks get CVD at younger ages than whites?
-Doesn't focusing on equalizing treatment assume that blacks and whites have the same timing of onset?*]
-
-*Literature review*
-
-BW disparities (period-based studies)
-- "higher death rates due to heart disease, cancer, homicide (men), diabetes (women), and perinatal conditions" - kochanek et al 2013, harper et al 2007; see also Orsi 2010
-- harper 2012; 
-- Macinko and Elo (2009): decline in disparity among working-aged people since late 1980s.
+This research extends previous studies that have estimated either the contribution of causes of death to the life expectancy gap on the national level, or the excess risk among blacks in cause of death specific mortality rates (references: Harper 2007, 2012; kochanek 2013, Orsi 2010).
 
@@
 Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.
-[Methodologically, using the Bayesian autoregressive model, allowed us to estimate mortality rates in strata that were sparsely populated...*something about methods?*]
 
@@
 As we are examine the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white.
-Migration between states can also affect life expectancy.
-Migration is particularly important for the District of Columbia, because of it's small geographical bounds (it is easier to be considered migration when regions are smaller - how to say this?), and substantial in-migration from New York, in- and out-migration to contiguous Maryland and Virginia.
+Migration between states can also affect the estimated life expectancy of a state.
+Migration is particularly important for the District of Columbia, because of it's small geographical bounds, and substantial in-migration from New York, in- and out-migration to Maryland and Virginia.
 Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
-These changes would both serve to increase the racial gap in DC.
+Such changes would both serve to increase the racial gap in DC.
 
-helpful: https://www.nytimes.com/2014/08/16/upshot/mapping-migration-in-the-united-states-since-1900.html
--movement of southern blacks to northern states, and migration out of rust belt.
--DC - influx in immigrants since 1980s, huge migration from other US states (NY, MD then Virginia), but also huge out flux to MD and Virginia. Should almost consider these three places as one.
--lots of migration into MD
--NY - 24% are immigrants
--Texas has influx of immigrants to 17%
--lots of migration into Virginia, as well as immigration
--Florida is the state most known for in-migration, both from states (Northeast particularly NY) and from outside the US
--Alabama gets migrants from other southern states
--Connecticut - influx in those immigrating from outside US since 1990s
+Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
+For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.
 
-Individuals who migrate are more likely to have completed a bachelor's degree than those who do not, suggesting that migration may contribute to heightened life expectancy.
+**Conclusions**
 
-Thus, states with black or white populations that have shifted over time will be affected.
-This limits the interpretation of results especially for Washington DC, as it has undergone substantial demographic shifts since X.
--foreign borns tend to have higher LE, so increase in immigration can lead to gains in LE in either blacks or whites, or both, depending on the country of origin - better because of migrant selectivity (select on traits related to good health) and better health (less likely to smoke or be obese or binge drink [ref], and return migration when they become sick)
-
-Problems specific to DC
--migration's effect on LE over the course of the time period
-
--migration - has been on the decline since the 1980s (Molloy 2011, 2017).
-migration happens more in people with bachelor's degrees in their early twenties (and declines thereafter) but is at a bit of a higher rate throughout lifetime than those without degree (Kaplan 2017)
--roughly 1.5% of the popn moves between census regions per year, and 1.3% move within states in a census region.
-
-Read: racial disparity in where people from DC move:
--https://ggwash.org/view/38326/theres-a-big-racial-disparity-in-where-people-born-in-dc-end-up-moving-or-not-moving
-
-*Conclusions*
-
-- DC exhibits the highest contributions for many causes, but they have the greatest gap and a widening gap - so more room for each cause to contribute.
-Goes to say that all causes are contributing to the gap (it isn't being driven by one cause in particular)
-
-Life Expectancy generally
-The spotlight on racial disparities in the US has intensified.
-Several recent papers examining US mortality by region, cause of death, or within population subgroups have received considerable attention while ignoring important black-white disparities.2, 3, 4 These disparities remain profound, and no previous paper has examined cause-specific contributions by state over an extended time period.
-Our analysis provides an important contribution to the literature by highlighting a high degree of heterogeneity in state-specific performance over time in eliminating these inequalities.
-from above:
-2. Chetty R, Stepner M, Abraham S, Lin S, Scuderi B, Turner N, Bergeron A, Cutler D. The association between income and life expectancy in the United States, 2001-2014. JAMA. 2016. 315(16):1750-1766.
-
-3.
-Dwyer-Lindgren L, Bertozzi-Villa A, Stubbs RW, Morozoff C, Kutz MJ, Huynh C, Barber RM, et al.
-US County-Level Trends in Mortality Rates for Major Causes of Death, 1980-2014.
-JAMA.
-2016.
-316(22):2385-2401.
-
-4. Case A, Deaton A. Rising morbidity and mortality in midlife among white non-Hispanic Americans in the 21st century. Proc Natl Acad Sci. 2015. 112(49):15078-83.
-
-Cancer
-
-Oliver. CEBP. 2008. Recent trends in BW disparities in Cancer Mortality. Note: higher mort rates in their paper likely because they restrict to >35yo.
--BW diff in lung cancer mort mimics trends in overall mort rate in males. In females, no diff in lung cancer (although minor difference in other cancer related to smoking)
--increasing disparity is seen in colorectal cancer in males between B and W
--hump also seen in prostate cancer, especially among blacks, but falling mort rates since mid 1990s.
--among females, increasing disparity in mort rates from colorectal cancer and breast cancer since 1975 (beginning of their study)
--not sure what explains the inverted-U in cancer mort in our data among females - this CEBP people doesn't seem to shed light on this.
-
-Screening - when is it proven effective vs. when does it affect the recorded COD? Corinne to review that literature.
--breast cancer: kills mostly among those 70+ yo.
-
--trends in incidence may be related to overdx (prostate cancer, for ex) - for prostate cancer don't know whether the decline is due to improvements in treatment or due to screening -- could be both or just one
-
-https://nccd.cdc.gov/uscs/toptencancers.aspx#White
--black women have higher rates of breast cancer than white women (28/100000 vs 20 for mortality)
--blacks have higher rates of prostate cancer than whites (39.1 vs 18)
-
-Smoking and lung cancer
-see: https://www.washingtonpost.com/news/to-your-health/wp/2015/11/12/smoking-among-u-s-adults-has-fallen-to-historic-lows-these-7-charts-show-who-still-lights-up-the-most/?utm_term=.f2b261e63acf -- uses CDC data for citation
--smoking rates have fallen steadily over the course of the study time period, but homogeneity in smoking still exists. used to be 42.4% in 1965 and down to 16.8% in 2014.
--by race: nearly same for blacks and whites - 17.5% in blacks vs. 18.2% in whites, but 11.2% in hispanics (which are mixed into BW designation in our study)
--by geography: highest in the midwest (20.7%), then south (17%), NE (15.3%), then west (13%), within heterogeneity within state (
-Dwyer-Lindgren  L, Mokdad  AH, Srebotnjak  T, Flaxman  AD, Hansen  GM, Murray  CJ.  Cigarette smoking prevalence in US counties: 1996-2012.  Popul Health Metr. 2014;12(1):5.)
-
--Delancey et al (2008) identified more rapid decreases in smoking related cancers among black males since 1990 as contributing to the decrease in the overall gap.
-
-Obesity and...
--high obesity in the south - ref 9
--women and blacks have highest rates of obesity - ref 8
-
-Opiods
-https://www.cdc.gov/nchs/products/databriefs/db273.htm - Although rates increased 63% for non-Hispanic black persons (from 7.5 per 100,000 in 1999 to 12.2 in 2015) and 43% for Hispanic persons (from 5.4 in 1999 to 7.7 in 2015), non-Hispanic white persons experienced the greatest increase between 1999 and 2015.
-The age-adjusted rate of drug overdose deaths among non-Hispanic white persons increased 240% from 6.2 per 100,000 in 1999 to 21.1 in 2015.
-
-Centers for Disease Control and Prevention.
-Vital signs: overdoses of prescription opioid pain relievers and other drugs among women—United States, 1999–2010.
-MMWR Morb Mortal Wkly Rep.
-2013;62(26):537–542: opioid-related mortality increased 415% among women and 265% among men between 1999 and 2010.38
-
-Paulozzi LJ, Xi Y.
-Recent changes in drug poisoning mortality in the United States by urban-rural status and by drug type.
-Pharmacoepidemiol Drug Saf.
-2008;17(10):997–1005.: in 1999, large central metropolitan areas had the highest opioid-related mortality rates, and noncore areas had the lowest rates; by 2004, noncore areas had the highest rates and had seen the largest relative increase during that time period.50
-
-Paulozzi LJ, Ryan GW.
-Opioid analgesics and rates of fatal drug poisoning in the United States.
-Am J Prev Med.
-2006;31(6):506–511: drug poisoning mortality rates varied 8-fold between states, as did opioid consumption (methadone, 13-fold; oxycodone, 7-fold; overall, 4-fold).17
-
-Centers for Disease Control and Prevention.
-Vital signs: overdoses of prescription opioid pain relievers—United States, 1999–2008.
-MMWR Morb Mortal Wkly Rep.
-2011;60(43):1487–1492.: Another study showed that prescription opioid sales ranged from 3.7 (Illinois) to 12.6 (Florida) kilograms per 10,000 population, and opioid-related mortality rates ranged from 5.5 (Nebraska) to 27 (New Mexico) per 100,000 population.2
-
--should the focus be prevention of specific condictions/diseases, treatment of those conditions/diseases, or more distal --- SES differences - edu, employment, income (Hummer Du Bois review argues we need to focus on SES distal exposures, while Firebaugh 2014 differentiates whhen to focus on treatmnet of condition vs.
-prevention of the condition)
-
--medicaid coverage varies among states.
-MS and NY have longer black LE than general trend and also have expanded medicare coverage (Bharmal 2012 citing the henry kaiser family foundation).
-CA NJ in middle for LE and coverage, AL TX lower for both -- so there is this observed ecological association
+This analysis provides a foundation for future work to investigate time-varying and time-fixed state-level characteristics that may be responsible for differences among state in risk factors leading to each cause of death.
+Differences in smoking policy, medicare coverage, access to care, segregation, and racism, for example, could all affect risk factors for ill health and vary in magnitude by state.
 
@@
 ###Bibliography
+
+to add to bibliography list:
+nchs_LE ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf
+
+seer stat: National Cancer Institute. Surveillance, Epidemiology, and End Results
+Program: SEER*Stat software: latest release: version 8.1.5 [Internet]. Bethesda (MD): NCI; 2013 Mar 31 [cited 2014 Jun 20]. Available from: http://www.seer.cancer.gov/seerstat
+
+nchs data: National Cancer Institute.
+Surveillance, Epidemiology, and End Results Program: US mortality data [Internet].
+Bethesda (MD): NCI; [cited 2014 Jun 20].
+Available from: http://www.seer.cancer.gov/mortality/
+
+cdc_hisp: [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf]
+
+census_map: https://www.census.gov/2010census/popmap/. retrived apr 25, 2017

### 2017-04-26 Manuscript/Manuscript_COD_Trends.Rmd
bib changes

@@
 
-To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [@selvin_text; @preston_demography_text].
+To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [@selvin_2004; @preston].
 Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex.
@@
 
-To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref].
-First, the model partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks.
-For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites.
-This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination.
-For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old.
-We summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference, for each state and for each sex.
+To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method [@arriaga1984measuring].
+This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher age-standardized cause-specific mortality rates among blacks.
+For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher age-standardized rate of mortality due to cardiovascular disease among blacks compared to whites.
+If the decomposition produces a negative estimate, say -0.5 years for injuries, this implies that whites have a higher age-standardized rate of mortality for injuries compared to blacks.
 Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 45 state-level trends and their credible intervals.
@@
 We also calculated age-standardized mortality rates for year-state-race-sex stratum, to aid the interpretation of the decomposition results.
-We used the US standard population from the year 2000 [ref].
+We used the US standard population from the year 2000 [ref census_stdpop].
 
@@
 For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males.
-*Don't know whether you want to note it here or not, but I find it is interesting to see that the gap is now lower in many US southern (or southeastern) states than in the Midwest, suggesting that progress has actually been somewhat greater in the South than elsewhere (particularly the upper Midwest and especially for women).
+
+*SH says: Don't know whether you want to note it here or not, but I find it is interesting to see that the gap is now lower in many US southern (or southeastern) states than in the Midwest, suggesting that progress has actually been somewhat greater in the South than elsewhere (particularly the upper Midwest and especially for women).
 I think this is interesting and somewhat counter to the narrative that health indicators are worse in the South.
@@
 Injury is one of the least common causes of death.
-Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure X**).
+Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][*better word?*] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure X**).
 For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure X**).
@@
 While we found that many states mimicked the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result.
+
 We found that cardiovascular disease contributes the most to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
@@
 
-This research extends previous studies that have estimated either the contribution of causes of death to the life expectancy gap on the national level, or the excess risk among blacks in cause of death specific mortality rates (references: Harper 2007, 2012; kochanek 2013, Orsi 2010).
+This research extends previous studies that have estimated either the contribution of causes of death to the life expectancy gap on the national level, or the excess risk among blacks in cause of death specific mortality rates [@harper2007trends; @harper2012trends; orsi2010black] [ref kochanek_2013].
 
@@
 
-census_map: https://www.census.gov/2010census/popmap/. retrived apr 25, 2017
+census_map: https://www.census.gov/2010census/popmap/. retrieved apr 25, 2017
+
+census_std_pop: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html. retrieved apr 25, 2017
+
+kochanek_2013: https://www.cdc.gov/nchs/data/databriefs/db125.pdf

### 2017-04-26 Manuscript/Appendix-agestd-mort-rates.Rmd
better appendix

@@
 ---
-title: "Appendix: Trends in age-adjusted mortality rates for blacks and whites by cause and sex, 1969-2013, United States"
+title: "Supplementary Appendix: Trends in age-standardized mortality rates for blacks and whites by cause and sex, 1969-2013, United States"
 output: html_document
@@
 
-Corinne A Riddell, PhD
+This is the supplementary appendix for the paper, "Historical trends in the contribution of six major causes of death to the black-white life expectancy gap by US state."
 
-Department of Epidemiology, Biostatistics and Occupational Health, McGill University, Montreal, Canada.
+Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Sam Harper, PhD; Jay S Kaufman, PhD
 
@@
   a <- ggplot(subset.data, aes(x = year, y = rate.per.100k_mean)) + 
-              geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, fill = Race), col = NA, alpha = 0.5) +
+              geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, group = Race), fill = "grey", col = NA, alpha = 0.5) +
               geom_line(aes(col = Race)) +
               facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
-              xlab(" ") + 
+              xlab("Year (1969-2013)") + 
               ylab("Age-standardized mortality rate (per 100,000)") + 
               geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +  
-              ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex1, " for ", COD1)) +
               theme_classic(base_size = 10) +
@@
   b <- ggplot(subset.data2, aes(x = year, y = rate.difference_mean)) +
-           # geom_hline(aes(yintercept = -300), lwd = 0.5, col = "grey") +
-           # geom_hline(aes(yintercept = -200), lwd = 0.5, col = "grey") +
-           # geom_hline(aes(yintercept = -100), lwd = 0.5, col = "grey") +
-           # geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey") +
-           # geom_hline(aes(yintercept = 200), lwd = 0.5, col = "grey") +
-           # geom_hline(aes(yintercept = 100), lwd = 0.5, col = "grey") +
               geom_line() +
@@
               facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
-              xlab(" ") + 
+              xlab("Year (1969-2013)") + 
               ylab("Excess mortality among blacks (per 100,000)") +
               geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +  
-              ggtitle(paste0("Excess mortality among blacks (per 100,000) in ", sex1, " for ", COD1)) +
+              #ggtitle(paste0("Excess mortality among blacks (per 100,000) in ", sex1, " for ", COD1)) +
               theme_classic(base_size = 10) +
@@
 
-    return(list(plot1 = a, plot2 = b))
+  c <- ggplotly(a)
+  d <- ggplotly(b)
+
+  for(i in 1:length(c$x$data)){
+    if (c$x$data[[i]]$line$color == "rgba(0,0,0,1)") {
+      c$x$data[[i]]$hoverinfo <- "none"
+    }
+
+    c$x$data[[i]]$text <- gsub("rate.per.100k_mean", "Mean mortality rate", c$x$data[[i]]$text)
+    c$x$data[[i]]$text <- gsub("rate.per.100k_lcl", "Lower credible limit", c$x$data[[i]]$text)
+    c$x$data[[i]]$text <- gsub("rate.per.100k_ucl", "Upper credible limit", c$x$data[[i]]$text)
+  }
+
+  for(i in 1:length(d$x$data)){
+    if (d$x$data[[i]]$line$color == "rgba(0,0,0,1)") {
+      d$x$data[[i]]$hoverinfo <- "none"
+    }
+
+    d$x$data[[i]]$text <- gsub("rate.difference_mean", "Mean excess mortality", d$x$data[[i]]$text)
+    d$x$data[[i]]$text <- gsub("rate.difference_LCL", "Lower credible limit", d$x$data[[i]]$text)
+    d$x$data[[i]]$text <- gsub("rate.difference_UCL", "Upper credible limit", d$x$data[[i]]$text)
+  }
+
+    return(list(plot1 = a, plot2 = b, plot3 = c, plot4 = d))
 }
@@
 
-**eFigure 1**
+**eFigure 1: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**
 
 ```{r print-em, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cardiovascular.Male.mort.plot[[1]]) %>% 
-  layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F)
-
-ggplotly(Cardiovascular.Male.mort.plot[[2]]) %>% 
-  layout(xaxis = list(title = "Year (1969-2013)"))
-
-#manipulating hover info:
-#p <- ggplotly(Cardiovascular.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
-#p$x$data[[1]]$hoverinfo = "none" #if want no hover info for a particular trace
-#p$x$data[[1]]$text #otherwise, modified what is stored here to display how you want it to look. 
+Cardiovascular.Male.mort.plot[[3]] %>% layout(margin = list(b = 100))
+Cardiovascular.Male.mort.plot[[4]]
 ```
 
-**eFigure 2**
+**eFigure 2: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**
 
 ```{r print-em-2, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cardiovascular.Female.mort.plot[[1]]) %>% 
-  layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F)
-
-ggplotly(Cardiovascular.Female.mort.plot[[2]]) %>% 
-  layout(xaxis = list(title = "Year (1969-2013)"))
+Cardiovascular.Female.mort.plot[[3]] %>% layout(margin = list(b = 100))
+Cardiovascular.Female.mort.plot[[4]]
 ```
 
-**eFigure 3**
+**eFigure 3: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**
 
 ```{r print-em-3, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cancers.Male.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
-ggplotly(Cancers.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+Cancers.Male.mort.plot[[3]] %>% layout(margin = list(b = 100))
+Cancers.Male.mort.plot[[4]]
 ```
 
-**eFigure 4**
+**eFigure 4: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**
 
 ```{r print-em-4, fig.width = 9, fig.height = 9, echo = F}
-ggplotly(Cancers.Female.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
-ggplotly(Cancers.Female.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+Cancers.Female.mort.plot[[3]] %>% layout(margin = list(b = 100))
+Cancers.Female.mort.plot[[4]]
 ```
 
+**eFigure 5: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**
+
+```{r, fig.width = 9, fig.height = 9, echo = F}
+`Non-communicable.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100))
+`Non-communicable.Male.mort.plot`[[4]]
+```
+
+**eFigure 6: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**
+
+```{r, fig.width = 9, fig.height = 9, echo = F}
+`Non-communicable.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100))
+`Non-communicable.Male.mort.plot`[[4]]
+```
+
+**eFigure 7: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**
+
+```{r, fig.width = 9, fig.height = 9, echo = F}
+Injuries.Male.mort.plot[[3]] %>% layout(margin = list(b = 100))
+Injuries.Male.mort.plot[[4]]
+```
+
+**eFigure 8: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**
+
+```{r, fig.width = 9, fig.height = 9, echo = F}
+Injuries.Female.mort.plot[[3]] %>% layout(margin = list(b = 100)) 
+Injuries.Female.mort.plot[[4]]
+```
+
+**eFigure 9: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**
+
 ```{r print-3, fig.width = 9, fig.height=9, echo = F}
-ggplotly(Communicable.Male.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
-ggplotly(Communicable.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+Communicable.Male.mort.plot[[3]] %>% layout(margin = list(b = 100))
+Communicable.Male.mort.plot[[4]]
+```
 
-ggplotly(Communicable.Female.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
-ggplotly(Communicable.Female.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+**eFigure 10: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**
 
-ggplotly(`Non-communicable.Male.mort.plot`[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
-ggplotly(`Non-communicable.Male.mort.plot`[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+```{r print-4, fig.width = 9, fig.height=9, echo = F}
+Communicable.Female.mort.plot[[3]] %>% layout(margin = list(b = 100))
+Communicable.Female.mort.plot[[4]]
+```
 
-ggplotly(`Non-communicable.Female.mort.plot`[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
-ggplotly(`Non-communicable.Male.mort.plot`[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+**eFigure 11: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**
 
-ggplotly(Injuries.Male.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
-ggplotly(Injuries.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+```{r, fig.width = 9, fig.height = 9, echo = F}
+`All other causes.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100))
+`All other causes.Male.mort.plot`[[4]]
+```
 
-ggplotly(Injuries.Female.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
-ggplotly(Injuries.Female.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+**eFigure 12: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**
 
-ggplotly(`All other causes.Male.mort.plot`[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
-ggplotly(`All other causes.Male.mort.plot`[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
-
-ggplotly(`All other causes.Female.mort.plot`[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
-ggplotly(`All other causes.Female.mort.plot`[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
+```{r, fig.width = 9, fig.height = 9, echo = F}
+`All other causes.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100)) 
+`All other causes.Female.mort.plot`[[4]]
 ```

### 2017-04-26 Manuscript/Manuscript_COD_Trends.Rmd
better appendix

--- 2017-04-25 Manuscript/Manuscript_COD_Trends.Rmd
@@
 
-Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**eFigure X**).
+Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**eFigure 5**).
 Recently, there has been a slight decrease in mortality rates, especially among black males.
-These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure Y**).
+These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure 6**).
 As such, the contribution of non-communicable disease to the mortality gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**).
@@
 Injury is one of the least common causes of death.
-Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure X**).
-For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure X**).
+Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure 7**).
+For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure 8**).
 Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white females demonstrating sharper increases.
@@
 
-Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available.
+Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**eFigures 9 and 10**).
 This spike was greater among males, especially blacks, and within states with larger urban centers.
@@
 
-Mortality from all other causes generally decreased earlier in the time period, especially among blacks.
+Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**eFigures 11 and 12**).
 White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, while black men have continued to experience decreasing rates within several states.

### 2017-04-26 Manuscript/Manuscript_COD_Trends.Rmd

--- 2017-04-26 Manuscript/Manuscript_COD_Trends.Rmd
@@
 Injury is one of the least common causes of death.
-Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][*better word?*] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure X**).
-For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure X**).
+Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure 7**).
+For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure 8**).
 Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white females demonstrating sharper increases.

### 2017-04-28 Manuscript/Manuscript_COD_Trends.Rmd
added infant contrib plots

@@
                                                                         "Communicable", "Non-communicable",  "Injuries"))
+
+age_cod_results_infants_AOC <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_infants_AOC.csv")
+age_cod_results_infants_AOC$stabbrs <- state.abb[match(age_cod_results_infants_AOC$state, state.name)]
+age_cod_results_infants_AOC$stabbrs[age_cod_results_infants_AOC$state == "Washington DC"] <- "DC"
+age_cod_results_infants_AOC <- reorder.as.map(age_cod_results_infants_AOC, "state", "stabbrs")
+
 ```
@@
     #geom_line(aes(y = fitted.RE), col = "#d7191c", lty = 3) +
-    ggtitle(paste0("Estimated smoothed contribution of ", COD1, " disease to the black-white life expectancy gap (blue 95% credible interval) vs.
+#    ggtitle(paste0("Estimated smoothed contribution of ", COD1, " disease to the black-white life expectancy gap (blue 95% credible interval) vs.
 the meta-estimated general pattern (red line) in ", sex1, " 1969-2013")) +
@@
 
+```{r, echo = F}
+male.endpoints.aoc <- cod_decomp_results %>% filter(year %in% c(1969, 2013), sex =="Male", COD == "All other causes", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )
+
+female.endpoints.aoc <- cod_decomp_results %>% filter(year %in% c(1969, 2013), sex =="Female", COD == "All other causes", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )
+
+male.infants.endpoints.aoc <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Male", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )
+
+female.infants.endpoints.aoc <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Female", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )
+```
+
 Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**eFigures 11 and 12**).
@@
 The most disparate states include Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.
+
+*CR note: I just added the following paragraph.
+Open to how/whether we want to talk about this.
+Thought we need to say something about how "all other causes partiallly reflects infant mortality.
+Also these additional figures (Fig 15 and 16) would be moved to the appendix.*
+
+To investigate how much of the gap in all other causes occurred during the first year of life, we conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred *during the first year of life* to the life expectancy gap (**Figures 15 and 16**).
+We investigated this because the majority of infant deaths are classified as being due to all other causes, conflating the all other causes category with inquities in infant mortality.
+We found that 43% of the contribution of all other causes to the gap was rooted in differences during the first year of life.
+In 2013, 53% of these deaths occurred during the first year of life.
+Thus, the trend in the all other causes contribution partially reflects a decreasing disparity in infant mortality.
 
@@
 
+**Figure 15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males younger than one**
+
+```{r age-cod-trend-infant-males, fig.width = 9, fig.height = 9, echo = F}
+ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Male"), aes(y = age_COD_cont_yrs_mean, x = year)) +
+  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
+  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
+  geom_line(col = "#08519c") +  
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (years)") +
+  xlab("Year (1969-2013)") +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
+```
+
+**Figure 16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in females younger than one**
+
+```{r age-cod-trend-infant-females, fig.width = 9, fig.height = 9, echo = F}
+ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Female"), aes(y = age_COD_cont_yrs_mean, x = year)) +
+  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
+  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
+  geom_line(col = "#08519c") +  
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (years)") +
+  xlab("Year (1969-2013)") +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
+```
+
 ###Discussion

### 2017-04-28 Code/4_combined-meta-regression.Rmd

--- 2017-04-07 Code/combined-meta-regression.Rmd
@@
 ```{r load-data}
-cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
+cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
 cod_decomp_results$COD_cont_yrs_width <- cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl
 cod_decomp_results$COD_cont_yrs_SE <- (cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl)/3.92
-```
 
-```{r FE_meta_regression_of_cod_contribution, echo = F}
-#Fixed effects meta analysis (fixed effects for state)
+age_cod_results_female <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_female.csv")
+age_cod_results_male <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_male.csv")
+age_cod_results <- rbind(age_cod_results_female, age_cod_results_male)
+rm(age_cod_results_female, age_cod_results_male)
 
-#run 12 stratified models. bs() created the b-spline basis
-models <- cod_decomp_results %>% 
-  group_by(COD, sex) %>%
-  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state,
-              weights = 1/(COD_cont_yrs_SE)^2, 
-              data = .))
-
-#calculated predicted values for all models
-dfPred <- broom::augment(models, fit)
-dfPred$year <- 1969:2013
-
-#merge the predictions with original dataset
-cod_decomp_results <- merge(cod_decomp_results, 
-                            dfPred %>% select(COD, sex, year, state, .fitted, .resid), 
-                            by = c("COD", "sex", "state", "year")) %>%
-                      rename(fitted.FE = .fitted, resid.FE = .resid)
-
-#calculated the sum of the residuals for each state-sex-cod strata (summed over years)
-cod_decomp_results <- cod_decomp_results %>% 
-  group_by(state, sex, COD) %>%
-  mutate(sum.residual = sum(abs(resid.FE)))
+age_cod_results$age_COD_cont_yrs_SE <- (age_cod_results$age_COD_cont_yrs_ucl - age_cod_results$age_COD_cont_yrs_lcl)/3.92
+age_cod_results_infants_AOC <- subset(age_cod_results, age_minbin == 0 & COD == "All other causes")
 ```
@@
 
+```{r FE_meta_regression_of_cod_contribution, echo = F}
+#Fixed effects meta analysis (fixed effects for state)
+
+#run 12 stratified models. bs() created the b-spline basis
+models <- cod_decomp_results %>% 
+  group_by(COD, sex) %>%
+  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state,
+              weights = 1/(COD_cont_yrs_SE)^2, 
+              data = .))
+
+#calculated predicted values for all models
+dfPred <- broom::augment(models, fit)
+dfPred$year <- 1969:2013
+
+#merge the predictions with original dataset
+cod_decomp_results <- merge(cod_decomp_results, 
+                            dfPred %>% select(COD, sex, year, state, .fitted, .resid), 
+                            by = c("COD", "sex", "state", "year")) %>%
+                      rename(fitted.FE = .fitted, resid.FE = .resid)
+
+#calculate the sum of the residuals for each state-sex-cod strata (summed over years)
+cod_decomp_results <- cod_decomp_results %>% 
+  group_by(state, sex, COD) %>%
+  mutate(sum.residual = sum(abs(resid.FE)))
+```
+
+```{r RE_meta_regression_agecod_infants_only, echo = F}
+RE.models.infants <- age_cod_results_infants_AOC %>% 
+  group_by(sex) %>%
+  do(fit = lmer(age_COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + (1|state),
+              weights = 1/(age_COD_cont_yrs_SE)^2, 
+              data = .))
+
+#creates a dataframe for males and females of the fitted estimates. Includes state variable, but we need to add on the year variable
+RE.pred.infants <- broom::augment(RE.models.infants, fit)
+RE.pred.infants$year <- 1969:2013
+
+RE.models.infants$RE.estimates <- purrr::map(RE.models.infants$fit, ~ ranef(.x))
+
+sex.infants <- c(rep("Female", 40), rep("Male", 40))
+COD.infants <- rep("All other causes", 80)
+RE.estimates.infants <- c(RE.models.infants$RE.estimates[[1]]$state$`(Intercept)`, RE.models.infants$RE.estimates[[2]]$state$`(Intercept)`)
+state.infants <- rep(levels(age_cod_results$state), 2)
+REs.infants <- data.frame(sex.infants, COD.infants, RE.estimates.infants, state.infants)
+REs.infants <- REs.infants %>% rename(sex = sex.infants, COD = COD.infants, RE.estimates = RE.estimates.infants, state = state.infants)
+
+#merge in the RE estimates to the original dataset
+age_cod_results_infants_AOC <- merge(age_cod_results_infants_AOC,
+                            REs.infants, by = c("sex", "state"), #state-specific REs; don't need time for merge (because time fixed)
+                            all.x = T)
+
+#merge other aspects, like the FE, with original dataset
+age_cod_results_infants_AOC <- merge(age_cod_results_infants_AOC, 
+                             RE.pred.infants %>% select(sex, state, year, .fitted, .resid, .fixed), 
+                             by = c("sex", "state", "year"))
+
+age_cod_results_infants_AOC <- age_cod_results_infants_AOC %>% rename(fitted.RE = .fitted, fixed.from.RE = .fixed, resid.RE = .resid)
+```
+
 ```{r save_data}
-write.csv(cod_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
+write.csv(cod_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
+
+write.csv(age_cod_results_infants_AOC, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_infants_AOC.csv")
 ```

### 2017-05-03 shiny_app/app.R
customized css

@@ -80,7 +80,7 @@ BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")
 
 
 
-ui1 <- fluidPage(theme = shinytheme("cosmo"),
+ui1 <- fluidPage(theme = "cosmo-customized.css",
                  pageWithSidebar(
                  
                  #shinythemes::themeSelector(), 
@@ -123,7 +123,7 @@ ui1 <- fluidPage(theme = shinytheme("cosmo"),
                                                  #              value = FALSE, col = "GB", type = "OO"),
                                                  radioButtons(inputId = "pop_model",
                                                               label = "Population curve:",
-                                                              choices = c("Hide", "State-specific", "Aggregated"),
+                                                              choices = c("Hide", "National", "State-specific"),
                                                               selected = "Hide"),
                                                  radioButtons(inputId = "contribution_type", 
                                                               label = "Contribution format:", 
@@ -187,8 +187,9 @@ ui1 <- fluidPage(theme = shinytheme("cosmo"),
                        
                      )
                    )
+
                  )
-                   
+  
 )
 
 server <- function(input, output) {
@@ -235,7 +236,11 @@ server <- function(input, output) {
                geom_text(data = subset(BlackWhite_results, 
                                        year == input$years_LEgap[2]-2 & sex == input$sex), 
                          aes(label = stabbrs), check_overlap = T, size = 2.5) +
-               theme_minimal() +  theme(axis.text.x = element_text(angle = 40)) +
+               theme_minimal() +  
+                 theme(axis.text.x = element_text(angle = 40),
+                       panel.background = element_rect(fill = "transparent", colour = NA), 
+                       plot.background = element_rect(fill = "transparent", colour = NA)
+                       ) +
                geom_hline(yintercept = 0)
              )
     
@@ -259,7 +264,10 @@ server <- function(input, output) {
     theme(axis.text.x = element_blank(),
           strip.background=element_blank(),
           axis.line=element_blank(),
-          axis.ticks=element_blank()) +
+          axis.ticks=element_blank(),
+          panel.background = element_rect(fill = "transparent", colour = NA), 
+          plot.background = element_rect(fill = "transparent", colour = NA),
+          legend.background = element_rect(fill = "transparent", colour = NA)) +
[…]

### 2017-05-03 conferences/ResearchDay/Riddell_ResearchDay_minimal.Rmd
research day stuff

---
title: "Changes in cause of death and the black-white life expectancy gap"
author: "Corinne Riddell, PhD"
date: "April 24, 2017"
output: ioslides_presentation
---

<head>
<style>
.greentext {
        color: green;
}
.redtext {
        color: red;
}
</style>
</head>

```{r initialize, echo=FALSE, warning=F, message=F}
library(plotly)
library(viridis)
library(scales)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
library(splines)
library(broom)
library(tidyverse)

library(png)
library(jpeg)
library(grid)
library(gridExtra)
library(knitr)

source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

## Thank you

```{r, echo=F}
img1 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/k_morrison.png")), interpolate = FALSE)
img2 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/s_harper2.png")), interpolate = FALSE)
img3 <-  rasterGrob(as.raster(readJPEG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/j_kaufman.jpg")), interpolate = FALSE)
grid.arrange(img3,
             img1,
             img2,
             ncol = 3)
```

----

```{r, echo = F, out.width = 500}
include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/Harper_HA_title.png")
```

```{r, echo=F, out.width = 750}
include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/Harper_HA_fig1.png")
```

----

```{r, echo = F, out.width = 250}
include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/Harper_HA_quote.png")
```

And also...

>- be broadly accessible,
>- be interactive,
>- be reproducible,
>- and allow pictures to speak for themselves.

## The data

```{r, echo = F, out.width = 400}
include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/alabama.data.ex.png")
```

- <b class = "greentext">40</b> "states": 39 states + DC
- <b class = "greentext">45</b> years: 1969--2013 
- <b class = "greentext">19</b> age groups: <1, 1-4, 5-9, ..., 85+
- <b class = "greentext">2</b> sexes
- <b class = "greentext">2</b> races (Black, White)
- <b class = "greentext">6</b> major causes of deaths: CVD, cancer, non-communicable, communicable, injuries, all other causes
- = <b class = "greentext">820,800 rows of data</b>

## The data

```{r, echo = F, out.width = 400}
include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/alabama.data.ex.png")
```

- When the number of deaths is between 1 and 9 it is <b class = "redtext">coarsened</b>.

## The analysis

**Estimate smoothed mortality rates and impute coarsened data.**

+ Bayesian autoregressive model to borrow information *over time* within strata
+ $log(\lambda_t) \sim N(\ log(\lambda_{t-1}),\ \tau)$
+ impute coarsened 1--9 values using a truncated Poisson
+ `rjags`

## The analysis

**Create 1,000 posterior samples**

+ Each sample contains estimated *smoothed* and *imputed* death counts
+ For each sample:
    1. Create a life table for every strata
    2. Calculate the <b style = "greentext">difference</b> in life expectancy between blacks and whites
    3. Decompose the difference by age and cause of death
    4. Calculate age-standardized mortality rates
+ Calculate the mean, 5^th^ and 95^th^ percentiles for all estimates.

## The analysis

**Synthesize state-specific findings**

+ Random-effects meta regression to estimate overall trend in contributions for each cause.
+ Use `ggplot2`, `plotly`, and `Shiny` to make results accessible as a web app.

----
# To the results!

### 2017-05-03 conferences/Max Planck/submission.Rmd

---
title: Exploring how changes in cause of death have narrowed the black-white life
  expectancy gap in the United States
output:
  word_document: default
  pdf_document: default
  html_document: default
always_allow_html: yes
---

```{r initialize, echo=FALSE, warning=F, message=F}
library(plotly)
library(scales)
library(dplyr)
library(purrr)
library(stringr)
library(ggplot2)
library(stringr)
library(reshape2)
library(grid)
library(gridExtra)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

```{r read_data, echo=F}
age_cod_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results.csv")
age_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_results.csv")
cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")

COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
                "Non-communicable", "Injuries", "All other causes")

cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
age_cod_results$COD <- factor(age_cod_results$COD, levels(age_cod_results$COD)[c(3, 2, 4, 5, 6, 1)])

age_decomp_results$age <- factor(age_decomp_results$age, levels = levels(age_decomp_results$age)[c(1, 2, 11, 3:10, 12:19)])
age_cod_results$age <- factor(age_cod_results$age, levels = levels(age_cod_results$age)[c(1, 2, 11, 3:10, 12:19)])                                                             
```

```{r prep_era_figure, echo = F}
year1 <- 1969
year2 <- 1983
year3 <- 1993
year4 <- 2013

bw_dat <- BlackWhite_results %>% 
  filter(year %in% c(year1, year2, year3, year4), is.na(LE_wbgap) == F) %>%
  select(state, year, LE_wbgap, sex) %>%
  arrange(year)

bw_dat <- tidyr::spread(bw_dat, year, LE_wbgap) 
names(bw_dat)[3:6] <- c("first.gap","second.gap", "third.gap", "fourth.gap")  
bw_dat.males <- bw_dat %>% filter(sex == "Male") %>%
  mutate(era1.diff = first.gap - second.gap,
         era2.diff = second.gap - third.gap,
         era3.diff = third.gap - fourth.gap,
         state.gdiff.order.1 = reorder(state, first.gap),
         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
         state.gdiff.order.2 = reorder(state, second.gap),
         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
         state.gdiff.order.3 = reorder(state, third.gap),
         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
  )

bw_dat.females <- bw_dat %>% filter(sex == "Female") %>%
  mutate(era1.diff = first.gap - second.gap,
         era2.diff = second.gap - third.gap,
         era3.diff = third.gap - fourth.gap,
         state.gdiff.order.1 = reorder(state, first.gap),
         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
         state.gdiff.order.2 = reorder(state, second.gap),
         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
         state.gdiff.order.3 = reorder(state, third.gap),
         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
  )

cod.df <- data.frame(state = factor(), sex = factor(), era = character(), 
                     COD = factor(), Contribution.to.change = numeric(), 
                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())

age.df <- data.frame(state = factor(), sex = factor(), era = character(), 
                     age_minbin = numeric(), Contribution.to.change = numeric(), 
                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())

years <- c(1969, 1983, 1993, 2013)

for (state.i in levels(BlackWhite_results$state)) {
  for (sex.i in c("Male", "Female")) {
    for (year.i in 1:3) {

      table1 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(COD)
      table2 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(COD)

      table3 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(age_minbin)
      table4 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(age_minbin)

      temp1 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
                          contribution.to.gap.change(type.of.decomp = "COD", 
                                                     decomp.table1 = table1, 
                                                     decomp.table2 = table2, 
                                                     type.of.data = "results"))

      temp2 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
                          contribution.to.gap.change(type.of.decomp = "Age", 
                                                     decomp.table1 = table3, 
                                                     decomp.table2 = table4,
                                                     type.of.data = "results"))

      cod.df <- rbind(cod.df, temp1)
      age.df <- rbind(age.df, temp2)
    }
  }
}

```

```{r more_figure_prep, echo = F}

cod.df <- make_pretty_decomp_plot(decomp.table = cod.df, strat.var.1 = "state", 
                        strat.var.2 = "sex", strat.var.3 = "era", 
                        sign.var = "narrowed_gap",
                        decomp.var = "Contribution.to.change",
                        decomp.var.prop = "Contrib.to.change.prop", 
                        partition.bar.var = "COD", type.of.data = "results")

age.df <- make_pretty_decomp_plot(decomp.table = age.df, strat.var.1 = "state", 
                        strat.var.2 = "sex", strat.var.3 = "era", 
                        sign.var = "narrowed_gap",
                        decomp.var = "Contribution.to.change",
                        decomp.var.prop = "Contrib.to.change.prop", 
                        partition.bar.var = "Ages", type.of.data = "results")

cod.df <- merge(cod.df, BlackWhite_results[BlackWhite_results$year == 1969 & BlackWhite_results$sex == "Male", 
                                           c("state", "Census_Region", "Census_Division", "stabbrs")], by = c("state"))
```

```{r, echo = F}
sum1 <- BlackWhite_results %>% filter(year == 1969 & sex == "Male") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)

diff <- bw_dat.males %>% arrange(era1.diff)

for(i in 1:5) {
  j <- i-1

  max.value <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-j, ]$state & 
                                            cod.df$era == "1969-1983" &
                                            cod.df$sex == "Male" ])

  max.cause <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-j, ]$state &
                    cod.df$sex == "Male" &
                    cod.df$era == "1969-1983" & 
                    cod.df$Contribution.to.change == max.value]

  min.value <- min(cod.df$Contribution.to.change[cod.df$state == diff[i, ]$state & 
                                            cod.df$era == "1969-1983" &
                                            cod.df$sex == "Male" ])

  min.cause <- cod.df$COD[cod.df$state == diff[i, ]$state &
                    cod.df$sex == "Male" &
                    cod.df$era == "1969-1983" & 
                    cod.df$Contribution.to.change == min.value]

  assign(paste0("max", i), max.value)
  assign(paste0("max.cause", i), max.cause)
  assign(paste0("min", i), min.value)
  assign(paste0("min.cause", i), min.cause)
}

cod.df.males <- merge(subset(cod.df, sex == "Male"), bw_dat.males, by = c("state"))
cod.df.males <- cod.df.males %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)

```

```{r, echo =F}
sum2 <- BlackWhite_results %>% filter(year == 1983 & sex == "Male") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)

  min.value.DC <- min(cod.df$Contribution.to.change[cod.df$state == "Washington DC" & 
                                            cod.df$era == "1983-1993" &
                                            cod.df$sex == "Male" ])

  min.cause.DC <- cod.df$COD[cod.df$state == "Washington DC" &
                    cod.df$sex == "Male" &
                    cod.df$era == "1983-1993" & 
                    cod.df$Contribution.to.change == min.value.DC]
```

```{r, echo =F}
sum3 <- BlackWhite_results %>% filter(year == 1993 & sex == "Male") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)

sum4 <- BlackWhite_results %>% filter(year == 2013 & sex == "Male") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
```

**Provision Title:** Exploring how changes in cause of death have narrowed the black-white life expectancy gap in the United States.

**Author list**:

Corinne A Riddell, PhD. <corinne.riddell@mail.mcgill.ca>

Kathryn T Morrison, PhD. <kt.morrison@mail.mcgill.ca>

Jay S Kaufman, PhD. <jay.kaufman@mcgill.ca>

Sam Harper, PhD. <sam.harper@mcgill.ca>

Department of Epidemiology, Biostatistics and Occupational Health, McGill University, Montreal, Canada.

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>

**Research Areas**: Regional mortality variation and its determinants, Socioeconomic differences in the face of death.

###Abstract

**Background**

The overall difference in life expectancy between blacks and whites in the US has generally decreased over time. States have fared differently in their progress towards eliminating the life expectancy gap, with states such as New York experiencing greater reductions to the gap compared to states such as California and Texas. The objective of this study is to compare how changes to cause-specific mortality between 1969 and 2013 have resulted in narrower life expectancy gaps in some states and  wider gaps in others.

**Methods**

Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes according to year (1969-2013), state, age, sex, race (black or white), and six cause of death categories (Cancer, cardiovascular disease, communicable disease, non-communicable disease, injuries, and all other causes). We excluded 10 states with the smallest black populations, but included the District of Columbia. A Bayesian autoregressive model was used to estimate smoothed mortality rates over time within each stratum, and a truncated Poisson model was used to impute death counts that were suppressed due to small cell sizes. We calculated life expectancy for blacks and whites within each stratum using traditional life table methods, and decomposed the difference in life expectancy by cause of death using Arriaga's method. In order to understand whether states fared similarly over time, we divided the time frame into three eras which roughly encapsulate an initial period of increases in life expectancy for both races and sexes (1969-1983), an intermediate period of decreases and stagnating life expectancy for blacks but not for whites (1983-1993) and a final period of increasing life expectancy for both races (1993-2013). We then decomposed the *change* in the *difference* in life expectancy between blacks and whites that occurred over each time frame for men and women in each state.

**Results**

**Fig. 1-6** depict the change in the life expectancy gap by states over the three time periods for men (**Fig. 1-3**) and for women (**Fig. 4-6**). For example, **Fig. 1** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs. their gap in 1983 (solid dashed line), with the change in the gap being represented by the number of years between these lines along the x-axis. This gap is decomposed by cause of death, where causes to the left of the dashed line contributed to narrowing the gap and causes to the right contributed to widening the gap. The length of these coloured bars represent the number of years attributable to the cause along the x-axis.

While many states follow the already-known trends during the US at the time, interesting insights regarding specific states can be made. For example, changes to the life expectancy gap in Washington DC are markedly different than changes that occurred in other states. During the first two time periods, the gap widened substantially, initially due to CDV and the residual category, followed by injuries and communicable disease during the HIV/AIDs and crime epidemics. In 1993, the gap in DC is 15.9 years for males, and this decreased to 10.5 years by 2013 -- a significant decrease that still puts the gap 2.5 years larger than all other states.

Another interesting insight is the heterogeneous effect of CVD on the gap across states. For example, during the second time period, CVD contributed to a wider gap in women in states such as DC, California, and Pennsylvania, but a narrower gap in Kentucky, Kansas, and Delaware (**Fig. 5**).

Changes to the difference in the life expectancy gap is the result of changes to age-specific mortality rates for blacks and whites over time. The **appendix** (found after Fig 6, at the end of the document) illustrates age-adjusted mortality rates for blacks and whites for each cause followed by the resulting difference in mortality. These figures aid the interpretation of the main figures. For example, CVD has the highest rate of mortality of all the causes in males, and has decreased for both races over time. However, some states experienced a "cross-over" in the black and white mortality rates (e.g., Wisconsin), with the rates changing from being initially lower among blacks to being higher for blacks, and resulted in CVD's contribution towards widening the life expectancy gap in these states.

Another interesting insight from the **appendix** is the cross-over of injury-related mortality rates for females during the 1990s, where white females, especially in the Southern states experienced increasingly higher rates of mortality due to injuries. While the life expectancy gaps narrowed for females during this time period, these figures show that the contribution of injuries to the narrowing was not solely due to decreasing injury-related mortality among blacks, but also due to increasing injury-related mortality among whites.

**Conclusion**

The United States underwent improvements in life expectancy for both blacks and whites since 1969. For males, much of this improvement was due to narrowing differences in mortality rates due to injuries, but these contributions were rivalled by the increasing gap in CVD-related mortality, and point to the cause-specific interventions necessary to decrease the gap further. In females, the striking cross-over in injury-related mortality in many states led to narrower life expectancy gaps that were realized partially by increasing mortality rates among whites (rather than the preferred mechanism of decreasing rates among both groups). Thus, this work emphasizes the importance of understanding how the direction of changes in mortality by race results in changes to the life expectancy gap.

---

**Fig 1: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**

```{r figure1, fig.align = "center", echo=F, fig.height=9, fig.width=7}
y.labels <- levels(factor(cod.df.males$state.gdiff.order.1))
y.num.breaks <- length(levels(factor(cod.df.males$state.gdiff.order.1)))

males.era1 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = COD), 
              color = "white") +
    geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
        geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
    theme_minimal() + 
    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
    scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era1
```

---

**Fig 2: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1983 and 1993 by US state**

```{r figure2, echo = F, fig.height=9, fig.width=11}
# - The gap in 1983 ranges from `r round(sum2[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum2[1, ]$LE_wbgap_lcl, 1)`, `r round(sum2[1, ]$LE_wbgap_ucl, 1)`] in `r sum2[1, ]$state ` to `r round(sum2[dim(sum2)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum2[dim(sum2)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum2[dim(sum2)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum2[dim(sum2)[1], ]$state `.

y.labels.2 <- levels(factor(cod.df.males$state.gdiff.order.2))
y.num.breaks.2 <- length(levels(factor(cod.df.males$state.gdiff.order.2)))

males.era2 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
                  ymax = state.gdiff.num.2 + 0.45, 
                  xmax = new.finish.2, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
    theme_minimal() + 
    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
    scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2), minor_breaks = NULL) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era2 
```

---

**Fig 3: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1993 and 2013 by US state**
```{r figure3, echo = F, fig.height=9, fig.width=11}
y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
y.num.breaks.3 <- length(levels(factor(cod.df.males$state.gdiff.order.3)))

males.era3 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
                  ymax = state.gdiff.num.3 + 0.45, 
                  xmax = new.finish.3, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
    theme_minimal() + 
    scale_x_continuous(breaks = seq(0, 16, by = 2), labels = seq(0, 16, by = 2)) +
    scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))

males.era3
```

---

**Fig 4: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**

```{r figure4, fig.align = "center", echo=F, fig.height=9, fig.width=7}

cod.df.females <- merge(subset(cod.df, sex == "Female"), bw_dat.females, by = c("state"))
cod.df.females <- cod.df.females %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)

y.labels <- levels(factor(cod.df.females$state.gdiff.order.1))
y.num.breaks <- length(levels(factor(cod.df.females$state.gdiff.order.1)))

females.era1 <- ggplotly(
  ggplot(subset(cod.df.females, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.females,
                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
    geom_segment(data = bw_dat.females,
                 aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

females.era1
```

---

**Fig 5: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1983 and 1993 by US state**

```{r figure2b, fig.align = "center", echo=F, fig.height=9, fig.width=8}
y.labels.2 <- levels(factor(cod.df.females$state.gdiff.order.2))
y.num.breaks.2 <- length(levels(factor(cod.df.females$state.gdiff.order.2)))

females.era2 <- ggplotly(
  ggplot(subset(cod.df.females, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
                  ymax = state.gdiff.num.2 + 0.45, 
                  xmax = new.finish.2, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.females,
                 aes(x = third.gap, xend = third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
    geom_segment(data = bw_dat.females,
                 aes(x = second.gap, xend = second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1983-1993"), yaxis = list(title = NA, autorange = "reversed"))

females.era2
```

---

**Fig 6: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1993 and 2013 by US state**

```{r fig6, fig.align = "center", echo=F, fig.height=9, fig.width=9}
y.labels.3 <- levels(factor(cod.df.females$state.gdiff.order.3))
y.num.breaks.3 <- length(levels(factor(cod.df.females$state.gdiff.order.3)))

females.era3 <- ggplotly(
  ggplot(subset(cod.df.females, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
                  ymax = state.gdiff.num.3 + 0.45, 
                  xmax = new.finish.3, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.females,
                 aes(x = fourth.gap, xend = fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
    geom_segment(data = bw_dat.females,
                 aes(x = third.gap, xend = third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))

females.era3
```

---  
---  
**Appendix: Trends in age-adjusted mortality rates for blacks and whites by cause and sex, 1969-2013, United States **

```{r, echo=F}
mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates.csv")
mortality.rates.wide <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/mortality_rates_wide.csv")

mortality.rates <- mortality.rates %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
mortality.rates.wide <- mortality.rates.wide %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division)))
```

```{r, echo=F, fig.width=11, fig.height=11}
for(sex_i in c("Male", "Female")) {
  for(cod in levels(mortality.rates$COD)) {
    data.sub <- subset(mortality.rates, sex == sex_i & post.samp == 1 & COD == cod)
    a <- ggplot(data.sub, 
                aes(y=rate.per.100k, x = year)) + 
      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
      geom_line(aes(col = Census_Division, lty = race)) +
      facet_wrap( ~ state.reorder, ncol = 5) +
      ylab("Age-standardized mortality rate (per 100,000)") +
      xlab("Year") + 
      ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))

    print(a)

    b <- ggplot(subset(mortality.rates.wide, sex == sex_i & post.samp == 1 & COD == cod), 
                aes(y=rate.difference, x = year)) + 
      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
      geom_hline(aes(yintercept = 0), col = "#636363") +
      geom_line(aes(col = Census_Division)) +
      facet_wrap( ~ state.reorder, ncol = 5) +
      ylab("Black-white difference (per 100,000)") +
      xlab("Year") + 
      ggtitle(paste0("Black-white difference in age-adjusted mortality (per 100,000) in ", 
                     sex_i, " for ", cod))
    print(b)

  }
}
```

### 2017-05-12 Manuscript/Manuscript_COD_Trends.Rmd

@@
 cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
-cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
+cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) 
+cvd.female.2013 <- format(round(cvd.female.2013, 1), nsmall = 1)
 
@@
 
-Although mortality from cardiovascular disease has decreased over time for both racial groups, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r cvd.female.2013` years).
-Non-communicable diseases’ contribution to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in 1988) and a stable contribution for women (approximately `r can.female.average` years).
-Both men and women exhibited a decreased contribution for injuries (men: `r inj.male.1969` to `r inj.male.2013` years, women: `r inj.female.1969` to `r inj.female.2013` years), with the contribution becoming negative for women.
+Although mortality from cardiovascular disease has decreased dramatically for both racial groups, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r formatC(as.numeric(cvd.female.2013), digits = 1)` years).
+The contribution of non-communicable diseases’ contribution to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in 1988) and a relatively stable contribution for women (approximately `r can.female.average` years).
+Both men and women exhibited a decreased contribution for injuries (men: `r inj.male.1969` to `r inj.male.2013` years), with the contribution becoming negative for women (women: `r inj.female.1969` to `r inj.female.2013` years).
 Several states did not follow these general trends.
@@
 
-The United States has shown improvement in life expectancy for both blacks and whites since 1969.
-For males, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivaled by an increasing gap in CVD-related mortality, and point to the cause-specific interventions necessary to decrease the black-white gap further.
-In women, a crossover in injury-related mortality led to a narrower gap that was realized partially by increasing injury-related mortality among whites.
-This work emphasizes the importance of understanding how the direction of changes in mortality by race results in changes to the life expectancy gap.
+Life expectancy for both blacks and whites has improved substantially in the US since 1969.
+For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivaled by an increasing gap in CVD-related mortality, and highlight the cause-specific interventions necessary to decrease the black-white gap further.
+In women, a crossover in injury-related mortality led to a narrower gap that was realized partially by increasing mortality among whites.
+This work emphasizes the importance of understanding how the direction of changes in specific causes of death by race results in changes to the life expectancy gap.
 
@@
 In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethnicity.
-Whites lived 3.4 years longer than Blacks [ref nchs_LE].
-While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.
+Whites lived 3.4 years longer than Blacks [ref MMWR].
+While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied [@fuchs2016black].
 
-During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [@harper2007trends].
+During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among women, due to differential improvements in heart disease [@harper2007trends].
 The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement.
-For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [@harper2014trends].
+For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years [@harper2014trends].
 
 The over-arching objective of this study is to describe the pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences.
-To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for males and females.
+To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for men and women.
 This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states.
@@
 The population data were derived by the NCHS using intercensal bridged-race population estimates.
-This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death.
+This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death.
 Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013).
@@
 These states contained less than 0.4% of the black population in the United States in 2010 [ref Census_map].
-We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since the quality of the coding of ethnicity varied by state over the time period included [ref cdc_hisp].
+We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.
+Furthermore, the agreement between self-reported ethinicity on the Census  and secondary-reported ethinicity on the death certificate has been shown to be poorer during the earlier time period and differential according to geography, which could induce measurement error into mortality and disparity estimates [ref cdc_validity_race].
 
@@
 
-The data structure produced two statistical challenges that were addressed by our quantitative approach.
+The data structure produced two statistical challenges that were addressed by our modeling strategy.
 First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9).
@@
 
-To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [@selvin_2004; @preston].
+To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [@preston; @selvin_2004].
 Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex.
@@
 If the decomposition produces a negative estimate, say -0.5 years for injuries, this implies that whites have a higher age-standardized rate of mortality for injuries compared to blacks.
-Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 45 state-level trends and their credible intervals.
+Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 40 state-level trends and their credible intervals.
 
@@
 
-**Figures 1 and 2** display trends in the life expectancy gap for males and females, respectively.
-For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males.
+**Figures 1 and 2** display trends in the life expectancy gap for men and women, respectively.
+For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among men.
 
-*SH says: Don't know whether you want to note it here or not, but I find it is interesting to see that the gap is now lower in many US southern (or southeastern) states than in the Midwest, suggesting that progress has actually been somewhat greater in the South than elsewhere (particularly the upper Midwest and especially for women).
-I think this is interesting and somewhat counter to the narrative that health indicators are worse in the South.
-I also wonder how this might relate to the Murray/IHME work on US counties with "declining" life expectancy among white women.*
-
-**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for males and females, in years.
+**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for men and women, in years.
 The blue line illustrates the mean of the smoothed contribution estimated for each state-year, and illustrates the heterogeneity across states in their contribution trends.
@@
 Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races.
-Mortality from CVD decreased markedly since 1969 for both sexes and races (**eFigures 1 and 2**).
-However, white males underwent relatively larger mortality decreases than black males, resulting in a modest increase to the contribution of CVD to the life expectancy gap for most states (**Figure 3**).
-For females, black women experienced relatively larger mortality decreases over time compared to white females, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**).
-For both genders, Michigan, Wisconsin, California, and DC exhibit recent contributions that are relatively larger than the general pattern, while Massachusetts and New York exhibit persistently lower levels of CVD contributions.
+Mortality from CVD decreased markedly since 1969 for all groups (**eFigures 1 and 2**).
+However, white men underwent relatively larger mortality decreases than black men, resulting in a modest increase to the contribution of CVD to the life expectancy gap for most states (**Figure 3**).
+Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**).
+For both genders, Michigan, Wisconsin, California, and DC exhibit recent contributions that are relatively greater than the general pattern, while Massachusetts and New York exhibit persistently lower CVD contributions.
 
@@
 
-For both races and both sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**).
-Among males, the peak mortality rate was much higher among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**).
-For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern in mortality is similar between blacks and whites resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**).
-For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among males, but not females), while New York and Massachusetts underwent large decreases in the amount that cancer contributed to the life expectancy gap.
+For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**).
+Among men, the peak mortality rate was much higher among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white men and in cancer's contribution to the life expectancy gap (**Figure 5**).
+For women, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern in mortality is similar between blacks and whites resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**).
+For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.
 
@@
 
-Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**eFigure 5**).
-Recently, there has been a slight decrease in mortality rates, especially among black males.
-These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure 6**).
+Mortality from non-communicable disease has been increasing over time for men in most states, with DC as a notable exception (**eFigure 5**).
+Recently, there has been a slight decrease in mortality rates, especially among black men.
+These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure 6**).
 As such, the contribution of non-communicable disease to the mortality gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**).
-As with previous causes, New York and Massachusetts exhibit decreasing contributions, and Nevada has also made substantial progress.
+As with other causes, New York and Massachusetts exhibit decreasing contributions, and Nevada has also made substantial progress.
 Furthermore, while DC's mortality rates decreased, they still have the highest contribution of non-communicable disease to the gap.
@@
 Injury is one of the least common causes of death.
-Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure 7**).
-For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure 8**).
-Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white females demonstrating sharper increases.
-Thus, the contribution of injuries to the mortality gap has generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures 9 and 10**).
-Among males, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions.
-Among females, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.
+Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of criminal assault resulted in heightened mortality from injuries among black men during the 1990s (**eFigure 7**).
+For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**eFigure 8**).
+Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white women demonstrating sharper increases.
+The contribution of injuries to the mortality gap has therefore generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures 9 and 10**).
+Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions.
+Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.
 
@@
 
-Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**eFigures 9 and 10**).
-This spike was greater among males, especially blacks, and within states with larger urban centers.
-Thus, the contribution of communicable disease to the life expectancy gap exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**).
-DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among females.
+Communicable disease is the rarest cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**eFigures 9 and 10**).
+This spike was greater among men, especially blacks, and within states with larger urban centers.
+The contribution of communicable disease to the life expectancy gap therefore exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**).
+DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.
 
@@
 
-*CR note: I just added the following paragraph.
-Open to how/whether we want to talk about this.
-Thought we need to say something about how "all other causes partiallly reflects infant mortality.
-Also these additional figures (Fig 15 and 16) would be moved to the appendix.*
-
 To investigate how much of the gap in all other causes occurred during the first year of life, we conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred *during the first year of life* to the life expectancy gap (**Figures 15 and 16**).
-We investigated this because the majority of infant deaths are classified as being due to all other causes, conflating the all other causes category with inquities in infant mortality.
-We found that 43% of the contribution of all other causes to the gap was rooted in differences during the first year of life.
+The motivation for this analysis is that the majority of infant deaths are classified as being due to all other causes, conflating the all other causes category with inequities in infant mortality.
+We found that 43% of the contribution of all other causes to the gap was due to differences during the first year of life in 1969.
 In 2013, 53% of these deaths occurred during the first year of life.
@@
 
-**Figure 1: State-level trends in the difference in life expectancy between blacks and whites in males, United States, 1969-2013**
+**Figure 1: State-level trends in the difference in life expectancy between blacks and whites in men, United States, 1969-2013**
 
@@
 
-**Figure 2: State-level trends in the difference in life expectancy between blacks and whites in females, United States, 1969-2013**
+**Figure 1 Footnote:** The bands illustrate the 95% credible interval around the mean trend line.
+States with more imprecise estimates (often due to small black populations) have wider credible intervals.
+
+**Figure 2: State-level trends in the difference in life expectancy between blacks and whites in women, United States, 1969-2013**
 
@@
 
-**Figure 3: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in males**
+**Figure 3: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in men, 1969-2013**
 
@@
 
-**Figure 4: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in females**
+**Figure 4: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in women, 1969-2013**
 
@@
 
-**Figure 5: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in males**
+**Figure 5: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in men, 1969-2013**
 
@@
 
-**Figure 6: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in females**
+**Figure 6: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in women, 1969-2013**
 
@@
 
-**Figure 7: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the overall pattern in males**
+**Figure 7: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the overall pattern in men, 1969-2013**
 
@@
 
-**Figure 8: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the overall pattern in females**
+**Figure 8: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the overall pattern in women, 1969-2013**
 
@@
 
-**Figure 9: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the overall pattern in males**
+**Figure 9: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the overall pattern in men, 1969-2013**
 
@@
 
-**Figure 10: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the overall pattern in females**
+**Figure 10: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the overall pattern in women, 1969-2013**
 
@@
 
-**Figure 11: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the overall pattern in males**
+**Figure 11: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the overall pattern in men, 1969-2013**
 
@@
 
-**Figure 12: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the overall pattern in females**
+**Figure 12: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the overall pattern in women, 1969-2013**
 
@@
 
-**Figure 13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males**
+**Figure 13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in men, 1969-2013**
 
@@
 
-**Figure 14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in females**
+**Figure 14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in women, 1969-2013**
 
@@
 
-**Figure 15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males younger than one**
+**Figure 15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in men younger than one, 1969-2013**
 
@@
 
-**Figure 16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in females younger than one**
+**Figure 16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in women younger than one, 1969-2013**
 
@@
 
-In this paper, we examined state-level trends in six major CODs' contributions to the black-white life expectancy gap.
-While we found that many states mimicked the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result.
+We examined state-level trends in six major cause of death contributions to the black-white life expectancy gaps over a 45 year period.
+While we found that many states conformed to the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation.
+In contrast to some recent work focused on county-level inequalities [@arora2016population; @dwyer2016us], states have some substantial advantages as analysis units.
+They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.
 
-We found that cardiovascular disease contributes the most to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
-Injuries continue to play a role in contributing to the current gap for males in several states, especially in the Midwest and DC, while contributing to a decreased gap among females in the South.
+Overall, the racial disparity in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century.
+[add: Lemann, 1991] Trends for women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
+On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy [@dwyer2016us].
+On the other hand, greater racial equality can be achieved by falling black rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.
+[add: Case and Deaton, 2017]
+
+Consistent with what has been reported at the national level, we found that cardiovascular disease contributes the most to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time [@harper2012trends].
+Injuries continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South.
 Illinois, Michigan and Wisconsin also show relatively high contributions from cancer.
@@
 The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contribution of more than two years.
-Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing minimal difference and point to the importance of viewing the most recent estimates in light of their earlier values.
+Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing pattern of equity and points to the importance of viewing the most recent estimates in light of their earlier values.
 
-**Literature review**
-
-This research extends previous studies that have estimated either the contribution of causes of death to the life expectancy gap on the national level, or the excess risk among blacks in cause of death specific mortality rates [@harper2007trends; @harper2012trends; orsi2010black] [ref kochanek_2013].
+The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.
+States pursue varying approaches to an array of relevant programs and laws including tobacco control [@cotti2016effects], Medicare coverage [@grogan2017racial], hand-gun violence [@humphreys2017evaluating] and nutrition [@taber2011state].
+These presumably affect health and longevity, and can have especially important impacts on racial disparities when the policies are targeted to disadvantaged groups or outcomes that differentially effect blacks [@adimora2014policies].
+The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
+Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.[add: Strumpf FE Chapter in MSE]
 
@@
 
-We used data over 45 years, across two race/ethnicities, and for all age groups.
+We used data over 45 years, across two racial groups, and for all ages.
 By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year.
-This allowed us to recognize when the estimated contributions for a particular cause were relatively low *because they started low* vs.
-low *because they had decreased over time*.
+This allowed us to recognize when the estimated disparity contributions for a particular cause were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time.
 Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
-This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group, which may present a distorted reality [Case & Deaton, Minton IJE 2016, ].
+This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group [@case2015rising; @minton2016two].
+The Arriaga decomposition method accounts for the fact that deaths occurring in one age or cause stratum are reflected in the absence of that death in another stratum, creating interactions between cells.
 Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.
@@
 Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state.
-We decided against excluding or adjusting for Hispanic ethnicity in this study because the indicator for ethnicity has been noted to have varying accuracy across states and during the earlier time period [ref].
-Thus, states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded [ref].
-In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results [ref Harper 2012], providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings.
+This may be especially consequential for blacks in states like Florida or New York that received many black Hispanic immigrants from Cuba, Puerto Rico and the Dominican Republic, potentially changing the composition of the state’s black population to some extent during the study period [@laveist2011black].
+But it also has implications for the white population, especially in states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded [@ruiz2013hispanic].
+We decided against excluding or adjusting for Hispanic ethnicity in this study because the indicator for ethnicity on the death certificate has been noted to have varying accuracy across states and during the earlier time period [ref].
+In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results [@harper2014trends], providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings overall.
+As the US population becomes increasingly Hispanic over time, however, this distinction grows more salient [@brazil2015spatial].
 
-Secondly, our analysis does not account for immigration, or migration between states.
-Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [ref Preston 2014].
+Secondly, our analysis does not account for immigration by the foriegn-born, or migration of native-born individuals between states.
+Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [@preston2014anatomy].
 As we are examine the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white.
@@
 Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
-Such changes would both serve to increase the racial gap in DC.
+Such changes would both serve to increase the racial gap in DC [add: US Census bureau, 2015].
 
@@
 For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.
+Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses [@kuehn2014cdc; @jones2013pharmaceutical].
 
@@
 
+Racial disparities have been a fixture of American life for centuries, but their continual evolution is a topic of great relevance for public policy and social justice.
+The careful surveillance of these patterns is therefore a crucial public health activity, and this paper presents a unique perspective on state-level variation over nearly half a century.
 This analysis provides a foundation for future work to investigate time-varying and time-fixed state-level characteristics that may be responsible for differences among state in risk factors leading to each cause of death.
-Differences in smoking policy, medicare coverage, access to care, segregation, and racism, for example, could all affect risk factors for ill health and vary in magnitude by state.
+Differences in smoking policy, Medicare coverage, access to care, segregation, and racism, for example, could all affect risk factors for ill health and vary in magnitude by state.
 
@@
 to add to bibliography list:
-nchs_LE ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf
+MMWR: Cunningham TJ, Croft JB, Liu Y, Lu H, Eke PI, Giles WH.
+Vital Signs: Racial Disparities in Age-Specific Mortality Among Blacks or African Americans - United States, 1999-2015.
+MMWR Morb Mortal Wkly Rep.
+2017 May 5;66(17):444-456.
+PMID: 28472021
 
@@
 
+cdc_validity_race: https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf
+
 census_map: https://www.census.gov/2010census/popmap/. retrieved apr 25, 2017
@@
 kochanek_2013: https://www.cdc.gov/nchs/data/databriefs/db125.pdf
+
+Lemann, Nicholas. The Promised Land: The Great Black Migration and How It Changed America. New York: Alfred A. Knopf, 1991. ISBN 0-394-56004-3.
+
+Case A, Deaton A. Mortality and morbidity in the 21st century. Brookings Working Paper, 2017.
+
+Strumpf E, Harper S, Kaufman J. Fixed Effects and Difference-in-Differences. In: Methods in Social Epidemiology (Oakes & Kaufman, Eds) Jossey-Bass: John Wiley & Sons, 2017.
+
+U.S.
+Census Bureau (2015).
+American Community Survey 1-year estimates.
+Retrieved from Census Reporter Profile page for Washington, DC <https://censusreporter.org/profiles/16000US1150000-washington-dc/>

### 2017-05-15 Manuscript/Manuscript_COD_Trends.Rmd
SH revisions

@@
 
-The life expectancy of a population is a barometer for overall health.
-In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethnicity.
+Life expectancy is a barometer for overall population health.
+In 2014, life expectancy at birth in the United States was 78.8 years, but hidden within this summary are important subgroups differences by self-defined racial ethnicity.
 Whites lived 3.4 years longer than Blacks [ref MMWR].
-While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied [@fuchs2016black].
+While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied [@fuchs2016black].
 
@@
 For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years [@harper2014trends].
+However, the extent to which different causes of death may have contributed to state-level changes in the black-white gap are unknown.
 
-The over-arching objective of this study is to describe the pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences.
+The over-arching objective of this study is to describe the contribution of selected causes of death to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences.
 To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for men and women.
@@
 The population data were derived by the NCHS using intercensal bridged-race population estimates.
-This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death.
+We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death.
 Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013).
 These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**).
-When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.
 
@@
 These states contained less than 0.4% of the black population in the United States in 2010 [ref Census_map].
-We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.
-Furthermore, the agreement between self-reported ethinicity on the Census  and secondary-reported ethinicity on the death certificate has been shown to be poorer during the earlier time period and differential according to geography, which could induce measurement error into mortality and disparity estimates [ref cdc_validity_race].
+We coulnd not consider Hispanic origin to identify race according to Hispanic ethnicity, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.
+*Furthermore, the agreement between self-reported ethnicity on the Census and secondary-reported ethinicity on the death certificate has been shown to be poorer during the earlier time period and differential according to geography, which could induce measurement error into mortality and disparity estimates [ref cdc_validity_race].*
 
@@
 The data structure produced two statistical challenges that were addressed by our modeling strategy.
-First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9).
+First, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9 [ref NCHS or NCI page on suppression], we needed to impute the suppressed death counts (for counts between 1 and 9).
 Second, several states had small black populations which produced sparse counts in each sex-age category.
 In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic.
-To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.
+To better estimate the underlying age-specific mortality rates, we used a method that produces smoother age-specific mortality trends.
 
@@
 
-To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts.
-Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population.
-The mortality rates over time within each state-age-sex-race category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year.
+We used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts.
+Specifically, we used Poisson regression to model the number of deaths in each state-year-sex-race-age-cause category as a function of the at-risk population.
+The mortality rates over time within each state-year-sex-race-age-cause category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year.
 Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9).
@@
 
-To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [@preston; @selvin_2004].
-Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex.
+To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied [@preston; @selvin_2004].
+Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender.
 The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.
@@
 
-To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimate from the previous step as a function of a state random effect and a linear spline for year (estimated using four internal knots).
+To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimated from the previous step as a function of a state random effect and a linear spline for year (estimated using four internal knots).
 The model was inversely-weighted by the standard error of the mean estimate in each state-year.
-To identify divergent states, we compared the state-specific trends estimate in the previous step to the national pattern estimate by setting the random effect to zero.
+To identify states divergent from the national trend, we compared the state-specific trends estimate in the previous step to the national pattern estimate by setting the random effect to zero.
 
@@
 **Figures 1 and 2** display trends in the life expectancy gap for men and women, respectively.
-For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among men.
+For men and women, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among men.
 
-**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for men and women, in years.
-The blue line illustrates the mean of the smoothed contribution estimated for each state-year, and illustrates the heterogeneity across states in their contribution trends.
+**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for men and women, in years, for each state.
+The blue line illustrates the mean of the smoothed contribution estimated for each state, and illustrates the heterogeneity across states in their contribution trends.
 These lines are overlaid with the meta-estimated common trend, shown with a red line.
@@
 Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races.
-Mortality from CVD decreased markedly since 1969 for all groups (**eFigures 1 and 2**).
-However, white men underwent relatively larger mortality decreases than black men, resulting in a modest increase to the contribution of CVD to the life expectancy gap for most states (**Figure 3**).
+Mortality from CVD decreased markedly since 1969 for all groups (**eFigures 1 and 2**), but white men underwent relatively larger mortality decreases than black men.
+As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states (**Figure 3**).
 Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**).
-For both genders, Michigan, Wisconsin, California, and DC exhibit recent contributions that are relatively greater than the general pattern, while Massachusetts and New York exhibit persistently lower CVD contributions.
+For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.
 
@@
 For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**).
-Among men, the peak mortality rate was much higher among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white men and in cancer's contribution to the life expectancy gap (**Figure 5**).
-For women, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern in mortality is similar between blacks and whites resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**).
+Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 5**).
+For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**).
 For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.
@@
 These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure 6**).
-As such, the contribution of non-communicable disease to the mortality gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**).
+As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**).
 As with other causes, New York and Massachusetts exhibit decreasing contributions, and Nevada has also made substantial progress.
@@
 Injury is one of the least common causes of death.
-Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of criminal assault resulted in heightened mortality from injuries among black men during the 1990s (**eFigure 7**).
+Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**eFigure 7**).
 For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**eFigure 8**).
-Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white women demonstrating sharper increases.
-The contribution of injuries to the mortality gap has therefore generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures 9 and 10**).
+Thus, injury mortality rates among blacks are equivalent, or lower than those among whites in many states, with white women demonstrating sharper recent increases.
+The contribution of injuries to the mortality gap has therefore generally decreased over time, and has become negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 9 and 10**).
 Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions.
@@
 Communicable disease is the rarest cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**eFigures 9 and 10**).
-This spike was greater among men, especially blacks, and within states with larger urban centers.
-The contribution of communicable disease to the life expectancy gap therefore exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**).
+This spike was greater among men, especially blacks, and among states with larger urban centers.
+The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**).
 DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.
@@
 Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**eFigures 11 and 12**).
-White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, while black men have continued to experience decreasing rates within several states.
-As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and reached equality in New York, Massachusetts, and Rhode Island for both genders (**Figures 13 and 14**).
+White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states.
+As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 13 and 14**).
 The most disparate states include Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.
 
-To investigate how much of the gap in all other causes occurred during the first year of life, we conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred *during the first year of life* to the life expectancy gap (**Figures 15 and 16**).
-The motivation for this analysis is that the majority of infant deaths are classified as being due to all other causes, conflating the all other causes category with inequities in infant mortality.
-We found that 43% of the contribution of all other causes to the gap was due to differences during the first year of life in 1969.
-In 2013, 53% of these deaths occurred during the first year of life.
-Thus, the trend in the all other causes contribution partially reflects a decreasing disparity in infant mortality.
+Because the majority of infant deaths are classified as being due to all other causes in our analysis, we also tried to separate any impact of all other causes from those of infant mortality more generally.
+We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (Figures 15 and 16).
+We found that 43% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969.
+In 2013, this proportion reached 53%.
+Thus, the trend in the all other causes contribution partially reflects a decreasing contribution from in infant mortality.
 
@@
 
-Overall, the racial disparity in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century.
-[add: Lemann, 1991] Trends for women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
+Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century.
+[add: Lemann, 1991] Mortality trends for women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
 On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy [@dwyer2016us].
@@
 
-Consistent with what has been reported at the national level, we found that cardiovascular disease contributes the most to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time [@harper2012trends].
-Injuries continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South.
+Consistent with what has been reported at the national level, we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time [@harper2012trends].
+Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South.
 Illinois, Michigan and Wisconsin also show relatively high contributions from cancer.
@@
 The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contribution of more than two years.
-Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing pattern of equity and points to the importance of viewing the most recent estimates in light of their earlier values.
+Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.
 
@@
 States pursue varying approaches to an array of relevant programs and laws including tobacco control [@cotti2016effects], Medicare coverage [@grogan2017racial], hand-gun violence [@humphreys2017evaluating] and nutrition [@taber2011state].
-These presumably affect health and longevity, and can have especially important impacts on racial disparities when the policies are targeted to disadvantaged groups or outcomes that differentially effect blacks [@adimora2014policies].
+These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially effect blacks [@adimora2014policies].
 The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
@@
 By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year.
-This allowed us to recognize when the estimated disparity contributions for a particular cause were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time.
+This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time.
 Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
 This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group [@case2015rising; @minton2016two].
-The Arriaga decomposition method accounts for the fact that deaths occurring in one age or cause stratum are reflected in the absence of that death in another stratum, creating interactions between cells.
+The Arriaga decomposition method accounts for the fact that deaths occurring in one age or cause stratum are reflected in the absence of that death in another stratum.
 Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.
@@
 But it also has implications for the white population, especially in states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded [@ruiz2013hispanic].
-We decided against excluding or adjusting for Hispanic ethnicity in this study because the indicator for ethnicity on the death certificate has been noted to have varying accuracy across states and during the earlier time period [ref].
+We were unable to adjust for Hispanic ethnicity in this study because the indicator for ethnicity on the death certificate has been noted to have varying accuracy across states and during the earlier time period [ref].
 In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results [@harper2014trends], providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings overall.
@@
 
-Secondly, our analysis does not account for immigration by the foriegn-born, or migration of native-born individuals between states.
+Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
 Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [@preston2014anatomy].
-As we are examine the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white.
-Migration between states can also affect the estimated life expectancy of a state.
+As we are examining the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white.
+Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death.
 Migration is particularly important for the District of Columbia, because of it's small geographical bounds, and substantial in-migration from New York, in- and out-migration to Maryland and Virginia.
@@
 
-Racial disparities have been a fixture of American life for centuries, but their continual evolution is a topic of great relevance for public policy and social justice.
+Racial differences have been a fixture of American life for centuries, but their continued evolution is a topic of great relevance for public policy and social justice.
 The careful surveillance of these patterns is therefore a crucial public health activity, and this paper presents a unique perspective on state-level variation over nearly half a century.

### 2017-05-16 Code/2_entire_analysis.R
more SH revisions

@@ -1,6 +1,6 @@
 entire.analysis <- function(state_i, chosen.seed) {
   
-  #load_libraries
+  #load libraries
   library(Rcpp)
   library(dplyr)
   library(purrr)
@@ -19,17 +19,17 @@ entire.analysis <- function(state_i, chosen.seed) {
   library(gridExtra)
   library(parallel)
   
-  #load_user_written_functions
+  #load user written functions
   source("~/repos/BlackWhiteMortalityGap/Code/2_smoothing_models/bayes_smoothing_functions.R")
   source('~/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R') 
   
-  #load_tidied_data
+  #load tidied data
   dat.clean = read.csv('~/dat_clean.csv') 
   
   #load US standard population
   std.pop <- read.csv("~/repos/BlackWhiteMortalityGap/Data/US_Standard_Population_for_model.csv") %>% select(US_Standard_2000, age.weight, age_minbin)
   
-  #fit_models
+  #fit models
   #run time: 32 mins
   #fit the bayesian models and put the 1000 posterior samples into a data frame
   for(sex_i in c("Male", "Female")){

### 2017-05-16 Manuscript/Manuscript_COD_Trends.Rmd

@@
 For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming.
-These states contained less than 0.4% of the black population in the United States in 2010 [ref Census_map].
+These states contained less than 0.4% of the black population in the United States in 2010 [@census_map].
 We coulnd not consider Hispanic origin to identify race according to Hispanic ethnicity, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.
@@
 
-These models were run separately for every combination of state, race, and sex.
+These models were run separately for every combination of state, race, gender, and cause of death.
 They were fit using two Markov chains of length 10,000.
-One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration (to reduce autocorrelation).
+One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration to reduce autocorrelation.
 Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition.
@@
 
-Injury is one of the least common causes of death.
+Trends in mortalilty from injuries differ for blacks and whites.
 Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**eFigure 7**).
@@
 
-Communicable disease is the rarest cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**eFigures 9 and 10**).
+Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**eFigures 9 and 10**).
 This spike was greater among men, especially blacks, and among states with larger urban centers.
@@
 
-Consistent with what has been reported at the national level, we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time [@harper2012trends].
+Consistent with what has been reported at the national level[@harper2012trends], we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
 Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South.
+This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic [ref MMWR_opioids].
+
 Illinois, Michigan and Wisconsin also show relatively high contributions from cancer.
@@
 The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
-Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.[add: Strumpf FE Chapter in MSE]
+Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.[@strumpfMSE]
 
@@
 
-census_map: https://www.census.gov/2010census/popmap/. retrieved apr 25, 2017
-
 census_std_pop: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html. retrieved apr 25, 2017
@@
 Retrieved from Census Reporter Profile page for Washington, DC <https://censusreporter.org/profiles/16000US1150000-washington-dc/>
+
+MMWR opioids: https://www.cdc.gov/mmwr/volumes/65/wr/mm655051e1.htm Increases in Drug and Opioid-Involved Overdose Deaths — United States, 2010–2015 Rose Rudd et al.

### 2017-05-16 Manuscript/Manuscript_COD_Trends.Rmd
manual bibliography

@@
 In 2014, life expectancy at birth in the United States was 78.8 years, but hidden within this summary are important subgroups differences by self-defined racial ethnicity.
-Whites lived 3.4 years longer than Blacks [ref MMWR].
-While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied [@fuchs2016black].
+Whites lived 3.4 years longer than Blacks.^1^ While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.^2^
 
-During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among women, due to differential improvements in heart disease [@harper2007trends].
-The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement.
-For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years [@harper2014trends].
-However, the extent to which different causes of death may have contributed to state-level changes in the black-white gap are unknown.
+During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among women, due to differential improvements in heart disease.^3^ The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement.
+For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years.^4^ However, the extent to which different causes of death may have contributed to state-level changes in the black-white gap are unknown.
 
@@
 
-We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013[ref: seer stat].
-The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System [ref: nchs data].
-The population data were derived by the NCHS using intercensal bridged-race population estimates.
+We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.^5^ The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.^6^ The population data were derived by the NCHS using intercensal bridged-race population estimates.
 We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death.
@@
 For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming.
-These states contained less than 0.4% of the black population in the United States in 2010 [@census_map].
-We coulnd not consider Hispanic origin to identify race according to Hispanic ethnicity, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.
-*Furthermore, the agreement between self-reported ethnicity on the Census and secondary-reported ethinicity on the death certificate has been shown to be poorer during the earlier time period and differential according to geography, which could induce measurement error into mortality and disparity estimates [ref cdc_validity_race].*
+These states contained less than 0.4% of the black population in the United States in 2010.^7^ We could not consider Hispanic origin to identify race according to Hispanic ethnicity, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.
+*Furthermore, the agreement between self-reported ethnicity on the Census and secondary-reported ethinicity on the death certificate has been shown to be poorer during the earlier time period and differential according to geography, which could induce measurement error into mortality and disparity estimates.^8^*
 
@@
 The data structure produced two statistical challenges that were addressed by our modeling strategy.
-First, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9 [ref NCHS or NCI page on suppression], we needed to impute the suppressed death counts (for counts between 1 and 9).
+First, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^9^ we needed to impute the suppressed death counts (for counts between 1 and 9).
 Second, several states had small black populations which produced sparse counts in each sex-age category.
-In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic.
+In these cases, the estimation of the gender-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic.
 To better estimate the underlying age-specific mortality rates, we used a method that produces smoother age-specific mortality trends.
@@
 
-To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied [@preston; @selvin_2004].
-Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender.
+To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied.^10, 11^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender.
 The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.
 
-To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method [@arriaga1984measuring].
-This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher age-standardized cause-specific mortality rates among blacks.
+To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method.^12^ This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher age-standardized cause-specific mortality rates among blacks.
 For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher age-standardized rate of mortality due to cardiovascular disease among blacks compared to whites.
@@
 
-We also calculated age-standardized mortality rates for year-state-race-sex stratum, to aid the interpretation of the decomposition results.
-We used the US standard population from the year 2000 [ref census_stdpop].
+We also calculated age-standardized mortality rates for year-state-race-sex stratum, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000.^13^
 
@@
 While we found that many states conformed to the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation.
-In contrast to some recent work focused on county-level inequalities [@arora2016population; @dwyer2016us], states have some substantial advantages as analysis units.
+In contrast to some recent work focused on county-level inequalities ,^14,15^ states have some substantial advantages as analysis units.
 They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.
 
-Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century.
-[add: Lemann, 1991] Mortality trends for women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
-On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy [@dwyer2016us].
-On the other hand, greater racial equality can be achieved by falling black rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.
-[add: Case and Deaton, 2017]
+Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century.^16^  Mortality trends for women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
+On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, greater racial equality can be achieved by falling black rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^17^
 
-Consistent with what has been reported at the national level[@harper2012trends], we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
+Consistent with what has been reported at the national level,^18^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
 Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South.
-This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic [ref MMWR_opioids].
+This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^19^
 
@@
 The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.
-States pursue varying approaches to an array of relevant programs and laws including tobacco control [@cotti2016effects], Medicare coverage [@grogan2017racial], hand-gun violence [@humphreys2017evaluating] and nutrition [@taber2011state].
-These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially effect blacks [@adimora2014policies].
-The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
-Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.[@strumpfMSE]
+States pursue varying approaches to an array of relevant programs and laws including tobacco control,^20^ Medicare coverage,^21^ hand-gun violence,^22^  and nutrition.^23^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially effect blacks.^24^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
+Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.^25^
 
@@
 Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
-This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group [@case2015rising; @minton2016two].
-The Arriaga decomposition method accounts for the fact that deaths occurring in one age or cause stratum are reflected in the absence of that death in another stratum.
+This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^26, 27^ The Arriaga decomposition method accounts for the fact that deaths occurring in one age or cause stratum are reflected in the absence of that death in another stratum.
 Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.
@@
 Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state.
-This may be especially consequential for blacks in states like Florida or New York that received many black Hispanic immigrants from Cuba, Puerto Rico and the Dominican Republic, potentially changing the composition of the state’s black population to some extent during the study period [@laveist2011black].
-But it also has implications for the white population, especially in states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded [@ruiz2013hispanic].
-We were unable to adjust for Hispanic ethnicity in this study because the indicator for ethnicity on the death certificate has been noted to have varying accuracy across states and during the earlier time period [ref].
-In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results [@harper2014trends], providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings overall.
-As the US population becomes increasingly Hispanic over time, however, this distinction grows more salient [@brazil2015spatial].
+This may be especially consequential for blacks in states like Florida or New York that received many black Hispanic immigrants from Cuba, Puerto Rico and the Dominican Republic, potentially changing the composition of the state’s black population to some extent during the study period.^28^  But it also has implications for the white population, especially in states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded.^29^ We were unable to adjust for Hispanic ethnicity in this study because the indicator for ethnicity on the death certificate has been noted to have varying accuracy across states and during the earlier time period.^8^ In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results,^4^ providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings overall.
+As the US population becomes increasingly Hispanic over time, however, this distinction grows more salient.^30^
 
 Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
-Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [@preston2014anatomy].
-As we are examining the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white.
+Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010.^31^ As we are examining the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white.
 Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death.
@@
 Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
-Such changes would both serve to increase the racial gap in DC [add: US Census bureau, 2015].
+Such changes would both serve to increase the racial gap in DC.^32^
 
@@
 For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.
-Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses [@kuehn2014cdc; @jones2013pharmaceutical].
+Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^33,34^
 
@@
 
-to add to bibliography list:
-MMWR: Cunningham TJ, Croft JB, Liu Y, Lu H, Eke PI, Giles WH.
+1.
+Cunningham TJ, Croft JB, Liu Y, Lu H, Eke PI, Giles WH.
 Vital Signs: Racial Disparities in Age-Specific Mortality Among Blacks or African Americans - United States, 1999-2015.
 MMWR Morb Mortal Wkly Rep.
-2017 May 5;66(17):444-456.
-PMID: 28472021
+2017;66(17):444-456.
+doi: 10.15585/mmwr.mm6617e1.
 
-seer stat: National Cancer Institute. Surveillance, Epidemiology, and End Results
-Program: SEER*Stat software: latest release: version 8.1.5 [Internet]. Bethesda (MD): NCI; 2013 Mar 31 [cited 2014 Jun 20]. Available from: http://www.seer.cancer.gov/seerstat
+2. Fuchs, VR. Black Gains in Life Expectancy. JAMA. 2016;316(18): 1869–70. doi: 10.1001/jama.2016.14398.
 
-nchs data: National Cancer Institute.
+3. Harper S, Lynch J, Burris S, Smith GD. Trends in the black-white life expectancy gap in the United States, 1983-2003. JAMA. 2007;297(11):1224-32.
+
+4. Harper, Sam, Richard F MacLehose, and Jay S Kaufman. 2014. “Trends in the Black-White Life Expectancy Gap Among Us States, 1990–2009.” Health Affairs 33 (8). Health Affairs: 1375–82.
+
+5. National Cancer Institute. Surveillance, Epidemiology, and End Results
+Program: SEER*Stat software: release: version 8.3.2 [Internet]. Bethesda (MD): NCI; 2017 Apr 14 [cited 2016 August 20]. Available from: http://www.seer.cancer.gov/seerstat.
+
+6.
+Centers for Disease Control and Prevention.
+National Center for Health Statistics: National Vital Statistics System [Internet].
+[cited 2017 May 16].
+Available from: https://www.cdc.gov/nchs/nvss/index.htm.
+
+7. U.S. Census Bureau: United States Census 2010 Interactive Population Map [Internet]. [cited 2017 May 16]. Available from: https://www.census.gov/2010census/popmap/.
+
+8.
+Arias E, Schauman WS, Eschbach K, Sorlie PD, Backlund E.
+The validity of race and Hispanic origin reporting on death certificates in the United States.
+National Center for Health Statistics.
+Vital Health Stat 2(148).
+2008.
+Available from: https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf
+
+9.
+National Cancer Institute.
 Surveillance, Epidemiology, and End Results Program: US mortality data [Internet].
-Bethesda (MD): NCI; [cited 2014 Jun 20].
+Bethesda (MD): NCI; [cited 2017 May 16].
 Available from: http://www.seer.cancer.gov/mortality/
 
-cdc_hisp: [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf]
+10. Preston SH, Heuveline P, and Guillot M. Demography: Measuring and Modeling Population Processes. Malden: Blackwell Publishing; 2001.
 
-cdc_validity_race: https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf
+11. Selvin, S. Statistical Analysis of Epidemiologic Data. 3rd ed. Oxford University Press; 2004.
 
-census_std_pop: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html. retrieved apr 25, 2017
+12. Arriaga EE. Measuring and explaining the change in life expectancies. Demography. 1984;21(1):83-96.
 
-kochanek_2013: https://www.cdc.gov/nchs/data/databriefs/db125.pdf
+13.
+National Cancer Institute.
+Surveillance, Epidemiology, and End Results Program: Standard Populations - 19 Age Groups [Internet].
+[cited 2017 April 25].
+Available from: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html.
 
-Lemann, Nicholas. The Promised Land: The Great Black Migration and How It Changed America. New York: Alfred A. Knopf, 1991. ISBN 0-394-56004-3.
+14.
+Arora A, Spatz E, Herrin J, Riley C, Roy B, Kell K, Coberley C, Rula E, Krumholz HM.
+Population Well-Being Measures Help Explain Geographic Disparities In Life Expectancy At The County Level.
+Health Affairs.
+2016;35(11):2075-82.
 
-Case A, Deaton A. Mortality and morbidity in the 21st century. Brookings Working Paper, 2017.
+15.
+Dwyer-Lindgren L, Bertozzi-Villa A, Stubbs RW, Morozoff C, Kutz MJ, Huynh C, Barber RM, et al.
+US County-Level Trends in Mortality Rates for Major Causes of Death, 1980-2014.
+JAMA.
+2016;316(22):2385–2401.
 
-Strumpf E, Harper S, Kaufman J. Fixed Effects and Difference-in-Differences. In: Methods in Social Epidemiology (Oakes & Kaufman, Eds) Jossey-Bass: John Wiley & Sons, 2017.
+16. Lemann, N. The Promised Land: The Great Black Migration and How It Changed America. New York: Alfred A. Knopf, 1991. ISBN 0-394-56004-3.
 
+17. Case A, Deaton A. Mortality and Morbidity in the 21st Century. In: Eberly J and Stock J, Editors. Brookings Papers on Economic Activity, March 23-24 2017; 2017.
+
+18. Harper S, Rushani D, and Kaufman JS. Trends in the Black-White Life Expectancy Gap, 2003-2008. JAMA. 2012;307(21):2257–9.
+
+19.
+Rudd RA, Seth P, David F, Scholl L.
+Increases in Drug and Opioid-Involved Overdose Deaths — United States, 2010–2015.
+MMWR Morb Mortal Wkly Rep 2016;65:1445–1452.
+DOI: http://dx.doi.org/10.15585/mmwr.mm655051e1.
+
+20.
+Cotti C, Nesson E, Tefft N.
+The Effects of Tobacco Control Policies on Tobacco Products, Tar, and Nicotine Purchases among Adults: Evidence from Household Panel Data.
+American Economic Journal: Economic Policy.
+2016;8(4):103-23.
+
+21. Grogan CM, Park SE. The Racial Divide in State Medicaid Expansions. Journal of Health Politics, Policy and Law. 2017 Jun 1;42(3):539-72.
+
+22.
+Humphreys DK, Gasparrini A, Wiebe DJ.
+Evaluating the Impact of Florida’s “Stand Your Ground” Self-defense Law on Homicide and Suicide by Firearm: An Interrupted Time Series Study.
+JAMA internal medicine.
+2017;177(1):44-50.
+
+23.
+Taber DR, Stevens J, Evenson KR, Ward DS, Poole C, Maciejewski ML, Murray DM, Brownson RC.
+State policies targeting junk food in schools: racial/ethnic differences in the effect of policy change on soda consumption.
+American Journal of Public Health.
+2011;101(9):1769-75.
+
+24. Adimora AA, Ramirez C, Schoenbach VJ, Cohen MS. Policies and politics that promote HIV infection in the Southern United States. AIDS (London, England). 2014;28(10):1393.
+
+25. Strumpf EC, Harper S, Kaufman JS. Fixed Effects and Difference in Differences. In: Methods in Social Epidemiology. 2nd Ed. (Oakes & Kaufman, Eds) San Francisco: Jossey-Bass, 2017.
+
+26. Case A, Deaton A. Rising Morbidity and Mortality in Midlife among White Non-Hispanic Americans in the 21st Century. Proceedings of the National Academy of Sciences. 2015 Dec 8;112(49):15078-83.
+
+27.
+Minton J, Green M, McCartney G, Shaw R, Vanderbloemen L, Pickett K.
+Two cheers for a small giant?
+Why we need better ways of seeing data: A commentary on:‘Rising morbidity and mortality in midlife among White non-Hispanic Americans in the 21st century’.
+International Journal of Epidemiology.
+2016 Sep 17:dyw095.
+
+28.
+LaVeist-Ramos TA, Galarraga J, Thorpe RJ, Bell CN, Austin CJ.
+Are black Hispanics black or Hispanic?
+Exploring disparities at the intersection of race and ethnicity.
+Journal of Epidemiology and Community Health.
+2011 Mar 3:jech-2009.
+
+29. Ruiz JM, Steffen P, Smith TB. Hispanic mortality paradox: a systematic review and meta-analysis of the longitudinal literature. American Journal of Public Health. 2013 Mar;103(3):e52-60.
+
+30. Brazil N. Spatial Variation in the Hispanic Paradox: Mortality Rates in New and Established Hispanic US Destinations. Population, Space and Place. 2017;23:e1968. DOI:10.1002/psp.1968.
+
+31. Preston SH, Elo IT. Anatomy of a municipal triumph: New York City's upsurge in life expectancy. Population and development review. 2014;40(1):1-29.
+
+32.
 U.S.
-Census Bureau (2015).
-American Community Survey 1-year estimates.
-Retrieved from Census Reporter Profile page for Washington, DC <https://censusreporter.org/profiles/16000US1150000-washington-dc/>
+Census Bureau.
+Census Reporter Profile Page: American Community Survey 1-year estimates [Internet].
+[Cited 16 May 2017].
+Available from: https://censusreporter.org/profiles/16000US1150000-washington-dc/.
 
-MMWR opioids: https://www.cdc.gov/mmwr/volumes/65/wr/mm655051e1.htm Increases in Drug and Opioid-Involved Overdose Deaths — United States, 2010–2015 Rose Rudd et al.
+33. Kuehn BM. CDC: major disparities in opioid prescribing among states: some states crack down on excess prescribing. JAMA. 2014;312(7):684-6.
+
+34. Jones CM, Mack KA, Paulozzi LJ. Pharmaceutical overdose deaths, United States, 2010. JAMA. 2013;309(7):657-9.

### 2017-05-19 Manuscript/Appendix-agestd-mort-rates.Rmd
modified supp app

@@
 
-mortality.rates <- reorder.as.map(mortality.rates, "state", "stabbrs")
-mortality.rates.diff <- reorder.as.map(mortality.rates.diff, "state", "stabbrs")
+mortality.rates <- reorder.as.map2(mortality.rates, "state", "stabbrs")
+mortality.rates.diff <- reorder.as.map2(mortality.rates.diff, "state", "stabbrs")
 ```
@@
 
+    c$x$data[[i]]$line$width <- 1 #make all the lines thin
     c$x$data[[i]]$text <- gsub("rate.per.100k_mean", "Mean mortality rate", c$x$data[[i]]$text)
@@
 
+    d$x$data[[i]]$line$width <- 1 #make all the lines thin
     d$x$data[[i]]$text <- gsub("rate.difference_mean", "Mean excess mortality", d$x$data[[i]]$text)
@@
 
-**eFigure 1: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**
+**Figure S1: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**
 
-```{r print-em, fig.width = 9, fig.height = 9, echo = F}
-Cardiovascular.Male.mort.plot[[3]] %>% layout(margin = list(b = 100))
-Cardiovascular.Male.mort.plot[[4]]
+```{r print-em, fig.width = 9, fig.height = 7, echo = F}
+Cardiovascular.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+#Cardiovascular.Male.mort.plot[[4]]
 ```
 
-**eFigure 2: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**
+**Figure S2: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**
 
-```{r print-em-2, fig.width = 9, fig.height = 9, echo = F}
-Cardiovascular.Female.mort.plot[[3]] %>% layout(margin = list(b = 100))
-Cardiovascular.Female.mort.plot[[4]]
+```{r print-em-2, fig.width = 9, fig.height = 7, echo = F}
+Cardiovascular.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+#Cardiovascular.Female.mort.plot[[4]]
 ```
 
-**eFigure 3: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**
+**Figure S3: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**
 
-```{r print-em-3, fig.width = 9, fig.height = 9, echo = F}
-Cancers.Male.mort.plot[[3]] %>% layout(margin = list(b = 100))
-Cancers.Male.mort.plot[[4]]
+```{r print-em-3, fig.width = 9, fig.height = 7, echo = F}
+Cancers.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+#Cancers.Male.mort.plot[[4]]
 ```
 
-**eFigure 4: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**
+**Figure S4: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**
 
-```{r print-em-4, fig.width = 9, fig.height = 9, echo = F}
-Cancers.Female.mort.plot[[3]] %>% layout(margin = list(b = 100))
-Cancers.Female.mort.plot[[4]]
+```{r print-em-4, fig.width = 9, fig.height = 7, echo = F}
+Cancers.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+#Cancers.Female.mort.plot[[4]]
 ```
 
-**eFigure 5: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**
+**Figure S5: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**
 
-```{r, fig.width = 9, fig.height = 9, echo = F}
-`Non-communicable.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100))
-`Non-communicable.Male.mort.plot`[[4]]
+```{r, fig.width = 9, fig.height = 7, echo = F}
+`Non-communicable.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
+#`Non-communicable.Male.mort.plot`[[4]]
 ```
 
-**eFigure 6: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**
+**Figure S6: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**
 
-```{r, fig.width = 9, fig.height = 9, echo = F}
-`Non-communicable.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100))
-`Non-communicable.Male.mort.plot`[[4]]
+```{r, fig.width = 9, fig.height = 7, echo = F}
+`Non-communicable.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
+#`Non-communicable.Male.mort.plot`[[4]]
 ```
 
-**eFigure 7: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**
+**Figure S7: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**
 
-```{r, fig.width = 9, fig.height = 9, echo = F}
-Injuries.Male.mort.plot[[3]] %>% layout(margin = list(b = 100))
-Injuries.Male.mort.plot[[4]]
+```{r, fig.width = 9, fig.height = 7, echo = F}
+Injuries.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+#Injuries.Male.mort.plot[[4]]
 ```
 
-**eFigure 8: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**
+**Figure S8: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**
 
-```{r, fig.width = 9, fig.height = 9, echo = F}
-Injuries.Female.mort.plot[[3]] %>% layout(margin = list(b = 100)) 
-Injuries.Female.mort.plot[[4]]
+```{r, fig.width = 9, fig.height = 7, echo = F}
+Injuries.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70)) 
+#Injuries.Female.mort.plot[[4]]
 ```
 
-**eFigure 9: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**
+**Figure S9: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**
 
-```{r print-3, fig.width = 9, fig.height=9, echo = F}
-Communicable.Male.mort.plot[[3]] %>% layout(margin = list(b = 100))
-Communicable.Male.mort.plot[[4]]
+```{r print-3, fig.width = 9, fig.height = 7, echo = F}
+Communicable.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+#Communicable.Male.mort.plot[[4]]
 ```
 
-**eFigure 10: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**
+**Figure S10: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**
 
-```{r print-4, fig.width = 9, fig.height=9, echo = F}
-Communicable.Female.mort.plot[[3]] %>% layout(margin = list(b = 100))
-Communicable.Female.mort.plot[[4]]
+```{r print-4, fig.width = 9, fig.height = 7, echo = F}
+Communicable.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+#Communicable.Female.mort.plot[[4]]
 ```
 
-**eFigure 11: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**
+**Figure S11: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**
 
-```{r, fig.width = 9, fig.height = 9, echo = F}
-`All other causes.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100))
-`All other causes.Male.mort.plot`[[4]]
+```{r, fig.width = 9, fig.height = 7, echo = F}
+`All other causes.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
+#`All other causes.Male.mort.plot`[[4]]
 ```
 
-**eFigure 12: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**
+**Figure S12: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**
 
-```{r, fig.width = 9, fig.height = 9, echo = F}
-`All other causes.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100)) 
-`All other causes.Female.mort.plot`[[4]]
+```{r, fig.width = 9, fig.height = 7, echo = F}
+`All other causes.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70)) 
+#`All other causes.Female.mort.plot`[[4]]
 ```

### 2017-05-19 Manuscript/Manuscript_COD_Trends.Rmd
further manu edits

@@
 
-Department of Epidemiology, Biostatistics and Occupational Health, McGill University, Montreal, Canada.
+All authors: Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada.
 
 Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada.
-Email: <corinne.riddell@mail.mcgill.ca>
+Email: <corinne.riddell@mail.mcgill.ca>.
+Phone: +1-438-863-0402
 
@@
 This work emphasizes the importance of understanding how the direction of changes in specific causes of death by race results in changes to the life expectancy gap.
+
+###Significance statement: Add 120 word max statement.
 
@@
 
-**Men**
-
-```{r stacked_trends_male, echo=F, fig.height=9, fig.width=11}
-plot1 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male"), aes(x = year, y = COD_cont_yrs_mean)) + 
-  geom_area(aes(fill = COD_rev), stat = "identity") +
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
-  geom_line(data = subset(BlackWhite_results, sex == "Male"), aes(x = year, y = LE_wbgap_mean)) +
-  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
-  theme_classic(base_size = 10) + 
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Contribution to the life expectancy gap (%)") +
-  xlab("Year (1969-2013)"))
-
-for(i in 1:400){
-  if(plot1$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
-    plot1$x$data[[i]]$line$width <- 1
-  }
-}
-
-plot1
-
-plot2 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
-  geom_area(aes(fill = COD_rev), stat = "identity") +
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
-  geom_line(data = subset(BlackWhite_results, sex == "Male" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
-  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
-  theme_classic(base_size = 10) + 
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Contribution to the life expectancy gap (%)") +
-  xlab("Year (2009-2013)"))
-
-for(i in 1:400){
-  if(plot2$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
-    plot2$x$data[[i]]$line$width <- 1
-  }
-}
-
-plot2
-```
-
-**Women**
-
-```{r stacked_trends_female, echo=F, fig.height=9, fig.width=11}
-plot3 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female"), aes(x = year, y = COD_cont_yrs_mean)) + 
-                      geom_area(aes(fill = COD_rev), stat = "identity") +
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
-  geom_line(data = subset(BlackWhite_results, sex == "Female"), aes(x = year, y = LE_wbgap_mean)) +
-  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
-  theme_classic(base_size = 10) + 
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Contribution to the life expectancy gap (%)") +
-  xlab("Year (1969-2013)"))
-
-for(i in 1:400){
-  if(plot3$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
-    plot3$x$data[[i]]$line$width <- 1
-  }
-}
-
-plot3
-
-plot4 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
-                      geom_area(aes(fill = COD_rev), stat = "identity") +
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
-  geom_line(data = subset(BlackWhite_results, sex == "Female" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
-  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
-  theme_classic(base_size = 10) + 
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Contribution to the life expectancy gap (%)") +
-  xlab("Year (2009-2013)"))
-
-for(i in 1:400){
-  if(plot4$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
-    plot4$x$data[[i]]$line$width <- 1
-  }
-}
-
-plot4
-```
-
 ```{r create-the-resid-plots, echo = F, message = F}
@@
 
-****
+###Discussion
+
+**Summary of findings**
+
+We examined state-level trends in six major cause of death contributions to the black-white life expectancy gaps over a 45 year period.
+While we found that many states conformed to the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation.
+In contrast to some recent work focused on county-level inequalities ,^14,15^ states have some substantial advantages as analysis units.
+They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.
+
+Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century.^16^  Mortality trends for women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
+On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, greater racial equality can be achieved by falling black rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^17^
+
+Consistent with what has been reported at the national level,^18^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
+Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South.
+This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^19^
+
+Illinois, Michigan and Wisconsin also show relatively high contributions from cancer.
+On the other hand, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women.
+However, both states began with only a minimal contribution of CVD and cancer to the mortality gap.
+The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contribution of more than two years.
+Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.
+
+The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.
+States pursue varying approaches to an array of relevant programs and laws including tobacco control,^20^ Medicare coverage,^21^ hand-gun violence,^22^  and nutrition.^23^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially effect blacks.^24^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
+Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.^25^
+
+*Strengths and weaknesses*
+
+We used data over 45 years, across two racial groups, and for all ages.
+By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year.
+This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time.
+Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
+This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^26, 27^ The Arriaga decomposition method accounts for the fact that deaths occurring in one age or cause stratum are reflected in the absence of that death in another stratum.
+Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.
+
+Our analysis has two main limitations.
+Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state.
+This may be especially consequential for blacks in states like Florida or New York that received many black Hispanic immigrants from Cuba, Puerto Rico and the Dominican Republic, potentially changing the composition of the state’s black population to some extent during the study period.^28^  But it also has implications for the white population, especially in states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded.^29^ We were unable to adjust for Hispanic ethnicity in this study because the indicator for ethnicity on the death certificate has been noted to have varying accuracy across states and during the earlier time period.^8^ In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results,^4^ providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings overall.
+As the US population becomes increasingly Hispanic over time, however, this distinction grows more salient.^30^
+
+Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
+Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010.^31^ As we are examining the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white.
+Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death.
+Migration is particularly important for the District of Columbia, because of it's small geographical bounds, and substantial in-migration from New York, in- and out-migration to Maryland and Virginia.
+Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
+Such changes would both serve to increase the racial gap in DC.^32^
+
+Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
+For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.
+Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^33,34^
+
+**Conclusions**
+
+Racial differences have been a fixture of American life for centuries, but their continued evolution is a topic of great relevance for public policy and social justice.
+The careful surveillance of these patterns is therefore a crucial public health activity, and this paper presents a unique perspective on state-level variation over nearly half a century.
+This analysis provides a foundation for future work to investigate time-varying and time-fixed state-level characteristics that may be responsible for differences among state in risk factors leading to each cause of death.
+Differences in smoking policy, Medicare coverage, access to care, segregation, and racism, for example, could all affect risk factors for ill health and vary in magnitude by state.
 
@@
 
-###Discussion
-
-**Summary of findings**
-
-We examined state-level trends in six major cause of death contributions to the black-white life expectancy gaps over a 45 year period.
-While we found that many states conformed to the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation.
-In contrast to some recent work focused on county-level inequalities ,^14,15^ states have some substantial advantages as analysis units.
-They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.
-
-Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century.^16^  Mortality trends for women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
-On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, greater racial equality can be achieved by falling black rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^17^
-
-Consistent with what has been reported at the national level,^18^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
-Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South.
-This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^19^
-
-Illinois, Michigan and Wisconsin also show relatively high contributions from cancer.
-On the other hand, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women.
-However, both states began with only a minimal contribution of CVD and cancer to the mortality gap.
-The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contribution of more than two years.
-Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.
-
-The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.
-States pursue varying approaches to an array of relevant programs and laws including tobacco control,^20^ Medicare coverage,^21^ hand-gun violence,^22^  and nutrition.^23^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially effect blacks.^24^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
-Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.^25^
-
-*Strengths and weaknesses*
-
-We used data over 45 years, across two racial groups, and for all ages.
-By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year.
-This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time.
-Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
-This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^26, 27^ The Arriaga decomposition method accounts for the fact that deaths occurring in one age or cause stratum are reflected in the absence of that death in another stratum.
-Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.
-
-Our analysis has two main limitations.
-Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state.
-This may be especially consequential for blacks in states like Florida or New York that received many black Hispanic immigrants from Cuba, Puerto Rico and the Dominican Republic, potentially changing the composition of the state’s black population to some extent during the study period.^28^  But it also has implications for the white population, especially in states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded.^29^ We were unable to adjust for Hispanic ethnicity in this study because the indicator for ethnicity on the death certificate has been noted to have varying accuracy across states and during the earlier time period.^8^ In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results,^4^ providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings overall.
-As the US population becomes increasingly Hispanic over time, however, this distinction grows more salient.^30^
-
-Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
-Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010.^31^ As we are examining the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white.
-Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death.
-Migration is particularly important for the District of Columbia, because of it's small geographical bounds, and substantial in-migration from New York, in- and out-migration to Maryland and Virginia.
-Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
-Such changes would both serve to increase the racial gap in DC.^32^
-
-Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
-For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.
-Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^33,34^
-
-**Conclusions**
-
-Racial differences have been a fixture of American life for centuries, but their continued evolution is a topic of great relevance for public policy and social justice.
-The careful surveillance of these patterns is therefore a crucial public health activity, and this paper presents a unique perspective on state-level variation over nearly half a century.
-This analysis provides a foundation for future work to investigate time-varying and time-fixed state-level characteristics that may be responsible for differences among state in risk factors leading to each cause of death.
-Differences in smoking policy, Medicare coverage, access to care, segregation, and racism, for example, could all affect risk factors for ill health and vary in magnitude by state.
-
 ###Appendix
@@
 |
+
+**Stacked charts - not to be included in the submission**
+
+**Men**
+
+```{r stacked_trends_male, echo=F, fig.height=9, fig.width=11}
+plot1 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male"), aes(x = year, y = COD_cont_yrs_mean)) + 
+  geom_area(aes(fill = COD_rev), stat = "identity") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
+  geom_line(data = subset(BlackWhite_results, sex == "Male"), aes(x = year, y = LE_wbgap_mean)) +
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (%)") +
+  xlab("Year (1969-2013)"))
+
+for(i in 1:400){
+  if(plot1$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
+    plot1$x$data[[i]]$line$width <- 1
+  }
+}
+
+plot1
+
+plot2 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
+  geom_area(aes(fill = COD_rev), stat = "identity") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
+  geom_line(data = subset(BlackWhite_results, sex == "Male" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (%)") +
+  xlab("Year (2009-2013)"))
+
+for(i in 1:400){
+  if(plot2$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
+    plot2$x$data[[i]]$line$width <- 1
+  }
+}
+
+plot2
+```
+
+**Women**
+
+```{r stacked_trends_female, echo=F, fig.height=9, fig.width=11}
+plot3 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female"), aes(x = year, y = COD_cont_yrs_mean)) + 
+                      geom_area(aes(fill = COD_rev), stat = "identity") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
+  geom_line(data = subset(BlackWhite_results, sex == "Female"), aes(x = year, y = LE_wbgap_mean)) +
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (%)") +
+  xlab("Year (1969-2013)"))
+
+for(i in 1:400){
+  if(plot3$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
+    plot3$x$data[[i]]$line$width <- 1
+  }
+}
+
+plot3
+
+plot4 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
+                      geom_area(aes(fill = COD_rev), stat = "identity") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
+  geom_line(data = subset(BlackWhite_results, sex == "Female" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
+  theme_classic(base_size = 10) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (%)") +
+  xlab("Year (2009-2013)"))
+
+for(i in 1:400){
+  if(plot4$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
+    plot4$x$data[[i]]$line$width <- 1
+  }
+}
+
+plot4
+```
 

### 2017-05-21 conferences/ResearchDay/Riddell_ResearchDay_minimal.Rmd
prep for berlin

@@
 ---
-title: "Changes in cause of death and the black-white life expectancy gap"
-author: "Corinne Riddell, PhD"
-date: "April 24, 2017"
+title: "Long-term trends in the contribution of major causes of death to the black-white life expectancy gap by US state"
+author: "Corinne Riddell, Post-doctoral Researcher, McGill University, Canada"
+date: "May 23, 2017"
 output: ioslides_presentation
@@
 source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
+BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")
+
 ```
@@
 ```{r, echo=F}
-img1 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/k_morrison.png")), interpolate = FALSE)
-img2 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/s_harper2.png")), interpolate = FALSE)
-img3 <-  rasterGrob(as.raster(readJPEG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/j_kaufman.jpg")), interpolate = FALSE)
+img1 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/k_morrison.png")), interpolate = FALSE)
+img2 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/s_harper2.png")), interpolate = FALSE)
+img3 <-  rasterGrob(as.raster(readJPEG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/j_kaufman.jpg")), interpolate = FALSE)
 grid.arrange(img3,
@@
 ```
+Jay Kaufman, Supervisor        Kathryn Morrison, Post-doc       Sam Harper, Professor
 
@@
 ```{r, echo = F, out.width = 500}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/Harper_HA_title.png")
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/Harper_HA_title.png")
 ```
@@
 ```{r, echo=F, out.width = 750}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/Harper_HA_fig1.png")
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/Harper_HA_fig1.png")
 ```
@@
 ```{r, echo = F, out.width = 250}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/Harper_HA_quote.png")
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/Harper_HA_quote.png")
 ```
@@
 >- be interactive,
->- be reproducible,
->- and allow pictures to speak for themselves.
+>- be reproducible.
 
@@
 ```{r, echo = F, out.width = 400}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/ResearchDay/images/alabama.data.ex.png")
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/alabama.data.ex.png")
 ```
@@
 - <b class = "greentext">19</b> age groups: <1, 1-4, 5-9, ..., 85+
-- <b class = "greentext">2</b> sexes
-- <b class = "greentext">2</b> races (Black, White)
-- <b class = "greentext">6</b> major causes of deaths: CVD, cancer, non-communicable, communicable, injuries, all other causes
+- <b class = "greentext">2</b> genders
+- <b class = "greentext">2</b> races (black, white)
+- <b class = "greentext">6</b> major causes of deaths: cardiovascular disease, cancer, non-communicable, communicable, injuries, all other causes
 - = <b class = "greentext">820,800 rows of data</b>
@@
 
-**Estimate smoothed mortality rates and impute coarsened data.**
-
-+ Bayesian autoregressive model to borrow information *over time* within strata
-+ $log(\lambda_t) \sim N(\ log(\lambda_{t-1}),\ \tau)$
-+ impute coarsened 1--9 values using a truncated Poisson
-+ `rjags`
-
-## The analysis
-
-**Create 1,000 posterior samples**
-
-+ Each sample contains estimated *smoothed* and *imputed* death counts
++ Estimate smoothed mortality rates and impute coarsened data using a Bayesian time series model 
++ 1,000 posterior samples containing estimated *smoothed* and *imputed* death counts
 + For each sample:
@@
 + Calculate the mean, 5^th^ and 95^th^ percentiles for all estimates.
++ Compute national trend using random-effects meta regression
 
-## The analysis
+## Male black-white life expectancy gap
 
-**Synthesize state-specific findings**
+```{r, echo = F}
+    gg.p <- ggplot(data = subset(BlackWhite_results, sex == "Male"),
+                          aes(y = LE_wbgap_mean, x = year)) +
+    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
+    geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
+    #geom_line(aes(col = Census_Division)) + 
+    facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
+    theme_classic(base_size = 10) +
+    theme(axis.text.x = element_blank(),
+          strip.background=element_blank(),
+          axis.line=element_blank(),
+          axis.ticks=element_blank(),
+          legend.position = "none",
+          panel.background = element_rect(fill = "transparent", colour = NA), 
+          plot.background = element_rect(fill = "transparent", colour = NA),
+          legend.background = element_rect(fill = "transparent", colour = NA)) +
+    ylab("Life expectancy gap (years)") +
+    xlab(paste0("Year (1969-2013)"))
 
-+ Random-effects meta regression to estimate overall trend in contributions for each cause.
-+ Use `ggplot2`, `plotly`, and `Shiny` to make results accessible as a web app.
+      interactive.p <- ggplotly(gg.p +     
+        geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
+        geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
+        geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5))
 
-----
-# To the results!
+      for(i in 1:length(interactive.p$x$data)){
+        if (interactive.p$x$data[[i]]$line$color %in% c("rgba(0,0,0,1)", "rgba(190,190,190,0.5)")) {
+          interactive.p$x$data[[i]]$hoverinfo <- "none"
+          interactive.p$x$data[[i]]$line$width <- 1
+        }
+
+        interactive.p$x$data[[i]]$text <- gsub("Census_Division", "Census Division", interactive.p$x$data[[i]]$text)
+        interactive.p$x$data[[i]]$text <- gsub("LE_wbgap_mean", "Mean life expectancy gap", interactive.p$x$data[[i]]$text)
+        interactive.p$x$data[[i]]$text <- gsub("LE_wbgap_ucl", "Upper credible limit", interactive.p$x$data[[i]]$text)
+        interactive.p$x$data[[i]]$text <- gsub("LE_wbgap_lcl", "Lower credible limit", interactive.p$x$data[[i]]$text)
+      }
+
+       interactive.p
+```
+
+## Female black-white life expectancy gap
+
+```{r, echo = F}
+    gg.p2 <- ggplot(data = subset(BlackWhite_results, sex == "Female"),
+                          aes(y = LE_wbgap_mean, x = year)) +
+    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
+    geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
+    #geom_line(aes(col = Census_Division)) + 
+    facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
+    theme_classic(base_size = 10) +
+    theme(axis.text.x = element_blank(),
+          strip.background=element_blank(),
+          axis.line=element_blank(),
+          axis.ticks=element_blank(),
+          legend.position = "none",
+          panel.background = element_rect(fill = "transparent", colour = NA), 
+          plot.background = element_rect(fill = "transparent", colour = NA),
+          legend.background = element_rect(fill = "transparent", colour = NA)) +
+    ylab("Life expectancy gap (years)") +
+    xlab(paste0("Year (1969-2013)"))
+
+      interactive.p2 <- ggplotly(gg.p2 +     
+        geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
+        geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
+        geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5))
+
+      for(i in 1:length(interactive.p2$x$data)){
+        if (interactive.p2$x$data[[i]]$line$color %in% c("rgba(0,0,0,1)", "rgba(190,190,190,0.5)")) {
+          interactive.p2$x$data[[i]]$hoverinfo <- "none"
+          interactive.p2$x$data[[i]]$line$width <- 1
+        }
+
+        interactive.p2$x$data[[i]]$text <- gsub("Census_Division", "Census Division", interactive.p$x$data[[i]]$text)
+        interactive.p2$x$data[[i]]$text <- gsub("LE_wbgap_mean", "Mean life expectancy gap", interactive.p$x$data[[i]]$text)
+        interactive.p2$x$data[[i]]$text <- gsub("LE_wbgap_ucl", "Upper credible limit", interactive.p$x$data[[i]]$text)
+        interactive.p2$x$data[[i]]$text <- gsub("LE_wbgap_lcl", "Lower credible limit", interactive.p$x$data[[i]]$text)
+      }
+
+       interactive.p2
+```
+
+##National patterns in cause contributions to the gap
+
+```{r, echo = F}
+cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
+cod_decomp_results$COD <- factor(cod_decomp_results$COD, levels(cod_decomp_results$COD)[c(3, 2, 4, 6, 5, 1)])
+
+cod_decomp_national <- cod_decomp_results %>% 
+  filter(state == "Alabama") %>%
+  mutate(national.trend = fitted.RE - RE.estimates) %>%
+  select(-state)
+
+gg.p3 <- ggplotly(ggplot(cod_decomp_national, aes(y = national.trend, x = year)) + 
+  geom_line(aes(col = sex)) + 
+  facet_wrap(~COD) + theme_minimal() +
+  xlab("Year") + ylab("National trend in cause of death contribution")) %>%
+  layout(margin(b = 50, l = 50))
+
+gg.p3
+```
+
+#A tale of two states
+
+##Tennessee
+```{r, echo = F}
+age_cod_results_female <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_female.csv")
+
+age_cod_results_male <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_male.csv")
+
+age_cod_results <- rbind(age_cod_results_female, age_cod_results_male)
+rm(age_cod_results_female, age_cod_results_male)
+
+age_cod_results$age <- forcats::fct_relevel(f = age_cod_results$age, "<1 year", "1-4 years", "5-9 years", "10-14 years", 
+                                            "15-19 years", "20-24 years", "25-29 years", "30-34 years", "35-39 years", 
+                                            "40-44 years", "45-49 years", "50-54 years", "55-59 years", "60-64 years", 
+                                            "65-69 years", "70-74 years", "75-79 years", "80-84 years", "85+ years")
+
+age_cod_results$COD <- factor(age_cod_results$COD, levels(age_cod_results$COD)[c(3, 2, 4, 6, 5, 1)])
+
+sub1 <- subset(age_cod_results, year %in% c(1969, 2013) & state == "Tennessee")
+ly1 <- ggplotly(
+  ggplot(data = sub1,
+         aes(x=age, y = age_COD_cont_yrs_mean, fill = COD)) +
+    geom_bar(stat = "identity") + coord_flip() + theme_minimal()  + geom_vline(aes(xintercept = 0)) +
+    theme(legend.title = element_blank(),
+          panel.background = element_rect(fill = "transparent", colour = NA), 
+          plot.background = element_rect(fill = "transparent", colour = NA),
+          legend.background = element_rect(fill = "transparent", colour = NA)) +
+    ylab("Contribution to the life expectancy gap (years)") +
+    xlab("") +
+    facet_wrap(year ~ sex) 
+)
+
+for(i in 1:length(ly1$x$data)){
+  ly1$x$data[[i]]$text <- gsub("age_COD_cont_yrs_mean", "Contribution to gap (yrs)", ly1$x$data[[i]]$text)
+}
+
+ly1 %>% layout(margin = list(b = 100))
+```
+
+##New York
+```{r, echo = F}
+
+sub1 <- subset(age_cod_results, year %in% c(1969, 2013) & state == "New York")
+ly1 <- ggplotly(
+  ggplot(data = sub1,
+         aes(x=age, y = age_COD_cont_yrs_mean, fill = COD)) +
+    geom_bar(stat = "identity") + coord_flip() + theme_minimal()  + geom_vline(aes(xintercept = 0)) +
+    theme(legend.title = element_blank(),
+          panel.background = element_rect(fill = "transparent", colour = NA), 
+          plot.background = element_rect(fill = "transparent", colour = NA),
+          legend.background = element_rect(fill = "transparent", colour = NA)) +
+    ylab("Contribution to the life expectancy gap (years)") +
+    xlab("") +
+    facet_wrap(year ~ sex) 
+)
+
+for(i in 1:length(ly1$x$data)){
+  ly1$x$data[[i]]$text <- gsub("age_COD_cont_yrs_mean", "Contribution to gap (yrs)", ly1$x$data[[i]]$text)
+}
+
+ly1 %>% layout(margin = list(b = 100))
+```
+
+##Similarities & differences
+
+- Substantial contribution of infant mortality to the gap in 1969
+- Most deaths from all other causes are during infancy
+- Similar pattern of injury contributions by gender in 1969, but larger contribution in Tennessee
+- Similar pattern of CVD contributions in 1969 and 2013, but Tennessee started worse off
+- Ditto for cancer
+- Injury plays important role in Tennessee in 2013 with opposing effect directions in men and women
+
+##Blah
+
+-

### 2017-05-21 conferences/ResearchDay/Riddell_ResearchDay_minimal.Rmd

@@
 ---
-title: "Long-term trends in the contribution of major causes of death to the black-white life expectancy gap by US state"
+title: "Trends in the contribution of major causes of death to the black-white life expectancy gap"
 author: "Corinne Riddell, Post-doctoral Researcher, McGill University, Canada"
@@
 ```
-Jay Kaufman, Supervisor        Kathryn Morrison, Post-doc       Sam Harper, Professor
+Jay Kaufman, Kathryn Morrison, and Sam Harper
 
@@
 ```
-
-And also...
-
->- be broadly accessible,
->- be interactive,
->- be reproducible.
 
@@
 
-+ Estimate smoothed mortality rates and impute coarsened data using a Bayesian time series model 
-+ 1,000 posterior samples containing estimated *smoothed* and *imputed* death counts
-+ For each sample:
++ Estimated smoothed mortality rates using a Bayesian time series model with Poisson likelihood
++ Used a truncated Poisson likelihood when number of deaths were between 1 and 9
++ 1,000 posterior samples. For each:
     1. Create a life table for every strata
@@
     4. Calculate age-standardized mortality rates
-+ Calculate the mean, 5^th^ and 95^th^ percentiles for all estimates.
++ Calculate the mean, 5^th^ and 95^th^ percentiles for all estimates
 + Compute national trend using random-effects meta regression
@@
 ##Tennessee
-```{r, echo = F}
+```{r, echo = F, out.height = 600, out.width = 800}
 age_cod_results_female <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_female.csv")
@@
 ##New York
-```{r, echo = F}
+```{r, echo = F, out.height = 600, out.width = 800}
 
@@
 
-##Blah
+##Learn more
+- Working paper submitted to bioRxiv - will be public soon!
+- Shiny app hosted online with our results here:  https://corinne-riddell.shinyapps.io/shiny_app/
 
--
+**Get in touch**
+
+- corinne.riddell@mail.mcgill.ca
+- @datavisitor on Twitter

### 2017-05-29 conferences/ResearchDay/Riddell_ResearchDay_minimal.Rmd
updated pres for HMD

@@
 ```
+
 Jay Kaufman, Kathryn Morrison, and Sam Harper
@@
 
-## The data
-
-```{r, echo = F, out.width = 400}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/alabama.data.ex.png")
-```
-
-- When the number of deaths is between 1 and 9 it is <b class = "redtext">coarsened</b>.
-
 ## The analysis
@@
 + Compute national trend using random-effects meta regression
+
+## The output
+
++ Working paper being screened at bioRxiv
++ R Shiny App: https://corinne-riddell.shinyapps.io/shiny_app/
+
+## The app
+
+```{r, echo = F, out.width = 800}
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/app_ex1.gif")
+```
+
+## The app
+
+```{r, echo = F, out.width = 800}
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/app_ex2.gif")
+```
+
+## The app
+
+```{r, echo = F, out.width = 800}
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/app_ex3.gif")
+```
 
@@
 
+#Three examples
+
+##CVD in men
+```{r, echo = F, out.width = 675}
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/CVD_men.png")
+```
+
+##Cancer in women
+```{r, echo = F, out.width = 675}
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/Cancer_women.png")
+```
+
+##Injury in women
+```{r, echo = F, out.width = 675}
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/Injury_women.png")
+```
+
 #A tale of two states
@@
 
-##Learn more
-- Working paper submitted to bioRxiv - will be public soon!
-- Shiny app hosted online with our results here:  https://corinne-riddell.shinyapps.io/shiny_app/
-
-**Get in touch**
-
+##Get in touch
 - corinne.riddell@mail.mcgill.ca

### 2017-05-29 Manuscript/Manuscript_COD_Trends_PNAS.Rmd
PNAS manu submission

---
title: Long-term trends in the contribution of major causes of death to the black-white life expectancy gap by US state
bibliography: bibliography.bib
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

Corinne A Riddell, PhD

Kathryn T Morrison, PhD

Sam Harper, PhD

Jay S Kaufman, PhD

All authors: Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 27, Montreal, QC H3A 1A2, Canada.

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 27, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>. Phone: +1-438-863-0402

```{r initialize, echo=FALSE, warning=F, message=F}
library(plotly)
library(viridis)
library(scales)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
library(splines)
library(broom)
library(tidyverse)

source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

```{r load_data, echo=F}
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")
cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
cod_decomp_results <- reorder.as.map2(cod_decomp_results, "state", "stabbrs")
cod_decomp_results$COD_rev <- factor(cod_decomp_results$COD, levels = c("Cardiovascular", "Cancers", "All other causes",
                                                                        "Communicable", "Non-communicable",  "Injuries"))

age_cod_results_infants_AOC <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_infants_AOC.csv")
age_cod_results_infants_AOC$stabbrs <- state.abb[match(age_cod_results_infants_AOC$state, state.name)]
age_cod_results_infants_AOC$stabbrs[age_cod_results_infants_AOC$state == "Washington DC"] <- "DC"
age_cod_results_infants_AOC <- reorder.as.map2(age_cod_results_infants_AOC, "state", "stabbrs")

```

###Abstract

States have fared differently in their progress towards eliminating the black-white life expectancy gap. Our objective is to describe the pattern of contributions of six major causes of deaths to the sex-specific black-white life expectancy gap across states over the last half-century, and identify divergent states.

Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major causes of death.

```{r summarize_results_for_abstract, echo = F}
cvd.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) 
cvd.female.2013 <- format(round(cvd.female.2013, 1), nsmall = 1)

ncd.male.average <- cod_decomp_results %>% filter(sex == "Male" & COD == "Non-communicable" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
ncd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
ncd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)

can.male.max <- cod_decomp_results %>% filter(sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(max = round(max(fixed.from.RE), 1))
can.female.average <- cod_decomp_results %>% filter(sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))

inj.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
```

Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r formatC(as.numeric(cvd.female.2013), digits = 1)` years). The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in 1988) and a stable contribution for women (approximately `r can.female.average` years). Both genders exhibited a decreased contribution from injuries (men: `r inj.male.1969` to `r inj.male.2013` years), that became negative for women (women: `r inj.female.1969` to `r inj.female.2013` years). Several states diverged from these general trends.

Life expectancy for both races has improved substantially in the US. For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality. In women, a crossover in injury-related mortality led to a narrower gap, realized partially by increasing mortality among whites.

----

###Significance

Racial differences have been a fixture of American life for centuries, but their continued evolution is a topic of great relevance for public policy and social justice. States have fared differently in eliminating the black-white life expectancy gap. Using vital statistics and census data, we describe how six major causes of death contributed to the life expectancy gap between blacks and whites from 1969 to 2013. We find that states diverged in their cause-of-death specific contributions to the life expectancy gap in ways that either resulted in the near or total elimination of the gap, or alternatively its exacerbation. This analysis provides a foundation for future work to investigate state characteristics that may result in inequalities in mortality by race.

----

Life expectancy is a barometer for overall population health. In 2014, life expectancy at birth in the United States was 78.8 years, but hidden within this summary are important subgroups differences by self-defined racial ethnicity. Whites lived 3.4 years longer than blacks.^1^ While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.^2^

During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV/AIDS, unintentional injuries, and, among women, due to differential improvements in heart disease.^3^ The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years.^4^ However, the extent to which different causes of death may have contributed to state-level changes in the black-white gap are unknown.

The over-arching objective of this study is to describe the contribution of selected causes of death to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences. To do so, we used publicly available data from 1969 to 2013 detailing the number of deaths and population size according to US state, year, gender, race (black or white), age, and cause of death. We excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010.^5^ We did not consider Hispanic ethnicity as part of our racial definition, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.^6^

Using these data, we estimated state-level trends in the black-white life expectancy gap measured in years separately for men and women. We then calculated the contribution of each of six major causes of death: cardiovascular disease, cancer, non-communicable disease, injury, communicable disease, and all other causes. This study lays the foundation for the identification of policies or social conditions in highly successful states, and the suggestion of broad targets for interventions to further reduce the gap in less successful states.

###Results
```{r life-expectancy-trends, echo=F}
LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  labs(fill = "Census Division") +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")

LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  labs(fill = "Census Division") +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")
```

```{r plot_both, eval=F, echo=F}
#Informative: male and females at the same time
ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)"))
```

```{r plotting_resid_fx, echo = F}
plot.resid.trends <- function(dataset, COD1, sex1) {
  subset.data <- subset(dataset, COD == COD1 & sex == sex1)
  subset.data$order <- reorder(subset.data$state, subset.data$sum.residual, first)

  a <- ggplot(subset.data,
         aes(x = year, y = COD_cont_yrs_mean)) + 
    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
    #geom_line(aes(col = "Observed")) + 
    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9") +
    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
    #scale_fill_manual(values = c("#b2e2e2")) +
    facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
    geom_line(aes(y = fitted.FE), col = "#d7191c") +
    ggtitle(paste0("Estimated smoothed contribution vs. meta-estimated common trend in ", COD1, " in ", sex1)) +
    theme_classic(base_size = 14) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
    ylab("estimated smoothed contribution vs. common trend (years)") +
    xlab("Year (1969-2013)")

  b <- ggplot(subset.data,
              aes(x = year, y = resid.FE)) + geom_line(aes(col = "Residual")) + 
    scale_color_manual(values = c("#d7191c")) +    
    facet_wrap( ~ order) +
    xlab(" ") + ylab("Residual") +
    geom_hline(yintercept = 0, col = "black") +
      geom_text(data = subset(subset.data, year == 2010), 
                aes(y = 1, label = round(sum.residual, 1)), size = 2.5) + 
    ggtitle(paste0("Residuals vs. year in ", COD1, " in ", sex1)) +
    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

  c <- ggplot(subset.data,
              aes(x = year, y = COD_cont_yrs_mean)) + 
    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
    geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
    geom_line(col = "#08519c") + 
    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
    #scale_fill_manual(values = c("#b2e2e2")) +
    facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
    #geom_line(aes(y = fitted.RE), col = "#d7191c", lty = 3) +
#    ggtitle(paste0("Estimated smoothed contribution of ", COD1, " disease to the black-white life expectancy gap (blue 95% credible interval) vs. the meta-estimated general pattern (red line) in ", sex1, " 1969-2013")) +
    theme_classic(base_size = 14) + 
    theme(axis.text.x = element_blank(),
          strip.background=element_blank(),
          axis.line=element_blank(),
          axis.ticks=element_blank()) +
    ylab("Contribution to the life expectancy gap (years)") +
    xlab("Year (1969-2013)")

  return(list(plot1 = a, plot2 = b, plot3 = c))
}
```

```{r create-the-resid-plots, echo = F, message = F}
for(COD_i in levels(cod_decomp_results$COD)) {
  for(sex_i in levels(cod_decomp_results$sex)) {
    assign(paste0(COD_i, ".", sex_i, ".plot"), plot.resid.trends(dataset = cod_decomp_results, 
                                                                         COD1 = COD_i, sex1 = sex_i))
  }
}
```

```{r refine_plots, eval = F, echo = F}
#need tp add the "grid lines post-hoc" because they depend on the max height of the plot

#turn hover off for grid lines and axis lines
refine.patterns <- function(plot) {
  for(i in 1:320){
  if(plot$x$data[[i]]$line$color == "rgba(190,190,190,1)"){
    plot$x$data[[i]]$line$width <- 0.7559055
    plot$x$data[[i]]$hoverinfo <- "none"
  }
    if(plot$x$data[[i]]$line$color == "rgba(0, 0, 0, 1)"){
    plot$x$data[[i]]$hoverinfo <- "none"
  }
  }
  return(plot)
}

#refine.patterns(Cardiovascular.Male.plot[[3]])
```

**Figures 1 and 2** display trends in the life expectancy gap for men and women, respectively. For men and women, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among men.

**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for men and women, in years, for each state. The blue line depicts the mean of the smoothed contribution estimated for each state, such that states can have very different trend lines. These lines are overlaid with the meta-estimated national trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**Figures S1-S12**).

**Cardiovascular disease**

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for all groups (**Figures S1 and S2**), but white men underwent relatively larger mortality decreases than black men. As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states (**Figure 3**). Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**). For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.

**Cancer**

For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Figures S3 and S4**). Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 5**). For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**). For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.

**Non-communicable disease**

Mortality from non-communicable disease increased over time for men in most states, with DC as a notable exception (**Figure S5**). Recently, there was a slight decrease in mortality rates, especially among black men. These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Figure S6**). As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**). As with other causes, New York and Massachusetts exhibited decreasing contributions, and Nevada also made substantial progress. Furthermore, while DC's mortality rates decreased, they presently have the highest contribution of non-communicable disease to the gap.

**Injury**

Trends in mortalilty from injuries differ for blacks and whites. Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S7**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S8**). Thus, injury mortality rates among blacks are equivalent, or lower than those among whites in many states, with white women demonstrating sharper recent increases. The contribution of injuries to the mortality gap has therefore generally decreased over time, and has become negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 9 and 10**). Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

**Communicable disease**

Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Figures S9 and S10**). This spike was greater among men, especially blacks, and among states with larger urban centers. The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.

**All other causes**

```{r, echo = F}
male.endpoints.aoc <- cod_decomp_results %>% filter(year %in% c(1969, 2013), sex =="Male", COD == "All other causes", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )

female.endpoints.aoc <- cod_decomp_results %>% filter(year %in% c(1969, 2013), sex =="Female", COD == "All other causes", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )

male.infants.endpoints.aoc <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Male", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )

female.infants.endpoints.aoc <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Female", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )
```

Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**Figures S11 and S12**). White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 13 and 14**). The most disparate states are Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.

Because the majority of infant deaths are classified as being due to all other causes in our analysis, we also tried to separate any impact of all other causes from those of infant mortality more generally. We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (Figures 15 and 16). We found that 43% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969. In 2013, this proportion reached 53%. Thus, the trend in the all other causes contribution partially reflects a decreasing contribution from infant mortality.

###Discussion

**Summary of findings**

We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45 year period. While we found that many states conformed to the national pattern of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation. In contrast to some recent work focused on county-level inequalities,^7,8^ states have some substantial advantages as analysis units; they have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.

Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West, which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century.^9^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi. On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^8^ On the other hand, greater racial equality can be achieved by falling black mortality rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^10^

Consistent with what has been reported at the national level,^11^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South. This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^12^ Communicable disease continues to contribute to the gap in most states, especially in DC, New Jersey, and Florida. In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 nationally,^11^ suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV. Cancer is also responsible for large contributions to the life expectancy gap in some states (Illinnois, Michigan, and Wisconsin) but not in others (New York, Massachusetts). Overall, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women.However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contributions of more than two years. Thus, the reason for New York’s and Massachusetts’ low CVD contributions does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.

The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.  States pursue varying approaches to an array of relevant programs and laws including tobacco control,^13^ Medicare coverage,^14^ hand-gun violence,^15^  and nutrition.^16^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^17^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others. Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.^18^

*Strengths and weaknesses*

We used data over 45 years, across two racial groups, and for all ages. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^19,20^ The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum. Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.

Our analysis has three main limitations. Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state. This may be especially consequential for blacks in states like Florida or New York that received many black Hispanic immigrants from Cuba, Puerto Rico and the Dominican Republic, potentially changing the composition of the state’s black population to some extent during the study period.^21^  But it also has implications for the white population, especially in states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas. Given that Hispanics have higher life expectanies than non-Hispanics, including Hispanics in the life expectancy calculations will lead to estimates that are likely higher than they would have been had Hispanics been excluded.^22^ We did not to adjust for Hispanic ethnicity because reporting began only in 1978 and even then, agreement between self-reported ethnicity on the Census and secondary-reported ethinicity on the death certificate has been shown to be poorer during the earlier time period and differential according to geography, which could induce measurement error into mortality and disparity estimates.^6^ In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results,^4^ providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings. As the US population becomes increasingly Hispanic over time, however, this distinction grows more salient.^23^

Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states. Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010.^24^ As we are examining the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white. Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death. Migration is particularly important for the District of Columbia, because of it's small geographical bounds, and substantial in-migration from New York and in- and out-migration to Maryland and Virginia. Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Both changes would serve to increase the racial gap in DC.^25^

Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes. For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time. Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^26,27^

**Conclusions**

Racial differences have been a fixture of American life for centuries, but their continued evolution is a topic of great relevance for public policy and social justice. The careful surveillance of these patterns is therefore a crucial public health activity, and this paper presents a unique perspective on state-level variation over nearly half a century. This analysis provides a foundation for future work to investigate time-varying and time-fixed state-level characteristics that may be responsible for differences among states in risk factors leading to each cause of death. Differences in smoking policy, Medicare coverage, access to care, segregation, and racism, for example, could all affect risk factors for ill health and vary in magnitude by state.

###Methods

**Description of the mortality data**

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.^28^ The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.^29^ The population data were derived by the NCHS using intercensal bridged-race population estimates. We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**Table S1**).

**Statistical challenges**

The data structure produced two statistical challenges that were addressed by our modeling strategy. Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^30^ that we needed to impute. Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata. In these cases, the estimation of the mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying mortality rates, we used a method that produces smoother age-specific mortality trends.

**Estimation of smoothed mortality rates using a Bayesian time-series model**

We used a Bayesian time-series model to smooth the death counts across years and estimate the underlying mortality rate. Specifically, we used a Bayesian generalized linear model with a Poisson likelihood to model the number of deaths in each state-year-sex-race-age-cause stratum as a function of the at-risk population, for strata with zero or more than 9 deaths. The mortality rates over time within each stratum were assigned an autoregressive prior, where the log mortality rate was assumed to be Gaussian centered at the previous year's mortality rate. A non-informative gamma prior was used to estimate the inverse variance. For censored strata with between one and nine deaths, the deaths were modeled using a truncated Poisson likelihood, and otherwise the model remained the same. The truncated Poisson was bound between one and the larger of nine or the stratum's population size, which could be less than nine.

Models were fit using Markov chain Monte Carlo (MCMC). These models were run separately for every combination of state, race, gender, and cause of death. They were fit using two chains of length 10,000, with a 2,000 iteration burn-in period to ensure convergence. One thousand iterations remained after burn-in and sampling every eighth iteration to reduce autocorrelation among the MCMC samples. Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition. These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.

**Decomposition of the life expectancy gap by cause of death**

To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied.^31,32^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method.^33^ This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher cause-specific mortality rates among blacks. For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher rate of mortality due to cardiovascular disease among blacks compared to whites. If the decomposition produces a negative estimate this implies that whites have a higher rate of mortality for injuries compared to blacks. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 40 state-level trends and their credible intervals.

We also calculated state-level time trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000.^34^

**Description of the trend in each cause of death's contribution to the gap and identification of divergent states**

To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimated from the previous step as a function of a state random effect and a linear spline for year (estimated using four internal knots). The model was inversely-weighted by the standard error of the mean estimate in each state-year. To identify states divergent from the national trend, we compared the state-specific trends estimated in the previous step to the national pattern estimated by setting the random effect to zero.

**Data Sharing**

A replication data set including the raw data and statistical code to reproduce the manuscript is publicly available on github: https://github.com/corinne-riddell/BlackWhiteMortalityGap/.

###Acknowledgements

We extend our thanks to Dr. Richard Maclehose, who developed the initial version of the code to perform the Bayesian analysis as part of the research article “Trends in the Black-White Life Expectancy Gap Among Us States, 1990–2009”, by Harper S, MacLehose RF, and Kaufman JS.

Corinne Riddell is a post-doctoral researcher at McGill University, whose salary was supported by Dr. Kaufman's Canada Research Chair in Health Disparities. Kathryn Morrison is a post-doctoral researcher at McGill University. Sam Harper was supported by a Chercheur Boursier Junior 2 from the Fonds de la Recherche en Santé du Québec. Jay Kaufman was supported by a Canada Research Chair in Health Disparities.

**Figure 1: State-level trends in the black-white life expectancy gap in men, United States, 1969-2013**

```{r plot_LE_trends_males, echo=F, fig.height=10, fig.width=13}
LE_trend_plot_males +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 2: State-level trends in the black-white life expectancy gap in women, United States, 1969-2013**

```{r plot_LE_trends_females, echo=F, fig.height=10, fig.width=13}
LE_trend_plot_females +
  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 3: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

```{r print-plot, fig.width = 12, fig.height = 10, echo = F}
Cardiovascular.Male.plot[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 4: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

```{r print-plot-2, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
Cardiovascular.Female.plot[[3]]  +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 5: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

```{r print-plot-3, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
Cancers.Male.plot[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 6: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

```{r print-plot-4, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
Cancers.Female.plot[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 7: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

```{r print-plot-5, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
`Non-communicable.Male.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 8: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

```{r print-plot-6, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
`Non-communicable.Female.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 9: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

```{r print-plot-7, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Injuries.Male.plot[[3]])

`Injuries.Male.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) + 
           geom_hline(aes(yintercept = 6), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 10: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

```{r print-plot-8, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Injuries.Female.plot[[3]])

`Injuries.Female.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)
```

**Figure 11: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

```{r print-plot-9, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Communicable.Male.plot[[3]])

`Communicable.Male.plot`[[3]] +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) 
```

**Figure 12: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

```{r print-plot-10, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Communicable.Female.plot[[3]])

`Communicable.Female.plot`[[3]] +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```

**Figure 13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

```{r print-plot-11, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(`All other causes.Male.plot`[[3]])

`All other causes.Male.plot`[[3]] +
  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) 
```

**Figure 14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

```{r print-plot-12, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(`All other causes.Female.plot`[[3]])

`All other causes.Female.plot`[[3]] +
  geom_hline(aes(yintercept = -2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)
```

**Figure 15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men younger than one, 1969-2013**

```{r age-cod-trend-infant-males, fig.width = 12, fig.height = 10, echo = F}
ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Male"), aes(y = age_COD_cont_yrs_mean, x = year)) +
  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
  geom_line(col = "#08519c") +  
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (years)") +
  xlab("Year (1969-2013)") +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```

**Figure 16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women younger than one, 1969-2013**

```{r age-cod-trend-infant-females, fig.width = 12, fig.height = 10, echo = F}
ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Female"), aes(y = age_COD_cont_yrs_mean, x = year)) +
  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
  geom_line(col = "#08519c") +  
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (years)") +
  xlab("Year (1969-2013)") +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```

###Appendix
**Table S1: Cause of death groupings**

| Broad cause of death     | Included causes of death                                                     |
|--------------------------|------------------------------------------------------------------------------|
| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer. |
| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries. | 
| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Syndrome and Nephrosis.  |
| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death. |

**Stacked charts - not to be included in the submission**

**Men**

```{r stacked_trends_male, echo=F, fig.height=9, fig.width=11}
plot1 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male"), aes(x = year, y = COD_cont_yrs_mean)) + 
  geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Male"), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (1969-2013)"))

for(i in 1:400){
  if(plot1$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot1$x$data[[i]]$line$width <- 1
  }
}

plot1

plot2 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
  geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Male" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (2009-2013)"))

for(i in 1:400){
  if(plot2$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot2$x$data[[i]]$line$width <- 1
  }
}

plot2
```

**Women**

```{r stacked_trends_female, echo=F, fig.height=9, fig.width=11}
plot3 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female"), aes(x = year, y = COD_cont_yrs_mean)) + 
                      geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Female"), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (1969-2013)"))

for(i in 1:400){
  if(plot3$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot3$x$data[[i]]$line$width <- 1
  }
}

plot3

plot4 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
                      geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Female" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (2009-2013)"))

for(i in 1:400){
  if(plot4$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot4$x$data[[i]]$line$width <- 1
  }
}

plot4
```

###Bibliography

1. Cunningham TJ, Croft JB, Liu Y, Lu H, Eke PI, Giles WH (2017) Vital Signs: Racial Disparities in Age-Specific Mortality Among Blacks or African Americans - United States, 1999-2015. *MMWR Morb Mortal Wkly Rep* 66(17):444-456.

2. Fuchs, VR (2016) Black Gains in Life Expectancy. *JAMA*. 316(18): 1869–70.

3. Harper S, Lynch J, Burris S, Smith GD (2007) Trends in the black-white life expectancy gap in the United States, 1983-2003. *JAMA* 297(11):1224-32.

4. Harper S, MacLehose RF, Kaufman JS (2014) Trends in the Black-White Life Expectancy Gap Among Us States, 1990–2009. *Health Aff* 33(8): 1375–82.

5. U.S. Census Bureau: United States Census 2010 Interactive Population Map [Internet]. [cited 2017 May 16]. Available from: https://www.census.gov/2010census/popmap/.

6. Arias E, Schauman WS, Eschbach K, Sorlie PD, Backlund E (2008) The validity of race and Hispanic origin reporting on death certificates in the United States. National Center for Health Statistics. Vital Health Stat 2(148). Available from: https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf

7. Arora A, Spatz E, Herrin J, Riley C, Roy B, Kell K, Coberley C, Rula E, Krumholz HM (2016) Population Well-Being Measures Help Explain Geographic Disparities In Life Expectancy At The County Level. *Health Aff* 35(11):2075-82.

8. Dwyer-Lindgren L, Bertozzi-Villa A, Stubbs RW, Morozoff C, Kutz MJ, Huynh C, Barber RM, et al (2016) US County-Level Trends in Mortality Rates for Major Causes of Death, 1980-2014. *JAMA* 316(22):2385–2401.

9. Lemann, N (1991) The Promised Land: The Great Black Migration and How It Changed America. New York: Alfred A. Knopf, ISBN 0-394-56004-3.

10. Case A, Deaton A (2017) Mortality and Morbidity in the 21st Century. In: Eberly J and Stock J, Editors. Brookings Papers on Economic Activity, March 23-24 2017.

11. Harper S, Rushani D, and Kaufman JS (2012) Trends in the Black-White Life Expectancy Gap, 2003-2008. *JAMA* 307(21):2257–9.

12. Rudd RA, Seth P, David F, Scholl L (2016) Increases in Drug and Opioid-Involved Overdose Deaths — United States, 2010–2015. *MMWR Morb Mortal Wkly Rep* 65:1445–1452.

13. Cotti C, Nesson E, Tefft N (2016) The Effects of Tobacco Control Policies on Tobacco Products, Tar, and Nicotine Purchases among Adults: Evidence from Household Panel Data. *American Economic Journal: Economic Policy* 8(4):103-23.

14. Grogan CM, Park SE (2017) The Racial Divide in State Medicaid Expansions. *J Health Polit Policy Law* 42(3):539-72.

15. Humphreys DK, Gasparrini A, Wiebe DJ (2017) Evaluating the Impact of Florida’s “Stand Your Ground” Self-defense Law on Homicide and Suicide by Firearm: An Interrupted Time Series Study. *JAMA Intern Med* 177(1):44-50.

16. Taber DR, Stevens J, Evenson KR, Ward DS, Poole C, Maciejewski ML, Murray DM, Brownson RC (2011) State policies targeting junk food in schools: racial/ethnic differences in the effect of policy change on soda consumption. *Am J Public Health* 101(9):1769-75.

17. Adimora AA, Ramirez C, Schoenbach VJ, Cohen MS (2014) Policies and politics that promote HIV infection in the Southern United States. *AIDS*
28(10):1393.

18. Strumpf EC, Harper S, Kaufman JS (2017) Fixed Effects and Difference in Differences. In: Methods in Social Epidemiology. 2nd Ed, eds Oakes M, Kaufman JS(Jossey-Bass, San Francisco).

19. Case A, Deaton A (2015) Rising Morbidity and Mortality in Midlife among White Non-Hispanic Americans in the 21st Century. *Proc Natl Acad Sci * 112(49):15078-83.

20. Minton J, Green M, McCartney G, Shaw R, Vanderbloemen L, Pickett K (2016) Two cheers for a small giant? Why we need better ways of seeing data: A commentary on:‘Rising morbidity and mortality in midlife among White non-Hispanic Americans in the 21st century’. *Int J Epidemiol* 2016 Sep 17:dyw095.

21. LaVeist-Ramos TA, Galarraga J, Thorpe RJ, Bell CN, Austin CJ (2011) Are black Hispanics black or Hispanic? Exploring disparities at the intersection of race and ethnicity. *J Epidemiol Community Health* Mar 3:jech-2009.

22. Ruiz JM, Steffen P, Smith TB (2013) Hispanic mortality paradox: a systematic review and meta-analysis of the longitudinal literature. *Am J Public Health* 103(3):e52-60.

23. Brazil N (2017) Spatial Variation in the Hispanic Paradox: Mortality Rates in New and Established Hispanic US Destinations. *Population, Space and Place* 23:e1968.

24. Preston SH, Elo IT (2014) Anatomy of a municipal triumph: New York City's upsurge in life expectancy. *Population and development review* 40(1):1-29.

25. U.S. Census Bureau. Census Reporter Profile Page: American Community Survey 1-year estimates [Internet]. [Cited 16 May 2017]. Available from: https://censusreporter.org/profiles/16000US1150000-washington-dc/.

26. Kuehn BM (2014) CDC: major disparities in opioid prescribing among states: some states crack down on excess prescribing. *JAMA* 312(7):684-6.

27. Jones CM, Mack KA, Paulozzi LJ (2010) Pharmaceutical overdose deaths, United States, 2010. *JAMA* 309(7):657-9.

28. National Cancer Institute. Surveillance, Epidemiology, and End Results
Program: SEER*Stat software: release: version 8.3.2 [Internet]. Bethesda (MD): NCI; 2017 Apr 14 [cited 2016 August 20]. Available from: http://www.seer.cancer.gov/seerstat.

29. Centers for Disease Control and Prevention. National Center for Health Statistics: National Vital Statistics System [Internet]. [cited 2017 May 16]. Available from: https://www.cdc.gov/nchs/nvss/index.htm.

30. National Cancer Institute. Surveillance, Epidemiology, and End Results Program: US mortality data [Internet]. Bethesda (MD): NCI; [cited 2017 May 16]. Available from: http://www.seer.cancer.gov/mortality/

31. Preston SH, Heuveline P, and Guillot M (2001) Demography: Measuring and Modeling Population Processes. (Blackwell Publishing, Malden)

32. Selvin, S (2004) Statistical Analysis of Epidemiologic Data. 3rd ed. (Oxford University Press, Oxford)

33. Arriaga EE (1985) Measuring and explaining the change in life expectancies. *Demography* 21(1):83-96.

34. National Cancer Institute. Surveillance, Epidemiology, and End Results Program: Standard Populations - 19 Age Groups [Internet]. [cited 2017 April 25]. Available from: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html.

**Figure Legends**

**Figure 1:** The bands illustrate the 95% credible interval around the mean trend line. States with more imprecise estimates (often due to small black populations) have wider credible intervals. The estimated life expectancy gap in California in 1972 is much higher than in surrounding years. This anomaly is also illustrated in Figure 3, and is a product of an unexpectedly low CVD-related mortality rate among white males in California in 1972, likely denoting a coding error. Thus the data for this state and year should be interpreted with caution.

**Figure 3:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of CVD to the black-white life expectancy gap over time. The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states. The early spike in California's trend line illustrates the unexpectedly high estimate of the CVD contribution to the black-white life expectancy gap among California males in 1972. This anomolous estimate reflects the unexpectedly low CVD-related mortality rate among white males in California in 1972, and likely denotes a coding error. Thus the data for this state and year should be interpreted with caution.

### 2017-06-02 Manuscript/Manuscript_COD_Trends_PNAS.Rmd
moved and deleted bits

@@
 
+**Figure 1 Legend:** The bands illustrate the 95% credible interval around the mean trend line.
+States with more imprecise estimates (often due to small black populations) have wider credible intervals.
+The estimated life expectancy gap in California in 1972 is much higher than in surrounding years.
+This anomaly is also illustrated in Figure 3, and is a product of an unexpectedly low CVD-related mortality rate among white males in California in 1972, likely denoting a coding error.
+Thus the data for this state and year should be interpreted with caution.
+
 **Figure 2: State-level trends in the black-white life expectancy gap in women, United States, 1969-2013**
@@
 ```
+
+**Figure 3:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of CVD to the black-white life expectancy gap over time.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
+The early spike in California's trend line illustrates the unexpectedly high estimate of the CVD contribution to the black-white life expectancy gap among California males in 1972.
+This anomalous estimate reflects the unexpectedly low CVD-related mortality rate among white males in California in 1972, and likely denotes a coding error.
+Thus the data for this state and year should be interpreted with caution.
 
@@
 Available from: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html.
-
-**Figure Legends**
-
-**Figure 1:** The bands illustrate the 95% credible interval around the mean trend line.
-States with more imprecise estimates (often due to small black populations) have wider credible intervals.
-The estimated life expectancy gap in California in 1972 is much higher than in surrounding years.
-This anomaly is also illustrated in Figure 3, and is a product of an unexpectedly low CVD-related mortality rate among white males in California in 1972, likely denoting a coding error.
-Thus the data for this state and year should be interpreted with caution.
-
-**Figure 3:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of CVD to the black-white life expectancy gap over time.
-The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
-The early spike in California's trend line illustrates the unexpectedly high estimate of the CVD contribution to the black-white life expectancy gap among California males in 1972.
-This anomolous estimate reflects the unexpectedly low CVD-related mortality rate among white males in California in 1972, and likely denotes a coding error.
-Thus the data for this state and year should be interpreted with caution.

### 2017-06-02 shiny_app/app.R

@@ -178,7 +178,7 @@ ui1 <- fluidPage(theme = "cosmo-customized.css",
                                           htmlOutput("Explain_LE_females"),
                                           htmlOutput("Explain_LE_males"),
                                           plotlyOutput("age_cod", height = 700, width = 1100),
-                                          uiOutput("Warning_age1"),
+                                        
                                           htmlOutput("Alabama")
                                  ),
                                  
@@ -206,7 +206,7 @@ ui1 <- fluidPage(theme = "cosmo-customized.css",
                                  tabPanel(title = "Cross-sectional age contribution", value = "Age.snapshot",                                
                                           htmlOutput("description_age_summary"),
                                           plotlyOutput("state_age_summary", height = 800),
-                                          uiOutput("Warning_age2"),
+                                         
                                           dataTableOutput("data.temp2")),
                                  
                                  tabPanel(title = "More information", value = "more",
@@ -732,13 +732,7 @@ server <- function(input, output) {
     ly %>% layout(xaxis = list(title = " ", side = "top", xpad = 20), yaxis = list(title = NA, autorange = "reversed"))
   })
   
-  output$Warning_age2 <- renderUI({
-    HTML(paste0("<b>Interpretation warning:</b> For several states and years, the oldest age group(s) is/are estimated to contribute to narrowing the black-white gap.",
-                " However, there is known age misclassification for older generations, especially among blacks, that is thought to lead to this apparent 'cross-over'. See",
-                " <a href='https://www.ncbi.nlm.nih.gov/books/NBK109843/#ch2.r35'>here</a> for more information."
-                
-    ))
-  })
+
   
 ##########################################
 ##            Explore a state           ##
@@ -969,13 +963,7 @@ output$age_cod <- renderPlotly({
   
 })
 
-output$Warning_age1 <- renderUI({
-  HTML(paste0("<b>Interpretation warning:</b> For several states and years, the oldest age group(s) is/are estimated to contribute to narrowing the black-white gap.",
-              " However, there is known age misclassification for older generations, especially among blacks, that is thought to lead to this apparent 'cross-over'. See",
-              " <a href='https://www.ncbi.nlm.nih.gov/books/NBK109843/#ch2.r35'>here</a> for more information.<br/><br/>"
-              
-  ))
-})
+
 
 output$Alabama <- renderUI({
   if(input$state != "Alabama"){

### 2017-06-02 dataverse_files/CR edit state-le-analysis.R
removing old files

@@ -1,357 +1,0 @@
- # Decomposition analysis
- #
- ##############################################################################################
- 
- #Functions
- #################################################################
- # a couple of functions for timing the runs
- tic <- function(gcFirst = TRUE, type=c("elapsed", "user.self", "sys.self"))
- {
-    type <- match.arg(type)
-    assign(".type", type, envir=baseenv())
-    if(gcFirst) gc(FALSE)
-    tic <- proc.time()[type]         
-    assign(".tic", tic, envir=baseenv())
-    invisible(tic)
- }
- 
- toc <- function()
- {
-    type <- get(".type", envir=baseenv())
-    toc <- proc.time()[type]
-    tic <- get(".tic", envir=baseenv())
-    print(toc - tic)
-    invisible(toc)
- }
-
-#functions to convert risks to rates and vice versa
-#risktorate<-function(x,t) {
-#    -log(1-x)/t
-#}
-ratetorisk<-function(x,t) {
-    1-exp(-x*t)
-}
-##################################################################################################
-
-library(BRugs)
-setwd('c:/bugsR/4')
-
-#Matrix to collect ouput of Bayes decomp
-Out <- matrix(0, 15200, 46)
-colnames(Out) <- c("state","gender","age","year",
-                "total","totall","totalu",
-                "direct","directl","directu",
-                "indirect","indirectl","indirectu",
-                "atotal","atotall","atotalu",
-                "adirect","adirectl","adirectu",
-                "aindirect","aindirectl","aindirectu",                
-                "lew","lewl","lewu",
-                "leb","lebl","lebu",
-                "lediff","lediffl","lediffu",
-                "rw","rwl","rwu",
-                "rb","rbl","rbu",
-                "changeb","changebl","changebu",
-                "changew","changewl","changewu",
-                "change","changel","changeu"                
-                )
-count <- 1
-#Flag any instances of risks > 1 among blacks and whites
-flagb <- rep(0, 3)
-flagw <- rep(0, 3)
-
[…]

### 2017-06-02 dataverse_files/le_ref_ktm.R

@@ -1,339 +1,0 @@
-
-
-
-# Rich's analysis w/ openbugs and getting the LE SE 
-  
-library(R2OpenBUGS)
-setwd('~/Documents/BlackWhiteMortalityGap/dataverse_files')
-
-ratetorisk=function(x,t) {
-  1-exp(-x*t)
-}
-
-
-#Matrix to collect ouput of Bayes decomp
-Out = matrix(0, 15200, 46)
-colnames(Out) = c("state","gender","age","year",
-                   "total","totall","totalu",
-                   "direct","directl","directu",
-                   "indirect","indirectl","indirectu",
-                   "atotal","atotall","atotalu",
-                   "adirect","adirectl","adirectu",
-                   "aindirect","aindirectl","aindirectu",                
-                   "lew","lewl","lewu",
-                   "leb","lebl","lebu",
-                   "lediff","lediffl","lediffu",
-                   "rw","rwl","rwu",
-                   "rb","rbl","rbu",
-                   "changeb","changebl","changebu",
-                   "changew","changewl","changewu",
-                   "change","changel","changeu"                
-)
-count = 1
-#Flag any instances of risks > 1 among blacks and whites
-flagb = rep(0, 3)
-flagw = rep(0, 3)
-
-# load data
-dtotal = read.table("deaths-total.txt", na.strings=".")
-colnames(dtotal) = c("state","age","gender","race","year","crude","deaths","pop")
-
-for (istate in 0:11){
-#for (istate in 13:25){
-#for (istate in c(27:33)) {
-#for (istate in c(35:40,42:44,46:50)) { #CR question: why running the for loop for istate over different ranges separately?
-  
-  for (igender in 0:1) {
-    
-    
-    #do this next section once for whites then for blacks
-    
-    #Temoporary dataset for whites in istate and igender
[…]

### 2017-06-02 dataverse_files/state-le-analysis-nhisp.R

@@ -1,356 +1,0 @@
- # Decomposition analysis
- # 2/21/2013 - fixed to use non hispanic data
- # ignoring arriaga decomp to speed things up a little. just putting zeros in those cells of output
- ##############################################################################################
- 
- #Functions
- #################################################################
- # a couple of functions for timing the runs
- tic <- function(gcFirst = TRUE, type=c("elapsed", "user.self", "sys.self"))
- {
-    type <- match.arg(type)
-    assign(".type", type, envir=baseenv())
-    if(gcFirst) gc(FALSE)
-    tic <- proc.time()[type]         
-    assign(".tic", tic, envir=baseenv())
-    invisible(tic)
- }
- 
- toc <- function()
- {
-    type <- get(".type", envir=baseenv())
-    toc <- proc.time()[type]
-    tic <- get(".tic", envir=baseenv())
-    print(toc - tic)
-    invisible(toc)
- }
-
-#functions to convert risks to rates and vice versa
-#risktorate<-function(x,t) {
-#    -log(1-x)/t
-#}
-ratetorisk<-function(x,t) {
-    1-exp(-x*t)
-}
-##################################################################################################
-
-library(BRugs)
-setwd('c:/bugsR/4')
-
-#Matrix to collect ouput of Bayes decomp
-Out<-matrix(0,15200,46)
-colnames(Out)<-c("state","gender","age","year",
-                "total","totall","totalu",
-                "direct","directl","directu",
-                "indirect","indirectl","indirectu",
-                "atotal","atotall","atotalu",
-                "adirect","adirectl","adirectu",
-                "aindirect","aindirectl","aindirectu",                
-                "lew","lewl","lewu",
-                "leb","lebl","lebu",
[…]

### 2017-06-02 dataverse_files/state-le-analysis.R

@@ -1,355 +1,0 @@
- # Decomposition analysis
- #
- ##############################################################################################
- 
- #Functions
- #################################################################
- # a couple of functions for timing the runs
- tic <- function(gcFirst = TRUE, type=c("elapsed", "user.self", "sys.self"))
- {
-    type <- match.arg(type)
-    assign(".type", type, envir=baseenv())
-    if(gcFirst) gc(FALSE)
-    tic <- proc.time()[type]         
-    assign(".tic", tic, envir=baseenv())
-    invisible(tic)
- }
- 
- toc <- function()
- {
-    type <- get(".type", envir=baseenv())
-    toc <- proc.time()[type]
-    tic <- get(".tic", envir=baseenv())
-    print(toc - tic)
-    invisible(toc)
- }
-
-#functions to convert risks to rates and vice versa
-#risktorate<-function(x,t) {
-#    -log(1-x)/t
-#}
-ratetorisk<-function(x,t) {
-    1-exp(-x*t)
-}
-##################################################################################################
-
-library(BRugs)
-setwd('c:/bugsR/4')
-
-#Matrix to collect ouput of Bayes decomp
-Out <- matrix(0, 15200, 46)
-colnames(Out) <- c("state","gender","age","year",
-                "total","totall","totalu",
-                "direct","directl","directu",
-                "indirect","indirectl","indirectu",
-                "atotal","atotall","atotalu",
-                "adirect","adirectl","adirectu",
-                "aindirect","aindirectl","aindirectu",                
-                "lew","lewl","lewu",
-                "leb","lebl","lebu",
-                "lediff","lediffl","lediffu",
-                "rw","rwl","rwu",
-                "rb","rbl","rbu",
-                "changeb","changebl","changebu",
-                "changew","changewl","changewu",
-                "change","changel","changeu"                
-                )
-count <- 1
-#Flag any instances of risks > 1 among blacks and whites
[…]

### 2017-06-02 dataverse_files/state-le-plots.R

@@ -1,108 +1,0 @@
-##  program:  state-le-plots.r
-##  task:     life expectancy plots
-## 	input:	  state-le-results.dta					
-##	output:		various figures
-##  project:  state differences in black-white life expectancy
-##  author:   sam harper \ 1aug2014
-
-## users will need to specify working directory with life expectancy output
-# setwd(".......")  
-
-# setup
-setwd("/Users/samharper/Dropbox/SRJ/Data/replication files")
-library(ggplot2)
-library(foreign)
-library(gridExtra)
-library(reshape2)
-library(directlabels)
-library(doBy)
-library(proto)
-    
-# Grab results from decomposition
-decomp <- read.dta("state-le-results.dta")
-
-# subset to life expectancy at birth
-tg <- subset(decomp,age=="00-00 years")
-
-stabblist <- unique(tg$stabb)
-stnamelist <- unique(tg$stname)
-stnum <- unique(tg$stseer)
-rnamelist <- unique(tg$regname)
-rnum <- unique(tg$reg)
-dnamelist <- unique(tg$divname)
-dnum <- unique(tg$div)
-
-# labels for state, region, division
-tg$gender <- as.numeric(tg$gender)
-tg$abb <- factor(tg$stseer,levels=c(stnum),labels=c(stabblist))
-tg$name <- factor(tg$stseer,levels=c(stnum),labels=c(stnamelist))
-tg$regn <- factor(tg$reg,levels=c(rnum),labels=c(rnamelist))
-tg$regn <- factor(tg$reg,levels=c(1L,2L,3L,4L),labels=c("Northeast","Midwest","South","West"))
-tg$divn <- factor(tg$div,levels=c(dnum),labels=c(dnamelist))
-tg$divn <- factor(tg$div,levels=c(1L,2L,3L,4L,5L,6L,7L,8L,9L),labels=c("New England","Middle Atlantic","East North Central","West North Central","South Atlantic","East South Central","West South Central","Mountain","Pacific"))
-
-### Need to just estimate a weighted average of each state-specific trend rather than using geom_smooth (too variable depending on scale)
-
-# aggregate by division, year and gender
-tg$lediffse <- abs(tg$lediffu - tg$lediffl)/(2*1.96) #standard err for gap in each state
[…]

### 2017-06-02 conferences/Riddell_presentation_max_planck.Rmd
moved conference files

--- 2017-05-21 conferences/ResearchDay/Riddell_ResearchDay_minimal.Rmd
@@
 ```{r, echo=F}
-img1 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/k_morrison.png")), interpolate = FALSE)
-img2 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/s_harper2.png")), interpolate = FALSE)
-img3 <-  rasterGrob(as.raster(readJPEG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/j_kaufman.jpg")), interpolate = FALSE)
+img1 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/k_morrison.png")), interpolate = FALSE)
+img2 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/s_harper2.png")), interpolate = FALSE)
+img3 <-  rasterGrob(as.raster(readJPEG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/j_kaufman.jpg")), interpolate = FALSE)
 grid.arrange(img3,
@@
 ```{r, echo = F, out.width = 500}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/Harper_HA_title.png")
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/Harper_HA_title.png")
 ```
@@
 ```{r, echo=F, out.width = 750}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/Harper_HA_fig1.png")
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/Harper_HA_fig1.png")
 ```
@@
 ```{r, echo = F, out.width = 250}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/Harper_HA_quote.png")
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/Harper_HA_quote.png")
 ```
@@
 ```{r, echo = F, out.width = 400}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/alabama.data.ex.png")
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/alabama.data.ex.png")
 ```
@@
 - = <b class = "greentext">820,800 rows of data</b>
-
-## The data
-
-```{r, echo = F, out.width = 400}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/alabama.data.ex.png")
-```
-
-- When the number of deaths is between 1 and 9 it is <b class = "redtext">coarsened</b>.
 
@@
 + Compute national trend using random-effects meta regression
+
+## The output
+
++ Working paper being screened at bioRxiv
++ R Shiny App: https://corinne-riddell.shinyapps.io/shiny_app/
+
+## The app
+
+```{r, echo = F, out.width = 800}
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/app_ex1.gif")
+```
+
+## The app
+
+```{r, echo = F, out.width = 800}
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/app_ex2.gif")
+```
+
+## The app
+
+```{r, echo = F, out.width = 800}
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/app_ex3.gif")
+```
 
@@
 
+#Three examples
+
+##CVD in men
+```{r, echo = F, out.width = 675}
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/CVD_men.png")
+```
+
+##Cancer in women
+```{r, echo = F, out.width = 675}
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/Cancer_women.png")
+```
+
+##Injury in women
+```{r, echo = F, out.width = 675}
+include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/images/Injury_women.png")
+```
+
 #A tale of two states

### 2017-06-02 conferences/Max_planck_submission.Rmd

--- 2017-05-03 conferences/Max Planck/submission.Rmd
@@
 ---
+
+**Note:** This file is based on **OLD** results that no longer exist.
+I have not updated this file to work with the new results since I have yet to re-run this code and check that all lines are correct.
+Thus, this file is kept for historical purposes and in case I would like to refer back to it to recreate some of the analysis/vis using the same results.
+
+This file contains the code/analysis for my submission to the Max Planck Institute of Demography's symposium on the Human Mortality Database.
+The symposium was held in Berlin on May 22 and 23.
+My final presentation slides for the symposium can be found in the same folder as this submission and are based on updated results.
 

### 2017-06-02 Code/archived/bayesian_smoothing/bayes_smoothing_CIs.R
moving files

@@ -1,0 +1,86 @@
+
+
+# RUN-SCRIPT
+# OBTAIN THE CREDIBLE INTERVALS FOR THE BAYESIAN SMOOTHING MODELS 
+# NOV 29, 2016 
+
+source('~/BlackWhiteMortalityGap/Code/bayesian_smoothing/bayes_smoothing_src.R')
+
+# CR wrote this, will need to double check how it works 
+
+setwd("~/bwm_results")
+
+time.a <- system.time(alabama_1990_males_thinned <- investigate_convergence(year = 1990, state = "Alabama", sex = "male", n_post_samp = 20, "alabama_1990_males_thinned", 10, 950))
+
+#note for KTM: see line 246 of 4_smoothing_time_fun.R -- I think that line needs to be generalized to also include males. The code below produces an error
+#if I change sex = "male".
+time.a <- system.time(alabama_1990_females_thinned <- investigate_convergence(year = 1990, state = "Alabama", sex = "female", n_post_samp = 1000, "alabama_1990_females_thinned", 10, 950))
+time.b <- system.time(alabama_2009_females_thinned <- investigate_convergence(year = 2009, state = "Alabama", sex = "female", n_post_samp = 1000, "alabama_2009_females_thinned", 10, 950))
+
+time.c <- system.time(missouri_1990_females_thinned <- investigate_convergence(year = 1990, state = "Missouri", sex = "female", n_post_samp = 1000, "missouri_1990_females_thinned", 10, 950))
+time.d <- system.time(missouri_2009_females_thinned <- investigate_convergence(year = 2009, state = "Missouri", sex = "female", n_post_samp = 1000, "missouri_2009_females_thinned", 10, 950))
+
+time.e <- system.time(utah_1990_females_thinned <- investigate_convergence(year = 1990, state = "Utah", sex = "female", n_post_samp = 1000, "utah_1990_females_thinned", 10, 950))
+time.f <- system.time(utah_2009_females_thinned <- investigate_convergence(year = 2009, state = "Utah", sex = "female", n_post_samp = 1000, "utah_2009_females_thinned", 10, 950))
+
+wv_1990_females_thinned <- investigate_convergence(year = 1990, state = "West Virginia", sex = "female", n_post_samp = 1000, "wv_1990_females_thinned", 10, 950)
[…]

### 2017-06-02 Code/archived/bayesian_smoothing/bayes_smoothing_functions.R

@@ -1,0 +1,327 @@
+
+# LIST OF FUNCTIONS
+# subset_data_jags
+# create_dataset
+# bwmort_smooth_time
+# combine_race_data
+# merge_data
+# run_smoothing_models
+# get_allcod
+# get_life_tables
+# run_smoothing_models_mulitple_states
+# extract_mcmc_dist
+
+subset_data_jags = function(ds, race, sex, cod, state) { 
+  
+  ds2 = ds[order(ds$censored), ] 
+  ds2 = ds2[ds2$state==state, ]
+  
+  ds_jags = list(deaths = ds2$deaths, 
+                 lnpop = log(ds2$population), 
+                 age.bin = ds2$agebins, 
+                 year = ds2$year, 
+                 upper_bound = ds2$upper_bound, 
+                 binned = is.na(ds2$deaths), 
+                 not.binned = !is.na(ds2$deaths), 
+                 n.binned = sum(is.na(ds2$deaths)), 
+                 n.not.binned = sum(!is.na(ds2$deaths)),
+                 n.not.binned.plus.1 = sum(!is.na(ds2$deaths)) + 1,
+                 n.rows = sum(is.na(ds2$deaths)) + sum(!is.na(ds2$deaths)),
+                 n.age.bins = length(unique(ds2$agebins)),
+                 n.years = length(unique(ds2$year)), 
+                 binned.id = ds2$censored) 
+}
+
+create_dataset = function(df,  colname_state, colname_pop, colname_deaths, colname_cod, colname_year, colname_sex, colname_agebins, colname_race) {  
+
+  selected_vars = c(colname_state, colname_pop, colname_deaths, colname_cod, colname_year, colname_sex, colname_agebins, colname_race)
+  ds = df[ , selected_vars] 
+  colnames(ds) = c('state', 'population', 'deaths', 'cod', 'year', 'sex', 'agebins', 'race')
+  ds$agebins = ds$agebins - (min(ds$agebins)-1)
+  ds$year = ds$year - (min(ds$year)-1)
+  ds$upper_bound = ifelse(ds$population<9, ds$population, 9) 
+  ds$censored = ifelse(is.na(ds$deaths),1,0)
+  return(ds) 
+} 
+
+bwmort_smooth_time = function(ds_jags_bw) { 
+  
[…]

### 2017-06-02 Code/archived/bayesian_smoothing/bayes_smoothing_run.R

@@ -1,0 +1,54 @@
+
+# RUN-SCRIPT
+# BAYESIAN SMOOTHING MODELS FOR A GIVEN SET OF STATES 
+# NOV 29, 2016 
+
+
+setwd("~/BlackWhiteMortalityGap/Code/bayesian_smoothing")
+source('bayes_smoothing_src.R') 
+
+# Run models for all age-sex-year-race categories for a given state 
+# variable 'states' is a vector of all 51 states; you can select any one or a subset or all
+
+#Use <=4 cores during work hours or if the server is in use (htop to check); use up to 8 overnight when no one is using it 
+
+# We will often want to run all states except the ones that have tiny populations, e.g.: 
+excluded_states = unique(dat.clean$State2[dat.clean$Population < 5]) 
+included_states = setdiff(states, excluded_states) 
+
+cod_list = c('Injuries', 'Cardiovascular', 'Cancers', 'Communicable', 
+             'Non-communicable', 'All other causes')
+
+load('~/black_white_mortality_project/main_datasets.Rdata') # change for SERVER 
+#load('/Users/kathryn/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata') # change for LOCAL 
+
+colname_sex = 'Sex2'
+colname_race = 'Race2'
+colname_pop = 'Population'
+colname_state = 'State2'
+colname_cod = 'COD2'
+colname_year = 'Year'
+colname_deaths = 'Count'
+colname_agebins = 'Age' 
+
+
+mclapply(X = "Alabama", FUN=run_smoothing_models_mulitple_states, mc.cores=1, df=dat.clean, 
+                     cod_list, colname_state, colname_pop, colname_deaths, colname_cod, 
+                     colname_year, colname_sex, colname_agebins, colname_race) 
+
+#mclapply(X=included_states, FUN=run_smoothing_models_mulitple_states, mc.cores=1, df=dat.clean, 
+#         cod_list, colname_state, colname_pop, colname_deaths, colname_cod, 
+#         colname_year, colname_sex, colname_agebins, colname_race)  
+
+
+load('~/black_white_mortality_project/validation_data_0610.Rdata') # change for SERVER 
+#load('/Users/kathryn/Dropbox/BlackWhiteGap/Data/validation_data_0610.Rdata') # change for LOCAL 
+states = unique(validation_data_0610$State2)
+
+setwd("~/black_white_mortality_project/validation_results")
+
[…]

### 2017-06-02 Code/archived/bayesian_smoothing/bayes_smoothing_src.R

@@ -1,0 +1,21 @@
+
+library(rjags)
+library(R2jags) 
+library(devtools)
+#install_github(repo='johnbaums/jagstools') 
+library(jagstools)
+library(ggplot2)
+library(reshape2)
+library(parallel)
+library(gtools)
+library(bayesboot)
+library(coda)
+library(ggplot2)
+library(grid)
+library(gridExtra)
+
+
+
+source('~/BlackWhiteMortalityGap/Code/bayesian_smoothing/bayes_smoothing_functions.R') 
+#source('life_expectancy_functions.R') 
+

### 2017-06-02 Code/validation/SH_comparison.R

@@ -1,0 +1,107 @@
+
+load('~/SH_HA_data.Rdata')
+source("~/BlackWhiteMortalityGap/Code/2_smoothing_models/bayes_smoothing_src.R")
+
+nrows <- 3*2*length(levels(dtotal$state))
+
+le.table <-  data.frame(state = factor(nrows), 
+                        le.white = numeric(nrows), le.white.lcl = numeric(nrows), le.white.ucl = numeric(nrows),
+                        le.black = numeric(nrows), le.black.lcl = numeric(nrows), le.black.ucl = numeric(nrows),
+                        le.gap = numeric(nrows), le.gap.lcl = numeric(nrows), le.gap.ucl = numeric(nrows)
+                        )
+
+le.table$state <- levels(dtotal$state)
+le.table$year <- c(rep("1990", nrows/3), rep("2009", nrows/3), rep("difference", nrows/3))
+le.table$sex <- c(rep("Male", nrows/6), rep("Female", nrows/6))
+
+#1990 and 2009
+system.time( #this took 7.5 hours on one core on the server.
+  for(state_i in setdiff(levels(dtotal$state), c("Idaho", "Montana", "North Dakota", "South Dakota", "Vermont", "Wyoming"))) {
+    for(sex_i in c("Male", "Female")) { 
+      for(race_i in c("White", "Black")) {
+        ds_sub = subset_data(ds = dtotal, state = state_i, sex = sex_i, race = race_i)
+        data = jagsify_data(ds_sub)
+        jags_model = run_smoothing_model(data)
+        
+        assign(paste0("r1990.", sex_i), clean_smoothing_results(data, jags_model, year = 1990, n.posterior.samples = 1000))
+        assign(paste0("r2009.", sex_i), clean_smoothing_results(data, jags_model, year = 2009, n.posterior.samples = 1000))
+        
+        #make a life table for each sample
+        
+        assign(paste0("life.tables.1990.", sex_i), map(get(paste0("r1990.", sex_i)), ~ life.table( .x, num.ages.in.group = "age_bin_size",
+                                                                                                   death.counts = "smoothed_deaths",
+                                                                                                   population.counts = "population",
+                                                                                                   ave.prop.le1.lived = 0.10)))
+        
+        assign(paste0("life.tables.2009.", sex_i), map(get(paste0("r2009.", sex_i)), ~ life.table( .x, num.ages.in.group = "age_bin_size",
+                                                                                                   death.counts = "smoothed_deaths",
+                                                                                                   population.counts = "population",
[…]

### 2017-06-02 Code/archived/old_2_Run_LE_and_Decomp_Calcs.Rmd

--- 2017-01-20 Code/2_Run_LE_and_Decomp_Calcs.Rmd
@@
 
-rm(les, le1, le5, le9)
+rm(les, le1, le5, le9, sid)
 ```
@@
 
-```{r save_objects}
-write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated2.csv", col_names = T)
-write_csv(BlackWhite, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/BlackWhite.csv", col_names = T)
-
-save(paired.ids, non.zero.stratums,
-     list.age.decomp.tables1, list.age.decomp.tables5, list.age.decomp.tables9,
-     list.life.tables1, list.life.tables5, list.life.tables9,  
-     file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/calculated_data.Rdata")
-```
-
 ```{r calculate_cod_decomp}
@@
                                                               cod.table = ct.both %>% select(age, COD, prop.white9, prop.black9), 
-                                                              age.colname.cod.table = "Age_factor", 
-                                                              COD.colname.cod.table = "COD_factor", 
+                                                              age.colname.cod.table = "age", 
+                                                              COD.colname.cod.table = "COD", 
                                                               prop1.colname.cod.table = "prop.black9", 
@@
 for(i in 1:dim(paired.ids)[1]) {
-    if(list.cod.decomp.tables[[i]][1] != "Couldn't compare life tables" & 
-       list.cod.decomp.tables[[i]][1] != "Encountered age group with no deaths") {
+    if(list.cod.decomp.tables1[[i]]$decomp.table[1] != "Couldn't compare life tables" & 
+       list.cod.decomp.tables1[[i]]$decomp.table[1] != "Encountered age group with no deaths") {
 
-       list.cod.decomp.tables[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables1[[i]],
+       list.cod.decomp.tables1[[i]]$decomp.table <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables1[[i]]$decomp.table,
                                                             age.groups = "Ages", 
@@
 
-       list.cod.decomp.tables5[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables5[[i]],
+       list.cod.decomp.tables5[[i]]$decomp.table <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables5[[i]]$decomp.table,
                                                             age.groups = "Ages", 
@@
 
-       list.cod.decomp.tables9[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables9[[i]],
+       list.cod.decomp.tables9[[i]]$decomp.table <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables9[[i]]$decomp.table,
                                                             age.groups = "Ages", 
@@
 ```{r pretty_cod}
-ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
+ggplot(list.cod.decomp.tables1[[1]]$decomp.table, aes(y = Ages, x = start)) + 
   geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
@@
 
-ggplot(list.cod.decomp.tables9[[1]], aes(y = Ages, x = start)) + 
+ggplot(list.cod.decomp.tables9[[1]]$decomp.table, aes(y = Ages, x = start)) + 
   geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
@@
 ```
+
+```{r save_objects}
+write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated2.csv", col_names = T)
+write_csv(BlackWhite, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/BlackWhite.csv", col_names = T)
+
+save(paired.ids, non.zero.stratums,
+     list.life.tables1, list.life.tables5, list.life.tables9,  
+     list.age.decomp.tables1, list.age.decomp.tables5, list.age.decomp.tables9,
+     list.cod.decomp.tables1, list.cod.decomp.tables5, list.cod.decomp.tables9,
+     file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/calculated_data.Rdata")
+```

### 2017-06-02 Code/archived/old_append_smoothed_results.Rmd

--- 2017-01-05 Code/append_smoothed_results.Rmd
@@
 BlackWhite$Census_Region[BlackWhite$State2 %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee", 
-                                                  "Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "West or East South Central"
+                                                  "Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "South Central"
+
+BlackWhite$Census_Region <- factor(BlackWhite$Census_Region, levels = c("West", "Midwest", "South Central", "South Atlantic", "Northeast"))
 
@@
 
+cod.contributions$State2 <- factor(cod.contributions$State2, levels = c("Alabama", "Arizona", "Arkansas",
+                                                                        "California", "Colorado", "Connecticut",
+                                                                        "Delaware", "Washington DC", "Florida", "Georgia", "Illinois",
+                                                                        "Indiana", "Iowa", "Kansas", "Kentucky", 
+                                                                        "Louisiana", "Maryland", "Massachusetts", 
+                                                                        "Michigan", "Minnesota", "Mississippi",
+                                                                        "Missouri", "Nebraska", "Nevada", "New Jersey", 
+                                                                        "New Mexico", "New York", "North Carolina", 
+                                                                        "Ohio", "Oklahoma", "Pennsylvania", "Rhode Island", 
+                                                                        "South Carolina", "Tennessee", "Texas", "Utah", 
+                                                                        "Virginia", "Washington", "West Virginia", "Wisconsin" ))
+
 cod.contributions$Census_Region <- "South"
@@
 cod.contributions$Census_Region[cod.contributions$State2 %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee", 
-                                                                "Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "West or East South Central"
+                                                                "Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "South Central"
 
+cod.contributions$Census_Region <- factor(cod.contributions$Census_Region, levels = c("West", "Midwest", "South Central", "South Atlantic", "Northeast"))
+
+cod.contributions$rand_year = 2000 + sample(0:10, 1) #used for jittering labels on the plot 
 ```

### 2017-06-02 Code/bayesian_smoothing/bayes_smoothing_postrun_analysis.R
deleting removed files

@@ -1,4 +1,0 @@
-
-# Cleaning up the results imported from server 
-
-

### 2017-06-02 Code/archived/Run_model_on_SH_data/Import_SH_HA_data.Rmd
reorg

---
title: "extract table from SH HA paper"
author: "Corinne Riddell"
date: "January 22, 2017"
output: html_document
---

```{r}

library(pdftools)

txt <- pdf_text("/Users/corinneriddell/Dropbox/BlackWhiteGap/papers/Harper-Health Affairs-2014.pdf")

#males
cat(txt[13])
cat(txt[14])

#females
cat(txt[15])
cat(txt[16])

#i then copied and pasted the above into excel and saved as SH_HA_supp_app_data.csv
#to bring back here...

SH.dat <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/SH_HA_supp_app_data.csv", header = T)
```

### 2017-06-02 Code/archived/Run_model_on_SH_data/Prep SH HA data for our analysis.Rmd
reorg

---
title: "Compare with SH HA paper"
author: "Corinne Riddell"
date: "December 13, 2016"
output: html_document
---

```{r}
library(foreign)
library(dplyr)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")

dtotal <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/dataverse_files/deaths-total.txt", header = F, sep="\t", stringsAsFactors = F, colClasses = c(rep("numeric", 8)),
                     na.strings = ".")

names(dtotal)
#no names affixed to dataset. But the names are stored in the data dictionary file found at:
#/Users/corinneriddell/Dropbox/BlackWhiteGap/dataverse_files/deaths-total.dic

str(dtotal)
table(dtotal$V1) #50 states + DC
table(dtotal$V2) #20 age groups
table(dtotal$V3) #sex
table(dtotal$V4) #race
table(dtotal$V5) #20 years

names(dtotal) <- c("state.n", "age.n", "sex.n", "race.n", "year.0", "crude.rate", "deaths", "population") #see the data dictionary - path is above.
```

```{r recode_variables}
dtotal <- dtotal %>%
  mutate(state = factor(state.n, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
                                                             "Colorado", "Connecticut", "Delaware", "Washington DC",
                                                             "Florida", "Georgia", "Hawaii","Idaho", "Illinois", 
                                                             "Indiana","Iowa", "Kansas", "Kentucky", "Louisiana", 
                                                             "Maine", "Maryland", "Massachusetts", "Michigan", 
                                                             "Minnesota", "Mississippi","Missouri", "Montana", 
                                                             "Nebraska", "Nevada", "New Hampshire", "New Jersey",
                                                             "New Mexico", "New York","North Carolina","North Dakota",
                                                             "Ohio", "Oklahoma","Oregon", "Pennsylvania", 
                                                             "Rhode Island", "South Carolina", "South Dakota", 
                                                             "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
                                                             "Washington", "West Virginia", "Wisconsin", "Wyoming")), 
         age = factor(age.n, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                         "15-19 years", "20-24 years", "25-29 years", "30-34 years",
                                                         "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
                                                         "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
                                                         "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
         age_minbin = 0*(age == "<1 year") + 1*(age == "1-4 years") + 5*(age == "5-9 years") +
                10*(age == "10-14 years") + 15*(age == "15-19 years") + 20*(age == "20-24 years") + 
                25*(age == "25-29 years") + 30*(age == "30-34 years") + 35*(age == "35-39 years") +
                40*(age == "40-44 years") + 45*(age == "45-49 years") + 50*(age == "50-54 years") + 
                55*(age == "55-59 years") + 60*(age == "60-64 years") + 65*(age == "65-69 years") +   
                70*(age == "70-74 years") + 75*(age == "75-79 years") + 80*(age == "80-84 years") + 
                85*(age == "85+ years"),
         age_bin_size = 1*(age == "<1 year") + 4*(age == "1-4 years") + 5*(age != "<1 year" & age != "1-4 years"),
         sex = factor(sex.n, levels = c(0, 1), labels = c("Male", "Female")),
         race = factor(race.n, levels = c(0, 1), labels = c("White", "Black")),
         year = factor(year.0, levels = c(0:19), labels = as.character(1990:2009)),
         year.n = as.numeric(as.character(year)),
         race.sex = interaction(race, sex, sep = " "),
         state.year.sex = interaction(state, year, sex),
         state.year.race.sex = interaction(state, year, race, sex),
         Crude_Rate = deaths/population, 
         COD = "Total")
```

```{r unknown_age_group}
#there is an unknown age group for each combo of state-year-race-sex-age:
table(dtotal$age)

#no rows with the unknown age group (guess: was removed when pulling the data from CDC wonder)
table(dtotal$age, dtotal$d==0)

#remove these data from the data we analyze:
dtotal <- dtotal[dtotal$age != "Unknown", ]
```

```{r exclude_states}
excluded.states <- c("Idaho", "Montana", "North Dakota", "South Dakota", "Vermont")
included.states <- setdiff(levels(dtotal$state), excluded.states)
dtotal <- dtotal %>% filter(state %in% included.states)

save(dtotal, file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/SH_HA_data.Rdata")
```

```{r}
prop.table(table(is.na(dtotal$deaths)))

prop.table(table(is.na(dtotal$deaths[dtotal$state == "New York"])))
prop.table(table(is.na(dtotal$deaths[dtotal$state == "Georgia"])))
prop.table(table(is.na(dtotal$deaths[dtotal$state == "Florida"])))
table(is.na(dtotal$deaths[dtotal$state == "Alabama" & dtotal$year.n == 1990]))

test.le.fx <- function(data, state, sex, race, year) {
  test.data <- data[data["state"] == state & 
                      data["sex"] == sex & 
                      data["race"] == race & 
                      data["year.n"] == year, ]

  lt <- life.table(data = test.data, num.ages.in.group = "age_bin_size",
                   death.counts = "deaths", population.counts = "population")
  return(round(lt$e_x[1], 1))
}

test.le.fx(dtotal, "Florida", "Male", "White", 1990)
test.le.fx(dtotal, "Florida", "Male", "White", 2009)

test.le.fx(dtotal, "Georgia", "Male", "White", 1990)
test.le.fx(dtotal, "Georgia", "Male", "White", 2009)

test.le.fx(dtotal, "New York", "Male", "White", 1990)
test.le.fx(dtotal, "New York", "Male", "White", 2009)

test.le.fx(dtotal, "Alabama", "Male", "White", 2009)

test.le.fx(dtotal, "Alabama", "Male", "White", 1990)
test.le.fx(dtotal, "Alabama", "Male", "Black", 1990)
#test.le.fx(dtotal, "Alabama", "Male", "Black", 2009)

```

### 2017-06-02 Code/mortality rate plots.R

@@ -1,41 +1,0 @@
-#random plots exploring the mortality rate
-load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
-
-ggplot(subset(dat.clean.alabama, Race2=="Black"), aes(x = Count, y=crude_new)) + geom_point(aes(col=Population)) + facet_wrap(~Age2, scales = "free")
-
-ggplot(subset(dat.clean.alabama, Race2=="Black" & Sex2=="Male"), aes(x = Population, y=crude_new)) + 
-  geom_point(aes(col=COD2)) + facet_wrap(~Age2, scales = "free") +
-  geom_smooth(col="black") + scale_y_continuous("Mortality Rate") + ggtitle("Black Men in Alabama (all years)")
-
-ggplot(subset(dat.clean.alabama, Race2=="Black" & Sex2=="Male"), aes(x = Population, y=crude_new)) + 
-  geom_point(aes(col=Year)) + facet_wrap(~Age2, scales = "free") +
-  geom_smooth(col="black") + scale_y_continuous("Mortality Rate") + ggtitle("Black Men in Alabama (all years)")
-
-
-ggplot(subset(dat.clean.alabama, Race2=="White" & Sex2=="Male"), aes(x = Population, y=crude_new)) + 
-  geom_point(aes(col=COD2)) + facet_wrap(~Age2, scales = "free") +
-  geom_smooth(col="black") + scale_y_continuous("Mortality Rate") + ggtitle("White Men in Alabama (all years)")
-
-ggplot(subset(dat.clean.alabama, Race2=="White" & Sex2=="Male"), aes(x = Population, y=crude_new)) + 
-  geom_point(aes(col=Year)) + facet_wrap(~Age2, scales = "free") +
-  geom_smooth(col="black") + scale_y_continuous("Mortality Rate") + ggtitle("White Men in Alabama (all years)")
-
-
-
-mean.rate = c()
-counts1 = unique(dat.clean.alabama$Count) ; length(counts)
-counts = counts1[2:length(counts1)]
-
-for(i in counts) {
-  mean.rate = c(mean.rate, mean(dat.clean.alabama$crude_new[dat.clean.alabama$Count==i], na.rm=T))
-}
-
-plot(counts, mean.rate, pch=16, cex=0.5)
-
-mean(dat.clean.alabama$crude_new[dat.clean.alabama$Count==10], na.rm = T)
[…]

### 2017-06-02 Code/validation/SH_examine_comparisons.Rmd
updated with new results

---
title: "Compare findings"
output: github_document
---

```{r load_libraries}
library(dplyr)
library(stringr)
library(ggplot2)
library(plotly)
```

```{r read_data}
BlackWhite_results <- read.csv("~/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
subset <- filter(BlackWhite_results, year %in% c(1990, 2009))
subset <- subset %>% select(state, year, sex, Census_Division, Census_Region, stabbrs, starts_with("LE"))

SH.dat <- read.csv("~/Documents/repos/BlackWhiteMortalityGap/Data/SH_HA_supp_app_data.csv", header = T)
```

```{r tidy_data}
SH.dat <- SH.dat %>% rename(state = State, 
                            le.black.SH = le.black,
                            le.white.SH = le.white,
                            le.gap.SH = le.gap, 
                            le.gap.ci.SH = le.gap.ci) %>%
  filter(year %in% c(1990, 2009))

compare.le <- merge(subset, SH.dat, by = c("state", "sex", "year"))

test <- strsplit(stringr::str_sub(compare.le$le.gap.ci.SH, 2, -2), ",")
test2 <- matrix(unlist(test), ncol=2, byrow=TRUE)

compare.le$le.gap.lcl.SH <- as.numeric(test2[ , 1])
compare.le$le.gap.ucl.SH <- as.numeric(test2[ , 2])
rm(test, test2)

compare.le <- compare.le %>% mutate(le.gap.width.SH = le.gap.ucl.SH - le.gap.lcl.SH,
                                    le.gap.width = LE_wbgap_ucl - LE_wbgap_lcl,
                                    small.black.state = state %in% c("Hawaii", "Alaska", "Maine", 
                                                                     "New Hampshire","Utah", "New Mexico")
                                    )
```

**Figure 1: Comparison of life expectancy between the original model and our model for whites**

```{r}
ggplotly(ggplot(compare.le, 
       aes(x = le.white.SH, y = LE_white_mean)) +
  geom_point(aes(col = state)) +
  facet_grid(year ~ sex ) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Life expectancy mean (years)") +
  xlab("Life expectancy estimate from Health Affairs paper (years)"))
```

**Figure 2: Comparison of life expectancy between the original model and our model for blacks**

```{r}
ggplotly(ggplot(compare.le, 
       aes(x = le.black.SH, y = LE_black_mean)) +
  geom_point(aes(col = state)) +
  facet_grid(year ~ sex ) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Life expectancy mean (years)") +
  xlab("Life expectancy estimate from Health Affairs paper (years)"))
```

**Figure 3: Comparison of life expectancy gap between the original model and our model**

```{r}
ggplotly(ggplot(compare.le, 
       aes(x = le.gap.SH, y = LE_wbgap_mean)) +
  geom_point(aes(col = state)) +
  facet_grid(year ~ sex ) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Life expectancy gap (years)") +
  xlab("Life expectancy gap from Health Affairs paper (years)")) 
```

**Figure 4: Comparison of CI width for LE gap between the original model and our model**

```{r}
ggplotly(ggplot(compare.le, 
       aes(x = le.gap.width.SH, y = le.gap.width)) +
  geom_point(aes(col = state)) +
  facet_grid(year ~ sex) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Width of CI for LE gap") +
  xlab("Original width of CI for LE gap") )
```

**OLD**
**Figure 2: Comparison of life expectancy change over time (1990 vs. 2009) between the original model and our model**

```{r}
ggplot(compare.le, 
       aes(x = le.white.original, y = le.white)) +
  geom_point(aes(x = le.black.original, y = le.black, col = "blacks")) + 
  geom_point(aes(col = "whites")) +
  facet_grid(~sex ) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Life expectancy change (years)") +
  xlab("Original life expectancy change (years)") +
  geom_text(data = subset(compare.le, small.black.state == T & year == "difference"),
            aes(x = le.black.original, y = le.black, label = state), size = 2.5)

```

**Figure 4: Comparison of change over time (1990 vs. 2009) in the life expectancy gap between the original model and our model**

```{r}
ggplot(subset(compare.le, state!= "Wyoming" & year == "difference"), 
       aes(x = le.gap.original, y = le.gap)) +
  geom_point() +
  facet_grid(.~sex ) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Change in life expectancy gap (years)") +
  xlab("Original change in life expectancy gap (years)") +
  geom_text(data = subset(compare.le, small.black.state == T & year == "difference"),
            aes(x = le.gap.original, y = le.gap, label = state), size = 2.5)
```

**Figure 6: Comparison of the change over time (1990 vs. 2009) in the CI width for LE gap between the original model and our model**

```{r}
ggplot(subset(compare.le, state!= "Wyoming" & year == "difference"), 
       aes(x = le.gap.width.original, y = le.gap.width)) +
  geom_point() +
  facet_grid(.~sex ) +
  geom_abline(intercept = 0, slope = 1) + 
  ylab("Width of CI for LE gap") +
  xlab("Original width of CI for LE gap") +
  geom_text(data = subset(compare.le, small.black.state == T & year == "difference"),
            aes(x = le.gap.width.original, y = le.gap.width, label = state), size = 2.5)
```

### 2017-06-02 shiny_app/app.R
improved LE trend plot

@@ -174,7 +174,7 @@ ui1 <- fluidPage(theme = "cosmo-customized.css",
                                           htmlOutput("description_state_snapshot"),
                                           plotlyOutput("population_trend", height = 300, width = 500),
                                           htmlOutput("title_LE_trends"),
-                                          plotlyOutput("life_expectancy"),
+                                          plotlyOutput("life_expectancy", height = 400, width = 770),
                                           htmlOutput("Explain_LE_females"),
                                           htmlOutput("Explain_LE_males"),
                                           plotlyOutput("age_cod", height = 700, width = 1100),
@@ -793,10 +793,10 @@ output$life_expectancy <- renderPlotly({
   
   p1 <- ggplotly(ggplot(subset(BlackWhite_results, state == input$state & year >= input$years_LEgap[1] &
                                  year <= input$years_LEgap[2]), aes(x = year, y = LE_white_mean)) +
-                   geom_line(col = "black") + 
-                   geom_ribbon(aes(ymin = LE_white_lcl, ymax = LE_white_ucl, fill = sex), alpha = 0.3) +
-                   geom_line(aes(y = LE_black_mean), lty = 2, col = "black") + 
-                   geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = sex), alpha = 0.3) +
+                   geom_ribbon(aes(ymin = LE_white_lcl, ymax = LE_white_ucl, fill = sex), alpha = 0.5) +
+                   geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = sex), alpha = 0.5) +
+                   geom_line(lty = 2) + 
+                   geom_line(aes(y = LE_black_mean), lty = 1) + 
                    facet_grid(. ~ sex, labeller = as_labeller(facet_names)) +
                    geom_text(data = subset(BlackWhite_results, year == 2013 & state == input$state),
                              aes(y = LE_white_mean - 1),
@@ -805,10 +805,8 @@ output$life_expectancy <- renderPlotly({
                              aes(y = LE_black_mean - 1),
                              label = "Black", check_overlap = T, size = 2.5) +
                    theme_minimal() +
-                   theme(legend.title = element_blank(),
-                         panel.background = element_rect(fill = "transparent", colour = NA), 
[…]

### 2017-06-04 Manuscript/Manuscript_COD_Trends_PNAS.Rmd

--- 2017-05-29 Manuscript/Manuscript_COD_Trends_PNAS.Rmd
@@
 ```
+
+**Figure 3 Legend:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of CVD to the black-white life expectancy gap over time.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
+The early spike in California's trend line illustrates the unexpectedly high estimate of the CVD contribution to the black-white life expectancy gap among California males in 1972.
+This anomolous estimate reflects the unexpectedly low CVD-related mortality rate among white males in California in 1972, and likely denotes a coding error.
+Thus the data for this state and year should be interpreted with caution.
 

### 2017-06-05 Code/validation/5 SH_second_comparison.Rmd
reconfig and new file

--- 2017-06-02 Code/validation/SH_examine_comparisons.Rmd
@@
 title: "Compare findings"
-output: github_document
+output: html_document
 ---
+
+In this file we compare our findings (using our data which is different because higher rate of missingness and more years of data) to Sam's Health Affairs appendix data.
 
@@
   ylab("Width of CI for LE gap") +
-  xlab("Original width of CI for LE gap") )
+  xlab("Width of CI for LE gap from Health Affairs paper") )
 ```
-
-**OLD**
-**Figure 2: Comparison of life expectancy change over time (1990 vs. 2009) between the original model and our model**
-
-```{r}
-ggplot(compare.le, 
-       aes(x = le.white.original, y = le.white)) +
-  geom_point(aes(x = le.black.original, y = le.black, col = "blacks")) + 
-  geom_point(aes(col = "whites")) +
-  facet_grid(~sex ) +
-  geom_abline(intercept = 0, slope = 1) + 
-  ylab("Life expectancy change (years)") +
-  xlab("Original life expectancy change (years)") +
-  geom_text(data = subset(compare.le, small.black.state == T & year == "difference"),
-            aes(x = le.black.original, y = le.black, label = state), size = 2.5)
-
-```
-
-**Figure 4: Comparison of change over time (1990 vs. 2009) in the life expectancy gap between the original model and our model**
-
-```{r}
-ggplot(subset(compare.le, state!= "Wyoming" & year == "difference"), 
-       aes(x = le.gap.original, y = le.gap)) +
-  geom_point() +
-  facet_grid(.~sex ) +
-  geom_abline(intercept = 0, slope = 1) + 
-  ylab("Change in life expectancy gap (years)") +
-  xlab("Original change in life expectancy gap (years)") +
-  geom_text(data = subset(compare.le, small.black.state == T & year == "difference"),
-            aes(x = le.gap.original, y = le.gap, label = state), size = 2.5)
-```
-
-**Figure 6: Comparison of the change over time (1990 vs. 2009) in the CI width for LE gap between the original model and our model**
-
-```{r}
-ggplot(subset(compare.le, state!= "Wyoming" & year == "difference"), 
-       aes(x = le.gap.width.original, y = le.gap.width)) +
-  geom_point() +
-  facet_grid(.~sex ) +
-  geom_abline(intercept = 0, slope = 1) + 
-  ylab("Width of CI for LE gap") +
-  xlab("Original width of CI for LE gap") +
-  geom_text(data = subset(compare.le, small.black.state == T & year == "difference"),
-            aes(x = le.gap.width.original, y = le.gap.width, label = state), size = 2.5)
-```

### 2017-06-05 Code/validation/4 SH_examine_comparisons.Rmd

--- 2017-01-24 Code/3_validation/SH_examine_comparisons.Rmd
@@
 title: "Compare findings"
-output: github_document
+output: html_document
 ---
+
+In this file we compare the LE estimates for Sam's Health Affairs paper to estimates we get from running our model on his data. 
+Thus the goal of this exercise is to answer the questions "are the models doing the same thing?"
 
@@
 ```{r read_data}
-le.table <- read.csv("~/BlackWhiteMortalityGap/Code/3_validation/le_table.csv", header = T)
-SH.dat <- read.csv("~/SH_HA_supp_app_data.csv", header = T)
+le.table <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/validation/le_table.csv", header = T) 
+#le_table was created in the SH_comparison.R file
+
+SH.dat <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/SH_HA_supp_app_data.csv",
+                   header = T) #supplementary appendix data -- so Sam's original estimates
 ```
@@
 
-write.csv(compare.le, "~/BlackWhiteMortalityGap/Code/3_validation/compare_LE.csv")
+write.csv(compare.le, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/validation/compare_LE.csv")
 ```

### 2017-06-06 Code/Prepare_Validation_Dataset.R
moving files

@@ -1,36 +1,0 @@
-
-not.suppressed <- foreign::read.dta("/Users/corinneriddell/Dropbox/VitalStats/notsuppressed_0610.dta")
-
-not.suppressed.grouped <- not.suppressed %>% 
-  filter(agegrp > 1) %>%
-  group_by(stabbrs, Year3, agegrp, Sex, Race, cod_cat) %>%
-  summarise(total_deaths = sum(death), 
-            region = first(region),
-            division = first(division)) %>%
-  mutate(COD2 = factor(cod_cat, levels = c("Cardiovascular", "Cancers", "Communicable", "Non-communicable", 
-                                           "Injuries","All other causes")))
-
-head(not.suppressed.grouped)
-
-load(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
-
-years0610 <- dat.clean[dat.clean$Year3 >= 2006 & dat.clean$Year3 <= 2010 & dat.clean$Age > 1, ] 
-years0610$agegrp <- years0610$Age
-
-years0610 <- years0610 %>% mutate(stabbrs = factor(State, levels = c(0:50), 
-                                                 labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
-                                                            "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", 
-                                                            "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", 
-                                                            "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", 
-                                                            "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
-                                                            "VT", "VA", "WA", "WV", "WI", "WY")))
-
-validation_data_0610 <- merge(years0610, not.suppressed.grouped, by = c("stabbrs", "Year3", "agegrp", "Sex", "Race", "COD2"))
-
-validation_data_0610 <- validation_data_0610 %>% mutate(diff.count2 = Count - total_deaths)
-summary(validation_data_0610$diff.count2)
-
-validation_data_0610 <- validation_data_0610[, -c(33:44, 49)]
-validation_data_0610 <- validation_data_0610[, -c(24:32)]
-
-save(validation_data_0610, file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/validation_data_0610.Rdata")

### 2017-06-06 Code/5_RI_sensitivity.Rmd
CR sensitivity analysis

---
title: "Rhode_Island_sensitivity"
output: html_document
---

```{r setup, include=FALSE}
dat.clean = read.csv('~/dat_clean.csv')

RI <- dat.clean %>% filter(state == "Rhode Island")

RI.1999 <- RI %>% filter(year >= 1999)

rm(dat.clean)
```

```{r}
entire.analysis <- function(state_i, chosen.seed) {

  #load libraries
  library(Rcpp)
  library(dplyr)
  library(purrr)
  library(ggplot2)
  library(stringr)
  library(rjags)
  library(R2jags) 
  library(devtools)
  library(jagstools)
  library(reshape2)
  library(parallel)
  library(gtools)
  library(bayesboot)
  library(coda)
  library(grid)
  library(gridExtra)
  library(parallel)

  #load user written functions
  source("~/repos/BlackWhiteMortalityGap/Code/2_smoothing_models/bayes_smoothing_functions.R")
  source('~/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R')

  #load tidied data
  dat.clean = RI.1999

  #load US standard population
  std.pop <- read.csv("~/repos/BlackWhiteMortalityGap/Data/US_Standard_Population_for_model.csv") %>% select(US_Standard_2000, age.weight, age_minbin)

  #fit models
  #run time: 32 mins
  #fit the bayesian models and put the 1000 posterior samples into a data frame
  for(sex_i in c("Male", "Female")){
    print(sex_i)
    assign(paste0("r.White.", sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "White", chosen.seed))
    assign(paste0("r.Black.", sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "Black", chosen.seed))
  }

  r.White.Male$race <- "White"
  r.White.Female$race <- "White"
  r.Black.Male$race <- "Black"
  r.Black.Female$race <- "Black"
  r.All <- rbind(r.White.Male, r.White.Female, r.Black.Male, r.Black.Female)
  rm(r.White.Male, r.White.Female, r.Black.Male, r.Black.Female)

  #calculate age-standardized mortality rates (for each posterior sample)
  r.All <- merge(r.All, std.pop, by = "age_minbin")

  mortality.rates <- r.All %>% 
    group_by(post.samp, state.n, year.n, race.n, sex.n, COD) %>% 
    arrange(age_minbin) %>%
    summarise(age.std.mortality.rate = sum(age.weight*smoothed_rate), #the sum is across age bins
              rate.per.100k = age.std.mortality.rate*100000,
              race = first(race),
              sex = first(sex),
              state = first(state),
              year = first(year))

  mortality.rates <- add_Census_Division(mortality.rates, state)
  mortality.rates <- add_Census_Region(mortality.rates, state)
  mortality.rates$stabbrs <- state.abb[match(mortality.rates$state, state.name)]

  #calculate median, mean, and CI for age-standardized mortality rates (across samples) 
  mortality.rates.summary <- mortality.rates %>% group_by(state.n, year.n, race.n, sex.n, COD) %>%
    summarise(race = first(race),
              sex = first(sex),
              state = first(state),
              year = first(year),
              Census_Division = first(Census_Division),
              Census_Region = first(Census_Region),
              stabbrs = first(stabbrs),
              rate.per.100k_lcl = quantile(rate.per.100k, 0.025), 
              rate.per.100k_med = quantile(rate.per.100k, 0.5),
              rate.per.100k_ucl = quantile(rate.per.100k, 0.975),
              rate.per.100k_mean = mean(rate.per.100k))

  #compute black-white difference in mortality
  mortality.rates.white <- mortality.rates %>% filter(race == "White") %>% 
    ungroup() %>%
    select(sex, state, year, COD, post.samp, rate.per.100k) %>%
    rename(rate.per.100k.white = rate.per.100k)

  mortality.rates.black <- mortality.rates %>% filter(race == "Black") %>% 
    rename(rate.per.100k.black = rate.per.100k)

  mortality.rates.wide <- merge(mortality.rates.black, mortality.rates.white, by = c("post.samp", "sex", "state", "year", "COD"))

  mortality.rates.wide <- mortality.rates.wide %>% mutate(rate.difference = rate.per.100k.black- rate.per.100k.white)

  mortality.rates.difference <- mortality.rates.wide %>% 
    group_by(state.n, year.n, sex.n, COD) %>%
    summarise(rate.difference_LCL = quantile(rate.difference, 0.025), #summarize across posterior samples (post.samp)
              rate.difference_med = quantile(rate.difference, 0.5),
              rate.difference_UCL = quantile(rate.difference, 0.975),
              rate.difference_mean = mean(rate.difference),
              Census_Division = first(Census_Division),
              Census_Region = first(Census_Region),
              stabbrs = first(stabbrs),
              sex = first(sex),
              state = first(state),
              year = first(year))

  r.All$smoothed_deaths <- r.All$smoothed_rate*r.All$population
  table(r.All$smoothed_deaths == 0) #check this -- if no deaths in an age-band (i.e., no death for any COD, then can't run the COD calculations for that stratum.)

  r.All <- r.All %>% group_by(post.samp, state, race.sex, year.n, age_minbin) %>%
    arrange(post.samp, state, race.sex, year.n, age_minbin)

  #RUN LIFE EXPECTANCY AND DECOMPOSITION CALCULATIONS
  #run time: 2.25 hours per state
  #run the life table and decomposition calculations for each of the 1000 posterior samples

  strata <- levels(droplevels(r.All$state.year.sex))
  n <- length(strata)

  le.calculations = vector(mode = "list", length = n)

  counter = 1

  for(stratum_i in strata){
    data_sub <- subset(r.All, state.year.sex == stratum_i)

    le.calculations[[counter]]$stratum.id <- stratum_i
    le.calculations[[counter]]$calcs <- by(data = data_sub,
                                           INDICES = list(data_sub$post.samp),
                                           FUN = function(x) life_expectancy_and_gap(data = x))

    counter <- counter + 1

    print(stratum_i)
    rm(data_sub)
  }

  save(r.All, file = paste0("~/RhodeIsland/r_All_", state_i, "_", chosen.seed, "_1999.Rdata")) 
  levels.state.year.sex <- levels(droplevels(r.All$state.year.sex))
  rm(r.All)

  #extract_life_expectancy_results
  BlackWhite <- data.frame(stratum.id = character(),
                           LE_white_lcl = numeric(), LE_white_med = numeric(), LE_white_ucl = numeric(), LE_white_mean = numeric(), 
                           LE_black_lcl = numeric(), LE_black_med = numeric(), LE_black_ucl = numeric(), LE_black_mean = numeric(), 
                           LE_wbgap_lcl = numeric(), LE_wbgap_med = numeric(), LE_wbgap_ucl = numeric(), LE_wbgap_mean = numeric(), 
                           stringsAsFactors = F)

  age_minbin <- le.calculations[[1]]$calcs[[1]]$age.decomp$Ages
  cods <- le.calculations[[1]]$calcs[[1]]$cod.marginal$Cause.of.death

  age.decomp.estimates <- data.frame(stratum.id = rep(levels.state.year.sex, 
                                                      each = length(age_minbin)),
                                     age_minbin = numeric(19*n),
                                     age_cont_yrs_lcl = numeric(19*n), 
                                     age_cont_yrs_med = numeric(19*n), 
                                     age_cont_yrs_ucl = numeric(19*n),
                                     age_cont_yrs_mean = numeric(19*n), 
                                     age_cont_prop_lcl = numeric(19*n), 
                                     age_cont_prop_med = numeric(19*n), 
                                     age_cont_prop_ucl = numeric(19*n),
                                     age_cont_prop_mean = numeric(19*n), 
                                     stringsAsFactors = F)

  cod.marginal.estimates <- data.frame(stratum.id = rep(levels.state.year.sex, 
                                                        each = length(cods)),
                                       COD = character(6*n),
                                       COD_cont_yrs_lcl = numeric(6*n), 
                                       COD_cont_yrs_med = numeric(6*n), 
                                       COD_cont_yrs_ucl = numeric(6*n),
                                       COD_cont_yrs_mean = numeric(6*n), 
                                       COD_cont_prop_lcl = numeric(6*n), 
                                       COD_cont_prop_med = numeric(6*n), 
                                       COD_cont_prop_ucl = numeric(6*n),
                                       COD_cont_prop_mean = numeric(6*n), 
                                       stringsAsFactors = F)

  age.cod.estimates <- data.frame(stratum.id = rep(levels.state.year.sex, 
                                                   each = length(cods)*length(age_minbin)),
                                  age_minbin = numeric(19*6*n),   
                                  COD = character(19*6*n),
                                  age_COD_cont_yrs_lcl = numeric(19*6*n), 
                                  age_COD_cont_yrs_med = numeric(19*6*n), 
                                  age_COD_cont_yrs_ucl = numeric(19*6*n),
                                  age_COD_cont_yrs_mean = numeric(19*6*n),
                                  age_COD_cont_prop_lcl = numeric(19*6*n), 
                                  age_COD_cont_prop_med = numeric(19*6*n), 
                                  age_COD_cont_prop_ucl = numeric(19*6*n), 
                                  age_COD_cont_prop_mean = numeric(19*6*n), 
                                  stringsAsFactors = F)

  for(i in 1:n) {

    BlackWhite[i, c("stratum.id")] <- le.calculations[[i]]$stratum.id

    #LIFE EXPECTANCY EXTRACT RESULTS
    BlackWhite[i , c("LE_white_lcl", "LE_white_med", "LE_white_ucl")] <- quantile(map_dbl(le.calculations[[i]]$calcs,
                                                                                          ~.x$le.df$LE_White), 
                                                                                  c(0.025, 0.5, 0.975))
    BlackWhite[i , c("LE_black_lcl", "LE_black_med", "LE_black_ucl")] <- quantile(map_dbl(le.calculations[[i]]$calcs,
                                                                                          ~.x$le.df$LE_Black), 
                                                                                  c(0.025, 0.5, 0.975))
    BlackWhite[i , c("LE_wbgap_lcl", "LE_wbgap_med", "LE_wbgap_ucl")] <- quantile(map_dbl(le.calculations[[i]]$calcs,
                                                                                          ~.x$le.df$LE_WBgap), 
                                                                                  c(0.025, 0.5, 0.975))

    BlackWhite[i , c("LE_white_mean")] <- mean(map_dbl(le.calculations[[i]]$calcs, ~.x$le.df$LE_White))
    BlackWhite[i , c("LE_black_mean")] <- mean(map_dbl(le.calculations[[i]]$calcs, ~.x$le.df$LE_Black))
    BlackWhite[i , c("LE_wbgap_mean")] <- mean(map_dbl(le.calculations[[i]]$calcs, ~.x$le.df$LE_WBgap))

    #AGE DECOMPOSITION EXTRACT RESULTS
    age.decomp.posteriors <- map_df(le.calculations[[i]]$calcs, ~ .x$age.decomp$C_x)
    age.decomp.posteriors$age_minbin <- age_minbin
    age.decomp.posteriors <- age.decomp.posteriors %>% tidyr::gather(`1`:`1000`, key = "post.samp", value = "C_x")

    age.decomp.estimates[age.decomp.estimates$stratum.id == le.calculations[[i]]$stratum.id, 
                         c("age_minbin", "age_cont_yrs_lcl", "age_cont_yrs_med", "age_cont_yrs_ucl", "age_cont_yrs_mean")] <- 
      age.decomp.posteriors %>% 
      group_by(age_minbin) %>%
      summarise(lcl_C_x = quantile(C_x, 0.025),
                median_C_x = quantile(C_x, 0.5), 
                ucl_C_x = quantile(C_x, 0.975),
                mean_C_x = mean(C_x))

    rm(age.decomp.posteriors)

    age.decomp.posteriors2 <- map_df(le.calculations[[i]]$calcs, ~ .x$age.decomp$C_x_proportion)
    age.decomp.posteriors2$age_minbin <- age_minbin
    age.decomp.posteriors2 <- age.decomp.posteriors2 %>% tidyr::gather(`1`:`1000`, key = "post.samp", value = "C_x_proportion")

    age.decomp.estimates[age.decomp.estimates$stratum.id == le.calculations[[i]]$stratum.id, 
                         c("age_minbin", "age_cont_prop_lcl", "age_cont_prop_med", "age_cont_prop_ucl", "age_cont_prop_mean")] <- 
      age.decomp.posteriors2 %>% 
      group_by(age_minbin) %>%
      summarise(lcl_C_x_proportion = quantile(C_x_proportion, 0.025),
                median_C_x_proportion = quantile(C_x_proportion, 0.5), 
                ucl_C_x_proportion = quantile(C_x_proportion, 0.975),
                mean_C_x_proportion = mean(C_x_proportion))

    rm(age.decomp.posteriors2)

    #COD DECOMPOSITION EXTRACT RESULTS
    cod.marginal.posteriors <- map_df(le.calculations[[i]]$calcs, ~ .x$cod.marginal$C_x)
    cod.marginal.posteriors$COD <- as.character(cods)
    cod.marginal.posteriors <- cod.marginal.posteriors %>% tidyr::gather(`1`:`1000`, key = "post.samp", value = "C_x")

    cod.marginal.estimates[cod.marginal.estimates$stratum.id == le.calculations[[i]]$stratum.id, 
                           c("COD", "COD_cont_yrs_lcl", "COD_cont_yrs_med", "COD_cont_yrs_ucl", "COD_cont_yrs_mean")] <- 
      cod.marginal.posteriors %>% 
      group_by(COD) %>%
      summarise(lcl_C_x = quantile(C_x, 0.025),
                median_C_x = quantile(C_x, 0.5), 
                ucl_C_x = quantile(C_x, 0.975),
                mean_C_x = mean(C_x))

    rm(cod.marginal.posteriors)

    cod.marginal.posteriors2 <- map_df(le.calculations[[i]]$calcs, ~ .x$cod.marginal$C_x_proportion)
    cod.marginal.posteriors2$COD <- as.character(cods)
    cod.marginal.posteriors2 <- cod.marginal.posteriors2 %>% tidyr::gather(`1`:`1000`, key = "post.samp", 
                                                                           value = "C_x_proportion")

    cod.marginal.estimates[cod.marginal.estimates$stratum.id == le.calculations[[i]]$stratum.id,
                           c("COD", "COD_cont_prop_lcl", "COD_cont_prop_med", "COD_cont_prop_ucl", "COD_cont_prop_mean")] <-
      cod.marginal.posteriors2 %>%
      group_by(COD) %>%
      summarise(lcl_C_x_proportion = quantile(C_x_proportion, 0.025),
                median_C_x_proportion = quantile(C_x_proportion, 0.5),
                ucl_C_x_proportion = quantile(C_x_proportion, 0.975),
                mean_C_x_proportion = mean(C_x_proportion))

    rm(cod.marginal.posteriors2)

    #AGE-COD EXTRACT RESULTS
    age.cod.decomp.posteriors <- map_df(le.calculations[[i]]$calcs, ~ .x$cod.decomp$C_xi)
    age.cod.decomp.posteriors$age_minbin <- le.calculations[[i]]$calcs[[1]]$cod.decomp$Ages
    age.cod.decomp.posteriors$COD <- le.calculations[[i]]$calcs[[1]]$cod.decomp$Cause.of.death
    age.cod.decomp.posteriors <- age.cod.decomp.posteriors %>% tidyr::gather(`1`:`1000`, key = "post.samp", value = "C_xi")

    age.cod.estimates[age.cod.estimates$stratum.id == le.calculations[[i]]$stratum.id,
                      c("age_minbin", "COD", "age_COD_cont_yrs_lcl", "age_COD_cont_yrs_med", "age_COD_cont_yrs_ucl", "age_COD_cont_yrs_mean")] <-
      age.cod.decomp.posteriors %>%
      group_by(age_minbin, COD) %>%
      summarise(lcl_C_x = quantile(C_xi, 0.025),
                median_C_x = quantile(C_xi, 0.5),
                ucl_C_x = quantile(C_xi, 0.975),
                mean_C_x = mean(C_xi))

    rm(age.cod.decomp.posteriors)

    age.cod.decomp.posteriors2 <- map_df(le.calculations[[i]]$calcs, ~ .x$cod.decomp$C_xi_proportion)
    age.cod.decomp.posteriors2$age_minbin <- le.calculations[[i]]$calcs[[1]]$cod.decomp$Ages
    age.cod.decomp.posteriors2$COD <- le.calculations[[i]]$calcs[[1]]$cod.decomp$Cause.of.death
    age.cod.decomp.posteriors2 <- age.cod.decomp.posteriors2 %>% tidyr::gather(`1`:`1000`, key = "post.samp", 
                                                                               value = "C_xi_proportion")

    age.cod.estimates[age.cod.estimates$stratum.id == le.calculations[[i]]$stratum.id,
                      c("age_minbin", "COD", "age_COD_cont_prop_lcl", "age_COD_cont_prop_med", "age_COD_cont_prop_ucl", "age_COD_cont_prop_mean")] <-
      age.cod.decomp.posteriors2 %>%
      group_by(age_minbin, COD) %>%
      summarise(lcl_C_x_proportion = quantile(C_xi_proportion, 0.025),
                median_C_x_proportion = quantile(C_xi_proportion, 0.5),
                ucl_C_x_proportion = quantile(C_xi_proportion, 0.975),
                mean_C_x_proportion = mean(C_xi_proportion))

    rm(age.cod.decomp.posteriors2)
  }

  #save_results
  save(mortality.rates, mortality.rates.wide, le.calculations,
       file = paste0("~/RhodeIsland/results_", state_i, "_", chosen.seed, "1999.Rdata")) 
  write.csv(x = BlackWhite, file = paste0("~/repos/BlackWhiteMortalityGap/sensitivity/BlackWhite_", state_i, "_", 1999,".csv"))
  write.csv(x = age.decomp.estimates, file = paste0("~/repos/BlackWhiteMortalityGap/sensitivity/age_decomp_", state_i, "_", 1999,".csv"))
  write.csv(x = cod.marginal.estimates, file = paste0("~/repos/BlackWhiteMortalityGap/sensitivity/cod_marginal_", state_i, "_", 1999,".csv"))
  write.csv(x = age.cod.estimates, file = paste0("~/repos/BlackWhiteMortalityGap/sensitivity/age_cod_", state_i, "_", 1999, ".csv"))
  write.csv(x = mortality.rates.summary, file = paste0("~/repos/BlackWhiteMortalityGap/sensitivity/mortality_rates_", state_i, "_", 1999, ".csv"))
  write.csv(x = mortality.rates.difference, file = paste0("~/repos/BlackWhiteMortalityGap/sensitivity/mortality_rates_diff_", state_i, "_", 1999, ".csv"))
}
```

```{r run_analysis}
system.time(entire.analysis(state_i = "Rhode Island", chosen.seed = 123))
```

### 2017-06-06 Code/4_combined-meta-regression.Rmd
made paths relativ

@@
 
-In the paper, we perform a random-effects meta regression, that includes random intercepts for each state and plot the general pattern (when the RE estimate is equal to 0).
+In the working paper, we perform a random-effects meta regression, that includes random intercepts for each state and plot the general pattern (when the RE estimate is equal to 0).
 In this file we also include code for the fixed effect estimation, for interested readers.
-This model directly estimates vertical shifts to the mean curve for each state, which may be of interest to some readers.
+This model directly estimates vertical shifts to the mean curve for each state, which may be of interest.
 
 ```{r initialize, warning=F, message=F}
-library(lme4)
+library(lme4) #for random effects models
 library(splines)
@@
 
-source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+source("./Code/life_expectancy_functions.R")
 ```
@@
 ```{r load-data}
-cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
+cod_decomp_results <- read.csv("./Results2/cod_decomp_results.csv")
 cod_decomp_results$COD_cont_yrs_width <- cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl
@@
 
-age_cod_results_female <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_female.csv")
-age_cod_results_male <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_male.csv")
+age_cod_results_female <- read.csv("./Results2/age_cod_results_female.csv")
+age_cod_results_male <- read.csv("./Results2/age_cod_results_male.csv")
 age_cod_results <- rbind(age_cod_results_female, age_cod_results_male)
@@
 age_cod_results$age_COD_cont_yrs_SE <- (age_cod_results$age_COD_cont_yrs_ucl - age_cod_results$age_COD_cont_yrs_lcl)/3.92
+
+#the following is for our sensitivity analysis when we look only at infants and consider only the cause that 
+#captures "all other causes" (since this is the most common cause for this age group).
 age_cod_results_infants_AOC <- subset(age_cod_results, age_minbin == 0 & COD == "All other causes")
@@
 ```{r RE_meta_regression_of_cod_contribution, echo = F}
-#Random effects meta analysis (fixed effects for state)
+#Random effects meta analysis (random intercepts for state)
 
@@
 ```{r save_data}
-write.csv(cod_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
+write.csv(cod_decomp_results, "./Results2/cod_decomp_results.csv")
 
-write.csv(age_cod_results_infants_AOC, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_infants_AOC.csv")
+write.csv(age_cod_results_infants_AOC, "./Results2/age_cod_results_infants_AOC.csv")
 ```

### 2017-06-06 Code/2_entire_analysis.R

@@ -1,3 +1,9 @@
+#This function runs the bayesian time series model on a specified state. 
+#It then performed the demographic calculations for each of 1000 posterior samples and compiles the results as the mean and CI.
+#NOTE: If you run this code (using the "2*_run-models.R file ) change the path directories on lines #161 and #508 or comment 
+#out these lines. These lines save large Rdata files that are not in the repo. They contain intermediate results that aren't necessary
+#for the anaysis, which you may or may not want to save.
+
 entire.analysis <- function(state_i, chosen.seed) {
   
   #load libraries
@@ -20,22 +26,22 @@ entire.analysis <- function(state_i, chosen.seed) {
   library(parallel)
   
   #load user written functions
-  source("~/repos/BlackWhiteMortalityGap/Code/2_smoothing_models/bayes_smoothing_functions.R")
-  source('~/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R') 
+  source("./Code/2_smoothing_models/bayes_smoothing_functions.R")
+  source('./Code/life_expectancy_functions.R') 
   
-  #load tidied data
-  dat.clean = read.csv('~/dat_clean.csv') 
+  #load state-specific tidied data
+  dat.clean = read.csv(paste0('./Data/dat_clean_', state_i, ".csv")) 
   
   #load US standard population
-  std.pop <- read.csv("~/repos/BlackWhiteMortalityGap/Data/US_Standard_Population_for_model.csv") %>% select(US_Standard_2000, age.weight, age_minbin)
+  std.pop <- read.csv("./Data/US_Standard_Population_for_model.csv") %>% select(US_Standard_2000, age.weight, age_minbin)
   
   #fit models
-  #run time: 32 mins
+  #run time: 32 mins for one state
   #fit the bayesian models and put the 1000 posterior samples into a data frame
   for(sex_i in c("Male", "Female")){
     print(sex_i)
-    assign(paste0("r.White.", sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "White", chosen.seed))
-    assign(paste0("r.Black.", sex_i), run_analysis(dataset = dat.clean, state1 = state_i, sex1 = sex_i, race1 = "Black", chosen.seed))
[…]

### 2017-06-06 Code/2_smoothing_models/bayes_smoothing_functions.R

@@ -1,7 +1,7 @@
 
 # User-defined subsets 
-subset_data = function(ds, state, sex, race) {
-  ds_sub = ds[ds$state==state & ds$sex==sex & ds$race==race, ]
+subset_data = function(ds, sex, race) {
+  ds_sub = ds[ds$sex==sex & ds$race==race, ]
   return(ds_sub)
 }
 
@@ -94,7 +94,7 @@ run_smoothing_model = function(data, chosen.seed) {
 }
 
 # Takes the jagsify_data output, the run_smoothing_model output, and the desired year as parameters
-# Outputs the original data, subset by sex, race, state, and year (not age or COD), with smoothed rates and smoothed/imputed deaths 
+# Outputs the original data, subset by sex, race, and year (not age or COD), with smoothed rates and smoothed/imputed deaths 
 clean_smoothing_results = function(data, jags_model) {
   ds_sub = data$sub
   cods = unique(ds_sub$COD)
@@ -129,8 +129,8 @@ clean_smoothing_results = function(data, jags_model) {
   return(merge(data$sub, smoothed.rates.1000, by = c("year.n", "age.n", "COD")))
 }
  
-run_analysis <- function(dataset, state1, sex1, race1, chosen.seed) {
-  ds_sub = subset_data(ds = dataset, state = state1, sex = sex1, race = race1)
+run_analysis <- function(dataset, sex1, race1, chosen.seed) {
+  ds_sub = subset_data(ds = dataset, sex = sex1, race = race1)
   data.jags = jagsify_data(ds_sub)
   jags.model = run_smoothing_model(data.jags, chosen.seed)  
   return(clean_smoothing_results(data = data.jags, jags_model = jags.model))

### 2017-06-06 Code/2a_run-models.R

@@ -1,3 +1,11 @@
+#In this file, we call the function to run the analysis on each state.
+
+#Originally, I ran the analysis in three separate tmux (terminal multiplex) windows, hence the three files for running the models.
+#This is a less sophisticated way of running the models in parallel, because when I tried to parallelize it within RStudio
+#it would take over the server and use all the cores, even when I told it to only use three.
+
+#NOTE: see the "NOTE" in 2_entire_analysis.R that specifies two lines of code within that file that you will need to change
+#to prevent errors!
 
 for(i in c(1:13)){
   
@@ -9,7 +17,7 @@ for(i in c(1:13)){
                        "South Carolina",  "Tennessee", "Texas", "Virginia", "Washington", "Washington DC",
                        "West Virginia", "Wisconsin") 
   
-  source("~/repos/BlackWhiteMortalityGap/Code/entire_analysis.R")
+  source("./Code/entire_analysis.R")
   system.time(entire.analysis(state_i = included.states[i], chosen.seed = 123))
   rm(list = ls(all = TRUE))
 }

### 2017-06-06 Code/2b_run-models-second-window.R

@@ -9,7 +9,7 @@ for(i in c(14:27)){
                        "South Carolina",  "Tennessee", "Texas", "Virginia", "Washington", "Washington DC",
                        "West Virginia", "Wisconsin") 
   
-  source("~/repos/BlackWhiteMortalityGap/Code/entire_analysis.R")
+  source("./Code/entire_analysis.R")
   system.time(entire.analysis(state_i = included.states[i], chosen.seed = 123))
   rm(list = ls(all = TRUE))
 }

### 2017-06-06 Code/1_Explore_Data.Rmd

@@
 ```{r read_raw_data}
-dat <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/deaths-cause-6913.txt", header = F,sep="\t")
+dat <- read.table("./Data/deaths-cause-6913.txt", header = F, sep="\t")
 names(dat) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
@@
 ```{r write_data_to_csvs}
-#save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
-write_csv(dat.clean, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.csv", col_names = T)
-write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.csv", col_names = T)
+write_csv(dat.aggregated, path = "./dat_aggregated.csv", col_names = T)
 
@@
   ds_temp = dat.clean[dat.clean$state == included.states[i], ]
-  write.csv(ds_temp, file = paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/", name))
+  write.csv(ds_temp, file = paste0("./Data/", name))
 }
@@
 ```
-
-```{r save_specified_files}
-# save(dat.clean, dat2, BlackWhite, BlackWhiteSmooth, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
-#      list.age.decomp.tables5, list.cod.decomp.tables5, list.age.decomp.tables9, list.cod.decomp.tables9,
-#      list.age.decomp.tables.smoothed, list.cod.decomp.tables.smoothed, list.cod.marginal.tables.smoothed,
-#      non.zero.stratums, non.zero.stratums.smoothed, cod.contributions,
-#      file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
-```

### 2017-06-06 Code/3_combine_results.Rmd

@@
 **Combine results across the sets**
-The analysis was run in 8 sets. In this file we combine the results and tidy the data for graph and presentation.
+The analysis was run in 3 sets of states. In this file we combine the results and tidy the data for graphing and presentation.
 
@@
 library(dplyr)
-source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+source("./Code/life_expectancy_functions.R")
 ```
@@
 "West Virginia", "Wisconsin") 
-
-completed.states <- included.states
 ```
@@
 ```{r combine_results_seed123}
-age_cod_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_", completed.states[1], "_123.csv"))
-age_decomp_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_decomp_", completed.states[1], "_123.csv"))
-BlackWhite_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_", completed.states[1], "_123.csv"))
-cod_decomp_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_marginal_", completed.states[1], "_123.csv"))
-cod_change_results <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_change_", completed.states[1], "_123.csv"))
+age_cod_results <- read.csv(paste0("./Results2/age_cod_", included.states[1], "_123.csv"))
+age_decomp_results <- read.csv(paste0("./Results2/age_decomp_", included.states[1], "_123.csv"))
+BlackWhite_results <- read.csv(paste0("./Results2/BlackWhite_", included.states[1], "_123.csv"))
+cod_decomp_results <- read.csv(paste0("./Results2/cod_marginal_", included.states[1], "_123.csv"))
+cod_change_results <- read.csv(paste0("./Results2/cod_change_", included.states[1], "_123.csv"))
 
-mortality.rates <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/mortality_rates_", completed.states[1], "_123.csv"))
-mortality.rates.diff <- read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/mortality_rates_diff_", completed.states[1], "_123.csv"))
+mortality.rates <- read.csv(paste0("./Results2/mortality_rates_", included.states[1], "_123.csv"))
+mortality.rates.diff <- read.csv(paste0("./Results2/mortality_rates_diff_", included.states[1], "_123.csv"))
 
-for(state_i in 2:length(completed.states)){
+for(state_i in 2:length(included.states)){
 
-    age_cod_results <- rbind(age_cod_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_", completed.states[state_i], "_123.csv")))
+    age_cod_results <- rbind(age_cod_results, read.csv(paste0("./Results2/age_cod_", included.states[state_i], "_123.csv")))
 
-    age_decomp_results <- rbind(age_decomp_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_decomp_", completed.states[state_i], "_123.csv")))
+    age_decomp_results <- rbind(age_decomp_results, read.csv(paste0("./Results2/age_decomp_", included.states[state_i], "_123.csv")))
 
-    BlackWhite_results <- rbind(BlackWhite_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_", completed.states[state_i], "_123.csv")))
+    BlackWhite_results <- rbind(BlackWhite_results, read.csv(paste0("./Results2/BlackWhite_", included.states[state_i], "_123.csv")))
 
-    cod_decomp_results <- rbind(cod_decomp_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_marginal_", completed.states[state_i], "_123.csv")))
+    cod_decomp_results <- rbind(cod_decomp_results, read.csv(paste0("./Results2/cod_marginal_", included.states[state_i], "_123.csv")))    
+    cod_change_results <- rbind(cod_change_results, read.csv(paste0("./Results2/cod_change_", included.states[state_i], "_123.csv")))    
+    mortality.rates <- rbind(mortality.rates, read.csv(paste0("./Results2/mortality_rates_", included.states[state_i], "_123.csv")))
 
-    cod_change_results <- rbind(cod_change_results, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_change_", completed.states[state_i], "_123.csv")))
-
-  mortality.rates <- rbind(mortality.rates, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/mortality_rates_", completed.states[state_i], "_123.csv")))
-
-    mortality.rates.diff <- rbind(mortality.rates.diff, read.csv(paste0("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/mortality_rates_diff_", completed.states[state_i], "_123.csv")))
+    mortality.rates.diff <- rbind(mortality.rates.diff, read.csv(paste0("./Results2/mortality_rates_diff_", included.states[state_i], "_123.csv")))
 }
@@
 
-age_cod_results$age <- factor(age_cod_results$age_minbin, levels = c(0, 1, seq(5, 85, by = 5)),
+age_cod_results$age <- factor(age_cod_results$age_minbin, 
+                              levels = c(0, 1, seq(5, 85, by = 5)),
                               labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
@@
 
-age_decomp_results$age <- factor(age_decomp_results$age_minbin, levels = c(0, 1, seq(5, 85, by = 5)),
+age_decomp_results$age <- factor(age_decomp_results$age_minbin, 
+                                 levels = c(0, 1, seq(5, 85, by = 5)),
                                  labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
@@
                                            type.of.data = "practice")
-
-#trying to get this to work 
-age_cod_results_test <- make_pretty_decomp_plot(decomp.table = subset(age_cod_results, stratum.id == "Alabama.1969.Female"), 
-                                                strat.var.1 = "stratum.id",
-                                           strat.var.2 = "state", strat.var.3 = "age_minbin",
-                                           partition.bar.var = "COD", sign.var = "widened_gap",
-                                           decomp.var = "age_COD_cont_yrs_mean", decomp.var.prop = "age_COD_cont_prop_mean",
-                                           type.of.data = "results")
-
-ggplot(data = age_cod_results_test, aes(y = age, x = start)) + geom_bar(fill = COD)
-
-age_cod_results_test$age.n <- as.numeric(age_cod_results_test$age) #need to fix the 5 year old when convert to nummeric
-    ggplotly(ggplot(age_cod_results_test,
-                    aes(y = age.n, x = start)) + 
-               geom_rect(aes(xmin = start, xmax = finish, ymin = age.n - 0.45, ymax = age.n + 0.45, fill = COD), col = "white") +
-               #xlab(paste0("Contribution to life expectancy gap ", xaxis.title())) +
-               #ylab("Age group\n") +
-               theme_minimal() +
-               geom_vline(xintercept = 0)  
-               #facet_wrap(~ year) +
-               #ggtitle(paste0(input$selected_sex, "s in ", input$state)) 
-             ) 
 ```
@@
 ```{r save_combined_datasets, echo = F}
-write.csv(age_cod_results_female, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_female.csv")
-write.csv(age_cod_results_male, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_male.csv")
-write.csv(age_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_decomp_results.csv")
-write.csv(cod_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
-write.csv(BlackWhite_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
-write.csv(cod_change_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_change_results.csv")
-
-write.csv(mortality.rates, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/mortality_rates_combined.csv")
-write.csv(mortality.rates.diff, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/mortality_rates_diff_combined.csv")
+write.csv(age_cod_results_female, "./Results2/age_cod_results_female.csv")
+write.csv(age_cod_results_male, "./Results2/age_cod_results_male.csv")
+write.csv(age_decomp_results, "./Results2/age_decomp_results.csv")
+write.csv(cod_decomp_results, "./Results2/cod_decomp_results.csv")
+write.csv(BlackWhite_results, "./Results2/BlackWhite_results.csv")
+write.csv(cod_change_results, "./Results2/cod_change_results.csv")
+write.csv(mortality.rates, "./Results2/mortality_rates_combined.csv")
+write.csv(mortality.rates.diff, "./Results2/mortality_rates_diff_combined.csv")
 ```

### 2017-06-06 Code/1_Explore_Data.Rmd

@@
 dat <- read.table("./Data/deaths-cause-6913.txt", header = F, sep="\t")
-names(dat) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
+#see ./Data/deaths-cause-6913.dic for the data dictionary.
+
 names(dat) <- c("state.n", "age.n", "sex.n", "race.n", "COD.n", "year.0", "crude.rate", "deaths", "population") 
-# names and order of variables in the dataset are noted in the .dic dictionary file found in the same dir as the dataset.
 ```
@@
 unique(dat$state[is.na(dat$stabbrs)==T]) #Washington DC not defined
-
 dat$stabbrs[dat$state == "Washington DC"] <- "DC"

### 2017-06-06 Contour-replacement/extract-for-contour-replacement.R
relative paths and deleting files

@@ -1,100 +1,0 @@
-
-source("~/BlackWhiteMortalityGap/Contour-replacement/scripts/decomp_contour.R")
-source("~/BlackWhiteMortalityGap/Contour-replacement/scripts/ex.R")
-
-load("~/BW_results/results_New York.Rdata")
-load("~/BW_results/r_All_New York.Rdata")
-
-r.All.1 <- r.All %>% filter(post.samp == 1)
-
-black <- r.All.1 %>% 
-  filter(race == "Black" & sex == "Male" & year %in% c(1969, 1983)) %>% 
-  group_by(year, age_minbin) %>%
-  summarise(total_deaths = sum(smoothed_deaths),
-            race = first(race), 
-            population = first(population)) %>%
-  mutate(mort.rate = total_deaths/population) %>%
-  select(age_minbin, year, mort.rate) %>%
-  tidyr::spread(year, mort.rate) %>%
-  rename(Age = age_minbin, mx1 = `1969`, mx2=`1983`)
-
-white <- r.All.1 %>% 
-  filter(race == "White" & sex == "Male" & year %in% c(1969, 1983)) %>% 
-  group_by(year, age_minbin) %>%
-  summarise(total_deaths = sum(smoothed_deaths),
-            race = first(race), 
-            population = first(population)) %>%
-  mutate(mort.rate = total_deaths/population) %>%
-  select(age_minbin, year, mort.rate) %>%
-  tidyr::spread(year, mort.rate) %>%
-  rename(Age = age_minbin, mx1 = `1969`, mx2 = `1983`)
-
-#this works
-results <- decomp.contour(black, white, 
-               ages = c(0, 15, 40, 65, 87),
-               FUN = ex.per, sex = "m")
-
-results <- decomp.contour(black, white, 
-               ages = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 87),
-               FUN = ex.per, sex = "m")
-
-#now get it to work using my life expectancy function rather than the Max Planck one.
-
-results <- decomp.contour(black, white, 
-                          ages = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 87),
-                          FUN = life.expectancy, )
-
-
-#decomp.contour function
-library(sqldf)
[…]

### 2017-06-06 Contour-replacement/scripts/sdmab.R

@@ -1,81 +1,0 @@
-# (c) Dmitri A. Jdanov and Vladimir M. Shkolnikov
-# jdanov@demogr.mpg.de
-# last revised 30.11.2014
-
-cdecomp.sdmab<-function(fname1, fname2, bo = 1, years = c(1955, 1970), agerange = c(12,39)) {
-  # SDMAB example
-  # input arguments: file names, birth order, years, and age range
-  # returns list with 2 components: $decomp - resuls of decomposition and $values - original values of SDMAB
-  # cdecomp.sdmab(fname1, fname2, bo = 1, years = c(1955, 1970), agerange = c(12,39))
-  # Example1: decomp.sdmab('CZEasfrVHbo.txt', 'RUSasfrVHbo.txt', 1) 
-  #                 calculates (contour) decomposition of the 1970 cohort difference in the SDMAB
-  #                 of first birth with split into initial condition (1955 cohort) and trend components
-  #                 for CZE and RUS using data from the original HFD files (cohort ASFR by birth order) 
-  # Example2: decomp.sdmab('CZEasfrVHbo.txt', 'RUSasfrVHbo.txt', bo = c(3:5), years = c(1950,1970)) 
-  #                 same as Example 1 but for cohorts 1970 and 1950 and birth order 3+
-  
-  # prepare the data
-  c1 = ASFR.select(fname1, years, bo, agerange)
-  c2 = ASFR.select(fname2, years, bo, agerange)
-
-  # decomposition
-  r = decomp.contour(c1, c2, ages=c(12,20,25,30,35,40), FUN = sdmab)
-
-  #calculate original SDMABs
-  v = matrix(, 2, 2)
-  
-  for (i in 1:2) {
-    v[1, i] = sdmab(c1[, i + 1], c1$Age)
-    v[2, i] = sdmab(c2[, i + 1], c2$Age)
-  }
-  v = data.frame(v)
-  rownames(v) = c(fname1, fname2)
-  colnames(v) = years
-  
-  # output
-  r = list(r, v)
-  names(r) = c("decomp", "values")
-  return(r)
-}
-
-sdmab<-function(fx, age) {
-  #calculates standard deviatian in the mean age at birth (see the HFD Methods protocol for details)
-  #input arguments: vectors of ASFR (fx) and ages (age, same dimension as fx)
-  #usage: sdmab(fx, age)
-  
-  age = age + 0.5
[…]

### 2017-06-06 Manuscript/Appendix-agestd-mort-rates.Rmd

@@
 library(forcats)
-source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+source("./Code/life_expectancy_functions.R")
 
-mortality.rates <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_combined.csv")
-mortality.rates.diff <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/mortality_rates_diff_combined.csv")
+mortality.rates <- read.csv("./Results/mortality_rates_combined.csv")
+mortality.rates.diff <- read.csv("./Results/mortality_rates_diff_combined.csv")
 

### 2017-06-06 Manuscript/Manuscript_COD_Trends.Rmd

@@
 ```{r load_data, echo=F}
-BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
+BlackWhite_results <- read.csv("./Results2/BlackWhite_results.csv")
 BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")
-cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
+cod_decomp_results <- read.csv("./Results2/cod_decomp_results.csv")
 cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
@@
 
-age_cod_results_infants_AOC <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_infants_AOC.csv")
+age_cod_results_infants_AOC <- read.csv("./Results2/age_cod_results_infants_AOC.csv")
 age_cod_results_infants_AOC$stabbrs <- state.abb[match(age_cod_results_infants_AOC$state, state.name)]

### 2017-06-06 Manuscript/Manuscript_COD_Trends_PNAS.Rmd

@@
 ```{r load_data, echo=F}
-BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
+BlackWhite_results <- read.csv("./Results2/BlackWhite_results.csv")
 BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")
-cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
+cod_decomp_results <- read.csv("./Results2/cod_decomp_results.csv")
 cod_decomp_results <- reorder.as.map2(cod_decomp_results, "state", "stabbrs")

### 2017-06-06 Manuscript/Manuscript_COD_decomp.Rmd

@@
 library(DT)
-source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+source("./Code/life_expectancy_functions.R")
 ```
@@
 ```{r read_data, echo=F}
-age_cod_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results.csv")
-age_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_results.csv")
-cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
-cod_change_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_change_results.csv")
-BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
+age_cod_results <- read.csv("./Results/age_cod_results.csv")
+age_decomp_results <- read.csv("./Results/age_decomp_results.csv")
+cod_decomp_results <- read.csv("./Results/cod_decomp_results.csv")
+cod_change_results <- read.csv("./Results/cod_change_results.csv")
+BlackWhite_results <- read.csv("./Results/BlackWhite_results.csv")
 
@@
 
-![Difference in life expectancy between black and white males, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Men.bmp)
+![Difference in life expectancy between black and white males, United States, 1969-2013.](./Manuscript/export.graph.Men.bmp)
 
@@
 
-![Difference in life expectancy between black and white females, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Women.bmp)
+![Difference in life expectancy between black and white females, United States, 1969-2013.](./Manuscript/export.graph.Women.bmp)
 

### 2017-06-06 conferences/Max_planck_submission.Rmd

@@
 library(gridExtra)
-source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
+source("./Code/life_expectancy_functions.R")
 ```
@@
 ```{r read_data, echo=F}
-age_cod_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results.csv")
-age_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_results.csv")
-cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
-BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
+age_cod_results <- read.csv("./Results/age_cod_results.csv")
+age_decomp_results <- read.csv("./Results/age_decomp_results.csv")
+cod_decomp_results <- read.csv("./Results/cod_decomp_results.csv")
+BlackWhite_results <- read.csv("./Results/BlackWhite_results.csv")
 

### 2017-06-06 conferences/Riddell_presentation_max_planck.Rmd

--- 2017-05-21 conferences/ResearchDay/Riddell_ResearchDay_minimal.Rmd
@@
 
-source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
-BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
+source("./Code/life_expectancy_functions.R")
+BlackWhite_results <- read.csv("./Results2/BlackWhite_results.csv")
 BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")
@@
 ```{r, echo=F}
-img1 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/k_morrison.png")), interpolate = FALSE)
-img2 <-  rasterGrob(as.raster(readPNG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/s_harper2.png")), interpolate = FALSE)
-img3 <-  rasterGrob(as.raster(readJPEG("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/j_kaufman.jpg")), interpolate = FALSE)
+img1 <-  rasterGrob(as.raster(readPNG("./conferences/images/k_morrison.png")), interpolate = FALSE)
+img2 <-  rasterGrob(as.raster(readPNG("./conferences/images/s_harper2.png")), interpolate = FALSE)
+img3 <-  rasterGrob(as.raster(readJPEG("./conferences/images/j_kaufman.jpg")), interpolate = FALSE)
 grid.arrange(img3,
@@
 ```{r, echo = F, out.width = 500}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/Harper_HA_title.png")
+include_graphics(path = "./conferences/images/Harper_HA_title.png")
 ```
@@
 ```{r, echo=F, out.width = 750}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/Harper_HA_fig1.png")
+include_graphics(path = "./conferences/images/Harper_HA_fig1.png")
 ```
@@
 ```{r, echo = F, out.width = 250}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/Harper_HA_quote.png")
+include_graphics(path = "./conferences/images/Harper_HA_quote.png")
 ```
@@
 ```{r, echo = F, out.width = 400}
-include_graphics(path = "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/conferences/ResearchDay/images/alabama.data.ex.png")
+include_graphics(path = "./conferences/images/alabama.data.ex.png")
 ```
@@
 + Compute national trend using random-effects meta regression
+
+## The output
+
++ Working paper being screened at bioRxiv
++ R Shiny App: https://corinne-riddell.shinyapps.io/black-white-life-expectancy/
+
+## The app
+
+```{r, echo = F, out.width = 800}
+include_graphics(path = "./conferences/images/app_ex1.gif")
+```
+
+## The app
+
+```{r, echo = F, out.width = 800}
+include_graphics(path = "./conferences/images/app_ex2.gif")
+```
+
+## The app
+
+```{r, echo = F, out.width = 800}
+include_graphics(path = "./conferences/images/app_ex3.gif")
+```
 
@@
 
+#Three examples
+
+##CVD in men
+```{r, echo = F, out.width = 675}
+include_graphics(path = "./conferences/images/CVD_men.png")
+```
+
+##Cancer in women
+```{r, echo = F, out.width = 675}
+include_graphics(path = "./conferences/images/Cancer_women.png")
+```
+
+##Injury in women
+```{r, echo = F, out.width = 675}
+include_graphics(path = "./conferences/images/Injury_women.png")
+```
+
 #A tale of two states
@@
 ```{r, echo = F, out.height = 600, out.width = 800}
-age_cod_results_female <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_female.csv")
+age_cod_results_female <- read.csv("./Results2/age_cod_results_female.csv")
 
-age_cod_results_male <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_male.csv")
+age_cod_results_male <- read.csv("./Results2/age_cod_results_male.csv")
 

### 2017-06-28 Code/archived/mortality rate plots.R
add to archibe

@@ -1,0 +1,41 @@
+#random plots exploring the mortality rate
+load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
+
+ggplot(subset(dat.clean.alabama, Race2=="Black"), aes(x = Count, y=crude_new)) + geom_point(aes(col=Population)) + facet_wrap(~Age2, scales = "free")
+
+ggplot(subset(dat.clean.alabama, Race2=="Black" & Sex2=="Male"), aes(x = Population, y=crude_new)) + 
+  geom_point(aes(col=COD2)) + facet_wrap(~Age2, scales = "free") +
+  geom_smooth(col="black") + scale_y_continuous("Mortality Rate") + ggtitle("Black Men in Alabama (all years)")
+
+ggplot(subset(dat.clean.alabama, Race2=="Black" & Sex2=="Male"), aes(x = Population, y=crude_new)) + 
+  geom_point(aes(col=Year)) + facet_wrap(~Age2, scales = "free") +
+  geom_smooth(col="black") + scale_y_continuous("Mortality Rate") + ggtitle("Black Men in Alabama (all years)")
+
+
+ggplot(subset(dat.clean.alabama, Race2=="White" & Sex2=="Male"), aes(x = Population, y=crude_new)) + 
+  geom_point(aes(col=COD2)) + facet_wrap(~Age2, scales = "free") +
+  geom_smooth(col="black") + scale_y_continuous("Mortality Rate") + ggtitle("White Men in Alabama (all years)")
+
+ggplot(subset(dat.clean.alabama, Race2=="White" & Sex2=="Male"), aes(x = Population, y=crude_new)) + 
+  geom_point(aes(col=Year)) + facet_wrap(~Age2, scales = "free") +
+  geom_smooth(col="black") + scale_y_continuous("Mortality Rate") + ggtitle("White Men in Alabama (all years)")
+
+
+
+mean.rate = c()
+counts1 = unique(dat.clean.alabama$Count) ; length(counts)
+counts = counts1[2:length(counts1)]
+
+for(i in counts) {
+  mean.rate = c(mean.rate, mean(dat.clean.alabama$crude_new[dat.clean.alabama$Count==i], na.rm=T))
+}
+
+plot(counts, mean.rate, pch=16, cex=0.5)
+
+mean(dat.clean.alabama$crude_new[dat.clean.alabama$Count==10], na.rm = T)
[…]

### 2017-06-28 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd
structured abstract

--- 2017-06-06 Manuscript/Manuscript_COD_Trends_PNAS.Rmd
@@
 ---
-title: Long-term trends in the contribution of major causes of death to the black-white life expectancy gap by US state
+title: Long-term trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013
 bibliography: bibliography.bib
@@
 
-States have fared differently in their progress towards eliminating the black-white life expectancy gap.
-Our objective is to describe the pattern of contributions of six major causes of deaths to the sex-specific black-white life expectancy gap across states over the last half-century, and identify divergent states.
-
-Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major causes of death.
-
 ```{r summarize_results_for_abstract, echo = F}
@@
 
-Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r formatC(as.numeric(cvd.female.2013), digits = 1)` years).
+Objectives: Our objective is to describe the pattern of contributions of six major causes of deaths to the sex-specific black-white life expectancy gap across states over the last half-century, and identify divergent states.
+
+Design: A repeated cross-sectional study of all deaths
+
+Setting: 39 US States and the District of Columbia
+
+Participants: Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major causes of death.
+
+Primary and Secondary outcome measures: We estimated the contribution (in years) that each of six major causes of death to the gap in life expectancy between black and white individuals.
+
+Results: Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r formatC(as.numeric(cvd.female.2013), digits = 1)` years).
 The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in 1988) and a stable contribution for women (approximately `r can.female.average` years).
@@
 
-Life expectancy for both races has improved substantially in the US.
+Conclusions: Life expectancy for both races has improved substantially in the US.
 For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality.

### 2017-06-29 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd
added reqd sections

@@
 
-###Abstract
+###Abstract (check if 300 words or fewer)
 
@@
 
-Objectives: Our objective is to describe the pattern of contributions of six major causes of deaths to the sex-specific black-white life expectancy gap across states over the last half-century, and identify divergent states.
+Objective: To describe the pattern of contributions of six major causes of deaths to the gender-specific black-white life expectancy gap across states over the last half-century, and identify divergent states.
 
-Design: A repeated cross-sectional study of all deaths
+Design: A retrospective, repeated cross-sectional study of all deaths.
 
-Setting: 39 US States and the District of Columbia
+Setting: 39 US States and the District of Columbia.
 
@@
 
-Primary and Secondary outcome measures: We estimated the contribution (in years) that each of six major causes of death to the gap in life expectancy between black and white individuals.
+Primary outcome measure: We estimated trends in the gender-specific contribution of six major causes of death to the gap in life expectancy between black and whites and their 95% credible bands.
 
@@
 
-###Significance
+###Strengths and limitations of this study
 
-Racial differences have been a fixture of American life for centuries, but their continued evolution is a topic of great relevance for public policy and social justice.
-States have fared differently in eliminating the black-white life expectancy gap.
-Using vital statistics and census data, we describe how six major causes of death contributed to the life expectancy gap between blacks and whites from 1969 to 2013.
-We find that states diverged in their cause-of-death specific contributions to the life expectancy gap in ways that either resulted in the near or total elimination of the gap, or alternatively its exacerbation.
-This analysis provides a foundation for future work to investigate state characteristics that may result in inequalities in mortality by race.
+- This study provides the foundation for future work in which the estimated trends for each state can be used to study the effects of state-level policy.
+
+OR
+
+- The estimated trends for each state are relatively stable compared to estimates from smaller units of analysis such as counties, and can be used to study the effects of state-level policy in subsequent work.
+
+- Examination of historical trends places the most recent differences in context, to disentangle whether the smallest state- and cause-specific contributions were historically small or underwent greater shifts over time compared to other states.
+
+- Due to data supression when death counts were small, we did not consider more specific causes of death such as homocide and drug overdose.
+
+- Migration between states and immigration to the US both affect state-level life expectancy and the black-white difference but were unmeasured.
+
+- We examined the concept of race (black or white), not ethnicity (i.e., Hispanic), since ethnicity was unmeasured in the data for the earliest years and both poorly and differentially measured by state for many years thereafter.
 
@@
 To identify states divergent from the national trend, we compared the state-specific trends estimated in the previous step to the national pattern estimated by setting the random effect to zero.
-
-**Data Sharing**
-
-A replication data set including the raw data and statistical code to reproduce the manuscript is publicly available on github: https://github.com/corinne-riddell/BlackWhiteMortalityGap/.
-
-###Acknowledgements
-
-We extend our thanks to Dr.
-Richard Maclehose, who developed the initial version of the code to perform the Bayesian analysis as part of the research article “Trends in the Black-White Life Expectancy Gap Among Us States, 1990–2009”, by Harper S, MacLehose RF, and Kaufman JS.
-
-Corinne Riddell is a post-doctoral researcher at McGill University, whose salary was supported by Dr.
-Kaufman's Canada Research Chair in Health Disparities.
-Kathryn Morrison is a post-doctoral researcher at McGill University.
-Sam Harper was supported by a Chercheur Boursier Junior 2 from the Fonds de la Recherche en Santé du Québec.
-Jay Kaufman was supported by a Canada Research Chair in Health Disparities.
 
@@
 Available from: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html.
+
+###Footnotes
+
+**Contributors**
+
+CAR: Extracted the data, performed data analysis, drafted the manuscript.
+
+KTM: Performed data analysis, reviewed the paper for intellectual content.
+
+SH: Initial conception for the study, interpreted the results, reviewed the paper for intellectual content.
+
+JK: Initial conception for the study, interpreted the results, drafted the manuscript, reviewed the paper for intellectual content.
+
+We extend our thanks to Dr.
+Richard Maclehose, who developed the initial version of the code to perform the Bayesian analysis as part of the research article “Trends in the Black-White Life Expectancy Gap Among Us States, 1990–2009”, by Harper S, MacLehose RF, and Kaufman JS.
+
+**Funding**
+
+This work was supported by Dr.
+Kaufman's Canada Research Chair in Health Disparities.
+Corinne Riddell is a post-doctoral researcher at McGill University, whose salary was supported by Dr.
+Kaufman's Canada Research Chair in Health Disparities.
+Kathryn Morrison is a post-doctoral researcher at McGill University.
+Sam Harper was supported by a Chercheur Boursier Junior 2 from the Fonds de la Recherche en Santé du Québec.
+Jay Kaufman was supported by a Canada Research Chair in Health Disparities.
+
+**Competing Interests**
+
+All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare: no support from any organisation for the submitted work; no financial relationships with any organisations that might have an interest in the submitted work in the previous three years; no other relationships or activities that could appear to have influenced the submitted work.”
+
+**Data Sharing Statement**
+
+A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on Github: https://github.com/corinne-riddell/BlackWhiteMortalityGap/ (DOI: 10.5281/zenodo.803472).

### 2017-06-29 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd

@@
 
+###Introduction
+
 Life expectancy is a barometer for overall population health.
@@
 The over-arching objective of this study is to describe the contribution of selected causes of death to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences.
-To do so, we used publicly available data from 1969 to 2013 detailing the number of deaths and population size according to US state, year, gender, race (black or white), age, and cause of death.
-We excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming.
-These states contained less than 0.4% of the black population in the United States in 2010.^5^ We did not consider Hispanic ethnicity as part of our racial definition, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.^6^
+To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for men and women.
+This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states.
+This study may therefore facilitate the identification of policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.
 
-Using these data, we estimated state-level trends in the black-white life expectancy gap measured in years separately for men and women.
-We then calculated the contribution of each of six major causes of death: cardiovascular disease, cancer, non-communicable disease, injury, communicable disease, and all other causes.
-This study lays the foundation for the identification of policies or social conditions in highly successful states, and the suggestion of broad targets for interventions to further reduce the gap in less successful states.
+###Methods
+
+**Description of the mortality data**
+
+We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in 39 US states and the District of Columbia between 1969 and 2013.^28^ We excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming.
+These states contained less than 0.4% of the black population in the United States in 2010.^5^ The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.^29^ The population data were derived by the NCHS using intercensal bridged-race population estimates.
+We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death.
+Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013).
+These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**Table S1**).
+We did not consider Hispanic ethnicity as part of our racial definition, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.^6^
+
+**Statistical challenges**
+
+The data structure produced two statistical challenges that were addressed by our modeling strategy.
+Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^30^ that we needed to impute.
+Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata.
+In these cases, the estimation of the mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic.
+To better estimate the underlying mortality rates, we used a method that produces smoother age-specific mortality trends.
+
+**Estimation of smoothed mortality rates using a Bayesian time-series model**
+
+We used a Bayesian time-series model to smooth the death counts across years and estimate the underlying mortality rate.
+Specifically, we used a Bayesian generalized linear model with a Poisson likelihood to model the number of deaths in each state-year-sex-race-age-cause stratum as a function of the at-risk population, for strata with zero or more than 9 deaths.
+The mortality rates over time within each stratum were assigned an autoregressive prior, where the log mortality rate was assumed to be Gaussian centered at the previous year's mortality rate.
+A non-informative gamma prior was used to estimate the inverse variance.
+For censored strata with between one and nine deaths, the deaths were modeled using a truncated Poisson likelihood, and otherwise the model remained the same.
+The truncated Poisson was bound between one and the larger of nine or the stratum's population size, which could be less than nine.
+
+Models were fit using Markov chain Monte Carlo (MCMC).
+These models were run separately for every combination of state, race, gender, and cause of death.
+They were fit using two chains of length 10,000, with a 2,000 iteration burn-in period to ensure convergence.
+One thousand iterations remained after burn-in and sampling every eighth iteration to reduce autocorrelation among the MCMC samples.
+Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition.
+These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.
+
+**Decomposition of the life expectancy gap by cause of death**
+
+To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied.^31,32^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender.
+The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.
+
+To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method.^33^ This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher cause-specific mortality rates among blacks.
+For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher rate of mortality due to cardiovascular disease among blacks compared to whites.
+If the decomposition produces a negative estimate this implies that whites have a higher rate of mortality for injuries compared to blacks.
+Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 40 state-level trends and their credible intervals.
+
+We also calculated state-level time trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results.
+We used the US standard population from the year 2000.^34^
+
+**Description of the trend in each cause of death's contribution to the gap and identification of divergent states**
+
+To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimated from the previous step as a function of a state random effect and a cubic spline for year (estimated using four internal knots).
+The model was inversely-weighted by the standard error of the mean estimate in each state-year.
+To identify states divergent from the national trend, we compared the state-specific trends estimated in the previous step to the national pattern estimated by setting the random effect to zero.
 
@@
 
-We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45 year period.
+We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45-year period.
 While we found that many states conformed to the national pattern of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation.
-In contrast to some recent work focused on county-level inequalities,^7,8^ states have some substantial advantages as analysis units; they have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.
-
-Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West, which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century.^9^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
+Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West, which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20th Century.^9^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
 On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^8^ On the other hand, greater racial equality can be achieved by falling black mortality rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^10^
@@
 Cancer is also responsible for large contributions to the life expectancy gap in some states (Illinnois, Michigan, and Wisconsin) but not in others (New York, Massachusetts).
-Overall, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women.However, both states began with only a minimal contribution of CVD and cancer to the mortality gap.
+Overall, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women.
+However, both states began with only a minimal contribution of CVD and cancer to the mortality gap.
 The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contributions of more than two years.
@@
 
-The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.
-States pursue varying approaches to an array of relevant programs and laws including tobacco control,^13^ Medicare coverage,^14^ hand-gun violence,^15^  and nutrition.^16^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^17^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
-Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.^18^
-
-*Strengths and weaknesses*
+**Strengths and weaknesses**
 
@@
 This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^19,20^ The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum.
-Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.
+In contrast to some recent work focused on county-level inequalities,^7,8^ states have some substantial advantages as analysis units; They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.
 
 Our analysis has three main limitations.
-Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state.
-This may be especially consequential for blacks in states like Florida or New York that received many black Hispanic immigrants from Cuba, Puerto Rico and the Dominican Republic, potentially changing the composition of the state’s black population to some extent during the study period.^21^  But it also has implications for the white population, especially in states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas.
-Given that Hispanics have higher life expectanies than non-Hispanics, including Hispanics in the life expectancy calculations will lead to estimates that are likely higher than they would have been had Hispanics been excluded.^22^ We did not to adjust for Hispanic ethnicity because reporting began only in 1978 and even then, agreement between self-reported ethnicity on the Census and secondary-reported ethinicity on the death certificate has been shown to be poorer during the earlier time period and differential according to geography, which could induce measurement error into mortality and disparity estimates.^6^ In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results,^4^ providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings.
-As the US population becomes increasingly Hispanic over time, however, this distinction grows more salient.^23^
-
-Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
-Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010.^24^ As we are examining the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white.
+Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years and our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results.^6^, 4 Had Hispanics been excluded from this study, the life expectancy estimates for both blacks and whites are expected to be lower, since Hispanics are known to have higher life expectancies than blacks and whites.^22^ Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
+US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.^24^ Thus, the black-white gap in a state may decrease as more immigrants identify as black.
 Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death.
-Migration is particularly important for the District of Columbia, because of it's small geographical bounds, and substantial in-migration from New York and in- and out-migration to Maryland and Virginia.
-Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
-Both changes would serve to increase the racial gap in DC.^25^
-
-Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
+Migration is particularly important for the District of Columbia, because of its small geographical bounds.
+Affluent black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
+Both changes would serve to increase the racial gap in DC.^25^ Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
 For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.
@@
 
-**Conclusions**
+**Implications for Policy & Future Work**
 
-Racial differences have been a fixture of American life for centuries, but their continued evolution is a topic of great relevance for public policy and social justice.
-The careful surveillance of these patterns is therefore a crucial public health activity, and this paper presents a unique perspective on state-level variation over nearly half a century.
-This analysis provides a foundation for future work to investigate time-varying and time-fixed state-level characteristics that may be responsible for differences among states in risk factors leading to each cause of death.
-Differences in smoking policy, Medicare coverage, access to care, segregation, and racism, for example, could all affect risk factors for ill health and vary in magnitude by state.
-
-###Methods
-
-**Description of the mortality data**
-
-We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.^28^ The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.^29^ The population data were derived by the NCHS using intercensal bridged-race population estimates.
-We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death.
-Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013).
-These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**Table S1**).
-
-**Statistical challenges**
-
-The data structure produced two statistical challenges that were addressed by our modeling strategy.
-Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^30^ that we needed to impute.
-Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata.
-In these cases, the estimation of the mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic.
-To better estimate the underlying mortality rates, we used a method that produces smoother age-specific mortality trends.
-
-**Estimation of smoothed mortality rates using a Bayesian time-series model**
-
-We used a Bayesian time-series model to smooth the death counts across years and estimate the underlying mortality rate.
-Specifically, we used a Bayesian generalized linear model with a Poisson likelihood to model the number of deaths in each state-year-sex-race-age-cause stratum as a function of the at-risk population, for strata with zero or more than 9 deaths.
-The mortality rates over time within each stratum were assigned an autoregressive prior, where the log mortality rate was assumed to be Gaussian centered at the previous year's mortality rate.
-A non-informative gamma prior was used to estimate the inverse variance.
-For censored strata with between one and nine deaths, the deaths were modeled using a truncated Poisson likelihood, and otherwise the model remained the same.
-The truncated Poisson was bound between one and the larger of nine or the stratum's population size, which could be less than nine.
-
-Models were fit using Markov chain Monte Carlo (MCMC).
-These models were run separately for every combination of state, race, gender, and cause of death.
-They were fit using two chains of length 10,000, with a 2,000 iteration burn-in period to ensure convergence.
-One thousand iterations remained after burn-in and sampling every eighth iteration to reduce autocorrelation among the MCMC samples.
-Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition.
-These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.
-
-**Decomposition of the life expectancy gap by cause of death**
-
-To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied.^31,32^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender.
-The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.
-
-To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method.^33^ This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher cause-specific mortality rates among blacks.
-For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher rate of mortality due to cardiovascular disease among blacks compared to whites.
-If the decomposition produces a negative estimate this implies that whites have a higher rate of mortality for injuries compared to blacks.
-Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 40 state-level trends and their credible intervals.
-
-We also calculated state-level time trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results.
-We used the US standard population from the year 2000.^34^
-
-**Description of the trend in each cause of death's contribution to the gap and identification of divergent states**
-
-To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimated from the previous step as a function of a state random effect and a linear spline for year (estimated using four internal knots).
-The model was inversely-weighted by the standard error of the mean estimate in each state-year.
-To identify states divergent from the national trend, we compared the state-specific trends estimated in the previous step to the national pattern estimated by setting the random effect to zero.
+The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.
+States pursue varying approaches to an array of relevant programs and laws including tobacco control,^13^ Medicare coverage,^14^ hand-gun violence,^15^ and nutrition.^16^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^17^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
+Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.
 
@@
 
+**needs to be reordered**
+
 1.

### 2017-06-29 Code/4_combined-meta-regression.Rmd
cubic splines because we live in a curvy world!

@@
   group_by(COD, sex) %>%
-  do(fit = lmer(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + (1|state),
+  do(fit = lmer(COD_cont_yrs_mean ~ bs(year, degree = 3, knots = c(1980, 1991, 2002)) + (1|state),
               weights = 1/(COD_cont_yrs_SE)^2, 
@@
   group_by(COD, sex) %>%
-  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state,
+  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 3, knots = c(1980, 1991, 2002)) + state,
               weights = 1/(COD_cont_yrs_SE)^2, 
@@
   group_by(sex) %>%
-  do(fit = lmer(age_COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + (1|state),
+  do(fit = lmer(age_COD_cont_yrs_mean ~ bs(year, degree = 3, df = 7) + (1|state),
               weights = 1/(age_COD_cont_yrs_SE)^2, 
@@
 COD.infants <- rep("All other causes", 80)
-RE.estimates.infants <- c(RE.models.infants$RE.estimates[[1]]$state$`(Intercept)`, RE.models.infants$RE.estimates[[2]]$state$`(Intercept)`)
+RE.estimates.infants <- c(RE.models.infants$RE.estimates[[1]]$state$`(Intercept)`, 
+                          RE.models.infants$RE.estimates[[2]]$state$`(Intercept)`)
 state.infants <- rep(levels(age_cod_results$state), 2)
@@
 ```{r save_data}
-write.csv(cod_decomp_results, "./Results2/cod_decomp_results.csv")
+write.csv(cod_decomp_results, "./Results2/cod_decomp_results_cubic.csv")
 
-write.csv(age_cod_results_infants_AOC, "./Results2/age_cod_results_infants_AOC.csv")
+write.csv(age_cod_results_infants_AOC, "./Results2/age_cod_results_infants_AOC_cubic.csv")
 ```

### 2017-06-29 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd
cubic splines because we live in a curvy world!

@@
 
-source("./Code/life_expectancy_functions.R")
+source("../Code/life_expectancy_functions.R")
 ```
@@
 ```{r load_data, echo=F}
-BlackWhite_results <- read.csv("./Results2/BlackWhite_results.csv")
+BlackWhite_results <- read.csv("../Results2/BlackWhite_results.csv")
 BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")
-cod_decomp_results <- read.csv("./Results2/cod_decomp_results.csv")
+cod_decomp_results <- read.csv("../Results2/cod_decomp_results_cubic.csv")
 cod_decomp_results <- reorder.as.map2(cod_decomp_results, "state", "stabbrs")
@@
 
-age_cod_results_infants_AOC <- read.csv("./Results2/age_cod_results_infants_AOC.csv")
+age_cod_results_infants_AOC <- read.csv("../Results2/age_cod_results_infants_AOC_cubic.csv")
 age_cod_results_infants_AOC$stabbrs <- state.abb[match(age_cod_results_infants_AOC$state, state.name)]
@@
 
-```{r plotting_resid_fx, echo = F}
-plot.resid.trends <- function(dataset, COD1, sex1) {
+```{r plotting_contributions_fx, echo = F}
+plot.contribution.trends <- function(dataset, COD1, sex1) {
   subset.data <- subset(dataset, COD == COD1 & sex == sex1)
@@
   for(sex_i in levels(cod_decomp_results$sex)) {
-    assign(paste0(COD_i, ".", sex_i, ".plot"), plot.resid.trends(dataset = cod_decomp_results, 
+    assign(paste0(COD_i, ".", sex_i, ".plot"), plot.contribution.trends(dataset = cod_decomp_results, 
                                                                          COD1 = COD_i, sex1 = sex_i))

### 2017-06-29 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd
changed ref order

@@
 
-We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in 39 US states and the District of Columbia between 1969 and 2013.^28^ We excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming.
-These states contained less than 0.4% of the black population in the United States in 2010.^5^ The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.^29^ The population data were derived by the NCHS using intercensal bridged-race population estimates.
+We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in 39 US states and the District of Columbia between 1969 and 2013.^5^ We excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming.
+These states contained less than 0.4% of the black population in the United States in 2010.^6^ The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.^7^ The population data were derived by the NCHS using intercensal bridged-race population estimates.
 We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death.
@@
 These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**Table S1**).
-We did not consider Hispanic ethnicity as part of our racial definition, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.^6^
+We did not consider Hispanic ethnicity as part of our racial definition, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.^8^
 
@@
 The data structure produced two statistical challenges that were addressed by our modeling strategy.
-Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^30^ that we needed to impute.
+Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^9^ that we needed to impute.
 Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata.
@@
 
-To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied.^31,32^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender.
+To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied.^10,11^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender.
 The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.
 
-To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method.^33^ This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher cause-specific mortality rates among blacks.
+To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method.^12^ This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher cause-specific mortality rates among blacks.
 For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher rate of mortality due to cardiovascular disease among blacks compared to whites.
@@
 We also calculated state-level time trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results.
-We used the US standard population from the year 2000.^34^
+We used the US standard population from the year 2000.^13^
 
@@
 While we found that many states conformed to the national pattern of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation.
-Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West, which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20th Century.^9^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
-On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^8^ On the other hand, greater racial equality can be achieved by falling black mortality rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^10^
+Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West, which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20th Century.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
+On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, greater racial equality can be achieved by falling black mortality rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^16^
 
-Consistent with what has been reported at the national level,^11^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
+Consistent with what has been reported at the national level,^17^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
 Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South.
-This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^12^ Communicable disease continues to contribute to the gap in most states, especially in DC, New Jersey, and Florida.
-In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 nationally,^11^ suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV.
+This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^18^ Communicable disease continues to contribute to the gap in most states, especially in DC, New Jersey, and Florida.
+In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 nationally,^17^ suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV.
 Cancer is also responsible for large contributions to the life expectancy gap in some states (Illinnois, Michigan, and Wisconsin) but not in others (New York, Massachusetts).
@@
 This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^19,20^ The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum.
-In contrast to some recent work focused on county-level inequalities,^7,8^ states have some substantial advantages as analysis units; They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.
+In contrast to some recent work focused on county-level inequalities,^21,15^ states have some substantial advantages as analysis units; They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.
 
 Our analysis has three main limitations.
-Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years and our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results.^6^, 4 Had Hispanics been excluded from this study, the life expectancy estimates for both blacks and whites are expected to be lower, since Hispanics are known to have higher life expectancies than blacks and whites.^22^ Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
-US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.^24^ Thus, the black-white gap in a state may decrease as more immigrants identify as black.
+Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years and our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results.^8^ Had Hispanics been excluded from this study, the life expectancy estimates for both blacks and whites are expected to be lower, since Hispanics are known to have higher life expectancies than blacks and whites.^22^ Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
+US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.^23^ Thus, the black-white gap in a state may decrease as more immigrants identify as black.
 Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death.
@@
 Affluent black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
-Both changes would serve to increase the racial gap in DC.^25^ Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
+Both changes would serve to increase the racial gap in DC.^24^ Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
 For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.
-Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^26,27^
+Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^25,26^
 
@@
 The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.
-States pursue varying approaches to an array of relevant programs and laws including tobacco control,^13^ Medicare coverage,^14^ hand-gun violence,^15^ and nutrition.^16^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^17^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
-Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.
+States pursue varying approaches to an array of relevant programs and laws including tobacco control,^27^ Medicare coverage,^28^ hand-gun violence,^29^ and nutrition.^30^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^31^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
+Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.^32^
 
@@
 
-5. U.S. Census Bureau: United States Census 2010 Interactive Population Map [Internet]. [cited 2017 May 16]. Available from: https://www.census.gov/2010census/popmap/.
+5. National Cancer Institute. Surveillance, Epidemiology, and End Results
+Program: SEER*Stat software: release: version 8.3.2 [Internet]. Bethesda (MD): NCI; 2017 Apr 14 [cited 2016 August 20]. Available from: http://www.seer.cancer.gov/seerstat.
 
-6.
+6. U.S. Census Bureau: United States Census 2010 Interactive Population Map [Internet]. [cited 2017 May 16]. Available from: https://www.census.gov/2010census/popmap/.
+
+7.
+Centers for Disease Control and Prevention.
+National Center for Health Statistics: National Vital Statistics System [Internet].
+[cited 2017 May 16].
+Available from: https://www.cdc.gov/nchs/nvss/index.htm.
+
+8.
 Arias E, Schauman WS, Eschbach K, Sorlie PD, Backlund E (2008) The validity of race and Hispanic origin reporting on death certificates in the United States.
@@
 
-7.
-Arora A, Spatz E, Herrin J, Riley C, Roy B, Kell K, Coberley C, Rula E, Krumholz HM (2016) Population Well-Being Measures Help Explain Geographic Disparities In Life Expectancy At The County Level.
-*Health Aff* 35(11):2075-82.
+9.
+National Cancer Institute.
+Surveillance, Epidemiology, and End Results Program: US mortality data [Internet].
+Bethesda (MD): NCI; [cited 2017 May 16].
+Available from: http://www.seer.cancer.gov/mortality/
 
-8.
+10. Preston SH, Heuveline P, and Guillot M (2001) Demography: Measuring and Modeling Population Processes. (Blackwell Publishing, Malden)
+
+11. Selvin, S (2004) Statistical Analysis of Epidemiologic Data. 3rd ed. (Oxford University Press, Oxford)
+
+12. Arriaga EE (1985) Measuring and explaining the change in life expectancies. *Demography* 21(1):83-96.
+
+13.
+National Cancer Institute.
+Surveillance, Epidemiology, and End Results Program: Standard Populations - 19 Age Groups [Internet].
+[cited 2017 April 25].
+Available from: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html.
+
+14. Lemann, N (1991) The Promised Land: The Great Black Migration and How It Changed America. New York: Alfred A. Knopf, ISBN 0-394-56004-3.
+
+15.
 Dwyer-Lindgren L, Bertozzi-Villa A, Stubbs RW, Morozoff C, Kutz MJ, Huynh C, Barber RM, et al (2016) US County-Level Trends in Mortality Rates for Major Causes of Death, 1980-2014.
@@
 
-9. Lemann, N (1991) The Promised Land: The Great Black Migration and How It Changed America. New York: Alfred A. Knopf, ISBN 0-394-56004-3.
+16. Case A, Deaton A (2017) Mortality and Morbidity in the 21st Century. In: Eberly J and Stock J, Editors. Brookings Papers on Economic Activity, March 23-24 2017.
 
-10. Case A, Deaton A (2017) Mortality and Morbidity in the 21st Century. In: Eberly J and Stock J, Editors. Brookings Papers on Economic Activity, March 23-24 2017.
+17. Harper S, Rushani D, and Kaufman JS (2012) Trends in the Black-White Life Expectancy Gap, 2003-2008. *JAMA* 307(21):2257–9.
 
-11. Harper S, Rushani D, and Kaufman JS (2012) Trends in the Black-White Life Expectancy Gap, 2003-2008. *JAMA* 307(21):2257–9.
-
-12. Rudd RA, Seth P, David F, Scholl L (2016) Increases in Drug and Opioid-Involved Overdose Deaths — United States, 2010–2015. *MMWR Morb Mortal Wkly Rep* 65:1445–1452.
-
-13.
-Cotti C, Nesson E, Tefft N (2016) The Effects of Tobacco Control Policies on Tobacco Products, Tar, and Nicotine Purchases among Adults: Evidence from Household Panel Data.
-*American Economic Journal: Economic Policy* 8(4):103-23.
-
-14. Grogan CM, Park SE (2017) The Racial Divide in State Medicaid Expansions. *J Health Polit Policy Law* 42(3):539-72.
-
-15.
-Humphreys DK, Gasparrini A, Wiebe DJ (2017) Evaluating the Impact of Florida’s “Stand Your Ground” Self-defense Law on Homicide and Suicide by Firearm: An Interrupted Time Series Study.
-*JAMA Intern Med* 177(1):44-50.
-
-16.
-Taber DR, Stevens J, Evenson KR, Ward DS, Poole C, Maciejewski ML, Murray DM, Brownson RC (2011) State policies targeting junk food in schools: racial/ethnic differences in the effect of policy change on soda consumption.
-*Am J Public Health* 101(9):1769-75.
-
-17. Adimora AA, Ramirez C, Schoenbach VJ, Cohen MS (2014) Policies and politics that promote HIV infection in the Southern United States. *AIDS*
-28(10):1393.
-
-18. Strumpf EC, Harper S, Kaufman JS (2017) Fixed Effects and Difference in Differences. In: Methods in Social Epidemiology. 2nd Ed, eds Oakes M, Kaufman JS(Jossey-Bass, San Francisco).
+18. Rudd RA, Seth P, David F, Scholl L (2016) Increases in Drug and Opioid-Involved Overdose Deaths — United States, 2010–2015. *MMWR Morb Mortal Wkly Rep* 65:1445–1452.
 
@@
 21.
-LaVeist-Ramos TA, Galarraga J, Thorpe RJ, Bell CN, Austin CJ (2011) Are black Hispanics black or Hispanic?
-Exploring disparities at the intersection of race and ethnicity.
-*J Epidemiol Community Health* Mar 3:jech-2009.
+Arora A, Spatz E, Herrin J, Riley C, Roy B, Kell K, Coberley C, Rula E, Krumholz HM (2016) Population Well-Being Measures Help Explain Geographic Disparities In Life Expectancy At The County Level.
+*Health Aff* 35(11):2075-82.
 
@@
 
-23. Brazil N (2017) Spatial Variation in the Hispanic Paradox: Mortality Rates in New and Established Hispanic US Destinations. *Population, Space and Place* 23:e1968.
+23. Preston SH, Elo IT (2014) Anatomy of a municipal triumph: New York City's upsurge in life expectancy. *Population and development review* 40(1):1-29.
 
-24. Preston SH, Elo IT (2014) Anatomy of a municipal triumph: New York City's upsurge in life expectancy. *Population and development review* 40(1):1-29.
-
-25.
+24.
 U.S.
@@
 
-26. Kuehn BM (2014) CDC: major disparities in opioid prescribing among states: some states crack down on excess prescribing. *JAMA* 312(7):684-6.
+25. Kuehn BM (2014) CDC: major disparities in opioid prescribing among states: some states crack down on excess prescribing. *JAMA* 312(7):684-6.
 
-27. Jones CM, Mack KA, Paulozzi LJ (2010) Pharmaceutical overdose deaths, United States, 2010. *JAMA* 309(7):657-9.
+26. Jones CM, Mack KA, Paulozzi LJ (2010) Pharmaceutical overdose deaths, United States, 2010. *JAMA* 309(7):657-9.
 
-28. National Cancer Institute. Surveillance, Epidemiology, and End Results
-Program: SEER*Stat software: release: version 8.3.2 [Internet]. Bethesda (MD): NCI; 2017 Apr 14 [cited 2016 August 20]. Available from: http://www.seer.cancer.gov/seerstat.
+27.
+Cotti C, Nesson E, Tefft N (2016) The Effects of Tobacco Control Policies on Tobacco Products, Tar, and Nicotine Purchases among Adults: Evidence from Household Panel Data.
+*American Economic Journal: Economic Policy* 8(4):103-23.
+
+28. Grogan CM, Park SE (2017) The Racial Divide in State Medicaid Expansions. *J Health Polit Policy Law* 42(3):539-72.
 
 29.
-Centers for Disease Control and Prevention.
-National Center for Health Statistics: National Vital Statistics System [Internet].
-[cited 2017 May 16].
-Available from: https://www.cdc.gov/nchs/nvss/index.htm.
+Humphreys DK, Gasparrini A, Wiebe DJ (2017) Evaluating the Impact of Florida’s “Stand Your Ground” Self-defense Law on Homicide and Suicide by Firearm: An Interrupted Time Series Study.
+*JAMA Intern Med* 177(1):44-50.
 
 30.
-National Cancer Institute.
-Surveillance, Epidemiology, and End Results Program: US mortality data [Internet].
-Bethesda (MD): NCI; [cited 2017 May 16].
-Available from: http://www.seer.cancer.gov/mortality/
+Taber DR, Stevens J, Evenson KR, Ward DS, Poole C, Maciejewski ML, Murray DM, Brownson RC (2011) State policies targeting junk food in schools: racial/ethnic differences in the effect of policy change on soda consumption.
+*Am J Public Health* 101(9):1769-75.
 
-31. Preston SH, Heuveline P, and Guillot M (2001) Demography: Measuring and Modeling Population Processes. (Blackwell Publishing, Malden)
+31. Adimora AA, Ramirez C, Schoenbach VJ, Cohen MS (2014) Policies and politics that promote HIV infection in the Southern United States. *AIDS*
+28(10):1393.
 
-32. Selvin, S (2004) Statistical Analysis of Epidemiologic Data. 3rd ed. (Oxford University Press, Oxford)
-
-33. Arriaga EE (1985) Measuring and explaining the change in life expectancies. *Demography* 21(1):83-96.
-
-34.
-National Cancer Institute.
-Surveillance, Epidemiology, and End Results Program: Standard Populations - 19 Age Groups [Internet].
-[cited 2017 April 25].
-Available from: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html.
+32. Strumpf EC, Harper S, Kaufman JS (2017) Fixed Effects and Difference in Differences. In: Methods in Social Epidemiology. 2nd Ed, eds Oakes M, Kaufman JS(Jossey-Bass, San Francisco).
 

### 2017-06-29 Manuscript/Appendix-agestd-mort-rates.Rmd
updated fig ref and appdx move

@@
 
-**Figure S1: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**
+**Figure S3: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**
 
@@
 
-**Figure S2: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**
+**Figure S4: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**
 
@@
 
-**Figure S3: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**
+**Figure S5: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**
 
@@
 
-**Figure S4: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**
+**Figure S6: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**
 
@@
 
-**Figure S5: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**
+**Figure S7: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**
 
@@
 
-**Figure S6: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**
+**Figure S8: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**
 
@@
 
-**Figure S7: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**
+**Figure S9: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**
 
@@
 
-**Figure S8: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**
+**Figure S10: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**
 
@@
 
-**Figure S9: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**
+**Figure S11: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**
 
@@
 
-**Figure S10: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**
+**Figure S12: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**
 
@@
 
-**Figure S11: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**
+**Figure S13: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**
 
@@
 
-**Figure S12: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**
+**Figure S14: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**
 

### 2017-06-29 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd
updated fig ref and appdx move

@@
 
-###Abstract (check if 300 words or fewer)
+###Abstract
 
@@
 These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.
+This modeling was done in R (version 3.3.2) using the *rjags* package.
 
@@
 
-We also calculated state-level time trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results.
-We used the US standard population from the year 2000.^13^
+We also calculated state-level trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results.
+We used the US standard population from the year 2000.^13^ The calculations described in this section were performed using functions written by the first author.
 
@@
 To identify states divergent from the national trend, we compared the state-specific trends estimated in the previous step to the national pattern estimated by setting the random effect to zero.
+These estimates relied on the R *lme4* package for random effects modeling, and the *splines* package to define cubic splines.
 
@@
 
-**Figures 1 and 2** display trends in the life expectancy gap for men and women, respectively.
+**Figure S1 and S2** display trends in the life expectancy gap for men and women, respectively.
 For men and women, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among men.
 
-**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for men and women, in years, for each state.
+**Figures 1 - 6** display trends in the contribution of each broad cause of death to the life expectancy gap for men and women, in years, for each state.
 The blue line depicts the mean of the smoothed contribution estimated for each state, such that states can have very different trend lines.
 These lines are overlaid with the meta-estimated national trend, shown with a red line.
-We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**Figures S1-S12**).
+We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**Figures S3-S14**).
 
@@
 Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races.
-Mortality from CVD decreased markedly since 1969 for all groups (**Figures S1 and S2**), but white men underwent relatively larger mortality decreases than black men.
-As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states (**Figure 3**).
-Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**).
+Mortality from CVD decreased markedly since 1969 for all groups (**Figures S3 and S4**), but white men underwent relatively larger mortality decreases than black men.
+As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states (**Figure 1A**).
+Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 1B**).
 For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.
@@
 
-For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Figures S3 and S4**).
-Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 5**).
-For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**).
+For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Figures S5 and S6**).
+Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 2A**).
+For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 2B**).
 For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.
@@
 
-Mortality from non-communicable disease increased over time for men in most states, with DC as a notable exception (**Figure S5**).
+Mortality from non-communicable disease increased over time for men in most states, with DC as a notable exception (**Figure S7**).
 Recently, there was a slight decrease in mortality rates, especially among black men.
-These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Figure S6**).
-As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**).
+These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Figure S8**).
+As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 3A and 3B**).
 As with other causes, New York and Massachusetts exhibited decreasing contributions, and Nevada also made substantial progress.
@@
 Trends in mortalilty from injuries differ for blacks and whites.
-Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S7**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S8**).
+Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S9**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S10**).
 Thus, injury mortality rates among blacks are equivalent, or lower than those among whites in many states, with white women demonstrating sharper recent increases.
-The contribution of injuries to the mortality gap has therefore generally decreased over time, and has become negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 9 and 10**).
+The contribution of injuries to the mortality gap has therefore generally decreased over time, and has become negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 4A and 4B**).
 Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions.
@@
 
-Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Figures S9 and S10**).
+Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Figures S11 and S12**).
 This spike was greater among men, especially blacks, and among states with larger urban centers.
-The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**).
+The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 5A and 5B**).
 DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.
@@
 
-Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**Figures S11 and S12**).
+Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**Figures S13 and S14**).
 White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states.
-As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 13 and 14**).
+As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 6A and 6B**).
 The most disparate states are Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.
@@
 Because the majority of infant deaths are classified as being due to all other causes in our analysis, we also tried to separate any impact of all other causes from those of infant mortality more generally.
-We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (Figures 15 and 16).
+We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (**Figures S15 and S16**).
 We found that 43% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969.
@@
 
-**Figure 1: State-level trends in the black-white life expectancy gap in men, United States, 1969-2013**
-
-```{r plot_LE_trends_males, echo=F, fig.height=10, fig.width=13}
-LE_trend_plot_males +
-  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)
-
-```
-
-**Figure 1 Legend:** The bands illustrate the 95% credible interval around the mean trend line.
-States with more imprecise estimates (often due to small black populations) have wider credible intervals.
-The estimated life expectancy gap in California in 1972 is much higher than in surrounding years.
-This anomaly is also illustrated in Figure 3, and is a product of an unexpectedly low CVD-related mortality rate among white males in California in 1972, likely denoting a coding error.
-Thus the data for this state and year should be interpreted with caution.
-
-**Figure 2: State-level trends in the black-white life expectancy gap in women, United States, 1969-2013**
-
-```{r plot_LE_trends_females, echo=F, fig.height=10, fig.width=13}
-LE_trend_plot_females +
-  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5)
-
-```
-
-**Figure 3: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
+**Figure 1A: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
 
@@
 
-**Figure 3 Legend:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of CVD to the black-white life expectancy gap over time.
+**Figure 1 Legend:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of CVD to the black-white life expectancy gap over time.
 The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
@@
 
-**Figure 4: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
+**Figure 1B: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
 
@@
 
-**Figure 5: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
+**Figure 2A: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
 
@@
 
-**Figure 6: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
+**Figure 2B: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
 
@@
 
-**Figure 7: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
+**Figure 3A: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
 
@@
 
-**Figure 8: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
+**Figure 3B: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
 
@@
 
-**Figure 9: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
+**Figure 4A: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
 
@@
 
-**Figure 10: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
+**Figure 4B: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
 
@@
 
-**Figure 11: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
+**Figure 5A: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
 
@@
 
-**Figure 12: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
+**Figure 5B: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
 
@@
 
-**Figure 13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
+**Figure 6A: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
 
@@
 
-**Figure 14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
+**Figure 6B: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
 
@@
   geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)
-```
-
-**Figure 15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men younger than one, 1969-2013**
-
-```{r age-cod-trend-infant-males, fig.width = 12, fig.height = 10, echo = F}
-ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Male"), aes(y = age_COD_cont_yrs_mean, x = year)) +
-  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
-  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
-  geom_line(col = "#08519c") +  
-  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 14) + 
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Contribution to the life expectancy gap (years)") +
-  xlab("Year (1969-2013)") +
-  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
-  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
-```
-
-**Figure 16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women younger than one, 1969-2013**
-
-```{r age-cod-trend-infant-females, fig.width = 12, fig.height = 10, echo = F}
-ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Female"), aes(y = age_COD_cont_yrs_mean, x = year)) +
-  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
-  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
-  geom_line(col = "#08519c") +  
-  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 14) + 
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Contribution to the life expectancy gap (years)") +
-  xlab("Year (1969-2013)") +
-  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
-  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
 ```
@@
 
-This work was supported by Dr.
-Kaufman's Canada Research Chair in Health Disparities.
+This research received no specific grant from any funding agency in the public, commercial or not-for-profit sectors.
+
 Corinne Riddell is a post-doctoral researcher at McGill University, whose salary was supported by Dr.
@@
 A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on Github: https://github.com/corinne-riddell/BlackWhiteMortalityGap/ (DOI: 10.5281/zenodo.803472).
+
+###Supplementary appendix
+
+**Figure S1: State-level trends in the black-white life expectancy gap in men, United States, 1969-2013**
+
+```{r plot_LE_trends_males, echo=F, fig.height=10, fig.width=13}
+LE_trend_plot_males +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)
+
+```
+
+**Figure S1 Legend:** The bands illustrate the 95% credible interval around the mean trend line.
+States with more imprecise estimates (often due to small black populations) have wider credible intervals.
+The estimated life expectancy gap in California in 1972 is much higher than in surrounding years.
+This anomaly is also illustrated in Figure 3, and is a product of an unexpectedly low CVD-related mortality rate among white males in California in 1972, likely denoting a coding error.
+Thus the data for this state and year should be interpreted with caution.
+
+**Figure S2: State-level trends in the black-white life expectancy gap in women, United States, 1969-2013**
+
+```{r plot_LE_trends_females, echo=F, fig.height=10, fig.width=13}
+LE_trend_plot_females +
+  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5)
+
+```
+
+**Figure S15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men younger than one, 1969-2013**
+
+```{r age-cod-trend-infant-males, fig.width = 12, fig.height = 10, echo = F}
+ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Male"), aes(y = age_COD_cont_yrs_mean, x = year)) +
+  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
+  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
+  geom_line(col = "#08519c") +  
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 14) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (years)") +
+  xlab("Year (1969-2013)") +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
+```
+
+**Figure S16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women younger than one, 1969-2013**
+
+```{r age-cod-trend-infant-females, fig.width = 12, fig.height = 10, echo = F}
+ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Female"), aes(y = age_COD_cont_yrs_mean, x = year)) +
+  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
+  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
+  geom_line(col = "#08519c") +  
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 14) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (years)") +
+  xlab("Year (1969-2013)") +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
+```

### 2017-07-04 Manuscript/Appendix-agestd-mort-rates.Rmd
updated manu and appendix

@@
 ---
-title: "Supplementary Appendix: Trends in age-standardized mortality rates for blacks and whites by cause and sex, 1969-2013, United States"
+title: "Supplementary Appendix: Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013"
 output: html_document
@@
 
-This is the supplementary appendix for the paper, "Historical trends in the contribution of six major causes of death to the black-white life expectancy gap by US state."
+This is the supplementary appendix for the paper, "Trendfs in the contribution of major causes of death to the black-white life expectancy gap by US state.
+A retrospective study of all deaths between 1969 and 2013."
 
@@
 library(forcats)
-source("./Code/life_expectancy_functions.R")
+source("../Code/life_expectancy_functions.R")
 
-mortality.rates <- read.csv("./Results/mortality_rates_combined.csv")
-mortality.rates.diff <- read.csv("./Results/mortality_rates_diff_combined.csv")
+BlackWhite_results <- read.csv("../Results2/BlackWhite_results.csv")
+BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")
+
+mortality.rates <- read.csv("../Results2/mortality_rates_combined.csv")
+mortality.rates.diff <- read.csv("../Results2/mortality_rates_diff_combined.csv")
 
@@
 mortality.rates.diff <- reorder.as.map2(mortality.rates.diff, "state", "stabbrs")
+
+age_cod_results_infants_AOC <- read.csv("../Results2/age_cod_results_infants_AOC_cubic.csv")
+age_cod_results_infants_AOC$stabbrs <- state.abb[match(age_cod_results_infants_AOC$state, state.name)]
+age_cod_results_infants_AOC$stabbrs[age_cod_results_infants_AOC$state == "Washington DC"] <- "DC"
+age_cod_results_infants_AOC <- reorder.as.map2(age_cod_results_infants_AOC, "state", "stabbrs")
 ```
@@
 }
+```
+
+**Table S1: Cause of death groupings**
+
+| Broad cause of death     | Included causes of death                                                     |
+|--------------------------|------------------------------------------------------------------------------|
+| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer.
+|
+| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries.
+|
+| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
+| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Syndrome and Nephrosis.
+|
+| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
+| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death.
+|
+
+**Figure S1: State-level trends in the black-white life expectancy gap in men, United States, 1969-2013**
+
+```{r life-expectancy-trends, echo=F}
+LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
+  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
+  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
+  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 14) +
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  labs(fill = "Census Division") +
+  ylab("Life expectancy gap (years)") +
+  xlab("Year (1969-2013)")
+
+LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
+  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
+  labs(fill = "Census Division") +
+  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
+  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 14) +
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Life expectancy gap (years)") +
+  xlab("Year (1969-2013)")
+```
+
+```{r plot_LE_trends_males, echo=F, fig.height=10, fig.width=13}
+LE_trend_plot_males +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)
+
+```
+
+**Figure S1 Legend:** The bands illustrate the 95% credible interval around the mean trend line.
+States with more imprecise estimates (often due to small black populations) have wider credible intervals.
+The estimated life expectancy gap in California in 1972 is much higher than in surrounding years.
+This anomaly is also illustrated in Figure 3, and is a product of an unexpectedly low CVD-related mortality rate among white males in California in 1972, likely denoting a coding error.
+Thus the data for this state and year should be interpreted with caution.
+
+**Figure S2: State-level trends in the black-white life expectancy gap in women, United States, 1969-2013**
+
+```{r plot_LE_trends_females, echo=F, fig.height=10, fig.width=13}
+LE_trend_plot_females +
+  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5)
+
+```
+
+```{r plot_both, eval=F, echo=F}
+#Informative: male and females at the same time
+ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
+  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
+  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
+  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 14) +
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Life expectancy gap (years)") +
+  xlab("Year (1969-2013)"))
 ```
@@
 ```
+
+**Figure S15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in males younger than one, 1969-2013**
+
+```{r age-cod-trend-infant-males, fig.width = 12, fig.height = 10, echo = F}
+ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Male"), aes(y = age_COD_cont_yrs_mean, x = year)) +
+  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
+  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
+  geom_line(col = "#08519c") +  
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 14) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (years)") +
+  xlab("Year (1969-2013)") +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
+```
+
+**Figure S16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in females younger than one, 1969-2013**
+
+```{r age-cod-trend-infant-females, fig.width = 12, fig.height = 10, echo = F}
+ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Female"), aes(y = age_COD_cont_yrs_mean, x = year)) +
+  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
+  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
+  geom_line(col = "#08519c") +  
+  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 14) + 
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Contribution to the life expectancy gap (years)") +
+  xlab("Year (1969-2013)") +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
+```

### 2017-07-04 Manuscript/StackedGraphs.Rmd
updated manu and appendix

---
title: "Alternative Visualizations for: Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013"
author: "Corinne Riddell"
date: "July 4, 2017"
output: html_document
---

These stacked bar charts illustrate an alternative visualization of the figures in "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013."

Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Sam Harper, PhD; Jay S Kaufman, PhD

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>

```{r stacked_trends_male, echo=F, fig.height=9, fig.width=11, warning=F, message=F}
library(dplyr)
library(purrr)
library(ggplot2)
library(stringr)
library(reshape2)
library(grid)
library(gridExtra)
library(plotly)
library(forcats)
source("../Code/life_expectancy_functions.R")

cod_decomp_results <- read.csv("../Results2/cod_decomp_results_cubic.csv")
cod_decomp_results <- reorder.as.map2(cod_decomp_results, "state", "stabbrs")
cod_decomp_results$COD_rev <- factor(cod_decomp_results$COD, levels = c("Cardiovascular", "Cancers", "All other causes",
                                                                        "Communicable", "Non-communicable",  "Injuries"))

BlackWhite_results <- read.csv("../Results2/BlackWhite_results.csv")
BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")
```

**Figure 1: Stacked contributions of major causes of death to the life expectancy gap among men by US state, 1969-2013**

```{r, echo=F, fig.height=9, fig.width=11}
plot1 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male"), aes(x = year, y = COD_cont_yrs_mean)) + 
  geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Male"), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (1969-2013)"))

for(i in 1:400){
  if(plot1$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot1$x$data[[i]]$line$width <- 1
  }
}

plot1 %>% layout(margin = list(b = 100))
```

**Figure 2: Stacked contributions of major causes of death to the life expectancy gap among men by US state, 2009-2013**

```{r, echo=F, fig.height=9, fig.width=11}
plot2 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
  geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Male" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (2009-2013)"))

for(i in 1:400){
  if(plot2$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot2$x$data[[i]]$line$width <- 1
  }
}

plot2 %>% layout(margin = list(b = 100))
```

**Figure 3: Stacked contributions of major causes of death to the life expectancy gap among women by US state, 1969-2013**

```{r stacked_trends_female, echo=F, fig.height=9, fig.width=11}
plot3 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female"), aes(x = year, y = COD_cont_yrs_mean)) + 
                      geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Female"), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (1969-2013)"))

for(i in 1:400){
  if(plot3$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot3$x$data[[i]]$line$width <- 1
  }
}

plot3 %>% layout(margin = list(b = 100))
```

**Figure 4: Stacked contributions of major causes of death to the life expectancy gap among women by US state, 2009-2013**

```{r, echo=F, fig.height=9, fig.width=11}
plot4 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
                      geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Female" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (2009-2013)"))

for(i in 1:400){
  if(plot4$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot4$x$data[[i]]$line$width <- 1
  }
}

plot4 %>% layout(margin = list(b = 100))
```

### 2017-07-04 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd

@@
 ---
-title: Long-term trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013
+title: Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013
 bibliography: bibliography.bib
@@
 ```{r load_data, echo=F}
-BlackWhite_results <- read.csv("../Results2/BlackWhite_results.csv")
-BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")
 cod_decomp_results <- read.csv("../Results2/cod_decomp_results_cubic.csv")
@@
 age_cod_results_infants_AOC <- read.csv("../Results2/age_cod_results_infants_AOC_cubic.csv")
-age_cod_results_infants_AOC$stabbrs <- state.abb[match(age_cod_results_infants_AOC$state, state.name)]
-age_cod_results_infants_AOC$stabbrs[age_cod_results_infants_AOC$state == "Washington DC"] <- "DC"
-age_cod_results_infants_AOC <- reorder.as.map2(age_cod_results_infants_AOC, "state", "stabbrs")
-
 ```
@@
 ```{r summarize_results_for_abstract, echo = F}
-cvd.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
-cvd.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
-cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
-cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) 
-cvd.female.2013 <- format(round(cvd.female.2013, 1), nsmall = 1)
+cvd.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>%
+  ungroup() %>% select(fixed.from.RE) %>% round(1)
+cvd.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
 
@@
   select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
-ncd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
-ncd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
+ncd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% ungroup() %>%  select(fixed.from.RE) %>% round(1)
+ncd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
 
 can.male.max <- cod_decomp_results %>% filter(sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
-  select(fixed.from.RE) %>% summarize(max = round(max(fixed.from.RE), 1))
+  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))
 can.female.average <- cod_decomp_results %>% filter(sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
@@
 
-inj.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
-inj.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
-inj.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
-inj.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
+inj.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+inj.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+inj.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+inj.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
 ```
 
-Objective: To describe the pattern of contributions of six major causes of deaths to the gender-specific black-white life expectancy gap across states over the last half-century, and identify divergent states.
+Objective: To describe the pattern of contributions of six major causes of deaths to the gender-specific black-white life expectancy gap across US states over the last half-century, and identify divergent states.
 
@@
 
-Results: Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r formatC(as.numeric(cvd.female.2013), digits = 1)` years).
-The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in 1988) and a stable contribution for women (approximately `r can.female.average` years).
+Results: Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years).
+The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in the early 1990s) and a stable contribution for women (approximately `r can.female.average` years).
 Both genders exhibited a decreased contribution from injuries (men: `r inj.male.1969` to `r inj.male.2013` years), that became negative for women (women: `r inj.female.1969` to `r inj.female.2013` years).
@@
 
-Conclusions: Life expectancy for both races has improved substantially in the US.
+Conclusions: Life expectancy for blacks and whites has improved substantially in the US.
 For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality.
@@
 
-- This study provides the foundation for future work in which the estimated trends for each state can be used to study the effects of state-level policy.
-
-OR
-
 - The estimated trends for each state are relatively stable compared to estimates from smaller units of analysis such as counties, and can be used to study the effects of state-level policy in subsequent work.
@@
 
+- The outcome of life expectancy integrates mortality rates across the entire age distribution making it sensitive to changes in mortality at any age.
+
+Need to remove one of these:
 - Due to data supression when death counts were small, we did not consider more specific causes of death such as homocide and drug overdose.
@@
 ###Results
-```{r life-expectancy-trends, echo=F}
-LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
-  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
-  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
-  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 14) +
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  labs(fill = "Census Division") +
-  ylab("Life expectancy gap (years)") +
-  xlab("Year (1969-2013)")
-
-LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
-  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
-  labs(fill = "Census Division") +
-  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
-  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 14) +
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Life expectancy gap (years)") +
-  xlab("Year (1969-2013)")
-```
-
-```{r plot_both, eval=F, echo=F}
-#Informative: male and females at the same time
-ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
-  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
-  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
-  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 14) +
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Life expectancy gap (years)") +
-  xlab("Year (1969-2013)"))
-```
 
@@
 
-**Figures 1 - 6** display trends in the contribution of each broad cause of death to the life expectancy gap for men and women, in years, for each state.
+**Figures 1 - 6** display trends in the contribution of each broad cause of death to the life expectancy gap for men (panel A) and women (panel B), in years, for each state.
 The blue line depicts the mean of the smoothed contribution estimated for each state, such that states can have very different trend lines.
@@
 Mortality from CVD decreased markedly since 1969 for all groups (**Figures S3 and S4**), but white men underwent relatively larger mortality decreases than black men.
-As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states (**Figure 1A**).
-Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 1B**).
+As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years (**Figure 1A**).
+Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years nationally (**Figure 1B**).
 For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.
@@
 For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Figures S5 and S6**).
-Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 2A**).
-For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 2B**).
+Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 2A**), with a peak contribution of `r can.male.max` in the early 1990s.
+For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a generally stable national pattern in the contribution of cancer to the life expectancy over time of approximately `r can.female.average` years (**Figure 2B**).
 For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.
@@
 These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Figure S8**).
-As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 3A and 3B**).
+As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men (approximately `r ncd.male.average` years), and exhibited a gradual decrease over time among women from `r ncd.female.1969` to `r ncd.female.2013` years (**Figures 3A and 3B**).
 As with other causes, New York and Massachusetts exhibited decreasing contributions, and Nevada also made substantial progress.
@@
 Thus, injury mortality rates among blacks are equivalent, or lower than those among whites in many states, with white women demonstrating sharper recent increases.
-The contribution of injuries to the mortality gap has therefore generally decreased over time, and has become negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 4A and 4B**).
+The contribution of injuries to the mortality gap has therefore decreased over time from `r inj.male.1969` to `r inj.male.2013` years among men.
+Among women, the contribution went from `r inj.female.1969` in 1969 to `r inj.female.2013` in 2013, becoming negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 4A and 4B**).
 Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions.
@@
 
+```{r communicable_dx_stats, echo=F}
+com.male.1969 <- cod_decomp_results %>% filter(year == 1969, sex == "Male" & COD == "Communicable" & state == "Alabama") %>% 
+  ungroup() %>% select(fixed.from.RE) %>% round(1)
+
+com.female.1969 <- cod_decomp_results %>% 
+  filter(year == 1969, sex == "Female" & COD == "Communicable" & state == "Alabama") %>% 
+  ungroup() %>% select(fixed.from.RE) %>% round(1)
+
+com.male.mid.max <- cod_decomp_results %>% filter(year > 1980, sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
+  ungroup() %>%
+  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))
+
+com.female.mid.max <- cod_decomp_results %>% filter(year > 1980, sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
+  ungroup() %>%
+  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))
+
+```
+
 Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Figures S11 and S12**).
 This spike was greater among men, especially blacks, and among states with larger urban centers.
-The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 5A and 5B**).
+The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the early 1990s on `r com.male.mid.max` years in men and `r com.female.mid.max` years in women, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 5A and 5B**).
 DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.
@@
 ```{r, echo = F}
-male.endpoints.aoc <- cod_decomp_results %>% filter(year %in% c(1969, 2013), sex =="Male", COD == "All other causes", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )
 
-female.endpoints.aoc <- cod_decomp_results %>% filter(year %in% c(1969, 2013), sex =="Female", COD == "All other causes", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )
+aoc.male.endpoints <- cod_decomp_results %>% 
+  filter(year %in% c(1969, 2013), sex == "Male" & COD == "All other causes" & state == "Alabama") %>% 
+  ungroup() %>% select(fixed.from.RE) %>% round(1)
 
-male.infants.endpoints.aoc <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Male", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )
+aoc.female.endpoints <- cod_decomp_results %>% 
+  filter(year %in% c(1969, 2013), sex == "Female" & COD == "All other causes" & state == "Alabama") %>% 
+  ungroup() %>% select(fixed.from.RE) %>% round(1)
 
-female.infants.endpoints.aoc <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Female", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )
+aoc.male.endpoints.infant <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Male", state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+
+aoc.female.endpoints.infant <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Female", state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+
+proportions.males <- aoc.male.endpoints.infant/aoc.male.endpoints
+proportions.females <- aoc.female.endpoints.infant/aoc.female.endpoints
 ```
@@
 White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states.
-As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 6A and 6B**).
+As such, the contribution of all other causes to the life expectancy gap has generally decreased over time (from `r sprintf("%.1f", aoc.male.endpoints[1, 1])` to `r aoc.male.endpoints[2, 1]` years among males and `r aoc.female.endpoints[1, 1]` to `r aoc.female.endpoints[2, 1]` years among females), and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 6A and 6B**).
 The most disparate states are Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.
@@
 We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (**Figures S15 and S16**).
-We found that 43% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969.
-In 2013, this proportion reached 53%.
+We found that `r proportions.males[1,1]*100`% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969 in males and `r round(proportions.females[1,1]*100, 0)`% in females.
+In 2013, this proportion reached `r proportions.males[2,1]*100`% in males and `r proportions.females[2,1]*100`% in females.
 Thus, the trend in the all other causes contribution partially reflects a decreasing contribution from infant mortality.
@@
 While we found that many states conformed to the national pattern of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation.
-Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West, which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20th Century.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
+Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the Midwest, which had been the receiving point of blacks from the South during “the Great Migration” for improved economic opportunities between 1916 and 1970.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
 On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, greater racial equality can be achieved by falling black mortality rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^16^
@@
 
-###Appendix
-**Table S1: Cause of death groupings**
-
-| Broad cause of death     | Included causes of death                                                     |
-|--------------------------|------------------------------------------------------------------------------|
-| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer.
-|
-| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries.
-|
-| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
-| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Syndrome and Nephrosis.
-|
-| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
-| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death.
-|
-
-**Stacked charts - not to be included in the submission**
-
-**Men**
-
-```{r stacked_trends_male, echo=F, fig.height=9, fig.width=11}
-plot1 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male"), aes(x = year, y = COD_cont_yrs_mean)) + 
-  geom_area(aes(fill = COD_rev), stat = "identity") +
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
-  geom_line(data = subset(BlackWhite_results, sex == "Male"), aes(x = year, y = LE_wbgap_mean)) +
-  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
-  theme_classic(base_size = 14) + 
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Contribution to the life expectancy gap (%)") +
-  xlab("Year (1969-2013)"))
-
-for(i in 1:400){
-  if(plot1$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
-    plot1$x$data[[i]]$line$width <- 1
-  }
-}
-
-plot1
-
-plot2 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
-  geom_area(aes(fill = COD_rev), stat = "identity") +
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
-  geom_line(data = subset(BlackWhite_results, sex == "Male" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
-  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
-  theme_classic(base_size = 14) + 
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Contribution to the life expectancy gap (%)") +
-  xlab("Year (2009-2013)"))
-
-for(i in 1:400){
-  if(plot2$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
-    plot2$x$data[[i]]$line$width <- 1
-  }
-}
-
-plot2
-```
-
-**Women**
-
-```{r stacked_trends_female, echo=F, fig.height=9, fig.width=11}
-plot3 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female"), aes(x = year, y = COD_cont_yrs_mean)) + 
-                      geom_area(aes(fill = COD_rev), stat = "identity") +
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
-  geom_line(data = subset(BlackWhite_results, sex == "Female"), aes(x = year, y = LE_wbgap_mean)) +
-  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
-  theme_classic(base_size = 14) + 
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Contribution to the life expectancy gap (%)") +
-  xlab("Year (1969-2013)"))
-
-for(i in 1:400){
-  if(plot3$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
-    plot3$x$data[[i]]$line$width <- 1
-  }
-}
-
-plot3
-
-plot4 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
-                      geom_area(aes(fill = COD_rev), stat = "identity") +
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
-  geom_line(data = subset(BlackWhite_results, sex == "Female" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
-  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
-  theme_classic(base_size = 14) + 
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Contribution to the life expectancy gap (%)") +
-  xlab("Year (2009-2013)"))
-
-for(i in 1:400){
-  if(plot4$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
-    plot4$x$data[[i]]$line$width <- 1
-  }
-}
-
-plot4
-```
-
 ###Bibliography
@@
 A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on Github: https://github.com/corinne-riddell/BlackWhiteMortalityGap/ (DOI: 10.5281/zenodo.803472).
-
-###Supplementary appendix
-
-**Figure S1: State-level trends in the black-white life expectancy gap in men, United States, 1969-2013**
-
-```{r plot_LE_trends_males, echo=F, fig.height=10, fig.width=13}
-LE_trend_plot_males +
-  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)
-
-```
-
-**Figure S1 Legend:** The bands illustrate the 95% credible interval around the mean trend line.
-States with more imprecise estimates (often due to small black populations) have wider credible intervals.
-The estimated life expectancy gap in California in 1972 is much higher than in surrounding years.
-This anomaly is also illustrated in Figure 3, and is a product of an unexpectedly low CVD-related mortality rate among white males in California in 1972, likely denoting a coding error.
-Thus the data for this state and year should be interpreted with caution.
-
-**Figure S2: State-level trends in the black-white life expectancy gap in women, United States, 1969-2013**
-
-```{r plot_LE_trends_females, echo=F, fig.height=10, fig.width=13}
-LE_trend_plot_females +
-  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5)
-
-```
-
-**Figure S15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men younger than one, 1969-2013**
-
-```{r age-cod-trend-infant-males, fig.width = 12, fig.height = 10, echo = F}
-ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Male"), aes(y = age_COD_cont_yrs_mean, x = year)) +
-  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
-  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
-  geom_line(col = "#08519c") +  
-  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 14) + 
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Contribution to the life expectancy gap (years)") +
-  xlab("Year (1969-2013)") +
-  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
-  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
-```
-
-**Figure S16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women younger than one, 1969-2013**
-
-```{r age-cod-trend-infant-females, fig.width = 12, fig.height = 10, echo = F}
-ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Female"), aes(y = age_COD_cont_yrs_mean, x = year)) +
-  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
-  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
-  geom_line(col = "#08519c") +  
-  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 14) + 
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Contribution to the life expectancy gap (years)") +
-  xlab("Year (1969-2013)") +
-  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
-  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
-```

### 2017-07-04 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd

@@
 
-Objective: To describe the pattern of contributions of six major causes of deaths to the gender-specific black-white life expectancy gap across US states over the last half-century, and identify divergent states.
+**Objective:** To describe the pattern of contributions of six major causes of deaths to the gender-specific black-white life expectancy gap across US states over the last half-century, and identify divergent states.
 
-Design: A retrospective, repeated cross-sectional study of all deaths.
+**Design:** A retrospective, repeated cross-sectional study of all deaths.
 
-Setting: 39 US States and the District of Columbia.
+**Setting:** 39 US States and the District of Columbia.
 
-Participants: Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major causes of death.
+**Participants:** Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major causes of death.
 
-Primary outcome measure: We estimated trends in the gender-specific contribution of six major causes of death to the gap in life expectancy between black and whites and their 95% credible bands.
+**Primary outcome measure:** We estimated trends in the gender-specific contribution of six major causes of death to the gap in life expectancy between black and whites and their 95% credible bands.
 
-Results: Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years).
+**Results:** Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years).
 The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in the early 1990s) and a stable contribution for women (approximately `r can.female.average` years).
@@
 
-Conclusions: Life expectancy for blacks and whites has improved substantially in the US.
+**Conclusions:** Life expectancy for blacks and whites has improved substantially in the US.
 For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality.
@@
 Need to remove one of these:
-- Due to data supression when death counts were small, we did not consider more specific causes of death such as homocide and drug overdose.
+
+- Due to data suppression when death counts were small, we did not consider more specific causes of death such as homicide and drug overdose.
 
@@
            geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
-           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)
+           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
+           scale_y_continuous(breaks = seq(-0.5, 1.5, by = 0.5), labels = seq(-0.5, 1.5, by = 0.5))
 
@@
 `Non-communicable.Female.plot`[[3]] +
+           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
            geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
            geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
-           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)
+           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
+           scale_y_continuous(breaks = seq(-1, 1.5, by = 0.5), labels = seq(-1, 1.5, by = 0.5))
 
@@
 `All other causes.Female.plot`[[3]] +
-  geom_hline(aes(yintercept = -2), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
   geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
   geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
   geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) 

### 2017-08-03 Manuscript/Manuscript_COD_Trends_HealthandPlace.Rmd

--- 2017-07-04 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd
@@
 ---
-title: Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013
+title: Trends in the contribution of major causes of death to the black-white life expectancy gap by US state
 bibliography: bibliography.bib
@@
 
-**Objective:** To describe the pattern of contributions of six major causes of deaths to the gender-specific black-white life expectancy gap across US states over the last half-century, and identify divergent states.
+**Objective:** To describe the pattern of contributions of six major causes of deaths to the black-white life expectancy gap across US states over the last half-century, and identify divergent states.
 
@@
 
-**Participants:** Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major causes of death.
+**Participants:** Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major cause-of-death groups.
 
@@
 
-**Results:** Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years).
-The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in the early 1990s) and a stable contribution for women (approximately `r can.female.average` years).
+**Results:** Although mortality from cardiovascular disease has decreased dramatically, its contribution to the black-white life expectancy gap increased over time for men (from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years).
+The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) and decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in the early 1990s) and a stable contribution for women (approximately `r can.female.average` years).
 Both genders exhibited a decreased contribution from injuries (men: `r inj.male.1969` to `r inj.male.2013` years), that became negative for women (women: `r inj.female.1969` to `r inj.female.2013` years).
@@
 
-**Conclusions:** Life expectancy for blacks and whites has improved substantially in the US.
+**Conclusions:** The black-white life expectancy gap has decreased substantially in the US since 1969.
 For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality.
-In women, a crossover in injury-related mortality led to a narrower gap, realized partially by increasing mortality among whites.
-
-----
-
-###Strengths and limitations of this study
-
-- The estimated trends for each state are relatively stable compared to estimates from smaller units of analysis such as counties, and can be used to study the effects of state-level policy in subsequent work.
-
-- Examination of historical trends places the most recent differences in context, to disentangle whether the smallest state- and cause-specific contributions were historically small or underwent greater shifts over time compared to other states.
-
-- The outcome of life expectancy integrates mortality rates across the entire age distribution making it sensitive to changes in mortality at any age.
-
-Need to remove one of these:
-
-- Due to data suppression when death counts were small, we did not consider more specific causes of death such as homicide and drug overdose.
-
-- Migration between states and immigration to the US both affect state-level life expectancy and the black-white difference but were unmeasured.
-
-- We examined the concept of race (black or white), not ethnicity (i.e., Hispanic), since ethnicity was unmeasured in the data for the earliest years and both poorly and differentially measured by state for many years thereafter.
+In women, a crossover in injury-related mortality led to a narrower gap, due partially to increasing mortality among whites.
 
@@
 Life expectancy is a barometer for overall population health.
-In 2014, life expectancy at birth in the United States was 78.8 years, but hidden within this summary are important subgroups differences by self-defined racial ethnicity.
-Whites lived 3.4 years longer than blacks.^1^ While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.^2^
+In 2014, life expectancy at birth in the United States was 78.8 years, but hidden within this summary are important subgroups differences.
+Whites lived 3.4 years longer than blacks.^1^ While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied over time.^2^
 
-During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV/AIDS, unintentional injuries, and, among women, due to differential improvements in heart disease.^3^ The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement.
+During the 1980s the life expectancy gap widened, but then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths from homicide, HIV/AIDS, unintentional injuries, and, among women, from differential improvements in heart disease.^3^ The gap continued to narrow during the 2000s, but US states differed markedly in the magnitude of improvement.
 For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years.^4^ However, the extent to which different causes of death may have contributed to state-level changes in the black-white gap are unknown.
 
-The over-arching objective of this study is to describe the contribution of selected causes of death to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences.
+The overarching objective of this study is to describe the contribution of selected causes of death to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences.
 To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for men and women.
@@
 The data structure produced two statistical challenges that were addressed by our modeling strategy.
-Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^9^ that we needed to impute.
+Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^9^ which we needed to impute.
 Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata.
@@
 
-To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied.^10,11^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender.
+To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total number of deaths in each age group, and standard life table methods were applied.^10,11^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender.
 The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.
@@
 
-To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimated from the previous step as a function of a state random effect and a cubic spline for year (estimated using four internal knots).
+To describe the national trend in each cause's contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimated from the previous step as a function of a state random effect and a cubic spline for year (estimated using four internal knots).
 The model was inversely-weighted by the standard error of the mean estimate in each state-year.
@@
 These estimates relied on the R *lme4* package for random effects modeling, and the *splines* package to define cubic splines.
+
+A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on Github: link [DOI: removed for double blind review]
 
@@
 As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years (**Figure 1A**).
-Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years nationally (**Figure 1B**).
+Black women experienced relatively larger CVD mortality decreases over time compared to white women, resulting in a decrease of CVD to the contribution to the life expectancy gap from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years nationally (**Figure 1B**).
 For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.
-
-**Cancer**
-
-For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Figures S5 and S6**).
-Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 2A**), with a peak contribution of `r can.male.max` in the early 1990s.
-For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a generally stable national pattern in the contribution of cancer to the life expectancy over time of approximately `r can.female.average` years (**Figure 2B**).
-For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.
-
-**Non-communicable disease**
-
-Mortality from non-communicable disease increased over time for men in most states, with DC as a notable exception (**Figure S7**).
-Recently, there was a slight decrease in mortality rates, especially among black men.
-These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Figure S8**).
-As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men (approximately `r ncd.male.average` years), and exhibited a gradual decrease over time among women from `r ncd.female.1969` to `r ncd.female.2013` years (**Figures 3A and 3B**).
-As with other causes, New York and Massachusetts exhibited decreasing contributions, and Nevada also made substantial progress.
-Furthermore, while DC's mortality rates decreased, they presently have the highest contribution of non-communicable disease to the gap.
-
-**Injury**
-
-Trends in mortalilty from injuries differ for blacks and whites.
-Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S9**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S10**).
-Thus, injury mortality rates among blacks are equivalent, or lower than those among whites in many states, with white women demonstrating sharper recent increases.
-The contribution of injuries to the mortality gap has therefore decreased over time from `r inj.male.1969` to `r inj.male.2013` years among men.
-Among women, the contribution went from `r inj.female.1969` in 1969 to `r inj.female.2013` in 2013, becoming negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 4A and 4B**).
-Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions.
-Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.
-
-**Communicable disease**
-
-```{r communicable_dx_stats, echo=F}
-com.male.1969 <- cod_decomp_results %>% filter(year == 1969, sex == "Male" & COD == "Communicable" & state == "Alabama") %>% 
-  ungroup() %>% select(fixed.from.RE) %>% round(1)
-
-com.female.1969 <- cod_decomp_results %>% 
-  filter(year == 1969, sex == "Female" & COD == "Communicable" & state == "Alabama") %>% 
-  ungroup() %>% select(fixed.from.RE) %>% round(1)
-
-com.male.mid.max <- cod_decomp_results %>% filter(year > 1980, sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
-  ungroup() %>%
-  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))
-
-com.female.mid.max <- cod_decomp_results %>% filter(year > 1980, sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
-  ungroup() %>%
-  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))
-
-```
-
-Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Figures S11 and S12**).
-This spike was greater among men, especially blacks, and among states with larger urban centers.
-The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the early 1990s on `r com.male.mid.max` years in men and `r com.female.mid.max` years in women, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 5A and 5B**).
-DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.
-
-**All other causes**
-
-```{r, echo = F}
-
-aoc.male.endpoints <- cod_decomp_results %>% 
-  filter(year %in% c(1969, 2013), sex == "Male" & COD == "All other causes" & state == "Alabama") %>% 
-  ungroup() %>% select(fixed.from.RE) %>% round(1)
-
-aoc.female.endpoints <- cod_decomp_results %>% 
-  filter(year %in% c(1969, 2013), sex == "Female" & COD == "All other causes" & state == "Alabama") %>% 
-  ungroup() %>% select(fixed.from.RE) %>% round(1)
-
-aoc.male.endpoints.infant <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Male", state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
-
-aoc.female.endpoints.infant <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Female", state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
-
-proportions.males <- aoc.male.endpoints.infant/aoc.male.endpoints
-proportions.females <- aoc.female.endpoints.infant/aoc.female.endpoints
-```
-
-Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**Figures S13 and S14**).
-White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states.
-As such, the contribution of all other causes to the life expectancy gap has generally decreased over time (from `r sprintf("%.1f", aoc.male.endpoints[1, 1])` to `r aoc.male.endpoints[2, 1]` years among males and `r aoc.female.endpoints[1, 1]` to `r aoc.female.endpoints[2, 1]` years among females), and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 6A and 6B**).
-The most disparate states are Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.
-
-Because the majority of infant deaths are classified as being due to all other causes in our analysis, we also tried to separate any impact of all other causes from those of infant mortality more generally.
-We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (**Figures S15 and S16**).
-We found that `r proportions.males[1,1]*100`% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969 in males and `r round(proportions.females[1,1]*100, 0)`% in females.
-In 2013, this proportion reached `r proportions.males[2,1]*100`% in males and `r proportions.females[2,1]*100`% in females.
-Thus, the trend in the all other causes contribution partially reflects a decreasing contribution from infant mortality.
-
-###Discussion
-
-**Summary of findings**
-
-We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45-year period.
-While we found that many states conformed to the national pattern of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation.
-Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the Midwest, which had been the receiving point of blacks from the South during “the Great Migration” for improved economic opportunities between 1916 and 1970.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
-On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, greater racial equality can be achieved by falling black mortality rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^16^
-
-Consistent with what has been reported at the national level,^17^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
-Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South.
-This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^18^ Communicable disease continues to contribute to the gap in most states, especially in DC, New Jersey, and Florida.
-In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 nationally,^17^ suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV.
-Cancer is also responsible for large contributions to the life expectancy gap in some states (Illinnois, Michigan, and Wisconsin) but not in others (New York, Massachusetts).
-Overall, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women.
-However, both states began with only a minimal contribution of CVD and cancer to the mortality gap.
-The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contributions of more than two years.
-Thus, the reason for New York’s and Massachusetts’ low CVD contributions does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.
-
-**Strengths and weaknesses**
-
-We used data over 45 years, across two racial groups, and for all ages.
-By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year.
-This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time.
-Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
-This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^19,20^ The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum.
-In contrast to some recent work focused on county-level inequalities,^21,15^ states have some substantial advantages as analysis units; They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.
-
-Our analysis has three main limitations.
-Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years and our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results.^8^ Had Hispanics been excluded from this study, the life expectancy estimates for both blacks and whites are expected to be lower, since Hispanics are known to have higher life expectancies than blacks and whites.^22^ Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
-US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.^23^ Thus, the black-white gap in a state may decrease as more immigrants identify as black.
-Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death.
-Migration is particularly important for the District of Columbia, because of its small geographical bounds.
-Affluent black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
-Both changes would serve to increase the racial gap in DC.^24^ Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
-For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.
-Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^25,26^
-
-**Implications for Policy & Future Work**
-
-The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.
-States pursue varying approaches to an array of relevant programs and laws including tobacco control,^27^ Medicare coverage,^28^ hand-gun violence,^29^ and nutrition.^30^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^31^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
-Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.^32^
 
@@
 
+**Cancer**
+
+For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Figures S5 and S6**).
+Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 2A**), with a peak contribution of `r can.male.max` in the early 1990s.
+For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a generally stable national pattern in the contribution of cancer to the life expectancy over time of approximately `r can.female.average` years (**Figure 2B**).
+For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.
+
 **Figure 2A: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
@@
 ```
+
+**Non-communicable disease**
+
+Mortality from non-communicable disease increased over time for men in most states, with DC as a notable exception (**Figure S7**).
+Recently, there was a slight decrease in mortality rates, especially among black men.
+These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Figure S8**).
+As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men (approximately `r ncd.male.average` years), and exhibited a gradual decrease over time among women from `r ncd.female.1969` to `r ncd.female.2013` years (**Figures 3A and 3B**).
+As with other causes, New York and Massachusetts exhibited decreasing contributions, and Nevada also made substantial progress.
+Furthermore, while DC's mortality rates decreased, they presently have the highest contribution of non-communicable disease to the gap.
 
@@
 ```
+
+**Injury**
+
+Trends in mortalilty from injuries differ for blacks and whites.
+Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S9**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S10**).
+Thus, injury mortality rates among blacks are equivalent to, or lower than those among whites in many states, with white women demonstrating sharper recent increases.
+The contribution of injuries to the mortality gap has therefore decreased substantially over time from `r inj.male.1969` to `r inj.male.2013` years among men.
+Among women, the contribution went from `r inj.female.1969` in 1969 to `r inj.female.2013` in 2013, becoming negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 4A and 4B**).
+Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions.
+Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.
 
@@
 
+**Communicable disease**
+
+```{r communicable_dx_stats, echo=F}
+com.male.1969 <- cod_decomp_results %>% filter(year == 1969, sex == "Male" & COD == "Communicable" & state == "Alabama") %>% 
+  ungroup() %>% select(fixed.from.RE) %>% round(1)
+
+com.female.1969 <- cod_decomp_results %>% 
+  filter(year == 1969, sex == "Female" & COD == "Communicable" & state == "Alabama") %>% 
+  ungroup() %>% select(fixed.from.RE) %>% round(1)
+
+com.male.mid.max <- cod_decomp_results %>% filter(year > 1980, sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
+  ungroup() %>%
+  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))
+
+com.female.mid.max <- cod_decomp_results %>% filter(year > 1980, sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
+  ungroup() %>%
+  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))
+
+```
+
+Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Figures S11 and S12**).
+This spike was greater among men, especially blacks, and among states with larger urban centers.
+The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the early 1990s on `r com.male.mid.max` years in men and `r com.female.mid.max` years in women, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 5A and 5B**).
+DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.
+
 **Figure 5A: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
@@
 ```
+**All other causes**
+
+```{r, echo = F}
+
+aoc.male.endpoints <- cod_decomp_results %>% 
+  filter(year %in% c(1969, 2013), sex == "Male" & COD == "All other causes" & state == "Alabama") %>% 
+  ungroup() %>% select(fixed.from.RE) %>% round(1)
+
+aoc.female.endpoints <- cod_decomp_results %>% 
+  filter(year %in% c(1969, 2013), sex == "Female" & COD == "All other causes" & state == "Alabama") %>% 
+  ungroup() %>% select(fixed.from.RE) %>% round(1)
+
+aoc.male.endpoints.infant <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Male", state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+
+aoc.female.endpoints.infant <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Female", state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+
+proportions.males <- aoc.male.endpoints.infant/aoc.male.endpoints
+proportions.females <- aoc.female.endpoints.infant/aoc.female.endpoints
+```
+
+Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**Figures S13 and S14**).
+White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states.
+As such, the contribution of all other causes to the life expectancy gap has generally decreased over time (from `r sprintf("%.1f", aoc.male.endpoints[1, 1])` to `r aoc.male.endpoints[2, 1]` years among males and `r aoc.female.endpoints[1, 1]` to `r aoc.female.endpoints[2, 1]` years among females), and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 6A and 6B**).
+The most disparate states are Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.
+
+Because the majority of infant deaths are classified as being due to all other causes in our analysis, we also tried to separate any impact of all other causes from those of infant mortality more generally.
+We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (**Figures S15 and S16**).
+We found that `r proportions.males[1,1]*100`% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969 in males and `r round(proportions.females[1,1]*100, 0)`% in females.
+In 2013, this proportion reached `r proportions.males[2,1]*100`% in males and `r proportions.females[2,1]*100`% in females.
+Thus, the trend in the all other causes contribution partially reflects a decreasing contribution from infant mortality.
 
@@
 
+###Discussion
+
+**Summary of findings**
+
+We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45-year period.
+While we found that many states conformed to a general pattern, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation.
+Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the Midwest, which had been the receiving point of blacks from the South during “the Great Migration” for improved economic opportunities between 1916 and 1970.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
+On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, because greater racial equality can be achieved by either falling black mortality rates or rising white rates, the reduced life expectancy gap cannot be celebrated without a careful examination of the two individual trends.^16^
+
+Consistent with what has been reported at the national level,^17^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
+Injuries continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South.
+This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^18^ Communicable disease continues to contribute to the gap in most states, especially in DC, New Jersey, and Florida.
+In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 nationally,^17^ suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV.
+Cancer is also responsible for large contributions to the life expectancy gap in some states (Illinnois, Michigan, and Wisconsin) but not in others (New York, Massachusetts).
+Overall, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women.
+However, both states began with only a minimal contribution of CVD and cancer to the mortality gap.
+The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contributions of more than two years.
+Thus, the reason for New York’s and Massachusetts’ low CVD contributions does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.
+
+**Strengths and weaknesses**
+
+We used data over 45 years, across two racial groups, and for all ages.
+By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year.
+This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time.
+Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
+This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^19,20^ The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum.
+In contrast to some recent work focused on county-level inequalities,^21,15^ states have some substantial advantages as analysis units; They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities with divergent social environments that are reflected in varying patterns of mortality by race.
+
+Our analysis has three main limitations.
+Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years.
+Our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results for the overall life expectancy gap.^8^ Had Hispanics been excluded from this study, the life expectancy estimates for both blacks and whites are expected to be lower, since Hispanics are known to have lower mortality than blacks and whites.^22^ Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
+US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.^23^ Thus, the black-white gap in a state may decrease as more immigrants identify as black.
+Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death.
+Migration is particularly important for the District of Columbia, because of its small geographical bounds.
+Affluent black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
+Both changes would serve to increase the racial gap in DC.^24^ Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
+For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.
+Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^25,26^
+
+**Implications for Policy & Future Work**
+
+The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.
+States pursue varying approaches to an array of relevant programs and laws including tobacco control,^27^ Medicare coverage,^28^ hand-gun violence,^29^ and nutrition.^30^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^31^ The analyses reported here are entirely descriptive, but could be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
+Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.^32^
+
 ###Bibliography
-
-**needs to be reordered**
 

### 2017-09-08 Manuscript/Abstract-DataVizDemo.Rmd
abstract added

---
title: Interactive decompositions of the life expectancy gap
output:
  html_document: default
  word_document: default
always_allow_html: yes
---

Corinne A Riddell, PhD

Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 27, Montreal, QC H3A 1A2, Canada.

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 27, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>.

```{r, echo=F, message=F, warning=F}
library(shiny) #, lib.loc = "/Library/Frameworks/R.framework/Versions/3.3/Resources/library/old_versions") #OLD version: 0.14.1, NEW: 1.0.3
library(ggplot2)#, lib.loc = "/Library/Frameworks/R.framework/Versions/3.3/Resources/library/old_versions") #OLD: 2.2.1, NEW: ggplot2_2.2.1.9000 (install dev version using devtools::install_github)
library(plotly)#, lib.loc = "/Library/Frameworks/R.framework/Versions/3.3/Resources/library/old_versions") #OLD: 4.5.6, NEW: 4.6.0
library(crosstalk)#, lib.loc = "/Library/Frameworks/R.framework/Versions/3.3/Resources/library/old_versions") #1.0.0
library(viridis)
library(scales)
library(shinythemes)
library(dplyr)
library(grid)
library(gridExtra)
library(png)
library(forcats)

source("../Code/life_expectancy_functions.R")
```

```{r, echo=F, message=F, warning=F}
age_decomp_results <- read.csv("../Results2/age_decomp_results.csv")

age_decomp_results$age <- forcats::fct_relevel(f = age_decomp_results$age, "<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                            "15-19 years", "20-24 years", "25-29 years", "30-34 years", "35-39 years", 
                                            "40-44 years", "45-49 years", "50-54 years", "55-59 years", "60-64 years", 
                                            "65-69 years", "70-74 years", "75-79 years", "80-84 years", "85+ years")

cod_decomp_results <- read.csv("../Results2/cod_decomp_results.csv")
cod_change_results <- read.csv("../Results2/cod_change_results.csv")
BlackWhite_results <- read.csv("../Results2/BlackWhite_results.csv")
dat.aggregated <- read.csv("../Data/dat_aggregated.csv")

excluded <- c("Alaska", "Hawaii", "Idaho", "Maine", "Montana", "New Hampshire", "North Dakota",
              "South Dakota", "Utah", "Vermont", "Wyoming")

dat.agg2.white <- dat.aggregated %>% filter(age == "<1 year" & race == "White" & !(state %in% excluded)) %>% select(state, year, sex, pop_across_age) %>% rename(pop_white = pop_across_age)
dat.agg2.black <- dat.aggregated %>% filter(age == "<1 year" & race == "Black" & !(state %in% excluded)) %>% select(state, year, sex, pop_across_age) %>% rename(pop_black = pop_across_age)
dat.agg2 <- merge(dat.agg2.black, dat.agg2.white, by = c("state", "year", "sex"))
rm(dat.agg2.black, dat.agg2.white)

cod_decomp_results$COD <- factor(cod_decomp_results$COD, levels(cod_decomp_results$COD)[c(3, 2, 4, 6, 5, 1)])

cod_marginal_results <- cod_decomp_results %>% 
  group_by(stratum.id) %>% 
  summarise(state = first(state), sex= first(sex), year = first(year), total_Cx = sum(COD_cont_yrs_mean))

cod_decomp_results <- merge(cod_decomp_results, BlackWhite_results %>% select(stratum.id, LE_white_mean, LE_black_mean, LE_wbgap_mean), 
      by = "stratum.id")
cod_decomp_results <- merge(cod_decomp_results, dat.agg2, by = c("state", "year", "sex"))

cod_decomp_results$new.start = cod_decomp_results$start + cod_decomp_results$LE_black_mean
cod_decomp_results$new.start2 = cod_decomp_results$start2 + cod_decomp_results$LE_black_mean
cod_decomp_results$new.finish = cod_decomp_results$finish + cod_decomp_results$LE_black_mean
cod_decomp_results$new.finish2 = cod_decomp_results$finish2 +cod_decomp_results$ LE_black_mean

cod_decomp_results$LE_black_mean[cod_decomp_results$COD != "Cardiovascular"] <- NA
cod_decomp_results$LE_white_mean[cod_decomp_results$COD != "Cardiovascular"] <- NA

cod_decomp_results <- reorder.as.map2(cod_decomp_results, "state", "stabbrs")

age_decomp_results <- merge(age_decomp_results, BlackWhite_results %>% select(stratum.id, LE_white_mean, LE_black_mean, LE_wbgap_mean), 
                            by = "stratum.id")

age_decomp_results <- merge(age_decomp_results, dat.agg2, by = c("state", "year", "sex"))

age_decomp_results$new.start = age_decomp_results$start + age_decomp_results$LE_black_mean
age_decomp_results$new.start2 = age_decomp_results$start2 + age_decomp_results$LE_black_mean
age_decomp_results$new.finish = age_decomp_results$finish + age_decomp_results$LE_black_mean
age_decomp_results$new.finish2 = age_decomp_results$finish2 + age_decomp_results$ LE_black_mean

age_decomp_results$LE_black_mean[age_decomp_results$age != "<1 year"] <- NA
age_decomp_results$LE_white_mean[age_decomp_results$age != "<1 year"] <- NA

BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")
```

###Abstract

 **Motivation**

Methods to decompose the difference in life expectancy between populations are well-developed. In particular, the life expectancy gap can be partitioned by age groups or cause of death. This allows researchers to highlight where there is an excess of deaths according to age or cause in a disadvantaged population vs. a comparison population. Most often, results from such studies are displayed in lengthy tables which are difficult for the reader to process and interpret. Sometimes, the results are displayed using static graphics in paper journals. The motivation for the visualizations presented in this article was to use web technology to allow the interactive display of data in order to enhance what can be learned from each display. Specifically, the objectives were to:

1. utilize hover text to display more information than is reasonably shown on static displays, 
2. layer graphical objects that enrich the visualization as the user toggles the layers on and off, and
3. use transparency of graphical features as a proxy for statistical precision.

 **Data**

These graphics were developed as part of a larger project that studies trends in the difference in life expectancy between black and white Americans. Mortality and population data was abstracted using the US National Cancer Institute's software, SEER-Stat, stratified by state, year, sex, race (black or white), age group, and according to six cause of death groupings. As described elsewhere, we imputed supressed death counts and smoothed sparse data using a Bayesian time series design. For this paper, we utilize only the 2013 data for females.

**Data visualizations**

Figure 1 depicts the difference in life expectancy between white females (vertical red line) and black females (dashed red line). Ages to the left of the dashed line narrow the gap whereas ages to the right exacerbate it. The primary benefit of the interactive display is the ability to toggle the contribution of each age group to the gap. For example, turning off the contribution for infants (<1 year old) allows the use to immediately visualize that, relative to other age groups, infant mortality still contributed substantially to the black-white gap in life expectancy. Toggling the subsequent age groups reveal relatively small contributions during childhood and adolescence. The secondary benefit of the interactive display is the hover text that lists the estimate of the contribution numerically. These are the same numbers that would usually be shown only in a table format, but hover text allows for both the graphical and exact numeric display of information to be contained in one interactive illustration.

**Figure 1: Contribution of age to the life expectancy gap between black and white females in 2013 by US state**

```{r figure_age, fig.width=7, fig.height=9, echo=F, message=F, warning=F}
data.subset <- data.frame(subset(age_decomp_results, sex == "Female" & year == 2013 & pop_black >= 1)) 
data.subset["state.reorder2"] <- reorder(data.subset$state, data.subset$LE_black_mean, max, na.rm = T)
data.subset["state.reorder2.n"] <- as.numeric(data.subset[["state.reorder2"]])

ly <- ggplotly(ggplot(data.subset, 
                      aes(x = new.start, y = state.reorder2.n)) + 
                 geom_segment(aes(x = LE_black_mean, xend = LE_white_mean, 
                                  y = state.reorder2.n + 0.46, yend = state.reorder2.n + 0.46), col = "red") + 
                 geom_rect(aes(xmin = new.start, 
                               ymin = state.reorder2.n - 0.44, 
                               ymax = state.reorder2.n + 0.44, 
                               xmax = new.finish, fill = age)) +
                 scale_y_continuous(breaks = 1:length(levels(factor(data.subset$state.reorder2))), 
                                    labels = levels(factor(data.subset$state.reorder2))) +
                 theme_minimal() + 
                 theme(legend.title = element_blank(),
                       panel.background = element_rect(fill = "transparent", colour = NA), 
                       plot.background = element_rect(fill = "transparent", colour = NA),
                       legend.background = element_rect(fill = "transparent", colour = NA)) +
                 geom_segment(aes(x = LE_white_mean, xend = LE_white_mean, y = state.reorder2.n - 0.46, yend = state.reorder2.n + 0.46), col = "red") +
                 geom_segment(aes(x = LE_black_mean, xend = LE_black_mean, y = state.reorder2.n - 0.46, yend = state.reorder2.n + 0.46), col = "red", lty = 3) +
                 scale_fill_viridis(discrete = T, direction = -1) 
)

    for(i in 1:length(ly$x$data)){
      if (ly$x$data[[i]]$line$color == "rgba(255,0,0,1)") {
        ly$x$data[[i]]$line$width <- 2
        if(ly$x$data[[i]]$line$dash == "dot"){
          ly$x$data[[i]]$text <- "Black Life Expectancy"  
        }
        if(ly$x$data[[i]]$line$dash == "solid"){
          ly$x$data[[i]]$text <- "White Life Expectancy"  
        }
        }
    }

    ly %>% layout(xaxis = list(title = " ", side = "top", xpad = 20), yaxis = list(title = NA, autorange = "reversed"))
```

Figure 2 partitions the black-white life expectancy gap according to six major cause of death groupings. This figure shares the advantages of the previous one in terms of interactivity. With the exception of Nevada and Rhode Island, injuries is the only cause of death that contributes to narrowing the life expectancy gap (to the left of the dashed line). This figure communicates that finding very clearly, and makes it easy to identify states for which the contribution of injury to narrowing the gap is most prominent. In particular, the southern states consistently show the largest contribution of injury towards narrowing the gap, which is likely due to increased overdose mortality among white women linked to the ongoing opoiod epidemic.

**Figure 2: Contribution of cause of death to the life expectancy gap between black and white females in 2013 by US state**

```{r figure_cod, fig.width=7, fig.height=9, echo=F, message=F, warning=F}
temp <- data.frame(subset(cod_decomp_results, sex == "Female" & 
                            year == 2013 & pop_black >= 1))
temp["state.reorder2"] <- reorder(temp$state, temp$LE_black_mean, max, na.rm = T)
temp["state.reorder2.n"] <- as.numeric(temp[["state.reorder2"]])

interactive.plot <- ggplotly(ggplot(temp, 
                                    aes(x = new.start, y = state.reorder2.n)) + 
                               geom_segment(aes(x = LE_black_mean, xend = LE_white_mean, 
                                                y = state.reorder2.n + 0.46, yend = state.reorder2.n + 0.46)) +

                               geom_rect(aes(xmin = new.start, 
                                             ymin = state.reorder2.n - 0.44, 
                                             ymax = state.reorder2.n + 0.44, 
                                             xmax = new.finish, fill = COD)) +
                               scale_y_continuous(breaks = 1:length(levels(factor(temp$state.reorder2))), 
                                                  labels = levels(factor(temp$state.reorder2))) +
                               theme_minimal() + xlab(" ") + 
                               theme(legend.title = element_blank(),
                                     panel.background = element_rect(fill = "transparent", colour = NA), 
                                     plot.background = element_rect(fill = "transparent", colour = NA),
                                     legend.background = element_rect(fill = "transparent", colour = NA)) +
                               geom_segment(aes(x = LE_white_mean, xend = LE_white_mean, y = state.reorder2.n - 0.46, yend = state.reorder2.n + 0.46)) +
                               geom_segment(aes(x = LE_black_mean, xend = LE_black_mean, 
                                                y = state.reorder2.n - 0.46, yend = state.reorder2.n + 0.46), lty = 3)
)

for(i in 1:length(interactive.plot$x$data)){
  if (interactive.plot$x$data[[i]]$line$color == "rgba(0,0,0,1)") {
    interactive.plot$x$data[[i]]$line$width <- 2
    if(interactive.plot$x$data[[i]]$line$dash == "dot"){
      interactive.plot$x$data[[i]]$text <- "Black Life Expectancy"  
    }
    if(interactive.plot$x$data[[i]]$line$dash == "solid"){
      interactive.plot$x$data[[i]]$text <- "White Life Expectancy"  
    }
  }
}

interactive.plot %>% 
  layout(xaxis = list(title = " ", side = "top"), yaxis = list(title = NA, autorange = "reversed"))

```

 **Applications and future extensions**

### 2018-02-25 Code/6-Excluded-states-description.Rmd

---
title: "Excluded states description"
author: "Corinne Riddell"
date: "2/21/2018"
output: html_document
---

```{r}
library(tidyverse)
library(ggrepel)
```

```{r}
dat.aggregated <- readr::read_csv("./Data/dat_aggregated.csv", col_names = T)

dat2 <- dat.aggregated %>% 
  group_by(state, year) %>% 
  filter(age == "<1 year") %>% 
  select(state, year, race, sex, race.sex, pop_across_age)

dat3 <- dat2 %>% 
  group_by(state, year, race) %>% 
  summarise(pop.total.across.race = sum(pop_across_age))

dat3 <- dat2 %>% 
  group_by(state, year, race) %>% 
  summarise(pop.total.across.race = sum(pop_across_age),
            prop.race = round((pop.total.across.race/pop.total)*100, 1) )

dat4 <- dat3 %>% filter(state %in% c("Alaska", "Hawaii", "Idaho", "Maine", "Montana", "New Hampshire", "North Dakota", "South Dakota", "Utah", "Vermont", "Wyoming"))

```

```{r}

ggplot(dat3 %>% filter(race == "Black"), aes(year, prop.race)) + 
  geom_line(aes(col = state, group = state)) + 
  ylab("% of population that is black") + 
  geom_text_repel(data = dat4 %>% filter(race == "Black", year == 2013), aes(group = state, label = state), size = 3) + 
  ggtitle("Proportion of blacks in the population")

ggplot(dat3 %>% filter(race == "Black"), aes(year, pop.total.across.race)) + 
  geom_line(aes(col = state, group = state)) + 
  ylab("% of population that is black") + 
  geom_text_repel(data = dat4 %>% filter(race == "Black", year == 2013), aes(group = state, label = state), size = 3) + 
  ggtitle("Absolute size of black population") + guides(col = F) + 
  scale_y_log10(breaks = c(10000, 100000, 500000, 1000000, 3000000), 
                      labels = c("10k", "100k", "500k", "1m", "3m")) +
  # scale_y_continuous(breaks = c(10000, 100000, 500000, 1000000, 3000000), 
  #                    labels = c("10k", "100k", "500k", "1m", "3m")) +
  theme(panel.grid.minor = element_blank())

```

### 2018-03-21 Code/4_combined-meta-regression.Rmd
added summary plots

@@
 
-cod_decomp_results <- cod_decomp_results %>% rename(fitted.RE = .fitted, fixed.from.RE = .fixed, resid.RE = .resid)
+cod_decomp_results <- cod_decomp_results %>% rename(fitted.RE = .fitted, 
+                                                    fixed.from.RE = .fixed, 
+                                                    resid.RE = .resid)
+
+cod_decomp_results <- cod_decomp_results %>% 
+  mutate(national.from.RE = fitted.RE - RE.estimates, #the national trend
+         residual.RE = COD_cont_yrs_mean - national.from.RE) #the difference between the observed state trends and the national trend
 ```

### 2018-03-21 Manuscript/Appendix-agestd-mort-rates.Rmd

@@
 
-This is the supplementary appendix for the paper, "Trendfs in the contribution of major causes of death to the black-white life expectancy gap by US state.
+This is the supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state.
 A retrospective study of all deaths between 1969 and 2013."
@@
 age_cod_results_infants_AOC <- reorder.as.map2(age_cod_results_infants_AOC, "state", "stabbrs")
+
+cod_decomp_results <- read.csv("../Results2/cod_decomp_results_cubic.csv")
+cod_decomp_results$COD <- factor(cod_decomp_results$COD, levels(cod_decomp_results$COD)[c(3, 2, 4, 6, 5, 1)])
 ```
@@
 ```
+
+**Figure X: Trend in the difference between each state's trend and the national trend, by causes of death and gender, 1969-2013**
+
+```{r, fig.width = 10, fig.height = 90, echo=F}
+ggplot(cod_decomp_results %>% filter(state != "Washington DC"), 
+       aes(x = year, y = residual.RE)) + geom_line(aes(col = COD)) +
+  facet_grid(state~sex) + geom_hline(yintercept = 0, lty = 2) +
+  ylab("Difference between state and national trend (years)") +
+  xlab("Year")
+```
+
+```{r, fig.width = 10, fig.height = 3, echo=F}
+ggplot(cod_decomp_results %>% filter(state == "Washington DC"), 
+       aes(x = year, y = residual.RE)) + geom_line(aes(col = COD)) +
+  facet_grid(state~sex) + geom_hline(yintercept = 0, lty = 2) +
+  ylab("Difference between state and national trend (years)") +
+  xlab("Year")
+```
+
+The purpose of these plots are to summarize the information presented in Figures 3A-7B of the main paper and allow the reader to easily make comparisons between men and women and across causes of death.
+Each trend line shows the difference between each state's estimate and the national average for each cause of death.
+Thus, values close to zero are closest to the national average, values below zero imply that the state's contribution was lower than the national average and values above zero imply that the state's contribution was higher than the national average.

### 2018-03-23 Manuscript/Appendix-agestd-mort-rates.Rmd

--- 2017-07-04 Manuscript/Appendix-agestd-mort-rates.Rmd
@@
 
-This is the supplementary appendix for the paper, "Trendfs in the contribution of major causes of death to the black-white life expectancy gap by US state.
+This is the supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state.
 A retrospective study of all deaths between 1969 and 2013."
 
-Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Sam Harper, PhD; Jay S Kaufman, PhD
+Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Jay S Kaufman, PhD; Sam Harper, PhD;
 

### 2018-03-23 Manuscript/Appendix-compare-COD-gender-within-state.Rmd

---
title: "Supplementary Appendix 2: Trend in the difference between each state's contribution trend and the overall trend, by causes of death and gender, 1969-2013"
output: word_document
---

This is the second supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013."

Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Jay S Kaufman, PhD; Sam Harper, PhD

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>

```{r load_libraries, echo=FALSE, warning=F, message=F}
library(dplyr)
library(ggplot2)

cod_decomp_results <- read.csv("../Results2/cod_decomp_results_cubic.csv")
cod_decomp_results$COD <- factor(cod_decomp_results$COD, levels(cod_decomp_results$COD)[c(3, 2, 4, 6, 5, 1)])
```

The purpose of these plots are to summarize the information presented in Figures 3A-7B of the main paper and allow the reader to easily make comparisons between men and women and across causes of death. Each trend line shows the difference between each state's estimate of the cause's contribution to the black-white life expectancy gap and the overall average. That is, the difference between the state-level blue line and red smoothed line in Figures 3A and 7B. Thus, values close to zero are closest to the overall average, values below zero imply that the state-cause contribution was lower than the overall average for that cause and values above zero imply that the state-cause contribution was higher than the overall average.

```{r, fig.width = 10, fig.height = 5, echo=F, warning=F, message=F, comment=NA, null_prefix=TRUE}
results <- cod_decomp_results %>% 
  filter(state != "Washington DC") %>%
  group_by(state) %>% 
  do(plots = ggplot(data = .) +
       aes(x = year, y = residual.RE) + 
       geom_line(aes(col = COD)) +
       facet_wrap( ~ sex) + 
       geom_hline(yintercept = 0, lty = 2) +
       scale_y_continuous(limits = c(-2, 2)) +
       ylab("Difference between state-specific and overall contribution trends (years)") +
       xlab("Year") + 
       ggtitle(first(.$state))
  )

results$plots
```

```{r, fig.width = 10, fig.height = 5, echo=F}
ggplot(cod_decomp_results %>% filter(state == "Washington DC"), 
       aes(x = year, y = residual.RE)) + geom_line(aes(col = COD)) +
  facet_grid(state~sex) + geom_hline(yintercept = 0, lty = 2) +
  ylab("Difference between state-specific and overall contribution trend (years)") +
  xlab("Year") + ggtitle("Washington DC")
```

### 2018-04-09 Code/7_Explore-Hispanic-origin.Rmd
explore hisp origin

---
title: "Hispanic origin"
author: "Corinne Riddell"
date: "April 9, 2018"
output: word_document
---

```{r cancer}
library(tidyverse)
nh.cancer <- read.table("./Data/hispanic/NH Black and White Neoplasms.txt", header = T, nrows = 360, sep="\t")

nh.cancer <- nh.cancer %>% select(Year, Gender, Race, State, Deaths, Population, Crude.Rate, Age.Adjusted.Rate) %>%
  mutate(Hispanic.Origin = "Non-hispanic")

all.cancer <- read.table("./Data/hispanic/All Black and White Neoplasms.txt", header = T, nrows = 360, sep="\t")

all.cancer <- all.cancer %>% select(Year, Gender, Race, State, Deaths, Population, Crude.Rate, Age.Adjusted.Rate) %>%
  mutate(Hispanic.Origin = "All")

cancer <- dplyr::bind_rows(nh.cancer, all.cancer)

cancer2 <- cancer %>% 
  mutate(Race = plyr::revalue(Race, c("White" = "White",
                                      "Black or African American" = "Black")))

fig1 <- ggplot(cancer, aes(x = Year, y = Age.Adjusted.Rate)) + geom_line(aes(col = Race, lty = Hispanic.Origin)) + facet_grid(State~Gender)
```

```{r circulatory}
nh.circulatory <- read.table("../Data/hispanic/NH Black and White Circulatory.txt", header = T, nrows = 360, sep="\t")

nh.circulatory <- nh.circulatory %>% select(Year, Gender, Race, State, Deaths, Population, Crude.Rate, Age.Adjusted.Rate) %>%
  mutate(Hispanic.Origin = "Non-hispanic")

all.circulatory <- read.table("../Data/hispanic/All Black and White Circulatory.txt", header = T, nrows = 360, sep="\t")

all.circulatory <- all.circulatory %>% select(Year, Gender, Race, State, Deaths, Population, Crude.Rate, Age.Adjusted.Rate) %>%
  mutate(Hispanic.Origin = "All")

circulatory <- dplyr::bind_rows(nh.circulatory, all.circulatory)

circulatory2 <- circulatory %>% 
  mutate(Race = plyr::revalue(Race, c("White" = "White",
                                      "Black or African American" = "Black")))

fig2 <- ggplot(circulatory, aes(x = Year, y = Age.Adjusted.Rate)) + geom_line(aes(col = Race, lty = Hispanic.Origin)) + facet_grid(State~Gender)
```

```{r injuries}
nh.injuries <- read.table("./Data/hispanic/NH Black and White Injuries.txt", header = T, nrows = 360, sep="\t")

nh.injuries <- nh.injuries %>% select(Year, Gender, Race, State, Deaths, Population, Crude.Rate, Age.Adjusted.Rate) %>%
  mutate(Hispanic.Origin = "Non-hispanic")

all.injuries <- read.table("./Data/hispanic/All Black and White Injuries.txt", header = T, nrows = 360, sep="\t")

all.injuries <- all.injuries %>% select(Year, Gender, Race, State, Deaths, Population, Crude.Rate, Age.Adjusted.Rate) %>%
  mutate(Hispanic.Origin = "All")

injuries <- dplyr::bind_rows(nh.injuries, all.injuries)

injuries2 <- injuries %>% 
  mutate(Race = plyr::revalue(Race, c("White" = "White",
                                      "Black or African American" = "Black")))

fig3 <- ggplot(injuries, aes(x = Year, y = Age.Adjusted.Rate)) + geom_line(aes(col = Race, lty = Hispanic.Origin)) + facet_grid(State~Gender)
```

```{r}
ggsave(filename = "../Plots/Cancer.jpeg", plot = fig1, device = "jpeg", height = 9.5, width = 8, units = "in")
ggsave(filename = "../Plots/Cardiovacular.jpeg", plot = fig2, device = "jpeg", height = 9.5, width = 8, units = "in")
ggsave(filename = "../Plots/Injuries.jpeg", plot = fig3, device = "jpeg", height = 9.5, width = 8, units = "in")
```

### 2018-04-09 Code/7_Explore-Hispanic-origin.Rmd

@@
 fig3 <- ggplot(injuries, aes(x = Year, y = Age.Adjusted.Rate)) + geom_line(aes(col = Race, lty = Hispanic.Origin)) + facet_grid(State~Gender)
+
+ggplot(injuries, aes(x = Year, y = Age.Adjusted.Rate)) + geom_point(aes(col = Race, shape = Hispanic.Origin)) + facet_grid(State~Gender)
 ```

### 2018-04-10 Manuscript/Appendix_Explore-Hispanic-origin.Rmd

---
title: "Supplementary Appendix 3: Trends in black and white mortality rates before and after excluding Hispanics, 1999-2013"
output: github_document
---

This is the third supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013."

Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Jay S Kaufman, PhD; Sam Harper, PhD

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>

In our main analyses, we included Hispanics as part of the black and white racial groups. To investigate the impact of including Hispanics on mortality, we inspected mortality trends for five key states with large Hispanic immigrant populations (California, New York, Florida, Texas, and Illinois). We extracted this mortality data using (CDC WONDER's)[https://wonder.cdc.gov/] compressed mortality file. Hispanic ethnicity is only available beginning in 1999, so we consider the time period from 1999 to 2013. We extract age-adjusted mortality rates related to the circulatory system ("CVD", ICD-10 codes I00-I99), neoplasms ("cancers", ICD-10 codes C00-D48) and external causes ("injuries", ICD-10 codes V01-Y89) by year, state, race, and gender, for (a) non-Hispanics and (b) regardless of Hispanic ethinicity.

**Trends in cancer mortality for all whites and blacks vs. non-Hispanic whites and blacks, 1999-2013**
```{r load-cancer-data, echo=F, fig.width = 8, fig.height=9.5}
library(tidyverse)
nh.cancer <- read.table("../Data/hispanic/NH Black and White Neoplasms.txt", header = T, nrows = 360, sep="\t")

nh.cancer <- nh.cancer %>%
  select(Year, Gender, Race, State, Deaths, Population, Crude.Rate, Age.Adjusted.Rate) %>%
  mutate(Hispanic.Origin = "Non-hispanic")

all.cancer <- read.table("../Data/hispanic/All Black and White Neoplasms.txt", header = T, nrows = 360, sep="\t")

all.cancer <- all.cancer %>% 
  select(Year, Gender, Race, State, Deaths, Population, Crude.Rate, Age.Adjusted.Rate) %>%
  mutate(Hispanic.Origin = "All")

cancer <- dplyr::bind_rows(nh.cancer, all.cancer) %>%
  mutate(Race = plyr::revalue(Race, c("White" = "White",
                                      "Black or African American" = "Black"))) %>%
  filter(Year < 2014)

fig1 <- ggplot(cancer, aes(x = Year, y = Age.Adjusted.Rate)) + 
  geom_line(aes(col = Race, lty = Hispanic.Origin)) + facet_grid(State ~ Gender) +
  ylab("Age-adjusted mortality rate (per 100,000)") + 
  guides(lty = guide_legend(title = "Hispanic Origin")) +
  theme(legend.position = "top")

fig1
```

**Trends in CVD mortality for all whites and blacks vs. non-Hispanic whites and blacks, 1999-2013**
```{r load-circulatory-data, echo=F, fig.width = 8, fig.height=9.5}
nh.circulatory <- read.table("../Data/hispanic/NH Black and White Circulatory.txt", header = T, nrows = 360, sep="\t")

nh.circulatory <- nh.circulatory %>% 
  select(Year, Gender, Race, State, Deaths, Population, Crude.Rate, Age.Adjusted.Rate) %>%
  mutate(Hispanic.Origin = "Non-hispanic")

all.circulatory <- read.table("../Data/hispanic/All Black and White Circulatory.txt", header = T, nrows = 360, sep="\t")

all.circulatory <- all.circulatory %>% 
  select(Year, Gender, Race, State, Deaths, Population, Crude.Rate, Age.Adjusted.Rate) %>%
  mutate(Hispanic.Origin = "All")

circulatory <- dplyr::bind_rows(nh.circulatory, all.circulatory)

circulatory2 <- circulatory %>% 
  mutate(Race = plyr::revalue(Race, c("White" = "White",
                                      "Black or African American" = "Black"))) %>%
  filter(Year < 2014)

fig2 <- ggplot(circulatory, aes(x = Year, y = Age.Adjusted.Rate)) + 
  geom_line(aes(col = Race, lty = Hispanic.Origin)) + facet_grid(State ~ Gender) +
  ylab("Age-adjusted mortality rate (per 100,000)") + 
  guides(lty = guide_legend(title = "Hispanic Origin")) +
  theme(legend.position = "top")

fig2
```

**Trends in injury mortality for all whites and blacks vs. non-Hispanic whites and blacks, 1999-2013**
```{r load-injuries-data, echo=F, fig.width = 8, fig.height=9.5}
nh.injuries <- read.table("../Data/hispanic/NH Black and White Injuries.txt", header = T, nrows = 360, sep="\t")

nh.injuries <- nh.injuries %>% 
  select(Year, Gender, Race, State, Deaths, Population, Crude.Rate, Age.Adjusted.Rate) %>%
  mutate(Hispanic.Origin = "Non-hispanic")

all.injuries <- read.table("../Data/hispanic/All Black and White Injuries.txt", header = T, nrows = 360, sep="\t")

all.injuries <- all.injuries %>% 
  select(Year, Gender, Race, State, Deaths, Population, Crude.Rate, Age.Adjusted.Rate) %>%
  mutate(Hispanic.Origin = "All")

injuries <- dplyr::bind_rows(nh.injuries, all.injuries)

injuries2 <- injuries %>% 
  mutate(Race = plyr::revalue(Race, c("White" = "White",
                                      "Black or African American" = "Black"))) %>%
  filter(Year < 2014)

fig3 <- ggplot(injuries, aes(x = Year, y = Age.Adjusted.Rate)) + 
  geom_line(aes(col = Race, lty = Hispanic.Origin)) + 
  facet_grid(State ~ Gender) +
  ylab("Age-adjusted mortality rate (per 100,000)") + 
  guides(lty = guide_legend(title = "Hispanic Origin")) +
  theme(legend.position = "top")

fig3
```

```{r, echo=F}
ggsave(filename = "../Plots/Cancer.jpeg", plot = fig1, device = "jpeg", height = 9.5, width = 8, units = "in")
ggsave(filename = "../Plots/Cardiovacular.jpeg", plot = fig2, device = "jpeg", height = 9.5, width = 8, units = "in")
ggsave(filename = "../Plots/Injuries.jpeg", plot = fig3, device = "jpeg", height = 9.5, width = 8, units = "in")
```

### 2018-04-10 Manuscript/Abstract-DataVizDemo.Rmd
writing

@@
 
-```{r, echo=F, message=F, warning=F}
+```{r load-libraries, echo=F, message=F, warning=F}
 library(shiny) #, lib.loc = "/Library/Frameworks/R.framework/Versions/3.3/Resources/library/old_versions") #OLD version: 0.14.1, NEW: 1.0.3
@@
 
-```{r, echo=F, message=F, warning=F}
+```{r load-data, echo=F, message=F, warning=F}
 age_decomp_results <- read.csv("../Results2/age_decomp_results.csv")

### 2018-04-11 Manuscript/Abstract-DataVizDemo.Rmd
writing

@@
 
-source("../Code/life_expectancy_functions.R")
+source("./Code/life_expectancy_functions.R")
 ```
@@
 ```{r load-data, echo=F, message=F, warning=F}
-age_decomp_results <- read.csv("../Results2/age_decomp_results.csv")
+age_decomp_results <- read.csv("./Results2/age_decomp_results.csv")
 
@@
 
-cod_decomp_results <- read.csv("../Results2/cod_decomp_results.csv")
-cod_change_results <- read.csv("../Results2/cod_change_results.csv")
-BlackWhite_results <- read.csv("../Results2/BlackWhite_results.csv")
-dat.aggregated <- read.csv("../Data/dat_aggregated.csv")
+cod_decomp_results <- read.csv("./Results2/cod_decomp_results_cubic.csv")
+cod_change_results <- read.csv("./Results2/cod_change_results_cubic.csv")
+BlackWhite_results <- read.csv("./Results2/BlackWhite_results.csv")
+dat.aggregated <- read.csv("./Data/dat_aggregated.csv")
 
@@
 
+ly1 <- ggplotly(
+  ggplot(data = temp,
+         aes(y = COD_cont_yrs_mean, x = state.reorder2.n, fill = COD)) + coord_flip() +
+    geom_bar(stat = "identity") + theme_minimal() + #geom_vline(aes(xintercept = 0)) +
+    theme(legend.title = element_blank(),
+          panel.background = element_rect(fill = "transparent", colour = NA), 
+          plot.background = element_rect(fill = "transparent", colour = NA),
+          legend.background = element_rect(fill = "transparent", colour = NA)) +
+    ylab("Contribution to the life expectancy gap (years)") +
+    xlab("") +
+    scale_x_continuous(breaks = 1:length(levels(factor(temp$state.reorder2))), 
+                       labels = levels(factor(temp$state.reorder2))) 
+  ) 
+ly1
+
 ```

### 2018-05-09 Manuscript/Appendix-agestd-mort-rates.Rmd
update plots

@@
 ```{r plot_LE_trends_males, echo=F, fig.height=10, fig.width=13}
-LE_trend_plot_males +
+le.gap.men <- LE_trend_plot_males +
   geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)
+  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  ggtitle("A) Black-white life expectancy gap in men")
 
@@
 ```{r plot_LE_trends_females, echo=F, fig.height=10, fig.width=13}
-LE_trend_plot_females +
+le.gap.women <- LE_trend_plot_females +
   geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5)
+  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
+  ggtitle("B) Black-white life expectancy gap in women")
 
+ggsave(filename = "../Plots/figure1a.jpeg", plot = le.gap.men, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure1b.jpeg", plot = le.gap.women, device = "jpeg", width = 12, height = 10, units = "in")
 ```

### 2018-05-10 Manuscript/Appendix-agestd-mort-rates.Rmd
update appendix

--- 2018-03-23 Manuscript/Appendix-agestd-mort-rates.Rmd
@@
 ---
-title: "Supplementary Appendix: Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013"
-output: html_document
+title: 'Supplementary Appendix: Trends in the contribution of major causes of death
+  to the black-white life expectancy gap by US state'
+output:
+  word_document: default
+  html_document: default
 ---
 
-This is the supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state.
-A retrospective study of all deaths between 1969 and 2013."
+This is the supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state".
 
-Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Jay S Kaufman, PhD; Sam Harper, PhD;
+For an interactive version of all our results, please visit XXXX.
 
-Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada.
+```{r load_libraries, echo=FALSE, warning=F, message=F}
+#Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Sam Harper, PhD; Jay S Kaufman, PhD
+
+#Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, #1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada.
 Email: <corinne.riddell@mail.mcgill.ca>
 
-```{r load_libraries, echo=FALSE, warning=F, message=F}
 library(dplyr)
@@
 age_cod_results_infants_AOC <- reorder.as.map2(age_cod_results_infants_AOC, "state", "stabbrs")
-
-cod_decomp_results <- read.csv("../Results2/cod_decomp_results_cubic.csv")
-cod_decomp_results$COD <- factor(cod_decomp_results$COD, levels(cod_decomp_results$COD)[c(3, 2, 4, 6, 5, 1)])
 ```
@@
 ```{r print-em, fig.width = 9, fig.height = 7, echo = F}
-Cardiovascular.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+Cardiovascular.Male.mort.plot[[1]]
+#Cardiovascular.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
 #Cardiovascular.Male.mort.plot[[4]]
@@
 ```{r print-em-2, fig.width = 9, fig.height = 7, echo = F}
-Cardiovascular.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+Cardiovascular.Female.mort.plot[[1]]
+#Cardiovascular.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
 #Cardiovascular.Female.mort.plot[[4]]
@@
 ```{r print-em-3, fig.width = 9, fig.height = 7, echo = F}
-Cancers.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+Cancers.Male.mort.plot[[1]]
+#Cancers.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
 #Cancers.Male.mort.plot[[4]]
@@
 ```{r print-em-4, fig.width = 9, fig.height = 7, echo = F}
-Cancers.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+Cancers.Female.mort.plot[[1]]
+#Cancers.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
 #Cancers.Female.mort.plot[[4]]
@@
 ```{r, fig.width = 9, fig.height = 7, echo = F}
-`Non-communicable.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
+`Non-communicable.Male.mort.plot`[[1]]
+#`Non-communicable.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
 #`Non-communicable.Male.mort.plot`[[4]]
@@
 ```{r, fig.width = 9, fig.height = 7, echo = F}
-`Non-communicable.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
+`Non-communicable.Female.mort.plot`[[1]]
+#`Non-communicable.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
 #`Non-communicable.Male.mort.plot`[[4]]
@@
 ```{r, fig.width = 9, fig.height = 7, echo = F}
-Injuries.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+Injuries.Male.mort.plot[[1]]
+#Injuries.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
 #Injuries.Male.mort.plot[[4]]
@@
 ```{r, fig.width = 9, fig.height = 7, echo = F}
-Injuries.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70)) 
+Injuries.Female.mort.plot[[1]]
+#Injuries.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70)) 
 #Injuries.Female.mort.plot[[4]]
@@
 ```{r print-3, fig.width = 9, fig.height = 7, echo = F}
-Communicable.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+Communicable.Male.mort.plot[[1]]
+#Communicable.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
 #Communicable.Male.mort.plot[[4]]
@@
 ```{r print-4, fig.width = 9, fig.height = 7, echo = F}
-Communicable.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
+Communicable.Female.mort.plot[[1]]
+#Communicable.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
 #Communicable.Female.mort.plot[[4]]
@@
 ```{r, fig.width = 9, fig.height = 7, echo = F}
-`All other causes.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
+`All other causes.Male.mort.plot`[[1]] 
+#`All other causes.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
 #`All other causes.Male.mort.plot`[[4]]
@@
 ```{r, fig.width = 9, fig.height = 7, echo = F}
-`All other causes.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70)) 
+`All other causes.Female.mort.plot`[[1]]
+#`All other causes.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70)) 
 #`All other causes.Female.mort.plot`[[4]]

### 2018-05-10 Manuscript/Manuscript_COD_Trends_HealthandPlace.Rmd
ch

@@
 
+Jay S Kaufman, PhD
+
 Sam Harper, PhD
-
-Jay S Kaufman, PhD
 
@@
 ```{r print-plot, fig.width = 12, fig.height = 10, echo = F}
-Cardiovascular.Male.plot[[3]] +
+cvd.men <- Cardiovascular.Male.plot[[3]] +
   geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
-           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)
+  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  ggtitle("A) Contribution of CVD to the gap in men")
 
@@
 #ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-Cardiovascular.Female.plot[[3]]  +
+cvd.women <- Cardiovascular.Female.plot[[3]]  +
   geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
-           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)
+  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  ggtitle("B) Contribution of CVD to the gap in women")
 
@@
 #ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-Cancers.Male.plot[[3]] +
+can.men <- Cancers.Male.plot[[3]] +
   geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
-           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  ggtitle("A) Contribution of cancer to the gap in men")
 
@@
 #ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
-Cancers.Female.plot[[3]] +
+can.women <- Cancers.Female.plot[[3]] +
   geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
-           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  ggtitle("B) Contribution of cancer to the gap in women")
 
@@
 #ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-`Non-communicable.Male.plot`[[3]] +
+ncd.men <- `Non-communicable.Male.plot`[[3]] +
   geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
-           scale_y_continuous(breaks = seq(-0.5, 1.5, by = 0.5), labels = seq(-0.5, 1.5, by = 0.5))
+  scale_y_continuous(breaks = seq(-0.5, 1.5, by = 0.5), labels = seq(-0.5, 1.5, by = 0.5)) +
+  ggtitle("A) Contribution of non-communicable disease to the gap in men")
 
@@
 #ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
-`Non-communicable.Female.plot`[[3]] +
+ncd.women <- `Non-communicable.Female.plot`[[3]] +
   geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
-           scale_y_continuous(breaks = seq(-1, 1.5, by = 0.5), labels = seq(-1, 1.5, by = 0.5))
+  scale_y_continuous(breaks = seq(-1, 1.5, by = 0.5), labels = seq(-1, 1.5, by = 0.5)) + 
+  ggtitle("B) Contribution of non-communicable disease to the gap in women")
 
@@
 
-`Injuries.Male.plot`[[3]] +
+inj.men <-`Injuries.Male.plot`[[3]] +
   geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) + 
-           geom_hline(aes(yintercept = 6), lwd = 0.5, col = "grey", alpha = 0.5)
+  geom_hline(aes(yintercept = 6), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  ggtitle("A) Contribution of injury to the gap in men")
 
@@
 
-`Injuries.Female.plot`[[3]] +
+inj.women <- `Injuries.Female.plot`[[3]] +
   geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
-           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)
+  geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  ggtitle("B) Contribution of injury to the gap in women")
 ```
@@
 
-`Communicable.Male.plot`[[3]] +
+com.men <- `Communicable.Male.plot`[[3]] +
   geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) 
+  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
+  ggtitle("A) Contribution of communicable disease to the gap in men")
 ```
@@
 
-`Communicable.Female.plot`[[3]] +
+com.women <- `Communicable.Female.plot`[[3]] +
   geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
+  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
+  ggtitle("B) Contribution of communicable disease to the gap in women")
 ```
@@
 
-`All other causes.Male.plot`[[3]] +
+aoc.men <- `All other causes.Male.plot`[[3]] +
   geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) 
+  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) +
+  ggtitle("A) Contribution of all other causes to the gap in men")
 ```
@@
 
-`All other causes.Female.plot`[[3]] +
+aoc.women <- `All other causes.Female.plot`[[3]] +
   geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
@@
   geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) 
+  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) +
+  ggtitle("B) Contribution of all other causes to the gap in women")
+```
+
+```{r save-figs}
+ggsave(filename = "../Plots/figure2a.jpeg", plot = cvd.men, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure2b.jpeg", plot = cvd.women, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure3a.jpeg", plot = can.men, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure3b.jpeg", plot = can.women, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure4a.jpeg", plot = ncd.men, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure4b.jpeg", plot = ncd.women, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure5a.jpeg", plot = inj.men, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure5b.jpeg", plot = inj.women, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure6a.jpeg", plot = com.men, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure6b.jpeg", plot = com.women, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure7a.jpeg", plot = aoc.men, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure7b.jpeg", plot = aoc.women, device = "jpeg", width = 12, height = 10, units = "in")
 ```

### 2018-05-10 Manuscript/Manuscript_COD_Trends_HealthandPlace.Rmd
updated abstract and intro

@@
 
-```{r summarize_results_for_abstract, echo = F}
-cvd.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>%
-  ungroup() %>% select(fixed.from.RE) %>% round(1)
-cvd.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
-cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
-cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
-
-ncd.male.average <- cod_decomp_results %>% filter(sex == "Male" & COD == "Non-communicable" & state == "Alabama") %>% 
-  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
-ncd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% ungroup() %>%  select(fixed.from.RE) %>% round(1)
-ncd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
-
-can.male.max <- cod_decomp_results %>% filter(sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
-  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))
-can.female.average <- cod_decomp_results %>% filter(sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
-  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
-
-inj.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
-inj.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
-inj.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
-inj.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
-```
-
-**Objective:** To describe the pattern of contributions of six major causes of deaths to the black-white life expectancy gap across US states over the last half-century, and identify divergent states.
-
-**Design:** A retrospective, repeated cross-sectional study of all deaths.
-
-**Setting:** 39 US States and the District of Columbia.
-
-**Participants:** Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major cause-of-death groups.
-
-**Primary outcome measure:** We estimated trends in the gender-specific contribution of six major causes of death to the gap in life expectancy between black and whites and their 95% credible bands.
-
-**Results:** Although mortality from cardiovascular disease has decreased dramatically, its contribution to the black-white life expectancy gap increased over time for men (from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years).
-The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) and decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in the early 1990s) and a stable contribution for women (approximately `r can.female.average` years).
-Both genders exhibited a decreased contribution from injuries (men: `r inj.male.1969` to `r inj.male.2013` years), that became negative for women (women: `r inj.female.1969` to `r inj.female.2013` years).
-Several states diverged from these general trends.
-
-**Conclusions:** The black-white life expectancy gap has decreased substantially in the US since 1969.
-For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality.
-In women, a crossover in injury-related mortality led to a narrower gap, due partially to increasing mortality among whites.
+Life expectancy has increased in the United States over many decades.
+The difference in life expectancy between black and white Americans has also decreased, but some states have made much more progress towards racial equality than others.
+This paper describes the pattern of contributions of six major causes of death to the black-white life expectancy gap within US states and the District of Columbia between 1969 and 2013, and identifies states diverging from the overall pattern.
+Across multiple causes, the District of Columbia, Illinois, Wisconsin, and Michigan had the highest contributions to black-white inequality, while New York, Massachusetts, and Rhode Island had the lowest contributions and have either achieved or are the closest to achieving black-white equality in life expectancy.
 
@@
 Life expectancy is a barometer for overall population health.
-In 2014, life expectancy at birth in the United States was 78.8 years, but hidden within this summary are important subgroups differences.
-Whites lived 3.4 years longer than blacks.^1^ While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied over time.^2^
+In 2014, life expectancy at birth in the United States (US) was 78.8 years, but hidden within this summary are important subgroup differences.
+Whites lived 3.4 years longer than blacks (Cunningham et al., 2017).
+While both black and white Americans have experienced increases in life expectancy since the beginning of the twentieth century, the difference between their life expectancies has varied over time (Fuchs, 2016).
 
-During the 1980s the life expectancy gap widened, but then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths from homicide, HIV/AIDS, unintentional injuries, and, among women, from differential improvements in heart disease.^3^ The gap continued to narrow during the 2000s, but US states differed markedly in the magnitude of improvement.
-For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years.^4^ However, the extent to which different causes of death may have contributed to state-level changes in the black-white gap are unknown.
+During the 1980s the life expectancy gap widened, but then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths from homicide, HIV/AIDS, unintentional injuries, and, among women, from differential improvements in heart disease (Harper et al., 2007).
+The overall black-white gap continued to narrow during the 2000s, but US states differed markedly in the magnitude of improvement.
+For example, Harper et al.
+estimated that the black-white life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years (Harper et al., 2014).
+While the previous publication estimated changes to the black-white life expectancy gap for each US state, it did not shed light on why some states progressed more than others.
 
-The overarching objective of this study is to describe the contribution of selected causes of death to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences.
+Race-specific mortality trajectories within a state are likely affected by both national and state-specific factors, some of which could be affected by state-level policymaking.
+Taking cardiovascular disease (CVD) as an example, several state policies have direct or indirect effects on CVD risk factors.
+For example, excise taxes on tobacco were introduced between 1921 and 1969 across all states, and the magnitude of increases have varied across states, leading to large present-day differences (e.g., \$0.17 in Missouri vs.
+\$4.35 in New York) (Centers for Disease Control and Preventaion, 2018).
+Policies targeting obesity through promotion of physical activity and nutrition also vary by state, (Trust for America's Health and the Robert Wood Johnson Foundation, 2018) as do restrictions on vehicle emissions, with twelve states following the stricter standards imposed in California (California Air Resources Board, 2018).
+Thus, policies and other state-level characteristics may be implicated in deaths from specific causes.
+However, a first step in understanding the potential impact of state policies is to examine the extent to which different causes of death have contributed to state-level changes in the black-white life expectancy gap.
+
+The overarching objective of this study is to describe the contribution of selected causes of death to the black-white life expectancy gap over forty-five years, and identify divergent states.
 To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for men and women.
-This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states.
-This study may therefore facilitate the identification of policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.
+This facilitates the identification of causes of death whose mortality trends for blacks and whites have changed in ways that contributed to narrowing or widening the black-white life expectancy gap, and the identification of states for which cause-specific contributions are particularly high or low compared to the overall average.
+The study may therefore facilitate the identification of policies or social conditions states with narrow black-white gaps, and suggest broad targets for interventions to further reduce the black-white gap in less states that have seen less progress in reducing racial inequality.
 
@@
 **Cardiovascular disease**
+```{r summarize_results, echo = F}
+cvd.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>%
+  ungroup() %>% select(fixed.from.RE) %>% round(1)
+cvd.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+
+ncd.male.average <- cod_decomp_results %>% filter(sex == "Male" & COD == "Non-communicable" & state == "Alabama") %>% 
+  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
+ncd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% ungroup() %>%  select(fixed.from.RE) %>% round(1)
+ncd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+
+can.male.max <- cod_decomp_results %>% filter(sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
+  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))
+can.female.average <- cod_decomp_results %>% filter(sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
+  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
+
+inj.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+inj.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+inj.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+inj.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
+```
 

### 2018-05-10 Manuscript/Manuscript_COD_Trends_HealthandPlace.Rmd
updated methods

@@
 
-We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in 39 US states and the District of Columbia between 1969 and 2013.^5^ We excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming.
-These states contained less than 0.4% of the black population in the United States in 2010.^6^ The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.^7^ The population data were derived by the NCHS using intercensal bridged-race population estimates.
-We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death.
+We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEER\*Stat) to extract data on the number of deaths and the population size in thirty-nine US states and the District of Columbia (DC) between 1969 and 2013 (National Cancer Institute, 2016).
+We excluded eleven states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming.
+These states contained less than 10,000 black residents in the first years' of data, which led to computational issues when we tested our models.
+Overall, these states comprised less 0.4% of the black population in the United States in 2010, implying that their removal will not significantly detract from the depiction of racial inequality in the US that we aim to provide (U.S.
+Census Bureau, 2017a).
+The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System (Centers for Disease Control and Prevention, 2017).
+A thorough description of the population denominators provided by SEER\*Stat is available online (National Cancer Institute, 2018).
+Briefly, the population denominators were derived from US Census data.
+Prior to 2000, US residents could only check one racial categorization (American Indian or Alaskan Native, Asian or Pacific Islander, Black, or White).
+In 2000, multiple racial categories could be checked on the Census but the four-race categorization is still in-use by vital statistics.
+Thus, the population denominators in these years were derived by the NCHS using intercensal bridged-race estimates (2000-2009) or postcensal bridged race estimates (2010-2013), which used algorithms to go from the multiple race categorizations to the four-race system (Ingram et al., 2003).
+We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969-2013), and cause of death.
 Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013).
-These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**Table S1**).
-We did not consider Hispanic ethnicity as part of our racial definition, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.^8^
+These codes were aggregated into six broad causes of death for this analysis (Supplementary Appendix 1 Table S1).
+Major categories are: Cardiovascular disease (the most common cause of death), Cancers (the second most common cause of death), Non-communicable disease (often related to diabetes, obesity, or aging, three risk factors that became more prevalent over the study time period), Communicable disease (capturing periods of increased infectious activity), Injuries (capturing external causes of death, and suicide, which are of increased importance in early adulthood), and All other causes (capturing infant mortality, maternal mortality, and ill-defined causes).
+We chose these six categories to balance the limitations of parsing the data too finely, which would lead to data suppression and grouping causes too broadly, which would limit their interpretability.
+These major causes also approximate what has been used in previous research and aid extension of these earlier findings (Harper et al., 2007; Harper et al., 2012).
+We did not consider Hispanic ethnicity as part of our racial definition, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census (Arial et al., 2008).
 
@@
 The data structure produced two statistical challenges that were addressed by our modeling strategy.
-Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^9^ which we needed to impute.
+Firstly, because of confidentiality, NCHS suppressed counts of deaths between one and nine, (National Cancer Institute, 2017a) which we needed to impute.
 Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata.
@@
 We used a Bayesian time-series model to smooth the death counts across years and estimate the underlying mortality rate.
-Specifically, we used a Bayesian generalized linear model with a Poisson likelihood to model the number of deaths in each state-year-sex-race-age-cause stratum as a function of the at-risk population, for strata with zero or more than 9 deaths.
+Specifically, we used a Bayesian generalized linear model with a Poisson likelihood to model the number of deaths in each state-year-sex-race-age-cause stratum as a function of the at-risk population, for strata with zero or more than nine deaths.
 The mortality rates over time within each stratum were assigned an autoregressive prior, where the log mortality rate was assumed to be Gaussian centered at the previous year's mortality rate.
@@
 Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition.
-These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.
+These 1,000 decompositions formed the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.
 This modeling was done in R (version 3.3.2) using the *rjags* package.
@@
 
-To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total number of deaths in each age group, and standard life table methods were applied.^10,11^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender.
+To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total number of deaths in each age group, and standard life table methods were applied (Preston et al., 2001; Selvin, 2004).
+Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender.
 The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.
 
-To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method.^12^ This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher cause-specific mortality rates among blacks.
-For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher rate of mortality due to cardiovascular disease among blacks compared to whites.
-If the decomposition produces a negative estimate this implies that whites have a higher rate of mortality for injuries compared to blacks.
-Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 40 state-level trends and their credible intervals.
+To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method (Arriaga, 1985).
+This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher cause-specific mortality rates among blacks.
+For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher rate of mortality due to CVD among blacks compared to whites.
+If the decomposition produces a negative estimate this implies that whites have a higher rate of CVD mortality for injuries compared to blacks.
+Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing forty trends (thirty-nine states and DC) and their credible intervals.
 
 We also calculated state-level trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results.
-We used the US standard population from the year 2000.^13^ The calculations described in this section were performed using functions written by the first author.
+We used the US standard population from the year 2000 (National Cancer Institute, 2017b).
+The calculations described in this section were performed using functions written by the first author.
 
@@
 
-A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on Github: link [DOI: removed for double blind review]
+A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on GitHub: https://github.com/corinne-riddell/BlackWhiteMortalityGap/.
+An online web application interactively hosts the findings (https://corinne-riddell.shinyapps.io/black-white-life-expectancy/).
 

### 2018-05-10 Manuscript/Appendix-agestd-mort-rates.Rmd
updated results

@@
 source("../Code/life_expectancy_functions.R")
-
-BlackWhite_results <- read.csv("../Results2/BlackWhite_results.csv")
-BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")
 
@@
 |
-
-**Figure S1: State-level trends in the black-white life expectancy gap in men, United States, 1969-2013**
-
-```{r life-expectancy-trends, echo=F}
-LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
-  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
-  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
-  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 14) +
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  labs(fill = "Census Division") +
-  ylab("Life expectancy gap (years)") +
-  xlab("Year (1969-2013)")
-
-LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
-  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
-  labs(fill = "Census Division") +
-  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
-  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 14) +
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Life expectancy gap (years)") +
-  xlab("Year (1969-2013)")
-```
-
-```{r plot_LE_trends_males, echo=F, fig.height=10, fig.width=13}
-le.gap.men <- LE_trend_plot_males +
-  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5) + 
-  ggtitle("A) Black-white life expectancy gap in men")
-
-```
-
-**Figure S1 Legend:** The bands illustrate the 95% credible interval around the mean trend line.
-States with more imprecise estimates (often due to small black populations) have wider credible intervals.
-The estimated life expectancy gap in California in 1972 is much higher than in surrounding years.
-This anomaly is also illustrated in Figure 3, and is a product of an unexpectedly low CVD-related mortality rate among white males in California in 1972, likely denoting a coding error.
-Thus the data for this state and year should be interpreted with caution.
-
-**Figure S2: State-level trends in the black-white life expectancy gap in women, United States, 1969-2013**
-
-```{r plot_LE_trends_females, echo=F, fig.height=10, fig.width=13}
-le.gap.women <- LE_trend_plot_females +
-  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
-  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
-  ggtitle("B) Black-white life expectancy gap in women")
-
-ggsave(filename = "../Plots/figure1a.jpeg", plot = le.gap.men, device = "jpeg", width = 12, height = 10, units = "in")
-ggsave(filename = "../Plots/figure1b.jpeg", plot = le.gap.women, device = "jpeg", width = 12, height = 10, units = "in")
-```
-
-```{r plot_both, eval=F, echo=F}
-#Informative: male and females at the same time
-ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
-  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
-  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
-  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
-  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
-  theme_classic(base_size = 14) +
-  theme(axis.text.x = element_blank(),
-        strip.background=element_blank(),
-        axis.line=element_blank(),
-        axis.ticks=element_blank()) +
-  ylab("Life expectancy gap (years)") +
-  xlab("Year (1969-2013)"))
-```
 

### 2018-05-10 Manuscript/Manuscript_COD_Trends_HealthandPlace.Rmd

@@
 age_cod_results_infants_AOC <- read.csv("../Results2/age_cod_results_infants_AOC_cubic.csv")
+
+BlackWhite_results <- read.csv("../Results2/BlackWhite_results.csv")
+BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")
 ```
@@
 
-**Figure S1 and S2** display trends in the life expectancy gap for men and women, respectively.
+**Figs 1A and 1B** display trends in the life expectancy gap for men and women, respectively.
+These tile maps are organized to approximate the shape of the US and the spatial relations among states.
 For men and women, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among men.
 
-**Figures 1 - 6** display trends in the contribution of each broad cause of death to the life expectancy gap for men (panel A) and women (panel B), in years, for each state.
+**Figure 1A: State-level trends in the black-white life expectancy gap in men, United States, 1969-2013**
+
+```{r life-expectancy-trends, echo=F}
+LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
+  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
+  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
+  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 14) +
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  labs(fill = "Census Division") +
+  ylab("Life expectancy gap (years)") +
+  xlab("Year (1969-2013)")
+
+LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
+  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
+  labs(fill = "Census Division") +
+  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
+  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 14) +
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Life expectancy gap (years)") +
+  xlab("Year (1969-2013)")
+```
+
+```{r plot_LE_trends_males, echo=F, fig.height=10, fig.width=12}
+le.gap.men <- LE_trend_plot_males +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5) + 
+  ggtitle("A) Black-white life expectancy gap in men")
+
+le.gap.men
+```
+
+**Figure 1A Caption:** The bands illustrate the 95% credible interval around the mean trend line for men.
+States with more imprecise estimates (often due to small black populations) have wider credible intervals.
+The estimated life expectancy gap in California in men in 1972 is much higher than in surrounding years.
+This anomaly is also illustrated in Fig.
+3, and is a product of an unexpectedly low CVD-related mortality rate among white males in California in 1972, likely denoting a coding error.
+Thus the data for this state and year should be interpreted with caution.
+
+**Figure 1B: State-level trends in the black-white life expectancy gap in women, United States, 1969-2013**
+
+```{r plot_LE_trends_females, echo=F, fig.height=10, fig.width=12}
+le.gap.women <- LE_trend_plot_females +
+  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
+  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
+  ggtitle("B) Black-white life expectancy gap in women")
+
+le.gap.women
+```
+
+**Figure 1B Caption:** The bands illustrate the 95% credible interval around the mean trend line for women.
+States with more imprecise estimates (often due to small black populations) have wider credible intervals.
+
+```{r plot_both, eval=F, echo=F}
+#Informative: male and females at the same time
+#Not used
+ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
+  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
+  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
+  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
+  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
+  theme_classic(base_size = 14) +
+  theme(axis.text.x = element_blank(),
+        strip.background=element_blank(),
+        axis.line=element_blank(),
+        axis.ticks=element_blank()) +
+  ylab("Life expectancy gap (years)") +
+  xlab("Year (1969-2013)"))
+```
+
+**Figures 2-7** display trends in the contribution of each broad cause of death to the life expectancy gap for men (panel A) and women (panel B), in years, for each state.
 The blue line depicts the mean of the smoothed contribution estimated for each state, such that states can have very different trend lines.
 These lines are overlaid with the meta-estimated national trend, shown with a red line.
-We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**Figures S3-S14**).
+
+For causes of death that have undergone mortality reductions over the time period (like CVD), narrowing contributions of causes to the black-white life expectancy gap will result from faster declines in cause-specific mortality among blacks vs.
+whites.
+However, narrowing gaps can also occur even if mortality increased for both groups, or if it increased for whites but decreased for blacks.
+Thus, we supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death for blacks and whites (**Supplementary Appendix Figures S1-S12**).
+To aid direct comparisons of each cause’s contribution to each other and between men and women in a given state, plots overlaying all causes onto a single graph for each state side-by-side for men and women can be found in **Supplementary Appendix 2**.
 
@@
 
-Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races.
-Mortality from CVD decreased markedly since 1969 for all groups (**Figures S3 and S4**), but white men underwent relatively larger mortality decreases than black men.
-As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years (**Figure 1A**).
-Black women experienced relatively larger CVD mortality decreases over time compared to white women, resulting in a decrease of CVD to the contribution to the life expectancy gap from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years nationally (**Figure 1B**).
+Cardiovascular disease (CVD) is the most common cause of death for black and white men and women.
+Mortality from CVD decreased markedly since 1969 for all groups (**Supplementary Appendix 1 Figs.
+S1 and S2**), but white men underwent relatively larger mortality decreases than black men.
+As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years (**Fig.
+2A**).
+Black women experienced relatively larger CVD mortality decreases over time compared to white women, resulting in a decrease of CVD to the contribution to the life expectancy gap from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years nationally (**Figure 2B**).
 For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.
 
-**Figure 1A: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
+**Figure 2A: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
 
@@
 
+cvd.men
 ```
 
-**Figure 1 Legend:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of CVD to the black-white life expectancy gap over time.
+**Figure 2A Caption:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of CVD to the black-white life expectancy gap over time for men.
 The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
@@
 
-**Figure 1B: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
+**Figure 2B: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
 
@@
 
+cvd.women
 ```
+
+**Figure 2B Caption:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of CVD to the black-white life expectancy gap over time for women.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
 
@@
 
-For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Figures S5 and S6**).
-Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 2A**), with a peak contribution of `r can.male.max` in the early 1990s.
-For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a generally stable national pattern in the contribution of cancer to the life expectancy over time of approximately `r can.female.average` years (**Figure 2B**).
+For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Supplementary Appendix 1 Figs.
+S3 and S4**).
+Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 3A**), with a peak contribution of `r can.male.max` in the early 1990s.
+For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a generally stable national pattern in the contribution of cancer to the life expectancy gap over time of approximately `r can.female.average` years (**Figure 3B**).
 For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.
 
-**Figure 2A: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
+**Figure 3A: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
 
@@
 
+can.men
 ```
 
-**Figure 2B: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
+**Figure 3A Caption:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of cancer to the black-white life expectancy gap over time for men.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
+
+**Figure 3B: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
 
@@
 
+can.women
 ```
+
+**Figure 3B Caption:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of cancer to the black-white life expectancy gap over time for women.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
 
@@
 
-Mortality from non-communicable disease increased over time for men in most states, with DC as a notable exception (**Figure S7**).
+Mortality from non-communicable disease increased over time for men in most states, with DC as a notable exception (**Supplementary Appendix 1 Figs.
+S5**).
 Recently, there was a slight decrease in mortality rates, especially among black men.
-These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Figure S8**).
-As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men (approximately `r ncd.male.average` years), and exhibited a gradual decrease over time among women from `r ncd.female.1969` to `r ncd.female.2013` years (**Figures 3A and 3B**).
+These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Supplementary Appendix 1 Figs.
+S6**).
+As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men (approximately `r ncd.male.average` years), and exhibited a gradual decrease over time among women from `r ncd.female.1969` to `r ncd.female.2013` years (**Figures 4A and 4B**).
 As with other causes, New York and Massachusetts exhibited decreasing contributions, and Nevada also made substantial progress.
@@
 
-**Figure 3A: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
+**Figure 4A: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
 
@@
 
+ncd.men
 ```
 
-**Figure 3B: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
+**Figure 4A Caption:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of non-communicable disease to the black-white life expectancy gap over time for men.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
+
+**Figure 4B: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
 
@@
 
+ncd.women
 ```
+
+**Figure 4B Caption:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of non-communicable disease to the black-white life expectancy gap over time for women.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
 
@@
 Trends in mortalilty from injuries differ for blacks and whites.
-Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S9**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S10**).
-Thus, injury mortality rates among blacks are equivalent to, or lower than those among whites in many states, with white women demonstrating sharper recent increases.
+Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Supplementary Appendix 1 Figs.
+S7**) (Harper et al., 2007).
+For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Supplementary Appendix 1 Figs.
+S8**).
+Thus, in the most recent data, injury mortality rates among blacks are equivalent to, or lower than those among whites in many states, with white women demonstrating sharper recent increases.
 The contribution of injuries to the mortality gap has therefore decreased substantially over time from `r inj.male.1969` to `r inj.male.2013` years among men.
-Among women, the contribution went from `r inj.female.1969` in 1969 to `r inj.female.2013` in 2013, becoming negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 4A and 4B**).
+Among women, the contribution went from `r inj.female.1969` in 1969 to `r inj.female.2013` in 2013, becoming negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 5A and 5B**).
 Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions.
@@
 
-**Figure 4A: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
+**Figure 5A: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
 
@@
 
+inj.men
 ```
 
-**Figure 4B: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
+**Figure 5A Caption:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of injury to the black-white life expectancy gap over time for men.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
+
+**Figure 5B: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
 
@@
   ggtitle("B) Contribution of injury to the gap in women")
+
+inj.women
 ```
+
+**Figure 5B Caption:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of injury to the black-white life expectancy gap over time for women.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
 
@@
 
-Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Figures S11 and S12**).
+Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Supplementary Appendix 1 Figs.
+S9 and S10**).
 This spike was greater among men, especially blacks, and among states with larger urban centers.
-The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the early 1990s on `r com.male.mid.max` years in men and `r com.female.mid.max` years in women, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 5A and 5B**).
+The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the early 1990s on `r com.male.mid.max` years in men and `r com.female.mid.max` years in women, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 6A and 6B**).
 DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.
 
-**Figure 5A: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
+**Figure 6A: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
 
@@
   ggtitle("A) Contribution of communicable disease to the gap in men")
+
+com.men
 ```
 
-**Figure 5B: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
+**Figure 6A Caption:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of communicable disease to the black-white life expectancy gap over time for men.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
+
+**Figure 6B: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
 
@@
   ggtitle("B) Contribution of communicable disease to the gap in women")
+
+com.women
 ```
+
+**Figure 6B Caption:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of communicable disease to the black-white life expectancy gap over time for women.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
+
 **All other causes**
@@
 
-Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**Figures S13 and S14**).
+Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**Supplementary Appendix 1 Figs.
+S11 and S12**).
 White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states.
-As such, the contribution of all other causes to the life expectancy gap has generally decreased over time (from `r sprintf("%.1f", aoc.male.endpoints[1, 1])` to `r aoc.male.endpoints[2, 1]` years among males and `r aoc.female.endpoints[1, 1]` to `r aoc.female.endpoints[2, 1]` years among females), and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 6A and 6B**).
+As such, the contribution of all other causes to the life expectancy gap has generally decreased over time (from `r sprintf("%.1f", aoc.male.endpoints[1, 1])` to `r aoc.male.endpoints[2, 1]` years among males and `r aoc.female.endpoints[1, 1]` to `r aoc.female.endpoints[2, 1]` years among females), and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 7A and 7B**).
 The most disparate states are Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.
@@
 Because the majority of infant deaths are classified as being due to all other causes in our analysis, we also tried to separate any impact of all other causes from those of infant mortality more generally.
-We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (**Figures S15 and S16**).
+We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (**Supplementary Appendix 1 Figs.
+S13 and S14**).
 We found that `r proportions.males[1,1]*100`% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969 in males and `r round(proportions.females[1,1]*100, 0)`% in females.
 In 2013, this proportion reached `r proportions.males[2,1]*100`% in males and `r proportions.females[2,1]*100`% in females.
-Thus, the trend in the all other causes contribution partially reflects a decreasing contribution from infant mortality.
+Thus, the trend in the all other causes contribution partially reflects a increasing contribution from infant mortality.
 
-**Figure 6A: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
+**Figure 7A: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
 
@@
   ggtitle("A) Contribution of all other causes to the gap in men")
+
+aoc.men
 ```
 
-**Figure 6B: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
+**Figure 7A Caption:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of all other causes to the black-white life expectancy gap over time for men.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
+
+**Figure 7B: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**
 
@@
   ggtitle("B) Contribution of all other causes to the gap in women")
+
+aoc.women
 ```
 
-```{r save-figs}
+**Figure 7B Caption:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of all other causes to the black-white life expectancy gap over time for women.
+The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
+
+```{r save-figs, echo=F, eval=F}
+ggsave(filename = "../Plots/figure1a.jpeg", plot = le.gap.men, device = "jpeg", width = 12, height = 10, units = "in")
+ggsave(filename = "../Plots/figure1b.jpeg", plot = le.gap.women, device = "jpeg", width = 12, height = 10, units = "in")
 ggsave(filename = "../Plots/figure2a.jpeg", plot = cvd.men, device = "jpeg", width = 12, height = 10, units = "in")

### 2018-05-10 Manuscript/Manuscript_COD_Trends_HealthandPlace.Rmd
updated up to S&W!

@@
 
-```{r save-figs, echo=F, eval=F}
+```{r save-figs-separately, echo=F, eval=F}
 ggsave(filename = "../Plots/figure1a.jpeg", plot = le.gap.men, device = "jpeg", width = 12, height = 10, units = "in")
@@
 
-**Summary of findings**
+**Key findings**
 
-We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45-year period.
+We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a period of forty-five years.
 While we found that many states conformed to a general pattern, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation.
-Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the Midwest, which had been the receiving point of blacks from the South during “the Great Migration” for improved economic opportunities between 1916 and 1970.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.
-On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, because greater racial equality can be achieved by either falling black mortality rates or rising white rates, the reduced life expectancy gap cannot be celebrated without a careful examination of the two individual trends.^16^
 
-Consistent with what has been reported at the national level,^17^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time.
-Injuries continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South.
-This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^18^ Communicable disease continues to contribute to the gap in most states, especially in DC, New Jersey, and Florida.
-In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 nationally,^17^ suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV.
-Cancer is also responsible for large contributions to the life expectancy gap in some states (Illinnois, Michigan, and Wisconsin) but not in others (New York, Massachusetts).
-Overall, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women.
-However, both states began with only a minimal contribution of CVD and cancer to the mortality gap.
-The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contributions of more than two years.
-Thus, the reason for New York’s and Massachusetts’ low CVD contributions does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.
+By 2013, the northeastern states of Massachusetts, Rhode Island, New York, and Connecticut had eliminated or nearly eliminated their black-white differences in life expectancy for both men and women.
+Each of these states had smaller contributions of each cause of death to the gap compared to the overall average, however, their contributions for CVD were most markedly lower.
+Notably, the contribution of CVD to the gap in these states has been lower than the overall average since 1969, reflecting that their low CVD contributions are not a consequence of recent changes, but rather a long-standing pattern of greater racial equality in CVD mortality.
+This points to the importance of viewing the most recent estimates in light of historical patterns.
+Kentucky, Nevada, New Mexico, and Washington state also performed better than the nation overall, with almost all causes of death contributing less than the overall average for both men and women.
+
+The rustbelt states of Illinois, Wisconsin, and Michigan, alongside DC had the highest black-white gaps for both men and women.
+These states all had much higher contributions of CVD to the black-white life expectancy gap.
+Injuries also made large contributions to black-white gaps in these states and DC, but for men only.
+Cancer and All other causes also played a role in Wisconsin and especially in DC, for which the black-white gap in 2013 was far higher than any other state.
+
+For women, several southern states made significant progress towards narrowing the black-white life expectancy gap.
+For example, Kentucky, Georgia, Alabama, and West Virginia had the smallest black-white gaps among southern states in 2013 for women, all less than two years.
+While injuries were not a large contributor to the black-white gap for women, its underlying race-specific mortality trends are concerning.
+Beginning in the mid ‘90s, trends in injury mortality increased for white women and decreased for black women, implying that a portion of the progress towards narrowing the black-white difference in several states is due to increasing injury mortality among white women.
+This likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic (Rudd et al., 2016).
+
+The racial gap in life expectancy is now smaller in many states in the South Atlantic and the East South Central census divisions than in the Midwest (e.g., less than 2.5 years in South Carolina, Delaware, Florida, Mississippi, North Carolina, Maryland, Tennessee, West Virginia, Alabama, Georgia, Kentucky, vs.
+greater than four years in Michigan, Illinois, Kanses, Iowa, Nebraska, and Wisconsin for women).
+This regional comparison is of particular interest because of the Great Migration between 1916 and 1970, during which a substantial number of blacks moved from the rural south to the urban north (Lemann et al., 1991).
+Generally speaking, migration to areas with increased economic prospects has been linked to raising health in the recipient locations and decreasing health in the origin locations because migrants are generally healthier than non-migrants (Normal et al., 2005).
+However, an econometric analysis has linked the Great Migration to increased mortality among migrants (Black et al., 2015).
+Thus, it may be that the heightened inequalities in rustbelt states partially reflects elevated black mortality associated with migration, though the Great Migration cohorts are less reflected in the later time period.
+
+Communicable disease continues to contribute to the gap in most states, but especially in DC, New Jersey, and Florida.
+In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 overall, (Harper et al., 2012) suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV.
+Cancer is also responsible for large contributions to the life expectancy gap in some states (DC, Wisconsin, Illinois) but not in others (New York, Massachusetts, Rhode Island), with most states roughly following the overall pattern.
 
@@
 Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
-This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^19,20^ The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum.
-In contrast to some recent work focused on county-level inequalities,^21,15^ states have some substantial advantages as analysis units; They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities with divergent social environments that are reflected in varying patterns of mortality by race.
+This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.24,25 The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum.
+In contrast to some recent work focused on county-level inequalities,26,27 states have some substantial advantages as analysis units.
+They have large populations to generate stable estimates, and tend to be more demographically stable over time.
 
@@
 Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years.
-Our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results for the overall life expectancy gap.^8^ Had Hispanics been excluded from this study, the life expectancy estimates for both blacks and whites are expected to be lower, since Hispanics are known to have lower mortality than blacks and whites.^22^ Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
-US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.^23^ Thus, the black-white gap in a state may decrease as more immigrants identify as black.
+In order to investigate the potential impact of Hispanic ethnicity, we inspected mortality rates in five states with large Hispanic populations (California, New York, Florida, Texas, and Illinois).
+Excluding Hispanics tended to shift the mortality rates up for both blacks and whites (Supplementary Appendix 3), since Hispanics overall tend to have lower mortality rates than non-Hispanics.28 New York was the exception: excluding Hispanics shifted mortality upwards for blacks but not for whites.
+This is difficult to interpret without further study, but may reflect heterogeneity in health among Hispanic subpopulations that have distinct migration patterns across US states.29,30,31 The other exception we noted was with injuries: injury mortality rates for all whites vs.
+non-Hispanic whites diverged since 1999, and may reflect the lower overdose mortality among Hispanics.20 Our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results for the overall life expectancy gap.4 Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
+US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.32 Thus, the black-white gap in a state may decrease as more immigrants identify as black.
 Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death.
 Migration is particularly important for the District of Columbia, because of its small geographical bounds.
-Affluent black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
-Both changes would serve to increase the racial gap in DC.^24^ Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
-For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.
-Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^25,26^
+Affluent black residents have moved to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
+Both changes would serve to increase the racial gap in DC.33 Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
+For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.34 Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.35,36
 
 **Implications for Policy & Future Work**
-
-The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.
-States pursue varying approaches to an array of relevant programs and laws including tobacco control,^27^ Medicare coverage,^28^ hand-gun violence,^29^ and nutrition.^30^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^31^ The analyses reported here are entirely descriptive, but could be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
-Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.^32^
+Given the scope of this estimation exercise, we did not attempt to make causal inferences between the changes in mortality and black-white inequality to changes that occurred in policy, migration, health care access, or the multitude of other characteristics that vary on the state level.
+Making connections between changes in state policy and changes in health outcome is an important next step in this research.
+States have pursued varying approaches to an array of relevant programs and laws including tobacco control,37 Medicare coverage,38 hand-gun violence,39 and nutrition40, to name a few examples.
+These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.41 The analyses reported here are entirely descriptive, but could be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
+Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.42
 

### 2018-05-11 Manuscript/Manuscript_COD_Trends_HealthandPlace.Rmd
finished manu update

@@
 title: Trends in the contribution of major causes of death to the black-white life expectancy gap by US state
-bibliography: bibliography.bib
 output:
@@
 
-We used data over 45 years, across two racial groups, and for all ages.
+We used data over forty-five years, across two racial groups, and for all ages.
 By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year.
@@
 Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages.
-This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.24,25 The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum.
-In contrast to some recent work focused on county-level inequalities,26,27 states have some substantial advantages as analysis units.
+This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group (Case and Deaton, 2015; Minton et al., 2016).
+The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of deaths in another stratum.
+In contrast to some recent work focused on county-level inequalities,(Dwyer-Lindgren et al., 2016; Arora et al., 2016) states have some substantial advantages as analysis units.
 They have large populations to generate stable estimates, and tend to be more demographically stable over time.
@@
 In order to investigate the potential impact of Hispanic ethnicity, we inspected mortality rates in five states with large Hispanic populations (California, New York, Florida, Texas, and Illinois).
-Excluding Hispanics tended to shift the mortality rates up for both blacks and whites (Supplementary Appendix 3), since Hispanics overall tend to have lower mortality rates than non-Hispanics.28 New York was the exception: excluding Hispanics shifted mortality upwards for blacks but not for whites.
-This is difficult to interpret without further study, but may reflect heterogeneity in health among Hispanic subpopulations that have distinct migration patterns across US states.29,30,31 The other exception we noted was with injuries: injury mortality rates for all whites vs.
-non-Hispanic whites diverged since 1999, and may reflect the lower overdose mortality among Hispanics.20 Our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results for the overall life expectancy gap.4 Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
-US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.32 Thus, the black-white gap in a state may decrease as more immigrants identify as black.
+Excluding Hispanics tended to shift the mortality rates up for both blacks and whites (Supplementary Appendix 3), since Hispanics overall tend to have lower mortality rates than non-Hispanics (Ruiz et al., 2013).
+New York was the exception: excluding Hispanics shifted mortality upwards for blacks but not for whites.
+This is difficult to interpret without further study, but may reflect heterogeneity in health among Hispanic subpopulations that have distinct migration patterns across US states (Borrell, 2005; Borrell and Lancet, 2012; Pew Research Center, 2018).
+The other exception we noted was with injuries: injury mortality rates for all whites vs.
+non-Hispanic whites diverged since 1999, and may reflect the lower overdose mortality among Hispanics (Rudd et al., 2016).
+Our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results for the overall life expectancy gap (Haper et al., 2014).
+Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states.
+US immigrants have been shown to have life expectancies at birth that were approximately five years longer in 2010 (Preston and Elo, 2014).
+Thus, the black-white gap in a state may have decreased as more immigrants identified as black.
 Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death.
@@
 Affluent black residents have moved to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC.
-Both changes would serve to increase the racial gap in DC.33 Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
-For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.34 Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.35,36
+Both changes would serve to increase the racial gap in DC (U.S.
+Census Burearu, 2017b).
+Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes.
+For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time (Bastian et al., 2017).
+Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses (Kuehn, 2014; Jones et al., 2010).
 
-**Implications for Policy & Future Work**
+**Implications for policy and future work**
 Given the scope of this estimation exercise, we did not attempt to make causal inferences between the changes in mortality and black-white inequality to changes that occurred in policy, migration, health care access, or the multitude of other characteristics that vary on the state level.
-Making connections between changes in state policy and changes in health outcome is an important next step in this research.
-States have pursued varying approaches to an array of relevant programs and laws including tobacco control,37 Medicare coverage,38 hand-gun violence,39 and nutrition40, to name a few examples.
-These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.41 The analyses reported here are entirely descriptive, but could be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
-Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.42
+Making connections between changes in state policy and changes in health outcomes is an important next step in this research.
+States have pursued varying approaches to an array of relevant programs and laws including tobacco control, (Cotti et al., 2016) Medicare coverage, (Grogan and Park, 2017) hand-gun violence, (Humphreys et al., 2017) and nutrition, (Taber et al., 2011) to name a few examples.
+These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks (Adimora et al., 2014).
+The analyses reported here are entirely descriptive, but could be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.
+Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations (Strumpf et al., 2017).
 
@@
 1.
-Cunningham TJ, Croft JB, Liu Y, Lu H, Eke PI, Giles WH (2017) Vital Signs: Racial Disparities in Age-Specific Mortality Among Blacks or African Americans - United States, 1999-2015.
-*MMWR Morb Mortal Wkly Rep* 66(17):444-456.
+Cunningham TJ, Croft JB, Liu Y, Lu H, Eke PI, Giles WH (2017) Vital signs: Racial disparities in age-specific mortality among Blacks or African Americans - United States, 1999-2015.
+*MMWR Morb.
+Mortal.
+Wkly Rep.* 66(17):444-456.
 
-2. Fuchs, VR (2016) Black Gains in Life Expectancy. *JAMA*. 316(18): 1869–70.
+2. Fuchs, VR (2016) Black gains in life expectancy. *JAMA*. 316(18): 1869–70.
 
@@
 
-4. Harper S, MacLehose RF, Kaufman JS (2014) Trends in the Black-White Life Expectancy Gap Among Us States, 1990–2009. *Health Aff* 33(8): 1375–82.
+4. Harper S, MacLehose RF, Kaufman JS (2014) Trends in the black-white life expectancy gap among US states, 1990–2009. *Health Aff.* 33(8): 1375–82.
 
-5. National Cancer Institute. Surveillance, Epidemiology, and End Results
+5.	Centers for Disease Control and Prevention.
+Office on Smoking and Health, National Center for Chronic Disease Prevention and Health Promotion: Excise tax rates on cigarettes [Internet].
+[cited 2018 March 22].
+Available from: https://www.cdc.gov/statesystem/excisetax.html.
+
+6.	Trust for America’s Health and Robert Wood Johnson Foundation. State policies to prevent obesity [Internet]. [cited 2018 March 22]. Available from: https://stateofobesity.org/state-policy/
+
+7.	California Air Resources Board.
+CARB finds vehicle standards are achievable and cost effective [Internet].[cited 2018 March 22].
+Available from: https://www.arb.ca.gov/newsrel/newsrelease.php?id=908
+
+8. National Cancer Institute. Surveillance, Epidemiology, and End Results
 Program: SEER*Stat software: release: version 8.3.2 [Internet]. Bethesda (MD): NCI; 2017 Apr 14 [cited 2016 August 20]. Available from: http://www.seer.cancer.gov/seerstat.
 
-6. U.S. Census Bureau: United States Census 2010 Interactive Population Map [Internet]. [cited 2017 May 16]. Available from: https://www.census.gov/2010census/popmap/.
+9. U.S. Census Bureau, 2017a. United States Census 2010 Interactive Population Map [Internet]. [cited 2017 May 16]. Available from: https://www.census.gov/2010census/popmap/.
 
-7.
+10.
 Centers for Disease Control and Prevention.
@@
 
-8.
+11.	National Cancer Institute.
+Surveillance, Epidemiology, and End Results Program: Population Estimates used in the NCI’s SEER*Stat Software [Internet].
+[cited 2018 March 22].
+Available from: https://seer.cancer.gov/popdata/methods.html.
+
+12.	Ingram D.D., Parker J.D., Schenker N., Weed J.A., Hamilton B., Arias E., Madans J.H.
+United States Census 2000 population with bridged race categories.
+Natl.
+Cent.
+Health Stat.
+Vital Health Stat 2(135).
+2003.
+
+13.	Harper S., Rushani D., and Kaufman J.S., 2012. Trends in the black-white life expectancy gap, 2003-2008. JAMA 307(21):2257–9.
+
+14.
 Arias E, Schauman WS, Eschbach K, Sorlie PD, Backlund E (2008) The validity of race and Hispanic origin reporting on death certificates in the United States.
-National Center for Health Statistics.
-Vital Health Stat 2(148).
+Natl.
+Cent.
+Health Stat.
+Vital.
+Health Stat.
+2(148).
 Available from: https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf
 
-9.
+15.
 National Cancer Institute.
@@
 
-10. Preston SH, Heuveline P, and Guillot M (2001) Demography: Measuring and Modeling Population Processes. (Blackwell Publishing, Malden)
+16. Preston SH, Heuveline P, and Guillot M (2001) Demography: Measuring and modeling population processes. (Blackwell Publishing, Malden)
 
-11. Selvin, S (2004) Statistical Analysis of Epidemiologic Data. 3rd ed. (Oxford University Press, Oxford)
+17. Selvin, S (2004) Statistical Analysis of Epidemiologic Data. 3rd ed. (Oxford University Press, Oxford)
 
-12. Arriaga EE (1985) Measuring and explaining the change in life expectancies. *Demography* 21(1):83-96.
+18. Arriaga EE (1985) Measuring and explaining the change in life expectancies. *Demography* 21(1):83-96.
 
-13.
-National Cancer Institute.
-Surveillance, Epidemiology, and End Results Program: Standard Populations - 19 Age Groups [Internet].
+19.
+National Cancer Institute, 2917b.
+Surveillance, Epidemiology, and End Results Program: Standard populations - 19 age groups [Internet].
 [cited 2017 April 25].
@@
 
-14. Lemann, N (1991) The Promised Land: The Great Black Migration and How It Changed America. New York: Alfred A. Knopf, ISBN 0-394-56004-3.
+20. Rudd RA, Seth P, David F, Scholl L (2016) Increases in drug and opioid-involved overdose deaths — United States, 2010–2015. *MMWR Morb. Mortal. Wkly Rep.* 65:1445–1452.
+
+21.	Norman P., Boyle P., Rees P., 2005. Selective migration, health and deprivation: a longitudinal analysis. Social Sci. Med. 60(12):2755-71.
+
+22.	Black D.A., Sanders S.G., Taylor E.J., Taylor L.J., 2015. The impact of the Great Migration on mortality of African Americans: evidence from the deep south. Am. Econ. Rev. 105(2):477-503.
+
+14. Lemann, N (1991) The promised land: The great black migration and how it changed America. New York: Alfred A. Knopf, ISBN 0-394-56004-3.
 
@@
 
-16. Case A, Deaton A (2017) Mortality and Morbidity in the 21st Century. In: Eberly J and Stock J, Editors. Brookings Papers on Economic Activity, March 23-24 2017.
-
-17. Harper S, Rushani D, and Kaufman JS (2012) Trends in the Black-White Life Expectancy Gap, 2003-2008. *JAMA* 307(21):2257–9.
-
-18. Rudd RA, Seth P, David F, Scholl L (2016) Increases in Drug and Opioid-Involved Overdose Deaths — United States, 2010–2015. *MMWR Morb Mortal Wkly Rep* 65:1445–1452.
-
-19. Case A, Deaton A (2015) Rising Morbidity and Mortality in Midlife among White Non-Hispanic Americans in the 21st Century. *Proc Natl Acad Sci * 112(49):15078-83.
+19. Case A, Deaton A (2015) Rising morbidity and mortality in midlife among white non-Hispanic Americans in the 21st century. *Proc Natl Acad Sci * 112(49):15078-83.
 
@@
 Why we need better ways of seeing data: A commentary on:‘Rising morbidity and mortality in midlife among White non-Hispanic Americans in the 21st century’.
-*Int J Epidemiol* 2016 Sep 17:dyw095.
+*Int.
+J.
+Epidemiol.* 2016 Sep 17:dyw095.
 
 21.
-Arora A, Spatz E, Herrin J, Riley C, Roy B, Kell K, Coberley C, Rula E, Krumholz HM (2016) Population Well-Being Measures Help Explain Geographic Disparities In Life Expectancy At The County Level.
-*Health Aff* 35(11):2075-82.
+Arora A, Spatz E, Herrin J, Riley C, Roy B, Kell K, Coberley C, Rula E, Krumholz HM (2016) Population well-being measures help explain geographic disparities in life expectancy at the county level.
+*Health Aff.* 35(11):2075-82.
 
@@
 
-23. Preston SH, Elo IT (2014) Anatomy of a municipal triumph: New York City's upsurge in life expectancy. *Population and development review* 40(1):1-29.
+23.	Borrell L.N., 2005. Racial identity among Hispanics: implications for health and well-being. Am. J. Public Health. 95(3):379-381.
 
-24.
+24.	Borrell L.N., Lancet E.A., 2012. Race/ethnicity and all-cause mortality in US adults: revisiting the Hispanic paradox. Am. J. of Public Health. 102(5):836-43.
+
+25.	U.S. Hispanic country of origin for nation, top 30 metropolitan areas [Internet]. [Cited April 11, 2018]. Available from: http://www.pewhispanic.org/2011/05/26/ii-metropolitan-area-diversity/
+
+26. Preston SH, Elo IT (2014) Anatomy of a municipal triumph: New York City's upsurge in life expectancy. *Population and development review* 40(1):1-29.
+
+27.
 U.S.
@@
 
-25. Kuehn BM (2014) CDC: major disparities in opioid prescribing among states: some states crack down on excess prescribing. *JAMA* 312(7):684-6.
+28.	Bastian B., Lu L., Miniño A., et al. Injury mortality, United States: 1999–2015. National Center for Health Statistics. National Vital Statistics System. 2017.
 
-26. Jones CM, Mack KA, Paulozzi LJ (2010) Pharmaceutical overdose deaths, United States, 2010. *JAMA* 309(7):657-9.
+29. Kuehn BM (2014) CDC: major disparities in opioid prescribing among states: some states crack down on excess prescribing. *JAMA* 312(7):684-6.
 
-27.
-Cotti C, Nesson E, Tefft N (2016) The Effects of Tobacco Control Policies on Tobacco Products, Tar, and Nicotine Purchases among Adults: Evidence from Household Panel Data.
-*American Economic Journal: Economic Policy* 8(4):103-23.
+30. Jones CM, Mack KA, Paulozzi LJ (2010) Pharmaceutical overdose deaths, United States, 2010. *JAMA* 309(7):657-9.
 
-28. Grogan CM, Park SE (2017) The Racial Divide in State Medicaid Expansions. *J Health Polit Policy Law* 42(3):539-72.
+31.
+Cotti C, Nesson E, Tefft N (2016) The effects of tobacco control policies on tobacco products, tar, and nicotine purchases among adults: Evidence from household panel data.
+*Am.
+Econ.
+J.: Econ.
+Policy* 8(4):103-23.
 
-29.
-Humphreys DK, Gasparrini A, Wiebe DJ (2017) Evaluating the Impact of Florida’s “Stand Your Ground” Self-defense Law on Homicide and Suicide by Firearm: An Interrupted Time Series Study.
-*JAMA Intern Med* 177(1):44-50.
+32. Grogan CM, Park SE (2017) The racial divide in state medicaid expansions. *J Health Polit Policy Law* 42(3):539-72.
 
-30.
+33.
+Humphreys DK, Gasparrini A, Wiebe DJ (2017) Evaluating the impact of Florida’s “stand your ground” self-defense law on homicide and suicide by firearm: An interrupted time aeries atudy.
+*JAMA Intern.
+Med.* 177(1):44-50.
+
+34.
 Taber DR, Stevens J, Evenson KR, Ward DS, Poole C, Maciejewski ML, Murray DM, Brownson RC (2011) State policies targeting junk food in schools: racial/ethnic differences in the effect of policy change on soda consumption.
-*Am J Public Health* 101(9):1769-75.
+*Am.
+J.
+Public Health* 101(9):1769-75.
 
-31. Adimora AA, Ramirez C, Schoenbach VJ, Cohen MS (2014) Policies and politics that promote HIV infection in the Southern United States. *AIDS*
+35. Adimora AA, Ramirez C, Schoenbach VJ, Cohen MS (2014) Policies and politics that promote HIV infection in the southern United States. *AIDS*
 28(10):1393.
 
-32. Strumpf EC, Harper S, Kaufman JS (2017) Fixed Effects and Difference in Differences. In: Methods in Social Epidemiology. 2nd Ed, eds Oakes M, Kaufman JS(Jossey-Bass, San Francisco).
+36. Strumpf EC, Harper S, Kaufman JS (2017) Fixed effects and difference in differences. In: Methods in Social Epidemiology. 2nd ed, eds Oakes M, Kaufman JS (Jossey-Bass, San Francisco).
 
@@
 
-A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on Github: https://github.com/corinne-riddell/BlackWhiteMortalityGap/ (DOI: 10.5281/zenodo.803472).
+A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on GitHub: https://github.com/corinne-riddell/BlackWhiteMortalityGap/ (DOI: 10.5281/zenodo.803472).

### 2018-05-11 Manuscript/Manuscript_COD_Trends_HealthandPlace.Rmd
format rmd

@@
 output: 
-  word_document: default
-  html_document: default
-always_allow_html: yes
+  word_document:
+      reference_docx: reference.docx
 ---
@@
 Phone: +1-438-863-0402
+
+##### PAGE BREAK
 
@@
 
-----
+##### PAGE BREAK
 
@@
 
+##### PAGE BREAK
+
 **Figure 1A: State-level trends in the black-white life expectancy gap in men, United States, 1969-2013**
@@
 Thus the data for this state and year should be interpreted with caution.
+
+##### PAGE BREAK
 
@@
 
+##### PAGE BREAK
+
 **Figure 2A: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
@@
 
+##### PAGE BREAK
+
 **Figure 3A: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
@@
 The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states.
+
+##### PAGE BREAK
 
@@
 
+##### PAGE BREAK
+
 **Figure 4A: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
@@
 
+##### PAGE BREAK
+
 **Figure 5A: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
@@
 
+##### PAGE BREAK
+
 **Figure 6A: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
@@
 
+##### PAGE BREAK
+
 **Figure 7A: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**
@@
 Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations (Strumpf et al., 2017).
+
+**Contributors**
+
+CAR: Extracted the data, performed data analysis, drafted the manuscript.
+
+KTM: Performed data analysis, reviewed the paper for intellectual content.
+
+SH: Initial conception for the study, interpreted the results, reviewed the paper for intellectual content.
+
+JK: Initial conception for the study, interpreted the results, drafted the manuscript, reviewed the paper for intellectual content.
+
+We extend our thanks to Dr.
+Richard Maclehose, who developed the initial version of the code to perform the Bayesian analysis as part of the research article “Trends in the Black-White Life Expectancy Gap Among Us States, 1990–2009”, by Harper S, MacLehose RF, and Kaufman JS.
+
+**Funding**
+
+This research received no specific grant from any funding agency in the public, commercial or not-for-profit sectors.
+
+Corinne Riddell is a post-doctoral researcher at McGill University, whose salary was supported by Dr.
+Kaufman's Canada Research Chair in Health Disparities.
+Kathryn Morrison is a post-doctoral researcher at McGill University.
+Sam Harper was supported by a Chercheur Boursier Junior 2 from the Fonds de la Recherche en Santé du Québec.
+Jay Kaufman was supported by a Canada Research Chair in Health Disparities.
+
+**Competing Interests**
+
+All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare: no support from any organisation for the submitted work; no financial relationships with any organisations that might have an interest in the submitted work in the previous three years; no other relationships or activities that could appear to have influenced the submitted work.”
+
+**Data Sharing Statement**
+
+A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on GitHub: https://github.com/corinne-riddell/BlackWhiteMortalityGap/ (DOI: 10.5281/zenodo.803472).
+
+##### PAGE BREAK
 
@@
 36. Strumpf EC, Harper S, Kaufman JS (2017) Fixed effects and difference in differences. In: Methods in Social Epidemiology. 2nd ed, eds Oakes M, Kaufman JS (Jossey-Bass, San Francisco).
-
-###Footnotes
-
-**Contributors**
-
-CAR: Extracted the data, performed data analysis, drafted the manuscript.
-
-KTM: Performed data analysis, reviewed the paper for intellectual content.
-
-SH: Initial conception for the study, interpreted the results, reviewed the paper for intellectual content.
-
-JK: Initial conception for the study, interpreted the results, drafted the manuscript, reviewed the paper for intellectual content.
-
-We extend our thanks to Dr.
-Richard Maclehose, who developed the initial version of the code to perform the Bayesian analysis as part of the research article “Trends in the Black-White Life Expectancy Gap Among Us States, 1990–2009”, by Harper S, MacLehose RF, and Kaufman JS.
-
-**Funding**
-
-This research received no specific grant from any funding agency in the public, commercial or not-for-profit sectors.
-
-Corinne Riddell is a post-doctoral researcher at McGill University, whose salary was supported by Dr.
-Kaufman's Canada Research Chair in Health Disparities.
-Kathryn Morrison is a post-doctoral researcher at McGill University.
-Sam Harper was supported by a Chercheur Boursier Junior 2 from the Fonds de la Recherche en Santé du Québec.
-Jay Kaufman was supported by a Canada Research Chair in Health Disparities.
-
-**Competing Interests**
-
-All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare: no support from any organisation for the submitted work; no financial relationships with any organisations that might have an interest in the submitted work in the previous three years; no other relationships or activities that could appear to have influenced the submitted work.”
-
-**Data Sharing Statement**
-
-A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on GitHub: https://github.com/corinne-riddell/BlackWhiteMortalityGap/ (DOI: 10.5281/zenodo.803472).

### 2018-05-11 Manuscript/Appendix-1-agestd-mort-rates.Rmd
update appendices

--- 2018-05-10 Manuscript/Appendix-agestd-mort-rates.Rmd
@@
 output: 
-  word_document: default
-  html_document: default
+  word_document:
+      reference_docx: reference.docx
 ---
 
-This is the supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state".
+This is supplementary appendix #1 for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state".
 
-For an interactive version of all our results, please visit XXXX.
+Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Jay S Kaufman, PhD; Sam Harper, PhD
+
+Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada.
+Email: <corinne.riddell@mail.mcgill.ca>
+
+For an interactive version of all our results, please visit https://corinne-riddell.shinyapps.io/black-white-life-expectancy/.
 
 ```{r load_libraries, echo=FALSE, warning=F, message=F}
-#Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Sam Harper, PhD; Jay S Kaufman, PhD
-
-#Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, #1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada.
-Email: <corinne.riddell@mail.mcgill.ca>
-
 library(dplyr)
@@
 
+#####
 **Table S1: Cause of death groupings**
@@
 
-**Figure S3: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**
+#####PAGE BREAK
+
+**Figure S1: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**
 
@@
 
-**Figure S4: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**
+#####PAGE BREAK
+
+**Figure S2: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**
 
@@
 
-**Figure S5: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**
+#####PAGE BREAK 
+**Figure S3: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**
 
@@
 
-**Figure S6: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**
+#####PAGE BREAK
+**Figure S4: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**
 
@@
 
-**Figure S7: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**
+#####PAGE BREAK
+**Figure S5: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**
 
@@
 
-**Figure S8: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**
+#####PAGE BREAK
+**Figure S6: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**
 
@@
 
-**Figure S9: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**
+#####PAGE BREAK
+**Figure S7: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**
 
@@
 
-**Figure S10: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**
+#####PAGE BREAK
+**Figure S8: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**
 
@@
 
-**Figure S11: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**
+#####PAGE BREAK
+**Figure S9: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**
 
@@
 
-**Figure S12: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**
+#####PAGE BREAK
+**Figure S10: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**
 
@@
 
-**Figure S13: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**
+#####PAGE BREAK
+**Figure S11: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**
 
@@
 
-**Figure S14: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**
+#####PAGE BREAK
+**Figure S12: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**
 
@@
 
-**Figure S15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in males younger than one, 1969-2013**
+#####PAGE BREAK
+**Figure S13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in males infants (<1 year old), 1969-2013**
 
@@
 
-**Figure S16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in females younger than one, 1969-2013**
+#####PAGE BREAK
+**Figure S14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in female infants (<1 year old), 1969-2013**
 

### 2018-05-11 Manuscript/Appendix-2-compare-COD-gender-within-state.Rmd
update appendices

--- 2018-03-23 Manuscript/Appendix-compare-COD-gender-within-state.Rmd
@@
 title: "Supplementary Appendix 2: Trend in the difference between each state's contribution trend and the overall trend, by causes of death and gender, 1969-2013"
-output: word_document
+output: 
+  word_document:
+      reference_docx: reference.docx
 ---
 
-This is the second supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state.
+This is the second supplementary appendix #2 for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state.
 A retrospective study of all deaths between 1969 and 2013."

### 2018-05-11 Manuscript/Appendix-3-Explore-Hispanic-origin.Rmd
update appendices

--- 2018-04-10 Manuscript/Appendix_Explore-Hispanic-origin.Rmd
@@
 title: "Supplementary Appendix 3: Trends in black and white mortality rates before and after excluding Hispanics, 1999-2013"
-output: github_document
+output: 
+  word_document:
+      reference_docx: reference.docx
 ---
 
-This is the third supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state.
+This is the third supplementary appendix #3 for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state.
 A retrospective study of all deaths between 1969 and 2013."
@@
 
+#####
 In our main analyses, we included Hispanics as part of the black and white racial groups.
@@
 
+#####
 **Trends in cancer mortality for all whites and blacks vs. non-Hispanic whites and blacks, 1999-2013**
-```{r load-cancer-data, echo=F, fig.width = 8, fig.height=9.5}
+```{r load-cancer-data, echo=F, fig.width = 8, fig.height=9.5, warning=F, message=F}
 library(tidyverse)

### 2019-07-15 Manuscript2/Supplementary-appendix.Rmd
supplementary appendix files

---
title: "Supplementary Appendix"
output:
  word_document:
      reference_docx: reference.docx
---

# Supplementary Appendix to: Black and white life expectancy differences in four US states: 1969-2013

By: Jay S Kaufman, Corinne A Riddell, Sam Harper

##### page break

Table of Contents

Table S1: Description of specific causes comprising the six major cause categories

Figure S1: Age-specific trends in mortality due to all other causes in black and white women, all age groups, 1969-2013

Figure S2: Age-specific trends in CVD mortality in black and white women, 20 years and older, 1969-2013

Figure S3: Age-specific trends in cancer mortality in black and white women, 20 years and older, 1969-2013

Figure S4: Age-specific trends in mortality due to non-communicable disease in black and white women, 20 years and older, 1969-2013

Figure S5: Age-specific trends in communicable disease mortality in black and white women, all age groups, 1969-2013

Figure S6: Age-specific trends in injury mortality in black and white women, all age groups, 1969-2013

Figure S7: Age-specific trends in mortality due to all other causes in black and white men, all age groups, 1969-2013

Figure S8: Age-specific trends in CVD-related mortality in black and white men, 20 years and older, 1969-2013

Figure S9: Age-specific trends in cancer mortality in black and white men, 20 years and older, 1969-2013

Figure S10: Age-specific trends in mortality due to non-communicable disease in black and white men, 20 years and older, 1969-2013

Figure S11: Age-specific trends in communicable disease mortality in black and white men, all age groups, 1969-2013

Figure S12: Age-specific trends in injury mortality in black and white men, all age groups, 1969-2013

Figure S13: The contribution of major causes of death and age at death to the black-white life expectancy gap in women, 2013

Figure S14: The contribution of major causes of death and age at death to the black-white life expectancy gap in men, 2013

Figure S15: Total population sizes in 4 states by race and year

##### page break

**Table S1: Description of specific causes comprising the six major cause categories**

| Broad cause of death | Included causes of death                               |
|----------------------|--------------------------------------------------------|
|Cancers               | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer.                         |
|Cardiovascular disease               |  Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries.                        |
|Communicable disease               |  Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza                        |
|Non-communicable disease               |   Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Syndrome and Nephrosis.                       |
|Injuries               | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention                         |
|All other causes               | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death.                         |

##### page break

**Figure S1: Age-specific trends in mortality due to all other causes in black and white women, all age groups, 1969-2013**

```{r figure-s1, echo = F, message=F}
library(here)
knitr::include_graphics(here("Manuscript2", "supp_figures", "other_female.jpeg"))
```

**Table S1** lists the specific causes of death that contribute to the “all other causes” major categorization. The majority of these deaths are among infants, as shown by the large magnitude of mortality rates for deaths among those < 1 year in **Figure S1**. Thus, the first row of this figure provides a rough approximation to the overall rate of infant mortality. Among the other age groups, the exact cause is most often “Symptoms, signs, and ill-defined conditions”. Each plot shows three lines per race, which are often overlapping (except for 1-19 year olds). The line type indicates the imputation strategy for the suppressed death counts, where we set all suppressed counts to 1 (solid line, lower bound), 5 (dotted line, midpoint) or 9 (dashed line, upper bound).

##### page break

**Figure S2: Age-specific trends in CVD mortality in black and white women, 20 years and older, 1969-2013**

```{r figure-s2, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "cvd_female.jpeg"))
```

Not shown: <1 year olds, and 1-19 year olds, because there were very few deaths in those age bands.  
Each plot shows three lines per race, which are often overlapping (except for 1-19 year olds). The line type indicates the imputation strategy for the suppressed death counts, where we set all suppressed counts to 1 (solid line, lower bound), 5 (dotted line, midpoint) or 9 (dashed line, upper bound).

##### page break

**Figure S3: Age-specific trends in cancer mortality in black and white women, 20 years and older, 1969-2013**

```{r figure-s3, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "cancer_female.jpeg"))
```

Not shown: <1 year olds, and 1-19 year olds, because there were very few deaths in those age bands.  
Each plot shows three lines per race, which are often overlapping (except for 1-19 year olds). The line type indicates the imputation strategy for the suppressed death counts, where we set all suppressed counts to 1 (solid line, lower bound), 5 (dotted line, midpoint) or 9 (dashed line, upper bound).

##### page break

**Figure S4: Age-specific trends in mortality due to non-communicable disease in black and white women, 20 years and older, 1969-2013**

```{r figure-s4, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "nc_female.jpeg"))
```

Not shown: <1 year olds, and 1-19 year olds, because there were very few deaths in those age bands.  
Each plot shows three lines per race, which are often overlapping (except for 1-19 year olds). The line type indicates the imputation strategy for the suppressed death counts, where we set all suppressed counts to 1 (solid line, lower bound), 5 (dotted line, midpoint) or 9 (dashed line, upper bound).

##### page break

**Figure S5: Age-specific trends in communicable disease mortality in black and white women, all age groups, 1969-2013**

```{r figure-s5, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "comm_female.jpeg"))
```

Each plot shows three lines per race, which are often overlapping (except for 1-19 year olds). The line type indicates the imputation strategy for the suppressed death counts, where we set all suppressed counts to 1 (solid line, lower bound), 5 (dotted line, midpoint) or 9 (dashed line, upper bound).

##### page break

**Figure S6: Age-specific trends in injury mortality in black and white women, all age groups, 1969-2013**

```{r figure-s6, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "inj_female.jpeg"))
```

While injury rates are relatively low in women compared to men, the injury category is comprised of many distinct sources of injury including motor vehicle accidents, homicide, suicide, and poisonings, each of having their own distinct trends that are unfortunately masked in aggregation. Each plot shows three lines per race, which are often overlapping (except for 1-19 year olds). The line type indicates the imputation strategy for the suppressed death counts, where we set all suppressed counts to 1 (solid line, lower bound), 5 (dotted line, midpoint) or 9 (dashed line, upper bound).

##### page break

**Figure S7: Age-specific trends in mortality due to all other causes in black and white men, all age groups, 1969-2013**

```{r figure-s7, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "other_male.jpeg"))
```

Each plot shows three lines per race, which are often overlapping (except for 1-19 year olds). The line type indicates the imputation strategy for the suppressed death counts, where we set all suppressed counts to 1 (solid line, lower bound), 5 (dotted line, midpoint) or 9 (dashed line, upper bound).

##### page break

**Figure S8: Age-specific trends in CVD mortality in black and white men, 20 years and older, 1969-2013**

```{r figure-s8, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "cvd_male.jpeg"))
```

Not shown: <1 year olds, and 1-19 year olds, because there were very few deaths in those age bands.  
Each plot shows three lines per race, which are often overlapping (except for 1-19 year olds). The line type indicates the imputation strategy for the suppressed death counts, where we set all suppressed counts to 1 (solid line, lower bound), 5 (dotted line, midpoint) or 9 (dashed line, upper bound).

##### page break

**Figure S9: Age-specific trends in cancer mortality in black and white men, 20 years and older, 1969-2013**

```{r figure-s9, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "cancer_male.jpeg"))
```

Not shown: <1 year olds, and 1-19 year olds, because there were very few deaths in those age bands.  
Each plot shows three lines per race, which are often overlapping (except for 1-19 year olds). The line type indicates the imputation strategy for the suppressed death counts, where we set all suppressed counts to 1 (solid line, lower bound), 5 (dotted line, midpoint) or 9 (dashed line, upper bound).

##### page break

**Figure S10: Age-specific trends in mortality due to non-communicable disease in black and white men, 20 years and older, 1969-2013**

```{r figure-s10, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "nc_male.jpeg"))
```

Not shown: <1 year olds, and 1-19 year olds, because there were very few deaths in those age bands.  
Each plot shows three lines per race, which are often overlapping (except for 1-19 year olds). The line type indicates the imputation strategy for the suppressed death counts, where we set all suppressed counts to 1 (solid line, lower bound), 5 (dotted line, midpoint) or 9 (dashed line, upper bound).

##### page break

**Figure S11: Age-specific trends in communicable disease mortality in black and white men, all age groups, 1969-2013**

```{r figure-s11, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "comm_male.jpeg"))
```

Each plot shows three lines per race, which are often overlapping (except for 1-19 year olds). The line type indicates the imputation strategy for the suppressed death counts, where we set all suppressed counts to 1 (solid line, lower bound), 5 (dotted line, midpoint) or 9 (dashed line, upper bound).

##### page break

**Figure S12: Age-specific trends in injury mortality in black and white men, all age groups, 1969-2013**

```{r figure-s12, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "inj_male.jpeg"))
```

Each plot shows three lines per race, which are often overlapping (except for 1-19 year olds). The line type indicates the imputation strategy for the suppressed death counts, where we set all suppressed counts to 1 (solid line, lower bound), 5 (dotted line, midpoint) or 9 (dashed line, upper bound).

##### page break

**Figure S13: The contribution of major causes of death and age at death to the black-white life expectancy gap in women, 2013**

```{r figure-s13, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "age-cod-females.jpeg"))
```

##### page break

**Figure S14: The contribution of major causes of death and age at death to the black-white life expectancy gap in men, 2013**

```{r figure-s14, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "age-cod-males.jpeg"))
```

##### page break

**Figure S15: Total population sizes in 4 states by race and year**

```{r figure-s15, echo = F}
knitr::include_graphics(here("Manuscript2", "supp_figures", "population_plot.jpg"))
```

### 2019-07-15 Manuscript2/ageCOD_manuscript.Rmd
supplementary appendix files

---
title: Black and White Life Expectancy Differences in 4 US States, 1969-2013
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

Below is the code for the paper entitled "Black and White Life Expectancy Differences in 4 US States, 1969-2013".
The analysis for this paper builds on the analysis from our previous paper that is described in this repository.
To conduct these analysis, we import and manipulate tables of results from the previous paper and web application.

Load the libraries and a set of functions created by C Riddell:

```{r initialize, warning=F, message=F}
library(viridis)
library(scales)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
library(splines)
library(broom)
library(tidyverse)
library(forcats)
library(magrittr)
library(gridExtra)
library(here)
library(ggthemes)

source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

Load the results' tables from the previous paper and tidy these data for analysis/visualization:

```{r load_modelled_data}
included_states <- c("Georgia", "California", "New York", "Illinois")
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
dat.aggregated <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/dat_aggregated.csv")

cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results_cubic.csv")
cod_decomp_results <- reorder.as.map2(cod_decomp_results, "state", "stabbrs")
cod_decomp_results$COD_rev <- factor(cod_decomp_results$COD, levels = c("Cardiovascular", "Cancers", "All other causes",
                                                                        "Communicable", "Non-communicable",  "Injuries"))

age_cod_results_female <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_female.csv")
age_cod_results_male <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_male.csv")
age_cod_results <- rbind(age_cod_results_female, age_cod_results_male)
rm(age_cod_results_female, age_cod_results_male)
age_cod_results$COD <- factor(age_cod_results$COD, levels(age_cod_results$COD)[c(3, 2, 4, 6, 5, 1)])

age_cod_results %<>% mutate(age = fct_relevel(age,
                                              "<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                              "15-19 years", "20-24 years", "25-29 years", "30-34 years", "35-39 years", 
                                              "40-44 years", "45-49 years", "50-54 years", "55-59 years", "60-64 years", 
                                              "65-69 years", "70-74 years", "75-79 years", "80-84 years", "85+ years"))
```

Load mortality tables for the states of interest:

```{r load_raw_data }
#these data are used for the mortality rates shown in the appendix

dat.California <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/dat_clean_California.csv") 
dat.Georgia <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/dat_clean_Georgia.csv") 
dat.NewYork<- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/dat_clean_New York.csv") 
dat.Illinois <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Data/dat_clean_Illinois.csv")

dat.four <- rbind(dat.California, dat.Georgia, dat.Illinois, dat.NewYork)
rm(dat.California, dat.Georgia, dat.Illinois, dat.NewYork)
```

Take a look at population counts for these states, and the proportion of missing data:

```{r, eval=F}
#NOTE:
summary(dat.four$population)
#minimum population size is 1399 - lots of data for these states

prop.table(table(is.na(dat.four$deaths)))
#still have 19% suppressed data though.
```

Previously, the age groups were coded using 5-year age bands. We collapse this to wider bands as follows:

```{r mutate_age_collapse}
dat.four <- dat.four %>% 
  mutate(age_collapse = fct_collapse(age, "1-19 years" = c("1-4 years", "5-9 years", "10-14 years", "15-19 years"),
                                     "20-39 years" = c("20-24 years", "25-29 years", "30-34 years", "35-39 years"),
                                     "40-64 years" = c("40-44 years", "45-49 years", "50-54 years", "55-59 years", "60-64 years"),
                                     "65-84 years" = c("65-69 years", "70-74 years", "75-79 years", "80-84 years")))

#table(dat.four$age_collapse, dat.four$age) #check coding

dat.collapse.age <- dat.four %>% group_by(state.year.race.sex, age_collapse, COD) %>%
  arrange(state.year.race.sex, age_collapse, COD) %>%
  summarise(state = first(state),
            year = first(year),
            race = first(race),
            sex = first(sex),
    tot_deaths1_age_collapse = sum(deaths_md1),
            tot_deaths5_age_collapse = sum(deaths_md5),
            tot_deaths9_age_collapse = sum(deaths_md9),
            population_age_collapse = sum(population)) %>%
  mutate(crude_rate_age_collapse1 = (tot_deaths1_age_collapse/population_age_collapse)*100000,
         crude_rate_age_collapse5 = (tot_deaths5_age_collapse/population_age_collapse)*100000,
         crude_rate_age_collapse9 = (tot_deaths9_age_collapse/population_age_collapse)*100000)

```

The following code prepares the data for plotting.

```{r prep_COD_data}
excluded <- c("Alaska", "Hawaii", "Idaho", "Maine", "Montana", "New Hampshire", "North Dakota",
              "South Dakota", "Utah", "Vermont", "Wyoming")

dat.agg2.white <- dat.aggregated %>% 
  filter(age == "<1 year" & race == "White" & !(state %in% excluded)) %>% 
  select(state, year, sex, pop_across_age) %>% 
  rename(pop_white = pop_across_age)

dat.agg2.black <- dat.aggregated %>% 
  filter(age == "<1 year" & race == "Black" & !(state %in% excluded)) %>% 
  select(state, year, sex, pop_across_age) %>% 
  rename(pop_black = pop_across_age)

dat.agg2 <- merge(dat.agg2.black, dat.agg2.white, by = c("state", "year", "sex"))
rm(dat.agg2.black, dat.agg2.white)

cod_decomp_results <- merge(cod_decomp_results,
                            BlackWhite_results %>% select(stratum.id, LE_white_mean, LE_black_mean, LE_wbgap_mean), 
                            by = "stratum.id")

cod_decomp_results <- merge(cod_decomp_results, dat.agg2, by = c("state", "year", "sex"))

cod_decomp_results$new.start = cod_decomp_results$start + cod_decomp_results$LE_black_mean
cod_decomp_results$new.start2 = cod_decomp_results$start2 + cod_decomp_results$LE_black_mean
cod_decomp_results$new.finish = cod_decomp_results$finish + cod_decomp_results$LE_black_mean
cod_decomp_results$new.finish2 = cod_decomp_results$finish2 +cod_decomp_results$ LE_black_mean

cod_decomp_results$LE_black_mean[cod_decomp_results$COD != "Cardiovascular"] <- NA
cod_decomp_results$LE_white_mean[cod_decomp_results$COD != "Cardiovascular"] <- NA

cod_decomp_results <- reorder.as.map2(cod_decomp_results, "state", "stabbrs")
cod_decomp_results$COD <- factor(cod_decomp_results$COD, levels(cod_decomp_results$COD)[c(3, 2, 4, 6, 5, 1)])
```

Here is the code to create the first figure:

**Figure 1: Black-white life expectancy gap in four US states, by gender 1969-2013**

```{r, fig.height = 3, fig.width = 6}
BlackWhite_results$sex2 <- factor(BlackWhite_results$sex, levels = c("Female", "Male"), labels = c("Women", "Men"))

data <- BlackWhite_results %>% filter(state %in% included_states)
data$blah <- 1
data$blah[data$state %in% c("Georgia", "New York")] <- 2

labels <- data.frame(sex2 = c("Women", "Men"),
                     labs = c("Females", "Males"),
                     year = c(1976, 1976), LE_wbgap_mean = c(11.25, 11.25))

le.fig <- ggplot(data, aes(x = year, y = LE_wbgap_mean)) + 
  #geom_vline(xintercept=seq(1970,2010,5), colour="#f0f0f0") +
  geom_line(aes(col = state), lwd = 1.5) + #, lty = as.factor(blah)
  #scale_color_manual(values = c("#969696", "#969696", "#252525", "#252525")) +
  #theme_minimal() + 
  facet_wrap(~ sex2) + 
  xlab("Year") + 
  ylab("Black-white life expectancy gap, y") +
  scale_x_continuous(breaks = c(1969, 1980, 1990, 2000, 2013), 
                     labels = c(1969, 1980, 1990, 2000, 2013)) +
  scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12), 
                     labels = c(0, 2, 4, 6, 8, 10, 12), 
                     limits = c(0, 12)) + 
  #geom_hline(aes(yintercept = 0), lty =2) +
  coord_cartesian(ylim = c(0, 12), 
                    expand = FALSE) +
  scale_color_colorblind() +
  theme_bw() +
  theme(panel.grid.minor.x = element_blank(), 
        panel.grid.major.x = element_blank(),
        strip.background = element_blank(), #element_rect(fill="white"), 
        strip.text.x = element_blank(), #element_text(size = 12),
        strip.text.y = element_text(size = 12),
        legend.background = element_rect(colour = 'grey', fill = 'white', linetype='solid'),
        panel.spacing = unit(2, "lines")) +
  guides(col = guide_legend(title = element_blank())) +
  geom_text(data = labels, aes(x = year, y = LE_wbgap_mean, label = labs), check_overlap = T, hjust = "left")

le.fig
```

Some code to prepare to create Figure 2:

```{r, fig-2-prep}
age_cod_results <- age_cod_results %>% 
  mutate(age_collapse = fct_collapse(age, "1-19 years" = c("1-4 years", "5-9 years", "10-14 years", "15-19 years"),
                                     "20-39 years" = c("20-24 years", "25-29 years", "30-34 years", "35-39 years"),
                                     "40-64 years" = c("40-44 years", "45-49 years", "50-54 years", "55-59 years", "60-64 years"),
                                     "65-84 years" = c("65-69 years", "70-74 years", "75-79 years", "80-84 years")))

age_cod_results <- age_cod_results %>% group_by(state, year, sex, COD, age_collapse) %>%
  mutate(age_COD_cont_yrs_mean_collapse = sum(age_COD_cont_yrs_mean))
```

**Figure 2: Contribution of age and cause categories to the black-white life expectancy gap in women, 1969-2013**

```{r, fig-2-code, fig.height = 11, fig.width = 9, echo=2}
labels_women <- data.frame(state = c(rep("California", 6), rep("Georgia", 6), rep("Illinois", 6), rep("New York", 6)),
                           age_collapse2 = rep(c("<1 year", "1-19 years", "20-39 years", "40-64 years", "65-84 years", "\u2265 85 years"), 4),
                           year = rep(1970, 24), age_COD_cont_yrs_mean_collapse = rep(2.25, 24))

age_cod_results <- age_cod_results %>% mutate(age_collapse2 = case_when(age_collapse == "85+ years" ~  "\u2265 85 years",
                                                                        age_collapse == "<1 year" ~ "<1 year",
                                                                        age_collapse == "1-19 years" ~ "1-19 years",
                                                                        age_collapse == "20-39 years" ~ "20-39 years",
                                                                        age_collapse == "40-64 years" ~ "40-64 years",
                                                                        age_collapse == "65-84 years" ~ "65-84 years"))

table(age_cod_results$age_collapse2)

age_cod_results$age_collapse2 <- forcats::fct_relevel(age_cod_results$age_collapse2,
                                                      "<1 year", "1-19 years", "20-39 years", "40-64 years", "65-84 years", "\u2265 85 years")

to_plot <- age_cod_results %>% ungroup() %>%
         filter(state %in% included_states, sex == "Female") %>% 
  arrange(state, year) %>% 
  group_by(state, year, sex, COD, age_collapse)

women.fig <- ggplot(to_plot, 
       aes(x = year, y = age_COD_cont_yrs_mean_collapse)) + 
  facet_grid(age_collapse2 ~ state) + 
  geom_vline(xintercept = seq(1970,2010,5), colour="#f0f0f0") +
  geom_hline(yintercept = 0) +
  geom_line(aes(group = COD, col = COD), alpha = 0.90) + 
  geom_text(data = labels_women, aes(x = year, y = age_COD_cont_yrs_mean_collapse, label = age_collapse2), hjust = "left", check_overlap = T) +
  ylab("Contribution to the black-white gap in life expectancy, y") + 
  xlab("Year") +
  theme_bw() +
  theme(legend.position = "bottom", 
        strip.background.x = element_rect(fill="white"), 
        strip.background.y = element_blank(), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_blank(), #element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank(),
        legend.background = element_rect(colour = 'grey', fill = 'white', linetype='solid'),
        panel.spacing = unit(1, "lines")) + 
  scale_color_colorblind() +
  guides(col = guide_legend(title = element_blank())) +
  scale_x_continuous(breaks = c(1969, 1980, 1990, 2000, 2013), 
                     labels = c(1969, 1980, 1990, 2000, 2013))

women.fig
```

**Figure 3: Contribution of age and cause categories to the black-white life expectancy gap in men, 1969-2013**

```{r, fig.height = 11, fig.width = 9}
male.fig <- ggplot(age_cod_results %>% 
         filter(state %in% included_states, sex == "Male") %>% 
         arrange(state, year), 
       aes(x = year, y = age_COD_cont_yrs_mean_collapse)) + 
  facet_grid(age_collapse2 ~ state) + 
  geom_vline(xintercept = seq(1970,2010,5), colour="#f0f0f0") +
  geom_hline(yintercept = 0) +
  geom_line(aes(group = COD, col = COD), alpha = 0.90) + 
  geom_text(data = labels_women, aes(x = year, y = age_COD_cont_yrs_mean_collapse, label = age_collapse2), hjust = "left", check_overlap = T) +
  ylab("Contribution to the black-white gap in life expectancy, y") + 
  xlab("Year") +
  theme_bw() +
  theme(legend.position = "bottom", 
        legend.background = element_rect(colour = 'grey', fill = 'white', linetype='solid'),
        strip.background.x = element_rect(fill="white"), 
        strip.background.y = element_blank(),
        strip.text.x = element_text(size = 12),
        strip.text.y = element_blank(),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank(),
        panel.spacing = unit(1, "lines")) + 
  scale_color_colorblind() +
  scale_y_continuous(limits = c(-1, 3)) +
  guides(col = guide_legend(title = element_blank())) +
  scale_x_continuous(breaks = c(1969, 1980, 1990, 2000, 2013), 
                     labels = c(1969, 1980, 1990, 2000, 2013))

male.fig
```

Save the first three figures.

```{r save-figures, echo=2}
ggsave(filename = here("Manuscript2", "Images", "figure1.jpeg"), 
       plot = le.fig, device = "jpeg", width = 6, height = 3, units = "in", dpi = 1200)
ggsave(filename = here("Manuscript2", "Images", "figure2.jpeg"),
       plot = women.fig, device = "jpeg", width = 8, height = 9, units = "in", dpi = 1200)
ggsave(filename = here("Manuscript2", "Images", "figure3.jpeg"), 
       plot = male.fig, device = "jpeg", width = 8, height = 9, units = "in", dpi = 1200)
```

Code to create Figure 4:

**Figure 4: ...**

*left panel*

```{r}

dat.2013 <- BlackWhite_results %>% filter(year == 2013 & state %in% c(included_states))

LE.2013 <- ggplot(dat.2013, aes(x = LE_white_mean, y = state)) + 
  facet_wrap(~sex, nrow = 2) +
  geom_segment(aes(xend = LE_black_mean, yend = state)) +
  geom_point(aes(col = "white")) +
  geom_point(aes(x = LE_black_mean, col = "black")) +
  scale_color_manual(values = c('#a6cee3','#1f78b4'), labels = c("Black", "White")) + 
  geom_text(aes(label = round(LE_white_mean, 1)), nudge_y = 0.25, size = 4) + 
  geom_text(aes(label = round(LE_black_mean, 1), x = LE_black_mean), nudge_y = -0.25, size = 4) +
  theme_minimal(base_size = 15) + xlab("") + ylab("") +
  theme(legend.position = c(0.9, 0.15), panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        legend.background = element_rect(colour = 'grey', fill = 'white', linetype='solid')) +
  ggtitle("(A) Gap in life expectancy, y") +
  scale_x_continuous(breaks =  seq(71, 83, 2), labels =  seq(71, 83, 2)) + 
  scale_y_discrete(limits = c("New York", "Illinois", "Georgia","California"))
```

*right panel*

```{r, fig.height = 9, fig.width = 8.5}
cod.2013 <- cod_decomp_results %>% filter(year == 2013 & state %in% c(included_states)) %>%
  select(-X.1, -X.2, -X.3, -X.4, -Census_Region, -Census_Division, -X)

cod.graph <- ggplot(data = cod.2013,
       aes(x=state, y = COD_cont_yrs_mean, fill = COD)) +
  geom_bar(stat = "identity", col = "white", width = 0.5) + 
  coord_flip() +
  theme_minimal(base_size = 15)  + #guides(fill=FALSE) + 
  geom_hline(yintercept = 0, col = "black") +
  theme(legend.title = element_blank(), #legend.position = "bottom",
        panel.background = element_rect(fill = "transparent", colour = NA), 
        plot.background = element_rect(fill = "transparent", colour = NA),
        #legend.background = element_rect(fill = "transparent", colour = NA), 
        panel.grid.minor = element_blank(),
        legend.background = element_rect(colour = 'grey', fill = 'white', linetype='solid')) +
  ylab("") +
  xlab("") +
  scale_y_continuous(breaks = c(-0.5, 0:7), labels = c(-0.5, 0:7)) +
  scale_fill_colorblind() +
  facet_wrap(~sex, nrow = 2) +
  ggtitle("(B) Cause-of-death contributions to the gap in life expectancy, y") + 
  scale_x_discrete(limits = c("New York", "Illinois", "Georgia","California"))

figure4 <- grid.arrange(LE.2013, cod.graph, ncol=1)

ggsave(filename = here("Manuscript2", "Images", "figure4.jpeg"), 
                       plot = figure4, device = "jpeg", width = 8.5, height = 9, units = "in", dpi = 1200)

```

**Men**
```{r, fig.height = 5, fig.width = 9, echo=2}
sub.males <- subset(age_cod_results, year==2013 & 
                      state %in% c("California", "Georgia", "Illinois", "New York") &
                      sex == "Male")

age.cod.males <- ggplot(data = sub.males,
       aes(x=age, y = age_COD_cont_yrs_mean, fill = COD)) +
  geom_bar(stat = "identity") + 
  coord_flip() +
  scale_fill_colorblind() +
  theme_minimal()  + 
  geom_hline(yintercept = 0, lwd = 1, col = "black") +
  theme(legend.title = element_blank(), legend.position = "bottom",
        panel.background = element_rect(fill = "transparent", colour = NA), 
        plot.background = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill = "transparent", colour = NA)) +
  ylab("Contribution to the life expectancy gap (years)") +
  xlab("") +
  facet_wrap(year ~ state)

age.cod.males
```

**Women**
```{r, fig.height = 5, fig.width = 9, echo=2}
sub.females <- subset(age_cod_results, year==2013 & 
                      state %in% c("California", "Georgia", "Illinois", "New York") &
                      sex == "Female")

age.cod.females <-    ggplot(data = sub.females,
                             aes(x=age, y = age_COD_cont_yrs_mean, fill = COD)) +
  geom_bar(stat = "identity") + 
  coord_flip() +
  scale_fill_colorblind() +
  theme_minimal()  + 
  geom_hline(yintercept = 0, lwd = 1, col = "black") +
  theme(legend.title = element_blank(), legend.position = "bottom",
        panel.background = element_rect(fill = "transparent", colour = NA), 
        plot.background = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill = "transparent", colour = NA)) +
  ylab("Contribution to the life expectancy gap (years)") +
  xlab("") +
  facet_wrap(year ~ state)

age.cod.females

ggsave(filename = here("Manuscript2", "supp_figures", "age-cod-males.jpeg"),
       plot = age.cod.males, device = "jpeg", width = 11, height = 8, units = "in")
ggsave(filename = here("Manuscript2", "supp_figures", "age-cod-females.jpeg"),
       plot = age.cod.females, device = "jpeg", width = 11, height = 8, units = "in")

```

```{r, echo=F, eval =F}
#Could print this very small above the legend of the previous plots:
sub <- BlackWhite_results %>% 
  filter(year == 2013, state %in% c("California", "Georgia", "Illinois", "New York")) %>% 
  select(state.map.order, sex, year, LE_white_mean, LE_black_mean, LE_wbgap_mean)

ggplot(sub %>% filter(sex == "Males"), aes(y = as.numeric(state.map.order), x = LE_white_mean)) + geom_point()
```

**Figure 5: second option: men**
```{r}

    temp <- data.frame(subset(cod_decomp_results, sex == "Male" & 
                                year == 2013 & state %in% c("California", "Georgia", "Illinois", "New York")))
    temp["state.reorder2"] <- reorder(temp$state, temp$LE_black_mean, max, na.rm = T)
    temp["state.reorder2.n"] <- as.numeric(temp[["state.reorder2"]])

ggplot(temp, aes(x = new.start, y = state.reorder2.n)) + 
  geom_rect(aes(xmin = new.start, 
                ymin = state.reorder2.n - 0.44, 
                ymax = state.reorder2.n + 0.44, 
                xmax = new.finish, fill = COD)) +
    geom_segment(aes(x = LE_black_mean, 
                   xend = LE_white_mean, 
                   y = state.reorder2.n, 
                   yend = state.reorder2.n), lwd = 1) + 
  scale_y_continuous(breaks = 1:length(levels(factor(temp$state.reorder2))), 
                     labels = levels(factor(temp$state.reorder2))) +
  theme_minimal() + xlab(" ") + 
  theme(legend.title = element_blank(),
        panel.background = element_rect(fill = "transparent", colour = NA), 
        plot.background = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill = "transparent", colour = NA)) +
  geom_segment(aes(x = LE_white_mean, 
                   xend = LE_white_mean, 
                   y = state.reorder2.n - 0.46, 
                   yend = state.reorder2.n + 0.46), lwd = 1) +
  geom_segment(aes(x = LE_black_mean, 
                   xend = LE_black_mean, 
                   y = state.reorder2.n - 0.46, 
                   yend = state.reorder2.n + 0.46), lty = 3, lwd = 1) +
  ylab("") + xlab("Life expectancy (years)")

```

**Figure 5: second option: women**
```{r}

temp <- data.frame(subset(cod_decomp_results, sex == "Female" & 
                            year == 2013 & state %in% c("California", "Georgia", "Illinois", "New York")))
temp["state.reorder2"] <- reorder(temp$state, temp$LE_black_mean, max, na.rm = T)
temp["state.reorder2.n"] <- as.numeric(temp[["state.reorder2"]])

ggplot(temp, aes(x = new.start, y = state.reorder2.n)) + 
  geom_rect(aes(xmin = new.start, 
                ymin = state.reorder2.n - 0.44, 
                ymax = state.reorder2.n + 0.44, 
                xmax = new.finish, fill = COD)) +
  geom_segment(aes(x = LE_black_mean, 
                   xend = LE_white_mean, 
                   y = state.reorder2.n, 
                   yend = state.reorder2.n), lwd = 1) + 
  scale_y_continuous(breaks = 1:length(levels(factor(temp$state.reorder2))), 
                     labels = levels(factor(temp$state.reorder2))) +
  theme_minimal() + xlab(" ") + 
  theme(legend.title = element_blank(),
        panel.background = element_rect(fill = "transparent", colour = NA), 
        plot.background = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill = "transparent", colour = NA)) +
  geom_segment(aes(x = LE_white_mean, 
                   xend = LE_white_mean, 
                   y = state.reorder2.n - 0.46, 
                   yend = state.reorder2.n + 0.46), lwd = 1) +
  geom_segment(aes(x = LE_black_mean, 
                   xend = LE_black_mean, 
                   y = state.reorder2.n - 0.46, 
                   yend = state.reorder2.n + 0.46), lty = 3, lwd = 1) +
  ylab("") + xlab("Life expectancy (years)")

```

# Appendix 1: Trends in age-specific mortality rate by sex, race, and cause of death

**Figure S15: Population sizes**

```{r, echo = F}
population.dat <- dat.aggregated %>% filter(age == "<1 year") %>%
  group_by(state, year, race) %>%
  summarise(population2 = sum(pop_across_age))
# 
# ggplot(dat.aggregated %>% filter(age == "<1 year" & state %in% included_states & race == "Black")) +
#   geom_line(aes(x = year, y = pop_across_age, col = state, linetype = race)) + 
#   facet_grid(~sex, scale="free_y") + 
#   #scale_y_continuous(breaks = seq(0, 15000000, by = 2000000)) +
#   ylab("Population") + xlab("Year") + theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

white.pop <- ggplot(population.dat %>% filter(state %in% included_states & race == "White")) +
  geom_vline(xintercept=seq(1970,2010,5), colour="#f0f0f0") +
  geom_line(aes(x = year, y = population2, col = state), size=2) + 
  scale_color_colorblind() +
  #scale_y_continuous(breaks = seq(0, 15000000, by = 2000000)) +
  ylab("Population") + xlab("Year") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(limits = c(0, 30000000), breaks = seq(0, 30000000, length.out = 4), 
                     labels = c("0", "10,000,000", "20,000,000","30,000,000")) +
  scale_x_continuous(breaks = seq(1969, 2013, 1), labels = c("", 1970, rep("", 9), 
                                                             1980, rep("", 9),
                                                             1990, rep("", 9),
                                                             2000, rep("", 9),
                                                             2010, rep("", 3))) +
  ggtitle("a) Whites") +
  theme_bw() +
  theme(axis.text.x = element_text(hjust = 1),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()))

black.pop  <-  ggplot(population.dat %>% filter(state %in% included_states & race == "Black")) +
  geom_vline(xintercept=seq(1970,2010,5), colour="#f0f0f0") +
  geom_line(aes(x = year, y = population2, col = state), size = 2) + 
  #scale_y_continuous(breaks = seq(0, 15000000, by = 2000000)) +
  scale_color_colorblind() +
  ylab("Population") + xlab("Year") +
  scale_y_continuous(limits = c(0, 4000000), breaks = seq(0, 4000000, length.out = 5), 
                     labels = c("0", "1,000,000", "2,000,000","3,000,000","4,000,000")) +
    scale_x_continuous(breaks = seq(1969, 2013, 1), labels = c("", 1970, rep("", 9), 
                                                             1980, rep("", 9),
                                                             1990, rep("", 9),
                                                             2000, rep("", 9),
                                                             2010, rep("", 3))) +
  ggtitle("b) Blacks") +
  theme_bw() + 
  theme(axis.text.x = element_text(hjust = 1),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()))

white.pop

black.pop

library(patchwork)

population_plot <- white.pop + black.pop + plot_layout(ncol = 2)

ggsave(filename = here("Manuscript2", "supp_figures", "population_plot.jpg"), 
       plot = population_plot, device = "jpeg", width = 8, height = 3, units = "in")
```

**Figure S1: Age-specific trends in CVD-related mortality in black and white men, 20 years and older, 1969-2013**

```{r, fig.height = 7.3, fig.width = 9, echo=2}
cvd_males <- ggplot(data = subset(dat.collapse.age, sex == "Male" & COD == "Cardiovascular" & 
       age_collapse %in% c("20-39 years","40-64 years", "65-84 years",   "85+ years")), 
       aes(x = year, y = crude_rate_age_collapse1)) + 
  geom_line(aes(col = race, lty = "impute 1")) + 
  geom_line(aes(y = crude_rate_age_collapse5, col = race, lty = "impute 5")) + 
  geom_line(aes(y = crude_rate_age_collapse9, col = race, lty = "impute 9")) + 
  facet_grid(age_collapse ~ state, scales = "free_y") +
  ylab("Mortality Rate (per 100,000)") + 
  xlab("Year") +
  ggtitle("CVD in males") +
  scale_color_manual(values = c('#a6cee3','#1f78b4')) +
  scale_x_continuous(breaks = seq(1969, 2013, 1), 
                     labels =  c("", 1970, rep("", 9), 
                                 "'80", rep("", 9),
                                 "'90", rep("", 9),
                                 "'00", rep("", 9),
                                 2010, rep("", 3))) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(hjust = 1),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank()))
```

**Figure S2: Age-specific trends in CVD-related mortality in black and white women, 20 years and older, 1969-2013**

```{r, fig.height = 7.3, fig.width = 9, echo=2}
cvd_females <- ggplot(data = subset(dat.collapse.age, sex == "Female" & COD == "Cardiovascular" & 
       age_collapse %in% c("20-39 years","40-64 years", "65-84 years",   "85+ years")), 
       aes(x = year, y = crude_rate_age_collapse1)) + 
  geom_line(aes(col = race, lty = "impute 1")) + 
  geom_line(aes(y = crude_rate_age_collapse5, col = race, lty = "impute 5")) + 
  geom_line(aes(y = crude_rate_age_collapse9, col = race, lty = "impute 9")) + 
  facet_grid(age_collapse ~ state, scales = "free_y") +
  ylab("Mortality Rate (per 100,000)") + 
  scale_color_manual(values = c('#a6cee3','#1f78b4')) +
  xlab("Year") +
  scale_x_continuous(breaks = seq(1969, 2013, 1), 
                     labels =  c("", 1970, rep("", 9), 
                                 "'80", rep("", 9),
                                 "'90", rep("", 9),
                                 "'00", rep("", 9),
                                 2010, rep("", 3))) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(hjust = 1),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank()))
```

Not shown: <1 year olds, and 1-19 year olds, because very few deaths in those age bands. Note that there appears to be a measurement error in the data for Illinois in 1972 and 1973 for both males and females showing a much higher rate of death.

**Figure S3: Age-specific trends in cancer-related mortality in black and white men, 20 years and older, 1969-2013**

```{r, fig.height = 7.3, fig.width = 9, echo=2}
cancer_males <- ggplot(data = subset(dat.collapse.age, sex == "Male" & COD == "Cancers" &
                     age_collapse %in% c("20-39 years","40-64 years", "65-84 years",   "85+ years")
                     ), 
       aes(x = year, y = crude_rate_age_collapse1)) + 
  geom_line(aes(col = race, lty = "impute 1")) + 
  geom_line(aes(y = crude_rate_age_collapse5, col = race, lty = "impute 5")) + 
  geom_line(aes(y = crude_rate_age_collapse9, col = race, lty = "impute 9")) + 
  facet_grid(age_collapse ~ state, scales = "free_y") +
  ylab("Mortality Rate (per 100,000)") + 
  xlab("Year") +
  scale_color_manual(values = c('#a6cee3','#1f78b4')) +
  scale_x_continuous(breaks = seq(1969, 2013, 1), 
                     labels =  c("", 1970, rep("", 9), 
                                 "'80", rep("", 9),
                                 "'90", rep("", 9),
                                 "'00", rep("", 9),
                                 2010, rep("", 3))) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(hjust = 1),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank()))
```

**Figure S4: Age-specific trends in cancer-related mortality in black and white women, 20 years and older, 1969-2013**

```{r, fig.height = 7.3, fig.width = 9, echo=2}
cancer_females <- ggplot(data = subset(dat.collapse.age, sex == "Female" & COD == "Cancers" & 
       age_collapse %in% c("20-39 years","40-64 years", "65-84 years",   "85+ years")
       ), 
       aes(x = year, y = crude_rate_age_collapse1)) + 
  geom_line(aes(col = race, lty = "impute 1")) + 
  geom_line(aes(y = crude_rate_age_collapse5, col = race, lty = "impute 5")) + 
  geom_line(aes(y = crude_rate_age_collapse9, col = race, lty = "impute 9")) + 
  facet_grid(age_collapse ~ state, scales = "free_y") +
  ylab("Mortality Rate (per 100,000)") + 
  xlab("Year") +
  scale_color_manual(values = c('#a6cee3','#1f78b4')) +
  scale_x_continuous(breaks = seq(1969, 2013, 1), 
                     labels =  c("", 1970, rep("", 9), 
                                 "'80", rep("", 9),
                                 "'90", rep("", 9),
                                 "'00", rep("", 9),
                                 2010, rep("", 3))) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(hjust = 1),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank()))
```

**Figure S5: Age-specific trends in mortality due to non communicable disease in black and white men, 20 years and older, 1969-2013**

```{r, fig.height = 7.3, fig.width = 9, echo=2}
nc_males <- ggplot(data = subset(dat.collapse.age, sex == "Male" & COD == "Non-communicable" &
                     age_collapse %in% c("20-39 years","40-64 years", "65-84 years",   "85+ years")
                     ), 
       aes(x = year, y = crude_rate_age_collapse1)) + 
  geom_line(aes(col = race, lty = "impute 1")) + 
  geom_line(aes(y = crude_rate_age_collapse5, col = race, lty = "impute 5")) + 
  geom_line(aes(y = crude_rate_age_collapse9, col = race, lty = "impute 9")) + 
  facet_grid(age_collapse ~ state, scales = "free_y") +
  ylab("Mortality Rate (per 100,000)") + 
  xlab("Year") +
  scale_color_manual(values = c('#a6cee3','#1f78b4')) +
  scale_x_continuous(breaks = seq(1969, 2013, 1), 
                     labels =  c("", 1970, rep("", 9), 
                                 "'80", rep("", 9),
                                 "'90", rep("", 9),
                                 "'00", rep("", 9),
                                 2010, rep("", 3))) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(hjust = 1),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank()))
```

**Figure S6: Age-specific trends in mortality due to non communicable disease in black and white women, 20 years and older, 1969-2013**

```{r, fig.height = 7.3, fig.width = 9, echo=2}
nc_female <- ggplot(data = subset(dat.collapse.age, sex == "Female" & COD == "Non-communicable" & 
       age_collapse %in% c("20-39 years","40-64 years", "65-84 years",   "85+ years")
       ), 
       aes(x = year, y = crude_rate_age_collapse1)) + 
  geom_line(aes(col = race, lty = "impute 1")) + 
  geom_line(aes(y = crude_rate_age_collapse5, col = race, lty = "impute 5")) + 
  geom_line(aes(y = crude_rate_age_collapse9, col = race, lty = "impute 9")) + 
  facet_grid(age_collapse ~ state, scales = "free_y") +
  ylab("Mortality Rate (per 100,000)") + 
  xlab("Year") +
  scale_color_manual(values = c('#a6cee3','#1f78b4')) +
  scale_x_continuous(breaks = seq(1969, 2013, 1), 
                     labels =  c("", 1970, rep("", 9), 
                                 "'80", rep("", 9),
                                 "'90", rep("", 9),
                                 "'00", rep("", 9),
                                 2010, rep("", 3))) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(hjust = 1),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank()))
```

**Figure S7: Age-specific trends in injury-related mortality in black and white men, all age groups, 1969-2013**

```{r, fig.height = 11, fig.width = 9, echo=2}
injury_males <- ggplot(data = subset(dat.collapse.age, sex == "Male" & COD == "Injuries"), 
       aes(x = year, y = crude_rate_age_collapse1)) + 
  geom_line(aes(col = race, lty = "impute 1")) + 
  geom_line(aes(y = crude_rate_age_collapse5, col = race, lty = "impute 5")) + 
  geom_line(aes(y = crude_rate_age_collapse9, col = race, lty = "impute 9")) + 
  facet_grid(age_collapse ~ state, scales = "free_y") +
  ylab("Mortality Rate (per 100,000)") + 
  xlab("Year") +
  scale_color_manual(values = c('#a6cee3','#1f78b4')) +
  scale_x_continuous(breaks = seq(1969, 2013, 1), 
                     labels =  c("", 1970, rep("", 9), 
                                 "'80", rep("", 9),
                                 "'90", rep("", 9),
                                 "'00", rep("", 9),
                                 2010, rep("", 3))) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(hjust = 1),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank()))
```

**Figure S8: Age-specific trends in injury-related mortality in black and white women, all age groups, 1969-2013**

```{r, fig.height = 11, fig.width = 9, echo=2}
inj_females <- ggplot(data = subset(dat.collapse.age, sex == "Female" & COD == "Injuries" #& 
       #age_collapse %in% c("20-39 years","40-64 years", "65-84 years",   "85+ years")
       ), 
       aes(x = year, y = crude_rate_age_collapse1)) + 
  geom_line(aes(col = race, lty = "impute 1")) + 
  geom_line(aes(y = crude_rate_age_collapse5, col = race, lty = "impute 5")) + 
  geom_line(aes(y = crude_rate_age_collapse9, col = race, lty = "impute 9")) + 
  facet_grid(age_collapse ~ state, scales = "free_y") +
  ylab("Mortality Rate (per 100,000)") + 
  xlab("Year") +
  scale_color_manual(values = c('#a6cee3','#1f78b4')) +
  scale_x_continuous(breaks = seq(1969, 2013, 1), 
                     labels =  c("", 1970, rep("", 9), 
                                 "'80", rep("", 9),
                                 "'90", rep("", 9),
                                 "'00", rep("", 9),
                                 2010, rep("", 3))) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(hjust = 1),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank()))
```

**Figure S9: Age-specific trends in communicable-related mortality in black and white men, all age groups, 1969-2013**

```{r, fig.height = 11, fig.width = 9, echo=2}
comm_males <- ggplot(data = subset(dat.collapse.age, sex == "Male" & COD == "Communicable"), 
       aes(x = year, y = crude_rate_age_collapse1)) + 
  geom_line(aes(col = race, lty = "impute 1")) + 
  geom_line(aes(y = crude_rate_age_collapse5, col = race, lty = "impute 5")) + 
  geom_line(aes(y = crude_rate_age_collapse9, col = race, lty = "impute 9")) + 
  facet_grid(age_collapse ~ state, scales = "free_y") +
  ylab("Mortality Rate (per 100,000)") + 
  xlab("Year") +  
  scale_color_manual(values = c('#a6cee3','#1f78b4')) +
  scale_x_continuous(breaks = seq(1969, 2013, 1), 
                     labels =  c("", 1970, rep("", 9), 
                                 "'80", rep("", 9),
                                 "'90", rep("", 9),
                                 "'00", rep("", 9),
                                 2010, rep("", 3))) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(hjust = 1),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank()))
```

**Figure S10: Age-specific trends in communicable-related mortality in black and white women, all age groups, 1969-2013**

```{r, fig.height = 11, fig.width = 9, echo=2}
comm_female <- ggplot(data = subset(dat.collapse.age, sex == "Female" & COD == "Communicable"), 
       aes(x = year, y = crude_rate_age_collapse1)) + 
  geom_line(aes(col = race, lty = "impute 1")) + 
  geom_line(aes(y = crude_rate_age_collapse5, col = race, lty = "impute 5")) + 
  geom_line(aes(y = crude_rate_age_collapse9, col = race, lty = "impute 9")) + 
  facet_grid(age_collapse ~ state, scales = "free_y") +
  ylab("Mortality Rate (per 100,000)") + 
  xlab("Year") +
  scale_color_manual(values = c('#a6cee3','#1f78b4')) +
  scale_x_continuous(breaks = seq(1969, 2013, 1), 
                     labels =  c("", 1970, rep("", 9), 
                                 "'80", rep("", 9),
                                 "'90", rep("", 9),
                                 "'00", rep("", 9),
                                 2010, rep("", 3))) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(hjust = 1),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank()))
```

**Figure S11: Age-specific trends in mortality due to all other causes in black and white men, all age groups, 1969-2013**

```{r, fig.height = 11, fig.width = 9, echo=2}
other_male <- ggplot(data = subset(dat.collapse.age, sex == "Male" & COD == "All other causes"), 
       aes(x = year, y = crude_rate_age_collapse1)) + 
  geom_line(aes(col = race, lty = "impute 1")) + 
  geom_line(aes(y = crude_rate_age_collapse5, col = race, lty = "impute 5")) + 
  geom_line(aes(y = crude_rate_age_collapse9, col = race, lty = "impute 9")) + 
  facet_grid(age_collapse ~ state, scales = "free_y") +
  ylab("Mortality Rate (per 100,000)") + 
  xlab("Year") +
  scale_color_manual(values = c('#a6cee3','#1f78b4')) +
  scale_x_continuous(breaks = seq(1969, 2013, 1), 
                     labels =  c("", 1970, rep("", 9), 
                                 "'80", rep("", 9),
                                 "'90", rep("", 9),
                                 "'00", rep("", 9),
                                 2010, rep("", 3))) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(hjust = 1),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank()))
```

**Figure S12: Age-specific trends in mortality  due to all other causes in black and white women, all age groups, 1969-2013**

```{r, fig.height = 11, fig.width = 9, echo=2}
other_female <- ggplot(data = subset(dat.collapse.age, sex == "Female" & COD == "All other causes"), 
       aes(x = year, y = crude_rate_age_collapse1)) + 
  geom_line(aes(col = race, lty = "impute 1")) + 
  geom_line(aes(y = crude_rate_age_collapse5, col = race, lty = "impute 5")) + 
  geom_line(aes(y = crude_rate_age_collapse9, col = race, lty = "impute 9")) + 
  facet_grid(age_collapse ~ state, scales = "free_y") +
  ylab("Mortality Rate (per 100,000)") + 
  xlab("Year") +
  scale_color_manual(values = c('#a6cee3','#1f78b4')) +
  scale_x_continuous(breaks = seq(1969, 2013, 1), 
                     labels =  c("", 1970, rep("", 9), 
                                 "'80", rep("", 9),
                                 "'90", rep("", 9),
                                 "'00", rep("", 9),
                                 2010, rep("", 3))) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(hjust = 1),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + 
  guides(col = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank()))
```

Save the supplementary figures files:

```{r save-supp-figures}

# women
ggsave(filename = here("Manuscript2", "supp_figures", "other_female.jpeg"), 
       plot = other_female, device = "jpeg", width = 7, height = 8, units = "in")
ggsave(filename = here("Manuscript2", "supp_figures", "cvd_female.jpeg"), 
       plot = cvd_females, device = "jpeg", width = 7, height = 6, units = "in")
ggsave(filename = here("Manuscript2", "supp_figures", "cancer_female.jpeg"), 
       plot = cancer_females, device = "jpeg", width = 7, height = 6, units = "in")
ggsave(filename = here("Manuscript2", "supp_figures", "nc_female.jpeg"), 
       plot = nc_males, device = "jpeg", width = 7, height = 6, units = "in")
ggsave(filename = here("Manuscript2", "supp_figures", "comm_female.jpeg"), 
       plot = comm_female, device = "jpeg", width = 7, height = 8, units = "in")
ggsave(filename = here("Manuscript2", "supp_figures", "inj_female.jpeg"), 
       plot = inj_females, device = "jpeg", width = 7, height = 8, units = "in")

# men
ggsave(filename = here("Manuscript2", "supp_figures", "other_male.jpeg"),
       plot = other_male, device = "jpeg", width = 7, height = 8, units = "in")
ggsave(filename = here("Manuscript2", "supp_figures", "cvd_male.jpeg"), 
       plot = cvd_males, device = "jpeg", width = 7, height = 6, units = "in")
ggsave(filename = here("Manuscript2", "supp_figures", "cancer_male.jpeg"), 
       plot = cancer_males, device = "jpeg", width = 7, height = 6, units = "in")
ggsave(filename = here("Manuscript2", "supp_figures", "nc_male.jpeg"), 
       plot = nc_males, device = "jpeg", width = 7, height = 6, units = "in")
ggsave(filename = here("Manuscript2", "supp_figures", "comm_male.jpeg"), 
       plot = comm_males, device = "jpeg", width = 7, height = 8, units = "in")
ggsave(filename = here("Manuscript2", "supp_figures", "inj_male.jpeg"), 
       plot = injury_males, device = "jpeg", width = 7, height = 8, units = "in")
```

```{r, fig.width = 10, fig.height = 5, eval = F, echo = F}
ca.1969 <- age_cod_results %>% filter(state == "California" & year == 1969 & sex == "Male") %>%
 plot_ly(x = ~COD, y = ~age_collapse, z = ~age_COD_cont_yrs_mean, type = "heatmap") %>% 
  layout(xaxis = list(title = ""), yaxis = list(title = ""), title = "California Males 1969", 
        margin = list(l = 100, b=100))

ca.2013 <- age_cod_results %>% filter(state == "California" & year == 2013 & sex == "Male") %>%
 plot_ly(x = ~COD, y = ~age_collapse, z = ~age_COD_cont_yrs_mean, type = "heatmap") %>% 
  layout(xaxis = list(title = ""), yaxis = list(title = ""), title = "California Males 2013", 
        margin = list(l = 100, b=100))

subplot(ca.1969, ca.2013) %>% layout(title = "California Males, 1969 vs. 2013")
```

```{r, fig.width = 10, fig.height = 5, eval = F, echo = F}
ca.1969 <- age_cod_results %>% filter(state == "California" & year == 1969 & sex == "Female") %>%
 plot_ly(x = ~COD, y = ~age_collapse, z = ~age_COD_cont_yrs_mean, type = "heatmap") %>% 
  layout(xaxis = list(title = ""), yaxis = list(title = ""), title = "California Females 1969", 
        margin = list(l = 100, b=100))

ca.2013 <- age_cod_results %>% filter(state == "California" & year == 2013 & sex == "Female") %>%
 plot_ly(x = ~COD, y = ~age_collapse, z = ~age_COD_cont_yrs_mean, type = "heatmap") %>% 
  layout(xaxis = list(title = ""), yaxis = list(title = ""), title = "California Females 2013", 
        margin = list(l = 100, b=100))

subplot(ca.1969, ca.2013) %>% layout(title = "California Females, 1969 vs. 2013")
```

```{r, fig.width = 10, fig.height = 5, eval = F, echo = F}
ny.1969 <- age_cod_results %>% filter(state == "New York" & year == 1969 & sex == "Male") %>%
 plot_ly(x = ~COD, y = ~age_collapse, z = ~age_COD_cont_yrs_mean, type = "heatmap") %>% 
  layout(xaxis = list(title = ""), yaxis = list(title = ""), title = "California Males 1969", 
        margin = list(l = 100, b=100))

ny.2013 <- age_cod_results %>% filter(state == "New York" & year == 2013 & sex == "Male") %>%
 plot_ly(x = ~COD, y = ~age_collapse, z = ~age_COD_cont_yrs_mean, type = "heatmap") %>% 
  layout(xaxis = list(title = ""), yaxis = list(title = ""), title = "California Males 2013", 
        margin = list(l = 100, b=100))

subplot(ny.1969, ny.2013) %>% layout(title = "New York Males, 1969 vs. 2013")
```

```{r, fig.width = 10, fig.height = 5, eval = F, echo = F}
ny.1969 <- age_cod_results %>% filter(state == "New York" & year == 1969 & sex == "Female") %>%
 plot_ly(x = ~COD, y = ~age_collapse, z = ~age_COD_cont_yrs_mean, type = "heatmap") %>% 
  layout(xaxis = list(title = ""), yaxis = list(title = ""), title = "California Males 1969", 
        margin = list(l = 100, b=100))

ny.2013 <- age_cod_results %>% filter(state == "New York" & year == 2013 & sex == "Female") %>%
 plot_ly(x = ~COD, y = ~age_collapse, z = ~age_COD_cont_yrs_mean, type = "heatmap") %>% 
  layout(xaxis = list(title = ""), yaxis = list(title = ""), title = "California Females 2013", 
        margin = list(l = 100, b=100))

subplot(ny.1969, ny.2013) %>% layout(title = "New York Females, 1969 vs. 2013")
```

```{r, fig.width = 10, fig.height = 5, eval = F, echo = F}
ny.1969 <- age_cod_results %>% filter(state == "New York" & year == 1969 & sex == "Female") %>%
 plot_ly(x = ~COD, y = ~age, z = ~age_COD_cont_yrs_mean, type = "heatmap") %>% 
  layout(xaxis = list(title = ""), yaxis = list(title = ""), title = "California Males 1969", 
        margin = list(l = 100, b=100))

ny.2013 <- age_cod_results %>% filter(state == "New York" & year == 2013 & sex == "Female") %>%
 plot_ly(x = ~COD, y = ~age, z = ~age_COD_cont_yrs_mean, type = "heatmap") %>% 
  layout(xaxis = list(title = ""), yaxis = list(title = ""), title = "California Females 2013", 
        margin = list(l = 100, b=100))

subplot(ny.1969, ny.2013) %>% layout(title = "New York Females, 1969 vs. 2013")
```

### 2019-07-18 Manuscript2/ageCOD_manuscript.Rmd

@@
   mutate(age_COD_cont_yrs_mean_collapse = sum(age_COD_cont_yrs_mean))
+
+men_cod_contributions <- age_cod_results %>% ungroup() %>% filter(sex == "Male", state %in% c("California", "New York", "Illinois", "Georgia"))
+
+women_cod_contributions <- age_cod_results %>% ungroup() %>% filter(sex == "Female", state %in% c("California", "New York", "Illinois", "Georgia"))
 ```
